### 2018-10-13 manuscript.rmd
added file stubs for R analyses

---
csl: plos.csl
output:
  github_document: default
  html_document: default
  pdf_document: default
bibliography: manuscript.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	echo=FALSE,
	dpi=300,
	cache=TRUE
	)
```

```{r preliminaries}
library( tidyverse )
library( magrittr )
```

# Rooting the animal tree of life

## Introduction

There has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].

## Published analyses

## References

### 2018-10-15 manuscript.rmd
minor table updates

@@
 
+```{r}
+cteno_color = "#F8766D"
+poro_color = "#00BA38"
+unresolved_color = "#6F94CD"
+
+papers = read_tsv( "tables/papers.tsv" )
+datasets = read_tsv( "tables/datasets.tsv" )
+analyses = read_tsv( "tables/analyses.tsv" )
+
+bootstrap_threshold = 90
+posterior_prob_threshold = 95
+
+analyses$Result = "Unresolved"
+analyses$Result[ ( analyses$Inference == "Bayesian" ) & (analyses$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
+analyses$Result[ ( analyses$Inference == "Bayesian" ) & (analyses$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
+analyses$Result = factor( analyses$Result )
+
+```
+
 # Rooting the animal tree of life

### 2018-10-15 scripts/iqtree.sh
iqtree script

@@ -1,0 +1,17 @@
+#!/bin/bash
+#SBATCH -J iqtree
+#SBATCH -p general   
+#SBATCH -C avx2
+#SBATCH -N 1
+#SBATCH -c 20
+#SBATCH -t 7-00:00:00
+#SBATCH --mem 100G
+#SBATCH --mail-type=END,FAIL
+#SBATCH --mail-user=yuanning.li@yale.edu
+#SBATCH --output=iqtree.%j.out
+#SBATCH --error=iqtree.%j.err
+
+for FILENAME in *.nex
+do
+iqtree -s $FILENAME -nt 20 -bb 1000 -madd GTR+C60+F+R9,LG+C60+F+R9,LG+FO*H4,GTR+FO*H4, -pre $FILENAME.iqtree 
+done

### 2018-10-15 manuscript.rmd
added landscape plot

--- 2018-10-13 manuscript.rmd
@@
 
+```{r}
+cteno_color = "#F8766D"
+poro_color = "#00BA38"
+unresolved_color = "#6F94CD"
+
+papers = read_tsv( "tables/papers.tsv" )
+datasets = read_tsv( "tables/datasets.tsv" )
+analyses = 
+	read_tsv( "tables/analyses.tsv", col_types = 'icccccccccclnnnnccc' ) %>%
+	mutate( taxon_sampling = factor( taxon_sampling, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
+
+bootstrap_threshold = 90
+posterior_prob_threshold = 95
+
+analyses$result = "Unresolved"
+analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
+analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
+analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_porifera_sister >= bootstrap_threshold) ] = "Porifera-sister"
+analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
+analyses$result = factor( analyses$result )
+
+analyses$model_summary = analyses$model_rate_matrix
+long_summary = paste(analyses$model_rate_matrix, analyses$model_equilibrium, sep="+")
+analyses$model_summary[ ! is.na(analyses$model_equilibrium) ] = long_summary[ ! is.na(analyses$model_equilibrium) ]
+analyses$model_summary[ grepl("WAG", analyses$model_summary) ] = "WAG" # Simplify WAG
+analyses$model_summary = factor( analyses$model_summary, levels=c("GTR+CAT", "F81+CAT", "LG", "WAG") )
+
+```
+
 # Rooting the animal tree of life
@@
 
+```{r support}
+
+analyses %>% 
+	ggplot( aes(x=model_summary, y=taxon_sampling, col=result) ) +
+		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		theme_classic()
+```
+
 ## References

### 2018-10-16 manuscript.rmd
updated plot with new data

@@
 analyses = 
-	read_tsv( "tables/analyses.tsv", col_types = 'icccccccccclnnnnccc' ) %>%
+	read_tsv( "tables/analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
 	mutate( taxon_sampling = factor( taxon_sampling, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
@@
 
+A total of `r nrow(analyses)` analyses were transcribed from the literature.
+
 ## References

### 2018-10-16 manuscript.rmd
started on text, added Pick

@@
 
-## Published analyses
+Defining Porifera-sister and Ctenophora-sister... XXX Explain that it is about support for calde of all animals except Ctenophora or Porifera...
+
+## Variation across studies
+
+### Models of molecular evolution
+
+Exchangability... F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
+
+Equilibrium heterogenaity across sites
+
+Rate heterogeneity across sites
+
+The models that are applied in practice area heavily influenced by engineering and computational costs , as well as convention.
+For example, on the questions considered here F81 and GTR exchangability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
+LG and WAG exchangability matrices have only been used with site homogenous estimates of equilibrium frequency.
+This is further confused by the abreviations that are used for models.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangability matrix.
+CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogenous equilibrium frequency estimation.
+One could, though, run a WAG+CAT model.
+
+To avoid confusion on this point, we always specify the exchangability matrix first, followed by modifiers that describe accomodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+If there are no modifiers, then it is implied that site homogenous models are used.
+
+### Gene sampling
+
+### Outgroup taxon sampling
+
+Choanimalia, Holozoa, Opisthokonta
+
+### Ingroup taxon sampling
+
+Sensitivity to ingroup sampling has recieved less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.
+
+## Overview of published analyses
 
@@
 
+## Details of published analyses
+
+### Dunn *et al.* 2008
+
+Dunn *et al.* [@Dunn:2008ky] added Expressed Sequence Tag (EST) data for 29 animals.
+It was the first phylogenomic analysis that included ctenophores, and therefore that could test the relationships of both Ctenophora and Porifera to the rest of animals.
+It was the first phylogenetic analysis to recover Ctenophora as the sister group to all other animals.
+
+The data matrix was constructed using a semi-automated approach.
+Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplciations.
+Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors).
+The remaining gene trees with more than one sequence for any taxon were then manually inspected.
+If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded.
+If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded.
+Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix.
+Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships.
+Unstable taxa were identified with leaf stability indeces [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix.
+This resulted in the 64-taxon matrix that is the focus of most of their analyses.
+Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+> The placement of ctenophores (comb jellies) as the sister group to all other sampled metazoans is strongly supported in all our analyses.
+This result, which has not been postulated before, should be viewed as provisional until more data are considered from placozoans and additional sponges.
+
+Note that there was, in fact, an exception to strong support.
+An analysis of the 40 ribosomal proteins in the matrix recovered Ctenophora-sister with only 69% support.
+This study did not include *Trichoplax*.
+
+### Philippe *et al.* 2009
+
+Philippe *et al.* 2009 [Philippe:2009hh]...
+
+### Hejnol et al. 2009
+
+### Pick *et al.* 2010
+
+Pick *et al.* [@Pick:2010eb] sought to test wether Ctenophora-sister was an artefact of insufficient taxon sampling.
+They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky].
+The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*.
+They firther modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
+They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
+
+All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a bayesian framework and with bootstrapping.
+All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
+Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
+With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
+This is an inveresting case where increased outgroup sampling leads to increased support for Porifera-sister.
+
+The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected".
+Thout the posterior probabilities supporting Porifera-sister are not strong, they conclude:
+
+> Results of our analyses indicate that sponges are the sister group to the remaining Metazoa, and Placozoa are sister group to the Bilateria
+
+They also investigated saturation, and coclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
+
 ## References

### 2018-10-16 manuscript.rmd
more on models

@@
 
-Exchangability... F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
+Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
+The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
 
-Equilibrium heterogenaity across sites
+The Exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangability matrices have been used in the studies under consideration here include:
 
-Rate heterogeneity across sites
+- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
+The F81 matrix is also sometimes referred to as the Poisson matrix.
+It has no free parameters to estimate since all off-diagonal elements are set to 1.
 
-The models that are applied in practice area heavily influenced by engineering and computational costs , as well as convention.
+- WAG [@Whelan:2001ds] is an emprically derived exchangability matrix based on a dataset of 182 globular protein families.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- LG [@Le:2008fp], like WAG, is an emprically derived exchangability matrix.
+It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- GTR, the General Time Reversible exchangability matrix, has free parameters for all off-diagonal elements that describe the exchangability of different amino acids.
+It is consrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
+This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
+This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
+
+While the exchancheability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+There are couple approaches that have been used in the studies considered here:
+
+- Site homogenous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
+
+- Gamma rate heterogeneity.
+Each site is assigned to a different rate class with its own rate value.
+This accommodates different rates of evolution across different sites.
+Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
+
+The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
+
+- Empirical site homogenous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogenously to all sites in the matrix.
+
+- Estimated site homogenous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
+
+- CAT site heterogeneous [@Lartillot:2004dq].
+Each site is assigned to a class with its own equilibrium frequencies.
+The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
+
+Models can be assembled by selecting different options for all these different components.
+The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
 For example, on the questions considered here F81 and GTR exchangability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.

### 2018-10-16 manuscript.rmd
added figure 1, made palette colorblind friendly

@@
 ```{r}
-cteno_color = "#F8766D"
-poro_color = "#00BA38"
-unresolved_color = "#6F94CD"
+
+# A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+result_colors = c( cbPalette[2], cbPalette[4], cbPalette[1] )
+names( result_colors ) = c( "Ctenophora-sister", "Porifera-sister", "Unresolved" )
 
@@
 
+Casey W. Dunn^1^*, Benjamin Evans^2^, and Yuanning Li^1^
+
+^1^Department of Ecology and Evolutionary Biology, Yale University
+
+^2^Yale Center for Research Computing, Yale University
+
+\* Corresponding author, casey.dunn@yale.edu
+
+## Abstract
+
 ## Introduction
@@
 Defining Porifera-sister and Ctenophora-sister... XXX Explain that it is about support for calde of all animals except Ctenophora or Porifera...
+
+![](figures/Figure_overview.png)
+**Fig XXOverview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
+(B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
+Testing these hypotheses requires evaluating the support for each of these alternative nodes.
+(C) A ctenophore.
+(D) A sponge.
 
@@
 
+### Alignment overlap
+
+### Support for Porifera-sister and Ctenophora-sister
+
 ```{r support}
@@
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		scale_colour_manual(values = result_colors) +
 		theme_classic()
@@
 A total of `r nrow(analyses)` analyses were transcribed from the literature.
+
+## New analyses of published matrices
+
+One of the challenges of interpretting support for the placement of the animal root across studies is that different programs, sofware versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
+Here we reanalyze the primary matrices from each study under consistent conditions with iqtree.
+We selected this tool because it has greater model flexibility than other tools and is very fast.
+
+We
+
+### Comparison of iqtree and phylobayes results
+
+## New analyses of new matrices
+
+## The current state of understanding
+
+## Next steps
+
+## Conclusion
+
+## Methods
+
+All files associated with this analysis are available at https://github.com/caseywdunn/animal_root
+
+### Data wrangling
+
+### Matrix comparison and annotation
+
+### Phylogenetic analyses
+
+## Ackowledgements
+
+Author contributions
+
+\pagebreak
+
+## Supplemental Information
 

### 2018-10-16 manuscript.rmd
more outline and stubs

@@
 
-There has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-
-Defining Porifera-sister and Ctenophora-sister... XXX Explain that it is about support for calde of all animals except Ctenophora or Porifera...
+In the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses.
+Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
+In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
+The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
+Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither.
+As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
+Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
+Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
+We hope that this provides an integrative overview of the challenge and provides direction for future studies.
+We also hope that the work we have done here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
 
@@
 Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
+The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
 
-The Exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangability matrices have been used in the studies under consideration here include:
+The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include:
 
@@
 
-- WAG [@Whelan:2001ds] is an emprically derived exchangability matrix based on a dataset of 182 globular protein families.
+- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
 It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
 
-- LG [@Le:2008fp], like WAG, is an emprically derived exchangability matrix.
+- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
 It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
@@
 
-- GTR, the General Time Reversible exchangability matrix, has free parameters for all off-diagonal elements that describe the exchangability of different amino acids.
-It is consrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
+- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
+It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
 This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
@@
 
-While the exchancheability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
 There are couple approaches that have been used in the studies considered here:
 
-- Site homogenous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
+- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
 
@@
 
-- Empirical site homogenous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogenously to all sites in the matrix.
+- Empirical site homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
 
-- Estimated site homogenous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
+- Estimated site homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
 
@@
 The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here F81 and GTR exchangability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
-LG and WAG exchangability matrices have only been used with site homogenous estimates of equilibrium frequency.
-This is further confused by the abreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangability matrix.
-CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogenous equilibrium frequency estimation.
+For example, on the questions considered here F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
+LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
+This is further confused by the abbreviations that are used for models.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix.
+CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
 One could, though, run a WAG+CAT model.
 
-To avoid confusion on this point, we always specify the exchangability matrix first, followed by modifiers that describe accomodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
-If there are no modifiers, then it is implied that site homogenous models are used.
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+If there are no modifiers, then it is implied that site homogeneous models are used.
 
@@
 
+XXX
+
+**Fig XXOutgroup.** The animals and their outgroups, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
+
 Choanimalia, Holozoa, Opisthokonta
@@
 
-Sensitivity to ingroup sampling has recieved less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.
+Sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.
 
@@
 
+### Alignment composition
+
+XXX
+
+**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by taxon sampling.
+Horizontal size is proportional to the number of genes (XXOr should it be sites?) sampled, vertical size to the number of taxa sampled.
+
+XXX
+
+**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc).
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+
 ### Alignment overlap
+
+XXX
+
+**Fig XXAlignment overlap.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
 
@@
 
-One of the challenges of interpretting support for the placement of the animal root across studies is that different programs, sofware versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
+One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
 Here we reanalyze the primary matrices from each study under consistent conditions with iqtree.
@@
 
-We
+We first tested a variety of models for each matrix, and inferred support under the selected model. We then analyzed every matrix under a panel of standard models, including XXX.
 
@@
 
+Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister.
+This has been addressed with CAT models.
+iqtree provides a new family of C models that also address site heterogeneity.
+Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
+This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
+
 ## New analyses of new matrices
+
+Based on the variation across analyses, we constructed new matrices with altered taxon and gene sampling to test specific hypotheses about differences in support.
 
@@
 
-Author contributions
+## Author contributions
 
@@
 
-## Supplemental Information
+# Supplemental Information
 
@@
 The data matrix was constructed using a semi-automated approach.
-Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplciations.
+Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications.
 Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors).
@@
 Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships.
-Unstable taxa were identified with leaf stability indeces [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix.
+Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix.
 This resulted in the 64-taxon matrix that is the focus of most of their analyses.
@@
 
-Pick *et al.* [@Pick:2010eb] sought to test wether Ctenophora-sister was an artefact of insufficient taxon sampling.
+Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling.
 They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky].
 The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*.
-They firther modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
+They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
 They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
 
-All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a bayesian framework and with bootstrapping.
+All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping.
 All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
@@
 With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
-This is an inveresting case where increased outgroup sampling leads to increased support for Porifera-sister.
+This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.
 
 The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected".
-Thout the posterior probabilities supporting Porifera-sister are not strong, they conclude:
+Although the posterior probabilities supporting Porifera-sister are not strong, they conclude:
 
@@
 
-They also investigated saturation, and coclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
 Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.

### 2018-10-18 blast/gen_fastas.sh
adding standardized, pre-BLAST phylip alignments and nexus partitions.

@@ -1,0 +1,19 @@
+echo phylips/Borowiec2015_Best108.phy; ../scripts/phylippart2multifasta.py -p phylips/Borowiec2015_Best108 -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Borowiec2015_Total1080.phy; ../scripts/phylippart2multifasta.py -p phylips/Borowiec2015_Total1080 -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Chang2015_Chang2015.phy; ../scripts/phylippart2multifasta.py -p phylips/Chang2015_Chang2015 -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Dunn2008_Dunn2008.phy; ../scripts/phylippart2multifasta.py -p phylips/Dunn2008_Dunn2008 -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Hejnol2009_Hejnol2009.phy; ../scripts/phylippart2multifasta.py -p phylips/Hejnol2009_Hejnol2009 -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Moroz2014_ED3a.phy; ../scripts/phylippart2multifasta.py -p phylips/Moroz2014_ED3a -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Moroz2014_ED3d.phy; ../scripts/phylippart2multifasta.py -p phylips/Moroz2014_ED3d -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Nosenko2013_nonribosomal_9187_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_nonribosomal_9187_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Nosenko2013_ribosomal_11057_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_ribosomal_11057_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
+echo phylips/Nosenko2013_ribosomal_14615_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_ribosomal_14615_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
[…]

### 2018-10-18 blast/gen_phylips.sh

@@ -9,6 +9,10 @@
 ../scripts/prepare_phylip.py -i ../considered_data/Nosenko2013/ribosomal_11057_smatrix.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
 ../scripts/prepare_phylip.py -i ../considered_data/Nosenko2013/ribosomal_14615_smatrix.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
 ../scripts/prepare_phylip.py -i ../considered_data/Philippe2009/Philippe2009.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Ryan2013/genome.opisthokonta.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
+../scripts/prepare_phylip.py -i ../considered_data/Ryan2013/est.opisthokonta.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
 ../scripts/prepare_phylip.py -i ../considered_data/Simion2017/supermatrix_97sp_401632pos_1719genes.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
 ../scripts/prepare_phylip.py -i ../considered_data/Whelan2015/Datset20_Uncertain_LBAgenesHeteroRemoved_LBAtaxa.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
+../scripts/prepare_phylip.py -i ../considered_data/Whelan2015/Dataset10_CertainPruned_LBAtaxa_LBAandHeteroGenesPruned.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
+../scripts/prepare_phylip.py -i ../considered_data/Whelan2015/Dataset1_FullData.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
+../scripts/prepare_phylip.py -i ../considered_data/Whelan2017/Metazoa_Choano_RCFV_strict.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
+../scripts/prepare_phylip.py -i ../considered_data/Whelan2017/Metazoa_full.phy -o phylips -t ../taxonomy_info/taxon_table.tsv

### 2018-10-18 manuscript.rmd
parsing partitions

@@
 library( magrittr )
+library(viridis)
+
+source( "functions.R" )
 ```
@@
 
+```{r taxonomy}
+
+clades = c( "Fungi", "Ichthyosporea", "Filasterea", "Choanoflagellida", "Ctenophora", "Porifera", "Placozoa", "Bilateria", "Cnidaria" )
+
+clade_colors = viridis( length(clades) )
+names( clade_colors ) = clades
+
+taxa = 
+	read_tsv("taxonomy_info/taxon_table.tsv") %>% 
+	distinct( relabelled_name, clade_assignment, ncbi_tax_id ) %>%
+	dplyr::rename( taxon=relabelled_name, clade=clade_assignment )
+
+```
+
 XXX
@@
 ## The current state of understanding
+
+### Interpretting variation support
+
+External criteria, eg posterior predictive scores, model fit etc
 

### 2018-10-18 scripts/phylippart2multifasta.py
fixed some partition files

@@ -7,12 +7,13 @@ from collections import defaultdict
 from os.path import  abspath, basename, dirname, join, splitext
 from Bio import AlignIO
 from Bio import SeqIO
+from Bio import Alphabet
 
 def get_parts(part_file_name):
     parts = []
     partnames = defaultdict(lambda: 0)
     for line in open(part_file_name):
-        match = re.match(r"CHARSET\W+([\w\.\-]+)\W+=\W+(\d+)\W*-\W*(\d+)", line)
+        match = re.match(r"CHARSET\W+([\w\.\-]+)\W*=\W*(\d+)\W*-\W*(\d+)", line, re.IGNORECASE)
         if match is not None:
             part_name, part_start, part_end = match.groups()
             # number non-unique partition names
@@ -62,12 +63,17 @@ manuscript, dataset  = basename(prefix).split('_', 1)
 
 name2taxid = read_taxon_table(args.taxon_table)
 parts = get_parts(part_file_name)
+gapped_protein = Alphabet.Gapped(Alphabet.IUPAC.protein, '-')
 
 alignment = AlignIO.read(phylip_file_name, 'phylip-relaxed')
-fasta_out = open(join(out_dir, prefix)+'.fa',  'w')
+fasta_out = open(join(out_dir, '{}_{}.fa'.format(manuscript, dataset)),  'w')
 for part_name, start, stop in parts:
-    for taxon in alignment:
+    for i, taxon in enumerate(alignment):
         tmp_seq = taxon[start:stop]
-        tmp_seq.id = ':'.join([manuscript, dataset, taxon.id, name2taxid[taxon.id], part_name])
-        SeqIO.write(tmp_seq, fasta_out, 'fasta')
+        tmp_seq.seq.alphabet = gapped_protein
+        old_seq = tmp_seq.seq
+        tmp_seq.seq = tmp_seq.seq.ungap()
+        if len(tmp_seq) > 1:
+            tmp_seq.id = ':'.join([manuscript, dataset, taxon.id, name2taxid[taxon.id], part_name])
+            SeqIO.write(tmp_seq, fasta_out, 'fasta')
 fasta_out.close()

### 2018-10-18 scripts/prepare_phylip.py

@@ -57,6 +57,7 @@ def process_phylip(manuscript_name, names_map, in_filename, out_phylip_fn, out_p
     alignment = AlignIO.read(in_filename, 'phylip-relaxed')
     for taxon in alignment:
         if taxon.id in names_map[manuscript_name] and names_map[manuscript_name][taxon.id] != 'na':
+                # print(taxon.id, ": ", names_map[manuscript_name][taxon.id])
                 taxon.id = names_map[manuscript_name][taxon.id]
     AlignIO.write(alignment, out_phylip_fn, 'phylip-relaxed')
 

### 2018-10-18 functions.R
nave partitioned alignment s4 object

@@ -7,44 +7,53 @@ library(tidyverse)
 setClass(
 	Class = "PartitionedMultipleAlignment",
 	representation = representation(
-		partitions = "data.frame"
+		partitions = "data.frame",
+		taxon_map = "vector",
+		partition_map = "vector"
 	),
 	contains = "MultipleAlignment"
 )
 
 #' Construct a PartitionedMultipleAlignment from an alignment file and a partition file
 #'
-#' @param data_list A list containing the expression data
-#' @return An Expression object
+#' @param alignment_file Path to the gene alignment
+#' @param partition_file Path to the partition file in nexus format
+#' @param taxon_map A dataframe where the first column corresponds to sequence names in the alignment and the second column to clade names for the sequences
+#' @param partition_map A dataframe where the first column corresponds to partition names and the second column to names for groups of partitions
+#' @return A PartitionedMultipleAlignment object
 #' @export
-PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, alignment_format="phylip" ) {
+PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, taxon_map=NULL, partition_map=NULL, alignment_format="phylip" ) {
 	object = readAAMultipleAlignment( filepath = alignment_file, format=alignment_format )
 	class(object) = "PartitionedMultipleAlignment"
 	
-	if( ! is.null(partition_file) ){
+	
 		
-		conn = file( partition_file, open="r")
-		lines = readLines( conn )
-		lines = lines[ grepl("CHARSET", lines) ]
-		
-		partitions = 
-			lapply( 
-				lines, 
-				function( line ){
-					fields = str_match(line, "CHARSET (\\w+)\\s*=\\s*(\\d+)\\s*-\\s*(\\d+)")
-					D = data.frame( 
-						partition=fields[2], 
-						start=as.integer(fields[3]), 
-						stop=as.integer(fields[4]),
-						stringsAsFactors = FALSE
-						)
-					return( D )
-				}
-			) %>%
-			bind_rows()
-		
-		object@partitions = partitions
-		
+	conn = file( partition_file, open="r")
+	lines = readLines( conn )
+	lines = lines[ grepl("CHARSET", lines) ]
+	close( conn )
+	
+	partitions = 
+		lapply( 
[…]

### 2018-10-18 manuscript.rmd
nave partitioned alignment s4 object
calculates clade rectangles and plots them, native ggplot extension not working

@@
 	distinct( relabelled_name, clade_assignment, ncbi_tax_id ) %>%
-	dplyr::rename( taxon=relabelled_name, clade=clade_assignment )
+	dplyr::rename( taxon=relabelled_name, clade=clade_assignment ) %>%
+	mutate( clade = factor( clade, levels=clades )  )
 
+```
+
+```{r}
+
+a_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.phy"
+p_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.nex"
+msa = PartitionedMultipleAlignment( a_file, p_file, taxa )
+
+D=clade_rects(msa)
+
+ggplot() + 
+	scale_x_continuous(name="x") + 
+	scale_y_continuous(name="y") +
+	geom_rect(data=D, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
+	scale_fill_discrete(guide = guide_legend(reverse=TRUE))
 ```

### 2018-10-18 functions.R

@@ -1,6 +1,11 @@
 library(Biostrings)
 library(tidyverse)
 
+
+#######################################
+## Classes
+
+
 #' An S4 class to store multiple sequence alignments with partition data
 #' 
 #' @slot partitions  The partitions
@@ -58,6 +63,62 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 	
 	object
 }
+
+
+#####################################################
+## GGPlot extension of multiple sequence alignments
+
+# Builds on https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html
+
+
+#' Calculates rectangles that represent clades in an alignment
+#'
+#' @param data A PartitionedMultipleAlignment object
+#' @return A dataframe
+#' @export
+clade_rects = function( data ){
+	n_genes = nrow(data@partitions)
+	clade_table = table( data@taxon_map ) %>% as.vector()
+	clade_sums = as.vector(clade_table)
+	clade_cumsums = cumsum(clade_sums)
+	
+	clades = factor( levels( data@taxon_map ), levels( data@taxon_map ))
+	
+	
+	# Create a dataframe where each row is a clade and the columns describe how to draw the rectangle for the clade
+	D = data.frame( clade=clades )
+	
+	D$xmin = rep( 0, length(clades) )
+	D$xmax = rep( n_genes, length(clades) )
+	D$ymin = c(0, clade_cumsums[1:length(clade_cumsums)-1])
+	D$ymax = clade_cumsums
+	
+	D
+}
+
+
+StatClades = 
+	ggproto(
+		"StatClades", 
+		Stat,
+		compute_group = 
+			function(data, scales) {
+				clade_rects( data )
+			},
+		required_aes = c( "x", "y" )
+)
+
+stat_clades = function(mapping = NULL, data = NULL, geom = "rect",
+											 position = "identity", na.rm = FALSE, show.legend = NA, 
+											 inherit.aes = TRUE, ...) {
+	layer(
+		stat = StatClades, data = data, mapping = mapping, geom = geom, 
+		position = position, show.legend = show.legend, inherit.aes = inherit.aes,
+		params = list(na.rm = na.rm, ...)
+	)
+}
+
+
 
 
 

### 2018-10-18 manuscript.rmd
changed palette, color blind friendly but hard to read

@@
 	geom_rect(data=D, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
-	scale_fill_discrete(guide = guide_legend(reverse=TRUE))
+	scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
+	#scale_fill_discrete(guide = guide_legend(reverse=TRUE))
 ```

### 2018-10-18 scripts/blast_results_pandas.py
post-processed blast results, output in  has corresponding partitions and species across blast results.

@@ -1,0 +1,32 @@
+#!/usr/bin/env python
+from glob import glob
+import pandas as pd
+
+colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore']
+qseq_colnames = ['mscript1', 'dataset1', 'taxon1', 'taxid1', 'part1']
+sseq_colnames = ['mscript2', 'dataset2', 'taxon2', 'taxid2', 'part2']
+pident_cutoff = 95.0
+
+for i, infile in enumerate(sorted(glob('blast/results/*_permissive.txt'))):
+    # first file
+    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
+    df_tmp = df_tmp[ df_tmp['pident']>pident_cutoff ]
+
+    qseq = df_tmp['qseqid'].str.split(':', expand=True)
+    qseq.columns = qseq_colnames
+    sseq = df_tmp['sseqid'].str.split(':', expand=True)
+    sseq.columns = sseq_colnames
+
+    df_tmp = pd.concat((qseq, sseq, df_tmp), axis=1)
+    if i ==0:
+        df = df_tmp
+    else:
+        df = pd.concat((df, df_tmp))
+
+parts_graph = df[ ['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2'] ].drop_duplicates()
+parts_graph.to_csv('blast/graphs/partitions_graph.tsv', sep='\t', index=False)
+taxon_graph = df[ ['mscript1', 'dataset1', 'taxon1', 'mscript2', 'dataset2', 'taxon2'] ].drop_duplicates()
+taxon_graph.to_csv('blast/graphs/taxon_graph.tsv', sep='\t', index=False)
+
+cross_mscript = df[(df['taxon1'] == df['taxon2']) & (df['mscript1'] != df['mscript2'])]
+cross_mscript.to_csv('blast/graphs/cross_mscript_taxa.tsv', sep='\t', index=False)

### 2018-10-18 scripts/blast_results_pandas.py
Updated graphs tables to have cutoffs: >99% pident and evalue < 1e-25. Added count columns for each row.

@@ -6,11 +6,12 @@ colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qsta
 qseq_colnames = ['mscript1', 'dataset1', 'taxon1', 'taxid1', 'part1']
 sseq_colnames = ['mscript2', 'dataset2', 'taxon2', 'taxid2', 'part2']
 pident_cutoff = 95.0
+evalue_cutoff = 1e-25
 
 for i, infile in enumerate(sorted(glob('blast/results/*_permissive.txt'))):
     # first file
     df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
-    df_tmp = df_tmp[ df_tmp['pident']>pident_cutoff ]
+    df_tmp = df_tmp[ (df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff) ]
 
     qseq = df_tmp['qseqid'].str.split(':', expand=True)
     qseq.columns = qseq_colnames
@@ -23,9 +24,9 @@ for i, infile in enumerate(sorted(glob('blast/results/*_permissive.txt'))):
     else:
         df = pd.concat((df, df_tmp))
 
-parts_graph = df[ ['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2'] ].drop_duplicates()
+parts_graph = df.groupby(['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2']).size().reset_index(name='count')
 parts_graph.to_csv('blast/graphs/partitions_graph.tsv', sep='\t', index=False)
-taxon_graph = df[ ['mscript1', 'dataset1', 'taxon1', 'mscript2', 'dataset2', 'taxon2'] ].drop_duplicates()
+taxon_graph = df.groupby(['mscript1', 'dataset1', 'taxon1', 'mscript2', 'dataset2', 'taxon2']).size().reset_index(name='count')
 taxon_graph.to_csv('blast/graphs/taxon_graph.tsv', sep='\t', index=False)
 
 cross_mscript = df[(df['taxon1'] == df['taxon2']) & (df['mscript1'] != df['mscript2'])]

### 2018-10-19 manuscript.rmd
checking taxon name reconciliation

--- 2018-10-18 manuscript.rmd
@@
 	geom_rect(data=D, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
+	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
+	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
 	scale_fill_discrete(guide = guide_legend(reverse=TRUE))
@@
 
+## Matrix mapping
+
+Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast.
+Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.
+
+### Taxa
+
+The primary intent of comparing taxa across matrices was to validate our taxon name reconciliation across studies.
+
+We first considered pairwise similarity between the same species from different matrices in different studies.
+
+```{r taxon_mapping_across_studies}
+
+Dtaxaxman = 
+	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/cross_mscript_taxa.tsv") %>%
+	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
+
+Dtaxaxman_raw_count = nrow(Dtaxaxman)
+
+Dtaxaxman %<>%
+	distinct( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2, .keep_all = TRUE )
+
+```
+
+`r Dtaxaxman_raw_count` raw pairwise comparisons passed the similarity thresholds under these sampling criteria.
+These included multiple hits for the same sequence pairs from different regions.
+When only a single hit is retained, there are `r nrow(Dtaxaxman)` retained sequence pairs.
+
+We next considered taxon mapping more generally.
+
+```{r taxon_mapping}
+
+Dtaxa = 
+	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/taxon_graph.tsv") %>%
+	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
+
+# Set aside species correspondense to same species in same matrix in same matrix
+Dtaxa_self =
+	Dtaxa %>%
+		filter( mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2  )
+
+# Remove species correspondense to same species in same matrix in same study
+Dtaxa %<>%
+	filter( !(mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2) )
+
+n_not_self = nrow(Dtaxa)
+
+# Remove hits in same matrix in same study
+Dtaxa %<>%
+	filter( !(mscript1==mscript2 & dataset1==dataset2))
+
+# Remove hits same study, presume that taxon resolution is OK within study.
+Dtaxa %<>%
+	filter( !(mscript1==mscript2))
+
+ggplot(Dtaxa) + geom_density(aes(count)) + xlim(0,50)
+
+ggplot(Dtaxa_self) + geom_density(aes(count))
+
+# These are candidates for renaming
+Dtaxa %>% filter(taxon1!=taxon2) %>% arrange( desc(count))
+
+```
+
+### Partitions
+
 ## References

### 2018-10-19 manuscript.rmd
finishing first species name pass re #5

@@
 
-# These are candidates for renaming
+# Top of this list are candidates for renaming because they are very similar but have different names
 Dtaxa %>% filter(taxon1!=taxon2) %>% arrange( desc(count))
+
+# Top of this list are candidates for renaming because they are very different but have same names
+
+Dtaxa %>% filter(taxon1==taxon2) %>% arrange( count )
+
+taxonomy_reference = read_tsv("taxonomy_info/taxon_table.tsv")
 

### 2018-10-24 scripts/blast_partitions_graph.py
used unique hit pairs of mscript:dataset:partition to construct clusters of partitions at pident>95 and evalue<1e-25, saved to file `blast/graphs/partition_clusters.tsv`.

@@ -1,0 +1,46 @@
+from glob import glob
+import pandas as pd
+import networkx as nx
+
+# set up for reading in the blast data
+colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore']
+qseq_colnames = ['mscript1', 'dataset1', 'taxon1', 'taxid1', 'part1']
+sseq_colnames = ['mscript2', 'dataset2', 'taxon2', 'taxid2', 'part2']
+# cutoffs
+pident_cutoff = 95.0
+evalue_cutoff = 1e-25
+
+# read in blast results
+for i, infile in enumerate(sorted(glob('blast/results/*_permissive.txt'))):
+    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
+    df_tmp = df_tmp[(df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff)]
+    if i ==0:
+        df = df_tmp
+    else:
+        df = pd.concat((df, df_tmp))
+
+# keep only lowest evalue for each pair of hits
+unique_best_evalue = df.groupby(['qseqid', 'sseqid'])['evalue'].min().reset_index()
+# explode the qseq columns
+qseq = unique_best_evalue['qseqid'].str.split(':', expand=True)
+sseq = unique_best_evalue['sseqid'].str.split(':', expand=True)
+qseq.columns = qseq_colnames
+sseq.columns = sseq_colnames
+unique_best_evalue = pd.concat((qseq, sseq, unique_best_evalue), axis=1)
+
+# only retain inter-manuscript hits
+inter_mscript = unique_best_evalue[unique_best_evalue['mscript1'] != unique_best_evalue['mscript2']]
+# count unique mscript:dataset:partition <-> mscript:dataset:partition 
+groupby_columns = ['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2']
+parts = inter_mscript.groupby(groupby_columns).size().reset_index(name='count')
+
+# construct graph from hits
+parts_graph = nx.Graph()
+parts_graph.add_edges_from([(tuple(x[:3]), tuple(x[3:6]), {'weight':int(x[6])}) for x in  parts.values])
+# divide graph into separate components, sorted by size
+components = nx.connected_components(parts_graph)
+out = open('blast/graphs/partition_clusters.tsv', 'w')
[…]

### 2018-10-24 scripts/blast_results_pandas.py

@@ -1,33 +1,0 @@
-#!/usr/bin/env python
-from glob import glob
-import pandas as pd
-
-colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore']
-qseq_colnames = ['mscript1', 'dataset1', 'taxon1', 'taxid1', 'part1']
-sseq_colnames = ['mscript2', 'dataset2', 'taxon2', 'taxid2', 'part2']
-pident_cutoff = 95.0
-evalue_cutoff = 1e-25
-
-for i, infile in enumerate(sorted(glob('blast/results/*_permissive.txt'))):
-    # first file
-    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
-    df_tmp = df_tmp[ (df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff) ]
-
-    qseq = df_tmp['qseqid'].str.split(':', expand=True)
-    qseq.columns = qseq_colnames
-    sseq = df_tmp['sseqid'].str.split(':', expand=True)
-    sseq.columns = sseq_colnames
-
-    df_tmp = pd.concat((qseq, sseq, df_tmp), axis=1)
-    if i ==0:
-        df = df_tmp
-    else:
-        df = pd.concat((df, df_tmp))
-
-parts_graph = df.groupby(['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2']).size().reset_index(name='count')
-parts_graph.to_csv('blast/graphs/partitions_graph.tsv', sep='\t', index=False)
-taxon_graph = df.groupby(['mscript1', 'dataset1', 'taxon1', 'mscript2', 'dataset2', 'taxon2']).size().reset_index(name='count')
-taxon_graph.to_csv('blast/graphs/taxon_graph.tsv', sep='\t', index=False)
-
-cross_mscript = df[(df['taxon1'] == df['taxon2']) & (df['mscript1'] != df['mscript2'])]
-cross_mscript.to_csv('blast/graphs/cross_mscript_taxa.tsv', sep='\t', index=False)

### 2018-10-24 manuscript.rmd

@@
 
+```{r partition_mapping}
+
+Dpart = 
+	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/partitions_graph.tsv") %>% 
+	mutate( node1 = paste(mscript1, dataset1, part1, sep = ':') ) %>% 
+	mutate( node2 = paste(mscript2, dataset2, part2, sep = ':') )
+
+Dpart_self = Dpart %>% 
+	filter( node1==node2 ) %>%
+	select( mscript1, dataset1, part1, node1, count )
+
+ggplot( Dpart_self ) + geom_density(aes(count))
+
+# Retain only nonself edges
+Dpart %<>% filter( node1!=node2 )
+
+# Network analysis
+# Builds on https://www.jessesadler.com/post/network-analysis-with-r/
+
+D_source = Dpart %>%
+  distinct(node1, .keep_all = TRUE ) %>%
+	select( node1, mscript1, dataset1, part1 ) %>%
+  dplyr::rename(label = node1, mscript = mscript1, dataset = dataset1, part = part1 )
+
+D_dest = Dpart %>%
+  distinct(node2, .keep_all = TRUE ) %>%
+	select( node2, mscript2, dataset2, part2 ) %>%
+  dplyr::rename(label = node2, mscript = mscript2, dataset = dataset2, part = part2 )
+
+nodes = 
+	bind_rows( D_source, D_dest ) %>% 
+	distinct( label, .keep_all = TRUE ) %>%
+	arrange( label ) %>% 
+	rowid_to_column("id")
+
+sources = Dpart %>%
+  distinct(node1) %>%
+  dplyr::rename(label = node1)
+
+destinations = Dpart %>%
+  distinct(node2) %>%
+  dplyr::rename(label = node2)
+
+nodes2 = 
+	full_join(sources, destinations, by = "label") %>% 
+	arrange( label ) %>% 
+	rowid_to_column("id")
+
+edges = Dpart %>% 
+	select( node1, node2, count ) %>%
+  left_join(nodes, by = c("node1" = "label")) %>% 
+  dplyr::rename(from = id) %>% 
+  left_join(nodes, by = c("node2" = "label")) %>% 
+  dplyr::rename(to = id) %>%
+	dplyr::rename(weight = count) %>%
+	select( from, to, weight )
+
+#library( network )
+
+#gene_network = network( edges, directed=FALSE, vertex.attr = nodes, matrix.type = "edgelist", ignore.eval = FALSE)
+#
+#gene_matrix = as.sociomatrix(gene_network)
+
+#library( MCL )
+
+#gene_clusters = mcl( x=gene_matrix, addLoops = TRUE )
+
+library(igraph)
+gene_igraph = graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
+gene_clusters = clusters(gene_igraph, mode=c("weak", "strong"))
+
+nodes$cluster = gene_clusters$membership
+
+# Investigate commencted components of genes (ie partitions) that include multiple members fromt he same matrix
+D = nodes %>% group_by( mscript, dataset, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+D %>% print(n=100)
+
+clusterstats = nodes %>% group_by( cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+clusterstats %>% print(n=100)
+
+```
+
+The count for a partition pair can be much arger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.
+
+There are `r length(unique(nodes$dataset))` datasets.
+A gene that is perfectly sampled would form a cluster with this size.
+Very few clusters, though, are this size.
+This suggests that intersection of genes between matrices is low
+
 ## References

### 2018-11-15 scripts/filter_swissprot.py
added swissprot results, LFS tracking for .tar.gz files

@@ -1,0 +1,18 @@
+#!/usr/bin/env python
+from glob import glob
+import pandas as pd
+
+# set up for reading in the blast data
+colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore', 'gene_info']
+# cutoffs
+pident_cutoff = 95.0
+evalue_cutoff = 1e-25
+
+for i, infile in enumerate(sorted(glob('./*_swissprot.txt'))):
+    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
+    df_tmp = df_tmp[(df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff)]
+    if i ==0:
+        df = df_tmp
+    else:
+        df = pd.concat((df, df_tmp))
+df.to_csv('swissprot_best_hits.tsv', sep='\t', index=False)

### 2018-11-15 functions.R
updated manuscript, not running

@@ -13,8 +13,9 @@ setClass(
 	Class = "PartitionedMultipleAlignment",
 	representation = representation(
 		partitions = "data.frame",
-		taxon_map = "vector",
-		partition_map = "vector"
+		manuscript_name = "character",
+		matrix_name = "character",
+		taxon_map = "vector"
 	),
 	contains = "MultipleAlignment"
 )
@@ -23,15 +24,16 @@ setClass(
 #'
 #' @param alignment_file Path to the gene alignment
 #' @param partition_file Path to the partition file in nexus format
-#' @param taxon_map A dataframe where the first column corresponds to sequence names in the alignment and the second column to clade names for the sequences
-#' @param partition_map A dataframe where the first column corresponds to partition names and the second column to names for groups of partitions
+#' @param taxon_map_global A dataframe where the first column corresponds to sequence names in the alignment and the second column to clade names for the sequences
+#' @param partition_map_global A dataframe with columns manuscript, matrix, partition, global_partition_id that maps the locally defined partition to a global partition
 #' @return A PartitionedMultipleAlignment object
 #' @export
-PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, taxon_map=NULL, partition_map=NULL, alignment_format="phylip" ) {
+PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, taxon_map_global=NULL, partition_map_global=NULL, manuscript_name=NULL, matrix_name=NULL, alignment_format="phylip" ) {
 	object = readAAMultipleAlignment( filepath = alignment_file, format=alignment_format )
 	class(object) = "PartitionedMultipleAlignment"
 	
-	
+	object@manuscript_name = manuscript_name
+	object@matrix_name = matrix_name
 		
 	conn = file( partition_file, open="r")
 	lines = readLines( conn )
@@ -54,11 +56,20 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 		) %>%
 		bind_rows()
 	
+	if( ! is.null(partition_map_global) ){
+		partition_map_global %<>%
+			filter( manuscript == manuscript_name ) %>%
+			filter( matrix == matrix_name )
+		
+		
+			
+	}
+	
 	object@partitions = partitions
 	
-	if( ! is.null(taxon_map) ){
+	if( ! is.null(taxon_map_global) ){
[…]

### 2018-11-15 manuscript/manuscript.rmd
Re-organized the repo as per #7 . Links to files are probably broken and need to be fixed

--- 2018-10-24 manuscript.rmd
@@
 
-### Alignment composition
+### Matrix taxon composition
 
@@
 
-a_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.phy"
-p_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.nex"
+phylip_file_names = list.files(pattern = "blast/phylips/.+\\.phy$")
+
+a_file = "blast/phylips/Chang2015_Chang2015.phy"
+p_file = "blast/phylips/Chang2015_Chang2015.nex"
 msa = PartitionedMultipleAlignment( a_file, p_file, taxa )
@@
 
-XXX
+### Matrix gene composition
 
-**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc).
+```{r gene_composition}
+
+partition_clusters = 
+	read_tsv( "blast/graphs/diamond_partition_clusters.tsv" ) %>% 
+	dplyr::rename( partition=partition_name ) %>%
+	dplyr::rename( matrix=dataset ) %>%
+	dplyr::rename( cluster_id=cluster_number )
+
+busco_results = 
+	read_tsv("blast/graphs/busco_metazoa_results.tsv") %>% 
+	filter( Status != "Missing" )
+
+sequences_all_txt = 
+	read_tsv("blast/all_parts_list.txt")
+
+names(sequences_all_txt) = c("full_name")
+sequences_all = str_split_fixed( sequences_all_txt$full_name, ":", 5 ) %>% 
+	as_tibble()
+names( sequences_all ) = c( "manuscript", "matrix", "species", "ncbi_taxon_id", "partition" )
+
+partitions_all = 
+	sequences_all %>% 
+	group_by( manuscript, matrix, partition ) %>% 
+	summarise( n_sequences =n() )
+
+# Multiple fields are in one colon delimited string. Need to parse them out.
+# Busco result example
+# "Moroz2014:ED3a:Capitella:51293:0241"  
+# manuscript:matrix:species:NCBI_taxon_id:partition
+
+Bs = 
+	str_split_fixed( busco_results$Sequence, ":", 5 ) %>% 
+	as_tibble()
+
+names( Bs ) = c( "manuscript", "matrix", "species", "ncbi_taxon_id", "partition" )
+
+Bs %<>% mutate( ncbi_taxon_id = as.integer(ncbi_taxon_id) )
+
+busco_results %<>% bind_cols( Bs )
+
+# Find overrepresented partitions, ie partitions with genes that hit more than one Busco
+busco_distinct =  busco_results %>% select( manuscript, matrix, partition, Description ) %>% distinct(  )
+busco_overrepresented = busco_distinct %>% group_by( manuscript, matrix, partition ) %>% summarise( n = n() ) %>% filter( n > 1 )
+busco_overrepresented
+busco_overrepresented_full = left_join( busco_overrepresented, busco_results )
+
+# Combine and summarize results
+partition_to_busco_map = 
+	busco_distinct %>% 
+	group_by( manuscript, matrix, partition ) %>% 
+	summarise( BUSCO=names(which(table(Description) == max(table(Description)))[1]) )
+
+partition_clusters %<>% left_join( partition_to_busco_map, by=c("manuscript", "matrix", "partition") )
+
+busco_summary = 
+	partition_to_busco_map %>% 
+	group_by( manuscript, matrix ) %>%
+	summarise( n_busco_partitions = n() )
+
+matrix_summary =
+	partitions_all %>%
+	group_by( manuscript, matrix ) %>%
+	summarise( n_partitions = n() ) %>%
+	left_join(
+		busco_summary,
+		by=c("manuscript", "matrix")
+	)
+
+ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) + geom_abline(slope=1,intercept=0) + coord_fixed()
+
+cluster_summary =
+	partition_clusters %>% 
+	group_by( cluster_id ) %>% 
+	summarise( 
+		n_partitions = n(), 
+		n_matrices= length(unique(matrix)),
+		n_with_busco = sum(! is.na(BUSCO)),
+		n_unique_busco = length(unique(na.omit(BUSCO)))
+	)
+
+```
+
+**Fig XXBUSCO_annotations.** The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+
+**Fig XXGene_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc).
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
 
-### Alignment overlap
+### Matrix overlap
 

### 2018-11-15 blast/gen_fastas.sh

@@ -1,19 +1,0 @@
-echo phylips/Borowiec2015_Best108.phy; ../scripts/phylippart2multifasta.py -p phylips/Borowiec2015_Best108 -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Borowiec2015_Total1080.phy; ../scripts/phylippart2multifasta.py -p phylips/Borowiec2015_Total1080 -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Chang2015_Chang2015.phy; ../scripts/phylippart2multifasta.py -p phylips/Chang2015_Chang2015 -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Dunn2008_Dunn2008.phy; ../scripts/phylippart2multifasta.py -p phylips/Dunn2008_Dunn2008 -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Hejnol2009_Hejnol2009.phy; ../scripts/phylippart2multifasta.py -p phylips/Hejnol2009_Hejnol2009 -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Moroz2014_ED3a.phy; ../scripts/phylippart2multifasta.py -p phylips/Moroz2014_ED3a -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Moroz2014_ED3d.phy; ../scripts/phylippart2multifasta.py -p phylips/Moroz2014_ED3d -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Nosenko2013_nonribosomal_9187_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_nonribosomal_9187_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Nosenko2013_ribosomal_11057_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_ribosomal_11057_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
-echo phylips/Nosenko2013_ribosomal_14615_smatrix.phy; ../scripts/phylippart2multifasta.py -p phylips/Nosenko2013_ribosomal_14615_smatrix -o queries -t ../taxonomy_info/taxon_table.tsv
[…]

### 2018-11-15 blast/gen_phylips.sh

@@ -1,18 +1,0 @@
-../scripts/prepare_phylip.py -i ../considered_data/Borowiec2015/Best108.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Borowiec2015/Total1080.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Chang2015/Chang2015.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Dunn2008/Dunn2008.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Hejnol2009/Hejnol2009.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Moroz2014/ED3a.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Moroz2014/ED3d.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Nosenko2013/nonribosomal_9187_smatrix.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Nosenko2013/ribosomal_11057_smatrix.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Nosenko2013/ribosomal_14615_smatrix.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Philippe2009/Philippe2009.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Ryan2013/est.opisthokonta.nex -o phylips -t ../taxonomy_info/taxon_table.tsv
-../scripts/prepare_phylip.py -i ../considered_data/Simion2017/supermatrix_97sp_401632pos_1719genes.phy -o phylips -t ../taxonomy_info/taxon_table.tsv
[…]

### 2018-11-15 considered_data/Simion2017/qmark2dash.sh

@@ -1,1 +1,0 @@
-cat supermatrix_97sp_401632pos_1719genes.fasta | tr ? - >supermatrix_97sp_401632pos_1719genes_gaps.fasta

### 2018-11-15 raw_data/Whelan2015/3/renameTaxa.sh

@@ -1,82 +1,0 @@
-#!/bin/bash
-sed -i 's/AMAC______/Allomyces_macrogynus/' *.phy
-sed -i 's/AFUM______/Aspergillus_fumigatus/' *.phy
-sed -i 's/MVER______/Mortierella_verticillata/' *.phy
-sed -i 's/ASPE______/Acanthoeca_spectabilis/' *.phy
-sed -i 's/NCRA______/Neurospora_crassa/' *.phy
-sed -i 's/RORY______/Rhizopus_oryzae/' *.phy
-sed -i 's/SCER______/Saccharomyces_cerevisiae/' *.phy
-sed -i 's/SPUN______/Spizellomyces_punctatus/' *.phy
-sed -i 's/APAR______/Amoebidium_parasiticum/' *.phy
-sed -i 's/COWC______/Capsaspora_owczarzaki/' *.phy
-sed -i 's/MVIB______/Ministeria_vibrans/' *.phy
-sed -i 's/SARC______/Sphaeroforma_arctica/' *.phy
-sed -i 's/MBRE______/Monosiga_brevicollis/' *.phy
-sed -i 's/MOVA______/Monosiga_ovata/' *.phy
-sed -i 's/SPYX______/Salpingoeca_pyxidium/' *.phy
-sed -i 's/SROS______/Salpingoeca_rosetta/' *.phy
-sed -i 's/TADH______/Trichoplax_adhaerens/' *.phy
-sed -i 's/ATRA______/Acropora_digitifera/' *.phy
-sed -i 's/ATET______/Abylopsis_tetragona/' *.phy
-sed -i 's/BTUE______/Bolocera_tuediae/' *.phy
-sed -i 's/AELE______/Agalma_elegans/' *.phy
-sed -i 's/NBIJ______/Nanomia_bijuga/' *.phy
-sed -i 's/APAL______/Aiptasia_pallida/' *.phy
-sed -i 's/AAUR______/Aurelia_aurita/' *.phy
-sed -i 's/CHEM______/Clytia_hemisphaerica/' *.phy
-sed -i 's/EVER______/Eunicella_verrucosa/' *.phy
-sed -i 's/GANT______/Gersemia_antarctica/' *.phy
-sed -i 's/HOLI______/Hydra_oligactis/' *.phy
-sed -i 's/HVIR______/Hydra_viridissima/' *.phy
-sed -i 's/HDIG______/Hormathia_digitata/' *.phy
[…]

### 2018-11-15 scripts/blast_partitions_graph.py

@@ -1,47 +1,0 @@
-#!/usr/bin/env python
-from glob import glob
-import pandas as pd
-import networkx as nx
-
-# set up for reading in the blast data
-colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore']
-qseq_colnames = ['mscript1', 'dataset1', 'taxon1', 'taxid1', 'part1']
-sseq_colnames = ['mscript2', 'dataset2', 'taxon2', 'taxid2', 'part2']
-# cutoffs
-pident_cutoff = 95.0
-evalue_cutoff = 1e-25
-
-# read in blast results
-for i, infile in enumerate(sorted(glob('blast/diamond_results/*_permissive.txt'))):
-    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
-    df_tmp = df_tmp[(df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff)]
-    if i ==0:
-        df = df_tmp
-    else:
-        df = pd.concat((df, df_tmp))
-
-# keep only lowest evalue for each pair of hits
-unique_best_evalue = df.groupby(['qseqid', 'sseqid'])['evalue'].min().reset_index()
-# explode the qseq columns
-qseq = unique_best_evalue['qseqid'].str.split(':', expand=True)
-sseq = unique_best_evalue['sseqid'].str.split(':', expand=True)
-qseq.columns = qseq_colnames
-sseq.columns = sseq_colnames
-unique_best_evalue = pd.concat((qseq, sseq, unique_best_evalue), axis=1)
-
-# only retain hits from different matrices/datasets
-inter_mscript = unique_best_evalue[unique_best_evalue['dataset1'] != unique_best_evalue['dataset2']]
-# count unique mscript:dataset:partition <-> mscript:dataset:partition 
-groupby_columns = ['mscript1', 'dataset1', 'part1', 'mscript2', 'dataset2', 'part2']
-parts = inter_mscript.groupby(groupby_columns).size().reset_index(name='count')
-
-# construct graph from hits
-parts_graph = nx.Graph()
-parts_graph.add_edges_from([(tuple(x[:3]), tuple(x[3:6]), {'weight':int(x[6])}) for x in  parts.values])
-# divide graph into separate components, sorted by size
-components = nx.connected_components(parts_graph)
[…]

### 2018-11-15 scripts/create_conda_env.sh

@@ -1,1 +1,0 @@
-conda env create -f ./conda_animal_root.yml

### 2018-11-15 scripts/fasta2phylip.py

@@ -1,5 +1,0 @@
-#!/usr/bin/env python
-import sys
-from Bio import AlignIO
-
-AlignIO.write(AlignIO.parse(sys.argv[1], 'fasta'), sys.argv[2], 'phylip-relaxed')

### 2018-11-15 scripts/filter_swissprot.py

@@ -1,18 +1,0 @@
-#!/usr/bin/env python
-from glob import glob
-import pandas as pd
-
-# set up for reading in the blast data
-colnames = ['qseqid', 'sseqid', 'pident', 'length', 'mismatch', 'gapopen', 'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore', 'gene_info']
-# cutoffs
-pident_cutoff = 95.0
-evalue_cutoff = 1e-25
-
-for i, infile in enumerate(sorted(glob('./*_swissprot.txt'))):
-    df_tmp = pd.read_csv(infile, sep='\t', names=colnames)
-    df_tmp = df_tmp[(df_tmp['pident']>pident_cutoff) & (df_tmp['evalue']<evalue_cutoff)]
-    if i ==0:
-        df = df_tmp
-    else:
-        df = pd.concat((df, df_tmp))
-df.to_csv('swissprot_best_hits.tsv', sep='\t', index=False)

### 2018-11-15 scripts/generate_provenance_table.py

@@ -1,146 +1,0 @@
-#!/usr/bin/env python3
-# adapted from :
-# https://stackoverflow.com/questions/16504238/attempting-to-obtain-taxonomic-information-from-biopython
-import re
-import os
-import sys
-import glob
-import urllib
-import pickle
-from collections import defaultdict
-from nexus import NexusReader # https://pypi.org/project/python-nexus/
-from Bio import SeqIO
-from Bio import Entrez
-Entrez.email = 'b.evans@yale.edu'
-
-clades = { # clades to bin samples into
-    'Cnidaria': 'Cnidaria',
-    'Bilateria': 'Bilateria',
-    'Placozoa': 'Placozoa',
-    'Porifera': 'Porifera',
-    'Ctenophora': 'Ctenophora',
-    'Choanoflagellida': 'Choanoflagellida',
-    'Ministeria': 'Filasterea',
-    'Capsaspora': 'Filasterea',
-    'Ichthyosporea': 'Ichthyosporea',
-    'Fungi': 'Fungi',
-}
-
-def get_pickles():
-    """be kind to API, don't ask for what we already know"""
-    if os.path.isfile('known_ids.pickle'):
-        k_ids = pickle.load(open('known_ids.pickle', 'rb'))
-    else:
-        k_ids = defaultdict(lambda: {})
-
-    if os.path.isfile('known_tax.pickle'):
-        k_tax = pickle.load(open('known_tax.pickle', 'rb'))
-    else:
-        k_tax = defaultdict(lambda: {})
-    return k_ids, k_tax
-
-def get_names(file_name):
-    """read in a phy or nex file, return the taxa names found"""
-    names = []
-    if file_name.endswith('.phy'):
-        phy = SeqIO.parse(file_name, 'phylip-relaxed')
-        names = [t.name for t in phy]
-    elif file_name.endswith('.nex'):
-        nex = NexusReader(file_name)
-        try:
-            names = [t for t in nex.taxa]
-        except AttributeError as e:
-            # when there isn't a separate taxa block, get names from sequences
-            names = [block[0] for block in nex.blocks['data']]
-    return names
-
-def get_tax_id(query_name):
-    """sanitize query name, search for and return NCBI ID"""
-    query_name = query_name.rstrip('_.0123456789')
-    for trim in ['_sp', '_species']:
-        if query_name.endswith(trim):
-            query_name = query_name[:-len(trim)]
[…]

### 2018-11-15 scripts/iqtree.sh

@@ -1,17 +1,0 @@
-#!/bin/bash
-#SBATCH -J iqtree
-#SBATCH -p general   
-#SBATCH -C avx2
-#SBATCH -N 1
-#SBATCH -c 20
-#SBATCH -t 7-00:00:00
-#SBATCH --mem 100G
-#SBATCH --mail-type=END,FAIL
-#SBATCH --mail-user=yuanning.li@yale.edu
-#SBATCH --output=iqtree.%j.out
-#SBATCH --error=iqtree.%j.err
-
-for FILENAME in *.nex
-do
-iqtree -s $FILENAME -nt 20 -bb 1000 -madd GTR+C60+F+R9,LG+C60+F+R9,LG+FO*H4,GTR+FO*H4, -pre $FILENAME.iqtree 
-done

### 2018-11-15 scripts/merge_busco_results.py

@@ -1,14 +1,0 @@
-#!/usr/bin/env python
-import pandas as pd
-
-metazoa_results_file = 'blast/busco/run_animal_root_metazoa_busco/full_table_animal_root_metazoa_busco.tsv'
-busco_ids_file = 'blast/busco/busco_datasets/metazoa_33208_OrthoDB9_orthogroup_info.txt'
-outfile = 'blast/graphs/busco_metazoa_results.tsv'
-
-metazoa_results = pd.read_csv(metazoa_results_file, sep='\t', 
-                 names = ['Busco_Id', 'Status', 'Sequence', 'Score', 'Length'],
-                 comment='#', index_col=0)
-busco_ids = pd.read_csv(busco_ids_file, sep='\t', index_col=0)
-
-combined = metazoa_results.join(busco_ids)
-combined.to_csv(outfile, sep='\t', index_label="Busco_Id")

### 2018-11-15 scripts/phylippart2multifasta.py

@@ -1,79 +1,0 @@
-#!/usr/bin/env python
-import os
-import re
-import sys
-import argparse
-from collections import defaultdict
-from os.path import  abspath, basename, dirname, join, splitext
-from Bio import AlignIO
-from Bio import SeqIO
-from Bio import Alphabet
-
-def get_parts(part_file_name):
-    parts = []
-    partnames = defaultdict(lambda: 0)
-    for line in open(part_file_name):
-        match = re.match(r"CHARSET\W+([\w\.\-]+)\W*=\W*(\d+)\W*-\W*(\d+)", line, re.IGNORECASE)
-        if match is not None:
-            part_name, part_start, part_end = match.groups()
-            # number non-unique partition names
-            if partnames[part_name] ==0:
-                partnames[part_name]+=1
-            else:
-                part_name = "{}_{:02d}".format(part_name, partnames[part_name])
-                partnames[part_name]+=1
-            # zero-index the intervals 
-            parts.append((part_name, int(part_start)-1, int(part_end)))
-    return parts
-
-def read_taxon_table(taxon_table_fn):
-    names_map = defaultdict(lambda: 'na')
-    table = open(taxon_table_fn, 'r')
-    header = table.readline().split()
-    tax_id = header.index('ncbi_tax_id')
-    new_name = header.index('relabelled_name')
-    for line in table:
-        tmp = line.rstrip('\n').split('\t')
-        names_map[tmp[new_name]] = tmp[tax_id]
-    return names_map
-
-desc = """phylippart_to_multifasta.py
-Simple script to take a phylip and nexus partitions file pair and convert it into a miltifasta where each partition for each taxon is a separate sequence.
-"""
-
-parser = argparse.ArgumentParser(description=desc,
-                                 formatter_class=argparse.RawTextHelpFormatter,
-                                 prog='prepare_phylip.py')
-parser.add_argument('-p', '--prefix',
-                    required=True,
-                    help='phylip & nexus file name prefix.')
-parser.add_argument('-o', '--out-dir',
-                    required=True,
-                    help='Directory to put converted fasta')
-parser.add_argument('-t', '--taxon-table',
-                    required=True,
-                    help='path to taxon_table.tsv')
-args = parser.parse_args()
-
[…]

### 2018-11-15 scripts/prepare_phylip.py

@@ -1,100 +1,0 @@
-#!/usr/bin/env python
-import os
-import sys
-import argparse
-from shutil import copy2 as copy
-from collections import defaultdict
-from os.path import  abspath, basename, dirname, join, splitext
-from nexus import NexusReader # https://pypi.org/project/python-nexus/
-from Bio import AlignIO
-
-def read_taxon_table(taxon_table_fn):
-    names_map = defaultdict(lambda: {})
-    
-    table = open(taxon_table_fn, 'r')
-    header = table.readline().split()
-    original_file = header.index('original_matrix')
-    old_name = header.index('matrix_name')
-    new_name = header.index('relabelled_name')
-    for line in table:
-        tmp = line.rstrip('\n').split('\t')
-        manuscript = basename(dirname(tmp[original_file]))
-        names_map[manuscript][tmp[old_name]] = tmp[new_name]
-    return names_map
-
-def process_nexus(manuscript_name, names_map, in_filename, out_phylip_fn, out_partition_fn):
-    n = NexusReader(in_filename)
-    out_phylip = open(out_phylip_fn, 'w')
-
-    # write phylip header
-    out_phylip.write('{} {}\n'.format(n.data.ntaxa, n.data.nchar))
-
-    # write character matrix
-    for taxon, characters in n.data:
-        if taxon in names_map[manuscript_name] and names_map[manuscript_name][taxon] != 'na':
-            out_phylip.write(names_map[manuscript_name][taxon]+' ')
-        else:
-            out_phylip.write(taxon+' ')
-        if 'missing' in n.data.format:
-            missing =  n.data.format['missing']
-            for i, character in enumerate(characters):
-                if character == missing:
-                    characters[i] = '-'
-        out_phylip.write(''.join(characters)+'\n')
-    out_partition = open(out_partition_fn, 'w')
-    out_partition.write('\n'.join(n.sets.block)+'\n')
-
-def process_phylip(manuscript_name, names_map, in_filename, out_phylip_fn, out_partition_fn):
-    
-    try:
-        in_partition_fn = splitext(in_filename)[0] + '.nex'
-        copy(in_partition_fn, out_partition_fn)
-    except IOError as e:
-        print("Error, couldn't find partition for file: {}".format(in_filename))
-        sys.exit(1)
-
-    phylip_out = open(out_phylip_fn,'w')
-    alignment = AlignIO.read(in_filename, 'phylip-relaxed')
-    for taxon in alignment:
[…]

### 2018-11-15 taxonomy_info/gen_prov_table.sh

@@ -1,1 +1,0 @@
-../scripts/generate_provenance_table.py > provenance_table.log 2>&1

### 2018-11-15 taxonomy_info/get_unique_ncbi_taxids_notfound.sh

@@ -1,2 +1,0 @@
-awk '$4 == "nan" {n=split($1,A,"/"); print A[3],$6}' taxon_table.tsv | sort | uniq > find_our_taxids.txt
-

### 2018-11-29 manuscript/manuscript.rmd
updated R code for new paths, now parse partition clusters

@@
 
-papers = read_tsv( "tables/papers.tsv" )
-datasets = read_tsv( "tables/datasets.tsv" )
+papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
+datasets = read_tsv( "../data_processed/tables/previously_published_datasets.tsv" )
 analyses = 
-	read_tsv( "tables/analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
-	mutate( taxon_sampling = factor( taxon_sampling, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
+	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
+	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
+
+taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
+
+partition_map_global = 
+	read_tsv("../reconciliation/blast/graphs/diamond_partition_clusters.tsv") %>%
+	dplyr::rename(partition=partition_name) %>%
+	dplyr::rename(matrix=dataset)
 
@@
 taxa = 
-	read_tsv("taxonomy_info/taxon_table.tsv") %>% 
+	taxonomy_reference %>% 
 	distinct( relabelled_name, clade_assignment, ncbi_tax_id ) %>%
@@
 
-a_file = "blast/phylips/Chang2015_Chang2015.phy"
-p_file = "blast/phylips/Chang2015_Chang2015.nex"
-msa = PartitionedMultipleAlignment( a_file, p_file, taxa )
+a_file = "../data_processed/matrices/Chang2015_Chang2015.phy"
+p_file = "../data_processed/matrices/Chang2015_Chang2015.nex"
+msa = PartitionedMultipleAlignment( a_file, p_file, taxa, partition_map_global=partition_map_global, manuscript_name="Chang2015", matrix_name="Chang2015")
 
@@
 partition_clusters = 
-	read_tsv( "blast/graphs/diamond_partition_clusters.tsv" ) %>% 
+	read_tsv( "../reconciliation/blast/graphs/diamond_partition_clusters.tsv" ) %>% 
 	dplyr::rename( partition=partition_name ) %>%
@@
 busco_results = 
-	read_tsv("blast/graphs/busco_metazoa_results.tsv") %>% 
+	read_tsv("../reconciliation/blast/graphs/busco_metazoa_results.tsv") %>% 
 	filter( Status != "Missing" )
@@
 sequences_all_txt = 
-	read_tsv("blast/all_parts_list.txt")
+	read_tsv("../reconciliation/blast/all_parts_list.txt")
 
@@
 analyses %>% 
-	ggplot( aes(x=model_summary, y=taxon_sampling, col=result) ) +
+	ggplot( aes(x=model_summary, y=clade, col=result) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
@@
 Dtaxaxman = 
-	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/cross_mscript_taxa.tsv") %>%
+	read_tsv("../reconciliation/blast/graphs/cross_mscript_taxa.tsv") %>%
 	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
@@
 Dtaxa = 
-	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/taxon_graph.tsv") %>%
+	read_tsv("../blast/graphs/taxon_graph.tsv") %>%
 	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
@@
 Dtaxa %>% filter(taxon1==taxon2) %>% arrange( count )
-
-taxonomy_reference = read_tsv("taxonomy_info/taxon_table.tsv")
 

### 2018-11-29 manuscript/manuscript.rmd
parses all matrices

@@
 	dplyr::rename(partition=partition_name) %>%
-	dplyr::rename(matrix=dataset)
+	dplyr::rename(matrix=dataset) %>% 
+	dplyr::rename(gene=cluster_number)
 
@@
 
-phylip_file_names = list.files(pattern = "blast/phylips/.+\\.phy$")
+matrix_path = "../data_processed/matrices/"
+phylip_file_names = list.files(path = matrix_path, pattern = ".+\\.phy$")
+
+msas = lapply(
+	phylip_file_names,
+	function( phylip_file ){
+		print(phylip_file)
+		partition_file = sub( "phy$", "nex", phylip_file )
+
+		phylip_path = paste(c(matrix_path, phylip_file), collapse="")
+		partition_path = paste(c(matrix_path, partition_file), collapse="")
+
+		base_name = sub( "\\.phy$", "", phylip_file )
+		elements = strsplit(base_name, "_") %>% unlist()
+		manuscript_name = elements[1]
+		matrix_name = paste(elements[2:length(elements)], collapse="_")
+
+		msa = 
+			PartitionedMultipleAlignment( 
+				phylip_path, 
+				partition_path, 
+				taxa, 
+				partition_map_global=partition_map_global, 
+				manuscript_name=manuscript_name, 
+				matrix_name=matrix_name
+			)
+
+		return( msa )
+	}
+)
 
@@
 p_file = "../data_processed/matrices/Chang2015_Chang2015.nex"
-msa = PartitionedMultipleAlignment( a_file, p_file, taxa, partition_map_global=partition_map_global, manuscript_name="Chang2015", matrix_name="Chang2015")
+msa = 
+	PartitionedMultipleAlignment( 
+		a_file, 
+		p_file, 
+		taxa, 
+		partition_map_global=partition_map_global, 
+		manuscript_name="Chang2015", 
+		matrix_name="Chang2015"
+	)
 
-D=clade_rects(msa)
+D=clade_rects(msas[[1]])
 

### 2018-11-30 manuscript/manuscript.rmd
flushing out errors

@@
 
+focal_matrices = 
+	read.table(text = 
+		"manuscript matrix
+		 Borowiec2015	Total1080
+		 Chang2015 Chang2015
+     Dunn2008  Dunn2008
+		 Hejnol2009 Hejnol2009
+		 Moroz2014 ED3a
+		 Nosenko2013 nonribosomal_9187_smatrix
+		 Nosenko2013 ribosomal_11057_smatrix
+		 Philippe2009	Philippe2009
+		Ryan2013	est.opisthokonta
+		Ryan2013  genome.opisthokonta
+	Simion2017	supermatrix_97sp_401632pos_1719genes
+Whelan2015	Metazoa_Choano_RCFV_strict
+
+		 ", 
+ 		header = TRUE, 
+    stringsAsFactors = FALSE)
+
 # A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
@@
 	dplyr::rename(partition=partition_name) %>%
-	dplyr::rename(matrix=dataset) %>% 
+	#dplyr::rename(matrix=dataset) %>% 
 	dplyr::rename(gene=cluster_number)
@@
 
-msas = lapply(
+sequence_matrices = lapply(
 	phylip_file_names,
@@
 
-		msa = 
+		sequence_matrix = 
 			PartitionedMultipleAlignment( 
@@
 
-		return( msa )
+		return( sequence_matrix )
 	}
@@
 
-a_file = "../data_processed/matrices/Chang2015_Chang2015.phy"
-p_file = "../data_processed/matrices/Chang2015_Chang2015.nex"
-msa = 
-	PartitionedMultipleAlignment( 
-		a_file, 
-		p_file, 
-		taxa, 
-		partition_map_global=partition_map_global, 
-		manuscript_name="Chang2015", 
-		matrix_name="Chang2015"
-	)
-
-D=clade_rects(msas[[1]])
+D=clade_rects(sequence_matrices[[1]])
 
@@
 ```{r gene_composition}
-
-partition_clusters = 
-	read_tsv( "../reconciliation/blast/graphs/diamond_partition_clusters.tsv" ) %>% 
-	dplyr::rename( partition=partition_name ) %>%
-	dplyr::rename( matrix=dataset ) %>%
-	dplyr::rename( cluster_id=cluster_number )
 
@@
 
-partition_clusters %<>% left_join( partition_to_busco_map, by=c("manuscript", "matrix", "partition") )
+partition_map_global %<>% left_join( partition_to_busco_map, by=c("manuscript", "matrix", "partition") )
 
@@
 cluster_summary =
-	partition_clusters %>% 
-	group_by( cluster_id ) %>% 
+	partition_map_global %>% 
+	group_by( gene ) %>% 
 	summarise( 
@@
 
-XXX
+```{r}
+
+matrix_overlap = 
+	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) matrix_overlap(x,y))) %>% 
+	unlist(recursive=FALSE) %>% 
+	bind_rows()
+
+# Remove reciprocal comparisons and comparisons to self
+n = nrow(matrix_overlap)
+mask = lower.tri(matrix(1:n^2, n, n))
+dim(mask) = NULL
+matrix_overlap = matrix_overlap[mask,]
+
+```
 
@@
 Dpart = 
-	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/partitions_graph.tsv") %>% 
+	read_tsv("../reconciliation/blast/graphs/partitions_graph.tsv") %>% 
 	mutate( node1 = paste(mscript1, dataset1, part1, sep = ':') ) %>% 

### 2018-11-30 manuscript/functions.R

@@ -134,5 +134,24 @@ stat_clades = function(mapping = NULL, data = NULL, geom = "rect",
 
 
 
+matrix_overlap = function( msa1, msa2 ){
+	
+	n_gene_overlap = sum( na.omit(msa1@partitions$gene) %in% na.omit(msa2@partitions$gene) )
 
+	n_species_overlap = sum( rownames(msa1) %in% rownames(msa2) )
+	
+	tibble(
+		manuscript_name_1 = msa1@manuscript_name,
+		matrix_name_1 = msa1@matrix_name,
+		n_species_1 = nrow( msa1 ),
+		n_partitions_1 = nrow( msa1@partitions ),
+		manuscript_name_2 = msa2@manuscript_name,
+		matrix_name_2 = msa2@matrix_name,
+		n_species_2 = nrow( msa2 ),
+		n_partitions_2 = nrow( msa2@partitions ),
+		n_species_overlap = n_species_overlap,
+		n_gene_overlap = n_gene_overlap
+		
+	)
+}
 

### 2018-11-30 manuscript/manuscript.rmd
facet figure of matrix clade composition

@@
 
-D=clade_rects(sequence_matrices[[1]])
+taxon_rectangles = lapply(
+	sequence_matrices,
+	function ( sequence_matrix ){
+		clade_rects(sequence_matrix)
+	}
+) %>%
+	bind_rows()
 
-ggplot() + 
+ggplot(data=taxon_rectangles) + 
 	scale_x_continuous(name="x") + 
 	scale_y_continuous(name="y") +
-	geom_rect(data=D, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
 	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
 	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
-	scale_fill_discrete(guide = guide_legend(reverse=TRUE))
+	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
+	facet_wrap( ~ manuscript_matrix )
+
 ```
@@
 matrix_overlap = 
-	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) matrix_overlap(x,y))) %>% 
+	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) compute_matrix_overlap(x,y))) %>% 
 	unlist(recursive=FALSE) %>% 

### 2018-11-30 manuscript/functions.R

@@ -106,6 +106,11 @@ clade_rects = function( data ){
 	D$xmax = rep( n_genes, length(clades) )
 	D$ymin = c(0, clade_cumsums[1:length(clade_cumsums)-1])
 	D$ymax = clade_cumsums
+	D$manuscript = data@manuscript_name
+	D$matrix = data@matrix_name
+	D$manuscript_matrix = paste(D$manuscript, D$matrix, sep="_")
+	
+	D %<>% mutate( manuscript_matrix = as.factor(manuscript_matrix) )
 	
 	D
 }
@@ -134,7 +139,7 @@ stat_clades = function(mapping = NULL, data = NULL, geom = "rect",
 
 
 
-matrix_overlap = function( msa1, msa2 ){
+compute_matrix_overlap = function( msa1, msa2 ){
 	
 	n_gene_overlap = sum( na.omit(msa1@partitions$gene) %in% na.omit(msa2@partitions$gene) )
 

### 2018-11-30 manuscript/manuscript.rmd
first overlap figure

@@
 
+matrix_overlap_rectangles = 
+	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) overlap_rects(x,y))) %>% 
+	unlist(recursive=FALSE) %>% 
+	bind_rows()
+
+ggplot(data=matrix_overlap_rectangles) + 
+	scale_x_continuous(name="x") + 
+	scale_y_continuous(name="y") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) + 
+	facet_grid( matrix_1 ~ matrix_2 )
+
 ```

### 2018-11-30 manuscript/functions.R

@@ -160,3 +160,25 @@ compute_matrix_overlap = function( msa1, msa2 ){
 	)
 }
 
+overlap_rects = function( msa1, msa2 ){
+	overlap = compute_matrix_overlap( msa1, msa2 )
+	
+	# Create a dataframe where each row is one of the matrices in the pairwise comparison and 
+	# the columns describe how to draw the rectangle for matrix
+	
+	# First row is msa1 rectangle, second row is msa2 rectangle
+	D = data.frame( MSA=c("Matrix_1", "Matrix_2") )
+	
+	D$xmin = c( 0, overlap$n_partitions_1 - overlap$n_gene_overlap )
+	D$xmax = c( overlap$n_partitions_1, overlap$n_partitions_1 + overlap$n_partitions_2 - overlap$n_gene_overlap )
+	D$ymin = c(0, overlap$n_species_1 - overlap$n_species_overlap )
+	D$ymax = c( overlap$n_species_1, overlap$n_species_1 + overlap$n_species_2 - overlap$n_species_overlap )
+	D$manuscript_1 = msa1@manuscript_name
+	D$manuscript_2 = msa2@manuscript_name
+	D$matrix_1 = paste(msa1@manuscript_name, msa1@matrix_name, sep="_")
+	D$matrix_2 = paste(msa2@manuscript_name, msa2@matrix_name, sep="_")
+	
+	D
+	
+	
+}

### 2018-11-30 manuscript/manuscript.rmd
stuff

@@
 	dplyr::rename(partition=partition_name) %>%
-	#dplyr::rename(matrix=dataset) %>% 
-	dplyr::rename(gene=cluster_number)
+	dplyr::rename(gene=cluster_number) %>%
+	mutate( gene = as.character(gene) )
 

### 2018-11-30 manuscript/functions.R

@@ -66,6 +66,11 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 			select( -manuscript ) %>% 
 			select( -matrix )
 		
+		# Add names to the partitions that are not assigned to genes. Take these from the original 
+		# partition names, but prepend matrix name to be sure they are unique across matrices
+		
+		partitions$gene[ is.na(partitions$gene) ] = paste( matrix_name, partitions$partition[ is.na(partitions$gene) ], collapse = '_' )
+		
 	}
 	
 	object@partitions = partitions

### 2018-12-06 manuscript/manuscript.rmd
parsing basic info from iqtree runs

@@
 library(viridis)
+library(ape)
 
@@
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
-datasets = read_tsv( "../data_processed/tables/previously_published_datasets.tsv" )
+datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
 analyses = 
@@
 
+```{r}
+
+trees_path = "../trees_new/"
+tree_file_names = list.files(path = trees_path, pattern = ".+\\.contree$")
+
+trees = lapply(
+	tree_file_names,
+	function( tree_file ){
+		print(tree_file)
+		log_file = sub( "contree$", "log", tree_file )
+
+		tree_file_path = paste(c(trees_path, tree_file), collapse="")
+		log_file_path = paste(c(trees_path, log_file), collapse="")
+
+		base_name = sub( "\\.contree$", "", tree_file )
+		elements = strsplit(base_name, "\\.") %>% unlist()
+		matrix_name = elements[1]
+		model_name = elements[2]
+
+		tree = read.tree( file = tree_file_path )
+
+		# Get the clades of the tips
+		clades = 
+			taxonomy_reference$clade_assignment[ 
+				match( tree$tip.label, taxonomy_reference$relabelled_name ) 
+				]
+
+		# Get the outgroup sampling
+
+		outgroup = NA
+
+		sampling = "Metazoa"
+		if( "Choanoflagellida" %in% clades ){
+			sampling = "Choanimalia"
+			outgroup = which( clades == "Choanoflagellida" )
+		}
+
+		if( ("Ichthyosporea" %in% clades) | ("Filasterea" %in% clades) ){
+			sampling = "Holozoa"
+			outgroup = c( which( clades == "Ichthyosporea" ),  which( clades == "Filasterea" ))
+		}
+
+		if( "Fungi" %in% clades ){
+			sampling = "Opisthokonta"
+			outgroup = which( clades == "Fungi" )
+		}
+
+		# Reroot the tree
+		# Use a single taxon from the outgroup since the outgroup is not guaranteed to be monophyletic
+		tree = root( tree, outgroup[1] )
+
+		# Get the clades of the tips again, since tip order has changed
+		clades = 
+			taxonomy_reference$clade_assignment[ 
+				match( tree$tip.label, taxonomy_reference$relabelled_name ) 
+				]
+
+		# Stuff a few other things into the tree object
+		tree$matrix = matrix_name
+		tree$model = model_name
+		tree$modelfinder = FALSE
+		tree$clades = clades
+		tree$sampling = sampling
+
+		# parse model
+		log_lines = read_lines( log_file_path )
+
+		best_line = log_lines[ grepl("Best-fit model:", log_lines) ]
+
+		if( length(best_line) > 1 ){
+			warning("More than one best model found")
+		}
+
+		if( length(best_line) > 0 ){
+			tree$modelfinder = TRUE
+			elements = strsplit( best_line[1], " " ) %>% unlist()
+			tree$model = elements[3]
+		}
+
+		return( tree )
+	}
+)
+
+tree_summary = lapply(
+	trees,
+	function( tree ){
+		data.frame(
+			matrix = tree$matrix,
+			model = tree$model,
+			modelfinder = tree$modelfinder,
+			sampling = tree$sampling,
+			stringsAsFactors = FALSE
+		)
+	}
+) %>% bind_rows()
+
+```
+
 One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.

### 2018-12-20 manuscript/manuscript.rmd

@@
 trees_path = "../trees_new/"
-tree_file_names = list.files(path = trees_path, pattern = ".+\\.contree$")
+contree_extension = "treefile"
+tree_file_names = list.files(path = trees_path, pattern = str_c( "\\.", contree_extension, "$" ))
 
@@
 		print(tree_file)
-		log_file = sub( "contree$", "log", tree_file )
+		log_file = sub( str_c( contree_extension, "$" ), "log", tree_file )
 
@@
 
-		base_name = sub( "\\.contree$", "", tree_file )
+		if(!file.exists(log_file_path)){
+			warning( str_c( "Missong log file: ", log_file_path ) )
+			return( NA )
+		}
+
+		base_name = sub( str_c( "\\.", contree_extension, "$" ), "", tree_file )
 		elements = strsplit(base_name, "\\.") %>% unlist()

### 2019-01-03 manuscript/manuscript.rmd
updated R code for new files, tweaked nexus formatting for consistency

@@
 		manuscript_name = elements[1]
-		matrix_name = paste(elements[2:length(elements)], collapse="_")
+		matrix_name = base_name
 
@@
 names(sequences_all_txt) = c("full_name")
-sequences_all = str_split_fixed( sequences_all_txt$full_name, ":", 5 ) %>% 
+sequences_all = str_split_fixed( sequences_all_txt$full_name, ":", 4 ) %>% 
 	as_tibble()
-names( sequences_all ) = c( "manuscript", "matrix", "species", "ncbi_taxon_id", "partition" )
+names( sequences_all ) = c( "matrix", "species", "ncbi_taxon_id", "partition" )
 
@@
 	sequences_all %>% 
-	group_by( manuscript, matrix, partition ) %>% 
+	group_by( matrix, partition ) %>% 
 	summarise( n_sequences =n() )
@@
 Bs = 
-	str_split_fixed( busco_results$Sequence, ":", 5 ) %>% 
+	str_split_fixed( busco_results$Sequence, ":", 4 ) %>% 
 	as_tibble()
 
-names( Bs ) = c( "manuscript", "matrix", "species", "ncbi_taxon_id", "partition" )
+names( Bs ) = c( "matrix", "species", "ncbi_taxon_id", "partition" )
 
@@
 # Find overrepresented partitions, ie partitions with genes that hit more than one Busco
-busco_distinct =  busco_results %>% select( manuscript, matrix, partition, Description ) %>% distinct(  )
-busco_overrepresented = busco_distinct %>% group_by( manuscript, matrix, partition ) %>% summarise( n = n() ) %>% filter( n > 1 )
+busco_distinct =  busco_results %>% select( matrix, partition, Description ) %>% distinct(  )
+busco_overrepresented = busco_distinct %>% group_by( matrix, partition ) %>% summarise( n = n() ) %>% filter( n > 1 )
 busco_overrepresented
@@
 	busco_distinct %>% 
-	group_by( manuscript, matrix, partition ) %>% 
+	group_by( matrix, partition ) %>% 
 	summarise( BUSCO=names(which(table(Description) == max(table(Description)))[1]) )
 
-partition_map_global %<>% left_join( partition_to_busco_map, by=c("manuscript", "matrix", "partition") )
+partition_map_global %<>% left_join( partition_to_busco_map, by=c("matrix", "partition") )
 
@@
 	partition_to_busco_map %>% 
-	group_by( manuscript, matrix ) %>%
+	group_by( matrix ) %>%
 	summarise( n_busco_partitions = n() )
@@
 	partitions_all %>%
-	group_by( manuscript, matrix ) %>%
+	group_by( matrix ) %>%
 	summarise( n_partitions = n() ) %>%
@@
 		busco_summary,
-		by=c("manuscript", "matrix")
+		by=c("matrix")
 	)
+
+matrix_summary$manuscript = str_split( matrix_summary$matrix, "_", simplify=TRUE )[,1]
 

### 2019-01-03 manuscript/functions.R

@@ -58,12 +58,10 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 	
 	if( ! is.null(partition_map_global) ){
 		partition_map_global %<>%
-			filter( manuscript == manuscript_name ) %>%
 			filter( matrix == matrix_name )
 		
 		partitions %<>% 
 			left_join( partition_map_global ) %>%
-			select( -manuscript ) %>% 
 			select( -matrix )
 		
 		# Add names to the partitions that are not assigned to genes. Take these from the original 

### 2019-01-03 manuscript/manuscript.rmd
updated partition comparison code

@@
 library(ape)
+library(igraph)
 
@@
 
-```{r taxon_mapping_across_studies}
+```{r taxon_mapping_across_studies, eval=FALSE}
 
@@
 
-```{r taxon_mapping}
+```{r taxon_mapping, eval=FALSE}
 
@@
 	read_tsv("../reconciliation/blast/graphs/partitions_graph.tsv") %>% 
-	mutate( node1 = paste(mscript1, dataset1, part1, sep = ':') ) %>% 
-	mutate( node2 = paste(mscript2, dataset2, part2, sep = ':') )
+	mutate( node1 = paste(matrix1, part1, sep = ':') ) %>% 
+	mutate( node2 = paste(matrix2, part2, sep = ':') )
 
-Dpart_self = Dpart %>% 
-	filter( node1==node2 ) %>%
-	select( mscript1, dataset1, part1, node1, count )
-
-ggplot( Dpart_self ) + geom_density(aes(count))
+# Hits to self not included anymore, so skip this bit
+# Dpart_self = Dpart %>% 
+#	 filter( node1==node2 ) %>%
+#	 select( matrix1, part1, node1, count )
+# ggplot( Dpart_self ) + geom_density(aes(count))
 
@@
   distinct(node1, .keep_all = TRUE ) %>%
-	select( node1, mscript1, dataset1, part1 ) %>%
-  dplyr::rename(label = node1, mscript = mscript1, dataset = dataset1, part = part1 )
+	select( node1, matrix1, part1 ) %>%
+  dplyr::rename(label = node1, matrix = matrix1, part = part1 )
 
@@
   distinct(node2, .keep_all = TRUE ) %>%
-	select( node2, mscript2, dataset2, part2 ) %>%
-  dplyr::rename(label = node2, mscript = mscript2, dataset = dataset2, part = part2 )
+	select( node2, matrix2, part2 ) %>%
+  dplyr::rename(label = node2, matrix = matrix2, part = part2 )
 
@@
 
-#library( network )
-
-#gene_network = network( edges, directed=FALSE, vertex.attr = nodes, matrix.type = "edgelist", ignore.eval = FALSE)
-#
-#gene_matrix = as.sociomatrix(gene_network)
-
-#library( MCL )
-
-#gene_clusters = mcl( x=gene_matrix, addLoops = TRUE )
-
-library(igraph)
 gene_igraph = graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
@@
 # Investigate commencted components of genes (ie partitions) that include multiple members fromt he same matrix
-D = nodes %>% group_by( mscript, dataset, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+D = nodes %>% group_by( matrix, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
 D %>% print(n=100)

### 2019-01-04 manuscript/manuscript.rmd
added parallel code

@@
 library(igraph)
+library(hutan)
+library(doParallel)
 
 source( "functions.R" )
+
+# Set system computational parameters
+cores = detectCores() - 1
+if ( cores < 1 ) {
+	cores = 1
+}
+
+# Register parallel workers for %dopar%
+registerDoParallel( cores )
 ```
@@
 
-sequence_matrices = lapply(
-	phylip_file_names,
-	function( phylip_file ){
+parse_phylip =	function( phylip_file ){
 		print(phylip_file)
@@
 	}
-)
+
+sequence_matrices = foreach( phylip_file=phylip_file_names) %dopar%
+	parse_phylip( phylip_file )
 
@@
 
-trees = lapply(
-	tree_file_names,
-	function( tree_file ){
+# Temp fix re issue #8
+tree_file_names = tree_file_names[ ! grepl("Philippe2009", tree_file_names) ]
+
+tree_file_names = tree_file_names[ ! grepl("Nosenko2013", tree_file_names) ]
+
+parse_tree = 	function( tree_file ){
 		print(tree_file)
 		log_file = sub( str_c( contree_extension, "$" ), "log", tree_file )
+		bootstrap_file = sub( str_c( contree_extension, "$" ), "ufboot", tree_file )
 
@@
 		if(!file.exists(log_file_path)){
-			warning( str_c( "Missong log file: ", log_file_path ) )
+			warning( str_c( "Missing log file: ", log_file_path ) )
+			return( NA )
+		}
+
+		bootstrap_file_path = paste(c(trees_path, bootstrap_file), collapse="")
+		if(!file.exists(bootstrap_file_path)){
+			warning( str_c( "Missing bootstrap file: ", bootstrap_file_path ) )
 			return( NA )
@@
 
+		if( is.na(outgroup) ){
+			stop( "There are no outgroups in the tree, can't test rooting topology" )
+		}
+
 		# Reroot the tree
@@
 
+		if( sum(is.na(clades)) > 0 ){
+			missing_clades = tree$tip.label[ is.na(clades) ]
+			stop( paste( c( "Some tips have missing clade names:", missing_clades ), collapse=" " ))
+		}
+
+		if( sum(clades == "Ctenophora") < 1 ){
+			stop( "There are no ctenophores in the tree, can't test rooting topology" )
+		}
+
+		if( sum(clades == "Porifera") < 1 ){
+			stop( "There are no sponges in the tree, can't test rooting topology" )
+		}
+
+		# Get the set of taxa in the group Ctenophora+Placozoa+Bilateria+Cnidaria, the clade that exists under Proifera-sister
+		CtPlBiCn = tree$tip.label[ clades %in% c("Ctenophora", "Placozoa", "Bilateria", "Cnidaria") ]
+
+		# Get the set of taxa in the group Porifera+Placozoa+Bilateria+Cnidaria, the clade that exists under Ctenophora-sister
+		PoPlBiCn = tree$tip.label[ clades %in% c("Porifera", "Placozoa", "Bilateria", "Cnidaria") ]
+
+		# Read the bootstrap trees
+		bootstrap_trees = read.tree(file = bootstrap_file_path)
+
+		Ctenophora_sister = lapply(
+			bootstrap_trees,
+			function( phy ){
+				is_monophyletic( phy, PoPlBiCn )
+			}
+		) %>% 
+		unlist() %>%
+		mean()
+
+		Porifera_sister = lapply(
+			bootstrap_trees,
+			function( phy ){
+				is_monophyletic( phy, CtPlBiCn )
+			}
+		) %>% 
+		unlist() %>%
+		mean()
+
 		# Stuff a few other things into the tree object
@@
 		tree$sampling = sampling
+		tree$ctenophora_sister = Ctenophora_sister
+		tree$porifera_sister = Porifera_sister
 
@@
 	}
-)
+
+trees = foreach( tree_file=tree_file_names) %dopar%
+	parse_tree( tree_file )
 

### 2019-01-04 manuscript/manuscript.rmd
cleaned up R loose ends, reknit successfully

@@
 
+		# Do some cludgey renaming to make it easy to step into function below for debugging
+		alignment_path = phylip_path
+		partition_file = partition_path
+		taxon_map_global = taxa
+
 		sequence_matrix = 
 			PartitionedMultipleAlignment( 
-				phylip_path, 
-				partition_path, 
-				taxa, 
+				alignment_path, 
+				partition_file, 
+				taxon_map_global, 
 				partition_map_global=partition_map_global, 
@@
 
-```{r taxon_mapping_across_studies, eval=FALSE}
-
-Dtaxaxman = 
-	read_tsv("../reconciliation/blast/graphs/cross_mscript_taxa.tsv") %>%
-	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
-
-Dtaxaxman_raw_count = nrow(Dtaxaxman)
-
-Dtaxaxman %<>%
-	distinct( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2, .keep_all = TRUE )
-
-```
-
-`r Dtaxaxman_raw_count` raw pairwise comparisons passed the similarity thresholds under these sampling criteria.
-These included multiple hits for the same sequence pairs from different regions.
-When only a single hit is retained, there are `r nrow(Dtaxaxman)` retained sequence pairs.
-
-We next considered taxon mapping more generally.
-
-```{r taxon_mapping, eval=FALSE}
-
-Dtaxa = 
-	read_tsv("../blast/graphs/taxon_graph.tsv") %>%
-	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
-
-# Set aside species correspondense to same species in same matrix in same matrix
-Dtaxa_self =
-	Dtaxa %>%
-		filter( mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2  )
-
-# Remove species correspondense to same species in same matrix in same study
-Dtaxa %<>%
-	filter( !(mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2) )
-
-n_not_self = nrow(Dtaxa)
-
-# Remove hits in same matrix in same study
-Dtaxa %<>%
-	filter( !(mscript1==mscript2 & dataset1==dataset2))
-
-# Remove hits same study, presume that taxon resolution is OK within study.
-Dtaxa %<>%
-	filter( !(mscript1==mscript2))
-
-ggplot(Dtaxa) + geom_density(aes(count)) + xlim(0,50)
-
-ggplot(Dtaxa_self) + geom_density(aes(count))
-
-# Top of this list are candidates for renaming because they are very similar but have different names
-Dtaxa %>% filter(taxon1!=taxon2) %>% arrange( desc(count))
-
-# Top of this list are candidates for renaming because they are very different but have same names
-
-Dtaxa %>% filter(taxon1==taxon2) %>% arrange( count )
-
-```
-
 ### Partitions

### 2019-01-04 manuscript/functions.R

@@ -37,6 +37,7 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 		
 	conn = file( partition_file, open="r")
 	lines = readLines( conn )
+	lines = str_replace( lines, "charset", "CHARSET" )
 	lines = lines[ grepl("CHARSET", lines) ]
 	close( conn )
 	

### 2019-01-04 manuscript/matrix_comparison.R

@@ -1,0 +1,74 @@
+# Comparison of LG and WAG matrices
+
+library( tidyverse )
+library( magrittr )
+
+# Order in iqtree matrix
+aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
+
+# Functional groups adjacent, from https://en.wikipedia.org/wiki/Amino_acid
+aa_functional = c( "R", "H", "K", "D", "E", "S", "T", "N", "Q", "C", "G", "P", "A", "V", "I", "L", "M", "F", "Y", "W" )
+
+triange_to_matrix = function( triangle ){
+	R_matrix = rep(NA, 400)
+	dim( R_matrix ) = c(20,20)
+	
+
+	rownames( R_matrix ) = aa
+	colnames( R_matrix ) = aa
+	
+	# Though iqtree specifies lower triangle, actually in the correct order for upper triangle as this R command reads it
+	R_matrix [ upper.tri( R_matrix ) ] = triangle 
+	
+	# Now fill in the lower half of the matrix
+	R_matrix[ is.na(R_matrix) ] = t( R_matrix )[ is.na(R_matrix) ]
+	
+	# Scale
+	R_matrix = R_matrix / mean( R_matrix, na.rm=TRUE )
+	
+	# Calculate diagonla so that sum of each row is 0
+	# diag( R_matrix ) = - rowSums ( R_matrix, na.rm=TRUE )
+	
+
+	
+	# Reorder by functional groupings
+	R_matrix = R_matrix[, match(aa_functional, colnames(R_matrix) )]
+	R_matrix = R_matrix[match(aa_functional, rownames(R_matrix)), ]
+	
+	R_matrix
+}
+
+plot_matrix = function( R_matrix ){
+	D = data.frame( R_matrix )
+	D$source = row.names(D)
+	Dg = gather( D, 1:20, key=dest, value=exch, -source )
+	Dg %<>%
+		mutate( source=factor(source, levels=aa_functional)  ) %>%
+		mutate( dest=factor(dest, levels=aa_functional)  )
+	
+	Dg %>%
+		ggplot(aes(x=dest, y=source)) + 
+		geom_tile(aes(fill=exch)) +
+		scale_fill_gradient2()
+}
+
+
+
+
[…]

### 2019-01-04 manuscript/manuscript.rmd
fixed warning
moved compute intensive code to manuscript_kernel.R, started modifying docker image to support manuscript execution

--- 2019-01-03 manuscript/manuscript.rmd
@@
 	dpi=300,
-	cache=TRUE
+	cache=FALSE
 	)
+```
+
+```{r load_data}
+
+	# Load data and analysis results generated by 
+	# manuscript_kernel.R
+	load("manuscript.RData")
 ```
@@
 library( magrittr )
-library(viridis)
-library(ape)
-library(igraph)
 
 source( "functions.R" )
-```
-
-```{r}
-
-focal_matrices = 
-	read.table(text = 
-		"manuscript matrix
-		 Borowiec2015	Total1080
-		 Chang2015 Chang2015
-     Dunn2008  Dunn2008
-		 Hejnol2009 Hejnol2009
-		 Moroz2014 ED3a
-		 Nosenko2013 nonribosomal_9187_smatrix
-		 Nosenko2013 ribosomal_11057_smatrix
-		 Philippe2009	Philippe2009
-		Ryan2013	est.opisthokonta
-		Ryan2013  genome.opisthokonta
-	Simion2017	supermatrix_97sp_401632pos_1719genes
-Whelan2015	Metazoa_Choano_RCFV_strict
-
-		 ", 
- 		header = TRUE, 
-    stringsAsFactors = FALSE)
 
@@
 names( result_colors ) = c( "Ctenophora-sister", "Porifera-sister", "Unresolved" )
-
-papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
-datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
-analyses = 
-	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
-	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
-
-taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
-
-partition_map_global = 
-	read_tsv("../reconciliation/blast/graphs/diamond_partition_clusters.tsv") %>%
-	dplyr::rename(partition=partition_name) %>%
-	dplyr::rename(gene=cluster_number) %>%
-	mutate( gene = as.character(gene) )
-
-bootstrap_threshold = 90
-posterior_prob_threshold = 95
-
-analyses$result = "Unresolved"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_porifera_sister >= bootstrap_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
-analyses$result = factor( analyses$result )
-
-analyses$model_summary = analyses$model_rate_matrix
-long_summary = paste(analyses$model_rate_matrix, analyses$model_equilibrium, sep="+")
-analyses$model_summary[ ! is.na(analyses$model_equilibrium) ] = long_summary[ ! is.na(analyses$model_equilibrium) ]
-analyses$model_summary[ grepl("WAG", analyses$model_summary) ] = "WAG" # Simplify WAG
-analyses$model_summary = factor( analyses$model_summary, levels=c("GTR+CAT", "F81+CAT", "LG", "WAG") )
 
@@
 
-```{r taxonomy}
-
-clades = c( "Fungi", "Ichthyosporea", "Filasterea", "Choanoflagellida", "Ctenophora", "Porifera", "Placozoa", "Bilateria", "Cnidaria" )
-
-clade_colors = viridis( length(clades) )
-names( clade_colors ) = clades
-
-taxa = 
-	taxonomy_reference %>% 
-	distinct( relabelled_name, clade_assignment, ncbi_tax_id ) %>%
-	dplyr::rename( taxon=relabelled_name, clade=clade_assignment ) %>%
-	mutate( clade = factor( clade, levels=clades )  )
-
-```
-
-```{r}
-
-matrix_path = "../data_processed/matrices/"
-phylip_file_names = list.files(path = matrix_path, pattern = ".+\\.phy$")
-
-sequence_matrices = lapply(
-	phylip_file_names,
-	function( phylip_file ){
-		print(phylip_file)
-		partition_file = sub( "phy$", "nex", phylip_file )
-
-		phylip_path = paste(c(matrix_path, phylip_file), collapse="")
-		partition_path = paste(c(matrix_path, partition_file), collapse="")
-
-		base_name = sub( "\\.phy$", "", phylip_file )
-		elements = strsplit(base_name, "_") %>% unlist()
-		manuscript_name = elements[1]
-		matrix_name = base_name
-
-		sequence_matrix = 
-			PartitionedMultipleAlignment( 
-				phylip_path, 
-				partition_path, 
-				taxa, 
-				partition_map_global=partition_map_global, 
-				manuscript_name=manuscript_name, 
-				matrix_name=matrix_name
-			)
-
-		return( sequence_matrix )
-	}
-)
+```{ r taxon_rectangles }
 
@@
 bind_rows()
+
+clade_colors = viridis( length(clades) )
+names( clade_colors ) = clades
 
@@
 
-busco_results = 
-	read_tsv("../reconciliation/blast/graphs/busco_metazoa_results.tsv") %>% 
-	filter( Status != "Missing" )
-
-sequences_all_txt = 
-	read_tsv("../reconciliation/blast/all_parts_list.txt")
-
-names(sequences_all_txt) = c("full_name")
-sequences_all = str_split_fixed( sequences_all_txt$full_name, ":", 4 ) %>% 
-	as_tibble()
-names( sequences_all ) = c( "matrix", "species", "ncbi_taxon_id", "partition" )
-
-partitions_all = 
-	sequences_all %>% 
-	group_by( matrix, partition ) %>% 
-	summarise( n_sequences =n() )
-
-# Multiple fields are in one colon delimited string. Need to parse them out.
-# Busco result example
-# "Moroz2014:ED3a:Capitella:51293:0241"  
-# manuscript:matrix:species:NCBI_taxon_id:partition
-
-Bs = 
-	str_split_fixed( busco_results$Sequence, ":", 4 ) %>% 
-	as_tibble()
-
-names( Bs ) = c( "matrix", "species", "ncbi_taxon_id", "partition" )
-
-Bs %<>% mutate( ncbi_taxon_id = as.integer(ncbi_taxon_id) )
-
-busco_results %<>% bind_cols( Bs )
-
-# Find overrepresented partitions, ie partitions with genes that hit more than one Busco
-busco_distinct =  busco_results %>% select( matrix, partition, Description ) %>% distinct(  )
-busco_overrepresented = busco_distinct %>% group_by( matrix, partition ) %>% summarise( n = n() ) %>% filter( n > 1 )
 busco_overrepresented
-busco_overrepresented_full = left_join( busco_overrepresented, busco_results )
-
-# Combine and summarize results
-partition_to_busco_map = 
-	busco_distinct %>% 
-	group_by( matrix, partition ) %>% 
-	summarise( BUSCO=names(which(table(Description) == max(table(Description)))[1]) )
-
-partition_map_global %<>% left_join( partition_to_busco_map, by=c("matrix", "partition") )
-
-busco_summary = 
-	partition_to_busco_map %>% 
-	group_by( matrix ) %>%
-	summarise( n_busco_partitions = n() )
-
-matrix_summary =
-	partitions_all %>%
-	group_by( matrix ) %>%
-	summarise( n_partitions = n() ) %>%
-	left_join(
-		busco_summary,
-		by=c("matrix")
-	)
-
-matrix_summary$manuscript = str_split( matrix_summary$matrix, "_", simplify=TRUE )[,1]
 
 ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) + geom_abline(slope=1,intercept=0) + coord_fixed()
-
-cluster_summary =
-	partition_map_global %>% 
-	group_by( gene ) %>% 
-	summarise( 
-		n_partitions = n(), 
-		n_matrices= length(unique(matrix)),
-		n_with_busco = sum(! is.na(BUSCO)),
-		n_unique_busco = length(unique(na.omit(BUSCO)))
-	)
+	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) + 
+	geom_abline(slope=1,intercept=0) + 
+	coord_fixed()
 
@@
 
-```{r}
-
-matrix_overlap = 
-	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) compute_matrix_overlap(x,y))) %>% 
-	unlist(recursive=FALSE) %>% 
-	bind_rows()
-
-# Remove reciprocal comparisons and comparisons to self
-n = nrow(matrix_overlap)
-mask = lower.tri(matrix(1:n^2, n, n))
-dim(mask) = NULL
-matrix_overlap = matrix_overlap[mask,]
+```{r matrix_overlap}
 
@@
 ```{r}
-
-trees_path = "../trees_new/"
-contree_extension = "treefile"
-tree_file_names = list.files(path = trees_path, pattern = str_c( "\\.", contree_extension, "$" ))
-
-trees = lapply(
-	tree_file_names,
-	function( tree_file ){
-		print(tree_file)
-		log_file = sub( str_c( contree_extension, "$" ), "log", tree_file )
-
-		tree_file_path = paste(c(trees_path, tree_file), collapse="")
-		log_file_path = paste(c(trees_path, log_file), collapse="")
-
-		if(!file.exists(log_file_path)){
-			warning( str_c( "Missong log file: ", log_file_path ) )
-			return( NA )
-		}
-
-		base_name = sub( str_c( "\\.", contree_extension, "$" ), "", tree_file )
-		elements = strsplit(base_name, "\\.") %>% unlist()
-		matrix_name = elements[1]
-		model_name = elements[2]
-
-		tree = read.tree( file = tree_file_path )
-
-		# Get the clades of the tips
-		clades = 
-			taxonomy_reference$clade_assignment[ 
-				match( tree$tip.label, taxonomy_reference$relabelled_name ) 
-				]
-
-		# Get the outgroup sampling
-
-		outgroup = NA
-
-		sampling = "Metazoa"
-		if( "Choanoflagellida" %in% clades ){
-			sampling = "Choanimalia"
-			outgroup = which( clades == "Choanoflagellida" )
-		}
-
-		if( ("Ichthyosporea" %in% clades) | ("Filasterea" %in% clades) ){
-			sampling = "Holozoa"
-			outgroup = c( which( clades == "Ichthyosporea" ),  which( clades == "Filasterea" ))
-		}
-
-		if( "Fungi" %in% clades ){
-			sampling = "Opisthokonta"
-			outgroup = which( clades == "Fungi" )
-		}
-
-		# Reroot the tree
-		# Use a single taxon from the outgroup since the outgroup is not guaranteed to be monophyletic
-		tree = root( tree, outgroup[1] )
-
-		# Get the clades of the tips again, since tip order has changed
-		clades = 
-			taxonomy_reference$clade_assignment[ 
-				match( tree$tip.label, taxonomy_reference$relabelled_name ) 
-				]
-
-		# Stuff a few other things into the tree object
-		tree$matrix = matrix_name
-		tree$model = model_name
-		tree$modelfinder = FALSE
-		tree$clades = clades
-		tree$sampling = sampling
-
-		# parse model
-		log_lines = read_lines( log_file_path )
-
-		best_line = log_lines[ grepl("Best-fit model:", log_lines) ]
-
-		if( length(best_line) > 1 ){
-			warning("More than one best model found")
-		}
-
-		if( length(best_line) > 0 ){
-			tree$modelfinder = TRUE
-			elements = strsplit( best_line[1], " " ) %>% unlist()
-			tree$model = elements[3]
-		}
-
-		return( tree )
-	}
-)
-
-tree_summary = lapply(
-	trees,
-	function( tree ){
-		data.frame(
-			matrix = tree$matrix,
-			model = tree$model,
-			modelfinder = tree$modelfinder,
-			sampling = tree$sampling,
-			stringsAsFactors = FALSE
-		)
-	}
-) %>% bind_rows()
 
@@
 
-### Taxa
+### Taxa comparison across matrices
 
@@
 
-```{r taxon_mapping_across_studies, eval=FALSE}
-
-Dtaxaxman = 
-	read_tsv("../reconciliation/blast/graphs/cross_mscript_taxa.tsv") %>%
-	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
-
-Dtaxaxman_raw_count = nrow(Dtaxaxman)
-
-Dtaxaxman %<>%
-	distinct( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2, .keep_all = TRUE )
-
-```
-
-`r Dtaxaxman_raw_count` raw pairwise comparisons passed the similarity thresholds under these sampling criteria.
-These included multiple hits for the same sequence pairs from different regions.
-When only a single hit is retained, there are `r nrow(Dtaxaxman)` retained sequence pairs.
-
-We next considered taxon mapping more generally.
-
-```{r taxon_mapping, eval=FALSE}
-
-Dtaxa = 
-	read_tsv("../blast/graphs/taxon_graph.tsv") %>%
-	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
-
-# Set aside species correspondense to same species in same matrix in same matrix
-Dtaxa_self =
-	Dtaxa %>%
-		filter( mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2  )
-
-# Remove species correspondense to same species in same matrix in same study
-Dtaxa %<>%
-	filter( !(mscript1 == mscript2 & dataset1==dataset2 & taxon1==taxon2) )
-
-n_not_self = nrow(Dtaxa)
-
-# Remove hits in same matrix in same study
-Dtaxa %<>%
-	filter( !(mscript1==mscript2 & dataset1==dataset2))
-
-# Remove hits same study, presume that taxon resolution is OK within study.
-Dtaxa %<>%
-	filter( !(mscript1==mscript2))
-
-ggplot(Dtaxa) + geom_density(aes(count)) + xlim(0,50)
-
-ggplot(Dtaxa_self) + geom_density(aes(count))
-
-# Top of this list are candidates for renaming because they are very similar but have different names
-Dtaxa %>% filter(taxon1!=taxon2) %>% arrange( desc(count))
-
-# Top of this list are candidates for renaming because they are very different but have same names
-
-Dtaxa %>% filter(taxon1==taxon2) %>% arrange( count )
-
-```
-
-### Partitions
+### Partition comparison across matrices
 
@@
 
-Dpart = 
-	read_tsv("../reconciliation/blast/graphs/partitions_graph.tsv") %>% 
-	mutate( node1 = paste(matrix1, part1, sep = ':') ) %>% 
-	mutate( node2 = paste(matrix2, part2, sep = ':') )
+# Investigate connected components of genes (ie partitions) that include multiple members fromt he same matrix
+clusterstats_within_manuscripts = nodes %>% group_by( matrix, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+clusterstats = nodes %>% group_by( cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
 
-# Hits to self not included anymore, so skip this bit
-# Dpart_self = Dpart %>% 
-#	 filter( node1==node2 ) %>%
-#	 select( matrix1, part1, node1, count )
-# ggplot( Dpart_self ) + geom_density(aes(count))
-
-# Retain only nonself edges
-Dpart %<>% filter( node1!=node2 )
-
-# Network analysis
-# Builds on https://www.jessesadler.com/post/network-analysis-with-r/
-
-D_source = Dpart %>%
-  distinct(node1, .keep_all = TRUE ) %>%
-	select( node1, matrix1, part1 ) %>%
-  dplyr::rename(label = node1, matrix = matrix1, part = part1 )
-
-D_dest = Dpart %>%
-  distinct(node2, .keep_all = TRUE ) %>%
-	select( node2, matrix2, part2 ) %>%
-  dplyr::rename(label = node2, matrix = matrix2, part = part2 )
-
-nodes = 
-	bind_rows( D_source, D_dest ) %>% 
-	distinct( label, .keep_all = TRUE ) %>%
-	arrange( label ) %>% 
-	rowid_to_column("id")
-
-sources = Dpart %>%
-  distinct(node1) %>%
-  dplyr::rename(label = node1)
-
-destinations = Dpart %>%
-  distinct(node2) %>%
-  dplyr::rename(label = node2)
-
-nodes2 = 
-	full_join(sources, destinations, by = "label") %>% 
-	arrange( label ) %>% 
-	rowid_to_column("id")
-
-edges = Dpart %>% 
-	select( node1, node2, count ) %>%
-  left_join(nodes, by = c("node1" = "label")) %>% 
-  dplyr::rename(from = id) %>% 
-  left_join(nodes, by = c("node2" = "label")) %>% 
-  dplyr::rename(to = id) %>%
-	dplyr::rename(weight = count) %>%
-	select( from, to, weight )
-
-gene_igraph = graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
-gene_clusters = clusters(gene_igraph, mode=c("weak", "strong"))
-
-nodes$cluster = gene_clusters$membership
-
-# Investigate commencted components of genes (ie partitions) that include multiple members fromt he same matrix
-D = nodes %>% group_by( matrix, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
-D %>% print(n=100)
-
-clusterstats = nodes %>% group_by( cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+clusterstats_within_manuscripts %>% print(n=100)
 clusterstats %>% print(n=100)
@@
 
-There are `r length(unique(nodes$dataset))` datasets.
+There are `r length(unique(nodes$matrix))` matrices.
 A gene that is perfectly sampled would form a cluster with this size.

### 2019-01-04 manuscript/manuscript_kernel.R

@@ -1,0 +1,467 @@
+# The computationally intensive analyses for the manuscript 
+# presented at https://github.com/caseywdunn/animal_root/tree/master/manuscript .
+#
+# Executing this code generates manuscript.RData, which contains analysis 
+# results. That file is then read by manuscript.rmd for rendering and 
+# presentation of the results.
+# 
+# The code presented here is roughly in the order of the analyses presented 
+# in the manuscript, though there are exceptions
+
+## Preliminaries
+
+time_start = Sys.time()
+
+library( tidyverse )
+library( magrittr )
+library( viridis )
+library( ape )
+library( igraph )
+library( hutan )
+library( doParallel )
+
+source( "functions.R" )
+
+# Set system computational parameters
+cores = detectCores() - 1
+if ( cores < 1 ) {
+	cores = 1
+}
+
+# Register parallel workers for %dopar%
+registerDoParallel( cores )
+
+# Set up constants
+focal_matrices = 
+	read.table(text = 
+		"manuscript matrix
+		Borowiec2015	Total1080
+		Chang2015 Chang2015
+		Dunn2008  Dunn2008
+		Hejnol2009 Hejnol2009
+		Moroz2014 ED3a
+		Nosenko2013 nonribosomal_9187_smatrix
+		Nosenko2013 ribosomal_11057_smatrix
+		Philippe2009	Philippe2009
+		Ryan2013	est.opisthokonta
+		Ryan2013  genome.opisthokonta
+		Simion2017	supermatrix_97sp_401632pos_1719genes
+		Whelan2015	Metazoa_Choano_RCFV_strict
+		", 
+		header = TRUE, 
+		stringsAsFactors = FALSE)
+
+bootstrap_threshold = 90
+posterior_prob_threshold = 95
+
+# Load data
+
+papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
+datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
+analyses = 
+	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
+	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
+
+taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
[…]

### 2019-01-04 manuscript/manuscript.rmd
added summary figure for new analyses

@@
 
-```{r support}
+```{r support_published_analyses}
 
-analyses %>% 
+analyses_published %>% 
 	ggplot( aes(x=model_summary, y=clade, col=result) ) +
@@
 		theme_classic()
+
 ```
 
-A total of `r nrow(analyses)` analyses were transcribed from the literature.
+A total of `r nrow(analyses_published)` analyses were transcribed from the literature.
 
@@
 
-```{r}
+```{r support_new_analyses}
 
+analyses_new %>% 
+	ggplot( aes(x=model_summary, y=clade, col=result, shape=modelfinder) ) +
+		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		scale_colour_manual(values = result_colors) +
+		theme_classic()
 ```

### 2019-01-04 manuscript/manuscript_kernel.R

@@ -58,7 +58,7 @@ posterior_prob_threshold = 95
 
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
-analyses = 
+analyses_published = 
 	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
 	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
 
@@ -70,18 +70,18 @@ partition_map_global =
 	dplyr::rename(gene=cluster_number) %>%
 	mutate( gene = as.character(gene) )
 
-analyses$result = "Unresolved"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_porifera_sister >= bootstrap_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
-analyses$result = factor( analyses$result )
+analyses_published$result = "Unresolved"
+analyses_published$result[ ( analyses_published$inference == "Bayesian" ) & (analyses_published$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
+analyses_published$result[ ( analyses_published$inference == "Bayesian" ) & (analyses_published$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
+analyses_published$result[ ( analyses_published$inference == "ML" ) &       (analyses_published$support_porifera_sister >= bootstrap_threshold) ] = "Porifera-sister"
+analyses_published$result[ ( analyses_published$inference == "ML" ) &       (analyses_published$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
+analyses_published$result = factor( analyses_published$result )
 
[…]

### 2019-01-04 manuscript/manuscript_kernel.R
fixed formatting

@@ -14,7 +14,6 @@ time_start = Sys.time()
 
 library( tidyverse )
 library( magrittr )
-library( viridis )
 library( ape )
 library( igraph )
 library( hutan )

### 2019-01-04 manuscript/manuscript_kernel.R
added constraint trees

@@ -222,6 +222,7 @@ matrix_overlap = matrix_overlap[mask,]
 # New analyses of published matrices
 
 trees_path = "../trees_new/"
+constraint_tree_path = "../trees_new/constraint_trees/"
 contree_extension = "treefile"
 tree_file_names = list.files(path = trees_path, pattern = str_c( "\\.", contree_extension, "$" ))
 
@@ -314,6 +315,21 @@ parse_tree = 	function( tree_file ){
 	
 	# Get the set of taxa in the group Porifera+Placozoa+Bilateria+Cnidaria, the clade that exists under Ctenophora-sister
 	PoPlBiCn = tree$tip.label[ clades %in% c("Porifera", "Placozoa", "Bilateria", "Cnidaria") ]
+	
+	# Write constraint trees for hypothesis testing
+	CtPlBiCn_not = tree$tip.label[ ! tree$tip.label %in% CtPlBiCn ]
+	porifera_sister_constraint_tree = generate_constaint_tree( CtPlBiCn, CtPlBiCn_not )
+	write.tree( 
+		porifera_sister_constraint_tree, 
+		file = paste( constraint_tree_path, base_name, ".porifera_sister_constraint.tree", sep="") 
+	)
+
+	PoPlBiCn_not = tree$tip.label[ ! tree$tip.label %in% PoPlBiCn ]
+	ctenophora_sister_constraint_tree = generate_constaint_tree( PoPlBiCn, PoPlBiCn_not )
+	write.tree( 
+		ctenophora_sister_constraint_tree, 
+		file = paste( constraint_tree_path, base_name, ".ctenophora_sister_constraint.tree", sep="") 
+	)	
 	
 	# Read the bootstrap trees
 	bootstrap_trees = read.tree(file = bootstrap_file_path)

### 2019-01-16 manuscript/manuscript.rmd
fixed formatting
moved matrix comparison tomanuscript

@@
 library( magrittr )
+library( viridis )
 
@@
 
-```{ r taxon_rectangles }
+```{r taxon_rectangles }
 
@@
 
+## Model matrix comparison
+
+WAG and LG are both fixed exchange matrices. Their differences are largely limitted to a few amino acid changes.
+
+```{r wagvlg}
+
+# Comparison of LG and WAG matrices
+
+library( tidyverse )
+library( magrittr )
+
+# Order in iqtree matrix
+aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
+
+# Functional groups adjacent, from https://en.wikipedia.org/wiki/Amino_acid
+aa_functional = c( "R", "H", "K", "D", "E", "S", "T", "N", "Q", "C", "G", "P", "A", "V", "I", "L", "M", "F", "Y", "W" )
+
+triange_to_matrix = function( triangle ){
+	R_matrix = rep(NA, 400)
+	dim( R_matrix ) = c(20,20)
+
+	rownames( R_matrix ) = aa
+	colnames( R_matrix ) = aa
+
+	# Though iqtree specifies lower triangle, actually in the correct order for upper triangle as this R command reads it
+	R_matrix [ upper.tri( R_matrix ) ] = triangle
+
+	# Now fill in the lower half of the matrix
+	R_matrix[ is.na(R_matrix) ] = t( R_matrix )[ is.na(R_matrix) ]
+
+	# Scale
+	R_matrix = R_matrix / mean( R_matrix, na.rm=TRUE )
+
+	# Reorder by functional groupings
+	R_matrix = R_matrix[, match(aa_functional, colnames(R_matrix) )]
+	R_matrix = R_matrix[match(aa_functional, rownames(R_matrix)), ]
+
+	R_matrix
+}
+
+plot_matrix = function( R_matrix ){
+	D = data.frame( R_matrix )
+	D$source = row.names(D)
+	Dg = gather( D, 1:20, key=dest, value=exch, -source )
+	Dg %<>%
+		mutate( source=factor(source, levels=aa_functional)  ) %>%
+		mutate( dest=factor(dest, levels=aa_functional)  )
+
+	Dg %>%
+		ggplot(aes(x=dest, y=source)) + 
+		geom_tile(aes(fill=exch)) +
+		scale_fill_gradient2()
+}
+
+# Triangle matrices from iqtree source code at https://github.com/Cibiv/IQ-TREE/blob/master/model/modelprotein.cpp
+
+wag_triangle = c(55.15710, 50.98480, 63.53460, 73.89980, 14.73040, 542.94200, 102.70400, 52.81910, 26.52560, 3.02949, 90.85980, 303.55000, 154.36400, 61.67830, 9.88179, 158.28500, 43.91570, 94.71980, 617.41600, 2.13520, 546.94700, 141.67200, 58.46650, 112.55600, 86.55840, 30.66740, 33.00520, 56.77170, 31.69540, 213.71500, 395.62900, 93.06760, 24.89720, 429.41100, 57.00250, 24.94100, 19.33350, 18.69790, 55.42360, 3.94370, 17.01350, 11.39170, 12.73950, 3.04501, 13.81900, 39.79150, 49.76710, 13.15280, 8.48047, 38.42870, 86.94890, 15.42630, 6.13037, 49.94620, 317.09700, 90.62650, 535.14200, 301.20100, 47.98550, 7.40339, 389.49000, 258.44300, 37.35580, 89.04320, 32.38320, 25.75550, 89.34960, 68.31620, 19.82210, 10.37540, 39.04820, 154.52600, 31.51240, 17.41000, 40.41410, 425.74600, 485.40200, 93.42760, 21.04940, 10.27110, 9.61621, 4.67304, 39.80200, 9.99208, 8.11339, 4.99310, 67.93710, 105.94700, 211.51700, 8.88360, 119.06300, 143.85500, 67.94890, 19.50810, 42.39840, 10.94040, 93.33720, 68.23550, 24.35700, 69.61980, 9.99288, 41.58440, 55.68960, 17.13290, 16.14440, 337.07900, 122.41900, 397.42300, 107.17600, 140.76600, 102.88700, 70.49390, 134.18200, 74.01690, 31.94400, 34.47390, 96.71300, 49.39050, 54.59310, 161.32800, 212.11100, 55.44130, 203.00600, 37.48660, 51.29840, 85.79280, 82.27650, 22.58330, 47.33070, 145.81600, 32.66220, 138.69800, 151.61200, 17.19030, 79.53840, 437.80200, 11.31330, 116.39200, 7.19167, 12.97670, 71.70700, 21.57370, 15.65570, 33.69830, 26.25690, 21.24830, 66.53090, 13.75050, 51.57060, 152.96400, 13.94050, 52.37420, 11.08640, 24.07350, 38.15330, 108.60000, 32.57110, 54.38330, 22.77100, 19.63030, 10.36040, 387.34400, 42.01700, 39.86180, 13.32640, 42.84370, 645.42800, 21.60460, 78.69930, 29.11480, 248.53900, 200.60100, 25.18490, 19.62460, 15.23350, 100.21400, 30.12810, 58.87310, 18.72470, 11.83580, 782.13000, 180.03400, 30.54340, 205.84500, 64.98920, 31.48870, 23.27390, 138.82300, 36.53690, 31.47300)
+wag_matrix = triange_to_matrix( wag_triangle ) %>% log()
+
+lg_triangle = c(0.425093, 0.276818, 0.751878, 0.395144, 0.123954, 5.076149, 2.489084, 0.534551, 0.528768, 0.062556, 0.969894, 2.807908, 1.695752, 0.523386, 0.084808, 1.038545, 0.363970, 0.541712, 5.243870, 0.003499, 4.128591, 2.066040, 0.390192, 1.437645, 0.844926, 0.569265, 0.267959, 0.348847, 0.358858, 2.426601, 4.509238, 0.927114, 0.640543, 4.813505, 0.423881, 0.311484, 0.149830, 0.126991, 0.191503, 0.010690, 0.320627, 0.072854, 0.044265, 0.008705, 0.108882, 0.395337, 0.301848, 0.068427, 0.015076, 0.594007, 0.582457, 0.069673, 0.044261, 0.366317, 4.145067, 0.536518, 6.326067, 2.145078, 0.282959, 0.013266, 3.234294, 1.807177, 0.296636, 0.697264, 0.159069, 0.137500, 1.124035, 0.484133, 0.371004, 0.025548, 0.893680, 1.672569, 0.173735, 0.139538, 0.442472, 4.273607, 6.312358, 0.656604, 0.253701, 0.052722, 0.089525, 0.017416, 1.105251, 0.035855, 0.018811, 0.089586, 0.682139, 1.112727, 2.592692, 0.023918, 1.798853, 1.177651, 0.332533, 0.161787, 0.394456, 0.075382, 0.624294, 0.419409, 0.196961, 0.508851, 0.078281, 0.249060, 0.390322, 0.099849, 0.094464, 4.727182, 0.858151, 4.008358, 1.240275, 2.784478, 1.223828, 0.611973, 1.739990, 0.990012, 0.064105, 0.182287, 0.748683, 0.346960, 0.361819, 1.338132, 2.139501, 0.578987, 2.000679, 0.425860, 1.143480, 1.080136, 0.604545, 0.129836, 0.584262, 1.033739, 0.302936, 1.136863, 2.020366, 0.165001, 0.571468, 6.472279, 0.180717, 0.593607, 0.045376, 0.029890, 0.670128, 0.236199, 0.077852, 0.268491, 0.597054, 0.111660, 0.619632, 0.049906, 0.696175, 2.457121, 0.095131, 0.248862, 0.140825, 0.218959, 0.314440, 0.612025, 0.135107, 1.165532, 0.257336, 0.120037, 0.054679, 5.306834, 0.232523, 0.299648, 0.131932, 0.481306, 7.803902, 0.089613, 0.400547, 0.245841, 3.151815, 2.547870, 0.170887, 0.083688, 0.037967, 1.959291, 0.210332, 0.245034, 0.076701, 0.119013, 10.649107, 1.702745, 0.185202, 1.898718, 0.654683, 0.296501, 0.098369, 2.188158, 0.189510, 0.249313)
+lg_matrix = triange_to_matrix( lg_triangle ) %>% log()
+
+plot_matrix(wag_matrix)
+plot_matrix(lg_matrix)
+
+lg_wag_diff = lg_matrix - wag_matrix
+
+plot_matrix(lg_wag_diff)
+
+lg_wag_diff_thresh =lg_wag_diff
+lg_wag_diff_thresh[abs(lg_wag_diff_thresh)<1] = 0
+
+plot_matrix( lg_wag_diff_thresh )
+
+```
+
 ## Matrix mapping

### 2019-01-16 manuscript/matrix_comparison.R

@@ -52,9 +52,6 @@ plot_matrix = function( R_matrix ){
 		scale_fill_gradient2()
 }
 
-
-
-
 wag_triangle = c(55.15710, 50.98480, 63.53460, 73.89980, 14.73040, 542.94200, 102.70400, 52.81910, 26.52560, 3.02949, 90.85980, 303.55000, 154.36400, 61.67830, 9.88179, 158.28500, 43.91570, 94.71980, 617.41600, 2.13520, 546.94700, 141.67200, 58.46650, 112.55600, 86.55840, 30.66740, 33.00520, 56.77170, 31.69540, 213.71500, 395.62900, 93.06760, 24.89720, 429.41100, 57.00250, 24.94100, 19.33350, 18.69790, 55.42360, 3.94370, 17.01350, 11.39170, 12.73950, 3.04501, 13.81900, 39.79150, 49.76710, 13.15280, 8.48047, 38.42870, 86.94890, 15.42630, 6.13037, 49.94620, 317.09700, 90.62650, 535.14200, 301.20100, 47.98550, 7.40339, 389.49000, 258.44300, 37.35580, 89.04320, 32.38320, 25.75550, 89.34960, 68.31620, 19.82210, 10.37540, 39.04820, 154.52600, 31.51240, 17.41000, 40.41410, 425.74600, 485.40200, 93.42760, 21.04940, 10.27110, 9.61621, 4.67304, 39.80200, 9.99208, 8.11339, 4.99310, 67.93710, 105.94700, 211.51700, 8.88360, 119.06300, 143.85500, 67.94890, 19.50810, 42.39840, 10.94040, 93.33720, 68.23550, 24.35700, 69.61980, 9.99288, 41.58440, 55.68960, 17.13290, 16.14440, 337.07900, 122.41900, 397.42300, 107.17600, 140.76600, 102.88700, 70.49390, 134.18200, 74.01690, 31.94400, 34.47390, 96.71300, 49.39050, 54.59310, 161.32800, 212.11100, 55.44130, 203.00600, 37.48660, 51.29840, 85.79280, 82.27650, 22.58330, 47.33070, 145.81600, 32.66220, 138.69800, 151.61200, 17.19030, 79.53840, 437.80200, 11.31330, 116.39200, 7.19167, 12.97670, 71.70700, 21.57370, 15.65570, 33.69830, 26.25690, 21.24830, 66.53090, 13.75050, 51.57060, 152.96400, 13.94050, 5
[…]

### 2019-01-16 manuscript/matrix_comparison.R
removed old comp file

@@ -1,71 +1,0 @@
-# Comparison of LG and WAG matrices
-
-library( tidyverse )
-library( magrittr )
-
-# Order in iqtree matrix
-aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
-
-# Functional groups adjacent, from https://en.wikipedia.org/wiki/Amino_acid
-aa_functional = c( "R", "H", "K", "D", "E", "S", "T", "N", "Q", "C", "G", "P", "A", "V", "I", "L", "M", "F", "Y", "W" )
-
-triange_to_matrix = function( triangle ){
-	R_matrix = rep(NA, 400)
-	dim( R_matrix ) = c(20,20)
-	
-
-	rownames( R_matrix ) = aa
-	colnames( R_matrix ) = aa
-	
-	# Though iqtree specifies lower triangle, actually in the correct order for upper triangle as this R command reads it
-	R_matrix [ upper.tri( R_matrix ) ] = triangle 
-	
-	# Now fill in the lower half of the matrix
-	R_matrix[ is.na(R_matrix) ] = t( R_matrix )[ is.na(R_matrix) ]
-	
-	# Scale
-	R_matrix = R_matrix / mean( R_matrix, na.rm=TRUE )
-	
-	# Calculate diagonla so that sum of each row is 0
-	# diag( R_matrix ) = - rowSums ( R_matrix, na.rm=TRUE )
-	
-
-	
-	# Reorder by functional groupings
-	R_matrix = R_matrix[, match(aa_functional, colnames(R_matrix) )]
-	R_matrix = R_matrix[match(aa_functional, rownames(R_matrix)), ]
-	
-	R_matrix
-}
-
-plot_matrix = function( R_matrix ){
-	D = data.frame( R_matrix )
-	D$source = row.names(D)
-	Dg = gather( D, 1:20, key=dest, value=exch, -source )
-	Dg %<>%
-		mutate( source=factor(source, levels=aa_functional)  ) %>%
-		mutate( dest=factor(dest, levels=aa_functional)  )
-	
-	Dg %>%
-		ggplot(aes(x=dest, y=source)) + 
-		geom_tile(aes(fill=exch)) +
-		scale_fill_gradient2()
-}
-
[…]

### 2019-01-17 manuscript/manuscript.rmd
added model results

@@
 library( viridis )
+library( knitr )
+library( gridExtra )
 
@@
 
+```{r modelfinder}
+
+analyses_new %>% 
+	filter(modelfinder==TRUE) %>% 
+	select( matrix, clade, result, model_summary ) %>%
+	kable()
+
+```
+
+Table XXModelfinder. The models selected by modelfinder for each matrix.
+
 ### Comparison of iqtree and phylobayes results
@@
 
-plot_matrix = function( R_matrix ){
+plot_matrix = function( R_matrix, title=NULL ){
 	D = data.frame( R_matrix )
@@
 		geom_tile(aes(fill=exch)) +
-		scale_fill_gradient2()
+		scale_fill_gradient2() +
+		ggtitle(title)
 }
@@
 
-plot_matrix(wag_matrix)
-plot_matrix(lg_matrix)
-
 lg_wag_diff = lg_matrix - wag_matrix
-
-plot_matrix(lg_wag_diff)
 
@@
 
-plot_matrix( lg_wag_diff_thresh )
+grid.arrange(
+	plot_matrix(lg_matrix, title="(A) LG matrix"),
+	plot_matrix(wag_matrix, title="(B) WAG matrix"),
+	plot_matrix(lg_wag_diff, title="(C) LG - WAG"),
+	plot_matrix( lg_wag_diff_thresh, title="(D) Thresholded LG - WAG" )
+)
 

### 2019-01-17 manuscript/manuscript_kernel.R
read previously_published_analyses 20 columns

@@ -58,7 +58,7 @@ posterior_prob_threshold = 95
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
 analyses_published = 
-	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
+	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnnccc' ) %>%
 	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
 
 taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")

### 2019-01-17 manuscript/manuscript.rmd
a few additons to methods

@@
 
+We retreived matrices from each publication (Table XX), storing the raw data in this manuscript's version control repository.
+We manually edited some minor formatting to make the batch processing of the matrices *en masse*, e.g.
+standardizing the formatting of charset blocks.
+All changes made are tracked with git.
+
 ### Matrix comparison and annotation
+
+#### Taxon name reconciliation
+
+We programatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
+We also use a table where manual entries were needed (Supp Table XX, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g.
+authors of original matrix indicate species name in original manuscript.
+For a table summarizing all samples and their new or lengthened names, see Table XX(reconciliation/taxonomy_info/taxon_table.tsv).
+
+#### All by all sequence comparison
+
+We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default `diamond blastp` parameters.
+We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 from these results (filters: pident > 95.0, eValue < 1e-25, no self->self) where each node is a sequence and each edge is a result.
+We extracted each component of connected nodes such that that that every component has no edges to any other nodes.
 
@@
 ## Ackowledgements
+
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam cluser.
 

### 2019-02-28 manuscript/manuscript.rmd
new pass at blast partition network; relaxed filters for diamond and blast/swissprot results to include more partitions

@@
 
-#### All by all sequence comparison
+#### Sequence Comparisons
 
-We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default `diamond blastp` parameters.
-We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 from these results (filters: pident > 95.0, eValue < 1e-25, no self->self) where each node is a sequence and each edge is a result.
-We extracted each component of connected nodes such that that that every component has no edges to any other nodes.
+We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond blastp parameters.
+We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self).
+We ran BUSCO witrh default parameters for all sequences against the provided metazoa gene set.
+We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering hit such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+
+#### Partition Network
+
+We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 where each node is a partition and each edge represents a DIAMOND sequence to sequence match.
+We extracted each component of connected nodes where every component has no edges to any other nodes.
+We decorated every node with the most numerous SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
 

### 2019-03-14 manuscript/manuscript.rmd
stuff

--- 2019-01-17 manuscript/manuscript.rmd
@@
 
-In the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses.
+Over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses.
 Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
 In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
+
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.

### 2019-03-21 manuscript/manuscript.rmd
added brief description of node splitting

@@
 We extracted each component of connected nodes where every component has no edges to any other nodes.
-We decorated every node with the most numerous SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+We further split these components if the the most connected node (i.e.
+most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
+We then decorated every node with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
 

### 2019-03-28 manuscript/manuscript.rmd
word smithing for networks

@@
 
+Using the partition files for each matrix, we isolated each sequence for each taxon from each partition.
+Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
+
 We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond blastp parameters.
@@
 We ran BUSCO witrh default parameters for all sequences against the provided metazoa gene set.
-We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering hit such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering hits such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
 
@@
 
-We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 where each node is a partition and each edge represents a DIAMOND sequence to sequence match.
-We extracted each component of connected nodes where every component has no edges to any other nodes.
+We next used the sequence similarity comparisons derived above to compare partitions.
+
+We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
+We extracted each connected component from this network.
 We further split these components if the the most connected node (i.e.
 most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
-We then decorated every node with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
 

### 2019-05-16 manuscript/manuscript_kernel.R
count number of ribosomal annotated genes properly

@@ -322,7 +322,7 @@ n_ribo =
 		group_by(matrix,gene) %>% 
 		tally(ribo_found) %>% 
 		group_by(matrix) %>% 
-		count(name="n_ribo")
+		tally()
 discarded_parts = 
 	read_tsv("../reconciliation/blast/graphs/discarded_nodes.tsv") %>%
 	group_by(matrix) %>%

### 2019-05-16 manuscript/manuscript.rmd
manuscript knits again. still needs phyo methods. figure and table org and polish, conclusion.

@@
 We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self).
-We ran BUSCO witrh default parameters for all sequences against the provided metazoa gene set.
-We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering hits such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+We ran BUSCO with default parameters for all sequences against the provided metazoa gene set.
+We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
 
@@
 
-We next used the sequence similarity comparisons derived above to compare partitions.
+We used the sequence similarity comparisons described above to compare partitions.
 
@@
 We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+See Table XX [partition_network_summary table] for a summary tally of each part of the comparison.
 
@@
 # Investigate connected components of genes (ie partitions) that include multiple members fromt he same matrix
-clusterstats_within_manuscripts = nodes %>% group_by( matrix, cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
-clusterstats = nodes %>% group_by( cluster ) %>% summarise(n=n()) %>% arrange(desc(n))
+clusterstats_within_manuscripts = partition_map_global %>% group_by( matrix, component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
+clusterstats = partition_map_global %>% group_by( component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
 
@@
 clusterstats %>% print(n=100)
+partition_network_summary %>% print()
 
@@
 
-There are `r length(unique(nodes$matrix))` matrices.
+There are `r length(unique(partition_network_summary$matrix))` matrices.
 A gene that is perfectly sampled would form a cluster with this size.

### 2019-05-16 manuscript/manuscript_kernel.R

@@ -69,8 +69,7 @@ taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
 partition_map_global = 
 	read_tsv("../reconciliation/blast/graphs/partition_components_split_annotated.tsv") %>%
 	dplyr::rename(partition=partition_name) %>%
-	dplyr::rename(gene=component_number) %>%
-	mutate( gene = as.character(gene) )
+	mutate( component_number = as.character(component_number) )
 
 analyses_published$result = "Unresolved"
 analyses_published$result[ ( analyses_published$inference == "Bayesian" ) & (analyses_published$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
@@ -227,7 +226,7 @@ matrix_summary$manuscript = str_split( matrix_summary$matrix, "_", simplify=TRUE
 
 cluster_summary =
 	partition_map_global %>% 
-	group_by( gene ) %>% 
+	group_by( component_number ) %>% 
 	summarise( 
 		n_partitions = n(), 
 		n_matrices= length(unique(matrix)),
@@ -307,19 +306,19 @@ n_compandBUSCO =
 	partition_map_global %>%
 		group_by(matrix) %>%
 		summarize(
-			"n_components"=n_distinct(gene),
+			"n_components"=n_distinct(component_number),
 			"n_distinct_BUSCO"=n_distinct(BUSCO_ID)
 		)
 n_components_with_BUSCO =
 	partition_map_global %>%
 		filter(BUSCO_ID != "") %>%
-		group_by(matrix,gene) %>%
+		group_by( matrix, component_number ) %>%
 		summarize(n()) %>%
 		group_by(matrix) %>%
 		tally(name="n_components_with_BUSCO")
 n_ribo = 
 	partition_map_global %>% 
-		group_by(matrix,gene) %>% 
+		group_by( matrix, component_number ) %>% 
 		tally(ribo_found) %>% 
 		group_by(matrix) %>% 
 		tally()

### 2019-09-12 manuscript/manuscript_kernel.R
added chain2 for ryan2013_est_only

@@ -213,17 +213,15 @@ file_names_iqtree = file_names_iqtree[ ! grepl("Philippe2009", file_names_iqtree
 file_names_iqtree = file_names_iqtree[ ! grepl("Nosenko2013", file_names_iqtree) ]
 
 # read iqtrees
-trees = foreach( tree_file = file_names_iqtree ) %dopar%
+trees_iq = foreach( tree_file = file_names_iqtree ) %dopar%
 	parse_tree_iqtree( tree_file, taxonomy_reference )
 
 # read pb trees
-append( trees, 
-        foreach( tree_file = file_names_pb ) %dopar%
+trees_pb = foreach( tree_file = file_names_pb ) %dopar%
            parse_tree_pb( tree_file, taxonomy_reference )
-)
 
 analyses_new = lapply(
-	trees,
+	c( trees_iq, trees_pb ),
 	function( tree ){
 		data.frame(
 			matrix = tree$matrix,
@@ -238,7 +236,7 @@ analyses_new = lapply(
 ) %>% 
 	bind_rows()
 
-analyses_new$inference = rep( "ML", nrow( analyses_new ) )
+analyses_new$inference = c( rep( "ML", length( trees_iq ) ), rep( "MCMC", length( trees_pb ) ))
 
 # Summarize result
 analyses_new$result = "Unresolved"
@@ -285,7 +283,7 @@ discarded_parts =
 
 partition_network_summary = 
 	n_total_partitions%>%
-	left_join(n_compandBUSCO,          by="matrix") %>%
+	# left_join(n_compandBUSCO,          by="matrix") %>%
 	left_join(n_components_with_BUSCO, by="matrix") %>%
 	left_join(n_ribo,                  by="matrix") %>%
 	left_join(discarded_parts,         by="matrix") %>% 

### 2019-09-12 manuscript/manuscript.rmd
A fresh knittr-ing of the manuscript
add iqtree results from choanozoa as outgroup
new bayesian trees working now

@@
 output:
-  github_document: default
   html_document: default
   pdf_document: default
+  github_document: default
 bibliography: manuscript.bib
@@
 analyses_published %>% 
-	ggplot( aes(x=model_summary, y=clade, col=result) ) +
+	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
@@
 analyses_new %>% 
-	ggplot( aes(x=model_summary, y=clade, col=result, shape=modelfinder) ) +
+	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +

### 2019-09-12 manuscript/functions.R

@@ -323,7 +323,7 @@ parse_tree_pb = function( tree_file, taxonomy_reference ){
 	  ":") %>% unlist()
 	burnins = as.numeric(burnin_line[2])
 	
-	keep_every = 10
+	keep_every = 100
 	treelist_text = c( unlist( mapply( read_pb_treelist, pb_treelist, burnins, keep_every ) ) )
 	sample_trees = read.tree( text = treelist_text )
 	

### 2019-09-12 manuscript/manuscript_kernel.R

@@ -236,7 +236,7 @@ analyses_new = lapply(
 ) %>% 
 	bind_rows()
 
-analyses_new$inference = c( rep( "ML", length( trees_iq ) ), rep( "MCMC", length( trees_pb ) ))
+analyses_new$inference = c( rep( "ML", length( trees_iq ) ), rep( "Bayesian", length( trees_pb ) ))
 
 # Summarize result
 analyses_new$result = "Unresolved"
@@ -248,7 +248,7 @@ analyses_new$result = factor( analyses_new$result )
 
 # Parse model components
 analyses_new$model_summary = analyses_new$model
-analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG+C60+F+G", "LG+C60+F+G", "poisson_C60", "GTR20", "WAG", "LG") )
+analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60+F+G", "CAT+F81")) #possibly combine WAG+C60+F+G and LG+C60+F+G
 
 # Partition comparison across matrices
 

### 2019-09-13 manuscript/functions.R
kernel re-ran, mscript knittrd

@@ -245,8 +245,8 @@ parse_tree = 	function( tree_file,  taxonomy_reference ){
 	
 	# Reroot the tree
 	# Use a single taxon from the outgroup since the outgroup is not guaranteed to be monophyletic
-	tree = root( tree, outgroup[1] )
-
+	tree = root( tree, outgroup[1], resolve.root=TRUE )
+	tree$outgroup = outgroup[1]
 	# Get the clades of the tips again, since tip order has changed
 	clades = 
 		taxonomy_reference$clade_assignment[ 

### 2019-09-13 manuscript/manuscript_kernel.R

@@ -248,7 +248,7 @@ analyses_new$result = factor( analyses_new$result )
 
 # Parse model components
 analyses_new$model_summary = analyses_new$model
-analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60+F+G", "CAT+F81")) #possibly combine WAG+C60+F+G and LG+C60+F+G
+analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60+F+G", "CAT+F81"))
 
 # Partition comparison across matrices
 

### 2019-10-10 manuscript/manuscript.rmd
kernel re-ran, mscript knittrd
angle x axis labels

@@
 output:
-  html_document: default
   pdf_document: default
   github_document: default
+  html_document: default
 bibliography: manuscript.bib
@@
 		scale_colour_manual(values = result_colors) +
-		theme_classic()
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
 ```

### 2019-10-17 manuscript/functions.R
use components to compute "gene" overlap

@@ -147,7 +147,7 @@ stat_clades = function(mapping = NULL, data = NULL, geom = "rect",
 
 compute_matrix_overlap = function( msa1, msa2 ){
 	
-	n_gene_overlap = sum( na.omit(msa1@partitions$gene) %in% na.omit(msa2@partitions$gene) )
+	n_gene_overlap = sum( na.omit(msa1@partitions$component_number) %in% na.omit(msa2@partitions$component_number) )
 
 	n_species_overlap = sum( rownames(msa1) %in% rownames(msa2) )
 	

### 2019-10-17 manuscript/functions.R
extract manuscript year to allow for year-sorting

@@ -113,6 +113,7 @@ clade_rects = function( data ){
 	D$ymin = c(0, clade_cumsums[1:length(clade_cumsums)-1])
 	D$ymax = clade_cumsums
 	D$manuscript = data@manuscript_name
+	D$year = as.numeric(str_extract(data@manuscript_name, '\\d\\d\\d\\d'))
 	D$matrix = data@matrix_name
 	D$manuscript_matrix = paste(D$manuscript, D$matrix, sep="_")
 	

### 2019-10-17 manuscript/manuscript.rmd
update rdata, reknitr'd with all new analyses working (I think)
another pass at taxon_rectangles figure

@@
 output:
+  github_document: default
   pdf_document: default
-  github_document: default
   html_document: default
@@
   ) %>%
-bind_rows()
-
-clade_colors = viridis( length(clades) )
-names( clade_colors ) = clades
+  bind_rows() %>%
+  filter(!grepl('only', matrix)) %>%
+  mutate(matrix = str_replace_all(matrix, "_", " "))  %>%
+  arrange(year, matrix)
 
 ggplot(data=taxon_rectangles) + 
-	scale_x_continuous(name="x") + 
-	scale_y_continuous(name="y") +
+	scale_x_continuous(name="Genes Sampled") + 
+	scale_y_continuous(name="Taxa Sampled") +
 	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
-	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
-	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
-	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
-	facet_wrap( ~ manuscript_matrix )
+	scale_fill_brewer(type="qual", direction=-1, palette="Set3", guide = guide_legend(reverse=TRUE)) +
+	facet_wrap( ~ matrix, labeller = label_wrap_gen(width = 18, multi_line = TRUE)) + 
+  theme_void() +
+  theme(
+    strip.background = element_rect(fill = "lightgrey"),
+    strip.text.x = element_text(size = 8),
+    panel.background = element_rect(fill ="#f9f9f9"),
+    # panel.grid.major = element_blank(),
+    # panel.grid.minor = element_blank()
+  )
 

### 2019-10-17 manuscript/manuscript.rmd
working on matrix_overlap figure

@@
 	dpi=300,
-	cache=FALSE
+	cache=FALSE,
+	fig.width=8
 	)
@@
 # A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
-cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
 
@@
 ```{r matrix_overlap}
-
+sequence_matrices_nosub = sequence_matrices[sapply(sequence_matrices, function(x) !grepl('only|subsam', x@matrix_name))]
 matrix_overlap_rectangles = 
-	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) overlap_rects(x,y))) %>% 
+	lapply(sequence_matrices_nosub, function(x) lapply(sequence_matrices_nosub, function(y) overlap_rects(x,y))) %>% 
 	unlist(recursive=FALSE) %>% 
@@
 	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) + 
-	facet_grid( matrix_1 ~ matrix_2 )
+	facet_grid( matrix_1 ~ matrix_2 ) +
+  theme_void() +
+  theme(
+    strip.background = element_rect(fill = "lightgrey"),
+    strip.text.x = element_text(size = 8),
+    panel.background = element_rect(fill ="#f9f9f9"),
+  )
 

### 2019-10-22 manuscript/manuscript.rmd
update manuscript

--- 2019-10-10 manuscript/manuscript.rmd
@@
 
-Casey W. Dunn^1^*, Benjamin Evans^2^, and Yuanning Li^1^
+Yuanning Li^1^, Xingxing Shen, Benjamin Evans^2^, Antonis Rokas, and Casey W. Dunn^1^*
 
@@
 The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
+
+- C10 to C60 [@si2008empirical]. 10 to 60-profile mixture models as variants of the CAT model under the maximum-likelihood framework.
+
+- Six-state amino acid recoding.
+Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
+Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding.
 
@@
 
-### Data wrangling
+### Data selection and wrangling
 
@@
 ### Phylogenetic analyses
+#### Full data matrices
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 16 data matrices from recent studies that were constructed from transcriptomic or genomic data (Table).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal (e.g.
+site heterogeneous CAT model and site homogeneous model), we first investigated four different sets of substitution models in IQtree [@nguyen2014iq], including C10 to C60 profile mixture models as variants of the CAT model in ML framework.
+Phylogenomic analyses of all the datasets were first conducted under WAG+G, GTR+G, associated best-fit substitution model (C10-C60 model were included for model comparison via -madd option) with ModelFinder [@kalyaanamoorthy2017modelfinder] and Poisson + C60 models using IQtree v [@nguyen2014iq].
+Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
+
+#### Outgroup taxa sampling with C10-C60 and CAT models
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
+These datasets include the same set of genes but differ in the composition of outgroup species.
+Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
+For each Choanozoa data matrice, both C10-C60 models in ML and CAT models in PhyloBayes were conducted.
+The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
+Moreover, bayesian inference with the site-heterogeneous CAT-Poisson substitution model was done with PhyloBayes MPI.
+To minimize computational burden, CAT-GTR models were not in this study.
+
+For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with CAT-Poisson model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For CAT-poisson analyses on all the data matrices, two independent chains were sampled every generation.
+Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
+Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
+A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+All analyses in PhyloBayes were run for at least one month computational time.
+
+#### phylogenetic signal in conflicted dataset
+To investigate the distribution of phylogenetic signal in data matrices, we considered four data matrices from four studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanozoa and Opisthokonta).
+For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above.
+The two different
+phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding
+tree that was modified to have sponges as the sister to all other metazoans.
+The constraint trees were conducted by R package.
+The numbers of genes and sites supporting each hypothesis were calculated with RAxML output and Perl scripts from Shen et al.
+For BI analysis, we only considered the Whelan_2017 dataset.
 
@@
 
-We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam cluser.
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam cluster.
 
@@
 It was the first phylogenomic analysis that included ctenophores, and therefore that could test the relationships of both Ctenophora and Porifera to the rest of animals.
-It was the first phylogenetic analysis to recover Ctenophora as the sister group to all other animals.
+It was also the first phylogenetic analysis to recover Ctenophora as the sister group to all other animals.
 
@@
 
-Philippe *et al.* 2009 [Philippe:2009hh]...
+Philippe *et al.* 2009 [Philippe:2009hh] assembled a 128 EST dataset for 55 species to explore phylogenetic relationship of early diverging animals.
 

### 2019-10-24 manuscript/manuscript_kernel.R
update R script

@@ -58,7 +58,7 @@ posterior_prob_threshold = 95
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
 analyses_published = 
-	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnnccc' ) %>%
+	read.delim( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnnccc' ) %>%
 	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
 
 taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
@@ -78,11 +78,7 @@ analyses_published$result[ ( analyses_published$inference == "ML" ) &       (ana
 analyses_published$result[ ( analyses_published$inference == "ML" ) &       (analyses_published$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
 analyses_published$result = factor( analyses_published$result )
 
-analyses_published$model_summary = analyses_published$model_rate_matrix
-long_summary = paste(analyses_published$model_rate_matrix, analyses_published$model_equilibrium, sep="+")
-analyses_published$model_summary[ ! is.na(analyses_published$model_equilibrium) ] = long_summary[ ! is.na(analyses_published$model_equilibrium) ]
-analyses_published$model_summary[ grepl("WAG", analyses_published$model_summary) ] = "WAG" # Simplify WAG
-analyses_published$model_summary = factor( analyses_published$model_summary, levels=c("GTR+CAT", "F81+CAT", "LG", "WAG") )
+analyses_published$model_combined = factor( analyses_published$model_combined, levels=c("GTR + CAT", "F81 + CAT", "Recoding + GTR + CAT", "Recoding + GTR", "Partitionfinder", "GTR", "LG", "WAG" ) )
 
 
 # Matrix taxon composition

### 2019-10-24 manuscript/manuscript.rmd
update philippe20019_choanoza_LG tree
update R script

@@
 output:
+  pdf_document: default
   github_document: default
-  pdf_document: default
   html_document: default
@@
 analyses_published %>%
-	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
+	ggplot( aes(x=model_combined, y=clade, col=result, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +

### 2019-10-24 manuscript/manuscript_kernel.R
re-run manuscript kernel

@@ -17,7 +17,7 @@ library( magrittr )
 library( igraph )
 library( doParallel )
 
-# setwd("/animal_root/manuscript")
+setwd("/animal_root/manuscript")
 
 source( "functions.R" )
 
@@ -58,7 +58,7 @@ posterior_prob_threshold = 95
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
 analyses_published = 
-	read.delim( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'iccccccccccclnnnnccc' ) %>%
+	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'icccccccccccclccccccc' ) %>%
 	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
 
 taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")

### 2019-10-24 manuscript/manuscript.rmd
fix published analyses axis labels

--- 2019-10-22 manuscript/manuscript.rmd
@@
 analyses_published %>%
-	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
+	ggplot( aes(x=model_combined, y=clade, col=result, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
 		scale_colour_manual(values = result_colors) +
-		theme_classic()
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
 

### 2019-10-30 manuscript/manuscript.rmd
update manuscript

@@
 
-Over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses.
+Over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
 Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
 In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
+An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occured early in animal evolution, espeically for complex structures such as cell types, nervous systems, and developmental patterns.
+However, a consensus has not been reached despite multiple efforts.
 
@@
 
-Philippe *et al.* 2009 [Philippe:2009hh] assembled a 128 EST dataset for 55 species to explore phylogenetic relationship of early diverging animals.
+Philippe *et al.* 2009 [Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals.
 
 ### Hejnol et al. 2009
+
+Hejnol *et al.* 2009 added
+
+### Ryan *et al.* 2013
+Ryan *et al.* 2013 sequenced the first ctenophore genome of *Mnemiopsis leidyi*.
+With the genome resources of *M.
+leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals.
+They analyzed both matrices by site-homogeneous GTR+Gamma model and Poisson-CAT model with three sets of outgroup sampling to evaluate the effect of outgoup selection to the ingroup topology.
+The results strongly supported Ctenophora-sister in all datasets they analysed using site-homogeneous model.
+The Poisson + CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days.
+
+### Whelan et al. 2015
+Wehlan *et al.* 2015 constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
+long-branch attraction, compositional bias, fast evolving genes, etc.).
+Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (~ 1000 orthologs) and subsequetly removal of genes with too much missing data and potential paralogs.
+The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups.
+All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister.
+CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora.
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+> Importantly, biases resulting from elevated compositional heterogeneity or elevated substitution rates are ruled out. Placement of ctenophores as
+sister to all other animals, and sponge monophyly, are strongly supported under multiple analyses, herein.
+
+### Pisani *et al.* 2015
+
+Pisani *et al.* 2015 reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013, Moroz2014 and Whelan2015 datasets.
+It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of animal-root position.
+The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction.
+They found strong support of porifera-sister when only include Choanozoa as outgroup in these datasets using site-heterogeous CAT model.
+Regarding the recovery of Porifera-sister, the authors concluded:
+>Our results reinforce a traditional scenario for the evolution of complexity in animals, and indicate that inferences about the evolution of Metazoa
+based on the Ctenophora-sister hypothesis are not supported by the currently available data.
+
+### Feuda *et al.* 2017
+
+Feuda *et al.* didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets).
+It was the first study phylogenomic analysis using recoding strategy to explore animal-root position.
+The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, PartitionFinder)and site-heterogeneous (CAT-GTR) models with non-recoding and recoding datasets.
+The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis.
+Regarding the recovery of Porifera-sister, the authors concluded:
+
+> Because adequate modeling of the evolutionary process that generated the data is fundamental to recovering an accurate phylogeny, our results strongly support sponges as the sister group of all other animals and provide further evidence that Ctenophora-sister represents a tree reconstruction artifact.
+
+### Whelan and Halalnych 2017a
+
+Whelan *et al.* is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework.
+The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets.
+More importantly, the authors also showed that both Poisson + CAT and GTR +  CAT models could overestimated substitutional heterogeneity in many cases.
+They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneious model.
+The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.
+The authors also concluded:
+
+> Practices such as removing constant sites and parsimony uninformative characters, or using CAT-F81 when CAT-GTR is deemed too computationally expensive,
+cannot be logically justified. Given clear problems with CAT-F81, phylogenies previously inferred with this model should be reassessed.
+
+### Whelan *et al.* 2017b
+
+Whelan *et al.* added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores.
+It significantly increased ctenophore taxon sampling than other studies.
+Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes.
+The subsequent filtering strategy was also similar to the previous study.
+All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in Choanoazoa dataset.
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+>Using datasets with reasonably high ctenophore and other non-bilaterian taxon sampling, our results strongly reject the hypothesis that sponges are the sister lineage to all other extant metazoans.
+
+### Simion *et al.* 2017
+
+Simion *et al.* [@Dunn:2008ky] added transcriptomic data for 21 new animals.
+The data matrix was constructed using a semi-automated approach to comprehensively detect and eliminate potential systematic errors.
+The resulting dataset comprises 1,719 genes and 97 species, including 61 non-bilaterian metazoan species.
+It was by far the largest phylogenomic dataset in terms of taxon and gene sampling related to animal-root question.
+
+The supermatrix was first analyzed using the F81-CAT model.
+Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
+Each jackknife is baed on a random selection of ~ 25% of the genes.
+The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in the supermatrix dataset.
+Importantly, the authors compared the behavior of long-branch effect between site-heterogenious and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+> Our dataset outperforms previous metazoan gene superalignments in terms of data quality and quantity.
+Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.
 

### 2019-10-31 manuscript/manuscript.rmd
update manuscript

@@
 Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
-Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding.
+Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
 
@@
 ### Gene sampling
+Gene sampling has no doubt
 
@@
 		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
-
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
+   xlab("model complexity")
 ```
@@
 Philippe *et al.* 2009 [Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals.
+The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
+The authors concluded:
+> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.
 
@@
 
-Hejnol *et al.* 2009 added
+Hejnol *et al.* 2009 added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix.
+The orthology inference was largely simialr to Dunn et al.
+2008.
+Two datsets were constructed, one with 844 genes and the other with 330 genes.
+Maximum likelihood analyses from both datasets strongly supported Ctenophora-sister.
+
+### Pick *et al.* 2010
+
+Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling.
+They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky].
+The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*.
+They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
+They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
+
+All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping.
+All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
+Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
+With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
+This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.
+
+The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected".
+Although the posterior probabilities supporting Porifera-sister are not strong, they conclude:
+
+> Results of our analyses indicate that sponges are the sister group to the remaining Metazoa, and Placozoa are sister group to the Bilateria
+
+They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
 
@@
 leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals.
-They analyzed both matrices by site-homogeneous GTR+Gamma model and Poisson-CAT model with three sets of outgroup sampling to evaluate the effect of outgoup selection to the ingroup topology.
+They analyzed both matrices by site-homogeneous GTR+Gamma model and Poisson-CAT model with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology.
 The results strongly supported Ctenophora-sister in all datasets they analysed using site-homogeneous model.
 The Poisson + CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days.
+
+### Moroz *et al.* 2014
+
+Moroz *et al.* 2014 sequenced the second ctenophore genome *Pleurobrachia bachei* to explore the phylogenetic relationship of Metazoa.
+All phylogenetic analyses strongly supported Ctenophora-sister with different taxon and gene sampling using WAG site-homogeneous model.
 
@@
 sister to all other animals, and sponge monophyly, are strongly supported under multiple analyses, herein.
+
+### Chang et al. 2015
+
+Chang *et al.* 2015 was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
+The authors constructed a dataset with 200 protein markers based on Philippe et al.
+2011 with 51,940 amino acids and 77 taxa.
+Both site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.
 
@@
 
-The supermatrix was first analyzed using the F81-CAT model.
+The supermatrix was first analyzed using the Poisson-CAT model.
 Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
@@
 Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.
-
-### Pick *et al.* 2010
-
-Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling.
-They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky].
-The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*.
-They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
-They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
-
-All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping.
-All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
-Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
-With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
-This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.
-
-The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected".
-Although the posterior probabilities supporting Porifera-sister are not strong, they conclude:
-
-> Results of our analyses indicate that sponges are the sister group to the remaining Metazoa, and Placozoa are sister group to the Bilateria
-
-They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
-Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
 

### 2019-10-31 manuscript/manuscript_kernel.R

@@ -1,11 +1,11 @@
-# The computationally intensive analyses for the manuscript 
+# The computationally intensive analyses for the manuscript
 # presented at https://github.com/caseywdunn/animal_root/tree/master/manuscript .
 #
-# Executing this code generates manuscript.RData, which contains analysis 
-# results. That file is then read by manuscript.rmd for rendering and 
+# Executing this code generates manuscript.RData, which contains analysis
+# results. That file is then read by manuscript.rmd for rendering and
 # presentation of the results.
-# 
-# The code presented here is roughly in the order of the analyses presented 
+#
+# The code presented here is roughly in the order of the analyses presented
 # in the manuscript, though there are exceptions
 
 ## Preliminaries
@@ -31,8 +31,8 @@ if ( cores < 1 ) {
 registerDoParallel( cores )
 
 # Set up constants
-focal_matrices = 
-	read.table(text = 
+focal_matrices =
+	read.table(text =
 						 	"manuscript matrix
 						 Borowiec2015	Total1080
 						 Chang2015 Chang2015
@@ -46,8 +46,8 @@ focal_matrices =
 						 Ryan2013  genome.opisthokonta
 						 Simion2017	supermatrix_97sp_401632pos_1719genes
 						 Whelan2015	Metazoa_Choano_RCFV_strict
-						 ", 
-						 header = TRUE, 
+						 ",
+						 header = TRUE,
 						 stringsAsFactors = FALSE)
 
 bootstrap_threshold = 90
@@ -57,16 +57,16 @@ posterior_prob_threshold = 95
 
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
-analyses_published = 
+analyses_published =
 	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'icccccccccccclccccccc' ) %>%
 	mutate( clade = factor( clade, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
 
 taxonomy_reference = read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv")
 
 # raw columns:
[…]

### 2019-10-31 manuscript/manuscript.rmd
keep updating

@@
 
-Over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
-Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
+Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
+However, over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
+Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested since .
 In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occured early in animal evolution, espeically for complex structures such as cell types, nervous systems, and developmental patterns.
+An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for complex structures such as cell types, nervous systems, and developmental patterns.
 However, a consensus has not been reached despite multiple efforts.
@@
 Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
-We hope that this provides an integrative overview of the challenge and provides direction for future studies.
-We also hope that the work we have done here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
+We hope that this effort provides an integrative overview of the challenge and provides direction for future studies.
+We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
 
@@
 (D) A sponge.
+(E)
 
@@
 ### Gene sampling
-Gene sampling has no doubt
 
-### Outgroup taxon sampling
+Whole-genome and transcriptomic sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes.
+Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be ultraconserved across animal genomes and most data matrices should share a core set of orthologs.
+Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for orthology determination.
 
-XXX
+### Outgroup and ingroup taxon sampling
+
+Outgroup selection (the taxa used to root the ingroup animals) can potentially affect phylogenetic inference of the root position of animal phylogeny.
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology.
+Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+It has been suggested the inclusion of more distant related (e.g.
+Opisthokonta) outgroup taxa tend to recover Ctenophora-sister using both site-homogeneous and site-heterogeneous models, whereas high support of Porifera-sister often recovered when site-heterogeneous models were used.
+In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling.
+This may be because results have tended to be more sensitive to outgroup sampling.
 
@@
 
-Choanimalia, Holozoa, Opisthokonta
+### Data-recoding
 
-### Ingroup taxon sampling
-
-Sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.
+Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the position of animal root.
+Feuda et al.
+2017 has proposed to recode the full set of twenty amino acids into six groups of amino aicds that have more frequent evolutionary changes within groups than between groups based on different substitution models.
+By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
+By reanalyzing Chang et al.
+2015 and Whelan et al.
+2015, they found that Porifera sister was favored under all recoding strategies using GTR-CAT models.
+However, the performance of the recoding methods has never been tested empirically against non-recoding method in the root of the animal phylogeny, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies.
 
@@
 
-ggplot(data=taxon_rectangles) +
+filtered_taxon_rectangles <- taxon_rectangles[!grepl("only", taxon_rectangles$matrix),]
+
+ggplot(data=filtered_taxon_rectangles) +
 	scale_x_continuous(name="x") +
@@
 	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
-	facet_wrap( ~ manuscript_matrix )
+	facet_wrap( ~ matrix )
 
 ```
-
-XXX
 
@@
 
-ggplot( matrix_summary ) +
+p1 = ggplot( matrix_summary ) +
 	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
@@
 
+matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "NA", "Whelan2017_full", "Whelan2017_strict") )
+
+p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
 ```
@@
 One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
-Here we reanalyze the primary matrices from each study under consistent conditions with iqtree.
-We selected this tool because it has greater model flexibility than other tools and is very fast.
+Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood framework.
 
-We first tested a variety of models for each matrix, and inferred support under the selected model. We then analyzed every matrix under a panel of standard models, including XXX.
+We first tested a variety of models for each matrix, and inferred support under the best-fit model from IQtree.
+We then analyzed every matrix under a panel of standard models, including WAG+G, GTR+G, Poisson+G.
 
@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-
-## New analyses of new matrices
-
-Based on the variation across analyses, we constructed new matrices with altered taxon and gene sampling to test specific hypotheses about differences in support.
 
@@
 
-The count for a partition pair can be much arger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.
+The count for a partition pair can be much larger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.
 
@@
 A gene that is perfectly sampled would form a cluster with this size.
-Very few clusters, though, are this size.
-This suggests that intersection of genes between matrices is low
+To our surprise, very few clusters, though, are this size.
+This suggests that intersection of genes between matrices is extreme low.
+This finding suggested genes used for exploring the position of the root of the animal phylogeny
 

### 2019-11-01 manuscript/manuscript.rmd
update manuscript

@@
 Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for orthology determination.
+It should be noted that some earlier studies including
 
@@
 Feuda et al.
-2017 has proposed to recode the full set of twenty amino acids into six groups of amino aicds that have more frequent evolutionary changes within groups than between groups based on different substitution models.
+2017 has proposed to recode the full set of twenty amino acids into six groups of amino aicds that have more frequent evolutionary changes within groups than between groups based on different exchangeability matrices.
 By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
@@
 However, the performance of the recoding methods has never been tested empirically against non-recoding method in the root of the animal phylogeny, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies.
+Thus, it is important to evaluate the performance of data recoding method verses non-recoding approaches in empritical animal-root data matrices.
 
@@
 ### Matrix taxon composition
+
+we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies.
+For taxon composition, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages.
+These skewed data matrices includes two studies constructed from EST dataset, such as Dunn2008 (), Hejnol2009.
 

### 2019-11-04 manuscript/manuscript.rmd
update manuscript

@@
 An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for complex structures such as cell types, nervous systems, and developmental patterns.
-However, a consensus has not been reached despite multiple efforts.
+However, a wide consensus of animal root position has not been reached despite multiple efforts.
 
@@
 Whole-genome and transcriptomic sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes.
-Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be ultraconserved across animal genomes and most data matrices should share a core set of orthologs.
-Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for orthology determination.
-It should be noted that some earlier studies including
+Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthologs.
+Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for determination of orthology genes (OGs).
 
@@
 
-Outgroup selection (the taxa used to root the ingroup animals) can potentially affect phylogenetic inference of the root position of animal phylogeny.
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology.
+Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect phylogenetic inference of the root position of animal phylogeny.
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology .
 Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
@@
 
-we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies.
+We synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies.
 For taxon composition, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages.
-These skewed data matrices includes two studies constructed from EST dataset, such as Dunn2008 (), Hejnol2009.
+These skewed data matrices includes two studies constructed from EST dataset and one study from the sequenced whole animal genome.
+The rest data matrices have relative balanced taxon composition, although the the number of sampled taxon varies a lot.
+Since increase taxon sampling is broadly accepted to improve phylogenetic inference an aid in the placement of animal root position, it should be noted that Whelan2017 and Simion2017 have much broader taxon sampling than other data matrices, with Simion2017 has the most taxon sampling and Whelan2017 contains greater ctenophore taxon sampling.
 
@@
 
+Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes to tackle the incongruence of the topology among individual gene trees.
+The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
+However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring].
+Thus, it is important to evaluate the gene composition and matrix overlap across difference data matrices to explore.
+
+We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected.
+The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely present in the common ancestor of animal genome, and are commonly used to assess the completeness of animal genome or transcriptomic assembly and frequently used as phylogenetic markers.
+Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most data matrices given their nature of conserveness and high-expression in transcriptomes.
+However, to our surprise, we found that OGs used in all data matrices comprised less than 50%, even with several data matrices contain less than 20 percent of BUSCO genes.
+
+We furthur
+
 ```{r gene_composition}
@@
 p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
+count =  busco_results %>% select( matrix,  Busco_Id ) %>% distinct(  )
+reference_busco=read_tsv("/animal_root/manuscript/busco.tsv")
+
+Simion2017_busco <- count[grep("Simion2017", count$matrix), ] %>% select(Busco_Id)
+names(Simion2017_busco)[1]<-"Simion2017"
+
 ```
@@
 ### Matrix overlap
+
+Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has never been investigated.
+Our initial intent of comparing matrix overlap across matrices was to constructed new matrices with shared taxon and gene sampling to test specific hypotheses about differences in support.
+The hypothesis is that although the bioinformatics pipeline used for OG determination is differt, most larger data matrices should share a core set of OGs.
+However, our results show that surprisingly little overlap were identified across all the data matrices from different studies (Fig XXAlignment overlap).
 
@@
 
+Here we summarized all the previous phylogenetic analyses from all studies and the results are shown in Fig.
+support_published_analyses and Table perviouly_published_analyses.tsv.
+We found that all phylogenomic analyses strongly support of Ctenophora-sister once site-homogeneous model were used, no matter what outgroup choice, phylogenetic inference, recoding method, data partitioning are used.
+We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups and the use of site-heterogeneious CAT models, although many of analyses are not converged or posterior probability is relatively low using PhyloBayes.
+It should be noted that most support of Porifera-sister are from analyses using Poisson + CAT and recoding methods with GTR + CAT models.
+
 ```{r support_published_analyses}
@@
 
-A total of `r nrow(analyses_published)` analyses were transcribed from the literature.
+**Fig support_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table perviouly_published_analyses.tsv).
 
@@
 # Investigate connected components of genes (ie partitions) that include multiple members fromt he same matrix
-clusterstats_within_manuscripts = partition_map_global %>% group_by( matrix, component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
-clusterstats = partition_map_global %>% group_by( component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
+#clusterstats_within_manuscripts = partition_map_global %>% group_by( matrix, component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
+#clusterstats = partition_map_global %>% group_by( component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
 
-clusterstats_within_manuscripts %>% print(n=100)
-clusterstats %>% print(n=100)
-partition_network_summary %>% print()
+#clusterstats_within_manuscripts %>% print(n=100)
+#clusterstats %>% print(n=100)
+#partition_network_summary %>% print()
 

### 2019-11-06 manuscript/manuscript.rmd
update manuscript

@@
 Whole-genome and transcriptomic sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes.
-Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthologs.
-Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for determination of orthology genes (OGs).
+Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthology genes (OGs).
+Surprisingly, gene composition and matrice overlap across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for determination of OGs.
 
@@
 We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups and the use of site-heterogeneious CAT models, although many of analyses are not converged or posterior probability is relatively low using PhyloBayes.
-It should be noted that most support of Porifera-sister are from analyses using Poisson + CAT and recoding methods with GTR + CAT models.
+It should be noted that most support of Porifera-sister are from analyses using Poisson + CAT and recoding methods with site-heterogeneous models.
+Moreover, only a few analyses utilizing the most complex GTR + CAT model are converged and have mixed support for Porifera-sister or Ctenophora-sister.
+The main reason the GTR + CAT model has only been employed in a few analyses is because it is computational untractable and can only be used in smaller or reduced data matrices.
+Interestingly, a CAT-like, site-heterogeneous C10 - C60 (C) models implemented in the maximum likelihood (ML) framework has never been tested and compared to other models.
+
+The use of site-heterogeneous CAT is an infinte mixture model has long been suggested as the best approach to account for systematic errors, particular for long-branch attraction and saturation.
+The CAT model often received the best-model fit from cross validation or posterior predictive analyses compared to site-homogeneous models with or without data partitioning.
+In contrast, a recent study has suggested that Poisson + CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in  cases can infer incorrect topology based on the simulation analyses.
+The author also raised the issue of the potential overparameterization in CAT model in some cases.
+Nevertheless，the debate over the site-heterogeneous vs.
+site-homogeneous models and/or inclusion or exclusion more distantly related outgroups, constitutes the probably one of the most controversial node in the animal tree of life.
 
@@
 
-We first tested a variety of models for each matrix, and inferred support under the best-fit model from IQtree.
-We then analyzed every matrix under a panel of standard models, including WAG+G, GTR+G, Poisson+G.
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree.
+We found in all cases, C60 + WAG/LG +F +G models fit these matrices bettern than the site-homogeneous models in under BIC criteria and inferred support under the best-fit model from IQtree.
+We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G.
+Moreover, we also used the the best-fit model identified above with the removal of all distantly related outgroups and only include Choanozoa as outgroup.
+With the exception of Moroz2014 matrix (), every analysis conducted herein with different models or outgroup choice strongly supported the ctenophore-sister hypothesis.
+
+It
 
@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
+For computational efficiency, we first only applied Poisson+CAT models in every matrice with only Choanoflagellata as outgroup.
+It should be noted that although most Phylobayes runs were converged, the convergence of several large matrices have not been reached after at least a month's running time.
+
+Three matrices with inclusion of Choanoflagellata as outgroup
 

### 2019-11-08 manuscript/manuscript.rmd
upload phylobayes results from choanozoa outgroup

@@
 Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthology genes (OGs).
-Surprisingly, gene composition and matrice overlap across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for determination of OGs.
+Surprisingly, gene composition and matrice overlap across different studies has been rarely assessed, since most studies relied on their own bioinformatic pipeline for determination of OGs.
 
@@
 
-The use of site-heterogeneous CAT is an infinte mixture model has long been suggested as the best approach to account for systematic errors, particular for long-branch attraction and saturation.
+The use of site-heterogeneous CAT is an infinite mixture model has long been suggested as the best approach to account for systematic errors, particular for long-branch attraction and saturation.
 The CAT model often received the best-model fit from cross validation or posterior predictive analyses compared to site-homogeneous models with or without data partitioning.
-In contrast, a recent study has suggested that Poisson + CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in  cases can infer incorrect topology based on the simulation analyses.
+In contrast, a recent study has suggested that Poisson + CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in some cases can infer incorrect topology based on the simulation analyses.
 The author also raised the issue of the potential overparameterization in CAT model in some cases.
@@
 Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood framework.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework.
 
 We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree.
-We found in all cases, C60 + WAG/LG +F +G models fit these matrices bettern than the site-homogeneous models in under BIC criteria and inferred support under the best-fit model from IQtree.
-We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G.
+We found in all cases, C60 + WAG/LG +F +G models fit these matrices better than the site-homogeneous models in under BIC criteria and inferred support under the best-fit model from IQtree.
+We didn't include GTR + C models since it needs to estimate over 10,000 free parameters in ModelFinder in some large datasets compared to several hundred in site-homogeneous + C models.
+then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G.
 Moreover, we also used the the best-fit model identified above with the removal of all distantly related outgroups and only include Choanozoa as outgroup.
-With the exception of Moroz2014 matrix (), every analysis conducted herein with different models or outgroup choice strongly supported the ctenophore-sister hypothesis.
 
-It
+With the exception of Moroz2014 matrix supported Placozoa-sister, every analysis conducted herein with different models or outgroup choice strongly supported the ctenophore-sister hypothesis under maximum likelihood framework in IQtree.
+
+The site-heterogeneous CAT model has been introduced in PhyloBayes and it is a non-parametric modeling based on a Dirichlet process mixture, which is often used in Bayesian studies (Lartillot and Philippe, 2004).
+It was designed to better accommodate data saturation or tackle long-branch attraction for deep phylogenetic question but only used in Bayesian inference (BI).
+In contrast, C models is 10 to 60-profile mixture models as variants of the CAT model can be used in both ML or BI framework.
+Thus, it is interesting to compare results between C model to CAT models since it has never been used in any matrices analyzed here.
 
@@
 For computational efficiency, we first only applied Poisson+CAT models in every matrice with only Choanoflagellata as outgroup.
-It should be noted that although most Phylobayes runs were converged, the convergence of several large matrices have not been reached after at least a month's running time.
+It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
 
-Three matrices with inclusion of Choanoflagellata as outgroup
+Interestingly, we found a mixed support when using Poisson-CAT model in Choanimalia matrices.
+Consistent with previous study, we recovered Porifera-sister in Philippe2008, Ryan2013, Nosenko2013 matrices with only.
+Moreover, we also found a strong support of Porifera-sister in that Whelan2017 matrices.
+However, with the exception of Moroz2013, we found strong support for Ctenophora-sister in all other matrices.
 

### 2019-11-12 manuscript/manuscript.rmd
update results in manuscript

--- 2019-11-04 manuscript/manuscript.rmd
@@
 ---
+bibliography: manuscript.bib
 csl: plos.csl
 output:
+  word_document: default
+  html_document: default
+  pdf_document: default
   github_document: default
-  pdf_document: default
-  html_document: default
-bibliography: manuscript.bib
 ---
@@
 
-Yuanning Li^1^, Xingxing Shen, Benjamin Evans^2^, Antonis Rokas, and Casey W. Dunn^1^*
+Yuanning Li^1,3^, Xingxing Shen^3^, Benjamin Evans^2^, Antonis Rokas^3^, and Casey W. Dunn^1^*
 
@@
 
+^3^Department of Biological Sceince, Vanderbilt University
 \* Corresponding author, casey.dunn@yale.edu
@@
 Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
-However, over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
-Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested since .
+However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
+Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
 In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
 An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for complex structures such as cell types, nervous systems, and developmental patterns.
-However, a wide consensus of animal root position has not been reached despite multiple efforts.
+However, a wide consensus of the position of animal has not been reached despite multiple efforts.
 
@@
 Testing these hypotheses requires evaluating the support for each of these alternative nodes.
-(C) A ctenophore.
-(D) A sponge.
-(E)
+(C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node).
+(D) A ctenophore.
+(E) A sponge.
 
@@
 While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
-There are couple approaches that have been used in the studies considered here:
+There are a couple approaches that have been used in the studies considered here:
 
@@
 
-- Empirical site homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
+- Empirical site-homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
 
-- Estimated site homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
+- Estimated site-homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
 
@@
 The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
+For example, on the questions considered here, F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
 LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
@@
 
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
 If there are no modifiers, then it is implied that site homogeneous models are used.
@@
 
-Whole-genome and transcriptomic sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes.
-Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthologs.
-Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for determination of orthology genes (OGs).
+High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]).
+Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes).
+Surprisingly, gene composition and matrice overlap across different studies has been rarely assessed, since most studies relied on their own bioinformatic pipeline for determination of OGs.
 
@@
 
-Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect phylogenetic inference of the root position of animal phylogeny.
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology .
+Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**).
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
 It has been suggested the inclusion of more distant related (e.g.
-Opisthokonta) outgroup taxa tend to recover Ctenophora-sister using both site-homogeneous and site-heterogeneous models, whereas high support of Porifera-sister often recovered when site-heterogeneous models were used.
+Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered only when site-heterogeneous CAT models were used.
 In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling.
@@
 
-**Fig XXOutgroup.** The animals and their outgroups, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
-
 ### Data-recoding
 
-Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the position of animal root.
+Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the animal root position.
 Feuda et al.
-2017 has proposed to recode the full set of twenty amino acids into six groups of amino aicds that have more frequent evolutionary changes within groups than between groups based on different exchangeability matrices.
+2017 [@feuda2017improved] has proposed to recode the full set of twenty amino acids into six groups that have more frequent evolutionary changes within groups than between groups based on different exchangeability matrices.
 By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
 By reanalyzing Chang et al.
-2015 and Whelan et al.
-2015, they found that Porifera sister was favored under all recoding strategies using GTR-CAT models.
-However, the performance of the recoding methods has never been tested empirically against non-recoding method in the root of the animal phylogeny, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies.
-Thus, it is important to evaluate the performance of data recoding method verses non-recoding approaches in empritical animal-root data matrices.
+2015 [@chang2015genomic] and Whelan et al.
+2015 [@whelan2015error], they found that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
+However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies [@hernandez2019six].
+Thus, it is important to evaluate the performance of data recoding method verses non-recoding approaches in past empritical matrices .
 
 ## Overview of published analyses
+Here, we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies (Table 1).
+
+Table 1.
+```{r overview data matrices}
+	kable(datasets)
+
+```
 
@@
 
-We synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies.
-For taxon composition, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages.
-These skewed data matrices includes two studies constructed from EST dataset and one study from the sequenced whole animal genome.
-The rest data matrices have relative balanced taxon composition, although the the number of sampled taxon varies a lot.
-Since increase taxon sampling is broadly accepted to improve phylogenetic inference an aid in the placement of animal root position, it should be noted that Whelan2017 and Simion2017 have much broader taxon sampling than other data matrices, with Simion2017 has the most taxon sampling and Whelan2017 contains greater ctenophore taxon sampling.
+Taxa sampling plays a critical role in phylogenetic inference, especially for the phylogenetic position that needs to be explored.
+However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
+These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Henjol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
+Importantly, some earlier studies only include less than five ctenophore species.
+The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
+In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position, and it should be noted that Whelan2017 [@whelan2017ctenophore] and Simion2017 [@simion2017large] have much broader taxon sampling than other data matrices: Simion2017 has the most taxon sampling, and Whelan2017 contains greater ctenophore taxon sampling.
 
@@
 **Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by taxon sampling.
-Horizontal size is proportional to the number of genes (XXOr should it be sites?) sampled, vertical size to the number of taxa sampled.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
 
-### Matrix gene composition
+### Matrix gene composition and overlap
 
@@
 However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring].
-Thus, it is important to evaluate the gene composition and matrix overlap across difference data matrices to explore.
+Thus, it is important to evaluate the gene composition and matrix overlap to explore congruence of gene sampling across different matrices.
+It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.
 
 We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected.
-The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely present in the common ancestor of animal genome, and are commonly used to assess the completeness of animal genome or transcriptomic assembly and frequently used as phylogenetic markers.
-Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most data matrices given their nature of conserveness and high-expression in transcriptomes.
-However, to our surprise, we found that OGs used in all data matrices comprised less than 50%, even with several data matrices contain less than 20 percent of BUSCO genes.
+The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies.
+Morever, the BUSCO dataset also are frequently used as phylogenomic markers in a varietly of studies.
+Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conserveness and high-expression properties.
+However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
+XXBUSCO_annotations).
 
-We furthur
+We furthered BLAST gene partitions to SwissProt database
+
+Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
+Our initial intent of comparing matrix overlap across matrices was to construct a new matrice with shared taxon and gene sampling to test specific hypotheses about differences in support.
+The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs.
+However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap).
+These results indicated that different studies essentially sampled different regions of the genomes with little overlap.
+One thing should be noted that
 
@@
 p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
-count =  busco_results %>% select( matrix,  Busco_Id ) %>% distinct(  )
-reference_busco=read_tsv("/animal_root/manuscript/busco.tsv")
-
-Simion2017_busco <- count[grep("Simion2017", count$matrix), ] %>% select(Busco_Id)
-names(Simion2017_busco)[1]<-"Simion2017"
 
@@
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-
-### Matrix overlap
-
-Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has never been investigated.
-Our initial intent of comparing matrix overlap across matrices was to constructed new matrices with shared taxon and gene sampling to test specific hypotheses about differences in support.
-The hypothesis is that although the bioinformatics pipeline used for OG determination is differt, most larger data matrices should share a core set of OGs.
-However, our results show that surprisingly little overlap were identified across all the data matrices from different studies (Fig XXAlignment overlap).
 
@@
 
-Here we summarized all the previous phylogenetic analyses from all studies and the results are shown in Fig.
-support_published_analyses and Table perviouly_published_analyses.tsv.
-We found that all phylogenomic analyses strongly support of Ctenophora-sister once site-homogeneous model were used, no matter what outgroup choice, phylogenetic inference, recoding method, data partitioning are used.
-We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups and the use of site-heterogeneious CAT models, although many of analyses are not converged or posterior probability is relatively low using PhyloBayes.
-It should be noted that most support of Porifera-sister are from analyses using Poisson + CAT and recoding methods with GTR + CAT models.
+Here we summarized all the 167 previous phylogenetic analyses from all studies (Fig.
+support_published_analyses and Table perviouly_published_analyses).
+We found that all past phylogenomic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference, recoding methods are used.
+We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT models.
+It should be noted that the convergence were not reached or posterior probability is not well supported in many BI analyses.
+Moreover, the strong support of Porifera-sister are mainly from analyses using Poisson+CAT and recoding methods with site-heterogeneous models.
+Moreover, only a few past analyses with the most complex GTR+CAT model are converged.
+However, for the ones that are converged with GTR+CAT model, Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister when choanozoa are used as outgroups.
+
+The use of site-heterogeneous CAT is an infinite mixture model has long been suggested as the best approach to account for systematic errors, particular for data saturation, especially in the case of long-branch attraction (LBA) (Lartillot and Philippe, 2004).
+The GTR-CAT model often received the best-model fit from cross validation or posterior predictive analyses compared to site-homogeneous or Poisson-CAT models.
+However, the GTR+CAT model is extremely computational demanding and not feasible in large matrices so many studies only include Poisson+CAT models in their analyses.
+In contrast, a recent study has suggested that Poisson+CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in some cases can infer incorrect topology based on the simulation analyses.
+The author also raised the issue of the potential overparameterization in CAT models in some cases.
+Nevertheless，the debate over the site-heterogeneous vs.
+site-homogeneous models and/or inclusion or exclusion more distantly related outgroups, constitutes the probably one of the most controversial nodes in the animal tree of life.
 
@@
 Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood framework.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework.
+Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models that are implemented in both ML framework in our analyses and it is interesting to compare results between C model to CAT models since it has never been used in any matrices analyzed here.
 
-We first tested a variety of models for each matrix, and inferred support under the best-fit model from IQtree.
-We then analyzed every matrix under a panel of standard models, including WAG+G, GTR+G, Poisson+G.
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree.
+We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
+It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive.
+The GTR+C models requires to estimate up to over 10,000 (e.g.
+GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models.
+We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G.
+Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
+
+With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister, every analysis conducted herein choice strongly supported the ctenophore-sister hypothesis, even with site-heterogeneous C models with the most closely related outgroups.
 
@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
+For computational efficiency, we first only applied Poisson+CAT models in every matrice with only Choanoflagellata as outgroup.
+It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.
+
+Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices.
+Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, and Whelan2015.
+Moreover, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrice of Simion2017.
+Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrice.
+For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model.
+Consistently with past study, we found strong support of Ctenophora-sister in all anaylzed Opisthokonta matrices, indicating that Ctenophora-sister is also favored in all matrices with more distantly related taxa are used as outgroups.
+
+###
 
@@
 
-All files associated with this analysis are available at https://github.com/caseywdunn/animal_root
+All files and bioinformatic commmands associated with this analysis are available at https://github.com/caseywdunn/animal_root
 
@@
 More importantly, the authors also showed that both Poisson + CAT and GTR +  CAT models could overestimated substitutional heterogeneity in many cases.
-They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneious model.
+They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site -homogeneious model.
 The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.

### 2019-11-12 manuscript/manuscript.rmd
revise methods

@@
 However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
-These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Henjol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
+These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
 Importantly, some earlier studies only include less than five ctenophore species.
@@
 ### Phylogenetic analyses
-#### Full data matrices
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 16 data matrices from recent studies that were constructed from transcriptomic or genomic data (Table).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal (e.g.
-site heterogeneous CAT model and site homogeneous model), we first investigated four different sets of substitution models in IQtree [@nguyen2014iq], including C10 to C60 profile mixture models as variants of the CAT model in ML framework.
-Phylogenomic analyses of all the datasets were first conducted under WAG+G, GTR+G, associated best-fit substitution model (C10-C60 model were included for model comparison via -madd option) with ModelFinder [@kalyaanamoorthy2017modelfinder] and Poisson + C60 models using IQtree v [@nguyen2014iq].
+
+#### Phylogenetic analyses in IQtree
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal tree of life (e.g.
+site-heterogeneous vs.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C models in the model testing.
+For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters.
+For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
+
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60 and associated best-fit model identified above.
 Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
+Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C model or CAT site-heterogeneous models.
+To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
+For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
+2009 [@hejnol2009assessing], since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+It should be noted that the main conclusion of Simion et al.
+2008 was also based on selection of 25% of genes for their jackknife approach.
 
@@
 These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanozoa data matrice, both C10-C60 models in ML and CAT models in PhyloBayes were conducted.
+Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
+For each Choanimalia data matrice, both C models IQtree and Poisson-CAT models in PhyloBayes were conducted.
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous CAT-Poisson substitution model was done with PhyloBayes MPI.
+Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI.
 To minimize computational burden, CAT-GTR models were not in this study.
 
-For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with CAT-Poisson model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
-For CAT-poisson analyses on all the data matrices, two independent chains were sampled every generation.
+For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
 Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
@@
 A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-All analyses in PhyloBayes were run for at least one month computational time.
+For those matrices that were not converged, PhyloByes analyses were run for at least one month.
 
 #### phylogenetic signal in conflicted dataset
-To investigate the distribution of phylogenetic signal in data matrices, we considered four data matrices from four studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanozoa and Opisthokonta).
+To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.
+We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
 For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above.
-The two different
-phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding
-tree that was modified to have sponges as the sister to all other metazoans.
-The constraint trees were conducted by R package.
-The numbers of genes and sites supporting each hypothesis were calculated with RAxML output and Perl scripts from Shen et al.
-For BI analysis, we only considered the Whelan_2017 dataset.
+The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
+The constraint trees were conducted in a customized R script.
+The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al.
+2017 [@shen2017contentious].
+For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
 

### 2019-11-14 manuscript/manuscript.rmd
update phylogenetic signal in manuscript_kernel and mnuascript

@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-For computational efficiency, we first only applied Poisson+CAT models in every matrice with only Choanoflagellata as outgroup.
+For computational efficiency, we first only applied Poisson+CAT models in every matrix with only Choanoflagellata as outgroup.
 It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.
@@
 Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, and Whelan2015.
-Moreover, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrice of Simion2017.
-Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrice.
+Moreover, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
+Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix.
 For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model.
@@
 
-###
-
+### Phylogenetic signal
+To further explore the effect of site-homogeneous or site-heterogeneous models and outgroup choices in on the animal-root position,  we next quantify the support of phylogenetic signal over two alternative hypotheses ()
 ## The current state of understanding
@@
 Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI.
-To minimize computational burden, CAT-GTR models were not in this study.
+To minimize computational burden, CAT-GTR models were only performed in the.
 

### 2019-11-14 manuscript/manuscript_kernel.R

@@ -284,6 +284,156 @@ partition_network_summary =
 	left_join(n_ribo,                  by="matrix") %>%
 	left_join(discarded_parts,         by="matrix") %>%
 	mutate(n_partitions_discarded = replace_na(n_partitions_discarded, 0))
+### Summarize phylogenetic signal by genes
+Philippe2009  <- read_tsv("../trees_new/AU_test/Philippe2009_C60_genewise_lnL.txt")
+Philippe2009$ABS <- abs(Philippe2009$`tr1_log-likelihood` - Philippe2009$`tr2_log-likelihood`)
+Philippe2009$result[ (Philippe2009$tree_supported == "tr1" ) & (Philippe2009$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
+Philippe2009$result[ (Philippe2009$tree_supported == "tr1" ) & (Philippe2009$ABS < 1) ] = "Ctenophore-sister-weak-signal"
+Philippe2009$result[ (Philippe2009$tree_supported == "tr2" ) & (Philippe2009$ABS >= 1) ] = "Porifera-sister-strong-signal"
+Philippe2009$result[ (Philippe2009$tree_supported == "tr2" ) & (Philippe2009$ABS <1) ] = "Porifera-sister-weak-signal"
+
+Philippe2009 %>%
+	count(result) %>%
+	mutate(perc = n/ nrow(Philippe2009)) -> Philippe2009_count
+Philippe2009_count$study <- 'Philippe2009_Opisthokonta_C60'
+
+
+Philippe2009_only_choanozoa  <- read_tsv("../trees_new/AU_test/Philippe2009_only_choanozoa_C60_genewise_lnL.txt")
+Philippe2009_only_choanozoa$ABS <- abs(Philippe2009_only_choanozoa$`tr1_log-likelihood` - Philippe2009_only_choanozoa$`tr2_log-likelihood`)
+Philippe2009_only_choanozoa$result[ (Philippe2009_only_choanozoa$tree_supported == "tr1" ) & (Philippe2009_only_choanozoa$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
[…]

### 2019-11-14 manuscript/functions.R
parse AU tests into one tibble instead of lots of separate objects

@@ -519,15 +519,21 @@ get_matrix_occupancy = function( seq_matrix ){
 #' @return A tibble with columns: gene_id tree_supported tr1_log.likelihood tr2_log.likelihood tr3_log.likelihood manuscript matrix. tr1 = ctenophora sister tr2 = porifera sister tr3 = cnidaria/ctenophora sister
 #' @export
 parse_au_gene_tests = function(path="../trees_new/AU_test"){
-  # tr1 = ctenophora sister tr2 = porifera sister tr3 = nideria/ctenophora sister
-  files_list = list.files(path, pattern = "genewise", full.names = TRUE, recursive = TRUE) 
-  filename_parts_list = strsplit( basename( files_list ), "_" )
-  
-  au_tibble = 
-    files_list %>%
-    map(read.delim) %>%
-    imap(~ transform(.x, manuscript = filename_parts_list[[.y]][[1]])) %>%
-    imap(~ transform(.x, matrix = filename_parts_list[[.y]][[2]])) %>%
-    bind_rows()
-  return(au_tibble)
+	files_list = list.files(path, pattern = "genewise", full.names = TRUE, recursive = TRUE) 
+	filename_parts_list = str_split(gsub('_genewise_lnL.txt', '', basename( files_list )), '_', n=2)
+
+	au_tibble = 
+		files_list %>%
+		map(read.delim) %>%
+		imap(~ transform(.x, manuscript = filename_parts_list[[.y]][[1]])) %>%
+		imap(~ transform(.x, matrix = filename_parts_list[[.y]][[2]])) %>%
+		bind_rows()
+
+	au_tibble$ABS <- abs(au_tibble$tr1_log.likelihood - au_tibble$tr2_log.likelihood)
+	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS < 1) ] = "Ctenophore-sister-weak-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS >= 1) ] = "Porifera-sister-strong-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS <1) ] = "Porifera-sister-weak-signal"
+	
+	return(au_tibble)
 }

### 2019-11-14 manuscript/manuscript_kernel.R

@@ -284,156 +284,9 @@ partition_network_summary =
 	left_join(n_ribo,                  by="matrix") %>%
 	left_join(discarded_parts,         by="matrix") %>%
 	mutate(n_partitions_discarded = replace_na(n_partitions_discarded, 0))
+
 ### Summarize phylogenetic signal by genes
-Philippe2009  <- read_tsv("../trees_new/AU_test/Philippe2009_C60_genewise_lnL.txt")
-Philippe2009$ABS <- abs(Philippe2009$`tr1_log-likelihood` - Philippe2009$`tr2_log-likelihood`)
-Philippe2009$result[ (Philippe2009$tree_supported == "tr1" ) & (Philippe2009$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
-Philippe2009$result[ (Philippe2009$tree_supported == "tr1" ) & (Philippe2009$ABS < 1) ] = "Ctenophore-sister-weak-signal"
-Philippe2009$result[ (Philippe2009$tree_supported == "tr2" ) & (Philippe2009$ABS >= 1) ] = "Porifera-sister-strong-signal"
-Philippe2009$result[ (Philippe2009$tree_supported == "tr2" ) & (Philippe2009$ABS <1) ] = "Porifera-sister-weak-signal"
-
-Philippe2009 %>%
-	count(result) %>%
-	mutate(perc = n/ nrow(Philippe2009)) -> Philippe2009_count
-Philippe2009_count$study <- 'Philippe2009_Opisthokonta_C60'
-
-
-Philippe2009_only_choanozoa  <- read_tsv("../trees_new/AU_test/Philippe2009_only_choanozoa_C60_genewise_lnL.txt")
-Philippe2009_only_choanozoa$ABS <- abs(Philippe2009_only_choanozoa$`tr1_log-likelihood` - Philippe2009_only_choanozoa$`tr2_log-likelihood`)
-Philippe2009_only_choanozoa$result[ (Philippe2009_only_choanozoa$tree_supported == "tr1" ) & (Philippe2009_only_choanozoa$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
[…]

### 2019-11-14 manuscript/manuscript_kernel.R
iqtree bootstrap cutoff to 95% as per Li and iqtree docs

@@ -50,7 +50,7 @@ focal_matrices =
 						 header = TRUE,
 						 stringsAsFactors = FALSE)
 
-bootstrap_threshold = 90
+bootstrap_threshold = 95
 posterior_prob_threshold = 95
 
 # Load data

### 2019-11-17 manuscript/manuscript.rmd
update abstract, methods and resluts to the manuscipt

@@
 ## Abstract
+Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages.
+However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
+Here, we systematically explore the mechanisms that potentially cause the incongruent results from different studies by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
+We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
+Moreover, We find  Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of more closely related outgroups, whereas Ctenophora-sister is recovered in other cases even with site-heterogeneous C60 models.
+We next quantify the support of phylogenetic signal over the two hypotheses and characterize substantial differences of phylogenetic distribution between site-homogenious and site-heterogeneous models.
+Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
 
@@
 ### Phylogenetic signal
-To further explore the effect of site-homogeneous or site-heterogeneous models and outgroup choices in on the animal-root position,  we next quantify the support of phylogenetic signal over two alternative hypotheses ()
+
+To further explore the effect of site-homogeneous or site-heterogeneous models and outgroup choices in on the animal-root position, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+1A); T2: Porifera-sister (Fig.
+1B)) to three key data matrices with two outgroup choice (Choanimalia and Opisthokonta), including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis.
+These results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis.
+
+Surprisingly, we found significant differences of phylogenetic distribution between site-homogenious and site-heterogeneous models.
+We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal.
+This finding is consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models [@shen2017contentious].
+
+In contrast to site-homogenious models, the distribution of phylogenetic signal over the animal-root position has never been quantified with site-heterogeneous models.
+Surprisingly, except for Whelan2017_Holozoa matrix, we found that the phylogenetic signal decreases in many genes using C models compared to WAG models, especially in two matrices from Ryan_2013 that over 30% of genes changed from strong to week signal (\Delta$lnL<2).
+It is not clear the reason why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing (control test will be conducted here, by changing monophyletic of bilaterian vs.
+non-monophyletic of bilaterians...).
+
+```{r matrix_overlap}
+
+phylogenetic_signal_by_genes <- rbind(Philippe2009_count, Philippe2009_WAG_count, Philippe2009_only_choanozoa_count, Philippe2009_WAG_only_choanozoa_count,  Ryan2013_count, Ryan2013_WAG_count, Ryan2013_only_choanozoa_count, Ryan2013_WAG_only_choanozoa_count, Whelan2017_full_count, Whelan2017_full_WAG_count,  Whelan2017_full_only_choanozoa_count, Whelan2017_full_WAG_only_choanozoa_count)
+ggplot(phylogenetic_signal_by_genes, aes(x = study, y = perc, fill=result)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle=45)) +  scale_fill_manual(values=c("#E69F00", "#999999","#009E73", "#999999" ))
+
+```
+**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage.
+Proportion of genes supporting each of two alternative hypotheses for each of
+
 ## The current state of understanding

### 2019-11-21 manuscript/manuscript.rmd
update manuscript

@@
 
-#### phylogenetic signal in conflicted dataset
+#### phylogenetic distribution of support
 To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.
 We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above.
+For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models.
 The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
@@
 For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
+
+#### Performance of data recoding
 

### 2019-11-21 manuscript/manuscript_kernel.R
renamed some pb trees for proper parsing, bringing in Nosenko2013 iqtree now

@@ -201,7 +201,7 @@ file_names_iqtree = list.files( path = trees_path_iqtree, pattern = iqtree_ext, 
 
 # Temp fix re issue #8
 file_names_iqtree = file_names_iqtree[ ! grepl("Philippe2009", file_names_iqtree) ]
-file_names_iqtree = file_names_iqtree[ ! grepl("Nosenko2013", file_names_iqtree) ]
+#file_names_iqtree = file_names_iqtree[ ! grepl("Nosenko2013", file_names_iqtree) ]
 
 # read iqtrees
 trees_iq = foreach( tree_file = file_names_iqtree ) %dopar%

### 2019-11-21 manuscript/manuscript_kernel.R
add LG+C60 to model factors

@@ -244,7 +244,7 @@ analyses_new$result = factor( analyses_new$result )
 
 # Parse model components
 analyses_new$model_summary = analyses_new$model
-analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60+F+G", "CAT+F81"))
+analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60", "LG+C60+F+G", "CAT+F81"))
 
 # Partition comparison across matrices
 

### 2019-11-21 manuscript/manuscript.rmd
pass at au test / phylogenetic signal plot

@@
 
-```{r matrix_overlap}
+```{r phylogenetic_signal}
+au_tests %>%
+  unite(Matrix, c(manuscript, matrix)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
 
-phylogenetic_signal_by_genes <- rbind(Philippe2009_count, Philippe2009_WAG_count, Philippe2009_only_choanozoa_count, Philippe2009_WAG_only_choanozoa_count,  Ryan2013_count, Ryan2013_WAG_count, Ryan2013_only_choanozoa_count, Ryan2013_WAG_only_choanozoa_count, Whelan2017_full_count, Whelan2017_full_WAG_count,  Whelan2017_full_only_choanozoa_count, Whelan2017_full_WAG_only_choanozoa_count)
-ggplot(phylogenetic_signal_by_genes, aes(x = study, y = perc, fill=result)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle=45)) +  scale_fill_manual(values=c("#E69F00", "#999999","#009E73", "#999999" ))
+ggplot(aes(x = Matrix, y = percent, fill=result)) + 
+  geom_bar(stat = "identity") +
+  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
 

### 2019-11-21 manuscript/functions.R

@@ -521,7 +521,7 @@ get_matrix_occupancy = function( seq_matrix ){
 parse_au_gene_tests = function(path="../trees_new/AU_test"){
 	files_list = list.files(path, pattern = "genewise", full.names = TRUE, recursive = TRUE) 
 	filename_parts_list = str_split(gsub('_genewise_lnL.txt', '', basename( files_list )), '_', n=2)
-
+	likelihood_cutoff = 2
 	au_tibble = 
 		files_list %>%
 		map(read.delim) %>%
@@ -530,10 +530,10 @@ parse_au_gene_tests = function(path="../trees_new/AU_test"){
 		bind_rows()
 
 	au_tibble$ABS <- abs(au_tibble$tr1_log.likelihood - au_tibble$tr2_log.likelihood)
-	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS >= 1) ] = "Ctenophore-sister-strong-signal"
-	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS < 1) ] = "Ctenophore-sister-weak-signal"
-	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS >= 1) ] = "Porifera-sister-strong-signal"
-	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS <1) ] = "Porifera-sister-weak-signal"
-	
+	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS >= likelihood_cutoff) ] = "Ctenophore-sister-strong-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr1" ) & (au_tibble$ABS < likelihood_cutoff) ] = "Ctenophore-sister-weak-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS >= likelihood_cutoff) ] = "Porifera-sister-strong-signal"
+	au_tibble$result[ (au_tibble$tree_supported == "tr2" ) & (au_tibble$ABS < likelihood_cutoff) ] = "Porifera-sister-weak-signal"
+	au_tibble$result <- factor(au_tibble$result, levels = c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
 	return(au_tibble)
 }

### 2019-11-21 manuscript/functions.R
simplified model names, re-ran kernel

@@ -417,7 +417,7 @@ parse_tree_iqtree = function( tree_file, taxonomy_reference ){
 	if( length(best_line) > 0 ){
 		tree$modelfinder = TRUE
 		elements = strsplit( best_line[1], " " ) %>% unlist()
-		tree$model = elements[3]
+		tree$model = str_remove(elements[3], '\\+F\\+G')
 	}
 	return(tree)
 }

### 2019-11-21 manuscript/manuscript_kernel.R

@@ -244,7 +244,7 @@ analyses_new$result = factor( analyses_new$result )
 
 # Parse model components
 analyses_new$model_summary = analyses_new$model
-analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "poisson_C60", "WAG+C60+F+G", "LG+C60", "LG+C60+F+G", "CAT+F81"))
+analyses_new$model_summary = factor( analyses_new$model_summary, levels=c("WAG", "GTR20", "Poisson_C60", "WAG+C60", "LG+C60", "LG+C60", "CAT+F81"))
 
 # Partition comparison across matrices
 

### 2019-12-05 manuscript/manuscript.rmd
matrix overlap figure now mostly legible

@@
 
+no_subsets = unlist(lapply(sequence_matrices, function(x) !grepl("only|_D", x@matrix_name)))
+
 taxon_rectangles = lapply(
-	sequence_matrices,
+	sequence_matrices[no_subsets],
 	function ( sequence_matrix ){
@@
 
-filtered_taxon_rectangles <- taxon_rectangles[!grepl("only", taxon_rectangles$matrix),]
-
-ggplot(data=filtered_taxon_rectangles) +
+ggplot(data=taxon_rectangles) +
 	scale_x_continuous(name="x") +
@@
 
-```{r matrix_overlap}
+```{r matrix_overlap, fig.height = 12, fig.width = 20}
 
 matrix_overlap_rectangles =
-	lapply(sequence_matrices, function(x) lapply(sequence_matrices, function(y) overlap_rects(x,y))) %>%
+	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
 	unlist(recursive=FALSE) %>%
@@
 ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="x") +
-	scale_y_continuous(name="y") +
+	scale_x_continuous(name="Genes") +
+	scale_y_continuous(name="Species") +
 	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2 )
+	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
 

### 2019-12-05 manuscript/functions.R

@@ -182,8 +182,8 @@ overlap_rects = function( msa1, msa2 ){
 	D$ymax = c( overlap$n_species_1, overlap$n_species_1 + overlap$n_species_2 - overlap$n_species_overlap )
 	D$manuscript_1 = msa1@manuscript_name
 	D$manuscript_2 = msa2@manuscript_name
-	D$matrix_1 = paste(msa1@manuscript_name, msa1@matrix_name, sep="_")
-	D$matrix_2 = paste(msa2@manuscript_name, msa2@matrix_name, sep="_")
+	D$matrix_1 = msa1@matrix_name
+	D$matrix_2 = msa2@matrix_name
 	
 	D
 	

### 2019-12-05 manuscript/functions.R
match matrix to partition map with grepl so our subsampled matrices still work

@@ -62,7 +62,7 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 	
 	if( ! is.null(partition_map_global) ){
 		partition_map_global %<>%
-			filter( matrix == matrix_name )
+			filter( grepl(matrix, matrix_name) )
 		
 		partitions %<>% 
 			left_join( partition_map_global ) %>%

### 2019-12-05 manuscript/functions.R
actually match matrix names this time

@@ -61,11 +61,9 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
 		bind_rows()
 	
 	if( ! is.null(partition_map_global) ){
-		partition_map_global %<>%
-			filter( grepl(matrix, matrix_name) )
-		
+
 		partitions %<>% 
-			left_join( partition_map_global ) %>%
+			left_join( partition_map_global[startsWith(matrix_name, partition_map_global$matrix),] ) %>%
 			select( -matrix )
 		
 		# Add names to the partitions that are not assigned to components. Take these from the original 

### 2019-12-05 manuscript/manuscript_kernel.R
blast network cutoff changed to be more permissive: pident 78 -> 50, evalue 1e-15- > 1e-5. The node splitting rules kept everything from glomming together.

@@ -252,13 +252,6 @@ n_total_partitions =
 	partitions_all %>%
 	group_by(matrix) %>%
 	summarize("n_total_partitions"=n_distinct(partition))
-n_compandBUSCO =
-	partition_map_global %>%
-		group_by(matrix) %>%
-		summarize(
-			"n_components"=n_distinct(component_number),
-			"n_distinct_BUSCO"=n_distinct(BUSCO_ID)
-		)
 n_components_with_BUSCO =
 	partition_map_global %>%
 		filter(BUSCO_ID != "") %>%
@@ -279,7 +272,6 @@ discarded_parts =
 
 partition_network_summary =
 	n_total_partitions%>%
-	# left_join(n_compandBUSCO,          by="matrix") %>%
 	left_join(n_components_with_BUSCO, by="matrix") %>%
 	left_join(n_ribo,                  by="matrix") %>%
 	left_join(discarded_parts,         by="matrix") %>%

### 2019-12-12 manuscript/manuscript.rmd
add Whelan and Ryan phylobayes results

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: default

### 2019-12-12 manuscript/manuscript.rmd
update manuscript

@@
 However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
-Here, we systematically explore the mechanisms that potentially cause the incongruent results from different studies by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
-We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
-Moreover, We find  Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of more closely related outgroups, whereas Ctenophora-sister is recovered in other cases even with site-heterogeneous C60 models.
-We next quantify the support of phylogenetic signal over the two hypotheses and characterize substantial differences of phylogenetic distribution between site-homogenious and site-heterogeneous models.
+Here, we systematically explore the mechanisms (evolutionary models, matrix composition and overlappin and outgroup choice) that potentially cause the incongruent results from different studies by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
+We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices from differnet studies.
+Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases.
+We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models.
+Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers.
 Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
@@
 In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for complex structures such as cell types, nervous systems, and developmental patterns.
+An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
 However, a wide consensus of the position of animal has not been reached despite multiple efforts.

### 2019-12-12 manuscript/functions.R
better AU test handling

@@ -518,13 +518,16 @@ get_matrix_occupancy = function( seq_matrix ){
 #' @export
 parse_au_gene_tests = function(path="../trees_new/AU_test"){
 	files_list = list.files(path, pattern = "genewise", full.names = TRUE, recursive = TRUE) 
-	filename_parts_list = str_split(gsub('_genewise_lnL.txt', '', basename( files_list )), '_', n=2)
+	filename_parts_list = str_split(gsub('_genewise_lnL.txt', '', basename( files_list )), '_')
 	likelihood_cutoff = 2
 	au_tibble = 
 		files_list %>%
 		map(read.delim) %>%
 		imap(~ transform(.x, manuscript = filename_parts_list[[.y]][[1]])) %>%
-		imap(~ transform(.x, matrix = filename_parts_list[[.y]][[2]])) %>%
+	  imap(~ transform(.x, matrix = ifelse( length(filename_parts_list[[.y]])>2,
+	                                        filename_parts_list[[.y]][2:(length(filename_parts_list[[.y]])-1)],
+	                                        filename_parts_list[[.y]][1]))) %>%
+		imap(~ transform(.x, model = filename_parts_list[[.y]][[length(filename_parts_list[[.y]])]])) %>%
 		bind_rows()
 
 	au_tibble$ABS <- abs(au_tibble$tr1_log.likelihood - au_tibble$tr2_log.likelihood)

### 2019-12-19 manuscript/manuscript.rmd
update manuscript

@@
 ^3^Department of Biological Sceince, Vanderbilt University
+
 \* Corresponding author, casey.dunn@yale.edu
@@
 However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
-Here, we systematically explore the mechanisms (evolutionary models, matrix composition and overlappin and outgroup choice) that potentially cause the incongruent results from different studies by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
-We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices from differnet studies.
+Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
+We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
 Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases.
 We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models.
+Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
 Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers.
@@
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
-Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither.
+Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither.
 As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
@@
 Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
-
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include:
-
-- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
-The F81 matrix is also sometimes referred to as the Poisson matrix.
-It has no free parameters to estimate since all off-diagonal elements are set to 1.
-
-- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
-It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
-It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
-This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
-This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
-
-While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
-There are a couple approaches that have been used in the studies considered here:
-
-- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
-
-- Gamma rate heterogeneity.
-Each site is assigned to a different rate class with its own rate value.
-This accommodates different rates of evolution across different sites.
-Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
-
-The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
-
-- Empirical site-homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
-
-- Estimated site-homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
-
-- CAT site heterogeneous [@Lartillot:2004dq].
-Each site is assigned to a class with its own equilibrium frequencies.
-The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
-
-- C10 to C60 [@si2008empirical]. 10 to 60-profile mixture models as variants of the CAT model under the maximum-likelihood framework.
-
-- Six-state amino acid recoding.
-Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
+The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
+A detailed description of each component is in Supplementary text.
+Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as Site homogeneous rates, Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models) .
+(4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
 Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
@@
 The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here, F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
+For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
 LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
@@
 CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-One could, though, run a WAG+CAT model.
-
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+One could, though, run a WAG+CAT model.To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
 If there are no modifiers, then it is implied that site homogeneous models are used.
@@
 
-High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]).
-Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be conserved across animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes).
-Surprisingly, gene composition and matrice overlap across different studies has been rarely assessed, since most studies relied on their own bioinformatic pipeline for determination of OGs.
+High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]) to explore the root of metazoan phylogeny.
+Although the number of sampled genes might be different across studies, it should expect that the sampled genes should be conserved across most animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes).
+Surprisingly, gene composition and matrix overlap across different studies has been rarely assessed, and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text).
 
@@
 Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**).
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [e.g.
+@ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
 It has been suggested the inclusion of more distant related (e.g.
-Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered only when site-heterogeneous CAT models were used.
+Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when site-heterogeneous CAT models were used [@pisani2015genomic].
 In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling.
@@
 
-Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the animal root position.
-Feuda et al.
-2017 [@feuda2017improved] has proposed to recode the full set of twenty amino acids into six groups that have more frequent evolutionary changes within groups than between groups based on different exchangeability matrices.
+Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving].
+Feuda *et al*.
+(2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices.
 By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
@@
 2015 [@chang2015genomic] and Whelan et al.
-2015 [@whelan2015error], they found that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
-However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies [@hernandez2019six].
-Thus, it is important to evaluate the performance of data recoding method verses non-recoding approaches in past empritical matrices .
+2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
+However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
 
@@
 
-Taxa sampling plays a critical role in phylogenetic inference, especially for the phylogenetic position that needs to be explored.
+Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored.
 However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
 These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
-Importantly, some earlier studies only include less than five ctenophore species.
+Importantly, everal earlier studies only include less than five ctenophore species.
 The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
-In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position, and it should be noted that Whelan2017 [@whelan2017ctenophore] and Simion2017 [@simion2017large] have much broader taxon sampling than other data matrices: Simion2017 has the most taxon sampling, and Whelan2017 contains greater ctenophore taxon sampling.
+In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position.
+Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices.
 
@@
 
-Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes to tackle the incongruence of the topology among individual gene trees.
+Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogney.
 The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
 However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring].
-Thus, it is important to evaluate the gene composition and matrix overlap to explore congruence of gene sampling across different matrices.
+Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices.
 It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.
@@
 We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected.
-The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies.
-Morever, the BUSCO dataset also are frequently used as phylogenomic markers in a varietly of studies.
-Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conserveness and high-expression properties.
+The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
+Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ).
+Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conservation and ubiquitous.
 However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
 XXBUSCO_annotations).
-
-We furthered BLAST gene partitions to SwissProt database
 
@@
 These results indicated that different studies essentially sampled different regions of the genomes with little overlap.
-One thing should be noted that
 
@@
 
-Here we summarized all the 167 previous phylogenetic analyses from all studies (Fig.
+Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
 support_published_analyses and Table perviouly_published_analyses).
-We found that all past phylogenomic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference, recoding methods are used.
-We found that support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT models.
-It should be noted that the convergence were not reached or posterior probability is not well supported in many BI analyses.
-Moreover, the strong support of Porifera-sister are mainly from analyses using Poisson+CAT and recoding methods with site-heterogeneous models.
-Moreover, only a few past analyses with the most complex GTR+CAT model are converged.
-However, for the ones that are converged with GTR+CAT model, Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister when choanozoa are used as outgroups.
-
-The use of site-heterogeneous CAT is an infinite mixture model has long been suggested as the best approach to account for systematic errors, particular for data saturation, especially in the case of long-branch attraction (LBA) (Lartillot and Philippe, 2004).
-The GTR-CAT model often received the best-model fit from cross validation or posterior predictive analyses compared to site-homogeneous or Poisson-CAT models.
-However, the GTR+CAT model is extremely computational demanding and not feasible in large matrices so many studies only include Poisson+CAT models in their analyses.
-In contrast, a recent study has suggested that Poisson+CAT model did not outperform site-homogeneous with data partitioning or GTR + CAT models and in some cases can infer incorrect topology based on the simulation analyses.
-The author also raised the issue of the potential overparameterization in CAT models in some cases.
-Nevertheless，the debate over the site-heterogeneous vs.
-site-homogeneous models and/or inclusion or exclusion more distantly related outgroups, constitutes the probably one of the most controversial nodes in the animal tree of life.
+Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table PAPERS).
+Moreover, three of them are reanalyses papers without generating any new data matrices.
+We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
+It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mianly due to computational limits of CAT models employed in PhyloBayes.
+The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models.
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
+support_published_analyses).
 
@@
 
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree.
+Here we reanalyzed `r nrow(analyses_datasets)` from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets).
+We also progressively trimmed each data matrix with different level of outgroup choice if applicable.
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices).
 We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
@@
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-
-With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister, every analysis conducted herein choice strongly supported the ctenophore-sister hypothesis, even with site-heterogeneous C models with the most closely related outgroups.
+With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.
 
@@
 This has been addressed with CAT models.
-iqtree provides a new family of C models that also address site heterogeneity.
+IQtree provides a new family of C models that also address site heterogeneity.
 Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
@@
 Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices.
-Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, and Whelan2015.
-Moreover, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
 Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix.
+Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Whelan2015_D20.
+In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
+Ctenophora-sister were strongly supported in all other matrices.
 For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model.
-Consistently with past study, we found strong support of Ctenophora-sister in all anaylzed Opisthokonta matrices, indicating that Ctenophora-sister is also favored in all matrices with more distantly related taxa are used as outgroups.
+We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups.
+Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.
 
@@
 
-To further explore the effect of site-homogeneous or site-heterogeneous models and outgroup choices in on the animal-root position, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+To further explore the effect of model and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
 1A); T2: Porifera-sister (Fig.
-1B)) to three key data matrices with two outgroup choice (Choanimalia and Opisthokonta), including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis.
-These results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis.
-
-Surprisingly, we found significant differences of phylogenetic distribution between site-homogenious and site-heterogeneous models.
+1B)) to three key data matrices that have incongruent results with two outgroup choice, including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal) and four different models in both ML and BI framework.
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis when using LG+GAMMA and C models and the results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis.
+However, we found significant differences of phylogenetic distribution between LG and C models.
 We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal.
@@
 non-monophyletic of bilaterians...).
+
+### Bias of data-recoding
 
@@
 # Supplemental Information
+## Models of molecular evolution
+
+The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR.
+While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+There are a couple approaches that have been used in the studies considered here:
+
+- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
+The F81 matrix is also sometimes referred to as the Poisson matrix.
+It has no free parameters to estimate since all off-diagonal elements are set to 1.
+
+- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
+It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
+It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
+This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
+This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
+
+While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+There are a couple approaches that have been used in the studies considered here:
+
+- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
+
+- Gamma rate heterogeneity.
+Each site is assigned to a different rate class with its own rate value.
+This accommodates different rates of evolution across different sites.
+Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
+
+The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
+
+- Empirical site-homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
+
+- Estimated site-homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
+
+- CAT site heterogeneous [@Lartillot:2004dq].
+Each site is assigned to a class with its own equilibrium frequencies.
+The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
+
+- C10 to C60 [@si2008empirical]. 10 to 60-profile mixture models as variants of the CAT model under the maximum-likelihood framework.
+
+- Six-state amino acid recoding.
+Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
+Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
 

### 2020-01-02 manuscript/manuscript.rmd
add recoding and update manuscript

@@
 bibliography: manuscript.bib
-csl: plos.csl
+csl: nature.csl
 output:
@@
 Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases.
-We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models.
+We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (in site-heterogeneous CAT models) .
 Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
-Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers.
+Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers.
 Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
@@
 
-Table 1.
 ```{r overview data matrices}
-	kable(datasets)
+table1 = select(datasets, -c(paths))
+	kable(table1, caption= "Table 1. Overview of data matrices used in this study")
 
@@
 
-clade_colors = viridis( length(clades) )
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
+
+clade_colors = c("#e31a1c","#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#a6cee3","#fdbf6f","#cab2d6", "#ff7f00")
 names( clade_colors ) = clades
@@
 ggplot(data=taxon_rectangles) +
-	scale_x_continuous(name="x") +
-	scale_y_continuous(name="y") +
+	scale_x_continuous(name="Gene sampling") +
+	scale_y_continuous(name="Taxa sampling") +
 	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
-	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
-	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
-	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
+	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
+ scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
+	#scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
 	facet_wrap( ~ matrix )
@@
 
-Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
-Our initial intent of comparing matrix overlap across matrices was to construct a new matrice with shared taxon and gene sampling to test specific hypotheses about differences in support.
+Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
+Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
 The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs.
@@
 
-Here we reanalyzed `r nrow(analyses_datasets)` from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets).
+Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets).
 We also progressively trimmed each data matrix with different level of outgroup choice if applicable.
@@
 Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix.
-Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Whelan2015_D20.
+Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015.
 In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
@@
 
-To further explore the effect of model and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models.
+To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
 1A); T2: Porifera-sister (Fig.
-1B)) to three key data matrices that have incongruent results with two outgroup choice, including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal) and four different models in both ML and BI framework.
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis when using LG+GAMMA and C models and the results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis.
-However, we found significant differences of phylogenetic distribution between LG and C models.
-We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal.
-This finding is consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models [@shen2017contentious].
+1B)) to three key data matrices with two outgroup choice and four models in both ML and BI framework (Fig XXPhylogenetic signal).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models.
+Moreover, the outgroup choice has little impact in general on the distribution of the support phylogenetic signal in site-homogeneous models.
+This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
 
-In contrast to site-homogenious models, the distribution of phylogenetic signal over the animal-root position has never been quantified with site-heterogeneous models.
-Surprisingly, except for Whelan2017_Holozoa matrix, we found that the phylogenetic signal decreases in many genes using C models compared to WAG models, especially in two matrices from Ryan_2013 that over 30% of genes changed from strong to week signal (\Delta$lnL<2).
-It is not clear the reason why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing (control test will be conducted here, by changing monophyletic of bilaterian vs.
-non-monophyletic of bilaterians...).
+In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+phylogenetic_signal B).
+It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny.
+In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal.
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled genes) when exclusive of more distantly related outgroups with the Poisson-CAT models.
+The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.
 
-### Bias of data-recoding
+We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices.
+Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
+In other words.
+our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs.
+site-heterogeneous models).
+More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these data matrices.
+Thus, it is clear that recovering Porifera-sister requires more constrains including model, outgroup choice, and even with gene sampling.
 
 ```{r phylogenetic_signal}
-au_tests %>%
-  unite(Matrix, c(manuscript, matrix)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
+#au_tests %>%
+#  unite(Matrix, c(manuscript, matrix, model)) %>%
+#  group_by(Matrix, result) %>%
+#  count() %>%
+#  group_by(Matrix) %>%
+#  mutate(percent = prop.table(n)) %>%
 
-ggplot(aes(x = Matrix, y = percent, fill=result)) +
+#ggplot(aes(x = Matrix, y = percent, fill=result)) +
+#  geom_bar(stat = "identity") +
+#  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+#  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+#  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+
+library(ggpubr)
+theme_set(theme_pubr())
+au_tests = read_tsv("../trees_new/AU_test/summary.tsv")
+
+ Philippe2009 = filter(au_tests,
+                   grepl('Philippe', Matrix))
+
+x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanozoa_WAG", "Philippe2009_LG", "Philippe2009_Choanozoa_LG", "Philippe2009_LG(phylobayes)", "Philippe2009_Choanozoa_LG(phylobayes)", "Philippe2009_C60", "Philippe2009_Choanozoa_C60", "Philippe2009_F81", "Philippe2009_Choanozoa_F81"))
+y1=factor(Philippe2009$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+
+ Ryan2013_est = filter(au_tests,
+                   grepl('Ryan', Matrix))
+
+y2=factor(Ryan2013_est$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+
+x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanozoa_WAG", "Ryan2013_est_LG", "Ryan2013_Choanozoa_LG",  "Ryan2013_est_C60", "Ryan2013_Choanozoa_C60"))
+
+Whelan2017 = filter(au_tests,
+                   grepl('Whelan', Matrix))
+y3=factor(Whelan2017$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+
+x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanozoa_WAG", "Whelan2017_full_LG", "Whelan2017_Choanozoa_LG", "Whelan2017_full_LG(phylobayes)", "Whelan2017_Choanozoa_LG(phylobayes)", "Whelan2017_full_C60", "Whelan2017_Choanozoa_C60", "Whelan2017_full_F81", "Whelan2017_Choanozoa_F81"))
+
+p1 <- ggplot(Philippe2009, aes(x = x1, y = percent, fill=y1)) +
   geom_bar(stat = "identity") +
@@
 
+p2 <- ggplot(Ryan2013_est, aes(x = x2, y = percent, fill=y2)) +
+  geom_bar(stat = "identity") +
+  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+
+p3 <- ggplot(Whelan2017, aes(x = x3, y = percent, fill=y3)) +
+  geom_bar(stat = "identity") +
+  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+
+figure <- ggarrange(p1, p2, p3,
+                    labels = c("A", "B", "C"),
+                    ncol = 1, nrow = 3, widths=c(2,1,2), heights=c(1,1,1), common.legend = TRUE, legend = "right")
 ```
+
+### Bias of data-recoding
+There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Figure ).
+Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure 1B).Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
+
 **Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage.
@@
 ## The current state of understanding
-
-### Interpretting variation support
-
-External criteria, eg posterior predictive scores, model fit etc
 
@@
 We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models.
+For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models.
 The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
@@
 
-#### Performance of data recoding
+#### Performance of data-recoding
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
+The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
+
+The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
+This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
+
+The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
+To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
+This creates new recodings that have the same sized groups as SR-6.
+The new recodings were, from `random-00` to `random-03`:
+
+    MPKE AY IDGQ TRC WNLF SVH
+    EIFT WL QVPG NKM SCHA RYD
+    LCQS GK WPTI VRD YEFN MAH
+    IWQV TY KDLM ESH APCF GRN
+
+To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
+
+The 20 state matrices are the same across all analyses since they are not recoded.
+Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.
+
+Each new matrix was analyzed with PhyloBayes-mpi.
+Analyses were run for 1000 generations, and a 200 generation burnin applied.
+The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
+
+The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 
-- Six-state amino acid recoding.
-Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
+- Recoding method:  Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
 Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
@@
 
-## Model matrix comparison
+## Data-recoding methods
 
-WAG and LG are both fixed exchange matrices. Their differences are largely limitted to a few amino acid changes.
+Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
+In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig ##Random_recodingA).
+Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
+Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister.
+Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies.
+In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids.
+These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
+The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
+Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
+Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister.
+Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister.
+They also show that recoding can be a problematic method for addressing compositional variation.
 
-```{r wagvlg}
+Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes.
+This provides 32 analyses that they report in their Table 3 and that we present graphically here as Supplemental Figure ##recoding_summary.
+There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Figure ##recoding_summary).
+Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
+They considered five different posterior predictive statistics that capture different types of variation in the data.
+From this they conclude that their "results strongly support sponges as the sister group of all other animals".
 
-# Comparison of LG and WAG matrices
+This conclusion does not follow from their own presented results.
+Only a single analysis with posterior predictive scores provides what could be considered strong support > 95 posterior probability) for Porifera-sister.
+Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices).
+Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
+This leaves eight CAT-GTR+G analyses (Figure ##recoding_summary).
+Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
+Of the six recoded analyses, five are unresolved.
+Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
+Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Figure ##recoding_summary).
+The only statistic that stands out for this one analysis is that it has the highest maxdiff (Figure ##recoding_summary), indicating that it did not converge as well as other analyses.
 
-library( tidyverse )
-library( magrittr )
+Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally.
+This is important given the growing interest in recoding [@Feuda:2017ew].
+Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies.
+They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding.
+An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information.
+These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange.
+Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended.
+Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity.
+Similar results would suggest that the impact of recoding is due to discarding information.
+Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data.
+we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig ##random_recodingB).
 
-# Order in iqtree matrix
-aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
+These analyses suggest that the major impact of recoding on phylogenetic analyses is data reduction, not accommodation of compositional heterogeneity across species.
+This indicates that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies.
+Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [@Blanquart:2008gl, @Foster:2004tw], if progress can be made on the considerable computational challenges of such complex models.
 
-# Functional groups adjacent, from https://en.wikipedia.org/wiki/Amino_acid
-aa_functional = c( "R", "H", "K", "D", "E", "S", "T", "N", "Q", "C", "G", "P", "A", "V", "I", "L", "M", "F", "Y", "W" )
-
-triange_to_matrix = function( triangle ){
-	R_matrix = rep(NA, 400)
-	dim( R_matrix ) = c(20,20)
-
-	rownames( R_matrix ) = aa
-	colnames( R_matrix ) = aa
-
-	# Though iqtree specifies lower triangle, actually in the correct order for upper triangle as this R command reads it
-	R_matrix [ upper.tri( R_matrix ) ] = triangle
-
-	# Now fill in the lower half of the matrix
-	R_matrix[ is.na(R_matrix) ] = t( R_matrix )[ is.na(R_matrix) ]
-
-	# Scale
-	R_matrix = R_matrix / mean( R_matrix, na.rm=TRUE )
-
-	# Reorder by functional groupings
-	R_matrix = R_matrix[, match(aa_functional, colnames(R_matrix) )]
-	R_matrix = R_matrix[match(aa_functional, rownames(R_matrix)), ]
-
-	R_matrix
-}
-
-plot_matrix = function( R_matrix, title=NULL ){
-	D = data.frame( R_matrix )
-	D$source = row.names(D)
-	Dg = gather( D, 1:20, key=dest, value=exch, -source )
-	Dg %<>%
-		mutate( source=factor(source, levels=aa_functional)  ) %>%
-		mutate( dest=factor(dest, levels=aa_functional)  )
-
-	Dg %>%
-		ggplot(aes(x=dest, y=source)) +
-		geom_tile(aes(fill=exch)) +
-		scale_fill_gradient2() +
-		ggtitle(title)
-}
-
-# Triangle matrices from iqtree source code at https://github.com/Cibiv/IQ-TREE/blob/master/model/modelprotein.cpp
-
-wag_triangle = c(55.15710, 50.98480, 63.53460, 73.89980, 14.73040, 542.94200, 102.70400, 52.81910, 26.52560, 3.02949, 90.85980, 303.55000, 154.36400, 61.67830, 9.88179, 158.28500, 43.91570, 94.71980, 617.41600, 2.13520, 546.94700, 141.67200, 58.46650, 112.55600, 86.55840, 30.66740, 33.00520, 56.77170, 31.69540, 213.71500, 395.62900, 93.06760, 24.89720, 429.41100, 57.00250, 24.94100, 19.33350, 18.69790, 55.42360, 3.94370, 17.01350, 11.39170, 12.73950, 3.04501, 13.81900, 39.79150, 49.76710, 13.15280, 8.48047, 38.42870, 86.94890, 15.42630, 6.13037, 49.94620, 317.09700, 90.62650, 535.14200, 301.20100, 47.98550, 7.40339, 389.49000, 258.44300, 37.35580, 89.04320, 32.38320, 25.75550, 89.34960, 68.31620, 19.82210, 10.37540, 39.04820, 154.52600, 31.51240, 17.41000, 40.41410, 425.74600, 485.40200, 93.42760, 21.04940, 10.27110, 9.61621, 4.67304, 39.80200, 9.99208, 8.11339, 4.99310, 67.93710, 105.94700, 211.51700, 8.88360, 119.06300, 143.85500, 67.94890, 19.50810, 42.39840, 10.94040, 93.33720, 68.23550, 24.35700, 69.61980, 9.99288, 41.58440, 55.68960, 17.13290, 16.14440, 337.07900, 122.41900, 397.42300, 107.17600, 140.76600, 102.88700, 70.49390, 134.18200, 74.01690, 31.94400, 34.47390, 96.71300, 49.39050, 54.59310, 161.32800, 212.11100, 55.44130, 203.00600, 37.48660, 51.29840, 85.79280, 82.27650, 22.58330, 47.33070, 145.81600, 32.66220, 138.69800, 151.61200, 17.19030, 79.53840, 437.80200, 11.31330, 116.39200, 7.19167, 12.97670, 71.70700, 21.57370, 15.65570, 33.69830, 26.25690, 21.24830, 66.53090, 13.75050, 51.57060, 152.96400, 13.94050, 52.37420, 11.08640, 24.07350, 38.15330, 108.60000, 32.57110, 54.38330, 22.77100, 19.63030, 10.36040, 387.34400, 42.01700, 39.86180, 13.32640, 42.84370, 645.42800, 21.60460, 78.69930, 29.11480, 248.53900, 200.60100, 25.18490, 19.62460, 15.23350, 100.21400, 30.12810, 58.87310, 18.72470, 11.83580, 782.13000, 180.03400, 30.54340, 205.84500, 64.98920, 31.48870, 23.27390, 138.82300, 36.53690, 31.47300)
-wag_matrix = triange_to_matrix( wag_triangle ) %>% log()
-
-lg_triangle = c(0.425093, 0.276818, 0.751878, 0.395144, 0.123954, 5.076149, 2.489084, 0.534551, 0.528768, 0.062556, 0.969894, 2.807908, 1.695752, 0.523386, 0.084808, 1.038545, 0.363970, 0.541712, 5.243870, 0.003499, 4.128591, 2.066040, 0.390192, 1.437645, 0.844926, 0.569265, 0.267959, 0.348847, 0.358858, 2.426601, 4.509238, 0.927114, 0.640543, 4.813505, 0.423881, 0.311484, 0.149830, 0.126991, 0.191503, 0.010690, 0.320627, 0.072854, 0.044265, 0.008705, 0.108882, 0.395337, 0.301848, 0.068427, 0.015076, 0.594007, 0.582457, 0.069673, 0.044261, 0.366317, 4.145067, 0.536518, 6.326067, 2.145078, 0.282959, 0.013266, 3.234294, 1.807177, 0.296636, 0.697264, 0.159069, 0.137500, 1.124035, 0.484133, 0.371004, 0.025548, 0.893680, 1.672569, 0.173735, 0.139538, 0.442472, 4.273607, 6.312358, 0.656604, 0.253701, 0.052722, 0.089525, 0.017416, 1.105251, 0.035855, 0.018811, 0.089586, 0.682139, 1.112727, 2.592692, 0.023918, 1.798853, 1.177651, 0.332533, 0.161787, 0.394456, 0.075382, 0.624294, 0.419409, 0.196961, 0.508851, 0.078281, 0.249060, 0.390322, 0.099849, 0.094464, 4.727182, 0.858151, 4.008358, 1.240275, 2.784478, 1.223828, 0.611973, 1.739990, 0.990012, 0.064105, 0.182287, 0.748683, 0.346960, 0.361819, 1.338132, 2.139501, 0.578987, 2.000679, 0.425860, 1.143480, 1.080136, 0.604545, 0.129836, 0.584262, 1.033739, 0.302936, 1.136863, 2.020366, 0.165001, 0.571468, 6.472279, 0.180717, 0.593607, 0.045376, 0.029890, 0.670128, 0.236199, 0.077852, 0.268491, 0.597054, 0.111660, 0.619632, 0.049906, 0.696175, 2.457121, 0.095131, 0.248862, 0.140825, 0.218959, 0.314440, 0.612025, 0.135107, 1.165532, 0.257336, 0.120037, 0.054679, 5.306834, 0.232523, 0.299648, 0.131932, 0.481306, 7.803902, 0.089613, 0.400547, 0.245841, 3.151815, 2.547870, 0.170887, 0.083688, 0.037967, 1.959291, 0.210332, 0.245034, 0.076701, 0.119013, 10.649107, 1.702745, 0.185202, 1.898718, 0.654683, 0.296501, 0.098369, 2.188158, 0.189510, 0.249313)
-lg_matrix = triange_to_matrix( lg_triangle ) %>% log()
-
-lg_wag_diff = lg_matrix - wag_matrix
-
-lg_wag_diff_thresh =lg_wag_diff
-lg_wag_diff_thresh[abs(lg_wag_diff_thresh)<1] = 0
-
-grid.arrange(
-	plot_matrix(lg_matrix, title="(A) LG matrix"),
-	plot_matrix(wag_matrix, title="(B) WAG matrix"),
-	plot_matrix(lg_wag_diff, title="(C) LG - WAG"),
-	plot_matrix( lg_wag_diff_thresh, title="(D) Thresholded LG - WAG" )
-)
-
-```
+![](figures/Figure_overview.png)
+**Fig ##Random_recoding.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
 

### 2020-01-02 manuscript/functions.R

@@ -516,7 +516,7 @@ get_matrix_occupancy = function( seq_matrix ){
 #' @param path A string that specifies the path to root of gene-level AU test files of form manuscript_matrix_genewise_lnL.txt
 #' @return A tibble with columns: gene_id tree_supported tr1_log.likelihood tr2_log.likelihood tr3_log.likelihood manuscript matrix. tr1 = ctenophora sister tr2 = porifera sister tr3 = cnidaria/ctenophora sister
 #' @export
-parse_au_gene_tests = function(path="../trees_new/AU_test"){
+parse_au_gene_tests = function(path="../trees_new/AU_test/genewise"){
 	files_list = list.files(path, pattern = "genewise", full.names = TRUE, recursive = TRUE) 
 	filename_parts_list = str_split(gsub('_genewise_lnL.txt', '', basename( files_list )), '_')
 	likelihood_cutoff = 2

### 2020-01-03 manuscript/manuscript.rmd
update manuscript

@@
 library( gridExtra )
-
+library(ggpubr)
+theme_set(theme_pubr())
 source( "functions.R" )
@@
 
-^3^Department of Biological Sceince, Vanderbilt University
+^3^Department of Biological Science, Vanderbilt University
 
@@
 We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
-Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases.
-We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (in site-heterogeneous CAT models) .
+More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
+We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
 Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
@@
 Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
-Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as Site homogeneous rates, Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models) .
+Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).
 (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
-Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
+Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models [@feuda2017improved].
 
@@
 The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
+For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency.
 LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
@@
 CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-One could, though, run a WAG+CAT model.To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+One could, though, run a WAG+CAT model.
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
 If there are no modifiers, then it is implied that site homogeneous models are used.
@@
 
-High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]) to explore the root of metazoan phylogeny.
-Although the number of sampled genes might be different across studies, it should expect that the sampled genes should be conserved across most animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes).
-Surprisingly, gene composition and matrix overlap across different studies has been rarely assessed, and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text).
+High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of metazoan phylogeny (Table 1).
+Although the number of sampled genes might be different across studies, it is reasonable to expect that a conserved core set of orthology genes (OGs) should be present across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
+Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text) for animal phylogeny.
 
@@
 Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**).
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [e.g.
-@ryan2013genome; @pisani2015genomic].
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
 It has been suggested the inclusion of more distant related (e.g.
-Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when site-heterogeneous CAT models were used [@pisani2015genomic].
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic].
 In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling.
@@
 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
-However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
+However, how the recoding methods impact phylogenetic inference has not been well explored.
+Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
 
@@
 These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
-Importantly, everal earlier studies only include less than five ctenophore species.
+Importantly, several earlier studies only include less than five ctenophore species.
 The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
 In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position.
-Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices.
+Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices (Table1, Fi.
+XXTaxon_composition).
 
@@
 
-Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogney.
-The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
+Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny.
+It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
 However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring].
 Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices.
-It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.
+It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones (*e.g.*, Dunn2008 and Hejnol2009).
 
@@
 Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ).
-Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conservation and ubiquitous.
+Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
 However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
@@
 However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap).
-These results indicated that different studies essentially sampled different regions of the genomes with little overlap.
+Thus, these results suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
 
@@
 
-busco_overrepresented
-
 p1 = ggplot( matrix_summary ) +
 	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) +
+	geom_abline(slope=1,intercept=0) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
 	coord_fixed()
@@
 
-p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
+p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of  BUSCO genes")
+
+figure <- ggarrange(p1, p2,
+                    labels = c("A", "B"),
+                    ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")
 
@@
 
-**Fig XXBUSCO_annotations.** The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+**Fig XXBUSCO_annotations.** BUSCO annotation and representation in each data matrix.
+(A).
+The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+(B).
+The percentage BUSCO annotations in each matrix.
 
@@
 support_published_analyses and Table perviouly_published_analyses).
-Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table PAPERS).
-Moreover, three of them are reanalyses papers without generating any new data matrices.
+Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (TablexxPAPERS).
+Moreover, three of them are studies of reanalysis without generating any new data matrices.
 We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
-It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mianly due to computational limits of CAT models employed in PhyloBayes.
-The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models.
+It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mainly due to computational limits of CAT models employed in PhyloBayes.
+The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices.
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
 support_published_analyses).
+
+```{r overview of studies}
+table_Papers = read_tsv("../data_processed/tables/previously_published_manuscripts.tsv")
+	kable(table_Papers, caption= "Table XX. Overview of previous phylogenomic studies related to the root position of animal phylogeny")
+
+```
 
@@
 We also progressively trimmed each data matrix with different level of outgroup choice if applicable.
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices).
-We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except only LG+C60+F+G models are used large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices).
+We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
 It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive.
@@
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
+
 With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.
@@
 IQtree provides a new family of C models that also address site heterogeneity.
-Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
+Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-For computational efficiency, we first only applied Poisson+CAT models in every matrix with only Choanoflagellata as outgroup.
-It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.
+For computational efficiency, we first only applied Poisson+CAT models in every matrix with Choanoflagellata as outgroup.
+It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
 
@@
 Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015.
-In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
+Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
 Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model.
-We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups.
+For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
+We found strong support of Porifera-sister in holozoa matrices and Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups.
 Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.
@@
 1A); T2: Porifera-sister (Fig.
-1B)) to three key data matrices with two outgroup choice and four models in both ML and BI framework (Fig XXPhylogenetic signal).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models.
-Moreover, the outgroup choice has little impact in general on the distribution of the support phylogenetic signal in site-homogeneous models.
+1B)) to three key data matrices with different outgroup choice and  models in both ML and BI framework (Fig XXPhylogenetic signal).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models.
 This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
@@
 In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal.
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled genes) when exclusive of more distantly related outgroups with the Poisson-CAT models.
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models.
 The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.
@@
 site-heterogeneous models).
-More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these data matrices.
-Thus, it is clear that recovering Porifera-sister requires more constrains including model, outgroup choice, and even with gene sampling.
+More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these key data matrices.
+Thus, it is clear that recovering Porifera-sister requires more constrains, including more stringent model selection, outgroup choice, and even with gene sampling.
 
-```{r phylogenetic_signal}
+```{r phylogenetic_signal, fig.width=8, fig.height=15}
 #au_tests %>%
@@
 
-library(ggpubr)
-theme_set(theme_pubr())
 au_tests = read_tsv("../trees_new/AU_test/summary.tsv")
@@
                     ncol = 1, nrow = 3, widths=c(2,1,2), heights=c(1,1,1), common.legend = TRUE, legend = "right")
+
+figure
 ```
+**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage.
+Proportion of genes supporting each of two alternative hypotheses for each of
 
@@
 They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Figure ).
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Fig XXRecoidng_summary).
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure 1B).Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding.
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure XXRandom_recodingB).
+Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
-
-**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage.
-Proportion of genes supporting each of two alternative hypotheses for each of
 
@@
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI.
+Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI [@Lartillot:2013fg].
 To minimize computational burden, CAT-GTR models were only performed in the.
@@
 Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
-In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig ##Random_recodingA).
+In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig XXRandom_recodingA).
 Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
@@
 Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes.
-This provides 32 analyses that they report in their Table 3 and that we present graphically here as Supplemental Figure ##recoding_summary.
-There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Figure ##recoding_summary).
+This provides 32 analyses that they report in their Table 3 and that we present graphically here as Supplemental Fig XXRecoidng_summary.
+There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Fig XXRecoidng_summary).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
-This leaves eight CAT-GTR+G analyses (Figure ##recoding_summary).
+This leaves eight CAT-GTR+G analyses (Fig XXRecoding_pp).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
@@
 Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
-Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Figure ##recoding_summary).
-The only statistic that stands out for this one analysis is that it has the highest maxdiff (Figure ##recoding_summary), indicating that it did not converge as well as other analyses.
+Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Fig XXRecoding_pp).
+The only statistic that stands out for this one analysis is that it has the highest maxdiff (Fig XXRecoding_pp), indicating that it did not converge as well as other analyses.
 
@@
 When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig ##random_recodingB).
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig XXRandom_recodingB).
 
@@
 
-![](figures/Figure_overview.png)
-**Fig ##Random_recoding.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+![](figures/Figure_random_rocoding.png)
+
+Fig XXRandom_recoding.
+(A) The two alternative hypotheses for deep animal relationships considered here.
+Relationships that are not part of these hypotheses are shown as unresolved polytomies.
+(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+![](figures/Figure_recoding_summary.png)
+**Fig XXRecoidng_summary.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
 Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
 Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+
+![](figures/Figure_recoding_pp.png)
+
+**Fig XXRecoding_pp** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
+The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
+The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
+Cells are color coded as in Supplemental Fig XXRecoidng_summary.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
+The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
 

### 2020-01-08 manuscript/manuscript.rmd
add pairwise comparison of gene overlap (figure)

@@
 
-```{r phylogenetic_signal, fig.width=8, fig.height=15}
+```{r phylogenetic_signal, fig.width=8, fig.height=12}
 #au_tests %>%
@@
 Proportion of genes supporting each of two alternative hypotheses for each of
+
+![](figures/Figure_Pairwise_heatmap_animal_root.png)
+**Fig XXPairwise_heatmap_animal_root.**
 

### 2020-01-11 manuscript/manuscript.rmd
update previous analyses

@@
 This resulted in the 64-taxon matrix that is the focus of most of their analyses.
-Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
+Phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
 
@@
 
-Philippe *et al.* 2009 [Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals.
-The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
-The authors concluded:
+Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species.
+Th data matrix were assembled based on The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
+Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70.
+Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained.
+The phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq].
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
 > The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.
@@
 
-Hejnol *et al.* 2009 added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix.
-The orthology inference was largely simialr to Dunn et al.
-2008.
-Two datsets were constructed, one with 844 genes and the other with 330 genes.
-Maximum likelihood analyses from both datasets strongly supported Ctenophora-sister.
+Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms.
+The orthology inference was largely similar to Dunn et al.
+2008, with the exception of orthology genes were clustered by MCL.
+The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy.
+With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).
 
@@
 
+### Nosenko *et al.* 2013
+Nosenko *et al.* [@nosenko2013deep] added Expressed Sequence Tag (EST) data for 9 species of non-bilaterian metazoans (7 sponges).
+They constructed a novel matrix containing 122 genes and parsed them into two non-overlaping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny.
+The other major finding was that ribosomal gene partitions showed  significantly lower saturation than the non-ribosomal ones.
+
+OGs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect].
+They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix.
+By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister).
+Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in phylobayes [@Lartillot:2004dq].
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+> we were able to reconstruct a metazoan phylogeny that is consistent with traditional, morphology-based views on the phylogeny of non-bilaterian metazoans, including monophyletic Porifera and ctenophores as a sister-group of cnidarians.
+
+Note the main conclusion of this paper is baed on the selection of genes that evolved slowly ribosomal RNA genes.
+The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].
+
 ### Ryan *et al.* 2013
-Ryan *et al.* 2013 sequenced the first ctenophore genome of *Mnemiopsis leidyi*.
+Ryan *et al.* sequenced the first ctenophore genome of *Mnemiopsis leidyi*.
 With the genome resources of *M.
 leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals.
-They analyzed both matrices by site-homogeneous GTR+Gamma model and Poisson-CAT model with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology.
-The results strongly supported Ctenophora-sister in all datasets they analysed using site-homogeneous model.
-The Poisson + CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days.
+They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson-CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix.
+The OGs were constructed based on the method of Hejnol *et al.* 2009.
+For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-baed analyses.
+Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model.
+The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+>Our phylogenetic analyses suggest that ctenophores are the sister group to the rest of the extant animals.
 
@@
 
-Moroz *et al.* 2014 sequenced the second ctenophore genome *Pleurobrachia bachei* to explore the phylogenetic relationship of Metazoa.
+Moroz *et al.* [@moroz2014ctenophore] sequenced the second ctenophore genome *Pleurobrachia bachei* to explore the phylogenetic relationship of Metazoa.
 All phylogenetic analyses strongly supported Ctenophora-sister with different taxon and gene sampling using WAG site-homogeneous model.
+Two phylogenomic matrices were generated, the first set was represented by two ctenophore species, whereas the other set contained improved ctenophore sampling (10 taxa, Moroz2013_3d).
+Orthology determination employed in HaMStR [@ebersberger2009hamstr] using 1,032 "model organism" single-copy OGs.
+Sequence were then trimmed and aligned.
+This resulted in a final matrix of 170,871 amino acid positions across 586 genes with 44 taxa for the first matrix, and 114 genes with 60 taxa for the second matrix.
+All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different than CAT models in PhyloBayes) to reduce the computational cost.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+>Our integrative analyses place Ctenophora as the earliest lineage within Metazoa.
+This hypothesis is supported by comparative analysis of multiple gene families, including the apparent absence of HOX genes, canonical microRNA machinery, and reduced immune complement in ctenophores.
+
+It should be noted that only the Moroz_3d matrix has been reanalyzed in other studies, although the support of Ctenophora-sister is quite low.
+
+### Borowiec *et al.* 2015
+
+Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 OGs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
+The OGS were constructed using OrthologID pipeline [@chiu2006orthologid].
+After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080 OGs (Total 1080).
+The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes.
+The main conclusion of the paper was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices.
+Phylogenetic analyses were conducted under the GTR+CAT model in PhyloBayes in selected matrices, and under the data-partitioning methods in RAxML for all matrices.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+>Our phylogeny
+supports the still-controversial position of ctenophores as sister group to all other metazoans. This study also
+provides a workflow and computational tools for minimizing systematic bias in genome-based phylogenetic
+analyses.
+
+It should be noted that the authors also employed recoding-method in the Borowiec_Best108 matrix and found neither support of Porifera-sister or Ctenophora-sister [@borowiec2015extracting].
 
 ### Whelan et al. 2015
-Wehlan *et al.* 2015 constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
+Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
 long-branch attraction, compositional bias, fast evolving genes, etc.).
-Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (~ 1000 orthologs) and subsequetly removal of genes with too much missing data and potential paralogs.
+Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.
 The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups.
@@
 CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora.
+
 Regarding the recovery of Ctenophora-sister, the authors concluded:
+
 > Importantly, biases resulting from elevated compositional heterogeneity or elevated substitution rates are ruled out. Placement of ctenophores as
@@
 
+Note that the authors also reanalyzed Philippe2009 matrix (with the removal of ribosomal genes) and recovered Porifera-sister with moderate support (pp=90).
+
 ### Chang et al. 2015
 
-Chang *et al.* 2015 was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
+Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
 The authors constructed a dataset with 200 protein markers based on Philippe et al.
-2011 with 51,940 amino acids and 77 taxa.
+2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
 Both site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.
+
+Note that this data matrix has been extensively reanalyzed in many studies.
 
@@
 
-Pisani *et al.* 2015 reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013, Moroz2014 and Whelan2015 datasets.
-It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of animal-root position.
+Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets.
+It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny.
 The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction.
-They found strong support of porifera-sister when only include Choanozoa as outgroup in these datasets using site-heterogeous CAT model.
+Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in phylobayes.
+They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
+Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
+
 Regarding the recovery of Porifera-sister, the authors concluded:
->Our results reinforce a traditional scenario for the evolution of complexity in animals, and indicate that inferences about the evolution of Metazoa
-based on the Ctenophora-sister hypothesis are not supported by the currently available data.
+
+>Our results reinforce a traditional scenario for the evolution of complexity in animals, and indicate that inferences about the evolution of Metazoa based on the Ctenophora-sister hypothesis are not supported by the currently available data.
 
@@
 
-Feuda *et al.* didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets).
-It was the first study phylogenomic analysis using recoding strategy to explore animal-root position.
-The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, PartitionFinder)and site-heterogeneous (CAT-GTR) models with non-recoding and recoding datasets.
-The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis.
+Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets).
+It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods in the relationships at the root of the animal tree.
+The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets.
+The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).
+
 Regarding the recovery of Porifera-sister, the authors concluded:
@@
 
-### Whelan and Halalnych 2017a
+### Whelan and Halanych 2016
 
-Whelan *et al.* is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework.
+Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework.
 The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets.
-More importantly, the authors also showed that both Poisson + CAT and GTR +  CAT models could overestimated substitutional heterogeneity in many cases.
-They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site -homogeneious model.
-The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.
-The authors also concluded:
+More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in many cases.
+They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model.
+The results indicated that Poisson + CAT model tends to recover less accurate trees and more importantly,  both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.
+
+The authors concluded:
 
@@
 
-### Whelan *et al.* 2017b
+### Whelan *et al.* 2017
 
-Whelan *et al.* added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores.
+Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores.
 It significantly increased ctenophore taxon sampling than other studies.
@@
 The subsequent filtering strategy was also similar to the previous study.
-All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in Choanoazoa dataset.
+All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset.
+The main conclusions of this paper were based on Whelan2017_full and Whelan2017_strict matrices.
+
 Regarding the recovery of Ctenophora-sister, the authors concluded:
@@
 
-Simion *et al.* [@Dunn:2008ky] added transcriptomic data for 21 new animals.
+Simion *et al.* [@simion2017large] added transcriptomic data for 21 new animals.
 The data matrix was constructed using a semi-automated approach to comprehensively detect and eliminate potential systematic errors.
-The resulting dataset comprises 1,719 genes and 97 species, including 61 non-bilaterian metazoan species.
-It was by far the largest phylogenomic dataset in terms of taxon and gene sampling related to animal-root question.
+The resulting dataset comprises 1,719 genes and 97 species, including 61 non-bilaterian species.
+It was by far the largest phylogenomic dataset in terms of taxon and gene sampling related to the relationship at the root of animal phylogeny.
 
-The supermatrix was first analyzed using the Poisson-CAT model.
+The final matrix was first analyzed using the Poisson-CAT model.
 Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
 Each jackknife is baed on a random selection of ~ 25% of the genes.
-The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in the supermatrix dataset.
-Importantly, the authors compared the behavior of long-branch effect between site-heterogenious and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
+The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets.
+Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
+
 Regarding the recovery of Ctenophora-sister, the authors concluded:
@@
 Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.
+
+It should be noted that all the PhyloBayes runs have not reached convergence due to the computational cost in these large matrices.
 

### 2020-01-13 manuscript/manuscript.rmd
update manuscipt

@@
 Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages.
-However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
-Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
+However, a considerable debate still remains regarding to the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
+Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
 We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
@@
 An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
-However, a wide consensus of the position of animal has not been reached despite multiple efforts.
+However, a wide consensus of animl origin has not been reached despite multiple efforts.
 
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
-Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither.
+Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
 As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
@@
 
-High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of metazoan phylogeny (Table 1).
-Although the number of sampled genes might be different across studies, it is reasonable to expect that a conserved core set of orthology genes (OGs) should be present across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
-Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text) for animal phylogeny.
+High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of animal phylogeny (Table 1).
+Although the number of sampled genes might be different across studies, it is reasonable to expect that a core set of orthology genes (OGs) should be present and conserved across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
+Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (for details see Supplementary text section ).
 
@@
 
-Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**).
+Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (**Fig XXOverview.C**).
 Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
@@
 It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic].
-In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling.
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic].
+In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than outgroup sampling.
 This may be because results have tended to be more sensitive to outgroup sampling.
@@
 Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving].
+Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie].
 Feuda *et al*.
-(2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices.
+[@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
 By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
@@
 Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
+Thus, it is important to reevalute the performance of the impact of data-recoding compared to non-recoding methods in animal phylogeny.
 
@@
 ```{r overview data matrices}
-table1 = select(datasets, -c(paths))
+table1 = read_tsv("../data_processed/tables/study_summary.tsv")
 	kable(table1, caption= "Table 1. Overview of data matrices used in this study")
@@
 However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
-These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ).
-Importantly, several earlier studies only include less than five ctenophore species.
+These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
+Importantly, several earlier studies only include less than five ctenophore species (Fig XXTaxon_composition).
 The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
-In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position.
-Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices (Table1, Fi.
+In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
+Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices (Table1, Fi.
 XXTaxon_composition).
@@
 
-Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny.
+Gene sampling also plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny.
 It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
@@
 Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices.
-It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones (*e.g.*, Dunn2008 and Hejnol2009).
 
-We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected.
+We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected.
 The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
-Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ).
+Moreover, the BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
 Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
-However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
+However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
 XXBUSCO_annotations).
 
-Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
+Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
 Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
@@
 However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap).
-Thus, these results suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
+Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
+Moreover, the little overlap among studies may potentially lead to incongruent phylogenetic results and difficult to directly compare results across studies.
 
@@
                     ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")
+figure
 
@@
 The percentage BUSCO annotations in each matrix.
-
-**Fig XXGene_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc).
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
 
@@
 Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
-support_published_analyses and Table perviouly_published_analyses).
-Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (TablexxPAPERS).
-Moreover, three of them are studies of reanalysis without generating any new data matrices.
-We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models.
-It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mainly due to computational limits of CAT models employed in PhyloBayes.
+XXsupport_published_analyses and Table XXperviouly_published_analyses).
+Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table 1).
+Moreover, three of them are reanalyzing studies without generating any new data matrices.
+Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models.
+It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
 The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices.
@@
 support_published_analyses).
-
-```{r overview of studies}
-table_Papers = read_tsv("../data_processed/tables/previously_published_manuscripts.tsv")
-	kable(table_Papers, caption= "Table XX. Overview of previous phylogenomic studies related to the root position of animal phylogeny")
-
-```
+Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
 
@@
 
-**Fig support_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table perviouly_published_analyses.tsv).
+**Fig XXsupport_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XXperviouly_published_analyses).
 
@@
 One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
-Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework.
-Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models that are implemented in both ML framework in our analyses and it is interesting to compare results between C model to CAT models since it has never been used in any matrices analyzed here.
+Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating].
+Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
 
-Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets).
-We also progressively trimmed each data matrix with different level of outgroup choice if applicable.
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except only LG+C60+F+G models are used large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices).
-We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
-It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive.
+Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1).
+We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview).
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60+F+G models are used in 3 large matrices).
+We found in all cases, WAG/LG+C60+G models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
+It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive.
 The GTR+C models requires to estimate up to over 10,000 (e.g.
@@
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-
-With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.
+Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.
 
@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-For computational efficiency, we first only applied Poisson+CAT models in every matrix with Choanoflagellata as outgroup.
+For computational efficiency, we first only applied Poisson+CAT+G (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups.
 It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
@@
 Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices.
-Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix.
-Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015.
-Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017.
-Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
-We found strong support of Porifera-sister in holozoa matrices and Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups.
-Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.
+Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices.
+Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015.
+Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large].
+The Ctenophora-sister were strongly supported in all other matrices.
+For the matrices have conflicted results between Poisson-CAT and site-homogeneous models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
+We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models.
+Thus, our reanalyzing results are largely consistent with previous analyses, both suggesting that the support of Porifera-sister is increasing with more complex models and closely related outgroups.
+More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.
 
@@
 
-The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models.
-To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
+To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
 1A); T2: Porifera-sister (Fig.
 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig XXPhylogenetic signal).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models.
@@
 
-In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
-phylogenetic_signal B).
-It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny.
-In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal.
+In contrast, the phylogenetic signal of site-heterogenous models changed dramatically comopared to the site-homogenous models.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+XXphylogenetic_signal B).
+It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
+In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig.
+XXphylogenetic_signalA, C).
 We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models.
-The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.
+The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
 
-We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices.
-Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
+We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig XXPairwise_heatmap_animal_root).
+Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
 In other words.
-our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs.
+our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs.
 site-heterogeneous models).
-More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these key data matrices.
-Thus, it is clear that recovering Porifera-sister requires more constrains, including more stringent model selection, outgroup choice, and even with gene sampling.
+More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in Whelan2017_full matrices.
 
@@
 ```
-**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage.
-Proportion of genes supporting each of two alternative hypotheses for each of
+**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
+(A).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
+(B).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
+(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
+The (\Delta$GLS\) values for the genes across each data matrix are provided in TableXX.
 
 ![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Fig XXPairwise_heatmap_animal_root.**
+**Fig XXPairwise_heatmap_animal_root.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices.
+(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
+(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
+The color scale indicates the
 
@@
 
-## The current state of understanding
+## The current state of understanding and future direction
+Resolving the root position of animal phylogeny that diverged ~ 600 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices.
+By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups.
+Three factors contribute significantly to the previous incongruent results.
+First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
+Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models.
+More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgoups.
+Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
+Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister, but that .
+It is still too early to Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.
 
-## Next steps
+## Future direction
 
-## Conclusion
+Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
+Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution.
+First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
+Second, a consensus set of OGs and analytical strages should be generated and fomalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
+Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
+More importantly, we also hope all investigators critically evaluate the competing hypotheses unbiasedly instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
+
+Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
+We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
 

### 2020-01-13 manuscript/manuscript_kernel.R

@@ -50,7 +50,7 @@ focal_matrices =
 						 header = TRUE,
 						 stringsAsFactors = FALSE)
 
-bootstrap_threshold = 95
+bootstrap_threshold = 90
 posterior_prob_threshold = 95
 
 # Load data

### 2020-01-14 manuscript/manuscript.rmd
remove some of Nosenko matrices

@@
 ## The current state of understanding and future direction
-Resolving the root position of animal phylogeny that diverged ~ 600 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices.
-By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups.
+Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices.
+By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
 Three factors contribute significantly to the previous incongruent results.
@@
 Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
-Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister, but that .
-It is still too early to Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.
+Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.
+Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.
 
@@
 First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
-Second, a consensus set of OGs and analytical strages should be generated and fomalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
+Second, a consensus set of OGs and analytical strateges should be generated and formalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
 Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
-More importantly, we also hope all investigators critically evaluate the competing hypotheses unbiasedly instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
-
-Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
+More importantly, we also hope all investigators critically evaluate the competing hypotheses unbiasedly instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
 We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

### 2020-01-16 manuscript/manuscript.rmd
update manuscript

@@
 ## Abstract
-Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages.
-However, a considerable debate still remains regarding to the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life.
-Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies.
-We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
-More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
-We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
+Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
+However, a considerable debate still remains regarding the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult to resolve nodes in animal tree of life.
+Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses with these data in a consistent manner to characterize the differences between studies.
+We find that gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
+More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
+We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
 Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
-Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers.
+Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
 Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
@@
 Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
-In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
+By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
 An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
@@
 Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
-As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
+It has become clear that this is a very difficult phylogenetic challenge, and as the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
 Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
@@
 Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
-Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).
+Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyses; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).
 (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
@@
 Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix.
-CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-One could, though, run a WAG+CAT model.
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
+CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
+One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model.
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).
 If there are no modifiers, then it is implied that site homogeneous models are used.
+Moreover, the Gamma-rate heterogeinity is used in almost every analysis.
 
@@
 
-Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (**Fig XXOverview.C**).
+Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (Fig XXOverview.C).
 Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
@@
 
-Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored.
+Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
 However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
@@
 
-```{r taxon_rectangles }
+```{r taxon_rectangles, fig.height = 12, fig.width = 20 }
 
@@
 Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
-However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig.
-XXBUSCO_annotations).
+However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig  XXBUSCO_annotations).
 
@@
 
-Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
-XXsupport_published_analyses and Table XXperviouly_published_analyses).
-Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table 1).
-Moreover, three of them are reanalyzing studies without generating any new data matrices.
+Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig  XXsupport_published_analyses and Table XXperviouly_published_analyses).
+Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
+Moreover, three studies are secondary studies without any new data matrices.
 Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
 It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
-The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
-support_published_analyses).
+The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig support_published_analyses).
 Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
@@
 ```{r support_published_analyses}
-
+analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
+x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
 analyses_published %>%
-	ggplot( aes(x=model_combined, y=clade, col=result, shape=inference) ) +
+	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
@@
 		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-   xlab("model complexity")
+   xlab("Model complexity") + ylab("Clade")
 ```
@@
 ```{r support_new_analyses}
-
+analyses_new=analyses_new %>%
+mutate(model=recode(model,
+                    `GTR20`="GTR",
+                    `CAT+F81`="Poisson+CAT"))
+
+x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
 analyses_new %>%
-	ggplot( aes(x=model_summary, y=clade, col=result, shape=inference) ) +
+	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
 		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
-```
-
-One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
+       xlab("Model complexity") + ylab("Clade")
+```
+
+One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
 Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
@@
 
-Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1).
+Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1).
 We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview).
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60+F+G models are used in 3 large matrices).
-We found in all cases, WAG/LG+C60+G models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
+We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
 It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive.
@@
 GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models.
-We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G.
+We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60+F.
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
@@
 
-### Comparison of iqtree and phylobayes results
+### Comparison of IQtree and PhyloBayes results
 
@@
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-For computational efficiency, we first only applied Poisson+CAT+G (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups.
+For computational efficiency, we first only applied Poisson+CAT (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups.
 It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
@@
 By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models.
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
 This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
 
-In contrast, the phylogenetic signal of site-heterogenous models changed dramatically comopared to the site-homogenous models.
+In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
 Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
-XXphylogenetic_signal B).
+XXphylogenetic_signal B; tableXXdistribution).
 It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
@@
 XXphylogenetic_signalA, C).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models.
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (tableXXdistribution).
 The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
@@
 
-```{r phylogenetic_signal, fig.width=8, fig.height=12}
+```{r phylogenetic_signal, fig.width=6, fig.height=12}
 #au_tests %>%
@@
 
-x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanozoa_WAG", "Philippe2009_LG", "Philippe2009_Choanozoa_LG", "Philippe2009_LG(phylobayes)", "Philippe2009_Choanozoa_LG(phylobayes)", "Philippe2009_C60", "Philippe2009_Choanozoa_C60", "Philippe2009_F81", "Philippe2009_Choanozoa_F81"))
-y1=factor(Philippe2009$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanimalia_WAG", "Philippe2009_LG", "Philippe2009_Choanimalia_LG", "Philippe2009_LG (PhyloBayes)", "Philippe2009_Choanimalia_LG (PhyloBayes)", "Philippe2009_WAG+C60", "Philippe2009_Choanimalia_WAG+C60", "Philippe2009_Poisson+CAT (PhyloBayes)", "Philippe2009_Choanimalia_Poisson+CAT (PhyloBayes)"))
+y1=factor(Philippe2009$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
 
@@
 
-y2=factor(Ryan2013_est$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
-
-x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanozoa_WAG", "Ryan2013_est_LG", "Ryan2013_Choanozoa_LG",  "Ryan2013_est_C60", "Ryan2013_Choanozoa_C60"))
+y2=factor(Ryan2013_est$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+
+x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanimalia_WAG", "Ryan2013_est_LG", "Ryan2013_Choanimalia_LG",  "Ryan2013_est_WAG+C60", "Ryan2013_Choanimalia_WAG+C60"))
 
@@
                    grepl('Whelan', Matrix))
-y3=factor(Whelan2017$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
-
-x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanozoa_WAG", "Whelan2017_full_LG", "Whelan2017_Choanozoa_LG", "Whelan2017_full_LG(phylobayes)", "Whelan2017_Choanozoa_LG(phylobayes)", "Whelan2017_full_C60", "Whelan2017_Choanozoa_C60", "Whelan2017_full_F81", "Whelan2017_Choanozoa_F81"))
-
-p1 <- ggplot(Philippe2009, aes(x = x1, y = percent, fill=y1)) +
+y3=factor(Whelan2017$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
+
+x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanimalia_WAG", "Whelan2017_full_LG", "Whelan2017_Choanimalia_LG", "Whelan2017_full_LG (PhyloBayes)", "Whelan2017_Choanimalia_LG (PhyloBayes)", "Whelan2017_full_LG+C60", "Whelan2017_Choanimalia_LG+C60", "Whelan2017_full_Poisson+CAT (PhyloBayes)", "Whelan2017_Choanimalia_Poisson+CAT (PhyloBayes)"))
+
+p1 <- ggplot(Philippe2009, aes(x = x1, y = Percentage, fill=y1)) +
+  geom_bar(stat = "identity") +
+  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+
+p2 <- ggplot(Ryan2013_est, aes(x = x2, y = Percentage, fill=y2)) +
+  geom_bar(stat = "identity") +
+  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
+  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+
+p3 <- ggplot(Whelan2017, aes(x = x3, y = Percentage, fill=y3)) +
   geom_bar(stat = "identity") +
   #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
-
-p2 <- ggplot(Ryan2013_est, aes(x = x2, y = percent, fill=y2)) +
-  geom_bar(stat = "identity") +
-  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
-
-p3 <- ggplot(Whelan2017, aes(x = x3, y = percent, fill=y3)) +
-  geom_bar(stat = "identity") +
-  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
   scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
@@
                     labels = c("A", "B", "C"),
-                    ncol = 1, nrow = 3, widths=c(2,1,2), heights=c(1,1,1), common.legend = TRUE, legend = "right")
+                    ncol = 1, nrow = 3,  common.legend = TRUE, legend = "right")
 
 figure
+```
+
+```{r phylogenetic signal distribution}
+table1 = read_tsv("../trees_new/AU_test/summary.tsv")
+	kable(tableXXdistribution, caption= "Table XX. Distribution of phylogenetic signal for different models and outgroup choice")
+
 ```
@@
 ![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Fig XXPairwise_heatmap_animal_root.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices.
+**Fig XXPairwise_heatmap_animal_root.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
 (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
 (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the
+The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
+The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
 
@@
 Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
-More importantly, we also hope all investigators critically evaluate the competing hypotheses unbiasedly instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
+More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
 We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
@@
 
-We retreived matrices from each publication (Table XX), storing the raw data in this manuscript's version control repository.
+We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
 We manually edited some minor formatting to make the batch processing of the matrices *en masse*, e.g.
@@
 
-We programatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
+We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
 We also use a table where manual entries were needed (Supp Table XX, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g.
@@
 
-We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond blastp parameters.
+We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
 We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self).
 We ran BUSCO with default parameters for all sequences against the provided metazoa gene set.
-We also ran a BLAST+ blastp search against the SwissProt [cite] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+We also ran a BLAST+ blastp search against the SwissProt [@boeckmann2003swiss] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
 
@@
 #### Phylogenetic analyses in IQtree
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root of animal tree of life (e.g.
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
 site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C models in the model testing.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C models in the model testing.
 For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters.
@@
 
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60 and associated best-fit model identified above.
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above.
 Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
@@
 2009 [@hejnol2009assessing], since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
-It should be noted that the main conclusion of Simion et al.
-2008 was also based on selection of 25% of genes for their jackknife approach.
+It should be noted that the main conclusion of Simion *et al*.
+was also based on selection of 25% of genes for their jackknife approach.
 
@@
 
-For results of choanozoa matrices indicated a strong support that sponges are the sister group to the remaining Metazoa using CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
 For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.

### 2020-01-16 manuscript/manuscript.rmd
update manuscript and change structures of trees_new

@@
 We find that gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
-More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
+More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model and inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
 We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
@@
 
-## Introduction
+## Main
 
 Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
-However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses.
+However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig.
+1) emerging as the two primary hypotheses.
 Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
@@
 An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
-However, a wide consensus of animl origin has not been reached despite multiple efforts.
+However, a wide consensus of animal origin has not been reached despite multiple efforts.
 
@@
 ![](figures/Figure_overview.png)
-**Fig XXOverview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
+**Fig.
+1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
 (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
@@
 The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
-A detailed description of each component is in Supplementary text.
+A detailed description of each component is in Supplementary Results.
 Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
 Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyses; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).
-(4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
-Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models [@feuda2017improved].
 
@@
 High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of animal phylogeny (Table 1).
-Although the number of sampled genes might be different across studies, it is reasonable to expect that a core set of orthology genes (OGs) should be present and conserved across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
-Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (for details see Supplementary text section ).
+Although the number of sampled genes might be different across studies, it is reasonable to expect that a core set of orthologs should be present and conserved across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
+Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of orthologs (for details see Supplementary Results).
 
@@
 
-Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (Fig XXOverview.C).
+Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (Fig.
+1C).
 Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
@@
 ```{r overview data matrices}
-table1 = read_tsv("../data_processed/tables/study_summary.tsv")
-	kable(table1, caption= "Table 1. Overview of data matrices used in this study")
-
-```
+tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
+	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
+
+```
+
+## Results and Discussion
 
@@
 Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
-However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition).
+However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Extended Data Fig.
+1).
 These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
-Importantly, several earlier studies only include less than five ctenophore species (Fig XXTaxon_composition).
+Importantly, several earlier studies only include less than five ctenophore species (Extended Data Fig.
+1).
 The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
 In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
-Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices (Table1, Fi.
-XXTaxon_composition).
-
-```{r taxon_rectangles, fig.height = 12, fig.width = 20 }
+Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.
+
+```{r taxon_rectangles, fig.height = 18, fig.width = 20 }
 
@@
  scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
-	#scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
+	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
 	facet_wrap( ~ matrix )
@@
 
-**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by taxon sampling.
+**Fig.
+Extended Data Fig.
+1.** Each of the primary matrices considered here, color coded by taxon sampling.
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
@@
 
-We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected.
-The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
-Moreover, the BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
+We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
+2A).
+The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
+The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
 Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
-However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig  XXBUSCO_annotations).
+However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig.
+2B).
+Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
+2C).
+We found Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
+2D).
+We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
 
@@
 Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
-The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs.
-However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap).
+The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of orthologs.
+However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig.
+2).
 Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
@@
 
-```{r gene_composition}
+```{r gene_composition,, fig.height = 6, fig.width = 10}
+matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
 
@@
 	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
+	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
+
+matrix_summary=matrix_summary %>%
+  mutate(matrix=recode(matrix,`NA`="Simion2017"))
+
+matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+
+p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
+
+p3 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
 	coord_fixed()
 
-matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "NA", "Whelan2017_full", "Whelan2017_strict") )
-
-p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of  BUSCO genes")
-
-figure <- ggarrange(p1, p2,
-                    labels = c("A", "B"),
-                    ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")
-figure
-
-```
-
-**Fig XXBUSCO_annotations.** BUSCO annotation and representation in each data matrix.
+p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
+
+ ggarrange(p1, p2,p3, p4,
+                    labels = c("A", "B", "C", "D"),
+                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
+
+```
+
+**Extended Data Fig.
+2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
 (A).
@@
 The percentage BUSCO annotations in each matrix.
+(C).
+The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
+(D).
+The percentage  ribosomal genes in each matrix.
 
@@
 
-**Fig XXAlignment overlap.** Pairwise overlap between each of the primary matrices considered here.
+**Fig.
+2.** Pairwise overlap between each of the primary matrices considered here.
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
@@
 
-Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig  XXsupport_published_analyses and Table XXperviouly_published_analyses).
+Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+3 and Supplementary Table 2).
 Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
@@
 The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig support_published_analyses).
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
+3).
 Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
@@
 
-**Fig XXsupport_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XXperviouly_published_analyses).
+**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).
 
@@
 ```
+**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table 3).
 
@@
 Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating].
-Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
+4; Supplementary Table 3).
+Importantly, we also include a CAT-like, site-heterogeneous C10-C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
 
 Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1).
-We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview).
+We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
 We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
-We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree.
+We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
 It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive.
@@
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.
-
+Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig.
+4; Supplementary Table 3).
+
+**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
 ```{r modelfinder}
-
 analyses_new %>%
@@
 ```
-
-Table XXModelfinder. The models selected by modelfinder for each matrix.
 
@@
 1A); T2: Porifera-sister (Fig.
-1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig XXPhylogenetic signal).
+1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig.
+5).
 By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
@@
 Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
-XXphylogenetic_signal B; tableXXdistribution).
+5B; Supplementary Table 4).
 It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
 In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig.
-XXphylogenetic_signalA, C).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (tableXXdistribution).
+5A, C).
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
 The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
 
-We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig XXPairwise_heatmap_animal_root).
+We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig.
+Extended Data Fig.
+3).
 Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
@@
   scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Philippe2009", y = "Catogary of phylgoenetic signal")
 
@@
   scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
+  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Ryan2013_est", y = "Catogary of phylgoenetic signal")
 
@@
   #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=6),axis.text.x = element_text(angle=45, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
-
-figure <- ggarrange(p1, p2, p3,
-                    labels = c("A", "B", "C"),
-                    ncol = 1, nrow = 3,  common.legend = TRUE, legend = "right")
-
-figure
-```
-
-```{r phylogenetic signal distribution}
-table1 = read_tsv("../trees_new/AU_test/summary.tsv")
-	kable(tableXXdistribution, caption= "Table XX. Distribution of phylogenetic signal for different models and outgroup choice")
-
-```
-**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
+  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Ryan2013_est", y = "Catogary of phylgoenetic signal")
+
+ggarrange(p1, p2, p3,
+                    labels = c("A", "B", "C"), widths=c(10,6,10),
+                    ncol = 1, nrow = 3,  common.legend = TRUE, legend = "right", align="hv")
+
+```
+
+**Fig.5** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
 (A).
@@
 (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
-The (\Delta$GLS\) values for the genes across each data matrix are provided in TableXX.
+The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
 
 ![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Fig XXPairwise_heatmap_animal_root.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
+**Extended Data Fig.
+3.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
 (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
@@
 They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Fig XXRecoidng_summary).
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
+4).
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
 When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure XXRandom_recodingB).
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
+5).
 Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
+
+![](figures/Figure_random_rocoding.png)
+
+**Extended Data Fig.
+4.** (A) The two alternative hypotheses for deep animal relationships considered here.
+Relationships that are not part of these hypotheses are shown as unresolved polytomies.
+(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+![](figures/Figure_recoding_summary.png)
+**Extended Data Fig.
+6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+
+![](figures/Figure_recoding_pp.png)
+
+**Extended Data Fig.
+5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
+The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
+The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
+Cells are color coded as in Supplemental Fig XXRecoidng_summary.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
+The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
 
@@
 First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
-Second, a consensus set of OGs and analytical strateges should be generated and formalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
+Second, a consensus set of orthologs and analytical strateges should be generated and formalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
 Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
@@
 
-All files and bioinformatic commmands associated with this analysis are available at https://github.com/caseywdunn/animal_root
+All files and bioinformatic commands associated with this analysis are available at https://github.com/caseywdunn/animal_root
 
@@
 We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also use a table where manual entries were needed (Supp Table XX, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g.
+We also use a table where manual entries were needed (Supplementary Table 5, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g.
 authors of original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see Table XX(reconciliation/taxonomy_info/taxon_table.tsv).
+For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (reconciliation/taxonomy_info/taxon_table.tsv).
 
@@
 We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See Table XX [partition_network_summary table] for a summary tally of each part of the comparison.
+See Supplementary Table 7 [partition_network_summary table] for a summary tally of each part of the comparison.
 
@@
 The 20 state matrices are the same across all analyses since they are not recoded.
-Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.
+Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
+4) give insight into the technical variance of the inference process.
 
@@
 
-The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 # Supplemental Information
-## Models of molecular evolution
+
+## Supplementary Results
+
+### Models of molecular evolution
 
@@
 
-## Details of published analyses
-
-### Dunn *et al.* 2008
+### Details of published analyses
+
+#### Dunn *et al.* 2008
 
@@
 
-### Philippe *et al.* 2009
+#### Philippe *et al.* 2009
 
@@
 
-### Hejnol et al. 2009
+#### Hejnol et al. 2009
 
@@
 
-### Pick *et al.* 2010
+#### Pick *et al.* 2010
 
@@
 
-### Nosenko *et al.* 2013
+#### Nosenko *et al.* 2013
 Nosenko *et al.* [@nosenko2013deep] added Expressed Sequence Tag (EST) data for 9 species of non-bilaterian metazoans (7 sponges).
@@
 
-OGs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect].
+Orthologs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect].
 They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix.
@@
 
-### Ryan *et al.* 2013
+#### Ryan *et al.* 2013
 Ryan *et al.* sequenced the first ctenophore genome of *Mnemiopsis leidyi*.
@@
 They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson-CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix.
-The OGs were constructed based on the method of Hejnol *et al.* 2009.
+The Orthologs were constructed based on the method of Hejnol *et al.* 2009.
 For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-baed analyses.
@@
 
-### Moroz *et al.* 2014
+#### Moroz *et al.* 2014
 
@@
 Two phylogenomic matrices were generated, the first set was represented by two ctenophore species, whereas the other set contained improved ctenophore sampling (10 taxa, Moroz2013_3d).
-Orthology determination employed in HaMStR [@ebersberger2009hamstr] using 1,032 "model organism" single-copy OGs.
+Orthology determination employed in HaMStR [@ebersberger2009hamstr] using 1,032 "model organism" single-copy orthologs.
 Sequence were then trimmed and aligned.
@@
 
-### Borowiec *et al.* 2015
-
-Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 OGs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
-The OGS were constructed using OrthologID pipeline [@chiu2006orthologid].
-After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080 OGs (Total 1080).
+#### Borowiec *et al.* 2015
+
+Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
+The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid].
+After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080  (Total 1080).
 The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes.
@@
 
-### Whelan et al. 2015
+#### Whelan et al. 2015
 Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
@@
 
-### Chang et al. 2015
+#### Chang et al. 2015
 
@@
 
-### Pisani *et al.* 2015
+#### Pisani *et al.* 2015
 
@@
 
-### Feuda *et al.* 2017
+#### Feuda *et al.* 2017
 
@@
 
-### Whelan and Halanych 2016
+#### Whelan and Halanych 2016
 
@@
 
-### Whelan *et al.* 2017
+#### Whelan *et al.* 2017
 
@@
 
-### Simion *et al.* 2017
+#### Simion *et al.* 2017
 
@@
 
-## Data-recoding methods
+### Data-recoding methods
 
 Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
-In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig XXRandom_recodingA).
+In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig.
+Extended Data Fig.
+4).
 Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
@@
 Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes.
-This provides 32 analyses that they report in their Table 3 and that we present graphically here as Supplemental Fig XXRecoidng_summary.
-There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Fig XXRecoidng_summary).
+This provides 32 analyses that they report in their Table 3 and that we present graphically here as Extended Data Fig.
+6.
+There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Extended Data Fig.
+6).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
-This leaves eight CAT-GTR+G analyses (Fig XXRecoding_pp).
+This leaves eight CAT-GTR+G analyses (Extended Data Fig.
+5).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
@@
 Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
-Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Fig XXRecoding_pp).
-The only statistic that stands out for this one analysis is that it has the highest maxdiff (Fig XXRecoding_pp), indicating that it did not converge as well as other analyses.
+Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
+5).
+The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig.
+5), indicating that it did not converge as well as other analyses.
 
@@
 When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig XXRandom_recodingB).
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
+4).
 
@@
 
-![](figures/Figure_random_rocoding.png)
-
-Fig XXRandom_recoding.
-(A) The two alternative hypotheses for deep animal relationships considered here.
-Relationships that are not part of these hypotheses are shown as unresolved polytomies.
-(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
-
-![](figures/Figure_recoding_summary.png)
-**Fig XXRecoidng_summary.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-
-![](figures/Figure_recoding_pp.png)
-
-**Fig XXRecoding_pp** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
-These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
-The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
-The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Cells are color coded as in Supplemental Fig XXRecoidng_summary.
-Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
-The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
-
-## Matrix mapping
+## Supplementary Methods
+
+### Matrix mapping
 
@@
 
+## Supplementary Tables
+```{r phylogenetic signal distribution}
+tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
+	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
+
+```
+
 ## References

### 2020-01-17 manuscript/manuscript.rmd
update the manuscript

@@
 
-Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated.
+Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated since all these studies rely on their own method for orthology inference.
 Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
@@
 
-Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
-3 and Supplementary Table 2).
-Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
-Moreover, three studies are secondary studies without any new data matrices.
-Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
-It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
-The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig.
-3).
-Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
-
 ```{r support_published_analyses}
@@
 ```
-
 **Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).
+
+Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+3 and Supplementary Table 2).
+Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
+Moreover, three studies are secondary studies without any new data matrices.
+Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
+It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
+The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
+3).
+Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
 
@@
 GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models.
-We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60+F.
+We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
@@
 5).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
@@
 **Extended Data Fig.
-3.** Pairwise congruent comparisons of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
+3.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
 (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
@@
 By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
-Three factors contribute significantly to the previous incongruent results.
+Three factors contribute significantly to the previous incongruent results based on our analyses.
 First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
 Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models.
-More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgoups.
+More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice.
 Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
@@
 First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
-Second, a consensus set of orthologs and analytical strateges should be generated and formalized to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
+Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
 Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
-More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
+More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
+Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
 We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
@@
 #### Phylogenetic analyses in IQtree
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root of animal position, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
 Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
@@
 
-Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
 For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
@@
 #### phylogenetic distribution of support
-To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices.
+To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
 We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models.
+For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
 The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.

### 2020-02-04 manuscript/manuscript.rmd
word smithing

@@
 output:
+  pdf_document: default
+  word_document: default
   html_document: default
-  word_document: default
-  pdf_document: default
   github_document: default
@@
 ## Abstract
-Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
-However, a considerable debate still remains regarding the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult to resolve nodes in animal tree of life.
-Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses with these data in a consistent manner to characterize the differences between studies.
-We find that gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices.
-More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model and inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases.
+Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
+However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
+This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution.
+Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes.
+These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used.
+As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool phylobayes.
+Most other conditions supported Ctenophora-sister.
+It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups.
+We perform new analyses with these data to characterize the differences between matrices under consistent conditions.
+Our reanalyses are generally consistent with previous results.
+We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister.
+This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
 We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
-Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses.
+We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic.
 Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
-Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
+This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
 
@@
 By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
+An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
 However, a wide consensus of animal origin has not been reached despite multiple efforts.
@@
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
-Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
-It has become clear that this is a very difficult phylogenetic challenge, and as the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
+Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
+It has become clear that this is a very difficult phylogenetic challenge.
+As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
 Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
-Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
-We hope that this effort provides an integrative overview of the challenge and provides direction for future studies.
+Here we synthesize data and results from all phylogenomic analyses published befor 2018 that tested Ctenophora-sister and Porifera-sister, and for which we could obtain matrix partition data.
+We reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
+This provides an integrative overview of the challenges posed by this question and provides direction for future studies.
 We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
@@
 Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
+The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the amino acid equilibrium frequencies $\Pi$.
 A detailed description of each component is in Supplementary Results.
-Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
-Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyses; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).
-
-Models can be assembled by selecting different options for all these different components.
-The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency.
+Briefly, the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand).
+The rate of excahnge can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites.
+The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary.
+Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software tool CAT and has been widely applied to questions of deep animal phylogeny, and the C models of iqtree that have until now not been applied to the problem.
+
+Models of molecular evolution are assembled by selecting different options for all these different components.
+The models that are applied in practice are heavily influenced by computational costs, as well as convention.
+For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices.
 LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
 This is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix.
-CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
+IN this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
 One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model.

### 2020-02-05 manuscript/manuscript.rmd
more editing

@@
 Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
-Here we synthesize data and results from all phylogenomic analyses published befor 2018 that tested Ctenophora-sister and Porifera-sister, and for which we could obtain matrix partition data.
+Here we synthesize data and results from phylogenomic analyses that tested Ctenophora-sister and Porifera-sister.
 We reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
@@
 
-High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of animal phylogeny (Table 1).
-Although the number of sampled genes might be different across studies, it is reasonable to expect that a core set of orthologs should be present and conserved across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]).
-Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of orthologs (for details see Supplementary Results).
+High-throughput sequencing allows investigators interested in the root of animal phylogeny to assemble matrices with hundreds or thousands of genes (Table 1).
+Surprisingly, gene overlap between studies has rarely been assessed.
+Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), animal root studies relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
+These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered.
+As a result, the genes selected for analysis can vary widely between studies.
 
@@
 
-Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (Fig.
+Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig.
 1C).
 Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated-- Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
 It has been suggested the inclusion of more distant related (e.g.
 Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic].
-In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than outgroup sampling.
-This may be because results have tended to be more sensitive to outgroup sampling.
+In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
 
@@
 
-Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving].
+Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving].
 Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie].
@@
 ## Overview of published analyses
-Here, we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies (Table 1).
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies (Table 1).
+These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.
 

### 2020-02-25 manuscript/manuscript.rmd
wording

@@
 output:
+  github_document: default
   pdf_document: default
@@
   html_document: default
-  github_document: default
 ---

### 2020-02-25 manuscript/manuscript.tex
add Poisson_C60 from phylobayes results

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\begin{document}

\hypertarget{rooting-the-animal-tree-of-life}{%
\section{Rooting the animal tree of
life}\label{rooting-the-animal-tree-of-life}}

Yuanning Li\textsuperscript{1,3}, Xingxing Shen\textsuperscript{3},
Benjamin Evans\textsuperscript{2}, Antonis Rokas\textsuperscript{3}, and
Casey W. Dunn\textsuperscript{1}*

\textsuperscript{1}Department of Ecology and Evolutionary Biology, Yale
University

\textsuperscript{2}Yale Center for Research Computing, Yale University

\textsuperscript{3}Department of Biological Science, Vanderbilt
University

* Corresponding author,
\href{mailto:casey.dunn@yale.edu}{\nolinkurl{casey.dunn@yale.edu}}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Phylogenomic studies based on hundreds to thousands of genes have
greatly improved our understanding of evolutionary relationships among
major animal lineages. However, a considerable debate still remains
regarding the animal origin, with Ctenophora-sister and Porifera-sister
emerging as the two primary hypotheses, comprising one of the most
difficult to resolve nodes in animal tree of life. Here, we
systematically explore the mechanisms that potentially cause the
incongruent results by synthesizing data and results from all previous
phylogenomic analyses, and perform new analyses with these data in a
consistent manner to characterize the differences between studies. We
find that gene sampling varies across studies, indicating that different
regions of the genomes are sampled with little overlap among data
matrices. More importantly, we find Porifera-sister can only be
recovered with more constraints, including using site-heterogeneous CAT
model and inclusive of the most closely related outgroups in several
matrices, whereas Ctenophora-sister is recovered in all other cases. We
next quantify the support of phylogenetic signal over the two hypotheses
and find substantial differences in the distribution of phylogenetic
signal between models and outgroup choice (especially CAT models). Last,
we show that recoding can be a problematic method for addressing
compositional variation in phylogenetic analyses. Our results provide a
comprehensive overview of the current understanding of the first
diverging animal branch and why different studies yield different
answers. Rather than embracing one hypothesis over another, we hope our
efforts provides an integrative overview of the challenge and provides
direction for future studies to address difficult phylogenetic problems.

\hypertarget{main}{%
\subsection{Main}\label{main}}

Phylogenomic analyses with genome-scale data have the potential to
revolutionize our understanding of the tree of life. However, over the
past decade, there has been considerable debate about the position of
the root of the animal phylogeny, with Ctenophora-sister (comb jelly)
and Porifera-sister (sponges) (Fig. 1) emerging as the two primary
hypotheses. Historically, there was little debate about the root of the
animal tree of life and Porifera-sister was widely accepted though
rarely tested. By contrast, there has long been uncertainty about the
relationship of Ctenophora to other animals\textsuperscript{1}. An
understanding of the first diverging animal lineage is critical for
scientists to reconstruct events that occurred early in animal
evolution, especially for the evolution of complex structures such as
cell types, nervous systems, and developmental patterns. However, a wide
consensus of animal origin has not been reached despite multiple
efforts.

The first phylogenomic study to include ctenophores\textsuperscript{2}
suggested a new hypothesis, now referred to as Ctenophora-sister, that
ctenophores are our most distant animal relative. Since then many more
studies have been published and extensive reanalyzed, some supporting
Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It
has become clear that this is a very difficult phylogenetic challenge,
and as the problem has become better characterized, it has become an
interesting test-case to phylogenetic biologists beyond those concerned
with this particular biological problem. Work has been hindered, though,
because it has been difficult to directly compare results across studies
and synthesize findings to understand the broader patterns of support.
Here we synthesize data and results from all previous phylogenomic
analyses that tested Ctenophora-sister and Porifera-sister, and
reanalyze these data using standardized methods, and perform new
analyses to characterize differences between studies. We hope that this
effort provides an integrative overview of the challenge and provides
direction for future studies. We also hope that the work we have
conducted here, including consolidating all the datasets in one place
with consistent formats and species names, will enhance the technical
value of this interesting question to methods-focused investigators that
look to develop methods to address difficult phylogenetic problems.

\includegraphics{figures/Figure_overview.png} \textbf{Fig. 1.} (A) The
Ctenophora-sister hypothesis posits that there is a clade (designated by
the orange node) that includes all animals except Ctenophora, and that
Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis
posits that there is a clade (designated by the green node) that
includes all animals except Porifera, and that Porifera is sister to
this clade. Testing these hypotheses requires evaluating the support for
each of these alternative nodes. (C) The animals and their outgroup
choice, showing the three progressively more inclusive clades
Choanimalia, Holozoa, and Opisthokonta (designated by the blue node).
(D) A ctenophore. (E) A sponge.

\hypertarget{variation-across-studies}{%
\subsection{Variation across studies}\label{variation-across-studies}}

\hypertarget{models-of-molecular-evolution}{%
\subsubsection{Models of molecular
evolution}\label{models-of-molecular-evolution}}

Models of molecular evolution have several components that each consider
different aspects of the evolutionary process. The models that have been
used to model protein evolution in studies of the deep phylogeny have
largely differed according to three components: the exchangeability
matrix \(E\), the rate of evolution, and the state equilibrium
frequencies \(\Pi\). A detailed description of each component is in
Supplementary Results. Briefly, (1) the exchangeability matrix \(E\)
describes the rate at which one amino acid changes to another.
Exchangeability matrices have been used in the studies under
consideration here include: Poisson (or F81), WAG, LG, GTR, as they have
been used regularly in phylogenomic analyses; (2) While the
exchangeability matrix describes the relative rate of different changes
between amino acids, the actual rate can be further scaled, such as site
homogeneous rates and Gamma-rate heterogeneity; (3) the vector of
equilibrium frequencies \(\Pi\) describes the stationary frequency of
amino acids, such as empirical site-homogeneous, estimated
site-homogeneous, CAT site-heterogeneous\textsuperscript{3}, and C10-C60
models (C models).

Models can be assembled by selecting different options for all these
different components. The models that are applied in practice area
heavily influenced by engineering and computational costs, as well as
convention. For example, on the questions considered here, Poisson and
GTR exchangeability matrices have only been used in combination with CAT
site-heterogeneous models of equilibrium frequency. LG and WAG
exchangeability matrices have only been used with site homogeneous
estimates of equilibrium frequency. This is further confused by the
abbreviations that are used for models. Papers often discuss CAT and WAG
models as if they are exclusive, but these particular terms apply to
non-exclusive model components-- CAT refers to variation across sites
and WAG a particular exchangeability matrix. CAT is generally shorthand
for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium
frequency estimation. One could, though, run a WAG+CAT model, although
the computation time is much longer than Poisson+CAT model. To avoid
confusion on this point, we always specify the exchangeability matrix
first, followed by modifiers that describe the accommodation of
heterogeneity in equilibrium frequencies (\emph{e.g.}, CAT). If there
are no modifiers, then it is implied that site homogeneous models are
used. Moreover, the Gamma-rate heterogeinity is used in almost every
analysis.

\hypertarget{gene-sampling}{%
\subsubsection{Gene sampling}\label{gene-sampling}}

High-throughput sequencing is leading to exciting new data matrices for
phylogenomic analyses, potentially including hundreds or thousands of
genes to explore the root of animal phylogeny (Table 1). Although the
number of sampled genes might be different across studies, it is
reasonable to expect that a core set of orthologs should be present and
conserved across most animal genomes (\emph{e.g.}, animal BUSCO
genes\textsuperscript{4}). Surprisingly, gene composition and matrix
overlap has been rarely assessed across studies and most studies relied
on their own bioinformatic pipeline for determination of orthologs (for
details see Supplementary Results).

\hypertarget{outgroup-and-ingroup-taxon-sampling}{%
\subsubsection{Outgroup and ingroup taxon
sampling}\label{outgroup-and-ingroup-taxon-sampling}}

Outgroup selection (the taxa used to root the ingroup taxa) can
potentially affect the phylogenetic inference of animal evolution (Fig.
1C). Several studies have progressively removed more distantly related
outgroup taxa to test the effect of outgroup selection to ingroup
topology\textsuperscript{5,6}. Three progressively more inclusive clades
have often been investigated, including Choanomalia (Metazoa plus most
closely related Choanoflagellata), Holoza (Choanimalia plus more
distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has
been suggested the inclusion of more distant related (e.g.~Fungi)
outgroup taxa tend to recover Ctenophora-sister, whereas increasing
support of Porifera-sister often recovered when more closely related
outgroup taxa were used with site-heterogeneous CAT
models\textsuperscript{6}. In contrast of outgroup selection,
sensitivity to ingroup sampling has received less attention than
outgroup sampling. This may be because results have tended to be more
sensitive to outgroup sampling.

\hypertarget{data-recoding}{%
\subsubsection{Data-recoding}\label{data-recoding}}

Compositional heterogeneity (lineage-specific differences in amino acid
frequencies) was long thought has a major effect in resolving the animal
tree of life, especially for the deep animal
phylogeny\textsuperscript{7}. Variation in evolutionary rates across
taxa can lead to compositional heterogeneity that may lead to long
branch attraction of sponges or ctenophores\textsuperscript{8}. Feuda
\emph{et al}.\textsuperscript{9} has proposed to use data-recoding
method to reduce compositional heterogeneity in matrices. By discarding
information on which specific amino acid is found at each site in each
species and instead focusing on which groups those amino acids belong
to, they sought to reduce the impact of differences between species in
which particular amino acids within those groups are most frequent. By
reanalyzing Chang et al.~2015\textsuperscript{10} and Whelan et
al.~2015\textsuperscript{11}, they report that posterior predictive
analyses indicate recoded analyses have better model adequacy than
non-recoding analyses, and that Porifera-sister was strongly favored
under all recoding strategies using GTR-CAT models\textsuperscript{9}.
However, how the recoding methods impact phylogenetic inference has not
been well explored. Moreover, a recent simulation study has suggested
that non-recoding approaches significantly outperformed recoding
strategies\textsuperscript{12}. Thus, it is important to reevalute the
performance of the impact of data-recoding compared to non-recoding
methods in animal phylogeny.

\hypertarget{overview-of-published-analyses}{%
\subsection{Overview of published
analyses}\label{overview-of-published-analyses}}

Here, we synthesize data and results from all previous phylogenomic
analyses that tested Ctenophora-sister and Porifera-sister, including 17
data matrices from 12 studies (Table 1).

\begin{longtable}[]{@{}lllrrrlll@{}}
\caption{Table 1. Overview of data matrices used in this
study}\tabularnewline
\toprule
Study & Matrix & Clade & Number of gene partitions & Number of sites &
Number of taxa & Infered sister lineage & Journal &
Citations\tabularnewline
\midrule
\endfirsthead
\toprule
Study & Matrix & Clade & Number of gene partitions & Number of sites &
Number of taxa & Infered sister lineage & Journal &
Citations\tabularnewline
\midrule
\endhead
Dunn2008 & Dunn2008 & Opisthokonta & 150 & 21152 & 64 &
Ctenophora-sister & Nature & NA\tabularnewline
Philippe2009 & Philippe2009 & Opisthokonta & 129 & 30257 & 55 &
Porifera-sister & Current Biology & NA\tabularnewline
Hejnol2009 & Hejnol2009 & Opisthokonta & 1487 & 270580 & 94 &
Ctenophora-sister & Proc. Biol. Sci. & NA\tabularnewline
NA & Hejnol2009\_subsampled & Opisthokonta & 844 & 55594 & 94 &
Ctenophora-sister & NA & NA\tabularnewline
Nosenko2013 & Nosenko2013\_nonribo\_9187 & Opisthokonta & 35 & 9187 & 63
& Ctenophora-sister & Molecular Phylogenetics and Evolution &
NA\tabularnewline
Nosenko2013 & Nosenko2013\_ribo\_14615 & Opisthokonta & 87 & 14615 & 71
& Porifera-sister & NA & NA\tabularnewline
Ryan2013 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 61 &
Ctenophora-sister & Science & NA\tabularnewline
Moroz2014 & Moroz2014\_3d & Opisthokonta & 114 & 22772 & 46 &
Ctenophora-sister & Nature & NA\tabularnewline
Whelan2015 & Whelan2015\_D1 & Opisthokonta & 251 & 59733 & 76 &
Ctenophora-sister & PNAS & NA\tabularnewline
Whelan2015 & Whelan2015\_D10 & Opisthokonta & 210 & 59733 & 70 &
Ctenophora-sister & NA & NA\tabularnewline
Whelan2015 & Whelan2015\_D20 & Opisthokonta & 178 & 81008 & 65 &
Ctenophora-sister & NA & NA\tabularnewline
Borowiec2015 & Borowiec2015\_Best108 & Choaimalia & 108 & 41808 & 36 &
Ctenophora-sister & BMC Genomics & NA\tabularnewline
Borowiec2015 & Borowiec2015\_Total1080 & Choaimalia & 1080 & 384981 & 36
& Ctenophora-sister & NA & NA\tabularnewline
Pisani2015 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 60 &
Porifera-sister & PNAS & NA\tabularnewline
NA & Whelan\_D6 & Opisthokonta & 115 & 33403 & 70 & Porifera-sister & NA
& NA\tabularnewline
Chang2015 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
Ctenophora-sister & PNAS & NA\tabularnewline
Simion2017 & Simion2017 & Opisthokonta & 1719 & 401632 & 97 &
Porifera-sister & Current Biology & NA\tabularnewline
NA & Simion2017\_subsampled & Opisthokonta & 1719 & 110602 & 97 &
Porifera-sister & NA & NA\tabularnewline
Feuda2017 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
Ctenophora-sister & Current Biology & NA\tabularnewline
NA & WhelanD20 & Opisthokonta & 178 & 81008 & 65 & Porifera-sister & NA
& NA\tabularnewline
Whelan2017 & Whelan2017b\_full & Holozoa & 212 & 68062 & 80 &
Ctenophora-sister & Nature Ecology and Evolution & NA\tabularnewline
Whelan2017 & Whelan2017b\_strict & Choaimalia & 59 & 49388 & 76 &
Ctenophora-sister & NA & NA\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{results-and-discussion}{%
\subsection{Results and Discussion}\label{results-and-discussion}}

\hypertarget{matrix-taxon-composition}{%
\subsubsection{Matrix taxon
composition}\label{matrix-taxon-composition}}

Taxa sampling plays a critical role in phylogenetic inference,
especially for the taxa whose phylogenetic position is poorly
understood. However, several data matrices have relatively skewed taxon
sampling, with more than 50\% taxa sampled in bilaterians compared to
other major animal lineages (Extended Data Fig. 1). These skewed data
matrices include two studies constructed from ESTs
(Dunn2008\textsuperscript{2} and Hejnol2009\textsuperscript{13}) and one
study from whole-genome data (Borrowiec2015\textsuperscript{14}).
Importantly, several earlier studies only include less than five
ctenophore species (Extended Data Fig. 1). The rest data matrices have
relatively balanced taxon composition, although the total number of
sampled taxon varies a lot. In general, increasing taxon sampling is
broadly accepted to improve phylogenetic inference and aid in the
placement of the earliest diverging animal lineage. Thus, it should be
noted that Whelan2017\textsuperscript{15} (has the most Ctenophore taxa)
and Simion2017\textsuperscript{16} (contains the most taxon sampling)
have much broader taxon sampling than other matrices.

\includegraphics{manuscript_files/figure-latex/taxon_rectangles-1.pdf}

\textbf{Fig. Extended Data Fig. 1.} Each of the primary matrices
considered here, color coded by taxon sampling. Horizontal size is
proportional to the number of genes sampled, vertical size to the number
of taxa sampled.

\hypertarget{matrix-gene-composition-and-overlap}{%
\subsubsection{Matrix gene composition and
overlap}\label{matrix-gene-composition-and-overlap}}

Gene sampling also plays an essential role in phylogenomic analyses, and
most often rely on the concatenation of sampled genes for deep animal
phylogeny. It has been suggested that the concatenation assumes that
phylogenetic signal from genes that do not share the species phylogeny
will be overwhelmed by the signal from the majority of genes whose
genealogy mirrors that of the species evolutionary
history\textsuperscript{17}. However, large phylogenomic data matrices
potentially contain genes with conflicting or weak phylogenetic signal
that can lead to incongruence phylogenomic results, especially when
analyzing deeper node represents ancient divergence in the
tree\textsuperscript{18}. Thus, it is important to evaluate the gene
composition and matrix overlap to explore level of congruence of gene
sampling across different matrices.

We first used the Metazoa BUSCO genes\textsuperscript{{\textbf{???}}} to
evaluate gene composition from all data matrices and the result is
unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains
978 single-copy orthologs (identified from 64 animal taxa) that are most
likely shared by most animal genomes, and are commonly used to assess
the completeness of genome or transcriptomic
assemblies\textsuperscript{4}. The BUSCO genes also are frequently used
as markers in many large-scale phylogenomic studies (e.g.,
yeasts\textsuperscript{19}, spiders\textsuperscript{20}). Thus, it is
reasonable to expect that these conserved genes should present at a
higher percentage than non-BUSCO genes in most matrices. However, we
found that gene partitions from all matrices comprised less than 50\% of
BUSCO genes, even with several data matrices contain less than 20
percent (Extended Data Fig. 2B). Moreover, we identified number of
ribosomal RNA genes in each data matrix (Extended Data Fig. 2C). We
found Philippe2009, Nosenko2013\_ribo\_14615 and Chang2015 matrices
possessed much higher content of ribosomal genes than other matrices
(Extended Data Fig. 2D). We also annotated each gene partition against
SwissProt\textsuperscript{21} and Gene Ontology databases, and results
are shown in Supplementary Table 1.

Previous analyses have mainly focused on outgroup selection and/or model
fits in different data matrices, and the matrix overlap across different
studies has been rarely investigated since all these studies rely on
their own method for orthology inference. Our initial intent of
comparing matrix overlap across matrices was to construct a new matrix
with shared taxon and gene sampling to test specific hypotheses about
differences in support. The hypothesis is that although the
bioinformatics pipeline used for OG determination is different, most
larger data matrices should share a core set of orthologs. However, our
results show that surprisingly little overlap across all the data
matrices from different studies (Fig. 2). Thus, these results strongly
suggested that gene sampling largely varies across studies, indicating
that different regions of the genomes are sampled with little overlap
among different data matrices. Moreover, the little overlap among
studies may potentially lead to incongruent phylogenetic results and
difficult to directly compare results across studies.

\begin{verbatim}
## <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
##     aspect: function
##     backtransform_range: function
##     clip: on
##     default: FALSE
##     distance: function
##     expand: TRUE
##     is_free: function
##     is_linear: function
##     labels: function
##     limits: list
##     modify_scales: function
##     range: function
##     ratio: 1
##     render_axis_h: function
##     render_axis_v: function
##     render_bg: function
##     render_fg: function
##     setup_data: function
##     setup_layout: function
##     setup_panel_params: function
##     setup_params: function
##     transform: function
##     super:  <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/gene_composition,-1.pdf}

\textbf{Extended Data Fig. 2.} Annotation and representation of BUSCO
and ribosomal genes in each data matrix. (A). The number of partitions
with BUSCO annotations in each matrix, relative to the number of
partitions. (B). The percentage BUSCO annotations in each matrix. (C).
The number of partitions with annotations of ribosomal genes in each
matrix, relative to the number of partitions. (D). The percentage
ribosomal genes in each matrix.

\includegraphics{manuscript_files/figure-latex/matrix_overlap-1.pdf}

\textbf{Fig. 2.} Pairwise overlap between each of the primary matrices
considered here. Horizontal size is proportional to the number of genes
sampled, vertical size to the number of taxa sampled. The horizontal
intersection shows the proportions of shared genes, the vertical
intersection shows the proportions of shared taxa.

\hypertarget{support-for-porifera-sister-and-ctenophora-sister}{%
\subsubsection{Support for Porifera-sister and
Ctenophora-sister}\label{support-for-porifera-sister-and-ctenophora-sister}}

\includegraphics{manuscript_files/figure-latex/support_published_analyses-1.pdf}
\textbf{Fig. 3.} A total of 163 analyses were transcribed from the
literature (Supplementary Table 2).

Here we summarized all the 163 previous phylogenetic analyses from 15
studies (Fig. 3 and Supplementary Table 2). Among these studies, the
main conclusions of five studies strongly in favor of Porifera-sister
and 10 in support of Ctenophora-sister (Table 1). Moreover, three
studies are secondary studies without any new data matrices.
Interestingly, we found that all analyses strongly support of
Ctenophora-sister once site-homogeneous models were used, no matter what
outgroup choice, phylogenetic inference and whether data-recoding are
used. The support for Porifera-sister increases with the exclusion of
more distantly related outgroups and with the use of site-heterogeneious
CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI
analyses with CAT model are not well resolved (either not reach
convergence or do not provide significant support for the key node of
animal evolution) mainly due to computational limits of CAT models
employed in PhyloBayes. The strong support of Porifera-sister can only
be recovered from analyses using Poisson+CAT and recoding methods with
GTR+CAT models in several matrices. Although GTR+CAT model generally has
better model adequacy than Poisson+CAT or site-homogeneous models from
previous studies\textsuperscript{6,9}, Porifera-sister is not always
supported since several analyses strongly in favor Ctenophora-sister
even with only choanofalgellates are used as outgroups (Fig. 3). Thus,
our summary of previous phylogenetic analyses strongly indicated that
the trend of the support of Porifera-sister is generally increasing with
the use of more complex CAT models and more closely related outgroups,
whereas Ctenophora-sister is supported in all other cases.

\hypertarget{new-analyses-of-published-matrices}{%
\subsection{New analyses of published
matrices}\label{new-analyses-of-published-matrices}}

\includegraphics{manuscript_files/figure-latex/support_new_analyses-1.pdf}
\textbf{Fig. 4.} A total of 84 new phylgoenomic analyses were conducted
in this study (Supplementary Table 3).

One of the challenges of interpreting support for the placement of the
animal root is that different programs, software versions, and settings
have been used across studies, and phylogenetic analysis decisions have
been approached in very different ways. Here we first reanalyze the
primary matrices from each study under consistent conditions with
IQtree\textsuperscript{22} under a panel of evolutionary models. We
selected this tool because it has greater model flexibility,
computational time and accuracy than other tools in maximum likelihood
(ML) framework\textsuperscript{23} (Fig. 4; Supplementary Table 3).
Importantly, we also include a CAT-like, site-heterogeneous C10-C60 (C)
models\textsuperscript{24} that are implemented in both ML and Bayesian
inference (BI) framework in our analyses and it is interesting to
compare results between C60 model to CAT models since it has never been
used in any matrices analyzed here.

Here we reanalyzed 17 matrices from 15 studies that are used as main
conclusions from their original paper (Table 1). We also progressively
trimmed each data matrix with three level of outgroup choice if
applicable (Fig 1C). We first tested a variety of models for each matrix
and compared the relative fit of site-homogeneous and site-heterogeneous
C models using ModelFinder\textsuperscript{25} in IQtree (except only
LG+C60 models are used in 3 large matrices). We found in all cases,
WAG/LG+C60 models fit these matrices better than the site-homogeneous
models under BIC criteria and we then inferred support under the
best-fit model from IQtree (Extended Data Table 1). It should be noted
that we didn't include GTR+C models in model testing since it is
computational too expensive. The GTR+C models requires to estimate up to
over 10,000 (e.g.~GTR+C60 contains 234*60 parameters) free parameters
inferred from matrices, whereas only several hundred parameters needs to
estimate in site-homogeneous + C models. We then analyzed every matrix
under a panel of standard site-heterogeneous and site-homogeneous
models, including WAG, GTR and Poisson+C60. Moreover, we also used the
best-fit model identified above with the removal of all distantly
related outgroups and only include Choanoflagellata as outgroups.
Interestingly, with the exception of Moroz2014\_3d and all Nosenko2013
matrices that supports neither Porifera-sister or Ctenophora-sister, all
analysis conducted herein in IQtree strongly supported the
ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C
models with the most closely related outgroups (Fig. 4; Supplementary
Table 3).

\textbf{Extended Data Table 1.} The models selected by ModelFinder for
each matrix.

\begin{longtable}[]{@{}llll@{}}
\toprule
matrix & clade & result & model\_summary\tabularnewline
\midrule
\endhead
Borowiec2015\_Best108 & Choanimalia & Ctenophora-sister &
WAG+C60\tabularnewline
Chang2015 & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
Dunn2008 & Opisthokonta & Ctenophora-sister & WAG+C60\tabularnewline
Moroz2014\_3d & Choanimalia & Unresolved & WAG+C60\tabularnewline
Nosenko2013\_nonribo\_9187 & Opisthokonta & Unresolved &
WAG+C60\tabularnewline
Nosenko2013\_ribo\_11057 & Choanimalia & Unresolved &
WAG+C60\tabularnewline
Nosenko2013\_ribo\_14615 & Opisthokonta & Unresolved &
WAG+C60\tabularnewline
Ryan2013\_est & Opisthokonta & Ctenophora-sister &
WAG+C60\tabularnewline
Whelan2015\_D1 & Opisthokonta & Ctenophora-sister &
LG+C60\tabularnewline
Whelan2015\_D10 & Opisthokonta & Ctenophora-sister &
WAG+C60\tabularnewline
Whelan2015\_D20 & Opisthokonta & Ctenophora-sister &
WAG+C60\tabularnewline
Whelan2017\_full & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
Whelan2017\_strict & Choanimalia & Ctenophora-sister &
LG+C60\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{comparison-of-iqtree-and-phylobayes-results}{%
\subsubsection{Comparison of IQtree and PhyloBayes
results}\label{comparison-of-iqtree-and-phylobayes-results}}

Site heterogeneity in equilibrium frequency has been a major concern in
tests of Ctenophora-sister and Porifera-sister. This has been addressed
with CAT models. IQtree provides a new family of C models that also
address site heterogeneity. Given the extensive computational cost and
concerns about overparameterization of CAT models\textsuperscript{26},
we compared iqtree C results to CAT results for a subset of matrices to
see if they give consistent results. This would be of technical interest
because it would reduce the cost of accommodating compositional
heterogeneity in future analyses. For computational efficiency, we first
only applied Poisson+CAT (Poisson+CAT) models in every matrix with
Choanoflagellata as outgroups. It should be noted that most Phylobayes
runs were converged, although several large matrices have not reached
convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using
Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we
found no strong support for either hypothesis in Moroz2013\_3d and
Nosenko2013 matrices. Consistent with previous
study\textsuperscript{27,6}, we recovered strong support of
Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015.
Interestingly, we also found a strong support of Porifera-sister in two
Whelan2017 matrices\textsuperscript{15} and a weaker support of
subsampled matrix of Simion2017\textsuperscript{16}. The
Ctenophora-sister were strongly supported in all other matrices. For the
matrices have conflicted results between Poisson-CAT and
site-homogeneous models, we further ran Poisson+CAT models with
inclusion of more distantly related outgroups to evaluate how outgroup
choice may affect phylogenetic inference under CAT model. We found a
weak support of Porifera-sister in Holozoa matrices and strong support
of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating
that the support of Porifera-sister is increasing with inclusive of the
more closely related outgroups and CAT models. Thus, our reanalyzing
results are largely consistent with previous analyses, both suggesting
that the support of Porifera-sister is increasing with more complex
models and closely related outgroups. More importantly, our results
indicated that both model choice and outgroup sampling play a major role
in incongruent results of animal origin in several key data matrices.

\hypertarget{phylogenetic-signal}{%
\subsubsection{Phylogenetic signal}\label{phylogenetic-signal}}

The distribution of phylogenetic signal has never been quantified in
site-heterogeneous models and compared with site-homogeneous models. To
further explore the effect of model selection and outgroup choices in
animal origin, we next quantify the support of phylogenetic signal over
two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2:
Porifera-sister (Fig. 1B)) to three key data matrices with different
outgroup choice and models in both ML and BI framework (Fig. 5). By
calculating gene-wise log-likelihood scores between T1 and T2 for every
gene (\(\Delta\)GLS) or site (\(\Delta\)SLS) in each matrix, we found
that Ctenophora-sister had the higher proportions of supporting genes in
every analysis when using site-homogeneous models (LG or WAG in both
IQtree or PhyloBayes). Moreover, the outgroup choice has little impact
on the distribution of the support of phylogenetic signal in
sifte-homogeneous models. This finding is largely consistent with the
past study that majority of phylogenetic signal is strongly favored in
Ctenophora-sister hypothesis with site-homogeneous models in other data
matrices\textsuperscript{28}.

In contrast, the phylogenetic signal of site-heterogenous models changed
dramatically compared to the site-homogenous models. Although
Ctenophora-sister also had the higher proportions of supporting genes
with C models, we found that the phylogenetic signal decreases in many
genes using C models compared to site-homogeneous models, especially in
matrices from Ryan2013\_est that nearly 30\% of genes changed from
strong to week signal (\Delta\$lnL\textless{}2) (Fig. 5B; Supplementary
Table 4). It is not clear why the phylogenetic signal decreases in many
genes with a better fitting model as suggested by model testing in
IQtree, but our results indicated the C models lack of phylogenetic
signal compared to site-homogeneous models in the node of the root
position of animal phylogeny. In contrast to the C models, the
phylogenetic signal largely increased in Poisson-CAT models and more
importantly, the outgroup choice has a major effect of the distribution
of phylogenetic signal (Fig. 5A, C). We found that support for
Ctenophora-sister largely decreases and Porifera-sister increases
(\textasciitilde{} 20\% of sampled gene partitions) when exclusive of
more distantly related outgroups with the Poisson-CAT models
(Supplementary Table 4). The results here are further corroborated by
our phylogenomic analyses in BI, which the support of Porifera-sister
generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are
supported in each hypothesis across different analyses in
Whelan2017\_full matrices (Fig. Extended Data Fig. 3). Consistent with
the summary of \Delta\$GLS~analysis, We found gene population that
supports in either hypothesis are largely congruent between
site-homogenous and C models, whereas little overlap was observed in
Poisson-CAT models (especially with different outgroup choice). In other
words. our results strongly indicated that the distribution of
phylogenetic signal of animal origin is largely sensitive to model
selection (site-homogeneous vs.~site-heterogeneous models). More
importantly, the site-heterogeneous CAT model is also more sensitive to
the outgroup choice than site-homogeneous and C models in
Whelan2017\_full matrices.

\includegraphics{manuscript_files/figure-latex/phylogenetic_signal-1.pdf}

\textbf{Fig.5} The distribution of phylogenetic signal for two
alternative topological hypotheses on the earliest-branching animal
lineage with different models and outgroup choice. (A). Proportion of
genes supporting each of two alternative hypotheses for each of two
alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes
supporting each of two alternative hypotheses for each of two
alternative hypotheses for Ryan2013\_est matrix. (C) Proportion of genes
supporting each of two alternative hypotheses for each of two
alternative hypotheses for Whelan2017\_full matrix. The (\Delta\$GLS)
values for the genes across each data matrix are provided in
Supplementary Table 4.

\includegraphics{figures/Figure_Pairwise_heatmap_animal_root.png}
\textbf{Extended Data Fig. 3.} Pairwise intersections of the level of
congruence of gene partitions that are supported in each hypothesis
across different analyses in Whelan2017\_full matrices calculated by
Euclidean distance. (A) Pairwise comparisons across genes that supported
Ctenophora-sister in different analyses. (B) Pairwise comparisons across
genes that supported Porifera-sister in different analyses. The color
scale indicates the Pearson correlation, where 1 is positive linear
correlation, 0 is no linear correlation, and −1 is negative linear
correlation. The pairwise intersection heatmap is generated by
Intervene\textsuperscript{29}.

\hypertarget{bias-of-data-recoding}{%
\subsubsection{Bias of data-recoding}\label{bias-of-data-recoding}}

There is growing interest in data-recoding and previous
study\textsuperscript{9} hoped that data-recoding would reduce potential
artifacts due to differences across species in amino acid frequencies.
They report that posterior predictive (PP) analyses\textsuperscript{30}
indicate 6-state recoded analyses have better model adequacy than
20-state amino acid analyses, and ``Porifera-sister was favored under
all recoding strategies'' in Whelan2015\_D20 and Chang2015 data
matrices. Here we focus on two aspects of Feuda \emph{et al.} First, we
found that many of their recoded analyses are actually unresolved
(\emph{i.e.}, without strong support for either Porifera-sister or
Ctenophora-sister), and that the analyses with the best posterior
predictive scores do not provide strong support for Porifera-sister
(Extended Data Fig. 4). Second, we created four new random recoding
schemes by shuffling the amino acids in the SR-6 scheme (see methods).
When we applied each of these randomized codes to the Whelan matrix and
analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed
similar results as for the empirical SR-6 recoding. Like SR-6 recoding,
random recoding increases support for Porifera-sister and improves the
apparent adequacy of models to explain heterogeneity of states across
taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our
analyses show the impact of recoding is largely due to discarding
information, not accommodating variation in amino acid composition.
These findings indicate that it is premature to accept Porifera-sister
and reject Ctenophora-sister using data-recoding methods and recoding
can be a problematic method for addressing compositional variation (see
more detailed discussion in supplementary text).

\includegraphics{figures/Figure_random_rocoding.png}

\textbf{Extended Data Fig. 4.} (A) The two alternative hypotheses for
deep animal relationships considered here. Relationships that are not
part of these hypotheses are shown as unresolved polytomies. (B) Each of
the six plots presents one statistic, which include Posterior
probability of Porifera-sister and the five Posterior Predictive (PP)
statistics considered by Feuda \emph{et al.} Within each plot, there are
six lines for six different analyses. These six analyses are the
published SR-6 analyses presented by Feuda \emph{et al.}
(SR6-published), analyses obtained by applying the same methods to the
same data to to confirm that I can reproduce their published results
(SR6-reproduced), and four analyses based on randomized recoding
matrices obtained by shuffling the SR-6 coding scheme (random-00 --
random-03). Each analysis includes results for 20 states (the raw amino
acid data, shown by the left point) and for 6 states (the 6-state
recoded data, shown by the right point). For each statistic, the results
obtained with the random recoding are similar to those of the SR6
recoding. This indicates that the impact of recoding is dominated by
discarding data when collapsing from 20 states to 6 states, not
accommodating compositional heterogeneity across lineages.

\includegraphics{figures/Figure_recoding_summary.png} \textbf{Extended
Data Fig. 6.} A graphical representation of the posterior probabilities
for the 32 analyses presented by Feuda \emph{et al.} in their Table 3.
Cells are color coded by whether posterior probability is \textgreater{}
95 for Porifera-sister, \textgreater{} 95 for Ctenophora-sister, or
neither (unresolved). Posterior predictive (PP) statistics were
estimated for the 16 analyses in the top two rows of this figure (the
Chang and full Whelan matrices), but not the bottom two (the Whelan
matrices with reduced outgroup sampling).

\includegraphics{figures/Figure_recoding_pp.png}

\textbf{Extended Data Fig. 5.} The subset of eight CAT-GTR+G analyses
with posterior predictive (PP) scores that is the focus of Feuda
\emph{et al.}'s primary conclusions. These are a subset of the 32
analyses presented in their Table 3 and graphically here in the upper
right pane. The eight analyses are for two datasets (Chang and Whelan)
and four coding schemes. The coding schemes are the original 20 state
amino acid data, and three different six state recodings that group
amino acids based on different criteria: KGB6, D6, and SR6. Cells are
color coded as in Supplemental Fig XXRecoidng\_summary. Only one of
these analyses, the SR6 coding of the Whelan matrix, has \textgreater{}
95 support for Porifera-sister. The 20-state and 6-state points on the
plots in Fig XXRecoidng\_summary of the main text correspond to the 20
and SR6 Whelan cells shown here. The presented statistics for these
cells are posterior probability of Porifera-sister, Maxdiff (with lower
scores indicating better convergence of runs), and five posterior
predictive statistics (where lower absolute value indicates better model
adequacy). The only one of these eight analyses that provides strong
support for Porifera-sister is not the most adequate analysis by any of
the posterior predictive scores, and showed the poorest convergence
according to Maxdiff.

\hypertarget{the-current-state-of-understanding-and-future-direction}{%
\subsection{The current state of understanding and future
direction}\label{the-current-state-of-understanding-and-future-direction}}

Resolving the root position of animal phylogeny that diverged
\textasciitilde{} 650 million years ago has proven difficult, even with
large genome-scale phylogenomic data matrices. By synthesizing all past
phylogenomic studies and perform new analyses to characterize difference
between studies, we found the support of Porifera-sister can only be
recovered by more complex CAT models with the most closely related
outgroups, and Ctenophora-sister in all other cases. Three factors
contribute significantly to the previous incongruent results based on
our analyses. First, the overlap of gene sampling and matrix is
extremely low across different studies, leading to the difficult of
directly compare the results across different studies. Second, the
distribution of phylogenetic signal are largely varied between
site-homogeneous and site-heterogeneous models. More importantly,
site-heterogeneous models are also more sensitive to outgroup choice
than site-homogeneous models, leading to different results even with the
same data matrix with different outgroup choice. Third, we demonstrated
that the support of Porifera-sister by data-recoding methods is
potentially resulted by data reduction, not accommodation of
compositional heterogeneity across species. Thus, the recovery of
Porifera-sister needs more constraints than Ctenophora-sister. Rather
than embracing one hypothesis over another, our results provide an
integrative overview of the challenge and uncertainty related to the
origin of animals.

\hypertarget{future-direction}{%
\subsection{Future direction}\label{future-direction}}

Many key positions of animal phylogeny can be reconstructed by
increasing genome-scale data within a phylogenomic framework, although
new challenges have also arised. Overcoming the challenges to inferring
the earliest diverging animal lineage is particular difficult and here
we discuss the several aspects that will help to improve our
understanding of animal evolution. First, a broader sampling of
high-quality sponge and ctenophore genomes are needed to shorten the
long branch since they are among the longest branches in the animal
phylogeny (currently only 2 ctenophore and 3 sponge genomes are
available). Second, accurate reconstruction of orthologs and a consensus
analytical strategy should be formalized with investigators to improve
communications and reproducibility of phylogenetic
workflow\textsuperscript{31}. Last, we also hope that the work we have
conducted here, including consolidating all the datasets in one place
with consistent formats and species names, will enhance the technical
value of this interesting question to methods-focused investigators that
look to develop and evaluate methods or evolutionary models to address
this difficult phylogenetic problem. More importantly, we also hope all
investigators critically evaluate the competing hypotheses without bias
instead of showing preferences to one hypothesis over as a priori due to
historical reasons or previous work. Finally, other sources of evidence
may also provide important insight to animal origin, including
morphological characters\textsuperscript{32,33} and fossil
records\textsuperscript{34}. We hope the future breakthroughs that will
allow us to move the field forward and a consensus viewpoint of early
animal evolution will be reached in the coming years.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

All files and bioinformatic commands associated with this analysis are
available at \url{https://github.com/caseywdunn/animal_root}

\hypertarget{data-selection-and-wrangling}{%
\subsubsection{Data selection and
wrangling}\label{data-selection-and-wrangling}}

We retrieved matrices from each publication (Table 1), storing the raw
data in this manuscript's version control repository. We manually edited
some minor formatting to make the batch processing of the matrices
\emph{en masse}, e.g.~standardizing the formatting of charset blocks.
All changes made are tracked with git.

\hypertarget{matrix-comparison-and-annotation}{%
\subsubsection{Matrix comparison and
annotation}\label{matrix-comparison-and-annotation}}

\hypertarget{taxon-name-reconciliation}{%
\paragraph{Taxon name reconciliation}\label{taxon-name-reconciliation}}

We programmatically queried the NCBI Taxonomy database to standardize
names of samples in each matrix. We also use a table where manual
entries were needed (Supplementary Table 5,
reconciliation/taxonomy\_info/manual\_taxonomy\_map.tsv), e.g.~authors
of original matrix indicate species name in original manuscript. For a
table summarizing all samples and their new or lengthened names, see
Supplementary Table 6 (reconciliation/taxonomy\_info/taxon\_table.tsv).

\hypertarget{sequence-comparisons}{%
\paragraph{Sequence Comparisons}\label{sequence-comparisons}}

Using the partition files for each matrix, we isolated each sequence for
each taxon from each partition. Because many of the matrices had been
processed by the original authors to remove columns that are poorly
sampled or highly variable, these matrix-derived sequences can have
deletions relative to the actual gene sequences.

We used DIAMOND\textsuperscript{35} to compare each sequence to all
others using default diamond Blastp parameters. We further filtered
DIAMOND results such that we retained hits for 90\% of partitions
(pident \textgreater{} 78.0, eValue \textless{} 1e-15, no
self-\textgreater{}self). We ran BUSCO with default parameters for all
sequences against the provided metazoa gene set. We also ran a BLAST+
blastp search against the SwissProt\textsuperscript{21} database,
filtering such that we retain at least one hit for \textasciitilde{}97\%
of partitions (pident \textgreater{} 50.0, eValue \textless{} 1e-15).

\hypertarget{partition-network}{%
\paragraph{Partition Network}\label{partition-network}}

We used the sequence similarity comparisons described above to compare
partitions.

We constructed a network with Python and NetworkX\textsuperscript{36}
v2.2 where each node is a partition and each edge represents a DIAMOND
sequence-to-sequence match between sequences in the partitions. We
extracted each connected component from this network. We further split
these components if the the most connected node (i.e.~most edges) had
two times more the standard deviation from the mean number of edges in
the component it is a member of and if removing that node splits the
component into two or more components. We then decorated every node in
the partition network with the most often found SwissProt BLAST+ result
and BUSCO results to see which components contain which classes and
families of genes. See Supplementary Table 7
{[}partition\_network\_summary table{]} for a summary tally of each part
of the comparison.

\hypertarget{phylogenetic-analyses}{%
\subsubsection{Phylogenetic analyses}\label{phylogenetic-analyses}}

\hypertarget{phylogenetic-analyses-in-iqtree}{%
\paragraph{Phylogenetic analyses in
IQtree}\label{phylogenetic-analyses-in-iqtree}}

To investigate the phylogenetic hypotheses and distribution of
phylogenetic signal in studies aiming to elucidate the root position of
animal phylogeny, we considered 17 data matrices from all phylogenomic
studies that were constructed from EST, transcriptomic or genomic data
(Table 1). Because different choices of substitution models could
largely influence phylogenetic inference of the placement of the root
position of animal phylogeny (e.g.~site-heterogeneous
vs.~site-homogeneous models), we first investigated model-fit from each
matrix using ModelFinder\textsuperscript{25} in
IQtree\textsuperscript{22}, including site-heterogenous C10 to C60
profile mixture models (C models) as variants of the CAT models in ML
framework (C10-C60 model were included for model comparison via -madd
option). We included models that are commonly used in previous analyses,
including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C
models in the model testing. For computational efficiency, the GTR with
C models were not included in model testing since it requires to
estimate over 10,000 parameters. For large matrices like Hejnol2009,
Borrowiec2015 and Simion2017, model testing is also not computational
feasible so only LG+C60 models were used since LG/WAG+C60 models were
suggested as the best-fit model in all other matrices.

We then reanalyzed each matrix under a panel of evolutionary models,
including WAG+G, GTR+G, poisson+C60+G and associated best-fit model
identified above. Nodal support was assessed with 1000 ultrafast
bootstrap replicates for each analysis. Because of the large size of
Hejnol2009 and Simion2017, it was not computationally feasible to
analyze the whole matrix using the C model or CAT site-heterogeneous
models. To circumvent his limitation, we reduced the data size from
their full matrices to facilitate computational efficiency for
site-heterogeneous models. For Hejnol2009 matrix, we instead used the
330-gene matrix constructed by Hejnol \emph{et al}.
2009\textsuperscript{13}, since the main conclusion for their study is
based on this subsampled matrix; For Simion2017 matrix, we only included
most complete 25\% of genes (genes that were present in less than 79
taxa were removed; 428 genes were kept). It should be noted that the
main conclusion of Simion \emph{et al}. was also based on selection of
25\% of genes for their jackknife approach.

\hypertarget{outgroup-taxa-sampling-with-c10-c60-and-cat-models}{%
\paragraph{Outgroup taxa sampling with C10-C60 and CAT
models}\label{outgroup-taxa-sampling-with-c10-c60-and-cat-models}}

Because different choices of outgroups could also affect phylogenetic
inference as suggested in previous analyses, we parsed the full data
matrices into three different types of outgroups: Choanimalia , Holozoa
and Opisthokonta. These datasets include the same set of genes but
differ in the composition of outgroup species. Choanimalia only includes
choanofagellates as outgroup; Holozoa also includes more distantly
related holozoans; Opistokonta also includes Fungi. For each Choanimalia
data matrice, both C models IQtree and Poisson-CAT models in PhyloBayes
were conducted. The maximum likelihood analysis was performed using the
best-fit substitution model identified as above and nodal support was
assessed with 1000 ultrafast bootstrap replicates using IQtree.
Moreover, bayesian inference with the site-heterogeneous poisson-CAT
model was done with PhyloBayes MPI\textsuperscript{37}. To minimize
computational burden, CAT-GTR models were only performed in the.

For several Choanozoa matrices indicated strong support for the
hypothesis that sponges are the sister group to the remaining Metazoa
using the CAT model, bayesian inference with Poisson-CAT model was also
conducted to Holozoa and Opisthokonta data matrices with the same
settings as above. For all the analyses with Poisson+CAT models in
PhyloBayes, two independent chains were sampled every generation. Tracer
plots of MCMC runs were visually inspected in Tracer v1.6 to assess
stationarity and appropriate burn-in. Chains were considered to have
reached convergence when the maxdiff statistic among chains was below
0.3 (as measured by bpcomp) and effective sample size \textgreater{} 50
for each parameter (as measured by tracecomp). A 50\% majority‐rule
consensus tree was computed with bpcomp, and nodal support was estimated
by posterior probability. For those matrices that were not converged,
PhyloByes analyses were run for at least one month.

\hypertarget{phylogenetic-distribution-of-support}{%
\paragraph{phylogenetic distribution of
support}\label{phylogenetic-distribution-of-support}}

To investigate the distribution of phylogenetic signal in data matrices,
we considered three major data matrices from three studies that had
different topology between ML and BI using CAT model in our reanalysis,
including Philippe2008, Ryan2013\_est, and Whelan2017\_full data
matrices. We examined two hypotheses: Ctenophora-sister and
Porifera-sister to rest of metazoans, under both ML and BI frameworks
with different outgroup schemes (Choanomalia and Opisthokonta). For ML
analysis in each dataset, site-wise likelihood scores were inferred for
both hypotheses using IQtree (option -g) with the same best-fit model
identified above, and site-homogeneous WAG and LG models. The two
different phylogenetic hypotheses passed to IQtree (via -z) were the
corresponding tree that the ctenophore as the sister lineage tree and
the corresponding tree that was modified to have sponges as the sister
to all other metazoans. The constraint trees were conducted in a
customized R script. The numbers of genes and sites supporting each
hypothesis were calculated with IQtree output and Perl scripts from Shen
et al.~2017\textsuperscript{28}. For BI analysis, we only considered the
Philippe\_2009 and Whelan\_2017 datasets.

\hypertarget{performance-of-data-recoding}{%
\paragraph{Performance of
data-recoding}\label{performance-of-data-recoding}}

All code used for the analyses presented here is available in a git
repository at \url{https://github.com/caseywdunn/feuda_2017_response}.
The randomized recoding analyses are in the
\texttt{../trees\_new/recoding/feuda\_2017\_response/alternative}
subdirectory of the git repository.

The original SR-6 recoding scheme is
``\texttt{APST\ CW\ DEGN\ FHY\ ILMV\ KQR}''\textsuperscript{38}, where
spaces separate amino acids that are placed into the same group. This
recoding is one member of a family of recodings, each with a different
number of groups, based on clustering of the JTT matrix. The other
recoding used by Feuda \emph{et al.}, KGB-6 and D-6, are based on
different matrices and methods\textsuperscript{39}.

The \texttt{alt\_recode.py} script was used to generate the randomized
recoding schemes and apply the recodings to the data. To create the
randomized recoding schemes, the amino acids in the SR-6 recoding were
randomly reshuffled. This creates new recodings that have the same sized
groups as SR-6. The new recodings were, from \texttt{random-00} to
\texttt{random-03}:

\begin{verbatim}
MPKE AY IDGQ TRC WNLF SVH
EIFT WL QVPG NKM SCHA RYD
LCQS GK WPTI VRD YEFN MAH
IWQV TY KDLM ESH APCF GRN
\end{verbatim}

To apply these to the data, each amino acid was replaced with the first
amino acid in the group. When applying \texttt{random-00}, for example,
each instance of \texttt{R} and \texttt{C} would be replaced by a
\texttt{T}.

The 20 state matrices are the same across all analyses since they are
not recoded. Since all the 20 state matrices are the same, variation
between 20-state results (as in the left side of each pane of Extended
Data Fig. 4) give insight into the technical variance of the inference
process.

Each new matrix was analyzed with PhyloBayes-mpi. Analyses were run for
1000 generations, and a 200 generation burnin applied. The resulting
tree files and posterior predictive scores were parsed for display with
the code in \texttt{manuscript.rmd}.

The statistics presented in Extended Data Figs. 4 and 5 were parsed from
the Feuda \emph{et al.} manuscript into the file
\texttt{tidy\_table\_3.tsv} and rendered for display with the code in
\texttt{manuscript.rmd}.

\hypertarget{ackowledgements}{%
\subsection{Ackowledgements}\label{ackowledgements}}

We thank the Yale Center for Research Computing for use of the research
computing infrastructure, specificaly the Farnam cluster.

\hypertarget{author-contributions}{%
\subsection{Author contributions}\label{author-contributions}}

\pagebreak

\hypertarget{supplemental-information}{%
\section{Supplemental Information}\label{supplemental-information}}

\hypertarget{supplementary-results}{%
\subsection{Supplementary Results}\label{supplementary-results}}

\hypertarget{models-of-molecular-evolution-1}{%
\subsubsection{Models of molecular
evolution}\label{models-of-molecular-evolution-1}}

The exchangeability matrix \(E\) describes the rate at which one amino
acid changes to another. Exchangeability matrices have been used in the
studies under consideration here include: Poisson (or F81), WAG, LG,
GTR. While the exchangeability matrix describes the relative rate of
different changes between amino acids, the actual rate can be further
scaled. There are a couple approaches that have been used in the studies
considered here:

\begin{itemize}
\item
  F81\textsuperscript{40} corresponds to equal rates between all states.
  The F81 matrix is also sometimes referred to as the Poisson matrix. It
  has no free parameters to estimate since all off-diagonal elements are
  set to 1.
\item
  WAG\textsuperscript{41} is an empirically derived exchangeability
  matrix based on a dataset of 182 globular protein families. It has no
  free parameters to estimate since all off-diagonal elements are set
  according to values estimated from this particular sample dataset.
\item
  LG\textsuperscript{42}, like WAG, is an empirically derived
  exchangeability matrix. It is based on a much larger set of genes, and
  variation in rates across sites was taken into consideration when it
  was calculated. It has no free parameters to estimate since all
  off-diagonal elements are set according to values estimated from this
  particular sample dataset.
\item
  GTR, the General Time Reversible exchangeability matrix, has free
  parameters for all off-diagonal elements that describe the
  exchangeability of different amino acids. It is constrained so that
  changes are reversible, \emph{i.e.} the rates above the diagonal are
  the same as those below the diagonal. This leaves 190 parameters that
  must be estimated from the data long with the other model parameters
  and the phylogenetic tree topology. This estimation requires a
  considerable amount of data and computational power, but if successful
  has the advantage of being based on the dataset at hand rather than a
  different dataset (as for LG and WAG).
\end{itemize}

While the exchangeability matrix describes the relative rate of
different changes between amino acids, the actual rate can be further
scaled. There are a couple approaches that have been used in the studies
considered here:

\begin{itemize}
\item
  Site homogeneous rates. The rates of evolution are assumed to be the
  same at all sites in the amino acid alignment.
\item
  Gamma rate heterogeneity. Each site is assigned to a different rate
  class with its own rate value. This accommodates different rates of
  evolution across different sites. Gamma is used so commonly that
  sometimes it isn't even specified, making it difficult at times to
  know if a study uses Gamma or not.
\end{itemize}

The vector of equilibrium frequencies \(\Pi\) describes the stationary
frequency of amino acids. There are a few approaches that have been used
across the studies considered here:

\begin{itemize}
\item
  Empirical site-homogeneous. The frequency of each amino acid is
  observed from the matrix under consideration and applied to
  homogeneously to all sites in the matrix.
\item
  Estimated site-homogeneous. The frequency of each amino acid is
  inferred along with other model parameters, under the assumption that
  it is the same at all sites.
\item
  CAT site heterogeneous\textsuperscript{3}. Each site is assigned to a
  class with its own equilibrium frequencies. The number of classes,
  assignment of sites to classes, and equilibrium frequencies within the
  data are all estimated in a Bayesian framework.
\item
  C10 to C60\textsuperscript{24}. 10 to 60-profile mixture models as
  variants of the CAT model under the maximum-likelihood framework.
\item
  Recoding method: Amino acids are recoded into six groups based on
  function to account for both compositional heterogeneity and
  substitution saturation. Several recoding strategies have been
  proposed, including Dayhoff 6-state recoding, S\&R 6-state recoding,
  KGB 6-state recoding based on exchangeability matrix of models.
\end{itemize}

\hypertarget{details-of-published-analyses}{%
\subsubsection{Details of published
analyses}\label{details-of-published-analyses}}

\hypertarget{dunn-et-al.-2008}{%
\paragraph{\texorpdfstring{Dunn \emph{et al.}
2008}{Dunn et al. 2008}}\label{dunn-et-al.-2008}}

Dunn \emph{et al.}\textsuperscript{2} added Expressed Sequence Tag (EST)
data for 29 animals. It was the first phylogenomic analysis that
included ctenophores, and therefore that could test the relationships of
both Ctenophora and Porifera to the rest of animals. It was also the
first phylogenetic analysis to recover Ctenophora as the sister group to
all other animals.

The data matrix was constructed using a semi-automated approach. Genes
were translated into proteins, promiscuous domains were masked, all gene
sequences from all species were compared to each other with blastp,
genes were clustered based on this similarity with
TribeMCL\textsuperscript{43}, and these clusters were filtered to remove
those with poor taxon sampling and high rates of lineage-specific
duplications. Gene trees were then constructed, and in clades of
sequences all from the same species all but one sequence were removed
(these groups are often due to assembly errors). The remaining gene
trees with more than one sequence for any taxon were then manually
inspected. If strongly supported deep nodes indicative of paralogy were
found, the entire gene was discarded. If the duplications for a a small
number of taxa were unresolved, all genes from those taxa were excluded.
Genes were then realigned and sites were filtered with
Gblocks\textsuperscript{44}, resulting in a 77 taxon matrix. Some taxa
in this matrix were quite unstable, which obscured other
strongly-supported relationships. Unstable taxa were identified with
leaf stability indices\textsuperscript{45}, as implemented in
phyutility\textsuperscript{46}, and removed from the matrix. This
resulted in the 64-taxon matrix that is the focus of most of their
analyses. Phylogenetic analyses were conducted under the Poisson+CAT
model in phylobayes\textsuperscript{3}, and under the WAG model in
MrBayes\textsuperscript{47} and RAxML\textsuperscript{48}.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
The placement of ctenophores (comb jellies) as the sister group to all
other sampled metazoans is strongly supported in all our analyses. This
result, which has not been postulated before, should be viewed as
provisional until more data are considered from placozoans and
additional sponges.
\end{quote}

Note that there was, in fact, an exception to strong support. An
analysis of the 40 ribosomal proteins in the matrix recovered
Ctenophora-sister with only 69\% support. This study did not include
\emph{Trichoplax}.

\hypertarget{philippe-et-al.-2009}{%
\paragraph{\texorpdfstring{Philippe \emph{et al.}
2009}{Philippe et al. 2009}}\label{philippe-et-al.-2009}}

Philippe \emph{et al.}\textsuperscript{27} assembled an EST dataset for
55 species with 128 genes to explore phylogenetic relationship of early
diverging animals by added 9 new species. Th data matrix were assembled
based on The phylogenetic analysis using Poisson-CAT model strongly
supported Porifera-sister, and ctenophores was sister to cnidarians to
form the ``coelenterate'' clade. Gene trees were then constructed, and
potentially paralogs were removed by a bootstrap threshold of 70.
Ambiguously aligned regions were trimemed and only genes sampled for at
least two-thirds of species were retained. The phylogenetic analyses
were conducted under the Poisson+CAT model in
phylobayes\textsuperscript{3}.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
The resulting phylogeny yields two significant conclusions reviving old
views that have been challenged in the molecular era: (1) that the
sponges (Porifera) are monophyletic and not paraphyletic as repeatedly
proposed, thus undermining the idea that ancestral metazoans had a
sponge-like body plan; (2) that the most likely position for the
ctenophores is together with the cnidarians in a ``coelenterate'' clade.
\end{quote}

\hypertarget{hejnol-et-al.2009}{%
\paragraph{Hejnol et al.~2009}\label{hejnol-et-al.2009}}

Hejnol \emph{et al.}\textsuperscript{13} added EST sequences from seven
taxa, and a total of 94 taxa were included in the final data matrix to
explore animal phylogeny, especially the position of acoelomorph
flatwroms. The orthology inference was largely similar to Dunn et
al.~2008, with the exception of orthology genes were clustered by MCL.
The final data matrix included 1497 genes, and then subsampled with 844,
330 and 53 gens by different thresholds of gene occupancy. With the
exception of 53 gene matrix, maximum likelihood analyses from all other
datasets strongly supported Ctenophora-sister (models were selected by
RaxML perl script).

\hypertarget{pick-et-al.-2010}{%
\paragraph{\texorpdfstring{Pick \emph{et al.}
2010}{Pick et al. 2010}}\label{pick-et-al.-2010}}

Pick \emph{et al.}\textsuperscript{49} sought to test whether
Ctenophora-sister was an artefact of insufficient taxon sampling. They
added new and additional published sequence data to the 64-taxon matrix
of Dunn \emph{et al.}\textsuperscript{2}. The new taxa included 12
sponges, 1 ctenophore, 5 cnidarians, and \emph{Trichoplax}. They further
modified the matrix by removing 2,150 sites that were poorly sampled or
aligned. They considered two different sets of outgroups:
Choanoflagellata (resulting in Choanimalia) and the same sampling as
Dunn \emph{et al.} (resulting in Opisthokonta).

All their analyses were conducted under the F81+CAT+Gamma model in
phylobayes\textsuperscript{3}, in both a Bayesian framework and with
bootstrapping. All analyses have the same ingroup sampling and site
removal so it isn't possible to independently assess the impact of these
factors. Analyses with Choanimalia sampling recovered Porifera-sister
with 72\% posterior probability (PP) and 91\% bootstrap support (BS).
With broader Opisthokonta sampling, support for Porifera-sister is 84\%
PP. This is an interesting case where increased outgroup sampling leads
to increased support for Porifera-sister.

The authors argue that previous results supporting Ctenophora-sister
``are artifacts stemming from insufficient taxon sampling and
long-branch attraction (LBA)'' and that ``this hypothesis should be
rejected''. Although the posterior probabilities supporting
Porifera-sister are not strong, they conclude:

\begin{quote}
Results of our analyses indicate that sponges are the sister group to
the remaining Metazoa, and Placozoa are sister group to the Bilateria
\end{quote}

They also investigated saturation, and conclude that Dunn \emph{et
al.}\textsuperscript{2} is more saturated than Philippe \emph{et al.}
2009 {[}Philippe:2009hh{]}. Note that the Pick \emph{et
al.}\textsuperscript{49} dataset is not reanalyzed here because
partition data are not available, and due to site filtering the
partition file from Dunn \emph{et al.}\textsuperscript{2} cannot be
applied to this matrix.

\hypertarget{nosenko-et-al.-2013}{%
\paragraph{\texorpdfstring{Nosenko \emph{et al.}
2013}{Nosenko et al. 2013}}\label{nosenko-et-al.-2013}}

Nosenko \emph{et al.}\textsuperscript{50} added Expressed Sequence Tag
(EST) data for 9 species of non-bilaterian metazoans (7 sponges). They
constructed a novel matrix containing 122 genes and parsed them into two
non-overlaping matrices (ribosomal and non-ribosomal genes) and found
incongruent results of deep metazoan phylogeny. The other major finding
was that ribosomal gene partitions showed significantly lower saturation
than the non-ribosomal ones.

Orthologs were constructed using the bioinfomatics pipeline
OrthoSelect\textsuperscript{51}. They also evaluated level of
saturation, leaf stability of sampled taxa, compositional heterogeinity
and model comparison of each matrix. By modifying gene sampling, ingroup
and outgroup sampling, three major topologies related to the position of
animal-root were constructed (including Porifera+Placozoa sister,
Ctenophora-sister and Porifera-sister). Phylogenetic analyses were
conducted under the Poisson+CAT, GTR+CAT and GTR models in
phylobayes\textsuperscript{3}.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
we were able to reconstruct a metazoan phylogeny that is consistent with
traditional, morphology-based views on the phylogeny of non-bilaterian
metazoans, including monophyletic Porifera and ctenophores as a
sister-group of cnidarians.
\end{quote}

Note the main conclusion of this paper is baed on the selection of genes
that evolved slowly ribosomal RNA genes. The Ctenophora-sister was
rejected based on the most saturated dataset and not supported by the
matrix with the most closely related outgroups (Choanomalia
matrix)\textsuperscript{50}.

\hypertarget{ryan-et-al.-2013}{%
\paragraph{\texorpdfstring{Ryan \emph{et al.}
2013}{Ryan et al. 2013}}\label{ryan-et-al.-2013}}

Ryan \emph{et al.} sequenced the first ctenophore genome of
\emph{Mnemiopsis leidyi}. With the genome resources of \emph{M. leidyi},
the authors constructed two phylogenomic datasets: a ``Genome set''
based on 13 animal genomes and a ``EST Set'' that also included 59
animals. They analyzed both matrices by site-homogeneous GTR+Gamma and
site-heterogeneous Poisson-CAT models with three sets of outgroup
sampling to evaluate the effect of outgroup selection to the ingroup
topology for the Ryan2013\_est matrix. The Orthologs were constructed
based on the method of Hejnol \emph{et al.} 2009. For the
Ryan2013\_genome matrix, they performed phylogenetic analyses with both
gene content and sequence-baed analyses. Overall, their results strongly
supported Ctenophora-sister in all datasets they analyzed using
site-homogeneous model. The Poisson+CAT model of the genome dataset
strongly supported of a clade of Ctenophora and Porifera as the sister
group to all other Metazoa and Bayesian analysis on the EST dataset did
not converge after 205 days (but strongly supported Porifera in
Choaimalia matrix).

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
Our phylogenetic analyses suggest that ctenophores are the sister group
to the rest of the extant animals.
\end{quote}

\hypertarget{moroz-et-al.-2014}{%
\paragraph{\texorpdfstring{Moroz \emph{et al.}
2014}{Moroz et al. 2014}}\label{moroz-et-al.-2014}}

Moroz \emph{et al.}\textsuperscript{52} sequenced the second ctenophore
genome \emph{Pleurobrachia bachei} to explore the phylogenetic
relationship of Metazoa. All phylogenetic analyses strongly supported
Ctenophora-sister with different taxon and gene sampling using WAG
site-homogeneous model. Two phylogenomic matrices were generated, the
first set was represented by two ctenophore species, whereas the other
set contained improved ctenophore sampling (10 taxa, Moroz2013\_3d).
Orthology determination employed in HaMStR\textsuperscript{53} using
1,032 ``model organism'' single-copy orthologs. Sequence were then
trimmed and aligned. This resulted in a final matrix of 170,871 amino
acid positions across 586 genes with 44 taxa for the first matrix, and
114 genes with 60 taxa for the second matrix. All the phylogenetic
analyses were analyzed in RAxML under the WAG+CAT+F models (different
than CAT models in PhyloBayes) to reduce the computational cost.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
Our integrative analyses place Ctenophora as the earliest lineage within
Metazoa. This hypothesis is supported by comparative analysis of
multiple gene families, including the apparent absence of HOX genes,
canonical microRNA machinery, and reduced immune complement in
ctenophores.
\end{quote}

It should be noted that only the Moroz\_3d matrix has been reanalyzed in
other studies, although the support of Ctenophora-sister is quite low.

\hypertarget{borowiec-et-al.-2015}{%
\paragraph{\texorpdfstring{Borowiec \emph{et al.}
2015}{Borowiec et al. 2015}}\label{borowiec-et-al.-2015}}

Borowiec \emph{et al.}\textsuperscript{14} assembled a genome dataset
comprised 1080 orthologs derived from 36 publicly available genomes
representing major lineages of animals, although only one genome of
sponge and ctenophore was included. The orthologs were constructed using
OrthologID pipeline\textsuperscript{54}. After removal of spurious
sequences and genes with more than 40\% of mission data, the final
matrix included 1080 (Total 1080). The authors further filtered the full
dataset to 9 sub-datasets by filtering genes with high long-branch
scores; genes with high saturation; gene occupancy; fast evolving genes.
The main conclusion of the paper was largely based on
BorowiecTotal\_1080 and Borowiec\_Best108 matrices. Phylogenetic
analyses were conducted under the GTR+CAT model in PhyloBayes in
selected matrices, and under the data-partitioning methods in RAxML for
all matrices.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
Our phylogeny supports the still-controversial position of ctenophores
as sister group to all other metazoans. This study also provides a
workflow and computational tools for minimizing systematic bias in
genome-based phylogenetic analyses.
\end{quote}

It should be noted that the authors also employed recoding-method in the
Borowiec\_Best108 matrix and found neither support of Porifera-sister or
Ctenophora-sister\textsuperscript{14}.

\hypertarget{whelan-et-al.2015}{%
\paragraph{Whelan et al.~2015}\label{whelan-et-al.2015}}

Whelan \emph{et al.} 2015\textsuperscript{11} constructed a new
phylogenomic dataset by eight new transcripomic data and investigated a
range of possible sources of systematic error under multiple analyses
(e.g.~long-branch attraction, compositional bias, fast evolving genes,
etc.). Putative orthologs were determined of each species using HaMStR
using the model organism core ortholog set (same as Moroz \emph{et al.}
2014) and subsequently removal of genes with too much missing data and
potential paralogs. The authors further filtered the full dataset to 24
sub-datasets by filtering genes with high long-branch scores; genes with
high RSFV values; genes that are potential paralogs; fast evolving genes
and progressively removal of outgroups. All the maximum likelihood
analyses with site-homogeneous model and PartitionFinder strongly
suggested Ctenophora-sister. CAT-GTR models only used in least saturated
dataset 6 and 16 also strongly supported Ctenophora.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
Importantly, biases resulting from elevated compositional heterogeneity
or elevated substitution rates are ruled out. Placement of ctenophores
as sister to all other animals, and sponge monophyly, are strongly
supported under multiple analyses, herein.
\end{quote}

Note that the authors also reanalyzed Philippe2009 matrix (with the
removal of ribosomal genes) and recovered Porifera-sister with moderate
support (pp=90).

\hypertarget{chang-et-al.2015}{%
\paragraph{Chang et al.~2015}\label{chang-et-al.2015}}

Chang \emph{et al.}\textsuperscript{55} was originally used to explore
phylogenetic position of Myxozoa in Cnidaria but also sampled broadly
across the breadth of animal diversity. The authors constructed a
dataset with 200 protein markers based on Philippe et
al.~2011\textsuperscript{7} with 51,940 amino acids and 77 taxa. Both
site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly
supported Ctenophora-sister.

Note that this data matrix has been extensively reanalyzed in many
studies.

\hypertarget{pisani-et-al.-2015}{%
\paragraph{\texorpdfstring{Pisani \emph{et al.}
2015}{Pisani et al. 2015}}\label{pisani-et-al.-2015}}

Pisani \emph{et al.}\textsuperscript{6} reanalyzed representative
datasets that supported Ctenophora-sister, including Ryan2013\_est,
Moroz2014\_3d and Whelan2015 datasets. It was the first study showing
that progressively removal of more distantly related outgroups could
largely affect phylogenomic inference of the position of the root of
animal phylogeny. The authors suggested that the inclusion of outgroups
very distant from the ingroup can cause systematic errors due to
long-branch attraction. Phylogenetic analyses were conducted under the
Poisson+CAT and GTR models in phylobayes. They found GTR-CAT and
Poisson-CAT models generally had better model-fit than site-homogeneous
GTR models in these data matrice. Moreover, they found the support of
Ctenophora-sister decreases when the exclusion of distantly related
outgroups and the use of site-heterogeneous CAT models.

Regarding the recovery of Porifera-sister, the authors concluded:

\begin{quote}
Our results reinforce a traditional scenario for the evolution of
complexity in animals, and indicate that inferences about the evolution
of Metazoa based on the Ctenophora-sister hypothesis are not supported
by the currently available data.
\end{quote}

\hypertarget{feuda-et-al.-2017}{%
\paragraph{\texorpdfstring{Feuda \emph{et al.}
2017}{Feuda et al. 2017}}\label{feuda-et-al.-2017}}

Feuda \emph{et al.}\textsuperscript{9} didn't generate any new data,
instead they used the data-recoding methods to reanalyze two key
datasets that support Ctenophora-sister (Whelan2015\_D20, Chang2015
datasets). It was the first phylogenomic study that suggested recoding
methods have better performance than non-recoding methods in the
relationships at the root of the animal tree. The authors compared model
adequacy using posterior predictive analyses from a set of
site-homogeneous (WAG, LG, GTR, data-partitioning) and
site-heterogeneous (CAT-GTR) models in non-recoding and recoding
datasets. The results showed that data-recoding can significant reduce
compositional heterogeneity in both datasets with CAT-GTR models and
strongly supported Porifera-sister hypothesis (see more details in
recoding section in supplementary text).

Regarding the recovery of Porifera-sister, the authors concluded:

\begin{quote}
Because adequate modeling of the evolutionary process that generated the
data is fundamental to recovering an accurate phylogeny, our results
strongly support sponges as the sister group of all other animals and
provide further evidence that Ctenophora-sister represents a tree
reconstruction artifact.
\end{quote}

\hypertarget{whelan-and-halanych-2016}{%
\paragraph{Whelan and Halanych 2016}\label{whelan-and-halanych-2016}}

Whelan \emph{et al.}\textsuperscript{26} is the only study to evaluate
performance of site-heterogeneous models and site-homogeneous model with
data partitioning under the simulation framework. The simulation results
suggested that Poisson+CAT model consistently performed worse than other
models in simulation datasets. More importantly, the authors also showed
that both Poisson+CAT and GTR+ CAT models could overestimated
substitutional heterogeneity in many cases. They also reanalyzed
datasets from Philippe 2009 and Nosenko 2013 using both CAT models and
data partitioning with site-homogeneous model. The results indicated
that Poisson + CAT model tends to recover less accurate trees and more
importantly, both GTR + CAT and data partitioning strongly supported
Ctenophora-sister in reanalyses.

The authors concluded:

\begin{quote}
Practices such as removing constant sites and parsimony uninformative
characters, or using CAT-F81 when CAT-GTR is deemed too computationally
expensive, cannot be logically justified. Given clear problems with
CAT-F81, phylogenies previously inferred with this model should be
reassessed.
\end{quote}

\hypertarget{whelan-et-al.-2017}{%
\paragraph{\texorpdfstring{Whelan \emph{et al.}
2017}{Whelan et al. 2017}}\label{whelan-et-al.-2017}}

Whelan \emph{et al.}\textsuperscript{15} added 27 new ctenophore
transcriptomic data to explore animal-root position as well as
relationships within Ctenophores. It significantly increased ctenophore
taxon sampling than other studies. Putative orthologs were determined
largely similar to Whelan2015, with the difference of a core Ctenophora
core datasets were constructed here with more than 2000 genes. The
subsequent filtering strategy was also similar to the previous study.
All analyses using site-homogeneous and site-heterogeneous models
strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model
in choanoazoa dataset. The main conclusions of this paper were based on
Whelan2017\_full and Whelan2017\_strict matrices.

Regarding the recovery of Ctenophora-sister, the authors concluded:
\textgreater{}Using datasets with reasonably high ctenophore and other
non-bilaterian taxon sampling, our results strongly reject the
hypothesis that sponges are the sister lineage to all other extant
metazoans.

\hypertarget{simion-et-al.-2017}{%
\paragraph{\texorpdfstring{Simion \emph{et al.}
2017}{Simion et al. 2017}}\label{simion-et-al.-2017}}

Simion \emph{et al.}\textsuperscript{16} added transcriptomic data for
21 new animals. The data matrix was constructed using a semi-automated
approach to comprehensively detect and eliminate potential systematic
errors. The resulting dataset comprises 1,719 genes and 97 species,
including 61 non-bilaterian species. It was by far the largest
phylogenomic dataset in terms of taxon and gene sampling related to the
relationship at the root of animal phylogeny.

The final matrix was first analyzed using the Poisson-CAT model.
Different that other phylobayes analyses, Simion \emph{et al.} used a
gene jackknife strategy based on 100 analyses to overcome the
computational limitation because of the large data size. Each jackknife
is baed on a random selection of \textasciitilde{} 25\% of the genes.
The Phylobayes with site-heterogeneous model strongly supported the
Porifera-sister, whereas site-homogeneous strongly supported
Ctenophora-sister in all datasets. Importantly, the authors compared the
behavior of long-branch effect between site-heterogeneous and
site-homogeneous models by progressively removal taxa and concluded
higher sensitivity of site-homogeneous models to LBA than CAT models.

Regarding the recovery of Ctenophora-sister, the authors concluded:

\begin{quote}
Our dataset outperforms previous metazoan gene superalignments in terms
of data quality and quantity. Analyses with a best-fitting
site-heterogeneous evolutionary model provide strong statistical support
for placing sponges as the sister-group to all other metazoans, with
ctenophores emerging as the second-earliest branching animal lineage.
\end{quote}

It should be noted that all the PhyloBayes runs have not reached
convergence due to the computational cost in these large matrices.

\hypertarget{data-recoding-methods}{%
\subsubsection{Data-recoding methods}\label{data-recoding-methods}}

Feuda \emph{et al.}\textsuperscript{39} tackled a difficult phylogenetic
problem -- placing the root of the animal phylogeny\textsuperscript{8}.
In the past decade, some analyses have placed Ctenophora (comb jellies)
and others Porifera (sponges) as the sister group to all other animals
(Fig. Extended Data Fig. 4). Feuda \emph{et al.}\textsuperscript{39}
present new phylogenetic analyses that they claim provide strong support
for Porifera-sister. Their analyses consider datasets from studies by
Chang \emph{et al.}\textsuperscript{55} and Whelan \emph{et
al.}\textsuperscript{56}, both of which found support for
Ctenophora-sister. Feuda \emph{et al.} were concerned that these
Ctenophora-sister results were artefacts of lineage-specific differences
in amino acid frequencies. In an attempt to reduce these differences,
they recoded the full set of twenty amino acids into six groups of amino
acids. These groups have more frequent evolutionary changes within them
than between them, based on empirical observations in large protein
datasets\textsuperscript{38}. The intent is to discard many
lineage-specific changes, which are expected to fall within these
groups. Rather than model compositional heterogeneity, as their title
suggests, this approach discards heterogeneous information so that much
simpler models with fewer states can be applied. They report that
posterior predictive (PP) analyses\textsuperscript{30} indicate 6-state
recoded analyses have better model adequacy than 20-state amino acid
analyses, and ``Porifera-sister was favored under all recoding
strategies''. Here we focus on two aspects of Feuda \emph{et al.} First,
we point out that many of their recoded analyses are actually unresolved
(\emph{i.e.}, without strong support for either Porifera-sister or
Ctenophora-sister), and that the analyses with the best posterior
predictive scores do not provide strong support for Porifera-sister.
Second, we present new analyses that show the impact of recoding is
largely due to discarding information, not accommodating variation in
amino acid composition. These findings indicate that it is premature to
accept Porifera-sister and reject Ctenophora-sister. They also show that
recoding can be a problematic method for addressing compositional
variation.

Feuda \emph{et al.} examine support for Ctenophora-sister and
Porifera-sister under all combinations of two models of molecular
evolution, four datasets, and four coding schemes. This provides 32
analyses that they report in their Table 3 and that we present
graphically here as Extended Data Fig. 6. There is striking variation in
support for Ctenophora-sister and Porifera-sister across these analyses
(Extended Data Fig. 6). Feuda \emph{et al.} accept the results of some
analyses and reject others based on posterior predictive (PP) analyses
of model adequacy, which assess how well a model explains variation in
the data\textsuperscript{30}. They considered five different posterior
predictive statistics that capture different types of variation in the
data. From this they conclude that their ``results strongly support
sponges as the sister group of all other animals''.

This conclusion does not follow from their own presented results. Only a
single analysis with posterior predictive scores provides what could be
considered strong support \textgreater{} 95 posterior probability) for
Porifera-sister. Of the 32 analyses, posterior predictive scores were
calculated for 16 (those for the full Whelan and Chang matrices). Based
on posterior predictive scores, Feuda \emph{et al.} reject eight of
these that were conducted under the GTR+G model (which all have strong
support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses
(Extended Data Fig. 5). Two of these eight are analyses of the original
20-state amino acid data, both of which provide strong support for
Ctenophora-sister. Of the six recoded analyses, five are unresolved.
Only a single analysis for which posterior predictive scores are
available provides strong support for Porifera-sister, the CAT-GTR+G
analysis of the SR-6\textsuperscript{38} recoded
Whelan\textsuperscript{56} matrix. Furthermore, this analysis does not
have the best score according to any of the five posterior predictive
statistics they considered (Extended Data Fig. 5). The only statistic
that stands out for this one analysis is that it has the highest maxdiff
(Extended Data Fig. 5), indicating that it did not converge as well as
other analyses.

Though their study does not provide strong support for Porifera-sister,
the sensitivity of their results to recoding provides an opportunity to
better understand and evaluate the impact of recoding more generally.
This is important given the growing interest in
recoding\textsuperscript{39}. Feuda \emph{et al.} hoped recoding would
reduce potential artefacts due to differences across species in amino
acid frequencies. They interpreted the fact that their analyses are
sensitive to recoding as evidence that such an artefact exists and that
they successfully addressed it by recoding. An alternative hypothesis is
that recoding impacts phylogenetic analyses because it discards so much
information. These two hypotheses can be tested by applying new recoding
schemes that also reduce twenty states down to six, but are based on
random grouping rather than empirical frequencies of amino acid
exchange. Empirical and random recodings both discard the same amount of
information, but only empirical recoding reduce the impact of amino-acid
frequency as intended. Different results between empirical and random
recoding would be consistent with the hypothesis that the empirical
approach works as intended to accommodate compositional heterogeneity.
Similar results would suggest that the impact of recoding is due to
discarding information. Here we focus on their single analysis with a
posterior predictive score that supports Porifera-sister, the CAT-GTR+G
analysis of the SR-6 recoded Whelan data. we created four new random
recoding schemes by shuffling the amino acids in the SR-6 scheme (see
Supplemental Methods and analysis code at
\url{https://github.com/caseywdunn/feuda_2017}). When we applied each of
these randomized codes to the Whelan matrix and analyzed them under the
CAT-GTR+G model with phylobayes-mpi\textsuperscript{37}, we observed
similar results as for the empirical SR-6 recoding. Like SR-6 recoding,
random recoding increases support for Porifera-sister and improves the
apparent adequacy of models to explain heterogeneity of states across
taxa (PP taxon hetero mean and max, Extended Data Fig. 4).

These analyses suggest that the major impact of recoding on phylogenetic
analyses is data reduction, not accommodation of compositional
heterogeneity across species. This indicates that recoding may not be an
effective tool for accommodating among-species differences in amino acid
frequencies. Compositional heterogeneity would be better addressed with
models of molecular evolution that explicitly describe such
differences\textsuperscript{57}, if progress can be made on the
considerable computational challenges of such complex models.

\hypertarget{supplementary-methods}{%
\subsection{Supplementary Methods}\label{supplementary-methods}}

\hypertarget{matrix-mapping}{%
\subsubsection{Matrix mapping}\label{matrix-mapping}}

Taxa and partition correspondence across manuscripts was assessed by
comparing all sequences for each taxon in each partition across all
matrices with diamond blast. Based on inspection of sequence similarity,
we excluded all comparisons with less than 99\% identity and greater
than 10\textsuperscript{-25} e-value.

\hypertarget{taxa-comparison-across-matrices}{%
\subsubsection{Taxa comparison across
matrices}\label{taxa-comparison-across-matrices}}

The primary intent of comparing taxa across matrices was to validate our
taxon name reconciliation across studies.

We first considered pairwise similarity between the same species from
different matrices in different studies.

\hypertarget{partition-comparison-across-matrices}{%
\subsubsection{Partition comparison across
matrices}\label{partition-comparison-across-matrices}}

The count for a partition pair can be much larger than the number of
genes in the matrix, which suggests that the count is the number of hsps
rather than the number of sequences with hits.

There are 17 matrices. A gene that is perfectly sampled would form a
cluster with this size. To our surprise, very few clusters, though, are
this size. This suggests that intersection of genes between matrices is
extreme low. This finding suggested genes used for exploring the
position of the root of the animal phylogeny

\hypertarget{supplementary-tables}{%
\subsection{Supplementary Tables}\label{supplementary-tables}}

\begin{longtable}[]{@{}llrr@{}}
\caption{Supplementary Table 4. Distribution of phylogenetic signal for
different models and outgroup choice}\tabularnewline
\toprule
Matrix & Catogary & Number of Gene Partitions &
Percentage\tabularnewline
\midrule
\endfirsthead
\toprule
Matrix & Catogary & Number of Gene Partitions &
Percentage\tabularnewline
\midrule
\endhead
Philippe2009\_Choanimalia\_WAG+C60 & Ctenophore-sister-strong-signal &
19 & 0.1484375\tabularnewline
Philippe2009\_Choanimalia\_WAG+C60 & Ctenophore-sister-weak-signal & 40
& 0.3125000\tabularnewline
Philippe2009\_Choanimalia\_WAG+C60 & Porifera-sister-weak-signal & 57 &
0.4453125\tabularnewline
Philippe2009\_Choanimalia\_WAG+C60 & Porifera-sister-strong-signal & 12
& 0.0937500\tabularnewline
Philippe2009\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-strong-signal & 17 & 0.1328125\tabularnewline
Philippe2009\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-weak-signal & 24 & 0.1875000\tabularnewline
Philippe2009\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Porifera-sister-weak-signal & 30 & 0.2343750\tabularnewline
Philippe2009\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Porifera-sister-strong-signal & 57 & 0.4453125\tabularnewline
Philippe2009\_Choanimalia\_LG & Ctenophore-sister-strong-signal & 43 &
0.3359375\tabularnewline
Philippe2009\_Choanimalia\_LG & Ctenophore-sister-weak-signal & 23 &
0.1796875\tabularnewline
Philippe2009\_Choanimalia\_LG & Porifera-sister-weak-signal & 30 &
0.2343750\tabularnewline
Philippe2009\_Choanimalia\_LG & Porifera-sister-strong-signal & 32 &
0.2500000\tabularnewline
Philippe2009\_Choanimalia\_LG (PhyloBayes) &
Ctenophore-sister-strong-signal & 48 & 0.3750000\tabularnewline
Philippe2009\_Choanimalia\_LG (PhyloBayes) &
Ctenophore-sister-weak-signal & 27 & 0.2109375\tabularnewline
Philippe2009\_Choanimalia\_LG (PhyloBayes) & Porifera-sister-weak-signal
& 19 & 0.1484375\tabularnewline
Philippe2009\_Choanimalia\_LG (PhyloBayes) &
Porifera-sister-strong-signal & 34 & 0.2656250\tabularnewline
Philippe2009\_Choanimalia\_WAG & Ctenophore-sister-strong-signal & 42 &
0.3281250\tabularnewline
Philippe2009\_Choanimalia\_WAG & Ctenophore-sister-weak-signal & 29 &
0.2265625\tabularnewline
Philippe2009\_Choanimalia\_WAG & Porifera-sister-weak-signal & 23 &
0.1796875\tabularnewline
Philippe2009\_Choanimalia\_WAG & Porifera-sister-strong-signal & 34 &
0.2656250\tabularnewline
Philippe2009\_WAG+C60 & Ctenophore-sister-strong-signal & 23 &
0.1796875\tabularnewline
Philippe2009\_WAG+C60 & Ctenophore-sister-weak-signal & 44 &
0.3437500\tabularnewline
Philippe2009\_WAG+C60 & Porifera-sister-weak-signal & 44 &
0.3437500\tabularnewline
Philippe2009\_WAG+C60 & Porifera-sister-strong-signal & 17 &
0.1328125\tabularnewline
Philippe2009\_Poisson+CAT (PhyloBayes) & Ctenophore-sister-strong-signal
& 40 & 0.3125000\tabularnewline
Philippe2009\_Poisson+CAT (PhyloBayes) & Ctenophore-sister-weak-signal &
31 & 0.2421875\tabularnewline
Philippe2009\_Poisson+CAT (PhyloBayes) & Porifera-sister-weak-signal &
34 & 0.2656250\tabularnewline
Philippe2009\_Poisson+CAT (PhyloBayes) & Porifera-sister-strong-signal &
23 & 0.1796875\tabularnewline
Philippe2009\_LG & Ctenophore-sister-strong-signal & 45 &
0.3515625\tabularnewline
Philippe2009\_LG & Ctenophore-sister-weak-signal & 22 &
0.1718750\tabularnewline
Philippe2009\_LG & Porifera-sister-weak-signal & 24 &
0.1875000\tabularnewline
Philippe2009\_LG & Porifera-sister-strong-signal & 37 &
0.2890625\tabularnewline
Philippe2009\_LG (PhyloBayes) & Ctenophore-sister-strong-signal & 35 &
0.2734375\tabularnewline
Philippe2009\_LG (PhyloBayes) & Ctenophore-sister-weak-signal & 39 &
0.3046875\tabularnewline
Philippe2009\_LG (PhyloBayes) & Porifera-sister-weak-signal & 45 &
0.3515625\tabularnewline
Philippe2009\_LG (PhyloBayes) & Porifera-sister-strong-signal & 9 &
0.0703125\tabularnewline
Philippe2009\_WAG & Ctenophore-sister-strong-signal & 48 &
0.3750000\tabularnewline
Philippe2009\_WAG & Ctenophore-sister-weak-signal & 19 &
0.1484375\tabularnewline
Philippe2009\_WAG & Porifera-sister-weak-signal & 19 &
0.1484375\tabularnewline
Philippe2009\_WAG & Porifera-sister-strong-signal & 42 &
0.3281250\tabularnewline
Ryan2013\_Choanimalia\_WAG+C60 & Ctenophore-sister-strong-signal & 12 &
0.0295566\tabularnewline
Ryan2013\_Choanimalia\_WAG+C60 & Ctenophore-sister-weak-signal & 202 &
0.4975370\tabularnewline
Ryan2013\_Choanimalia\_WAG+C60 & Porifera-sister-weak-signal & 184 &
0.4532020\tabularnewline
Ryan2013\_Choanimalia\_WAG+C60 & Porifera-sister-strong-signal & 8 &
0.0197044\tabularnewline
Ryan2013\_Choanimalia\_LG & Ctenophore-sister-strong-signal & 89 &
0.2192118\tabularnewline
Ryan2013\_Choanimalia\_LG & Ctenophore-sister-weak-signal & 142 &
0.3497537\tabularnewline
Ryan2013\_Choanimalia\_LG & Porifera-sister-weak-signal & 124 &
0.3054187\tabularnewline
Ryan2013\_Choanimalia\_LG & Porifera-sister-strong-signal & 51 &
0.1256158\tabularnewline
Ryan2013\_Choanimalia\_WAG & Ctenophore-sister-strong-signal & 92 &
0.2266010\tabularnewline
Ryan2013\_Choanimalia\_WAG & Ctenophore-sister-weak-signal & 138 &
0.3399015\tabularnewline
Ryan2013\_Choanimalia\_WAG & Porifera-sister-weak-signal & 118 &
0.2906404\tabularnewline
Ryan2013\_Choanimalia\_WAG & Porifera-sister-strong-signal & 58 &
0.1428571\tabularnewline
Ryan2013\_est\_WAG+C60 & Ctenophore-sister-strong-signal & 25 &
0.0615764\tabularnewline
Ryan2013\_est\_WAG+C60 & Ctenophore-sister-weak-signal & 206 &
0.5073892\tabularnewline
Ryan2013\_est\_WAG+C60 & Porifera-sister-weak-signal & 169 &
0.4162562\tabularnewline
Ryan2013\_est\_WAG+C60 & Porifera-sister-strong-signal & 6 &
0.0147783\tabularnewline
Ryan2013\_est\_LG & Ctenophore-sister-strong-signal & 87 &
0.2142857\tabularnewline
Ryan2013\_est\_LG & Ctenophore-sister-weak-signal & 156 &
0.3842364\tabularnewline
Ryan2013\_est\_LG & Porifera-sister-weak-signal & 133 &
0.3275862\tabularnewline
Ryan2013\_est\_LG & Porifera-sister-strong-signal & 30 &
0.0738916\tabularnewline
Ryan2013\_est\_WAG & Ctenophore-sister-strong-signal & 119 &
0.2931034\tabularnewline
Ryan2013\_est\_WAG & Ctenophore-sister-weak-signal & 123 &
0.3029557\tabularnewline
Ryan2013\_est\_WAG & Porifera-sister-weak-signal & 118 &
0.2906404\tabularnewline
Ryan2013\_est\_WAG & Porifera-sister-strong-signal & 46 &
0.1133005\tabularnewline
Whelan2017\_full\_LG+C60 & Ctenophore-sister-strong-signal & 37 &
0.1745283\tabularnewline
Whelan2017\_full\_LG+C60 & Ctenophore-sister-weak-signal & 96 &
0.4528302\tabularnewline
Whelan2017\_full\_LG+C60 & Porifera-sister-weak-signal & 65 &
0.3066038\tabularnewline
Whelan2017\_full\_LG+C60 & Porifera-sister-strong-signal & 14 &
0.0660377\tabularnewline
Whelan2017\_full\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-strong-signal & 102 & 0.4811321\tabularnewline
Whelan2017\_full\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-weak-signal & 20 & 0.0943396\tabularnewline
Whelan2017\_full\_Poisson+CAT (PhyloBayes) & Porifera-sister-weak-signal
& 16 & 0.0754717\tabularnewline
Whelan2017\_full\_Poisson+CAT (PhyloBayes) &
Porifera-sister-strong-signal & 74 & 0.3490566\tabularnewline
Whelan2017\_full\_LG & Ctenophore-sister-strong-signal & 82 &
0.3867924\tabularnewline
Whelan2017\_full\_LG & Ctenophore-sister-weak-signal & 54 &
0.2547170\tabularnewline
Whelan2017\_full\_LG & Porifera-sister-weak-signal & 55 &
0.2594340\tabularnewline
Whelan2017\_full\_LG & Porifera-sister-strong-signal & 21 &
0.0990566\tabularnewline
Whelan2017\_full\_LG (PhyloBayes) & Ctenophore-sister-strong-signal & 81
& 0.3820755\tabularnewline
Whelan2017\_full\_LG (PhyloBayes) & Ctenophore-sister-weak-signal & 59 &
0.2783019\tabularnewline
Whelan2017\_full\_LG (PhyloBayes) & Porifera-sister-weak-signal & 52 &
0.2452830\tabularnewline
Whelan2017\_full\_LG (PhyloBayes) & Porifera-sister-strong-signal & 20 &
0.0943396\tabularnewline
Whelan2017\_full\_WAG & Ctenophore-sister-strong-signal & 84 &
0.3962264\tabularnewline
Whelan2017\_full\_WAG & Ctenophore-sister-weak-signal & 55 &
0.2594340\tabularnewline
Whelan2017\_full\_WAG & Porifera-sister-weak-signal & 49 &
0.2311321\tabularnewline
Whelan2017\_full\_WAG & Porifera-sister-strong-signal & 24 &
0.1132076\tabularnewline
Whelan2017 only F81+C60 & Ctenophore-sister-strong-signal & 30 &
0.1415094\tabularnewline
Whelan2017\_Choanimalia\_LG+C60 & Ctenophore-sister-weak-signal & 102 &
0.4811321\tabularnewline
Whelan2017\_Choanimalia\_LG+C60 & Porifera-sister-weak-signal & 62 &
0.2924528\tabularnewline
Whelan2017\_Choanimalia\_LG+C60 & Porifera-sister-strong-signal & 18 &
0.0849057\tabularnewline
Whelan2017\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-strong-signal & 52 & 0.2452830\tabularnewline
Whelan2017\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Ctenophore-sister-weak-signal & 23 & 0.1084906\tabularnewline
Whelan2017\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Porifera-sister-weak-signal & 31 & 0.1462264\tabularnewline
Whelan2017\_Choanimalia\_Poisson+CAT (PhyloBayes) &
Porifera-sister-strong-signal & 106 & 0.5000000\tabularnewline
Whelan2017\_Choanimalia\_LG & Ctenophore-sister-strong-signal & 80 &
0.3773585\tabularnewline
Whelan2017\_Choanimalia\_LG & Ctenophore-sister-weak-signal & 60 &
0.2830189\tabularnewline
Whelan2017\_Choanimalia\_LG & Porifera-sister-weak-signal & 42 &
0.1981132\tabularnewline
Whelan2017\_Choanimalia\_LG & Porifera-sister-strong-signal & 30 &
0.1415094\tabularnewline
Whelan2017\_Choanimalia\_LG (PhyloBayes) &
Ctenophore-sister-strong-signal & 86 & 0.4056604\tabularnewline
Whelan2017\_Choanimalia\_LG (PhyloBayes) & Ctenophore-sister-weak-signal
& 56 & 0.2641509\tabularnewline
Whelan2017\_Choanimalia\_LG (PhyloBayes) & Porifera-sister-weak-signal &
53 & 0.2500000\tabularnewline
Whelan2017\_Choanimalia\_LG (PhyloBayes) & Porifera-sister-strong-signal
& 17 & 0.0801887\tabularnewline
Whelan2017\_Choanimalia\_WAG & Ctenophore-sister-strong-signal & 91 &
0.4292453\tabularnewline
Whelan2017\_Choanimalia\_WAG & Ctenophore-sister-weak-signal & 50 &
0.2358491\tabularnewline
Whelan2017\_Choanimalia\_WAG & Porifera-sister-weak-signal & 42 &
0.1981132\tabularnewline
Whelan2017\_Choanimalia\_WAG & Porifera-sister-strong-signal & 29 &
0.1367924\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Wallberg:2004ws}{}%
1.Wallberg, A., Thollesson, M., Farris, J. \& Jondelius, U. The
phylogenetic position of the comb jellies (Ctenophora) and the
importance of taxonomic sampling. \emph{Cladistics} \textbf{20},
558--578 (2004).

\leavevmode\hypertarget{ref-Dunn:2008ky}{}%
2.Dunn, C.W., Hejnol, A., Matus, D.Q., Pang, K., Browne, W.E., Smith,
S.A., Seaver, E., Rouse, G.W., Obst, M., Edgecombe, G.D., Sørensen,
M.V., Haddock, S.H.D., Schmidt-Rhaesa, A., Okusu, A., Kristensen, R.M.,
Wheeler, W.C., Martindale, M.Q. \& Giribet, G. Broad phylogenomic
sampling improves resolution of the animal tree of life. \emph{Nature}
\textbf{452}, 745--749 (2008).

\leavevmode\hypertarget{ref-Lartillot:2004dq}{}%
3.Lartillot, N. A Bayesian Mixture Model for Across-Site Heterogeneities
in the Amino-Acid Replacement Process. \emph{Molecular Biology and
Evolution} \textbf{21}, 1095--1109 (2004).

\leavevmode\hypertarget{ref-waterhouse2017busco}{}%
4.Waterhouse, R.M., Seppey, M., Simão, F.A., Manni, M., Ioannidis, P.,
Klioutchnikov, G., Kriventseva, E.V. \& Zdobnov, E.M. BUSCO applications
from quality assessments to gene prediction and phylogenomics.
\emph{Molecular biology and evolution} \textbf{35}, 543--548 (2017).

\leavevmode\hypertarget{ref-ryan2013genome}{}%
5.Ryan, J.F., Pang, K., Schnitzler, C.E., Nguyen, A.-D., Moreland, R.T.,
Simmons, D.K., Koch, B.J., Francis, W.R., Havlak, P., Smith, S.A. \&
others The genome of the ctenophore mnemiopsis leidyi and its
implications for cell type evolution. \emph{Science} \textbf{342},
1242592 (2013).

\leavevmode\hypertarget{ref-pisani2015genomic}{}%
6.Pisani, D., Pett, W., Dohrmann, M., Feuda, R., Rota-Stabelli, O.,
Philippe, H., Lartillot, N. \& Wörheide, G. Genomic data do not support
comb jellies as the sister group to all other animals. \emph{Proceedings
of the National Academy of Sciences} \textbf{112}, 15402--15407 (2015).

\leavevmode\hypertarget{ref-philippe2011resolving}{}%
7.Philippe, H., Brinkmann, H., Lavrov, D.V., Littlewood, D.T.J., Manuel,
M., Wörheide, G. \& Baurain, D. Resolving difficult phylogenetic
questions: Why more sequences are not enough. \emph{PLoS biology}
\textbf{9}, e1000602 (2011).

\leavevmode\hypertarget{ref-King:2017ie}{}%
8.King, N. \& Rokas, A. Embracing Uncertainty in Reconstructing Early
Animal Evolution. \emph{Current biology : CB} \textbf{27}, R1081--R1088
(2017).

\leavevmode\hypertarget{ref-feuda2017improved}{}%
9.Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O.,
Lartillot, N., Wörheide, G. \& Pisani, D. Improved modeling of
compositional heterogeneity supports sponges as sister to all other
animals. \emph{Current Biology} \textbf{27}, 3864--3870 (2017).

\leavevmode\hypertarget{ref-chang2015genomic}{}%
10.Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H.,
Huchon, D. \& Cartwright, P. Genomic insights into the evolutionary
origin of myxozoa within cnidaria. \emph{Proceedings of the National
Academy of Sciences} \textbf{112}, 14912--14917 (2015).

\leavevmode\hypertarget{ref-whelan2015error}{}%
11.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
signal, and the placement of ctenophora sister to all other animals.
\emph{Proceedings of the National Academy of Sciences} \textbf{112},
5773--5778 (2015).

\leavevmode\hypertarget{ref-hernandez2019six}{}%
12.Hernandez, A.M. \& Ryan, J.F. Six-state amino acid recoding is not an
effective strategy to offset the effects of compositional heterogeneity
and saturation in phylogenetic analyses. \emph{BioRxiv} 729103 (2019).

\leavevmode\hypertarget{ref-hejnol2009assessing}{}%
13.Hejnol, A., Obst, M., Stamatakis, A., Ott, M., Rouse, G.W.,
Edgecombe, G.D., Martinez, P., Baguñà, J., Bailly, X., Jondelius, U. \&
others Assessing the root of bilaterian animals with scalable
phylogenomic methods. \emph{Proceedings of the Royal Society B:
Biological Sciences} \textbf{276}, 4261--4270 (2009).

\leavevmode\hypertarget{ref-borowiec2015extracting}{}%
14.Borowiec, M.L., Lee, E.K., Chiu, J.C. \& Plachetzki, D.C. Extracting
phylogenetic signal and accounting for bias in whole-genome data sets
supports the ctenophora as sister to remaining metazoa. \emph{BMC
genomics} \textbf{16}, 987 (2015).

\leavevmode\hypertarget{ref-whelan2017ctenophore}{}%
15.Whelan, N.V., Kocot, K.M., Moroz, T.P., Mukherjee, K., Williams, P.,
Paulay, G., Moroz, L.L. \& Halanych, K.M. Ctenophore relationships and
their placement as the sister group to all other animals. \emph{Nature
ecology \& evolution} \textbf{1}, 1737 (2017).

\leavevmode\hypertarget{ref-simion2017large}{}%
16.Simion, P., Philippe, H., Baurain, D., Jager, M., Richter, D.J., Di
Franco, A., Roure, B., Satoh, N., Queinnec, E., Ereskovsky, A. \& others
A large and consistent phylogenomic dataset supports sponges as the
sister group to all other animals. \emph{Current Biology} \textbf{27},
958--967 (2017).

\leavevmode\hypertarget{ref-lanier2012recombination}{}%
17.Lanier, H.C. \& Knowles, L.L. Is recombination a problem for
species-tree analyses? \emph{Systematic Biology} \textbf{61}, 691--701
(2012).

\leavevmode\hypertarget{ref-salichos2013inferring}{}%
18.Salichos, L. \& Rokas, A. Inferring ancient divergences requires
genes with strong phylogenetic signals. \emph{Nature} \textbf{497}, 327
(2013).

\leavevmode\hypertarget{ref-shen2018tempo}{}%
19.Shen, X.-X., Opulente, D.A., Kominek, J., Zhou, X., Steenwyk, J.L.,
Buh, K.V., Haase, M.A., Wisecaver, J.H., Wang, M., Doering, D.T. \&
others Tempo and mode of genome evolution in the budding yeast
subphylum. \emph{Cell} \textbf{175}, 1533--1545 (2018).

\leavevmode\hypertarget{ref-fernandez2018phylogenomics}{}%
20.Fernández, R., Kallal, R.J., Dimitrov, D., Ballesteros, J.A., Arnedo,
M.A., Giribet, G. \& Hormiga, G. Phylogenomics, diversification
dynamics, and comparative transcriptomics across the spider tree of
life. \emph{Current Biology} \textbf{28}, 1489--1497 (2018).

\leavevmode\hypertarget{ref-boeckmann2003swiss}{}%
21.Boeckmann, B., Bairoch, A., Apweiler, R., Blatter, M.-C., Estreicher,
A., Gasteiger, E., Martin, M.J., Michoud, K., O'Donovan, C., Phan, I. \&
others The swiss-prot protein knowledgebase and its supplement trembl in
2003. \emph{Nucleic acids research} \textbf{31}, 365--370 (2003).

\leavevmode\hypertarget{ref-nguyen2014iq}{}%
22.Nguyen, L.-T., Schmidt, H.A., Haeseler, A. von \& Minh, B.Q. IQ-tree:
A fast and effective stochastic algorithm for estimating
maximum-likelihood phylogenies. \emph{Molecular biology and evolution}
\textbf{32}, 268--274 (2014).

\leavevmode\hypertarget{ref-zhou2017evaluating}{}%
23.Zhou, X., Shen, X.-X., Hittinger, C.T. \& Rokas, A. Evaluating fast
maximum likelihood-based phylogenetic programs using empirical
phylogenomic data sets. \emph{Molecular biology and evolution}
\textbf{35}, 486--503 (2017).

\leavevmode\hypertarget{ref-si2008empirical}{}%
24.Si Quang, L., Gascuel, O. \& Lartillot, N. Empirical profile mixture
models for phylogenetic reconstruction. \emph{Bioinformatics}
\textbf{24}, 2317--2323 (2008).

\leavevmode\hypertarget{ref-kalyaanamoorthy2017modelfinder}{}%
25.Kalyaanamoorthy, S., Minh, B.Q., Wong, T.K., Haeseler, A. von \&
Jermiin, L.S. ModelFinder: Fast model selection for accurate
phylogenetic estimates. \emph{Nature methods} \textbf{14}, 587 (2017).

\leavevmode\hypertarget{ref-whelan2016let}{}%
26.Whelan, N.V. \& Halanych, K.M. Who let the cat out of the bag?
Accurately dealing with substitutional heterogeneity in phylogenomic
analyses. \emph{Systematic biology} \textbf{66}, 232--255 (2016).

\leavevmode\hypertarget{ref-Philippe:2009hh}{}%
27.Philippe, H., Derelle, R., Lopez, P., Pick, K., Borchiellini, C.,
Boury-Esnault, N., Vacelet, J., Renard, E., Houliston, E., QuEinnec, E.,
Da Silva, C., Wincker, P., Le Guyader, H., Leys, S., Jackson, D.J.,
Schreiber, F., Erpenbeck, D., Morgenstern, B., WOrheide, G. \& Manuel,
M. Phylogenomics revives traditional views on deep animal relationships.
\emph{Current biology : CB} \textbf{19}, 706--712 (2009).

\leavevmode\hypertarget{ref-shen2017contentious}{}%
28.Shen, X.-X., Hittinger, C.T. \& Rokas, A. Contentious relationships
in phylogenomic studies can be driven by a handful of genes.
\emph{Nature Ecology \& Evolution} \textbf{1}, 0126 (2017).

\leavevmode\hypertarget{ref-khan2017intervene}{}%
29.Khan, A. \& Mathelier, A. Intervene: A tool for intersection and
visualization of multiple gene or genomic region sets. \emph{BMC
bioinformatics} \textbf{18}, 287 (2017).

\leavevmode\hypertarget{ref-Bollback:2002to}{}%
30.Bollback, J.P. Bayesian model adequacy and choice in phylogenetics.
\emph{Molecular Biology and Evolution} \textbf{19}, 1171--1180 (2002).

\leavevmode\hypertarget{ref-guang2016integrated}{}%
31.Guang, A., Zapata, F., Howison, M., Lawrence, C.E. \& Dunn, C.W. An
integrated perspective on phylogenetic workflows. \emph{Trends in
ecology \& evolution} \textbf{31}, 116--126 (2016).

\leavevmode\hypertarget{ref-mah2014choanoflagellate}{}%
32.Mah, J.L., Christensen-Dalsgaard, K.K. \& Leys, S.P. Choanoflagellate
and choanocyte collar-flagellar systems and the assumption of homology.
\emph{Evolution \& development} \textbf{16}, 25--37 (2014).

\leavevmode\hypertarget{ref-nielsen2019early}{}%
33.Nielsen, C. Early animal evolution: A morphologist's view.
\emph{Royal Society open science} \textbf{6}, 190638 (2019).

\leavevmode\hypertarget{ref-zhao2019cambrian}{}%
34.Zhao, Y., Vinther, J., Parry, L.A., Wei, F., Green, E., Pisani, D.,
Hou, X., Edgecombe, G.D. \& Cong, P. Cambrian sessile, suspension
feeding stem-group ctenophores and evolution of the comb jelly body
plan. \emph{Current Biology} \textbf{29}, 1112--1125 (2019).

\leavevmode\hypertarget{ref-Buchfink:2014}{}%
35.Buchfink, B., Xie, C. \& Huson, D.H. Fast and sensitive protein
alignment using diamond. \emph{Nature Methods} \textbf{12}, 59 EP
(2014).

\leavevmode\hypertarget{ref-Hagberg:2008}{}%
36.Hagberg, A.A., Schult, D.A. \& Swart, P.J. Exploring network
structure, dynamics, and function using networkx. \emph{Proceedings of
the 7th python in science conference} 11--15 (2008).

\leavevmode\hypertarget{ref-Lartillot:2013fg}{}%
37.Lartillot, N., Rodrigue, N., Stubbs, D. \& Richer, J. PhyloBayes MPI:
phylogenetic reconstruction with infinite mixtures of profiles in a
parallel environment. \emph{Systematic biology} \textbf{62}, 611--615
(2013).

\leavevmode\hypertarget{ref-Susko:2007ds}{}%
38.Susko, E. \& Roger, A.J. On reduced amino acid alphabets for
phylogenetic inference. \emph{Molecular Biology and Evolution}
\textbf{24}, 2139--2150 (2007).

\leavevmode\hypertarget{ref-Feuda:2017ew}{}%
39.Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O.,
Lartillot, N., Wörheide, G. \& Pisani, D. Improved Modeling of
Compositional Heterogeneity Supports Sponges as Sister to All Other
Animals. \emph{Current Biology} \textbf{27}, 1--12 (2017).

\leavevmode\hypertarget{ref-Felsenstein:1981vk}{}%
40.Felsenstein, J. Evolutionary trees from DNA sequences: a maximum
likelihood approach. \emph{Journal of Molecular Evolution} \textbf{17},
368--376 (1981).

\leavevmode\hypertarget{ref-Whelan:2001ds}{}%
41.Whelan, S. \& Goldman, N. A General Empirical Model of Protein
Evolution Derived from Multiple Protein Families Using a
Maximum-Likelihood Approach. \emph{Molecular Biology and Evolution}
\textbf{18}, 691--699 (2001).

\leavevmode\hypertarget{ref-Le:2008fp}{}%
42.Le, S.Q. \& Gascuel, O. An improved general amino acid replacement
matrix. \emph{Molecular Biology and Evolution} \textbf{25}, 1307--1320
(2008).

\leavevmode\hypertarget{ref-Enright:2002uq}{}%
43.Enright, A., Van Dongen, S. \& Ouzounis, C. An efficient algorithm
for large-scale detection of protein families. \emph{Nucleic Acids
Research} \textbf{30}, 1575--1584 (2002).

\leavevmode\hypertarget{ref-Castresana:2000vy}{}%
44.Castresana, J. Selection of conserved blocks from multiple alignments
for their use in phylogenetic analysis. \emph{Molecular Biology and
Evolution} \textbf{17}, 540--552 (2000).

\leavevmode\hypertarget{ref-Thorley:1999kg}{}%
45.Thorley, J. \& Wilkinson, M. Testing the phylogenetic stability of
early tetrapods. \emph{Journal of Theoretical Biology} \textbf{200},
343--344 (1999).

\leavevmode\hypertarget{ref-Smith:2008gb}{}%
46.Smith, S.A. \& Dunn, C.W. Phyutility: a phyloinformatics tool for
trees, alignments and molecular data. \emph{Bioinformatics} \textbf{24},
715--716 (2008).

\leavevmode\hypertarget{ref-Ronquist:2003hx}{}%
47.Ronquist, F. \& Huelsenbeck, J.P. MrBayes 3: Bayesian phylogenetic
inference under mixed models. \emph{Bioinformatics} \textbf{19},
1572--1574 (2003).

\leavevmode\hypertarget{ref-Stamatakis:2006wc}{}%
48.Stamatakis, A. RAxML-VI-HPC: maximum likelihood-based phylogenetic
analyses with thousands of taxa and mixed models. \emph{Bioinformatics}
\textbf{22}, 2688--2690 (2006).

\leavevmode\hypertarget{ref-Pick:2010eb}{}%
49.Pick, K.S., Philippe, H., Schreiber, F., Erpenbeck, D., Jackson,
D.J., Wrede, P., Wiens, M., Alie, A., Morgenstern, B., Manuel, M. \&
Worheide, G. Improved phylogenomic taxon sampling noticeably affects
nonbilaterian relationships. \emph{Molecular Biology and Evolution}
\textbf{27}, 1983--1987 (2010).

\leavevmode\hypertarget{ref-nosenko2013deep}{}%
50.Nosenko, T., Schreiber, F., Adamska, M., Adamski, M., Eitel, M.,
Hammel, J., Maldonado, M., Müller, W.E., Nickel, M., Schierwater, B. \&
others Deep metazoan phylogeny: When different genes tell different
stories. \emph{Molecular phylogenetics and evolution} \textbf{67},
223--233 (2013).

\leavevmode\hypertarget{ref-schreiber2009orthoselect}{}%
51.Schreiber, F., Pick, K., Erpenbeck, D., Wörheide, G. \& Morgenstern,
B. OrthoSelect: A protocol for selecting orthologous groups in
phylogenomics. \emph{BMC bioinformatics} \textbf{10}, 219 (2009).

\leavevmode\hypertarget{ref-moroz2014ctenophore}{}%
52.Moroz, L.L., Kocot, K.M., Citarella, M.R., Dosung, S., Norekian,
T.P., Povolotskaya, I.S., Grigorenko, A.P., Dailey, C., Berezikov, E.,
Buckley, K.M. \& others The ctenophore genome and the evolutionary
origins of neural systems. \emph{Nature} \textbf{510}, 109 (2014).

\leavevmode\hypertarget{ref-ebersberger2009hamstr}{}%
53.Ebersberger, I., Strauss, S. \& Haeseler, A. von HaMStR: Profile
hidden markov model based search for orthologs in ests. \emph{BMC
evolutionary biology} \textbf{9}, 157 (2009).

\leavevmode\hypertarget{ref-chiu2006orthologid}{}%
54.Chiu, J.C., Lee, E.K., Egan, M.G., Sarkar, I.N., Coruzzi, G.M. \&
DeSalle, R. OrthologID: Automation of genome-scale ortholog
identification within a parsimony framework. \emph{Bioinformatics}
\textbf{22}, 699--707 (2006).

\leavevmode\hypertarget{ref-Chang:2015hl}{}%
55.Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H.,
Huchon, D. \& Cartwright, P. Genomic insights into the evolutionary
origin of Myxozoa within Cnidaria. \emph{Proceedings of the National
Academy of Sciences} 1--6
(2015).doi:\href{https://doi.org/10.1073/pnas.1511468112}{10.1073/pnas.1511468112}

\leavevmode\hypertarget{ref-Whelan:2015jj}{}%
56.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
signal, and the placement of Ctenophora sister to all other animals.
\emph{Proceedings of the National Academy of Sciences} \textbf{112},
201503453--6 (2015).

\leavevmode\hypertarget{ref-Blanquart:2008gl}{}%
57.Blanquart, S. \& Lartillot, N. A Site- and Time-Heterogeneous Model
of Amino Acid Replacement. \emph{Molecular Biology and Evolution}
\textbf{25}, 842--858 (2008).

\leavevmode\hypertarget{ref-Foster:2004tw}{}%
58.Foster, P.G. Modeling compositional heterogeneity. \emph{Systematic
biology} \textbf{53}, 485--495 (2004).

\end{document}

### 2020-03-09 manuscript/manuscript.rmd
update manuscirpt with sensitive analyses

@@
 output:
-  github_document: default
+  html_document: default
   pdf_document: default
   word_document: default
-  html_document: default
+  github_document: default
 ---
@@
 
-# Rooting the animal tree of life
+# Rooting the animal tree of life - more is less for Ctenophora-sister?
 
@@
 
-## Abstract
+## Abstract (ideally less than 200 words)
+
+Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
+However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
+This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution.
+Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions.
+We next quantify the  support of phylogenetic signal and conduct sensitive analyses in several representative datasets.
+We found that the Porifera-sister can only be recovered with a combined factors in certain data matrices, with inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories and more closely related outgroups, whereas Ctenophora-sister is strongly supported in most other cases.
+We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic.
+Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
+This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
+
 Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
@@
 These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used.
-As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool phylobayes.
+As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
 Most other conditions supported Ctenophora-sister.
@@
 Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand).
-The rate of excahnge can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites.
+The rate of exchange can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites.
 The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary.
-Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software tool CAT and has been widely applied to questions of deep animal phylogeny, and the C models of iqtree that have until now not been applied to the problem.
+Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software PhyloBayes  and has been widely applied to questions of deep animal phylogeny, and the CAT-like C10 – C60 models (C models) employed in maximum likelihood (ML) inference until now has not been applied to the problem.
 
@@
 It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic].
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
 In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
@@
 Feuda *et al*.
-[@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
+2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
 By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
 By reanalyzing Chang et al.
-2015 [@chang2015genomic] and Whelan et al.
+2015 [@chang2015genomic] and Whelan *et al*.
 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
@@
 ## Overview of published analyses
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 17 data matrices from 12 studies (Table 1).
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 1 data matrices from 12 studies (Table 1).
 These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.
@@
 Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
-However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Extended Data Fig.
+However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig.
 1).
@@
 1).
-The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot.
+The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies.
 In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
@@
 3).
-Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.
+Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in all other cases.
 
@@
 4; Supplementary Table 3).
-Importantly, we also include a CAT-like, site-heterogeneous C10-C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
+Importantly, we also include a CAT-like, site-heterogeneous C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
 
@@
 We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
 We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
-It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive.
-The GTR+C models requires to estimate up to over 10,000 (e.g.
-GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models.
+It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive.
+The GTR+C60 models requires to estimate up to over 10,000 (e.g.
+GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous+C60 models.
 We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
@@
 This has been addressed with CAT models.
-IQtree provides a new family of C models that also address site heterogeneity.
-Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
+IQtree provides a new family of C60 models that also address site heterogeneity.
+Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared IQtree C60 results to CAT results for a subset of matrices to see if they give consistent results.
 This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-For computational efficiency, we first only applied Poisson+CAT (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups.
 It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
 
-Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices.
+Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices.
 Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices.
-Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015.
+Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015.
 Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large].
 The Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and site-homogeneous models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
+For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
 We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models.
-Thus, our reanalyzing results are largely consistent with previous analyses, both suggesting that the support of Porifera-sister is increasing with more complex models and closely related outgroups.
+These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups.
 More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.
@@
 1A); T2: Porifera-sister (Fig.
-1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig.
+1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Fig.
 5).
@@
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
-This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
+This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
 
 In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
 5B; Supplementary Table 4).
-It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
-In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig.
+It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C60 models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs.
 5A, C).
@@
 site-heterogeneous models).
-More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in Whelan2017_full matrices.
+More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Whelan2017_full matrices.
 
@@
 The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
+
+### Sensitive test of CAT categories {place holder for sensitive results}
+The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
+CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
+Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis ().
+The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
+Although the true number of substitutional categories in empirical data matrices is impossible with current bioinformatic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results,  potentially indicating that the C60 models might be already overfitting to most matrices.
+These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses.
+
+We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices.
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices.Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories.
+The major transition occurred at 90 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four choanimalia matrices.
+Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than site-homogeneous models.
+
+The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model.
+Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups.
+Interestingly, we found both Whelan2017_full and Whelan2017_strict matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
+Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
+Thus, our results also strongly suggested that the exchangeability matrix of evolutionary model (a more simplistic Poisson model and more realistic GTR model) also has strong impact to the rooting position of animal phylogeny in PhyloBayes.
 
@@
 By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
-Three factors contribute significantly to the previous incongruent results based on our analyses.
+Four factors contribute significantly to the previous incongruent results based on our analyses.
 First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
@@
 More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice.
-Third, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
+Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny.
+Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
 Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.
@@
 
-All files and bioinformatic commands associated with this analysis are available at https://github.com/caseywdunn/animal_root
+All tree files, intermidiate results and bioinformatic commands associated with this study are available at https://github.com/caseywdunn/animal_root.
 
@@
 We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self).
-We ran BUSCO with default parameters for all sequences against the provided metazoa gene set.
+We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
 We also ran a BLAST+ blastp search against the SwissProt [@boeckmann2003swiss] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
@@
 #### Phylogenetic analyses in IQtree
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 17 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
 Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
 site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C models in the model testing.
-For computational efficiency, the GTR with C models were not included in model testing since it requires to estimate over 10,000 parameters.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree v1.6.7 [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
+For computational efficiency, the GTR with C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
 For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
@@
 Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
-Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C model or CAT site-heterogeneous models.
+Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
 To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
@@
 
-#### Outgroup taxa sampling with C10-C60 and CAT models
+#### Outgroup taxa sampling with C60 and CAT models
 Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
@@
 Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C models IQtree and Poisson-CAT models in PhyloBayes were conducted.
+For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted.
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI [@Lartillot:2013fg].
-To minimize computational burden, CAT-GTR models were only performed in the.
+Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI v1.8 [@Lartillot:2013fg].
+To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 
@@
 A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-For those matrices that were not converged, PhyloByes analyses were run for at least one month.
+For those matrices that were not converged, PhyloByes analyses were run for at least two weeks.
 
@@
 For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
+
+#### Sensitive analyses with different number of substitutional categories
 

### 2020-03-09 manuscript/manuscript.rmd
update manuscript and add senstive analyses

@@
 
-Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
+Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
 However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
@@
 Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions.
-We next quantify the  support of phylogenetic signal and conduct sensitive analyses in several representative datasets.
-We found that the Porifera-sister can only be recovered with a combined factors in certain data matrices, with inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories and more closely related outgroups, whereas Ctenophora-sister is strongly supported in most other cases.
+Our reanalyses are generally consistent with previous results.
+We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices.
+We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
 We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic.
@@
 Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
-IN this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
+In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
 One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model.
 To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).
+Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here.
 If there are no modifiers, then it is implied that site homogeneous models are used.
-Moreover, the Gamma-rate heterogeinity is used in almost every analysis.
 
@@
 
-High-throughput sequencing allows investigators interested in the root of animal phylogeny to assemble matrices with hundreds or thousands of genes (Table 1).
+High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
 Surprisingly, gene overlap between studies has rarely been assessed.
-Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), animal root studies relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
-These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered.
+Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
+These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
 As a result, the genes selected for analysis can vary widely between studies.
@@
 Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated-- Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
 It has been suggested the inclusion of more distant related (e.g.
@@
 ## Overview of published analyses
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 1 data matrices from 12 studies (Table 1).
-These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 12 studies (Table 1, Table ./data_processed/tables/previously_published_analyses.tsv).
+These matrices include all relevant studies published before 2018 for which we could obtain data matrices with gene partitions that described boundaries between genes.
 
@@
 These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
-Importantly, several earlier studies only include less than five ctenophore species (Extended Data Fig.
+Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
 1).
@@
 2A).
-The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
+The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
 The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
@@
 Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
-2C).
-We found Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
+2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
 2D).
@@
 Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015.
 It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
@@
 3).
-Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in all other cases.
+Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.
 
@@
 One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
+Thus, it easily causes confusions with the increase number of studies.
 Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
@@
 4; Supplementary Table 3).
-Importantly, we also include a CAT-like, site-heterogeneous C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
+Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
 
@@
 We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
-It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive.
-The GTR+C60 models requires to estimate up to over 10,000 (e.g.
-GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous+C60 models.
+It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive with large number of parameters that need to be estimated.
 We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig.
+Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
 4; Supplementary Table 3).
@@
 For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
-We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models.
-These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups.
-More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.
+We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices.
+These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT.
+Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.
 
@@
 In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
 5B; Supplementary Table 4).
@@
 site-heterogeneous models).
-More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Whelan2017_full matrices.
+More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.
 
@@
 The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
-CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
-Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis ().
+CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix [@Lartillot:2004dq] .
+Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX).
 The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
-Although the true number of substitutional categories in empirical data matrices is impossible with current bioinformatic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results,  potentially indicating that the C60 models might be already overfitting to most matrices.
-These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses.
+Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices.
+These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
 
 We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices.
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices.Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories.
-The major transition occurred at 90 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four choanimalia matrices.
-Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than site-homogeneous models.
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar models.
+Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories.
+The major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four representative Choanimalia matrices.
+Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
 
@@
 Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
-Thus, our results also strongly suggested that the exchangeability matrix of evolutionary model (a more simplistic Poisson model and more realistic GTR model) also has strong impact to the rooting position of animal phylogeny in PhyloBayes.
+Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
 
@@
 ## Future direction
-
 Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
@@
 #### Sensitive analyses with different number of substitutional categories
+To explore how the number of categories used in CAT model
 

### 2020-03-24 manuscript/manuscript.rmd
progress on fig5, not finished though.

@@
 library( gridExtra )
-library(ggpubr)
+library( ggpubr )
+library( grid )
+library( gtable )
 theme_set(theme_pubr())
@@
 ```{r phylogenetic_signal, fig.width=6, fig.height=12}
-#au_tests %>%
-#  unite(Matrix, c(manuscript, matrix, model)) %>%
-#  group_by(Matrix, result) %>%
-#  count() %>%
-#  group_by(Matrix) %>%
-#  mutate(percent = prop.table(n)) %>%
-
-#ggplot(aes(x = Matrix, y = percent, fill=result)) +
-#  geom_bar(stat = "identity") +
-#  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-#  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-#  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))
-
-au_tests = read_tsv("../trees_new/AU_test/summary.tsv")
-
- Philippe2009 = filter(au_tests,
-                   grepl('Philippe', Matrix))
-
-x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanimalia_WAG", "Philippe2009_LG", "Philippe2009_Choanimalia_LG", "Philippe2009_LG (PhyloBayes)", "Philippe2009_Choanimalia_LG (PhyloBayes)", "Philippe2009_WAG+C60", "Philippe2009_Choanimalia_WAG+C60", "Philippe2009_Poisson+CAT (PhyloBayes)", "Philippe2009_Choanimalia_Poisson+CAT (PhyloBayes)"))
-y1=factor(Philippe2009$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
-
- Ryan2013_est = filter(au_tests,
-                   grepl('Ryan', Matrix))
-
-y2=factor(Ryan2013_est$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
-
-x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanimalia_WAG", "Ryan2013_est_LG", "Ryan2013_Choanimalia_LG",  "Ryan2013_est_WAG+C60", "Ryan2013_Choanimalia_WAG+C60"))
-
-Whelan2017 = filter(au_tests,
-                   grepl('Whelan', Matrix))
-y3=factor(Whelan2017$Catogary, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))
-
-x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanimalia_WAG", "Whelan2017_full_LG", "Whelan2017_Choanimalia_LG", "Whelan2017_full_LG (PhyloBayes)", "Whelan2017_Choanimalia_LG (PhyloBayes)", "Whelan2017_full_LG+C60", "Whelan2017_Choanimalia_LG+C60", "Whelan2017_full_Poisson+CAT (PhyloBayes)", "Whelan2017_Choanimalia_Poisson+CAT (PhyloBayes)"))
-
-p1 <- ggplot(Philippe2009, aes(x = x1, y = Percentage, fill=y1)) +
+fig5_bars <- au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+
+ggplot(aes(x = Matrix, y = percent, fill=result)) +
   geom_bar(stat = "identity") +
-  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Philippe2009", y = "Catogary of phylgoenetic signal")
-
-p2 <- ggplot(Ryan2013_est, aes(x = x2, y = Percentage, fill=y2)) +
-  geom_bar(stat = "identity") +
-  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Ryan2013_est", y = "Catogary of phylgoenetic signal")
-
-p3 <- ggplot(Whelan2017, aes(x = x3, y = Percentage, fill=y3)) +
-  geom_bar(stat = "identity") +
-  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
-  theme(text = element_text(size=8),axis.text.x = element_text(angle=30, hjust=1)) +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))+ labs(x ="Ryan2013_est", y = "Catogary of phylgoenetic signal")
-
-ggarrange(p1, p2, p3,
-                    labels = c("A", "B", "C"), widths=c(10,6,10),
-                    ncol = 1, nrow = 3,  common.legend = TRUE, legend = "right", align="hv")
-
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
+  coord_flip() +
+  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
+
+fig5_tbl <- au_tests %>%
+select(c(manuscript, matrix, model)) %>%
+distinct()
+
+fig_5_ylabs_grob <- tableGrob(fig5_tbl)
+fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
+fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
+
+fig5_bars_grob <- ggplotGrob(fig5_bars)
+#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
+
+grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
+# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
+
+# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid, 
+#                    widths = unit(c(4, 4), "cm"), 
+#                    heights = unit(c(2, 5), c("in", "lines")))
+# 
+# grid.newpage()
+# grid.draw(fig5)
 ```

### 2020-03-31 manuscript/manuscript.rmd
standin for fig5 for now

@@
 ```{r phylogenetic_signal, fig.width=6, fig.height=12}
-fig5_bars <- au_tests %>%
+# fig5_bars <- au_tests %>%
+au_tests %>%
   unite(Matrix, c(manuscript, matrix, model)) %>%
@@
 
-fig5_tbl <- au_tests %>%
-select(c(manuscript, matrix, model)) %>%
-distinct()
-
-fig_5_ylabs_grob <- tableGrob(fig5_tbl)
-fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
-fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
-
-fig5_bars_grob <- ggplotGrob(fig5_bars)
+#fig5_tbl <- au_tests %>%
+#select(c(manuscript, matrix, model)) %>%
+#distinct()
+
+#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
+#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
+#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
+
+#fig5_bars_grob <- ggplotGrob(fig5_bars)
 #fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
@@
 
-grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
+#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
 # https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html

### 2020-04-07 manuscript/manuscript.rmd

@@
 ## Abstract (ideally less than 200 words)
-
+(Current version)
 Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
@@
 
+(Casey's edit)
 Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
@@
 A detailed description of each component is in Supplementary Results.
-Briefly, the exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
 Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand).
@@
 ## Overview of published analyses
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 12 studies (Table 1, Table ./data_processed/tables/previously_published_analyses.tsv).
-These matrices include all relevant studies published before 2018 for which we could obtain data matrices with gene partitions that described boundaries between genes.
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
+These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partitions that described boundaries between genes.
 
@@
 **Fig.
-Extended Data Fig.
-1.** Each of the primary matrices considered here, color coded by taxon sampling.
+2.** Matrix taxa and gene sampling.
+Each of the primary matrices considered here, color coded by taxon sampling.
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
@@
 ### Matrix gene composition and overlap
-
-Gene sampling also plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny.
-It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination].
-However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring].
-Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices.
 
@@
 Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
-The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of orthologs.
-However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig.
+However, our results show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig.
 2).
@@
 
-Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1).
+Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1).
 We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
@@
 
-### Sensitive test of CAT categories {place holder for sensitive results}
+### Sensitive test of CAT categories
 The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
@@
 
-We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices.
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar models.
-Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories.
-The major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four representative Choanimalia matrices.
+We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories ( in the representative data matrices (Fig.
+XX).
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models.
+Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
+XX).
+More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
+For example, the major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
 Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
@@
 Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
+
+```{r phylogenetic_signal, fig.width=6, fig.height=12}
+df2=read_tsv("/animal_root/data_processed/tables/parameters.tsv")
+df2$Runs= factor( df2$Runs, levels=c("Poisson_CAT60", "Poisson_CAT70", "Poisson_CAT80", "Poisson_CAT90","Poisson_CAT100","Poisson_CAT110","Poisson_CAT120", "Poisson_CAT150", "Poisson_CAT180", "Poisson_CAT360") )
+
+p<-ggplot(df2, aes(x=Runs, y=Likelihood, group=Topology)) +
+    geom_line(aes(color=Topology))+
+    geom_point(aes(color=Topology)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="CAT categories", y = "Log-likelihood")
+```
 
@@
 ## The current state of understanding and future direction
-Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven difficult, even with large genome-scale phylogenomic data matrices.
-By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by more complex CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
+Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
+By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by site-heterogeneous CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
 Four factors contribute significantly to the previous incongruent results based on our analyses.
@@
 We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
-For computational efficiency, the GTR with C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
+For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
 For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
@@
 
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
 For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
@@
 For those matrices that were not converged, PhyloByes analyses were run for at least two weeks.
+We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
 
@@
 2017 [@shen2017contentious].
-For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.
+
+For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [Place holder for Xingxing]
 
 #### Sensitive analyses with different number of substitutional categories
-To explore how the number of categories used in CAT model
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+All phyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
 

### 2020-04-07 manuscript/manuscript.rmd
add burnin and change names for sensitve analyses

@@
 
-### Sensitive test of CAT categories
+### Sensitive test of CAT categories {place holder for sensitive results}
 The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
@@
 Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
-
-```{r phylogenetic_signal, fig.width=6, fig.height=12}
-df2=read_tsv("/animal_root/data_processed/tables/parameters.tsv")
-df2$Runs= factor( df2$Runs, levels=c("Poisson_CAT60", "Poisson_CAT70", "Poisson_CAT80", "Poisson_CAT90","Poisson_CAT100","Poisson_CAT110","Poisson_CAT120", "Poisson_CAT150", "Poisson_CAT180", "Poisson_CAT360") )
-
-p<-ggplot(df2, aes(x=Runs, y=Likelihood, group=Topology)) +
-    geom_line(aes(color=Topology))+
-    geom_point(aes(color=Topology)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="CAT categories", y = "Log-likelihood")
-```
 

### 2020-04-07 manuscript/manuscript.rmd

@@
 
-We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
+We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
 2A).

### 2020-04-14 manuscript/manuscript.rmd
renamed code blocks, moved some text around

@@
 
-# Rooting the animal tree of life - more is less for Ctenophora-sister?
+# Rooting the animal tree of life
 
@@
 
-## Abstract (ideally less than 200 words)
-(Current version)
-Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
-However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
-This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution.
-Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions.
-Our reanalyses are generally consistent with previous results.
-We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices.
-We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
-We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic.
-Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
-This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
-
-(Casey's edit)
+## Abstract
+
 Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
@@
 These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used.
-As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
+As other authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
 Most other conditions supported Ctenophora-sister.
@@
 This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
-We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
+We also quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
 We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic.
@@
 
-## Main
+## Introduction
 
@@
 
-```{r overview data matrices}
-tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
-	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
-
-```
-
 ## Results and Discussion
@@
 
-```{r taxon_rectangles, fig.height = 18, fig.width = 20 }
+```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
 
@@
 
-```{r gene_composition,, fig.height = 6, fig.width = 10}
-matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
-
-p1 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
-
-matrix_summary=matrix_summary %>%
-  mutate(matrix=recode(matrix,`NA`="Simion2017"))
-
-matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-
-p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
-
-p3 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
-	coord_fixed()
-
-p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
-
- ggarrange(p1, p2,p3, p4,
-                    labels = c("A", "B", "C", "D"),
-                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
-
-```
-
-**Extended Data Fig.
-2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
-(B).
-The percentage BUSCO annotations in each matrix.
-(C).
-The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
-(D).
-The percentage  ribosomal genes in each matrix.
-
-```{r matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
-	unlist(recursive=FALSE) %>%
-	bind_rows()
-
-ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="Genes") +
-	scale_y_continuous(name="Species") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
-
-```
-
-**Fig.
-2.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
 ### Support for Porifera-sister and Ctenophora-sister
 
-```{r support_published_analyses}
+```{r figure_support_analyses_published}
 analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
@@
 ```
-**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Supplementary Table 2).
+**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XX).
 
@@
 
-```{r support_new_analyses}
+```{r figure_support_analyses_new}
 analyses_new=analyses_new %>%
@@
 ```
-**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table 3).
+**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table XX).
 
@@
 
-**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
-```{r modelfinder}
-analyses_new %>%
-	filter(modelfinder==TRUE) %>%
-	select( matrix, clade, result, model_summary ) %>%
-	kable()
-
-```
-
 ### Comparison of IQtree and PhyloBayes results
@@
 
-```{r phylogenetic_signal, fig.width=6, fig.height=12}
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
 # fig5_bars <- au_tests %>%
@@
 The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
-
-![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Extended Data Fig.
-3.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
-(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
-(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
-The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
 
@@
 
-## Future direction
 Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
@@
 
+## Supplementary Methods
+
+### Matrix mapping
+
+Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast.
+Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.
+
+### Taxa comparison across matrices
+
+The primary intent of comparing taxa across matrices was to validate our taxon name reconciliation across studies.
+
+We first considered pairwise similarity between the same species from different matrices in different studies.
+
 ## Ackowledgements
@@
 
-# Supplemental Information
+# Supplementary Information
 
@@
 
-## Supplementary Methods
-
-### Matrix mapping
-
-Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast.
-Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.
-
-### Taxa comparison across matrices
-
-The primary intent of comparing taxa across matrices was to validate our taxon name reconciliation across studies.
-
-We first considered pairwise similarity between the same species from different matrices in different studies.
-
 ### Partition comparison across matrices
@@
 
-## Supplementary Tables
-```{r phylogenetic signal distribution}
+## Extended Tables
+
+```{r table_study_summary}
+tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
+	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
+
+```
+
+```{r table_phylogenetic_signal}
 tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
@@
 
+**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
+```{r table_modelfinder}
+analyses_new %>%
+	filter(modelfinder==TRUE) %>%
+	select( matrix, clade, result, model_summary ) %>%
+	kable()
+
+```
+
+## Extended Figures
+
+```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
+matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
+
+p1 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
+
+matrix_summary=matrix_summary %>%
+  mutate(matrix=recode(matrix,`NA`="Simion2017"))
+
+matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+
+p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
+
+p3 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
+	coord_fixed()
+
+p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
+
+ ggarrange(p1, p2,p3, p4,
+                    labels = c("A", "B", "C", "D"),
+                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
+
+```
+
+**Extended Data Fig.
+2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+(A).
+The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+(B).
+The percentage BUSCO annotations in each matrix.
+(C).
+The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
+(D).
+The percentage  ribosomal genes in each matrix.
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
+	unlist(recursive=FALSE) %>%
+	bind_rows()
+
+ggplot(data=matrix_overlap_rectangles) +
+	scale_x_continuous(name="Genes") +
+	scale_y_continuous(name="Species") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
+	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
+
+```
+
+**Fig.
+2.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
+
+![](figures/Figure_Pairwise_heatmap_animal_root.png)
+**Extended Data Fig.
+3.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
+(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
+(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
+The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
+The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
+
+## Supplementary Information
+
 ## References

### 2020-04-14 manuscript/manuscript.rmd
merging dev back to master

--- 2020-04-07 manuscript/manuscript.rmd
@@
 ```{r phylogenetic_signal, fig.width=6, fig.height=12}
-# fig5_bars <- au_tests %>%
-au_tests %>%
+fig5_bars = au_tests %>%
   unite(Matrix, c(manuscript, matrix, model)) %>%
@@
 
-#fig5_tbl <- au_tests %>%
-#select(c(manuscript, matrix, model)) %>%
-#distinct()
-
-#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
-#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
-#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
-
-#fig5_bars_grob <- ggplotGrob(fig5_bars)
-#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
-
-#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
-# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
-
-# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
-#                    widths = unit(c(4, 4), "cm"),
-#                    heights = unit(c(2, 5), c("in", "lines")))
-#
-# grid.newpage()
-# grid.draw(fig5)
+fig5_tbl = au_tests %>%
+select(c(manuscript, matrix, model)) %>%
+distinct()
+
+fig_5_ylabs = tableGrob(fig5_tbl, rows = NULL)
+fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
+fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
+
+fig5 = ggplotGrob(fig5_bars)
+ncol_bars = ncol(fig5)
+# fig5 = gtable_add_rows(gtable(widths = fig5$widths, heights = fig_5_ylabs$heights[0]), 
+#                        heights = unit.c(fig5$heights[1:4],  fig_5_ylabs$heights[0], fig5$heights[5:nrow(fig5)]),
+#                        pos=5)
+fig5 = gtable_add_cols(fig5, sum(fig_5_ylabs$widths))
+fig5 = gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol_bars, b=nrow(fig5)-5, r=ncol(fig5))
+#fig5 = gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
+
+grid.newpage()
+grid.draw(fig5)
 ```

### 2020-04-14 manuscript/functions.R

@@ -298,11 +298,26 @@ parse_tree_pb = function( tree_file, taxonomy_reference ){
 	bpdiff_file_path = sub( tree_ext, ".bpdiff", tree_file_path )
 	filename_parts = strsplit( basename( tree_file ), "\\." ) %>% unlist()
 	matrix_name = filename_parts[1]
-	model_name = "CAT+F81"
-	pb_treelist = sprintf(
-		sub( str_c("bpcomp",tree_ext), "phy_Chain%d.treelist", tree_file_path ),
-		1:2
-	)
+	if (filename_parts[2] == 'bpcomp' || filename_parts[2] == 'phy') {
+	  model_name = "CAT+F81"
+	  pb_treelist = sprintf(
+	    sub( str_c("bpcomp",tree_ext), "phy_Chain%d.treelist", tree_file_path ),
+	    1:2
+	  )
+	} else if (filename_parts[2] == 'phy') {
+	  model_name = "CAT+F81"
+	  pb_treelist = sprintf(
+	    sub( str_c("phy",tree_ext), "phy_Chain%d.treelist", tree_file_path ),
+	    1:2
+	  )
+	} else {
+	  model_name = sub('phy_', '', filename_parts[2])
+	  pb_treelist = sprintf(
+	    sub( tree_ext, "_Chain_%d.treelist", tree_file_path ),
+	    1:2
+	  )
+	}
+
 	
 	if(!file.exists(bpdiff_file_path)){
 		warning( str_c( "Missing bpdiff file: ", bpdiff_file_path ) )
@@ -323,7 +338,7 @@ parse_tree_pb = function( tree_file, taxonomy_reference ){
 	burnins = as.numeric(burnin_line[2])
 	
 	keep_every = 100
-	treelist_text = c( unlist( mapply( read_pb_treelist, pb_treelist, burnins, keep_every ) ) )
+	treelist_text = c( unlist( mapply( read_pb_treelist, pb_treelist, burnin=burnins, subsample_every=keep_every ) ) )
 	sample_trees = read.tree( text = treelist_text )
 	
 	Ctenophora_sister = lapply(
@@ -525,8 +540,8 @@ parse_au_gene_tests = function(path="../trees_new/AU_test/genewise"){
 		map(read.delim) %>%
 		imap(~ transform(.x, manuscript = filename_parts_list[[.y]][[1]])) %>%
 	  imap(~ transform(.x, matrix = ifelse( length(filename_parts_list[[.y]])>2,
[…]

### 2020-04-14 manuscript/manuscript_kernel.R

@@ -195,23 +195,30 @@ matrix_overlap = matrix_overlap[mask,]
 
 # New analyses of published matrices
 
+# read iqtrees
 trees_path_iqtree = "../trees_new/iqtree"
 iqtree_ext = "\\.treefile$"
 file_names_iqtree = list.files( path = trees_path_iqtree, pattern = iqtree_ext, full.names = TRUE )
 
 # Temp fix re issue #8
-file_names_iqtree = file_names_iqtree[ ! grepl("Philippe2009", file_names_iqtree) ]
+#file_names_iqtree = file_names_iqtree[ ! grepl("Philippe2009", file_names_iqtree) ]
 #file_names_iqtree = file_names_iqtree[ ! grepl("Nosenko2013", file_names_iqtree) ]
 
-# read iqtrees
 trees_iq = foreach( tree_file = file_names_iqtree ) %dopar%
 	parse_tree_iqtree( tree_file, taxonomy_reference )
 
-trees_path_pb = "../trees_new/phylobayes"
+# read phylobayes sensitivity analyses
+trees_path_sensitive = "../trees_new/sensitive"
 pb_tree_ext = "\\.con\\.tre$"
-file_names_pb = list.files( path = trees_path_pb, pattern = pb_tree_ext, full.names = TRUE )
+
+file_names_sensitive = list.files( path = trees_path_sensitive, pattern = pb_tree_ext, full.names = TRUE )
+trees_sensitive = foreach( tree_file = file_names_sensitive ) %dopar%
+  parse_tree_pb( tree_file, taxonomy_reference )
 
 # read pb trees
+trees_path_pb = "../trees_new/phylobayes"
+file_names_pb = list.files( path = trees_path_pb, pattern = pb_tree_ext, full.names = TRUE )
+
 trees_pb = foreach( tree_file = file_names_pb ) %dopar%
            parse_tree_pb( tree_file, taxonomy_reference )
 

### 2020-04-14 manuscript/manuscript.rmd

--- 2020-04-14 manuscript/manuscript.rmd
@@
 ## Abstract
-
+XX(Current Version)
+Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
+However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
+This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution.
+Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions.
+Our reanalyses are generally consistent with previous results.
+We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices.
+We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
+We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic.
+Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
+This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
+
+XX(Previous version)
 Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
@@
 These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used.
-As other authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
+As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
 Most other conditions supported Ctenophora-sister.
@@
 This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
-We also quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
+We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
 We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic.
@@
 
-## Introduction
-
+## Main
 Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
@@
 **Fig.
-1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
+XXFigure_overview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
 (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
@@
 
-## Results and Discussion
-
 ### Matrix taxon composition
@@
 **Fig.
-2.** Matrix taxa and gene sampling.
+XXtaxon_rectangles.** Matrix taxa and gene sampling.
 Each of the primary matrices considered here, color coded by taxon sampling.
@@
 ```{r figure_support_analyses_published}
-analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
-x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
-analyses_published %>%
+aanalyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
+analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
+x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
+p1 = analyses_published_filter %>%
 	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
@@
    xlab("Model complexity") + ylab("Clade")
-```
-**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XX).
+
+analyses_new=analyses_new %>%
+mutate(model=recode(model,
+                    `GTR20`="GTR",
+                    `CAT+F81`="Poisson+CAT"))
+
+x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
+p2 = analyses_new %>%
+	ggplot( aes(x=x2, y=clade, col=result, shape=inference) ) +
+		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		scale_colour_manual(values = result_colors) +
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
+       xlab("Model complexity") + ylab("Clade")
+
+ggarrange(p1, p2, ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")
+
+```
+**Fig.
+XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalyses conducted in this study.
+(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (Table XX).
+(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (Supplementary Table XX).
 
 Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
-3 and Supplementary Table 2).
+3A and Supplementary Table 2).
 Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
@@
 Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
-3).
+3A).
 Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.
@@
 ## New analyses of published matrices
-
-```{r figure_support_analyses_new}
-analyses_new=analyses_new %>%
-mutate(model=recode(model,
-                    `GTR20`="GTR",
-                    `CAT+F81`="Poisson+CAT"))
-
-x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
-analyses_new %>%
-	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
-		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-       xlab("Model complexity") + ylab("Clade")
-```
-**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table XX).
 
@@
 We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
-4; Supplementary Table 3).
+3B; Supplementary Table 3).
 Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
@@
 Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
-4; Supplementary Table 3).
+3B; Supplementary Table 3).
 
@@
 Extended Data Fig.
-3).
+3A).
 Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
@@
 More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.
-
-```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-# fig5_bars <- au_tests %>%
-au_tests %>%
-  unite(Matrix, c(manuscript, matrix, model)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
-
-ggplot(aes(x = Matrix, y = percent, fill=result)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
-  coord_flip() +
-  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
-
-#fig5_tbl <- au_tests %>%
-#select(c(manuscript, matrix, model)) %>%
-#distinct()
-
-#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
-#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
-#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
-
-#fig5_bars_grob <- ggplotGrob(fig5_bars)
-#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
-
-#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
-# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
-
-# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
-#                    widths = unit(c(4, 4), "cm"),
-#                    heights = unit(c(2, 5), c("in", "lines")))
-#
-# grid.newpage()
-# grid.draw(fig5)
-```
-
-**Fig.5** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
-(A).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
-(B).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
-(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
-The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
 
@@
 
+``` {r sensitive, fig.width=8, fig.height=8}
+sensitive = read_tsv("/animal_root/data_processed/tables/sensitive.tsv")
+Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )
+Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT60" , "Poisson_CAT90","Poisson_CAT120","Poisson_CAT150","Poisson_CAT180"))
+p1<-ggplot(Philippe2009, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanozoa")
+
+Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_Holozoa")
+Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT360", "Poisson_CAT400","Poisson_CAT440","Poisson_CAT480"))
+p2<-ggplot(Philippe2009, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Holozoa")
+
+Whelan2017_full_only_choanozoa = sensitive %>% filter(Matrix == "Whelan2017_full_only_choanozoa")
+Whelan2017_full_only_choanozoa$Model = factor(Whelan2017_full_only_choanozoa$Model, levels=c("Poisson_CAT60" , "Poisson_CAT340","Poisson_CAT380","Poisson_CAT420","Poisson_CAT460"))
+p3<-ggplot(Whelan2017_full_only_choanozoa, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Whelan2017_full_Choanozoa")
+
+Whelan2017_strict = sensitive %>% filter(Matrix == "Whelan2017_strict" )
+Whelan2017_strict$Model = factor(Whelan2017_strict$Model, levels=c("Poisson_CAT60" , "Poisson_CAT70","Poisson_CAT80","Poisson_CAT90","Poisson_CAT100","Poisson_CAT110", "Poisson_CAT120","Poisson_CAT150","Poisson_CAT180", "Poisson_CAT360"))
+p4<-ggplot(Whelan2017_strict, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support")+ ggtitle("Whelan2017_strict")
+
+ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
+          p3,p4, labels = c("C", "D"))
+```
+**Fig. XXsensitive.**
+
 ### Bias of data-recoding
@@
 Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
-4).
+3B).
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
@@
 These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
-
-![](figures/Figure_random_rocoding.png)
-
-**Extended Data Fig.
-4.** (A) The two alternative hypotheses for deep animal relationships considered here.
-Relationships that are not part of these hypotheses are shown as unresolved polytomies.
-(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
-
-![](figures/Figure_recoding_summary.png)
-**Extended Data Fig.
-6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-
-![](figures/Figure_recoding_pp.png)
-
-**Extended Data Fig.
-5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
-These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
-The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
-The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Cells are color coded as in Supplemental Fig XXRecoidng_summary.
-Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
-The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
 
@@
 
-### Data selection and wrangling
+### Data selection and wrangling (Ben)
 
@@
 
-### Matrix comparison and annotation
-
-#### Taxon name reconciliation
+### Matrix comparison and annotation (Ben)
+
+#### Taxon name reconciliation(Ben)
 
@@
 
-#### Sequence Comparisons
+#### Sequence Comparisons (Ben)
 
@@
 
-#### Partition Network
+#### Partition Network (Ben)
 
@@
 Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-4) give insight into the technical variance of the inference process.
+3B) give insight into the technical variance of the inference process.
 
@@
 
-## Supplementary Methods
-
-### Matrix mapping
+### Matrix mapping (Ben)
 
@@
 
-### Taxa comparison across matrices
+### Taxa comparison across matrices (Ben)
 
@@
 ## Author contributions
+
+## Extended Tables
+
+```{r table_study_summary}
+tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
+	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
+
+```
+
+```{r table_phylogenetic_signal}
+tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
+	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
+
+```
+
+**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
+```{r table_modelfinder}
+analyses_new %>%
+	filter(modelfinder==TRUE) %>%
+	select( matrix, clade, result, model_summary ) %>%
+	kable()
+
+```
+
+## Extended Figures
+
+```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
+matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
+
+p1 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
+
+matrix_summary=matrix_summary %>%
+  mutate(matrix=recode(matrix,`NA`="Simion2017"))
+
+matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+
+p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
+
+p3 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
+	coord_fixed()
+
+p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
+
+ ggarrange(p1, p2,p3, p4,
+                    labels = c("A", "B", "C", "D"),
+                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
+
+```
+
+**Extended Data Fig.
+XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+(A).
+The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+(B).
+The percentage BUSCO annotations in each matrix.
+(C).
+The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
+(D).
+The percentage  ribosomal genes in each matrix.
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
+	unlist(recursive=FALSE) %>%
+	bind_rows()
+
+ggplot(data=matrix_overlap_rectangles) +
+	scale_x_continuous(name="Genes") +
+	scale_y_continuous(name="Species") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
+	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
+
+```
+
+**Extended Data Fig.
+XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
+
+![](figures/Figure_Pairwise_heatmap_animal_root.png)
+**Extended Data Fig.
+XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
+(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
+(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
+The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
+The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
+
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+# fig5_bars <- au_tests %>%
+au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+
+ggplot(aes(x = Matrix, y = percent, fill=result)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
+  coord_flip() +
+  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
+
+#fig5_tbl <- au_tests %>%
+#select(c(manuscript, matrix, model)) %>%
+#distinct()
+
+#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
+#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
+#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
+
+#fig5_bars_grob <- ggplotGrob(fig5_bars)
+#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
+
+#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
+# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
+
+# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
+#                    widths = unit(c(4, 4), "cm"),
+#                    heights = unit(c(2, 5), c("in", "lines")))
+#
+# grid.newpage()
+# grid.draw(fig5)
+```
+**Extended Data Fig.
+XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
+(A).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
+(B).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
+(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
+The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
+
+![](figures/Figure_random_rocoding.png)
+
+**Extended Data Fig.
+XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here.
+Relationships that are not part of these hypotheses are shown as unresolved polytomies.
+(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+![](figures/Figure_recoding_pp.png)
+
+**Extended Data Fig.
+XXrecoding_pp.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
+The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
+The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
+Cells are color coded as in Supplemental Fig XXRecoidng_summary.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
+The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
+
+![](figures/Figure_recoding_summary.png)
+**Extended Data Fig.
+XXrecoding_summary.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
 
@@
 Extended Data Fig.
-4).
+3B).
 Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
@@
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-4).
+3B).
 
@@
 
-### Partition comparison across matrices
+### Partition comparison across matrices (XXBen, do we need this?)
 
@@
 
-## Extended Tables
-
-```{r table_study_summary}
-tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
-	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
-
-```
-
-```{r table_phylogenetic_signal}
-tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
-	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
-
-```
-
-**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
-```{r table_modelfinder}
-analyses_new %>%
-	filter(modelfinder==TRUE) %>%
-	select( matrix, clade, result, model_summary ) %>%
-	kable()
-
-```
-
-## Extended Figures
-
-```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
-matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
-
-p1 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
-
-matrix_summary=matrix_summary %>%
-  mutate(matrix=recode(matrix,`NA`="Simion2017"))
-
-matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-
-p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
-
-p3 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
-	coord_fixed()
-
-p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
-
- ggarrange(p1, p2,p3, p4,
-                    labels = c("A", "B", "C", "D"),
-                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
-
-```
-
-**Extended Data Fig.
-2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
-(B).
-The percentage BUSCO annotations in each matrix.
-(C).
-The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
-(D).
-The percentage  ribosomal genes in each matrix.
-
-```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
-	unlist(recursive=FALSE) %>%
-	bind_rows()
-
-ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="Genes") +
-	scale_y_continuous(name="Species") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
-
-```
-
-**Fig.
-2.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
-![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Extended Data Fig.
-3.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
-(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
-(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
-The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
-
-## Supplementary Information
-
 ## References

### 2020-04-14 manuscript/manuscript.rmd
pass at methods

@@
 The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
-CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix [@Lartillot:2004dq] .
+CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
 Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX).
@@
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding.
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
@@
 
-All tree files, intermidiate results and bioinformatic commands associated with this study are available at https://github.com/caseywdunn/animal_root.
-
-### Data selection and wrangling (Ben)
+All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.
+
+### Data selection and wrangling
 
 We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
-We manually edited some minor formatting to make the batch processing of the matrices *en masse*, e.g.
-standardizing the formatting of charset blocks.
+We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
 All changes made are tracked with git.
 
-### Matrix comparison and annotation (Ben)
-
-#### Taxon name reconciliation(Ben)
+### Matrix comparison and annotation
+
+#### Taxon name reconciliation
 
 We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also use a table where manual entries were needed (Supplementary Table 5, reconciliation/taxonomy_info/manual_taxonomy_map.tsv), e.g.
-authors of original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (reconciliation/taxonomy_info/taxon_table.tsv).
-
-#### Sequence Comparisons (Ben)
-
-Using the partition files for each matrix, we isolated each sequence for each taxon from each partition.
+We also used a table where manual entries were needed (Supplementary Table 5, manual_taxonomy_map.tsv), e.g.
+authors of the original matrix indicate species name in original manuscript.
+For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (taxon_table.tsv).
+
+#### Sequence comparisons
+
+Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
 Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
 
-We used DIAMOND [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
-We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 78.0, eValue < 1e-15, no self->self).
+We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
+We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
 We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
-We also ran a BLAST+ blastp search against the SwissProt [@boeckmann2003swiss] database, filtering such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
-
-#### Partition Network (Ben)
+We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+
+#### Partition network
 
@@
 
-We constructed a network with Python and NetworkX [@Hagberg:2008] v2.2 where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
+We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
 We extracted each connected component from this network.
@@
 We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See Supplementary Table 7 [partition_network_summary table] for a summary tally of each part of the comparison.
+See Supplementary Table 7 [XX partition_network_summary table] for a summary tally of each part of the comparison.
 
@@
 #### Phylogenetic analyses in IQtree
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to elucidate the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic or genomic data (Table 1).
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
 Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
 site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree v1.6.7 [@nguyen2014iq], including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
 We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
 For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
-For large matrices like Hejnol2009, Borrowiec2015 and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
+For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
 
@@
 For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
-2009 [@hejnol2009assessing], since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
 It should be noted that the main conclusion of Simion *et al*.
@@
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes MPI v1.8 [@Lartillot:2013fg].
+Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8.
 To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
@@
 A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-For those matrices that were not converged, PhyloByes analyses were run for at least two weeks.
+For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
 We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
 
-#### phylogenetic distribution of support
+#### Phylogenetic distribution of support
 To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
@@
 
-For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [Place holder for Xingxing]
+For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
 
@@
 To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All phyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
 
@@
 
-Each new matrix was analyzed with PhyloBayes-mpi.
+Each new matrix was analyzed with PhyloBayes-MPI.
 Analyses were run for 1000 generations, and a 200 generation burnin applied.
@@
 
-### Matrix mapping (Ben)
-
-Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast.
-Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.
-
-### Taxa comparison across matrices (Ben)
-
-The primary intent of comparing taxa across matrices was to validate our taxon name reconciliation across studies.
-
-We first considered pairwise similarity between the same species from different matrices in different studies.
-
 ## Ackowledgements
 
-We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam cluster.
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
 
@@
 
-- CAT site heterogeneous [@Lartillot:2004dq].
+- CAT site heterogeneous.
 Each site is assigned to a class with its own equilibrium frequencies.
@@
 This resulted in the 64-taxon matrix that is the focus of most of their analyses.
-Phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
+Phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes, and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
 
@@
 Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained.
-The phylogenetic analyses were conducted under the Poisson+CAT model in phylobayes [@Lartillot:2004dq].
+The phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes.
 
@@
 
-All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping.
+All their analyses were conducted under the F81+CAT+Gamma model in PhyloBayes, in both a Bayesian framework and with bootstrapping.
 All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
@@
 By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister).
-Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in phylobayes [@Lartillot:2004dq].
+Phylogenetic analyses were conducted under the Poisson+CAT, GTR+CAT and GTR models in PhyloBayes.
 
@@
 The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction.
-Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in phylobayes.
+Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes.
 They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
@@
 we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding.
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
@@
 
-### Partition comparison across matrices (XXBen, do we need this?)
-
-```{r partition_mapping}
-
-# Investigate connected components of genes (ie partitions) that include multiple members fromt he same matrix
-#clusterstats_within_manuscripts = partition_map_global %>% group_by( matrix, component_number ) %>%  summarise(n=n()) %>% arrange(desc(n))
-#clusterstats = partition_map_global %>% group_by( component_number ) %>% summarise(n=n()) %>% arrange(desc(n))
-
-#clusterstats_within_manuscripts %>% print(n=100)
-#clusterstats %>% print(n=100)
-#partition_network_summary %>% print()
-
-```
-
-The count for a partition pair can be much larger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.
-
-There are `r length(unique(partition_network_summary$matrix))` matrices.
-A gene that is perfectly sampled would form a cluster with this size.
-To our surprise, very few clusters, though, are this size.
-This suggests that intersection of genes between matrices is extreme low.
-This finding suggested genes used for exploring the position of the root of the animal phylogeny
-
 ## References

### 2020-04-26 manuscript/manuscript.rmd
first pass to the manuscript

@@
 output:
+  word_document: default
+  pdf_document: default
   html_document: default
-  pdf_document: default
-  word_document: default
   github_document: default
@@
 
-## Abstract
-XX(Current Version)
+## Summary (ideally no more than 200 words, currently 251 words)
+XX This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards
+
 Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
@@
 
-XX(Previous version)
-Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages.
-However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
-This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution.
-Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes.
-These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used.
-As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes.
-Most other conditions supported Ctenophora-sister.
-It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups.
-We perform new analyses with these data to characterize the differences between matrices under consistent conditions.
-Our reanalyses are generally consistent with previous results.
-We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister.
-This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
-We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models).
-We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic.
-Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
-This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
-
-## Main
+## Main (The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text), currently is ~4500 words...)
 Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
@@
 Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the amino acid equilibrium frequencies $\Pi$.
-A detailed description of each component is in Supplementary Results.
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
-Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand).
-The rate of exchange can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites.
-The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary.
-Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny, and the CAT-like C10 – C60 models (C models) employed in maximum likelihood (ML) inference until now has not been applied to the problem.
-
+The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$ , the rate of evolution, and the amino acid equilibrium frequencies $\Pi$.
+The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR).
+The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites.
+Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
+Models of molecular evolution are assembled by selecting different options for all these different components.
+The models that are applied in practice are heavily influenced by computational costs, as well as convention.
+For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices.
+LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
+This is further confused by the abbreviations that are used for models.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components– CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
+In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT).
+Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here.
+If there are no modifiers, then it is implied that site homogeneous models are used.
+A detailed description of each component is provided in Supplementary Results.
+
+### Gene and taxa sampling
+
+High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
+Surprisingly, gene overlap between studies has rarely been assessed.
+Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
+These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
+As a result, the genes selected for analysis can vary widely between studies.
+
+Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig.
+1C).
+Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
+Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+It has been suggested the inclusion of more distant related (e.g.
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
+In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
+
+### Data-recoding
+
+Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving].
+Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie].
+Feuda *et al*.
+2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
+By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
+However, how the recoding methods impact phylogenetic inference has not been well explored.
+Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
+
+## Overview of published analyses
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
+These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partitions that described boundaries between genes.
+
+### Matrix taxon composition
+
+Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
+However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig.
+1).
+These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
+Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
+1).
+The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies.
+In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
+Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.
+
+```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
+
+no_subsets = unlist(lapply(sequence_matrices, function(x) !grepl("only|_D", x@matrix_name)))
+
+taxon_rectangles = lapply(
+	sequence_matrices[no_subsets],
+	function ( sequence_matrix ){
+		clade_rects(sequence_matrix)
+	}
+) %>%
+bind_rows()
+
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
+
+clade_colors = c("#e31a1c","#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#a6cee3","#fdbf6f","#cab2d6", "#ff7f00")
+names( clade_colors ) = clades
+
+ggplot(data=taxon_rectangles) +
+	scale_x_continuous(name="Gene sampling") +
+	scale_y_continuous(name="Taxa sampling") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
+	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
+ scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
+	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
+	facet_wrap( ~ matrix )
+
+```
+
+**Fig.
+XXtaxon_rectangles.** Matrix taxa and gene sampling.
+Each of the primary matrices considered here, color coded by taxon sampling.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+
+### Matrix gene composition and overlap
+
+We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
+2A).
+The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco].
+The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
+Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
+However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig.
+2B).
+Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
+2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
+2D).
+We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
+Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig.
+2).
+Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
+
+### Support for Porifera-sister and Ctenophora-sister
+
+```{r figure_support_analyses_published}
+aanalyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
+analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
+x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
+p1 = analyses_published_filter %>%
+	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
+		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		scale_colour_manual(values = result_colors) +
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
+   xlab("Model complexity") + ylab("Clade")
+
+analyses_new=analyses_new %>%
+mutate(model=recode(model,
+                    `GTR20`="GTR",
+                    `CAT+F81`="Poisson+CAT"))
+
+x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
+p2 = analyses_new %>%
+	ggplot( aes(x=x2, y=clade, col=result, shape=inference) ) +
+		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+		scale_colour_manual(values = result_colors) +
+		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
+       xlab("Model complexity") + ylab("Clade")
+
+ggarrange(p1, p2, ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")
+
+```
+**Fig.
+XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalyses conducted in this study.
+(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (Table XX).
+(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (Supplementary Table XX).
+
+Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+3A and Supplementary Table 2).
+Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
+Moreover, three studies are secondary studies without any new data matrices.
+Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
+The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015.
+It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
+The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
+Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
+3A).
+Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.
+
+## New analyses of published matrices
+
+One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
+Thus, it easily causes confusions with the increase number of studies.
+Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
+We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
+3B; Supplementary Table 3).
+Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
+
+Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1).
+We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
+We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
+We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
+Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
+Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
+3B; Supplementary Table 3).
+
+### Comparison of IQtree and PhyloBayes results
+
+Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister.
+This has been addressed with CAT models.
+IQtree provides a new family of C60 models that also address site heterogeneity.
+Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared C60 results to CAT results for a subset of matrices to see if they give consistent results.
+This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
+It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
+
+Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices.
+Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices.
+Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015.
+Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large].
+The Ctenophora-sister were strongly supported in all other matrices.
+For the matrices have conflicted results between Poisson-CAT and  IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
+We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices.
+These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT.
+Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.
+
+### Phylogenetic signal
+
+The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
+To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+1A); T2: Porifera-sister (Fig.
+1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Fig.
+5).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
+This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
+
+In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+5B; Supplementary Table 4).
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs.
+5A, C).
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
+The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
+
+### Sensitive test of CAT categories {place holder for sensitive results}
+The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
+CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
+Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX).
+The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
+Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices.
+These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
+
+We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories in the representative data matrices (Fig.
+XX).
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models.
+Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
+XX).
+More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
+For example, the major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
+Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
+
+The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model.
+Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups.
+Interestingly, we found both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
+Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
+Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
+
+``` {r sensitive, fig.width=8, fig.height=8}
+sensitive = read_tsv("/animal_root/data_processed/tables/sensitive.tsv")
+Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )
+Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT60" , "Poisson_CAT90","Poisson_CAT120","Poisson_CAT150","Poisson_CAT180"))
+p1<-ggplot(Philippe2009, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanozoa")
+
+Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_Holozoa")
+Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT360", "Poisson_CAT400","Poisson_CAT440","Poisson_CAT480"))
+p2<-ggplot(Philippe2009, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Holozoa")
+
+Whelan2017_full_only_choanozoa = sensitive %>% filter(Matrix == "Whelan2017_full_only_choanozoa")
+Whelan2017_full_only_choanozoa$Model = factor(Whelan2017_full_only_choanozoa$Model, levels=c("Poisson_CAT60" , "Poisson_CAT340","Poisson_CAT380","Poisson_CAT420","Poisson_CAT460"))
+p3<-ggplot(Whelan2017_full_only_choanozoa, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Whelan2017_full_Choanozoa")
+
+Whelan2017_strict = sensitive %>% filter(Matrix == "Whelan2017_strict" )
+Whelan2017_strict$Model = factor(Whelan2017_strict$Model, levels=c("Poisson_CAT60" , "Poisson_CAT70","Poisson_CAT80","Poisson_CAT90","Poisson_CAT100","Poisson_CAT110", "Poisson_CAT120","Poisson_CAT150","Poisson_CAT180", "Poisson_CAT360"))
+p4<-ggplot(Whelan2017_strict, aes(Model))+
+  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle=45, hjust=1)) +
+  labs(x ="Number of CAT substitutional categories", y = "Support")+ ggtitle("Whelan2017_strict")
+
+ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
+          p3,p4, labels = c("C", "D"))
+```
+**Fig. XXsensitive.**
+
+### Bias of data-recoding
+There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
+3B).
+Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
+5).
+Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
+
+## The current state of understanding and future direction
+Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
+By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by site-heterogeneous CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
+Four factors contribute significantly to the previous incongruent results based on our analyses.
+First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
+Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models.
+More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice.
+Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny.
+Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
+Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.
+
+Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
+Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution.
+First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
+Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
+Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
+More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
+Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
+We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
+
+## Methods
+
+All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.
+
+### Data selection and wrangling
+
+We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
+We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
+All changes made are tracked with git.
+
+### Matrix comparison and annotation
+
+#### Taxon name reconciliation
+
+We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
+We also used a table where manual entries were needed (Supplementary Table 5, manual_taxonomy_map.tsv), e.g.
+authors of the original matrix indicate species name in original manuscript.
+For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (taxon_table.tsv).
+
+#### Sequence comparisons
+
+Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
+Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
+
+We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
+We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
+We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
+We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+
+#### Partition network
+
+We used the sequence similarity comparisons described above to compare partitions.
+
+We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
+We extracted each connected component from this network.
+We further split these components if the the most connected node (i.e.
+most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
+We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+See Supplementary Table 7 [XX partition_network_summary table] for a summary tally of each part of the comparison.
+
+### Phylogenetic analyses
+
+#### Phylogenetic analyses in IQtree
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
+site-heterogeneous vs.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
+For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
+For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
+
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above.
+Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
+Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
+To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
+For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
+2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+It should be noted that the main conclusion of Simion *et al*.
+was also based on selection of 25% of genes for their jackknife approach.
+
+#### Outgroup taxa sampling with C60 and CAT models
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
+These datasets include the same set of genes but differ in the composition of outgroup species.
+Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
+For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted.
+The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
+Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8.
+To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
+Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
+Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
+A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
+We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
+
+#### Phylogenetic distribution of support
+To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
+We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
+For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
+The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
+The constraint trees were conducted in a customized R script.
+The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al.
+2017 [@shen2017contentious].
+
+For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
+
+#### Sensitive analyses with different number of substitutional categories
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
+
+#### Performance of data-recoding
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
+The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
+
+The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
+This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
+
+The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
+To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
+This creates new recodings that have the same sized groups as SR-6.
+The new recodings were, from `random-00` to `random-03`:
+
+    MPKE AY IDGQ TRC WNLF SVH
+    EIFT WL QVPG NKM SCHA RYD
+    LCQS GK WPTI VRD YEFN MAH
+    IWQV TY KDLM ESH APCF GRN
+
+To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
+
+The 20 state matrices are the same across all analyses since they are not recoded.
+Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
+3B) give insight into the technical variance of the inference process.
+
+Each new matrix was analyzed with PhyloBayes-MPI.
+Analyses were run for 1000 generations, and a 200 generation burnin applied.
+The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
+
+The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+
+## Ackowledgements
+
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
+
+## Author contributions
+
+## Extended Tables
+
+```{r table_study_summary}
+tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
+	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
+
+```
+
+```{r table_phylogenetic_signal}
+tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
+	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
+
+```
+
+**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
+```{r table_modelfinder}
+analyses_new %>%
+	filter(modelfinder==TRUE) %>%
+	select( matrix, clade, result, model_summary ) %>%
+	kable()
+
+```
+
+## Extended Figures
+
+```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
+matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
+
+p1 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
+
+matrix_summary=matrix_summary %>%
+  mutate(matrix=recode(matrix,`NA`="Simion2017"))
+
+matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+
+p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
+
+p3 = ggplot( matrix_summary ) +
+	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
+	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
+	coord_fixed()
+
+p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
+
+ ggarrange(p1, p2,p3, p4,
+                    labels = c("A", "B", "C", "D"),
+                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
+
+```
+
+**Extended Data Fig.
+XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+(A).
+The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
+(B).
+The percentage BUSCO annotations in each matrix.
+(C).
+The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
+(D).
+The percentage  ribosomal genes in each matrix.
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
+	unlist(recursive=FALSE) %>%
+	bind_rows()
+
+ggplot(data=matrix_overlap_rectangles) +
+	scale_x_continuous(name="Genes") +
+	scale_y_continuous(name="Species") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
+	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
+
+```
+
+**Extended Data Fig.
+XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
+
+![](figures/Figure_Pairwise_heatmap_animal_root.png)
+**Extended Data Fig.
+XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
+(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
+(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
+The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
+The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
+
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+# fig5_bars <- au_tests %>%
+au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+
+ggplot(aes(x = Matrix, y = percent, fill=result)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
+  coord_flip() +
+  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
+
+#fig5_tbl <- au_tests %>%
+#select(c(manuscript, matrix, model)) %>%
+#distinct()
+
+#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
+#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
+#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
+
+#fig5_bars_grob <- ggplotGrob(fig5_bars)
+#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
+#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
+
+#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
+# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
+
+# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
+#                    widths = unit(c(4, 4), "cm"),
+#                    heights = unit(c(2, 5), c("in", "lines")))
+#
+# grid.newpage()
+# grid.draw(fig5)
+```
+**Extended Data Fig.
+XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
+(A).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
+(B).
+Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
+(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
+The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
+
+![](figures/Figure_random_rocoding.png)
+
+**Extended Data Fig.
+XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here.
+Relationships that are not part of these hypotheses are shown as unresolved polytomies.
+(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+![](figures/Figure_recoding_pp.png)
+
+**Extended Data Fig.
+XXrecoding_pp.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
+The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
+The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
+Cells are color coded as in Supplemental Fig XXRecoidng_summary.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
+The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
+
+![](figures/Figure_recoding_summary.png)
+**Extended Data Fig.
+XXrecoding_summary.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+
+\pagebreak
+
+# Supplementary Information
+
+## Supplementary Results
+
+### Models of molecular evolution
 Models of molecular evolution are assembled by selecting different options for all these different components.
@@
 If there are no modifiers, then it is implied that site homogeneous models are used.
-
-### Gene sampling
-
-High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
-Surprisingly, gene overlap between studies has rarely been assessed.
-Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
-These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
-As a result, the genes selected for analysis can vary widely between studies.
-
-### Outgroup and ingroup taxon sampling
-
-Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig.
-1C).
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
-It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
-In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
-
-### Data-recoding
-
-Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving].
-Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie].
-Feuda *et al*.
-2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
-By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
-By reanalyzing Chang et al.
-2015 [@chang2015genomic] and Whelan *et al*.
-2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved].
-However, how the recoding methods impact phylogenetic inference has not been well explored.
-Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
-Thus, it is important to reevalute the performance of the impact of data-recoding compared to non-recoding methods in animal phylogeny.
-
-## Overview of published analyses
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
-These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partitions that described boundaries between genes.
-
-### Matrix taxon composition
-
-Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
-However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig.
-1).
-These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
-Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
-1).
-The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies.
-In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
-Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.
-
-```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
-
-no_subsets = unlist(lapply(sequence_matrices, function(x) !grepl("only|_D", x@matrix_name)))
-
-taxon_rectangles = lapply(
-	sequence_matrices[no_subsets],
-	function ( sequence_matrix ){
-		clade_rects(sequence_matrix)
-	}
-) %>%
-bind_rows()
-
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
-
-clade_colors = c("#e31a1c","#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#a6cee3","#fdbf6f","#cab2d6", "#ff7f00")
-names( clade_colors ) = clades
-
-ggplot(data=taxon_rectangles) +
-	scale_x_continuous(name="Gene sampling") +
-	scale_y_continuous(name="Taxa sampling") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
-	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
- scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
-	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
-	facet_wrap( ~ matrix )
-
-```
-
-**Fig.
-XXtaxon_rectangles.** Matrix taxa and gene sampling.
-Each of the primary matrices considered here, color coded by taxon sampling.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-
-### Matrix gene composition and overlap
-
-We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
-2A).
-The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco].
-The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
-Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
-However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig.
-2B).
-Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
-2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
-2D).
-We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
-
-Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated since all these studies rely on their own method for orthology inference.
-Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support.
-However, our results show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig.
-2).
-Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
-Moreover, the little overlap among studies may potentially lead to incongruent phylogenetic results and difficult to directly compare results across studies.
-
-### Support for Porifera-sister and Ctenophora-sister
-
-```{r figure_support_analyses_published}
-aanalyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
-analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
-x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
-p1 = analyses_published_filter %>%
-	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
-		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-   xlab("Model complexity") + ylab("Clade")
-
-analyses_new=analyses_new %>%
-mutate(model=recode(model,
-                    `GTR20`="GTR",
-                    `CAT+F81`="Poisson+CAT"))
-
-x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
-p2 = analyses_new %>%
-	ggplot( aes(x=x2, y=clade, col=result, shape=inference) ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
-		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-       xlab("Model complexity") + ylab("Clade")
-
-ggarrange(p1, p2, ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")
-
-```
-**Fig.
-XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalyses conducted in this study.
-(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (Table XX).
-(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (Supplementary Table XX).
-
-Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
-3A and Supplementary Table 2).
-Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
-Moreover, three studies are secondary studies without any new data matrices.
-Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015.
-It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
-The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
-3A).
-Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.
-
-## New analyses of published matrices
-
-One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
-Thus, it easily causes confusions with the increase number of studies.
-Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
-3B; Supplementary Table 3).
-Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
-
-Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1).
-We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
-We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
-It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive with large number of parameters that need to be estimated.
-We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
-Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
-3B; Supplementary Table 3).
-
-### Comparison of IQtree and PhyloBayes results
-
-Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister.
-This has been addressed with CAT models.
-IQtree provides a new family of C60 models that also address site heterogeneity.
-Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared IQtree C60 results to CAT results for a subset of matrices to see if they give consistent results.
-This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
-
-Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices.
-Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices.
-Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015.
-Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large].
-The Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
-We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices.
-These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT.
-Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.
-
-### Phylogenetic signal
-
-The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
-To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
-1A); T2: Porifera-sister (Fig.
-1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Fig.
-5).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
-This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
-
-In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
-5B; Supplementary Table 4).
-It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C60 models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny.
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs.
-5A, C).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
-The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
-
-We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig.
-Extended Data Fig.
-3A).
-Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice).
-In other words.
-our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs.
-site-heterogeneous models).
-More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.
-
-### Sensitive test of CAT categories {place holder for sensitive results}
-The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
-CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
-Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX).
-The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
-Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices.
-These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
-
-We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories ( in the representative data matrices (Fig.
-XX).
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models.
-Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
-XX).
-More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
-For example, the major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
-Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
-
-The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model.
-Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups.
-Interestingly, we found both Whelan2017_full and Whelan2017_strict matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
-Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
-Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
-
-``` {r sensitive, fig.width=8, fig.height=8}
-sensitive = read_tsv("/animal_root/data_processed/tables/sensitive.tsv")
-Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )
-Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT60" , "Poisson_CAT90","Poisson_CAT120","Poisson_CAT150","Poisson_CAT180"))
-p1<-ggplot(Philippe2009, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanozoa")
-
-Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_Holozoa")
-Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT360", "Poisson_CAT400","Poisson_CAT440","Poisson_CAT480"))
-p2<-ggplot(Philippe2009, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Holozoa")
-
-Whelan2017_full_only_choanozoa = sensitive %>% filter(Matrix == "Whelan2017_full_only_choanozoa")
-Whelan2017_full_only_choanozoa$Model = factor(Whelan2017_full_only_choanozoa$Model, levels=c("Poisson_CAT60" , "Poisson_CAT340","Poisson_CAT380","Poisson_CAT420","Poisson_CAT460"))
-p3<-ggplot(Whelan2017_full_only_choanozoa, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Whelan2017_full_Choanozoa")
-
-Whelan2017_strict = sensitive %>% filter(Matrix == "Whelan2017_strict" )
-Whelan2017_strict$Model = factor(Whelan2017_strict$Model, levels=c("Poisson_CAT60" , "Poisson_CAT70","Poisson_CAT80","Poisson_CAT90","Poisson_CAT100","Poisson_CAT110", "Poisson_CAT120","Poisson_CAT150","Poisson_CAT180", "Poisson_CAT360"))
-p4<-ggplot(Whelan2017_strict, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support")+ ggtitle("Whelan2017_strict")
-
-ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
-          p3,p4, labels = c("C", "D"))
-```
-**Fig. XXsensitive.**
-
-### Bias of data-recoding
-There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
-They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
-3B).
-Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-5).
-Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
-These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
-
-## The current state of understanding and future direction
-Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
-By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by site-heterogeneous CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
-Four factors contribute significantly to the previous incongruent results based on our analyses.
-First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
-Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models.
-More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice.
-Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny.
-Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
-Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.
-Rather than embracing one hypothesis over another, our results provide an integrative overview of the challenge and uncertainty related to the origin of animals.
-
-Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
-Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution.
-First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
-Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
-Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
-More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
-Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
-We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
-
-## Methods
-
-All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.
-
-### Data selection and wrangling
-
-We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
-We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
-All changes made are tracked with git.
-
-### Matrix comparison and annotation
-
-#### Taxon name reconciliation
-
-We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also used a table where manual entries were needed (Supplementary Table 5, manual_taxonomy_map.tsv), e.g.
-authors of the original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (taxon_table.tsv).
-
-#### Sequence comparisons
-
-Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
-Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
-
-We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
-We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
-We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
-We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
-
-#### Partition network
-
-We used the sequence similarity comparisons described above to compare partitions.
-
-We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
-We extracted each connected component from this network.
-We further split these components if the the most connected node (i.e.
-most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
-We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See Supplementary Table 7 [XX partition_network_summary table] for a summary tally of each part of the comparison.
-
-### Phylogenetic analyses
-
-#### Phylogenetic analyses in IQtree
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
-site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
-For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
-For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
-
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above.
-Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
-Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
-To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
-For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
-2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
-It should be noted that the main conclusion of Simion *et al*.
-was also based on selection of 25% of genes for their jackknife approach.
-
-#### Outgroup taxa sampling with C60 and CAT models
-Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
-These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted.
-The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
-For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
-Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
-Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
-A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
-We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
-
-#### Phylogenetic distribution of support
-To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
-The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
-The constraint trees were conducted in a customized R script.
-The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al.
-2017 [@shen2017contentious].
-
-For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
-
-#### Sensitive analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
-
-#### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
-The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
-
-The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
-This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
-The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
-
-The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
-To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
-This creates new recodings that have the same sized groups as SR-6.
-The new recodings were, from `random-00` to `random-03`:
-
-    MPKE AY IDGQ TRC WNLF SVH
-    EIFT WL QVPG NKM SCHA RYD
-    LCQS GK WPTI VRD YEFN MAH
-    IWQV TY KDLM ESH APCF GRN
-
-To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
-
-The 20 state matrices are the same across all analyses since they are not recoded.
-Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-3B) give insight into the technical variance of the inference process.
-
-Each new matrix was analyzed with PhyloBayes-MPI.
-Analyses were run for 1000 generations, and a 200 generation burnin applied.
-The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
-
-The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
-
-## Ackowledgements
-
-We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
-
-## Author contributions
-
-## Extended Tables
-
-```{r table_study_summary}
-tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
-	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
-
-```
-
-```{r table_phylogenetic_signal}
-tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
-	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
-
-```
-
-**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
-```{r table_modelfinder}
-analyses_new %>%
-	filter(modelfinder==TRUE) %>%
-	select( matrix, clade, result, model_summary ) %>%
-	kable()
-
-```
-
-## Extended Figures
-
-```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
-matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
-
-p1 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
-
-matrix_summary=matrix_summary %>%
-  mutate(matrix=recode(matrix,`NA`="Simion2017"))
-
-matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-
-p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
-
-p3 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
-	coord_fixed()
-
-p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
-
- ggarrange(p1, p2,p3, p4,
-                    labels = c("A", "B", "C", "D"),
-                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
-
-```
-
-**Extended Data Fig.
-XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
-(B).
-The percentage BUSCO annotations in each matrix.
-(C).
-The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
-(D).
-The percentage  ribosomal genes in each matrix.
-
-```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
-	unlist(recursive=FALSE) %>%
-	bind_rows()
-
-ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="Genes") +
-	scale_y_continuous(name="Species") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
-
-```
-
-**Extended Data Fig.
-XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
-![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Extended Data Fig.
-XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
-(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
-(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
-The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
-
-```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-# fig5_bars <- au_tests %>%
-au_tests %>%
-  unite(Matrix, c(manuscript, matrix, model)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
-
-ggplot(aes(x = Matrix, y = percent, fill=result)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
-  coord_flip() +
-  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
-
-#fig5_tbl <- au_tests %>%
-#select(c(manuscript, matrix, model)) %>%
-#distinct()
-
-#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
-#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
-#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
-
-#fig5_bars_grob <- ggplotGrob(fig5_bars)
-#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
-
-#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
-# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
-
-# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
-#                    widths = unit(c(4, 4), "cm"),
-#                    heights = unit(c(2, 5), c("in", "lines")))
-#
-# grid.newpage()
-# grid.draw(fig5)
-```
-**Extended Data Fig.
-XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
-(A).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
-(B).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
-(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
-The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
-
-![](figures/Figure_random_rocoding.png)
-
-**Extended Data Fig.
-XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here.
-Relationships that are not part of these hypotheses are shown as unresolved polytomies.
-(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
-
-![](figures/Figure_recoding_pp.png)
-
-**Extended Data Fig.
-XXrecoding_pp.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
-These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
-The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
-The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Cells are color coded as in Supplemental Fig XXRecoidng_summary.
-Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
-The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
-
-![](figures/Figure_recoding_summary.png)
-**Extended Data Fig.
-XXrecoding_summary.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-
-\pagebreak
-
-# Supplementary Information
-
-## Supplementary Results
-
-### Models of molecular evolution
 

### 2020-04-26 manuscript/manuscript.rmd
change section names

@@
 
-### Matrix gene composition and overlap
+### Matrix gene sampling are largely different across studies
 
@@
 
-## New analyses of published matrices
+## New analyses of published matrices are largely consistent with previous
 
@@
 3B; Supplementary Table 3).
-
-### Comparison of IQtree and PhyloBayes results
 
@@
 
-### Phylogenetic signal
+### Drastic changes of phylogenetic signal between site- homogeneous and heterogeneous models
 
@@
 
-### Sensitive test of CAT categories {place holder for sensitive results}
+### Increase CAT categories lead to different topologies
 The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.

### 2020-04-28 manuscript/manuscript.rmd
reknit after fixing path

@@
 output:
+  pdf_document: default
   word_document: default
-  pdf_document: default
   html_document: default
@@
 ``` {r sensitive, fig.width=8, fig.height=8}
-sensitive = read_tsv("/animal_root/data_processed/tables/sensitive.tsv")
+sensitive = read_tsv("../data_processed/tables/sensitive.tsv")
 Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )

### 2020-04-28 manuscript/manuscript.tex
reknit after fixing path

@@
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
+\hypertarget{summary-ideally-no-more-than-200-words-currently-251-words}{%
+\subsection{Summary (ideally no more than 200 words, currently 251
+words)}\label{summary-ideally-no-more-than-200-words-currently-251-words}}
+
+XX This summary paragraph should be structured as follows: 2-3 sentences
+of basic-level introduction to the field; a brief account of the
+background and rationale of the work; a statement of the main
+conclusions (introduced by the phrase `Here we show' or its equivalent);
+and finally, 2-3 sentences putting the main findings into general
+context so it is clear how the results described in the paper have moved
+the field forwards
 
 Phylogenomic studies based on hundreds to thousands of genes have
-greatly improved our understanding of evolutionary relationships among
-major animal lineages. However, a considerable debate still remains
-regarding the animal origin, with Ctenophora-sister and Porifera-sister
-emerging as the two primary hypotheses, comprising one of the most
-difficult to resolve nodes in animal tree of life. Here, we
-systematically explore the mechanisms that potentially cause the
-incongruent results by synthesizing data and results from all previous
-phylogenomic analyses, and perform new analyses with these data in a
-consistent manner to characterize the differences between studies. We
-find that gene sampling varies across studies, indicating that different
-regions of the genomes are sampled with little overlap among data
-matrices. More importantly, we find Porifera-sister can only be
-recovered with more constraints, including using site-heterogeneous CAT
-model and inclusive of the most closely related outgroups in several
-matrices, whereas Ctenophora-sister is recovered in all other cases. We
-next quantify the support of phylogenetic signal over the two hypotheses
-and find substantial differences in the distribution of phylogenetic
-signal between models and outgroup choice (especially CAT models). Last,
-we show that recoding can be a problematic method for addressing
-compositional variation in phylogenetic analyses. Our results provide a
-comprehensive overview of the current understanding of the first
-diverging animal branch and why different studies yield different
-answers. Rather than embracing one hypothesis over another, we hope our
-efforts provides an integrative overview of the challenge and provides
-direction for future studies to address difficult phylogenetic problems.
-
-\hypertarget{main}{%
-\subsection{Main}\label{main}}
+greatly improved our understanding of evolutionary relationships in the
+animal tree of life. However, considerable debate still remains
+regarding the deepest split in the animal tree of life, with
+Ctenophora-sister and Porifera-sister emerging as the two primary
+alternative hypotheses. This has been one of the most difficult to
+resolve nodes in animal tree of life, and has major implications for our
+understanding of the earliest events in animal evolution. Here we
+systematically explore this question by synthesizing data and results
+from all previous phylogenomic analyses and perform new analyses with
+these data to characterize the differences between matrices under
+consistent conditions. Our reanalyses are generally consistent with
+previous results. We next quantify the support of phylogenetic signal
+and conduct sensitive analyses in several representative data matrices.
+We found that the Porifera-sister can only be recovered with several
+combined factors, including inclusive of a more simplistic evolutionary
+model, an unrealistic large number of substitution categories, more
+closely related outgroups and certain gene samplings, whereas
+Ctenophora-sister is supported in most other conditions. We also show
+that recoding amino acid data into reduced state space, another approach
+that has been suggested to address variation in amino acid frequencies,
+is problematic. Thus, our results provide a comprehensive overview of
+the current understanding of the first diverging animal branch and why
+different studies yield different answers. This synthesis provides an
+integrative overview of the challenge and provides direction for future
+studies to address difficult phylogenetic problems.
+
+\hypertarget{main-the-typical-length-of-an-article-with-3-4-modest-display-items-figures-and-tables-is-2000-2500-words-summary-paragraph-plus-body-text-currently-is-4500-words}{%
+\subsection{Main (The typical length of an article with 3-4 modest
+display items (figures and tables) is 2000-2500 words (summary paragraph
+plus body text), currently is \textasciitilde{}4500
+words\ldots{})}\label{main-the-typical-length-of-an-article-with-3-4-modest-display-items-figures-and-tables-is-2000-2500-words-summary-paragraph-plus-body-text-currently-is-4500-words}}
 
@@
 relationship of Ctenophora to other animals\textsuperscript{1}. An
-understanding of the first diverging animal lineage is critical for
+understanding of the placement of these lineages is critical for
 scientists to reconstruct events that occurred early in animal
@@
 ctenophores are our most distant animal relative. Since then many more
-studies have been published and extensive reanalyzed, some supporting
-Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It
-has become clear that this is a very difficult phylogenetic challenge,
-and as the problem has become better characterized, it has become an
-interesting test-case to phylogenetic biologists beyond those concerned
-with this particular biological problem. Work has been hindered, though,
-because it has been difficult to directly compare results across studies
-and synthesize findings to understand the broader patterns of support.
-Here we synthesize data and results from all previous phylogenomic
-analyses that tested Ctenophora-sister and Porifera-sister, and
-reanalyze these data using standardized methods, and perform new
-analyses to characterize differences between studies. We hope that this
-effort provides an integrative overview of the challenge and provides
-direction for future studies. We also hope that the work we have
-conducted here, including consolidating all the datasets in one place
-with consistent formats and species names, will enhance the technical
-value of this interesting question to methods-focused investigators that
-look to develop methods to address difficult phylogenetic problems.
-
-\includegraphics{figures/Figure_overview.png} \textbf{Fig. 1.} (A) The
-Ctenophora-sister hypothesis posits that there is a clade (designated by
-the orange node) that includes all animals except Ctenophora, and that
-Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis
-posits that there is a clade (designated by the green node) that
-includes all animals except Porifera, and that Porifera is sister to
-this clade. Testing these hypotheses requires evaluating the support for
-each of these alternative nodes. (C) The animals and their outgroup
-choice, showing the three progressively more inclusive clades
-Choanimalia, Holozoa, and Opisthokonta (designated by the blue node).
-(D) A ctenophore. (E) A sponge.
+studies have been published, some supporting Ctenophora-sister, some
+Porifera-sister, and some neither (Table 1). It has become clear that
+this is a very difficult phylogenetic challenge. As the problem has
+become better characterized, it has become an interesting test-case to
+phylogenetic biologists beyond those concerned with this particular
+biological problem. Work has been hindered, though, because it has been
+difficult to directly compare results across studies and synthesize
+findings to understand the broader patterns of support. Here we
+synthesize data and results from phylogenomic analyses that tested
+Ctenophora-sister and Porifera-sister. We reanalyze these data using
+standardized methods, and perform new analyses to characterize
+differences between studies. This provides an integrative overview of
+the challenges posed by this question and provides direction for future
+studies. We also hope that the work we have conducted here, including
+consolidating all the datasets in one place with consistent formats and
+species names, will enhance the technical value of this interesting
+question to methods-focused investigators that look to develop methods
+to address difficult phylogenetic problems.
+
+\includegraphics{figures/Figure_overview.png} \textbf{Fig.
+XXFigure\_overview.} (A) The Ctenophora-sister hypothesis posits that
+there is a clade (designated by the orange node) that includes all
+animals except Ctenophora, and that Ctenophora is sister to this clade.
+(B) The Porifera-sister hypothesis posits that there is a clade
+(designated by the green node) that includes all animals except
+Porifera, and that Porifera is sister to this clade. Testing these
+hypotheses requires evaluating the support for each of these alternative
+nodes. (C) The animals and their outgroup choice, showing the three
+progressively more inclusive clades Choanimalia, Holozoa, and
+Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A
+sponge.
 
@@
 largely differed according to three components: the exchangeability
-matrix \(E\), the rate of evolution, and the state equilibrium
-frequencies \(\Pi\). A detailed description of each component is in
-Supplementary Results. Briefly, (1) the exchangeability matrix \(E\)
-describes the rate at which one amino acid changes to another.
-Exchangeability matrices have been used in the studies under
-consideration here include: Poisson (or F81), WAG, LG, GTR, as they have
-been used regularly in phylogenomic analyses; (2) While the
-exchangeability matrix describes the relative rate of different changes
-between amino acids, the actual rate can be further scaled, such as site
-homogeneous rates and Gamma-rate heterogeneity; (3) the vector of
-equilibrium frequencies \(\Pi\) describes the stationary frequency of
-amino acids, such as empirical site-homogeneous, estimated
-site-homogeneous, CAT site-heterogeneous\textsuperscript{3}, and C10-C60
-models (C models).
-
-Models can be assembled by selecting different options for all these
-different components. The models that are applied in practice area
-heavily influenced by engineering and computational costs, as well as
-convention. For example, on the questions considered here, Poisson and
-GTR exchangeability matrices have only been used in combination with CAT
-site-heterogeneous models of equilibrium frequency. LG and WAG
-exchangeability matrices have only been used with site homogeneous
-estimates of equilibrium frequency. This is further confused by the
-abbreviations that are used for models. Papers often discuss CAT and WAG
-models as if they are exclusive, but these particular terms apply to
-non-exclusive model components-- CAT refers to variation across sites
-and WAG a particular exchangeability matrix. CAT is generally shorthand
-for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium
-frequency estimation. One could, though, run a WAG+CAT model, although
-the computation time is much longer than Poisson+CAT model. To avoid
-confusion on this point, we always specify the exchangeability matrix
-first, followed by modifiers that describe the accommodation of
-heterogeneity in equilibrium frequencies (\emph{e.g.}, CAT). If there
+matrix \(E\) , the rate of evolution, and the amino acid equilibrium
+frequencies \(\Pi\). The exchangeability matrix \(E\) describes the rate
+at which one amino acid changes to another (\emph{e.g.} Poisson/F81, WAG
+or LG, and GTR). The rate of exchange can be further scaled, for example
+with a Gamma parameter that accommodates rate variation across sites.
+Heterogeneous equilibrium approaches include CAT\textsuperscript{3},
+which is implemented in the software PhyloBayes and has been widely
+applied to questions of deep animal phylogeny. Models of molecular
+evolution are assembled by selecting different options for all these
+different components. The models that are applied in practice are
+heavily influenced by computational costs, as well as convention. For
+example, on the questions considered here, CAT site-heterogeneous models
+of equilibrium frequency have only been used in combination with Poisson
+and GTR exchangeability matrices. LG and WAG exchangeability matrices
+have only been used with site homogeneous estimates of equilibrium
+frequency. This is further confused by the abbreviations that are used
+for models. Papers often discuss CAT and WAG models as if they are
+exclusive, but these particular terms apply to non-exclusive model
+components-- CAT refers to equilibrium frequency variation across sites
+and WAG a particular exchangeability matrix. In this literature, CAT is
+generally shorthand for Poisson+CAT and WAG is shorthand for
+WAG+homogeneous equilibrium frequency estimation. To avoid confusion on
+this point, we always specify the exchangeability matrix first, followed
+by modifiers that describe the accommodation of heterogeneity in
+equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate
+heterogeneity is used in almost every analysis conducted here. If there
 are no modifiers, then it is implied that site homogeneous models are
-used. Moreover, the Gamma-rate heterogeinity is used in almost every
-analysis.
-
-\hypertarget{gene-sampling}{%
-\subsubsection{Gene sampling}\label{gene-sampling}}
-
-High-throughput sequencing is leading to exciting new data matrices for
-phylogenomic analyses, potentially including hundreds or thousands of
-genes to explore the root of animal phylogeny (Table 1). Although the
-number of sampled genes might be different across studies, it is
-reasonable to expect that a core set of orthologs should be present and
-conserved across most animal genomes (\emph{e.g.}, animal BUSCO
-genes\textsuperscript{4}). Surprisingly, gene composition and matrix
-overlap has been rarely assessed across studies and most studies relied
-on their own bioinformatic pipeline for determination of orthologs (for
-details see Supplementary Results).
-
-\hypertarget{outgroup-and-ingroup-taxon-sampling}{%
-\subsubsection{Outgroup and ingroup taxon
-sampling}\label{outgroup-and-ingroup-taxon-sampling}}
-
-Outgroup selection (the taxa used to root the ingroup taxa) can
-potentially affect the phylogenetic inference of animal evolution (Fig.
-1C). Several studies have progressively removed more distantly related
+used. A detailed description of each component is provided in
+Supplementary Results.
+
+\hypertarget{gene-and-taxa-sampling}{%
+\subsubsection{Gene and taxa sampling}\label{gene-and-taxa-sampling}}
+
+High-throughput sequencing allows investigators interested in the
+phylogeny of early diverging animals to assemble matrices with hundreds
+or thousands of genes (Table 1). Surprisingly, gene overlap between
+studies has rarely been assessed. Though there are many genes with
+homologs present in most known animal genomes (\emph{e.g.}, animal BUSCO
+genes\textsuperscript{4}), studies of animal phylogeny often relied on a
+wide variety of different bioinformatic approaches to identifying and
+selecting genes for their matrices. These take different approaches to
+sequence comparison, filtering, and orthology assessment, and gene
+selection can be impacted by which datasets are considered (see
+supplementary text for more details). As a result, the genes selected
+for analysis can vary widely between studies.
+
+Outgroup selection (the non-animal taxa used to root the animal tree)
+can affect the phylogenetic inference of animal evolution (Fig. 1C).
+Several studies have progressively removed more distantly related
 outgroup taxa to test the effect of outgroup selection to ingroup
 topology\textsuperscript{5,6}. Three progressively more inclusive clades
-have often been investigated, including Choanomalia (Metazoa plus most
-closely related Choanoflagellata), Holoza (Choanimalia plus more
-distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has
-been suggested the inclusion of more distant related (e.g.~Fungi)
-outgroup taxa tend to recover Ctenophora-sister, whereas increasing
-support of Porifera-sister often recovered when more closely related
-outgroup taxa were used with site-heterogeneous CAT
-models\textsuperscript{6}. In contrast of outgroup selection,
-sensitivity to ingroup sampling has received less attention than
-outgroup sampling. This may be because results have tended to be more
-sensitive to outgroup sampling.
+have often been investigated: Choanomalia (animals plus most closely
+related Choanoflagellata), Holoza (Choanimalia plus more distantly
+related holozoans) and Opisthokonta (Holozoa + Fungi). It has been
+suggested the inclusion of more distant related (e.g.~Fungi) outgroup
+taxa tend to recover Ctenophora-sister, whereas increasing support of
+Porifera-sister often recovered when more closely related outgroup taxa
+were used with site-heterogeneous Poisson-CAT models\textsuperscript{6}.
+In contrast of outgroup selection, sensitivity to ingroup sampling has
+received less attention.
 
@@
 
-Compositional heterogeneity (lineage-specific differences in amino acid
-frequencies) was long thought has a major effect in resolving the animal
-tree of life, especially for the deep animal
+Compositional heterogeneity (clade-specific and site-specific
+differences in amino acid frequencies) have been a concern in the study
+of phylogenetics, and has been of particular concern for the deep animal
 phylogeny\textsuperscript{7}. Variation in evolutionary rates across
@@
 branch attraction of sponges or ctenophores\textsuperscript{8}. Feuda
-\emph{et al}.\textsuperscript{9} has proposed to use data-recoding
+\emph{et al}. 2017\textsuperscript{9} has proposed to use data-recoding
 method to reduce compositional heterogeneity in matrices. By discarding
@@
 to, they sought to reduce the impact of differences between species in
-which particular amino acids within those groups are most frequent. By
-reanalyzing Chang et al.~2015\textsuperscript{10} and Whelan et
-al.~2015\textsuperscript{11}, they report that posterior predictive
-analyses indicate recoded analyses have better model adequacy than
-non-recoding analyses, and that Porifera-sister was strongly favored
-under all recoding strategies using GTR-CAT models\textsuperscript{9}.
+which particular amino acids within those groups are most frequent.
 However, how the recoding methods impact phylogenetic inference has not
@@
 that non-recoding approaches significantly outperformed recoding
-strategies\textsuperscript{12}. Thus, it is important to reevalute the
-performance of the impact of data-recoding compared to non-recoding
-methods in animal phylogeny.
+strategies\textsuperscript{10}.
 
@@
 
-Here, we synthesize data and results from all previous phylogenomic
-analyses that tested Ctenophora-sister and Porifera-sister, including 17
-data matrices from 12 studies (Table 1).
-
-\begin{longtable}[]{@{}lllrrrlll@{}}
-\caption{Table 1. Overview of data matrices used in this
-study}\tabularnewline
-\toprule
-Study & Matrix & Clade & Number of gene partitions & Number of sites &
-Number of taxa & Infered sister lineage & Journal &
-Citations\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-Study & Matrix & Clade & Number of gene partitions & Number of sites &
-Number of taxa & Infered sister lineage & Journal &
-Citations\tabularnewline
-\midrule
-\endhead
-Dunn2008 & Dunn2008 & Opisthokonta & 150 & 21152 & 64 &
-Ctenophora-sister & Nature & NA\tabularnewline
-Philippe2009 & Philippe2009 & Opisthokonta & 129 & 30257 & 55 &
-Porifera-sister & Current Biology & NA\tabularnewline
-Hejnol2009 & Hejnol2009 & Opisthokonta & 1487 & 270580 & 94 &
-Ctenophora-sister & Proc. Biol. Sci. & NA\tabularnewline
-NA & Hejnol2009\_subsampled & Opisthokonta & 844 & 55594 & 94 &
-Ctenophora-sister & NA & NA\tabularnewline
-Nosenko2013 & Nosenko2013\_nonribo\_9187 & Opisthokonta & 35 & 9187 & 63
-& Ctenophora-sister & Molecular Phylogenetics and Evolution &
-NA\tabularnewline
-Nosenko2013 & Nosenko2013\_ribo\_14615 & Opisthokonta & 87 & 14615 & 71
-& Porifera-sister & NA & NA\tabularnewline
-Ryan2013 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 61 &
-Ctenophora-sister & Science & NA\tabularnewline
-Moroz2014 & Moroz2014\_3d & Opisthokonta & 114 & 22772 & 46 &
-Ctenophora-sister & Nature & NA\tabularnewline
-Whelan2015 & Whelan2015\_D1 & Opisthokonta & 251 & 59733 & 76 &
-Ctenophora-sister & PNAS & NA\tabularnewline
-Whelan2015 & Whelan2015\_D10 & Opisthokonta & 210 & 59733 & 70 &
-Ctenophora-sister & NA & NA\tabularnewline
-Whelan2015 & Whelan2015\_D20 & Opisthokonta & 178 & 81008 & 65 &
-Ctenophora-sister & NA & NA\tabularnewline
-Borowiec2015 & Borowiec2015\_Best108 & Choaimalia & 108 & 41808 & 36 &
-Ctenophora-sister & BMC Genomics & NA\tabularnewline
-Borowiec2015 & Borowiec2015\_Total1080 & Choaimalia & 1080 & 384981 & 36
-& Ctenophora-sister & NA & NA\tabularnewline
-Pisani2015 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 60 &
-Porifera-sister & PNAS & NA\tabularnewline
-NA & Whelan\_D6 & Opisthokonta & 115 & 33403 & 70 & Porifera-sister & NA
-& NA\tabularnewline
-Chang2015 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
-Ctenophora-sister & PNAS & NA\tabularnewline
-Simion2017 & Simion2017 & Opisthokonta & 1719 & 401632 & 97 &
-Porifera-sister & Current Biology & NA\tabularnewline
-NA & Simion2017\_subsampled & Opisthokonta & 1719 & 110602 & 97 &
-Porifera-sister & NA & NA\tabularnewline
-Feuda2017 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
-Ctenophora-sister & Current Biology & NA\tabularnewline
-NA & WhelanD20 & Opisthokonta & 178 & 81008 & 65 & Porifera-sister & NA
-& NA\tabularnewline
-Whelan2017 & Whelan2017b\_full & Holozoa & 212 & 68062 & 80 &
-Ctenophora-sister & Nature Ecology and Evolution & NA\tabularnewline
-Whelan2017 & Whelan2017b\_strict & Choaimalia & 59 & 49388 & 76 &
-Ctenophora-sister & NA & NA\tabularnewline
-\bottomrule
-\end{longtable}
-
-\hypertarget{results-and-discussion}{%
-\subsection{Results and Discussion}\label{results-and-discussion}}
+Here, we synthesize data and results from previous phylogenomic analyses
+that tested Ctenophora-sister and Porifera-sister from 14 studies (Table
+1). These matrices include all relevant studies published before 2020
+for which we could obtain data matrices with gene partitions that
+described boundaries between genes.
 
@@
 understood. However, several data matrices have relatively skewed taxon
-sampling, with more than 50\% taxa sampled in bilaterians compared to
+sampling, with more than 50\% taxa sampled in Bilateria compared to
 other major animal lineages (Extended Data Fig. 1). These skewed data
 matrices include two studies constructed from ESTs
-(Dunn2008\textsuperscript{2} and Hejnol2009\textsuperscript{13}) and one
-study from whole-genome data (Borrowiec2015\textsuperscript{14}).
-Importantly, several earlier studies only include less than five
-ctenophore species (Extended Data Fig. 1). The rest data matrices have
-relatively balanced taxon composition, although the total number of
-sampled taxon varies a lot. In general, increasing taxon sampling is
-broadly accepted to improve phylogenetic inference and aid in the
-placement of the earliest diverging animal lineage. Thus, it should be
-noted that Whelan2017\textsuperscript{15} (has the most Ctenophore taxa)
-and Simion2017\textsuperscript{16} (contains the most taxon sampling)
-have much broader taxon sampling than other matrices.
-
-\includegraphics{manuscript_files/figure-latex/taxon_rectangles-1.pdf}
-
-\textbf{Fig. Extended Data Fig. 1.} Each of the primary matrices
-considered here, color coded by taxon sampling. Horizontal size is
-proportional to the number of genes sampled, vertical size to the number
-of taxa sampled.
-
-\hypertarget{matrix-gene-composition-and-overlap}{%
-\subsubsection{Matrix gene composition and
-overlap}\label{matrix-gene-composition-and-overlap}}
-
-Gene sampling also plays an essential role in phylogenomic analyses, and
-most often rely on the concatenation of sampled genes for deep animal
-phylogeny. It has been suggested that the concatenation assumes that
-phylogenetic signal from genes that do not share the species phylogeny
-will be overwhelmed by the signal from the majority of genes whose
-genealogy mirrors that of the species evolutionary
-history\textsuperscript{17}. However, large phylogenomic data matrices
-potentially contain genes with conflicting or weak phylogenetic signal
-that can lead to incongruence phylogenomic results, especially when
-analyzing deeper node represents ancient divergence in the
-tree\textsuperscript{18}. Thus, it is important to evaluate the gene
-composition and matrix overlap to explore level of congruence of gene
-sampling across different matrices.
-
-We first used the Metazoa BUSCO genes\textsuperscript{{\textbf{???}}} to
-evaluate gene composition from all data matrices and the result is
-unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains
-978 single-copy orthologs (identified from 64 animal taxa) that are most
+(Dunn2008\textsuperscript{2} and Hejnol2009\textsuperscript{11}) and one
+study from whole-genome data (Borrowiec2015\textsuperscript{12}).
+Importantly, several earlier studies before 2015 only include less than
+five ctenophore species (Extended Data Fig. 1). The rest data matrices
+have relatively balanced taxon composition, although the total number of
+sampled taxon varies. In general, increasing taxon sampling is broadly
+accepted to improve phylogenetic inference and aid in the placement of
+the earliest diverging animal lineage. Thus, it should be noted that
+Whelan2017\textsuperscript{13} (has the most Ctenophore taxa) and
+Simion2017\textsuperscript{14} (contains the most taxon sampling) have
+much broader taxon sampling than other matrices.
+
+\includegraphics{manuscript_files/figure-latex/figure_taxon_rectangles-1.pdf}
+
+\textbf{Fig. XXtaxon\_rectangles.} Matrix taxa and gene sampling. Each
+of the primary matrices considered here, color coded by taxon sampling.
+Horizontal size is proportional to the number of genes sampled, vertical
+size to the number of taxa sampled.
+
+\hypertarget{matrix-gene-sampling-are-largely-different-across-studies}{%
+\subsubsection{Matrix gene sampling are largely different across
+studies}\label{matrix-gene-sampling-are-largely-different-across-studies}}
+
+We first used the Metazoa BUSCO genes\textsuperscript{4} to evaluate
+gene composition from all data matrices and the result is unexpected
+(Extended Data Fig. 2A). The animal BUSCO dataset contains 978
+single-copy orthologs (identified from 64 animal genomes) that are most
 likely shared by most animal genomes, and are commonly used to assess
-the completeness of genome or transcriptomic
-assemblies\textsuperscript{4}. The BUSCO genes also are frequently used
-as markers in many large-scale phylogenomic studies (e.g.,
-yeasts\textsuperscript{19}, spiders\textsuperscript{20}). Thus, it is
-reasonable to expect that these conserved genes should present at a
-higher percentage than non-BUSCO genes in most matrices. However, we
-found that gene partitions from all matrices comprised less than 50\% of
-BUSCO genes, even with several data matrices contain less than 20
-percent (Extended Data Fig. 2B). Moreover, we identified number of
-ribosomal RNA genes in each data matrix (Extended Data Fig. 2C). We
-found Philippe2009, Nosenko2013\_ribo\_14615 and Chang2015 matrices
-possessed much higher content of ribosomal genes than other matrices
-(Extended Data Fig. 2D). We also annotated each gene partition against
-SwissProt\textsuperscript{21} and Gene Ontology databases, and results
-are shown in Supplementary Table 1.
-
-Previous analyses have mainly focused on outgroup selection and/or model
-fits in different data matrices, and the matrix overlap across different
-studies has been rarely investigated since all these studies rely on
-their own method for orthology inference. Our initial intent of
-comparing matrix overlap across matrices was to construct a new matrix
-with shared taxon and gene sampling to test specific hypotheses about
-differences in support. The hypothesis is that although the
-bioinformatics pipeline used for OG determination is different, most
-larger data matrices should share a core set of orthologs. However, our
-results show that surprisingly little overlap across all the data
-matrices from different studies (Fig. 2). Thus, these results strongly
-suggested that gene sampling largely varies across studies, indicating
-that different regions of the genomes are sampled with little overlap
-among different data matrices. Moreover, the little overlap among
-studies may potentially lead to incongruent phylogenetic results and
-difficult to directly compare results across studies.
-
-\begin{verbatim}
-## <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
-##     aspect: function
-##     backtransform_range: function
-##     clip: on
-##     default: FALSE
-##     distance: function
-##     expand: TRUE
-##     is_free: function
-##     is_linear: function
-##     labels: function
-##     limits: list
-##     modify_scales: function
-##     range: function
-##     ratio: 1
-##     render_axis_h: function
-##     render_axis_v: function
-##     render_bg: function
-##     render_fg: function
-##     setup_data: function
-##     setup_layout: function
-##     setup_panel_params: function
-##     setup_params: function
-##     transform: function
-##     super:  <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
-\end{verbatim}
-
-\includegraphics{manuscript_files/figure-latex/gene_composition,-1.pdf}
-
-\textbf{Extended Data Fig. 2.} Annotation and representation of BUSCO
-and ribosomal genes in each data matrix. (A). The number of partitions
-with BUSCO annotations in each matrix, relative to the number of
-partitions. (B). The percentage BUSCO annotations in each matrix. (C).
-The number of partitions with annotations of ribosomal genes in each
-matrix, relative to the number of partitions. (D). The percentage
-ribosomal genes in each matrix.
-
-\includegraphics{manuscript_files/figure-latex/matrix_overlap-1.pdf}
-
-\textbf{Fig. 2.} Pairwise overlap between each of the primary matrices
-considered here. Horizontal size is proportional to the number of genes
-sampled, vertical size to the number of taxa sampled. The horizontal
-intersection shows the proportions of shared genes, the vertical
-intersection shows the proportions of shared taxa.
+the completeness of assemblies from genomic data\textsuperscript{4}. The
+BUSCO genes also are frequently used as markers in many large-scale
+phylogenomic studies (e.g., yeasts\textsuperscript{15},
+spiders\textsuperscript{16}). Thus, it is reasonable to expect that
+these conserved genes should present at a higher percentage than
+non-BUSCO genes in most matrices. However, we found that gene partitions
+from all matrices comprised less than 50\% of BUSCO genes, even with
+several data matrices contain less than 20 percent (Extended Data Fig.
+2B). Moreover, we identified number of ribosomal RNA genes in each data
+matrix (Extended Data Fig. 2C) and found that Philippe2009,
+Nosenko2013\_ribo\_14615 and Chang2015 matrices possessed much higher
+content of ribosomal genes than other matrices (Extended Data Fig. 2D).
+We also annotated each gene partition against
+SwissProt\textsuperscript{{\textbf{???}}} and Gene Ontology databases,
+and results are shown in Supplementary Table 1. Moreover, our network
+analyses show that surprisingly little overlap of gene samplings across
+all the data matrices from different studies (Fig. 2). Thus, these
+results strongly suggested that gene sampling largely varies across
+studies, indicating that different regions of the genomes are sampled
+with little overlap among different data matrices.
 
@@
 
-\includegraphics{manuscript_files/figure-latex/support_published_analyses-1.pdf}
-\textbf{Fig. 3.} A total of 163 analyses were transcribed from the
-literature (Supplementary Table 2).
+\includegraphics{manuscript_files/figure-latex/figure_support_analyses_published-1.pdf}
+\textbf{Fig. XXfigure\_support\_analyses.} A summary of phylogenomic
+results from the previous studies and reanalyses conducted in this
+study. (A) A total of 135 analyses were transcribed from the literature
+(Table XX). (B) A total of 102 new phylogenomic analyses were conducted
+in this study (Supplementary Table XX).
 
 Here we summarized all the 163 previous phylogenetic analyses from 15
-studies (Fig. 3 and Supplementary Table 2). Among these studies, the
+studies (Fig. 3A and Supplementary Table 2). Among these studies, the
 main conclusions of five studies strongly in favor of Porifera-sister
@@
 more distantly related outgroups and with the use of site-heterogeneious
-CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI
-analyses with CAT model are not well resolved (either not reach
+CAT (Poisson+CAT or GTR+CAT) models in several matrices, including
+Philippe2009, Ryan2013\_est and Whelan2015. It should be noted that many
+BI analyses with CAT model are not well resolved (either not reach
 convergence or do not provide significant support for the key node of
@@
 supported since several analyses strongly in favor Ctenophora-sister
-even with only choanofalgellates are used as outgroups (Fig. 3). Thus,
+even with only choanofalgellates are used as outgroups (Fig. 3A). Thus,
 our summary of previous phylogenetic analyses strongly indicated that
 the trend of the support of Porifera-sister is generally increasing with
-the use of more complex CAT models and more closely related outgroups,
-whereas Ctenophora-sister is supported in all other cases.
-
-\hypertarget{new-analyses-of-published-matrices}{%
-\subsection{New analyses of published
-matrices}\label{new-analyses-of-published-matrices}}
-
-\includegraphics{manuscript_files/figure-latex/support_new_analyses-1.pdf}
-\textbf{Fig. 4.} A total of 84 new phylgoenomic analyses were conducted
-in this study (Supplementary Table 3).
+the use of more complex CAT models and more closely related outgroups in
+several representative matrices, whereas Ctenophora-sister is supported
+in most other conditions.
+
+\hypertarget{new-analyses-of-published-matrices-are-largely-consistent-with-previous}{%
+\subsection{New analyses of published matrices are largely consistent
+with
+previous}\label{new-analyses-of-published-matrices-are-largely-consistent-with-previous}}
 
@@
 have been used across studies, and phylogenetic analysis decisions have
-been approached in very different ways. Here we first reanalyze the
-primary matrices from each study under consistent conditions with
-IQtree\textsuperscript{22} under a panel of evolutionary models. We
+been approached in very different ways. Thus, it easily causes
+confusions with the increase number of studies. Here we first reanalyze
+the primary matrices from each study under consistent conditions with
+IQtree\textsuperscript{17} under a panel of evolutionary models. We
 selected this tool because it has greater model flexibility,
 computational time and accuracy than other tools in maximum likelihood
-(ML) framework\textsuperscript{23} (Fig. 4; Supplementary Table 3).
-Importantly, we also include a CAT-like, site-heterogeneous C10-C60 (C)
-models\textsuperscript{24} that are implemented in both ML and Bayesian
-inference (BI) framework in our analyses and it is interesting to
-compare results between C60 model to CAT models since it has never been
-used in any matrices analyzed here.
-
-Here we reanalyzed 17 matrices from 15 studies that are used as main
+(ML) framework\textsuperscript{18} (Fig. 3B; Supplementary Table 3).
+Importantly, we also include a CAT-like, site-heterogeneous C10 -C60
+(C60) models\textsuperscript{19} that are implemented in both ML
+framework in our analyses and it is interesting to compare results
+between C60 model to CAT models since it has never been used in any
+matrices analyzed here.
+
+Here we reanalyzed 17 matrices from 12 studies that are used as main
 conclusions from their original paper (Table 1). We also progressively
@@
 and compared the relative fit of site-homogeneous and site-heterogeneous
-C models using ModelFinder\textsuperscript{25} in IQtree (except only
+C60 models using ModelFinder\textsuperscript{20} in IQtree (except only
 LG+C60 models are used in 3 large matrices). We found in all cases,
@@
 models under BIC criteria and we then inferred support under the
-best-fit model from IQtree (Extended Data Table 1). It should be noted
-that we didn't include GTR+C models in model testing since it is
-computational too expensive. The GTR+C models requires to estimate up to
-over 10,000 (e.g.~GTR+C60 contains 234*60 parameters) free parameters
-inferred from matrices, whereas only several hundred parameters needs to
-estimate in site-homogeneous + C models. We then analyzed every matrix
-under a panel of standard site-heterogeneous and site-homogeneous
-models, including WAG, GTR and Poisson+C60. Moreover, we also used the
-best-fit model identified above with the removal of all distantly
-related outgroups and only include Choanoflagellata as outgroups.
-Interestingly, with the exception of Moroz2014\_3d and all Nosenko2013
-matrices that supports neither Porifera-sister or Ctenophora-sister, all
-analysis conducted herein in IQtree strongly supported the
-ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C
-models with the most closely related outgroups (Fig. 4; Supplementary
-Table 3).
-
-\textbf{Extended Data Table 1.} The models selected by ModelFinder for
-each matrix.
-
-\begin{longtable}[]{@{}llll@{}}
-\toprule
-matrix & clade & result & model\_summary\tabularnewline
-\midrule
-\endhead
-Borowiec2015\_Best108 & Choanimalia & Ctenophora-sister &
-WAG+C60\tabularnewline
-Chang2015 & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
-Dunn2008 & Opisthokonta & Ctenophora-sister & WAG+C60\tabularnewline
-Moroz2014\_3d & Choanimalia & Unresolved & WAG+C60\tabularnewline
-Nosenko2013\_nonribo\_9187 & Opisthokonta & Unresolved &
-WAG+C60\tabularnewline
-Nosenko2013\_ribo\_11057 & Choanimalia & Unresolved &
-WAG+C60\tabularnewline
-Nosenko2013\_ribo\_14615 & Opisthokonta & Unresolved &
-WAG+C60\tabularnewline
-Ryan2013\_est & Opisthokonta & Ctenophora-sister &
-WAG+C60\tabularnewline
-Whelan2015\_D1 & Opisthokonta & Ctenophora-sister &
-LG+C60\tabularnewline
-Whelan2015\_D10 & Opisthokonta & Ctenophora-sister &
-WAG+C60\tabularnewline
-Whelan2015\_D20 & Opisthokonta & Ctenophora-sister &
-WAG+C60\tabularnewline
-Whelan2017\_full & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
-Whelan2017\_strict & Choanimalia & Ctenophora-sister &
-LG+C60\tabularnewline
-\bottomrule
-\end{longtable}
-
-\hypertarget{comparison-of-iqtree-and-phylobayes-results}{%
-\subsubsection{Comparison of IQtree and PhyloBayes
-results}\label{comparison-of-iqtree-and-phylobayes-results}}
+best-fit model from IQtree (Extended Data Table 1). We then analyzed
+every matrix under a panel of standard site-heterogeneous and
+site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover,
+we also used the best-fit model identified above with the removal of all
+distantly related outgroups and only include Choanoflagellata as
+outgroups. Interestingly, with the exception of Moroz2014\_3d and all
+Nosenko2013 matrices that supports neither Porifera-sister or
+Ctenophora-sister, all analysis conducted herein in IQtree strongly
+supported the ctenophore-sister hypothesis, even with CAT-like,
+site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most
+closely related outgroups (Fig. 3B; Supplementary Table 3).
 
@@
 tests of Ctenophora-sister and Porifera-sister. This has been addressed
-with CAT models. IQtree provides a new family of C models that also
+with CAT models. IQtree provides a new family of C60 models that also
 address site heterogeneity. Given the extensive computational cost and
-concerns about overparameterization of CAT models\textsuperscript{26},
-we compared iqtree C results to CAT results for a subset of matrices to
-see if they give consistent results. This would be of technical interest
+concerns about overparameterization of CAT models\textsuperscript{21},
+we compared C60 results to CAT results for a subset of matrices to see
+if they give consistent results. This would be of technical interest
 because it would reduce the cost of accommodating compositional
-heterogeneity in future analyses. For computational efficiency, we first
-only applied Poisson+CAT (Poisson+CAT) models in every matrix with
-Choanoflagellata as outgroups. It should be noted that most Phylobayes
-runs were converged, although several large matrices have not reached
-convergence after at least a month's computational time.
+heterogeneity in future analyses. It should be noted that most
+Phylobayes runs were converged, although several large matrices have not
+reached convergence after at least a month's computational time.
 
 Interestingly, we found strong supports for both hypotheses when using
-Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we
-found no strong support for either hypothesis in Moroz2013\_3d and
-Nosenko2013 matrices. Consistent with previous
-study\textsuperscript{27,6}, we recovered strong support of
-Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015.
-Interestingly, we also found a strong support of Porifera-sister in two
-Whelan2017 matrices\textsuperscript{15} and a weaker support of
-subsampled matrix of Simion2017\textsuperscript{16}. The
-Ctenophora-sister were strongly supported in all other matrices. For the
-matrices have conflicted results between Poisson-CAT and
-site-homogeneous models, we further ran Poisson+CAT models with
-inclusion of more distantly related outgroups to evaluate how outgroup
-choice may affect phylogenetic inference under CAT model. We found a
-weak support of Porifera-sister in Holozoa matrices and strong support
-of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating
-that the support of Porifera-sister is increasing with inclusive of the
-more closely related outgroups and CAT models. Thus, our reanalyzing
-results are largely consistent with previous analyses, both suggesting
-that the support of Porifera-sister is increasing with more complex
-models and closely related outgroups. More importantly, our results
-indicated that both model choice and outgroup sampling play a major role
-in incongruent results of animal origin in several key data matrices.
-
-\hypertarget{phylogenetic-signal}{%
-\subsubsection{Phylogenetic signal}\label{phylogenetic-signal}}
+Poisson-CAT model in analyzed Choanimalia matrices. Similar to ML
+analyses, we found no strong support for either hypothesis in
+Moroz2013\_3d and Nosenko2013 matrices. Consistent with previous
+study\textsuperscript{22,6}, we recovered support of Porifera-sister in
+matrices of Philippe2009, Ryan2013\_est and Whelan2015. Interestingly,
+we also found a strong support of Porifera-sister in two Whelan2017
+matrices\textsuperscript{13} and a weaker support of subsampled matrix
+of Simion2017\textsuperscript{14}. The Ctenophora-sister were strongly
+supported in all other matrices. For the matrices have conflicted
+results between Poisson-CAT and IQtree, we further ran Poisson+CAT
+models with inclusion of more distantly related outgroups to evaluate
+how outgroup choice may affect phylogenetic inference under CAT model.
+We found a weak support of Porifera-sister in all Holozoa matrices and
+strong support of Ctenophora-sister in all analyzed Opisthokonta
+matrices. These results suggest that Porifera-sister can only be
+recovered in certain gene sampling in several representative matrices
+with the exclusive of more distantly related outgroups when analyzed
+using Poisson-CAT. Thus, our results from reanalyses are largely similar
+to previous studies and indicate that both model choice and outgroup
+sampling play a major role in incongruent results of animal origin in
+several representative data matrices.
+
+\hypertarget{drastic-changes-of-phylogenetic-signal-between-site--homogeneous-and-heterogeneous-models}{%
+\subsubsection{Drastic changes of phylogenetic signal between site-
+homogeneous and heterogeneous
+models}\label{drastic-changes-of-phylogenetic-signal-between-site--homogeneous-and-heterogeneous-models}}
 
@@
 two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2:
-Porifera-sister (Fig. 1B)) to three key data matrices with different
-outgroup choice and models in both ML and BI framework (Fig. 5). By
-calculating gene-wise log-likelihood scores between T1 and T2 for every
-gene (\(\Delta\)GLS) or site (\(\Delta\)SLS) in each matrix, we found
-that Ctenophora-sister had the higher proportions of supporting genes in
-every analysis when using site-homogeneous models (LG or WAG in both
-IQtree or PhyloBayes). Moreover, the outgroup choice has little impact
-on the distribution of the support of phylogenetic signal in
+Porifera-sister (Fig. 1B)) to three representative data matrices with
+different outgroup choice and models in both ML and BI framework (Fig.
+5). By calculating gene-wise log-likelihood scores between T1 and T2 for
+every gene (\(\Delta\)GLS) or site (\(\Delta\)SLS) in each matrix, we
+found that Ctenophora-sister had the higher proportions of supporting
+genes in every analysis when using site-homogeneous models (LG or WAG in
+both IQtree or PhyloBayes). Moreover, the outgroup choice has little
+impact on the distribution of the support of phylogenetic signal in
 sifte-homogeneous models. This finding is largely consistent with the
-past study that majority of phylogenetic signal is strongly favored in
-Ctenophora-sister hypothesis with site-homogeneous models in other data
-matrices\textsuperscript{28}.
+results from past study that majority of phylogenetic signal is strongly
+favored in Ctenophora-sister hypothesis with site-homogeneous models in
+other data matrices\textsuperscript{23}.
 
@@
 Ctenophora-sister also had the higher proportions of supporting genes
-with C models, we found that the phylogenetic signal decreases in many
-genes using C models compared to site-homogeneous models, especially in
-matrices from Ryan2013\_est that nearly 30\% of genes changed from
+with C60 models, we found that the phylogenetic signal decreases in many
+genes using C60 models compared to site-homogeneous models, especially
+in matrices from Ryan2013\_est that nearly 30\% of genes changed from
 strong to week signal (\Delta\$lnL\textless{}2) (Fig. 5B; Supplementary
-Table 4). It is not clear why the phylogenetic signal decreases in many
-genes with a better fitting model as suggested by model testing in
-IQtree, but our results indicated the C models lack of phylogenetic
-signal compared to site-homogeneous models in the node of the root
-position of animal phylogeny. In contrast to the C models, the
-phylogenetic signal largely increased in Poisson-CAT models and more
-importantly, the outgroup choice has a major effect of the distribution
-of phylogenetic signal (Fig. 5A, C). We found that support for
-Ctenophora-sister largely decreases and Porifera-sister increases
-(\textasciitilde{} 20\% of sampled gene partitions) when exclusive of
-more distantly related outgroups with the Poisson-CAT models
-(Supplementary Table 4). The results here are further corroborated by
-our phylogenomic analyses in BI, which the support of Porifera-sister
-generally increases in Holozoa and Choanimalia matrices.
-
-We also examined the level of congruence of gene partitions that are
-supported in each hypothesis across different analyses in
-Whelan2017\_full matrices (Fig. Extended Data Fig. 3). Consistent with
-the summary of \Delta\$GLS~analysis, We found gene population that
-supports in either hypothesis are largely congruent between
-site-homogenous and C models, whereas little overlap was observed in
-Poisson-CAT models (especially with different outgroup choice). In other
-words. our results strongly indicated that the distribution of
-phylogenetic signal of animal origin is largely sensitive to model
-selection (site-homogeneous vs.~site-heterogeneous models). More
-importantly, the site-heterogeneous CAT model is also more sensitive to
-the outgroup choice than site-homogeneous and C models in
-Whelan2017\_full matrices.
-
-\includegraphics{manuscript_files/figure-latex/phylogenetic_signal-1.pdf}
-
-\textbf{Fig.5} The distribution of phylogenetic signal for two
-alternative topological hypotheses on the earliest-branching animal
-lineage with different models and outgroup choice. (A). Proportion of
-genes supporting each of two alternative hypotheses for each of two
-alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes
-supporting each of two alternative hypotheses for each of two
-alternative hypotheses for Ryan2013\_est matrix. (C) Proportion of genes
-supporting each of two alternative hypotheses for each of two
-alternative hypotheses for Whelan2017\_full matrix. The (\Delta\$GLS)
-values for the genes across each data matrix are provided in
-Supplementary Table 4.
-
-\includegraphics{figures/Figure_Pairwise_heatmap_animal_root.png}
-\textbf{Extended Data Fig. 3.} Pairwise intersections of the level of
-congruence of gene partitions that are supported in each hypothesis
-across different analyses in Whelan2017\_full matrices calculated by
-Euclidean distance. (A) Pairwise comparisons across genes that supported
-Ctenophora-sister in different analyses. (B) Pairwise comparisons across
-genes that supported Porifera-sister in different analyses. The color
-scale indicates the Pearson correlation, where 1 is positive linear
-correlation, 0 is no linear correlation, and −1 is negative linear
-correlation. The pairwise intersection heatmap is generated by
-Intervene\textsuperscript{29}.
+Table 4). In contrast to the C60 models, the phylogenetic signal largely
+increased in Poisson-CAT models and more importantly, the outgroup
+choice has a major effect of the distribution of phylogenetic signal
+(Figs. 5A, C). We found that support for Ctenophora-sister largely
+decreases and Porifera-sister increases (\textasciitilde{} 20\% of
+sampled gene partitions) when exclusive of more distantly related
+outgroups with the Poisson-CAT models (Supplementary Table 4). The
+results here are further corroborated by our phylogenomic analyses in
+BI, which the support of Porifera-sister generally increases in Holozoa
+and Choanimalia matrices.
+
+\hypertarget{increase-cat-categories-lead-to-different-topologies}{%
+\subsubsection{Increase CAT categories lead to different
+topologies}\label{increase-cat-categories-lead-to-different-topologies}}
+
+The major difference between site-heterogeneous CAT and C60 models are
+the number of substitutional categories that are used in different data
+matrices. CAT models employ a Dirichlet process prior to infer number of
+substitutional categories or profiles with different nucleotide or amino
+acid frequencies to account for substitutional heterogeneity across a
+phylogenetic data matrix. Here, we reported the number of categories
+inferred from CAT model for each PhyloBayes analysis (Table XX). The
+number of categories is considerable larger than 60 categories used in
+C60 models, and in most cases even larger than the gene partitions
+within each data matrix (Table xx). Although the true number of
+substitutional categories in empirical data matrices is impossible to
+determine with current phylogenetic tools, we found that the amino
+frequencies of several categories inferred by C60 model are close to
+zero from IQtree results, potentially indicating that the C60 models
+might be already overfitting to most matrices. These results are
+consistent with the findings that CAT model often overestimate the true
+number of substitutional categories, especially for large data matrix
+from simulation analyses\textsuperscript{21}.
+
+We next explore how the increase of number of categories may affect the
+phylogenetic inference related to the animal phylogeny with a panel of
+different substitutional categories in the representative data matrices
+(Fig. XX). Surprisingly, we found Poisson+nCAT60 showed exact same
+topology as Poisson+C60 or WAG+C60 with strong support of
+Ctenophora-sister, suggesting that the different phylogenetic analysis
+software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI)
+has limited impact on phylogenetic inference to these matrices with
+similar evolutioanry models. Importantly, our results gave decreased
+support for Ctenophora-sister and increased support of Porifera-sister
+with the increase of number of substitutional categories (Fig. XX). More
+importantly, we are able to identify the transition of topology between
+the two alternative hypotheses in these three matrices. For example, the
+major transition occurred at 60 -- 120 categories, and the support for
+Ctenophora-sister disappears after 120 categories in the
+Whelan2017\_strict matrix. Interestingly, the results of our sensitive
+analyses further corroborated by the recent simulation study that the
+tendency of overestimate the true number of categories of Poisson-CAT or
+GTR-CAT model can potentially lead to a less accurate tree than the
+site-homogeneous models\textsuperscript{21}.
+
+The GTR-CAT model often has a better model-fitting than Poisson-CAT
+model in previous analyses, and it is also suggested a better
+performance than Poisson-CAT model. Due to computational efficiency, we
+only ran GTR-CAT and GTR-nCAT60 models in the representative data
+matrices with the most closely related choanofalgellates as outgroups.
+Interestingly, we found both Whelan2017 and Philippe matrices shifted
+from Porifera-sister with Poisson-CAT model to Ctenophora-sister using
+GTR-CAT model. Moreover, we also found all results strongly supported
+the Ctenophora-sister with GTR-nCAT60 models. Thus, these results
+suggested that the Porifera-sister can only be recovered with several
+combined factors, including inclusive of a more simplistic evolutionary
+model, an unrealistic large number of substitution categories, more
+closely related outgroups and certain gene samplings, whereas
+Ctenophora-sister is supported in most other conditions.
+
+\includegraphics{manuscript_files/figure-latex/sensitive-1.pdf}
+\textbf{Fig. XXsensitive.}
 
@@
 artifacts due to differences across species in amino acid frequencies.
-They report that posterior predictive (PP) analyses\textsuperscript{30}
+They report that posterior predictive (PP) analyses\textsuperscript{24}
 indicate 6-state recoded analyses have better model adequacy than
@@
 predictive scores do not provide strong support for Porifera-sister
-(Extended Data Fig. 4). Second, we created four new random recoding
+(Extended Data Fig. 3B). Second, we created four new random recoding
 schemes by shuffling the amino acids in the SR-6 scheme (see methods).
 When we applied each of these randomized codes to the Whelan matrix and
-analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed
-similar results as for the empirical SR-6 recoding. Like SR-6 recoding,
-random recoding increases support for Porifera-sister and improves the
-apparent adequacy of models to explain heterogeneity of states across
-taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our
-analyses show the impact of recoding is largely due to discarding
-information, not accommodating variation in amino acid composition.
-These findings indicate that it is premature to accept Porifera-sister
-and reject Ctenophora-sister using data-recoding methods and recoding
-can be a problematic method for addressing compositional variation (see
-more detailed discussion in supplementary text).
-
-\includegraphics{figures/Figure_random_rocoding.png}
-
-\textbf{Extended Data Fig. 4.} (A) The two alternative hypotheses for
-deep animal relationships considered here. Relationships that are not
-part of these hypotheses are shown as unresolved polytomies. (B) Each of
-the six plots presents one statistic, which include Posterior
-probability of Porifera-sister and the five Posterior Predictive (PP)
-statistics considered by Feuda \emph{et al.} Within each plot, there are
-six lines for six different analyses. These six analyses are the
-published SR-6 analyses presented by Feuda \emph{et al.}
-(SR6-published), analyses obtained by applying the same methods to the
-same data to to confirm that I can reproduce their published results
-(SR6-reproduced), and four analyses based on randomized recoding
-matrices obtained by shuffling the SR-6 coding scheme (random-00 --
-random-03). Each analysis includes results for 20 states (the raw amino
-acid data, shown by the left point) and for 6 states (the 6-state
-recoded data, shown by the right point). For each statistic, the results
-obtained with the random recoding are similar to those of the SR6
-recoding. This indicates that the impact of recoding is dominated by
-discarding data when collapsing from 20 states to 6 states, not
-accommodating compositional heterogeneity across lineages.
-
-\includegraphics{figures/Figure_recoding_summary.png} \textbf{Extended
-Data Fig. 6.} A graphical representation of the posterior probabilities
-for the 32 analyses presented by Feuda \emph{et al.} in their Table 3.
-Cells are color coded by whether posterior probability is \textgreater{}
-95 for Porifera-sister, \textgreater{} 95 for Ctenophora-sister, or
-neither (unresolved). Posterior predictive (PP) statistics were
-estimated for the 16 analyses in the top two rows of this figure (the
-Chang and full Whelan matrices), but not the bottom two (the Whelan
-matrices with reduced outgroup sampling).
-
-\includegraphics{figures/Figure_recoding_pp.png}
-
-\textbf{Extended Data Fig. 5.} The subset of eight CAT-GTR+G analyses
-with posterior predictive (PP) scores that is the focus of Feuda
-\emph{et al.}'s primary conclusions. These are a subset of the 32
-analyses presented in their Table 3 and graphically here in the upper
-right pane. The eight analyses are for two datasets (Chang and Whelan)
-and four coding schemes. The coding schemes are the original 20 state
-amino acid data, and three different six state recodings that group
-amino acids based on different criteria: KGB6, D6, and SR6. Cells are
-color coded as in Supplemental Fig XXRecoidng\_summary. Only one of
-these analyses, the SR6 coding of the Whelan matrix, has \textgreater{}
-95 support for Porifera-sister. The 20-state and 6-state points on the
-plots in Fig XXRecoidng\_summary of the main text correspond to the 20
-and SR6 Whelan cells shown here. The presented statistics for these
-cells are posterior probability of Porifera-sister, Maxdiff (with lower
-scores indicating better convergence of runs), and five posterior
-predictive statistics (where lower absolute value indicates better model
-adequacy). The only one of these eight analyses that provides strong
-support for Porifera-sister is not the most adequate analysis by any of
-the posterior predictive scores, and showed the poorest convergence
-according to Maxdiff.
+analyzed them under the CAT-GTR+G model with
+PhyloBayes-MPI\textsuperscript{25}, we observed similar results as for
+the empirical SR-6 recoding. Like SR-6 recoding, random recoding
+increases support for Porifera-sister and improves the apparent adequacy
+of models to explain heterogeneity of states across taxa (PP taxon
+hetero mean and max, Extended Data Fig. 5). Thus, our analyses show the
+impact of recoding is largely due to discarding information, not
+accommodating variation in amino acid composition. These findings
+indicate that it is premature to accept Porifera-sister and reject
+Ctenophora-sister using data-recoding methods and recoding can be a
+problematic method for addressing compositional variation (see more
+detailed discussion in supplementary text).
 
@@
 Resolving the root position of animal phylogeny that diverged
-\textasciitilde{} 650 million years ago has proven difficult, even with
-large genome-scale phylogenomic data matrices. By synthesizing all past
-phylogenomic studies and perform new analyses to characterize difference
-between studies, we found the support of Porifera-sister can only be
-recovered by more complex CAT models with the most closely related
-outgroups, and Ctenophora-sister in all other cases. Three factors
-contribute significantly to the previous incongruent results based on
-our analyses. First, the overlap of gene sampling and matrix is
-extremely low across different studies, leading to the difficult of
-directly compare the results across different studies. Second, the
-distribution of phylogenetic signal are largely varied between
-site-homogeneous and site-heterogeneous models. More importantly,
-site-heterogeneous models are also more sensitive to outgroup choice
-than site-homogeneous models, leading to different results even with the
-same data matrix with different outgroup choice. Third, we demonstrated
-that the support of Porifera-sister by data-recoding methods is
-potentially resulted by data reduction, not accommodation of
-compositional heterogeneity across species. Thus, the recovery of
-Porifera-sister needs more constraints than Ctenophora-sister. Rather
-than embracing one hypothesis over another, our results provide an
-integrative overview of the challenge and uncertainty related to the
-origin of animals.
-
-\hypertarget{future-direction}{%
-\subsection{Future direction}\label{future-direction}}
+\textasciitilde{} 650 million years ago has proven particularly
+difficult, even with large genome-scale phylogenomic data matrices. By
+synthesizing all past phylogenomic studies and perform new analyses to
+characterize difference between studies, we found the support of
+Porifera-sister can only be recovered by site-heterogeneous CAT models
+with the most closely related outgroups, and Ctenophora-sister in all
+other cases. Four factors contribute significantly to the previous
+incongruent results based on our analyses. First, the overlap of gene
+sampling and matrix is extremely low across different studies, leading
+to the difficult of directly compare the results across different
+studies. Second, the distribution of phylogenetic signal are largely
+varied between site-homogeneous and site-heterogeneous models. More
+importantly, site-heterogeneous models are also more sensitive to
+outgroup choice than site-homogeneous models, leading to different
+results even with the same data matrix with different outgroup choice.
+Third, we found that increase of number of substitutional categories
+coupled with a more simplistic model can change the root position of
+animal phylogeny. Last, we demonstrated that the support of
+Porifera-sister by data-recoding methods is potentially resulted by data
+reduction, not accommodation of compositional heterogeneity across
+species. Thus, the recovery of Porifera-sister needs more constraints
+than Ctenophora-sister.
 
@@
 communications and reproducibility of phylogenetic
-workflow\textsuperscript{31}. Last, we also hope that the work we have
+workflow\textsuperscript{26}. Last, we also hope that the work we have
 conducted here, including consolidating all the datasets in one place
@@
 may also provide important insight to animal origin, including
-morphological characters\textsuperscript{32,33} and fossil
-records\textsuperscript{34}. We hope the future breakthroughs that will
+morphological characters\textsuperscript{27,28} and fossil
+records\textsuperscript{29}. We hope the future breakthroughs that will
 allow us to move the field forward and a consensus viewpoint of early
@@
 
-All files and bioinformatic commands associated with this analysis are
-available at \url{https://github.com/caseywdunn/animal_root}
+All tree files, intermidiate results and scripts/commands associated
+with this study are available at
+\url{https://github.com/caseywdunn/animal_root}, doi will be provided
+upon publicaton.
 
@@
 We retrieved matrices from each publication (Table 1), storing the raw
-data in this manuscript's version control repository. We manually edited
-some minor formatting to make the batch processing of the matrices
-\emph{en masse}, e.g.~standardizing the formatting of charset blocks.
+data in this manuscript's version control repository. We manually made
+some formatting changes to make the batch processing of the matrices
+work well, \emph{e.g.} standardizing the format of Nexus CHARSET blocks.
 All changes made are tracked with git.
@@
 We programmatically queried the NCBI Taxonomy database to standardize
-names of samples in each matrix. We also use a table where manual
-entries were needed (Supplementary Table 5,
-reconciliation/taxonomy\_info/manual\_taxonomy\_map.tsv), e.g.~authors
-of original matrix indicate species name in original manuscript. For a
-table summarizing all samples and their new or lengthened names, see
-Supplementary Table 6 (reconciliation/taxonomy\_info/taxon\_table.tsv).
+names of samples in each matrix. We also used a table where manual
+entries were needed (Supplementary Table 5, manual\_taxonomy\_map.tsv),
+e.g.~authors of the original matrix indicate species name in original
+manuscript. For a table summarizing all samples and their new or
+lengthened names, see Supplementary Table 6 (taxon\_table.tsv).
 
 \hypertarget{sequence-comparisons}{%
-\paragraph{Sequence Comparisons}\label{sequence-comparisons}}
-
-Using the partition files for each matrix, we isolated each sequence for
-each taxon from each partition. Because many of the matrices had been
-processed by the original authors to remove columns that are poorly
-sampled or highly variable, these matrix-derived sequences can have
-deletions relative to the actual gene sequences.
-
-We used DIAMOND\textsuperscript{35} to compare each sequence to all
-others using default diamond Blastp parameters. We further filtered
+\paragraph{Sequence comparisons}\label{sequence-comparisons}}
+
+Using the original partition files for each matrix, we separated each
+sequence for each taxon from each partition. Because many of the
+matrices had been processed by the original authors to remove columns
+that are poorly sampled or highly variable, these matrix-derived
+sequences can have deletions relative to the actual gene sequences.
+
+We used DIAMOND v0.9.26\textsuperscript{30} to compare each sequence to
+all others using default diamond Blastp parameters. We further filtered
 DIAMOND results such that we retained hits for 90\% of partitions
-(pident \textgreater{} 78.0, eValue \textless{} 1e-15, no
-self-\textgreater{}self). We ran BUSCO with default parameters for all
-sequences against the provided metazoa gene set. We also ran a BLAST+
-blastp search against the SwissProt\textsuperscript{21} database,
-filtering such that we retain at least one hit for \textasciitilde{}97\%
-of partitions (pident \textgreater{} 50.0, eValue \textless{} 1e-15).
+(pident \textgreater{} 50.0, eValue \textless{} 1e-5, no self vs self).
+We ran BUSCO with default parameters for all sequences against the
+provided Metazoa gene set. We also ran a BLAST+
+v2.8.1\textsuperscript{31} blastp search against the
+SwissProt\textsuperscript{32} database, filtering results such that we
+retain at least one hit for \textasciitilde{}97\% of partitions (pident
+\textgreater{} 50.0, eValue \textless{} 1e-15).
 
 \hypertarget{partition-network}{%
-\paragraph{Partition Network}\label{partition-network}}
+\paragraph{Partition network}\label{partition-network}}
 
@@
 
-We constructed a network with Python and NetworkX\textsuperscript{36}
-v2.2 where each node is a partition and each edge represents a DIAMOND
-sequence-to-sequence match between sequences in the partitions. We
-extracted each connected component from this network. We further split
-these components if the the most connected node (i.e.~most edges) had
-two times more the standard deviation from the mean number of edges in
-the component it is a member of and if removing that node splits the
-component into two or more components. We then decorated every node in
-the partition network with the most often found SwissProt BLAST+ result
-and BUSCO results to see which components contain which classes and
-families of genes. See Supplementary Table 7
-{[}partition\_network\_summary table{]} for a summary tally of each part
-of the comparison.
+We constructed a network with Python and NetworkX
+v2.2\textsuperscript{33} where each node is a partition and each edge
+represents a DIAMOND sequence-to-sequence match between sequences in the
+partitions. We extracted each connected component from this network. We
+further split these components if the the most connected node (i.e.~most
+edges) had two times more the standard deviation from the mean number of
+edges in the component it is a member of and if removing that node
+splits the component into two or more components. We then decorated
+every node in the partition network with the most often found SwissProt
+BLAST+ result and BUSCO results to see which components contain which
+classes and families of genes. See Supplementary Table 7 {[}XX
+partition\_network\_summary table{]} for a summary tally of each part of
+the comparison.
 
@@
 To investigate the phylogenetic hypotheses and distribution of
-phylogenetic signal in studies aiming to elucidate the root position of
-animal phylogeny, we considered 17 data matrices from all phylogenomic
-studies that were constructed from EST, transcriptomic or genomic data
+phylogenetic signal in studies aiming to find the root position of
+animal phylogeny, we considered 16 data matrices from all phylogenomic
+studies that were constructed from EST, transcriptomic, or genomic data
 (Table 1). Because different choices of substitution models could
@@
 vs.~site-homogeneous models), we first investigated model-fit from each
-matrix using ModelFinder\textsuperscript{25} in
-IQtree\textsuperscript{22}, including site-heterogenous C10 to C60
-profile mixture models (C models) as variants of the CAT models in ML
-framework (C10-C60 model were included for model comparison via -madd
-option). We included models that are commonly used in previous analyses,
-including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C
-models in the model testing. For computational efficiency, the GTR with
-C models were not included in model testing since it requires to
-estimate over 10,000 parameters. For large matrices like Hejnol2009,
-Borrowiec2015 and Simion2017, model testing is also not computational
-feasible so only LG+C60 models were used since LG/WAG+C60 models were
-suggested as the best-fit model in all other matrices.
+matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous
+C10 to C60 profile mixture models (C60 models) as variants of the CAT
+models in ML framework (C10-C60 model were included for model comparison
+via -madd option). We included models that are commonly used in previous
+analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models
+plus C10-C60 models in the model testing. For computational efficiency,
+the GTR+C60 models were not included in model testing since it requires
+to estimate over 10,000 parameters. For large matrices like those from
+Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not
+computational feasible so only LG+C60 models were used since LG/WAG+C60
+models were suggested as the best-fit model in all other matrices.
 
@@
 Hejnol2009 and Simion2017, it was not computationally feasible to
-analyze the whole matrix using the C model or CAT site-heterogeneous
+analyze the whole matrix using the C60 model or CAT site-heterogeneous
 models. To circumvent his limitation, we reduced the data size from
@@
 site-heterogeneous models. For Hejnol2009 matrix, we instead used the
-330-gene matrix constructed by Hejnol \emph{et al}.
-2009\textsuperscript{13}, since the main conclusion for their study is
-based on this subsampled matrix; For Simion2017 matrix, we only included
-most complete 25\% of genes (genes that were present in less than 79
-taxa were removed; 428 genes were kept). It should be noted that the
-main conclusion of Simion \emph{et al}. was also based on selection of
-25\% of genes for their jackknife approach.
-
-\hypertarget{outgroup-taxa-sampling-with-c10-c60-and-cat-models}{%
-\paragraph{Outgroup taxa sampling with C10-C60 and CAT
-models}\label{outgroup-taxa-sampling-with-c10-c60-and-cat-models}}
+330-gene matrix constructed by Hejnol \emph{et al}. 2009, since the main
+conclusion for their study is based on this subsampled matrix; For
+Simion2017 matrix, we only included most complete 25\% of genes (genes
+that were present in less than 79 taxa were removed; 428 genes were
+kept). It should be noted that the main conclusion of Simion \emph{et
+al}. was also based on selection of 25\% of genes for their jackknife
+approach.
+
+\hypertarget{outgroup-taxa-sampling-with-c60-and-cat-models}{%
+\paragraph{Outgroup taxa sampling with C60 and CAT
+models}\label{outgroup-taxa-sampling-with-c60-and-cat-models}}
 
@@
 related holozoans; Opistokonta also includes Fungi. For each Choanimalia
-data matrice, both C models IQtree and Poisson-CAT models in PhyloBayes
-were conducted. The maximum likelihood analysis was performed using the
-best-fit substitution model identified as above and nodal support was
-assessed with 1000 ultrafast bootstrap replicates using IQtree.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT
-model was done with PhyloBayes MPI\textsuperscript{37}. To minimize
-computational burden, CAT-GTR models were only performed in the.
+data matrice, both C60 models in IQtree and Poisson-CAT models in
+PhyloBayes were conducted. The maximum likelihood analysis was performed
+using the best-fit substitution model identified as above and nodal
+support was assessed with 1000 ultrafast bootstrap replicates using
+IQtree. Moreover, bayesian inference with the site-heterogeneous
+poisson-CAT model was done with PhyloBayes-MPI v1.8. To minimize
+computational burden, CAT-GTR models were only performed in the
+representative Chaonimalia matrices from Philippe2009, Ryan2013\_est and
+Whelan2017\_full.
 
@@
 hypothesis that sponges are the sister group to the remaining Metazoa
-using the CAT model, bayesian inference with Poisson-CAT model was also
-conducted to Holozoa and Opisthokonta data matrices with the same
-settings as above. For all the analyses with Poisson+CAT models in
+using the Poisson-CAT model, bayesian inference with Poisson-CAT model
+was also conducted to Holozoa and Opisthokonta data matrices with the
+same settings as above. For all the analyses with Poisson+CAT models in
 PhyloBayes, two independent chains were sampled every generation. Tracer
@@
 by posterior probability. For those matrices that were not converged,
-PhyloByes analyses were run for at least one month.
+PhyloBayes analyses were run for at least two weeks. We also summarized
+the average number of substitutional categories inferred for each
+PhyloBayes analysis using Tracer.
 
 \hypertarget{phylogenetic-distribution-of-support}{%
-\paragraph{phylogenetic distribution of
+\paragraph{Phylogenetic distribution of
 support}\label{phylogenetic-distribution-of-support}}
@@
 hypothesis were calculated with IQtree output and Perl scripts from Shen
-et al.~2017\textsuperscript{28}. For BI analysis, we only considered the
-Philippe\_2009 and Whelan\_2017 datasets.
+et al.~2017\textsuperscript{23}.
+
+For BI analysis, we only considered the Philippe\_2009 and Whelan\_2017
+datasets due to computational efficiency. {[}XX Place holder for
+Xingxing{]}
+
+\hypertarget{sensitive-analyses-with-different-number-of-substitutional-categories}{%
+\paragraph{Sensitive analyses with different number of substitutional
+categories}\label{sensitive-analyses-with-different-number-of-substitutional-categories}}
+
+To explore how the number of substitutional categories may affect the
+phylogenetic inference related to the animal phylogeny, we conducted
+PhyloBayes analyses with a panel of different substitutional categories
+in the Whelan2017\_strict (ncat=60,70,80,90,110,120,150,180,360),
+Whelan2017\_full\_choanozoa (ncat=60,340,380,420,460), Philippe2009
+(ncat=60,90,120,150,180) Philippe2009\_holozoa (ncat=360,400,440,480)
+and Ryan2013\_est (ncat=60). To compare the results between Poisson-CAT
+and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60
+models were only performed in the representative Chaonimalia matrices
+from Philippe2009, Ryan2013\_est and Whelan2017\_full. All PhyloBayes
+analyses were used as the same settings as above (see Outgroup taxa
+sampling with C60 and CAT models section), with the exception of
+different numbers fo categories were used.
 
@@
 The original SR-6 recoding scheme is
-``\texttt{APST\ CW\ DEGN\ FHY\ ILMV\ KQR}''\textsuperscript{38}, where
+``\texttt{APST\ CW\ DEGN\ FHY\ ILMV\ KQR}''\textsuperscript{34}, where
 spaces separate amino acids that are placed into the same group. This
@@
 recoding used by Feuda \emph{et al.}, KGB-6 and D-6, are based on
-different matrices and methods\textsuperscript{39}.
+different matrices and methods\textsuperscript{35}.
 
@@
 between 20-state results (as in the left side of each pane of Extended
-Data Fig. 4) give insight into the technical variance of the inference
+Data Fig. 3B) give insight into the technical variance of the inference
 process.
 
-Each new matrix was analyzed with PhyloBayes-mpi. Analyses were run for
+Each new matrix was analyzed with PhyloBayes-MPI. Analyses were run for
 1000 generations, and a 200 generation burnin applied. The resulting
@@
 We thank the Yale Center for Research Computing for use of the research
-computing infrastructure, specificaly the Farnam cluster.
+computing infrastructure, specificaly the Farnam HPC cluster.
 
@@
 
-\pagebreak
-
-\hypertarget{supplemental-information}{%
-\section{Supplemental Information}\label{supplemental-information}}
-
-\hypertarget{supplementary-results}{%
-\subsection{Supplementary Results}\label{supplementary-results}}
-
-\hypertarget{models-of-molecular-evolution-1}{%
-\subsubsection{Models of molecular
-evolution}\label{models-of-molecular-evolution-1}}
-
-The exchangeability matrix \(E\) describes the rate at which one amino
-acid changes to another. Exchangeability matrices have been used in the
-studies under consideration here include: Poisson (or F81), WAG, LG,
-GTR. While the exchangeability matrix describes the relative rate of
-different changes between amino acids, the actual rate can be further
-scaled. There are a couple approaches that have been used in the studies
-considered here:
-
-\begin{itemize}
-\item
-  F81\textsuperscript{40} corresponds to equal rates between all states.
-  The F81 matrix is also sometimes referred to as the Poisson matrix. It
-  has no free parameters to estimate since all off-diagonal elements are
-  set to 1.
-\item
-  WAG\textsuperscript{41} is an empirically derived exchangeability
-  matrix based on a dataset of 182 globular protein families. It has no
-  free parameters to estimate since all off-diagonal elements are set
-  according to values estimated from this particular sample dataset.
-\item
-  LG\textsuperscript{42}, like WAG, is an empirically derived
-  exchangeability matrix. It is based on a much larger set of genes, and
-  variation in rates across sites was taken into consideration when it
-  was calculated. It has no free parameters to estimate since all
-  off-diagonal elements are set according to values estimated from this
-  particular sample dataset.
-\item
-  GTR, the General Time Reversible exchangeability matrix, has free
-  parameters for all off-diagonal elements that describe the
-  exchangeability of different amino acids. It is constrained so that
-  changes are reversible, \emph{i.e.} the rates above the diagonal are
-  the same as those below the diagonal. This leaves 190 parameters that
-  must be estimated from the data long with the other model parameters
-  and the phylogenetic tree topology. This estimation requires a
-  considerable amount of data and computational power, but if successful
-  has the advantage of being based on the dataset at hand rather than a
-  different dataset (as for LG and WAG).
-\end{itemize}
-
-While the exchangeability matrix describes the relative rate of
-different changes between amino acids, the actual rate can be further
-scaled. There are a couple approaches that have been used in the studies
-considered here:
-
-\begin{itemize}
-\item
-  Site homogeneous rates. The rates of evolution are assumed to be the
-  same at all sites in the amino acid alignment.
-\item
-  Gamma rate heterogeneity. Each site is assigned to a different rate
-  class with its own rate value. This accommodates different rates of
-  evolution across different sites. Gamma is used so commonly that
-  sometimes it isn't even specified, making it difficult at times to
-  know if a study uses Gamma or not.
-\end{itemize}
-
-The vector of equilibrium frequencies \(\Pi\) describes the stationary
-frequency of amino acids. There are a few approaches that have been used
-across the studies considered here:
-
-\begin{itemize}
-\item
-  Empirical site-homogeneous. The frequency of each amino acid is
-  observed from the matrix under consideration and applied to
-  homogeneously to all sites in the matrix.
-\item
-  Estimated site-homogeneous. The frequency of each amino acid is
-  inferred along with other model parameters, under the assumption that
-  it is the same at all sites.
-\item
-  CAT site heterogeneous\textsuperscript{3}. Each site is assigned to a
-  class with its own equilibrium frequencies. The number of classes,
-  assignment of sites to classes, and equilibrium frequencies within the
-  data are all estimated in a Bayesian framework.
-\item
-  C10 to C60\textsuperscript{24}. 10 to 60-profile mixture models as
-  variants of the CAT model under the maximum-likelihood framework.
-\item
-  Recoding method: Amino acids are recoded into six groups based on
-  function to account for both compositional heterogeneity and
-  substitution saturation. Several recoding strategies have been
-  proposed, including Dayhoff 6-state recoding, S\&R 6-state recoding,
-  KGB 6-state recoding based on exchangeability matrix of models.
-\end{itemize}
-
-\hypertarget{details-of-published-analyses}{%
-\subsubsection{Details of published
-analyses}\label{details-of-published-analyses}}
-
-\hypertarget{dunn-et-al.-2008}{%
-\paragraph{\texorpdfstring{Dunn \emph{et al.}
-2008}{Dunn et al. 2008}}\label{dunn-et-al.-2008}}
-
-Dunn \emph{et al.}\textsuperscript{2} added Expressed Sequence Tag (EST)
-data for 29 animals. It was the first phylogenomic analysis that
-included ctenophores, and therefore that could test the relationships of
-both Ctenophora and Porifera to the rest of animals. It was also the
-first phylogenetic analysis to recover Ctenophora as the sister group to
-all other animals.
-
-The data matrix was constructed using a semi-automated approach. Genes
-were translated into proteins, promiscuous domains were masked, all gene
-sequences from all species were compared to each other with blastp,
-genes were clustered based on this similarity with
-TribeMCL\textsuperscript{43}, and these clusters were filtered to remove
-those with poor taxon sampling and high rates of lineage-specific
-duplications. Gene trees were then constructed, and in clades of
-sequences all from the same species all but one sequence were removed
-(these groups are often due to assembly errors). The remaining gene
-trees with more than one sequence for any taxon were then manually
-inspected. If strongly supported deep nodes indicative of paralogy were
-found, the entire gene was discarded. If the duplications for a a small
-number of taxa were unresolved, all genes from those taxa were excluded.
-Genes were then realigned and sites were filtered with
-Gblocks\textsuperscript{44}, resulting in a 77 taxon matrix. Some taxa
-in this matrix were quite unstable, which obscured other
-strongly-supported relationships. Unstable taxa were identified with
-leaf stability indices\textsuperscript{45}, as implemented in
-phyutility\textsuperscript{46}, and removed from the matrix. This
-resulted in the 64-taxon matrix that is the focus of most of their
-analyses. Phylogenetic analyses were conducted under the Poisson+CAT
-model in phylobayes\textsuperscript{3}, and under the WAG model in
-MrBayes\textsuperscript{47} and RAxML\textsuperscript{48}.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-The placement of ctenophores (comb jellies) as the sister group to all
-other sampled metazoans is strongly supported in all our analyses. This
-result, which has not been postulated before, should be viewed as
-provisional until more data are considered from placozoans and
-additional sponges.
-\end{quote}
-
-Note that there was, in fact, an exception to strong support. An
-analysis of the 40 ribosomal proteins in the matrix recovered
-Ctenophora-sister with only 69\% support. This study did not include
-\emph{Trichoplax}.
-
-\hypertarget{philippe-et-al.-2009}{%
-\paragraph{\texorpdfstring{Philippe \emph{et al.}
-2009}{Philippe et al. 2009}}\label{philippe-et-al.-2009}}
-
-Philippe \emph{et al.}\textsuperscript{27} assembled an EST dataset for
-55 species with 128 genes to explore phylogenetic relationship of early
-diverging animals by added 9 new species. Th data matrix were assembled
-based on The phylogenetic analysis using Poisson-CAT model strongly
-supported Porifera-sister, and ctenophores was sister to cnidarians to
-form the ``coelenterate'' clade. Gene trees were then constructed, and
-potentially paralogs were removed by a bootstrap threshold of 70.
-Ambiguously aligned regions were trimemed and only genes sampled for at
-least two-thirds of species were retained. The phylogenetic analyses
-were conducted under the Poisson+CAT model in
-phylobayes\textsuperscript{3}.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-The resulting phylogeny yields two significant conclusions reviving old
-views that have been challenged in the molecular era: (1) that the
-sponges (Porifera) are monophyletic and not paraphyletic as repeatedly
-proposed, thus undermining the idea that ancestral metazoans had a
-sponge-like body plan; (2) that the most likely position for the
-ctenophores is together with the cnidarians in a ``coelenterate'' clade.
-\end{quote}
-
-\hypertarget{hejnol-et-al.2009}{%
-\paragraph{Hejnol et al.~2009}\label{hejnol-et-al.2009}}
-
-Hejnol \emph{et al.}\textsuperscript{13} added EST sequences from seven
-taxa, and a total of 94 taxa were included in the final data matrix to
-explore animal phylogeny, especially the position of acoelomorph
-flatwroms. The orthology inference was largely similar to Dunn et
-al.~2008, with the exception of orthology genes were clustered by MCL.
-The final data matrix included 1497 genes, and then subsampled with 844,
-330 and 53 gens by different thresholds of gene occupancy. With the
-exception of 53 gene matrix, maximum likelihood analyses from all other
-datasets strongly supported Ctenophora-sister (models were selected by
-RaxML perl script).
-
-\hypertarget{pick-et-al.-2010}{%
-\paragraph{\texorpdfstring{Pick \emph{et al.}
-2010}{Pick et al. 2010}}\label{pick-et-al.-2010}}
-
-Pick \emph{et al.}\textsuperscript{49} sought to test whether
-Ctenophora-sister was an artefact of insufficient taxon sampling. They
-added new and additional published sequence data to the 64-taxon matrix
-of Dunn \emph{et al.}\textsuperscript{2}. The new taxa included 12
-sponges, 1 ctenophore, 5 cnidarians, and \emph{Trichoplax}. They further
-modified the matrix by removing 2,150 sites that were poorly sampled or
-aligned. They considered two different sets of outgroups:
-Choanoflagellata (resulting in Choanimalia) and the same sampling as
-Dunn \emph{et al.} (resulting in Opisthokonta).
-
-All their analyses were conducted under the F81+CAT+Gamma model in
-phylobayes\textsuperscript{3}, in both a Bayesian framework and with
-bootstrapping. All analyses have the same ingroup sampling and site
-removal so it isn't possible to independently assess the impact of these
-factors. Analyses with Choanimalia sampling recovered Porifera-sister
-with 72\% posterior probability (PP) and 91\% bootstrap support (BS).
-With broader Opisthokonta sampling, support for Porifera-sister is 84\%
-PP. This is an interesting case where increased outgroup sampling leads
-to increased support for Porifera-sister.
-
-The authors argue that previous results supporting Ctenophora-sister
-``are artifacts stemming from insufficient taxon sampling and
-long-branch attraction (LBA)'' and that ``this hypothesis should be
-rejected''. Although the posterior probabilities supporting
-Porifera-sister are not strong, they conclude:
-
-\begin{quote}
-Results of our analyses indicate that sponges are the sister group to
-the remaining Metazoa, and Placozoa are sister group to the Bilateria
-\end{quote}
-
-They also investigated saturation, and conclude that Dunn \emph{et
-al.}\textsuperscript{2} is more saturated than Philippe \emph{et al.}
-2009 {[}Philippe:2009hh{]}. Note that the Pick \emph{et
-al.}\textsuperscript{49} dataset is not reanalyzed here because
-partition data are not available, and due to site filtering the
-partition file from Dunn \emph{et al.}\textsuperscript{2} cannot be
-applied to this matrix.
-
-\hypertarget{nosenko-et-al.-2013}{%
-\paragraph{\texorpdfstring{Nosenko \emph{et al.}
-2013}{Nosenko et al. 2013}}\label{nosenko-et-al.-2013}}
-
-Nosenko \emph{et al.}\textsuperscript{50} added Expressed Sequence Tag
-(EST) data for 9 species of non-bilaterian metazoans (7 sponges). They
-constructed a novel matrix containing 122 genes and parsed them into two
-non-overlaping matrices (ribosomal and non-ribosomal genes) and found
-incongruent results of deep metazoan phylogeny. The other major finding
-was that ribosomal gene partitions showed significantly lower saturation
-than the non-ribosomal ones.
-
-Orthologs were constructed using the bioinfomatics pipeline
-OrthoSelect\textsuperscript{51}. They also evaluated level of
-saturation, leaf stability of sampled taxa, compositional heterogeinity
-and model comparison of each matrix. By modifying gene sampling, ingroup
-and outgroup sampling, three major topologies related to the position of
-animal-root were constructed (including Porifera+Placozoa sister,
-Ctenophora-sister and Porifera-sister). Phylogenetic analyses were
-conducted under the Poisson+CAT, GTR+CAT and GTR models in
-phylobayes\textsuperscript{3}.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-we were able to reconstruct a metazoan phylogeny that is consistent with
-traditional, morphology-based views on the phylogeny of non-bilaterian
-metazoans, including monophyletic Porifera and ctenophores as a
-sister-group of cnidarians.
-\end{quote}
-
-Note the main conclusion of this paper is baed on the selection of genes
-that evolved slowly ribosomal RNA genes. The Ctenophora-sister was
-rejected based on the most saturated dataset and not supported by the
-matrix with the most closely related outgroups (Choanomalia
-matrix)\textsuperscript{50}.
-
-\hypertarget{ryan-et-al.-2013}{%
-\paragraph{\texorpdfstring{Ryan \emph{et al.}
-2013}{Ryan et al. 2013}}\label{ryan-et-al.-2013}}
-
-Ryan \emph{et al.} sequenced the first ctenophore genome of
-\emph{Mnemiopsis leidyi}. With the genome resources of \emph{M. leidyi},
-the authors constructed two phylogenomic datasets: a ``Genome set''
-based on 13 animal genomes and a ``EST Set'' that also included 59
-animals. They analyzed both matrices by site-homogeneous GTR+Gamma and
-site-heterogeneous Poisson-CAT models with three sets of outgroup
-sampling to evaluate the effect of outgroup selection to the ingroup
-topology for the Ryan2013\_est matrix. The Orthologs were constructed
-based on the method of Hejnol \emph{et al.} 2009. For the
-Ryan2013\_genome matrix, they performed phylogenetic analyses with both
-gene content and sequence-baed analyses. Overall, their results strongly
-supported Ctenophora-sister in all datasets they analyzed using
-site-homogeneous model. The Poisson+CAT model of the genome dataset
-strongly supported of a clade of Ctenophora and Porifera as the sister
-group to all other Metazoa and Bayesian analysis on the EST dataset did
-not converge after 205 days (but strongly supported Porifera in
-Choaimalia matrix).
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-Our phylogenetic analyses suggest that ctenophores are the sister group
-to the rest of the extant animals.
-\end{quote}
-
-\hypertarget{moroz-et-al.-2014}{%
-\paragraph{\texorpdfstring{Moroz \emph{et al.}
-2014}{Moroz et al. 2014}}\label{moroz-et-al.-2014}}
-
-Moroz \emph{et al.}\textsuperscript{52} sequenced the second ctenophore
-genome \emph{Pleurobrachia bachei} to explore the phylogenetic
-relationship of Metazoa. All phylogenetic analyses strongly supported
-Ctenophora-sister with different taxon and gene sampling using WAG
-site-homogeneous model. Two phylogenomic matrices were generated, the
-first set was represented by two ctenophore species, whereas the other
-set contained improved ctenophore sampling (10 taxa, Moroz2013\_3d).
-Orthology determination employed in HaMStR\textsuperscript{53} using
-1,032 ``model organism'' single-copy orthologs. Sequence were then
-trimmed and aligned. This resulted in a final matrix of 170,871 amino
-acid positions across 586 genes with 44 taxa for the first matrix, and
-114 genes with 60 taxa for the second matrix. All the phylogenetic
-analyses were analyzed in RAxML under the WAG+CAT+F models (different
-than CAT models in PhyloBayes) to reduce the computational cost.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-Our integrative analyses place Ctenophora as the earliest lineage within
-Metazoa. This hypothesis is supported by comparative analysis of
-multiple gene families, including the apparent absence of HOX genes,
-canonical microRNA machinery, and reduced immune complement in
-ctenophores.
-\end{quote}
-
-It should be noted that only the Moroz\_3d matrix has been reanalyzed in
-other studies, although the support of Ctenophora-sister is quite low.
-
-\hypertarget{borowiec-et-al.-2015}{%
-\paragraph{\texorpdfstring{Borowiec \emph{et al.}
-2015}{Borowiec et al. 2015}}\label{borowiec-et-al.-2015}}
-
-Borowiec \emph{et al.}\textsuperscript{14} assembled a genome dataset
-comprised 1080 orthologs derived from 36 publicly available genomes
-representing major lineages of animals, although only one genome of
-sponge and ctenophore was included. The orthologs were constructed using
-OrthologID pipeline\textsuperscript{54}. After removal of spurious
-sequences and genes with more than 40\% of mission data, the final
-matrix included 1080 (Total 1080). The authors further filtered the full
-dataset to 9 sub-datasets by filtering genes with high long-branch
-scores; genes with high saturation; gene occupancy; fast evolving genes.
-The main conclusion of the paper was largely based on
-BorowiecTotal\_1080 and Borowiec\_Best108 matrices. Phylogenetic
-analyses were conducted under the GTR+CAT model in PhyloBayes in
-selected matrices, and under the data-partitioning methods in RAxML for
-all matrices.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-Our phylogeny supports the still-controversial position of ctenophores
-as sister group to all other metazoans. This study also provides a
-workflow and computational tools for minimizing systematic bias in
-genome-based phylogenetic analyses.
-\end{quote}
-
-It should be noted that the authors also employed recoding-method in the
-Borowiec\_Best108 matrix and found neither support of Porifera-sister or
-Ctenophora-sister\textsuperscript{14}.
-
-\hypertarget{whelan-et-al.2015}{%
-\paragraph{Whelan et al.~2015}\label{whelan-et-al.2015}}
-
-Whelan \emph{et al.} 2015\textsuperscript{11} constructed a new
-phylogenomic dataset by eight new transcripomic data and investigated a
-range of possible sources of systematic error under multiple analyses
-(e.g.~long-branch attraction, compositional bias, fast evolving genes,
-etc.). Putative orthologs were determined of each species using HaMStR
-using the model organism core ortholog set (same as Moroz \emph{et al.}
-2014) and subsequently removal of genes with too much missing data and
-potential paralogs. The authors further filtered the full dataset to 24
-sub-datasets by filtering genes with high long-branch scores; genes with
-high RSFV values; genes that are potential paralogs; fast evolving genes
-and progressively removal of outgroups. All the maximum likelihood
-analyses with site-homogeneous model and PartitionFinder strongly
-suggested Ctenophora-sister. CAT-GTR models only used in least saturated
-dataset 6 and 16 also strongly supported Ctenophora.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-Importantly, biases resulting from elevated compositional heterogeneity
-or elevated substitution rates are ruled out. Placement of ctenophores
-as sister to all other animals, and sponge monophyly, are strongly
-supported under multiple analyses, herein.
-\end{quote}
-
-Note that the authors also reanalyzed Philippe2009 matrix (with the
-removal of ribosomal genes) and recovered Porifera-sister with moderate
-support (pp=90).
-
-\hypertarget{chang-et-al.2015}{%
-\paragraph{Chang et al.~2015}\label{chang-et-al.2015}}
-
-Chang \emph{et al.}\textsuperscript{55} was originally used to explore
-phylogenetic position of Myxozoa in Cnidaria but also sampled broadly
-across the breadth of animal diversity. The authors constructed a
-dataset with 200 protein markers based on Philippe et
-al.~2011\textsuperscript{7} with 51,940 amino acids and 77 taxa. Both
-site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly
-supported Ctenophora-sister.
-
-Note that this data matrix has been extensively reanalyzed in many
-studies.
-
-\hypertarget{pisani-et-al.-2015}{%
-\paragraph{\texorpdfstring{Pisani \emph{et al.}
-2015}{Pisani et al. 2015}}\label{pisani-et-al.-2015}}
-
-Pisani \emph{et al.}\textsuperscript{6} reanalyzed representative
-datasets that supported Ctenophora-sister, including Ryan2013\_est,
-Moroz2014\_3d and Whelan2015 datasets. It was the first study showing
-that progressively removal of more distantly related outgroups could
-largely affect phylogenomic inference of the position of the root of
-animal phylogeny. The authors suggested that the inclusion of outgroups
-very distant from the ingroup can cause systematic errors due to
-long-branch attraction. Phylogenetic analyses were conducted under the
-Poisson+CAT and GTR models in phylobayes. They found GTR-CAT and
-Poisson-CAT models generally had better model-fit than site-homogeneous
-GTR models in these data matrice. Moreover, they found the support of
-Ctenophora-sister decreases when the exclusion of distantly related
-outgroups and the use of site-heterogeneous CAT models.
-
-Regarding the recovery of Porifera-sister, the authors concluded:
-
-\begin{quote}
-Our results reinforce a traditional scenario for the evolution of
-complexity in animals, and indicate that inferences about the evolution
-of Metazoa based on the Ctenophora-sister hypothesis are not supported
-by the currently available data.
-\end{quote}
-
-\hypertarget{feuda-et-al.-2017}{%
-\paragraph{\texorpdfstring{Feuda \emph{et al.}
-2017}{Feuda et al. 2017}}\label{feuda-et-al.-2017}}
-
-Feuda \emph{et al.}\textsuperscript{9} didn't generate any new data,
-instead they used the data-recoding methods to reanalyze two key
-datasets that support Ctenophora-sister (Whelan2015\_D20, Chang2015
-datasets). It was the first phylogenomic study that suggested recoding
-methods have better performance than non-recoding methods in the
-relationships at the root of the animal tree. The authors compared model
-adequacy using posterior predictive analyses from a set of
-site-homogeneous (WAG, LG, GTR, data-partitioning) and
-site-heterogeneous (CAT-GTR) models in non-recoding and recoding
-datasets. The results showed that data-recoding can significant reduce
-compositional heterogeneity in both datasets with CAT-GTR models and
-strongly supported Porifera-sister hypothesis (see more details in
-recoding section in supplementary text).
-
-Regarding the recovery of Porifera-sister, the authors concluded:
-
-\begin{quote}
-Because adequate modeling of the evolutionary process that generated the
-data is fundamental to recovering an accurate phylogeny, our results
-strongly support sponges as the sister group of all other animals and
-provide further evidence that Ctenophora-sister represents a tree
-reconstruction artifact.
-\end{quote}
-
-\hypertarget{whelan-and-halanych-2016}{%
-\paragraph{Whelan and Halanych 2016}\label{whelan-and-halanych-2016}}
-
-Whelan \emph{et al.}\textsuperscript{26} is the only study to evaluate
-performance of site-heterogeneous models and site-homogeneous model with
-data partitioning under the simulation framework. The simulation results
-suggested that Poisson+CAT model consistently performed worse than other
-models in simulation datasets. More importantly, the authors also showed
-that both Poisson+CAT and GTR+ CAT models could overestimated
-substitutional heterogeneity in many cases. They also reanalyzed
-datasets from Philippe 2009 and Nosenko 2013 using both CAT models and
-data partitioning with site-homogeneous model. The results indicated
-that Poisson + CAT model tends to recover less accurate trees and more
-importantly, both GTR + CAT and data partitioning strongly supported
-Ctenophora-sister in reanalyses.
-
-The authors concluded:
-
-\begin{quote}
-Practices such as removing constant sites and parsimony uninformative
-characters, or using CAT-F81 when CAT-GTR is deemed too computationally
-expensive, cannot be logically justified. Given clear problems with
-CAT-F81, phylogenies previously inferred with this model should be
-reassessed.
-\end{quote}
-
-\hypertarget{whelan-et-al.-2017}{%
-\paragraph{\texorpdfstring{Whelan \emph{et al.}
-2017}{Whelan et al. 2017}}\label{whelan-et-al.-2017}}
-
-Whelan \emph{et al.}\textsuperscript{15} added 27 new ctenophore
-transcriptomic data to explore animal-root position as well as
-relationships within Ctenophores. It significantly increased ctenophore
-taxon sampling than other studies. Putative orthologs were determined
-largely similar to Whelan2015, with the difference of a core Ctenophora
-core datasets were constructed here with more than 2000 genes. The
-subsequent filtering strategy was also similar to the previous study.
-All analyses using site-homogeneous and site-heterogeneous models
-strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model
-in choanoazoa dataset. The main conclusions of this paper were based on
-Whelan2017\_full and Whelan2017\_strict matrices.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-\textgreater{}Using datasets with reasonably high ctenophore and other
-non-bilaterian taxon sampling, our results strongly reject the
-hypothesis that sponges are the sister lineage to all other extant
-metazoans.
-
-\hypertarget{simion-et-al.-2017}{%
-\paragraph{\texorpdfstring{Simion \emph{et al.}
-2017}{Simion et al. 2017}}\label{simion-et-al.-2017}}
-
-Simion \emph{et al.}\textsuperscript{16} added transcriptomic data for
-21 new animals. The data matrix was constructed using a semi-automated
-approach to comprehensively detect and eliminate potential systematic
-errors. The resulting dataset comprises 1,719 genes and 97 species,
-including 61 non-bilaterian species. It was by far the largest
-phylogenomic dataset in terms of taxon and gene sampling related to the
-relationship at the root of animal phylogeny.
-
-The final matrix was first analyzed using the Poisson-CAT model.
-Different that other phylobayes analyses, Simion \emph{et al.} used a
-gene jackknife strategy based on 100 analyses to overcome the
-computational limitation because of the large data size. Each jackknife
-is baed on a random selection of \textasciitilde{} 25\% of the genes.
-The Phylobayes with site-heterogeneous model strongly supported the
-Porifera-sister, whereas site-homogeneous strongly supported
-Ctenophora-sister in all datasets. Importantly, the authors compared the
-behavior of long-branch effect between site-heterogeneous and
-site-homogeneous models by progressively removal taxa and concluded
-higher sensitivity of site-homogeneous models to LBA than CAT models.
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-\begin{quote}
-Our dataset outperforms previous metazoan gene superalignments in terms
-of data quality and quantity. Analyses with a best-fitting
-site-heterogeneous evolutionary model provide strong statistical support
-for placing sponges as the sister-group to all other metazoans, with
-ctenophores emerging as the second-earliest branching animal lineage.
-\end{quote}
-
-It should be noted that all the PhyloBayes runs have not reached
-convergence due to the computational cost in these large matrices.
-
-\hypertarget{data-recoding-methods}{%
-\subsubsection{Data-recoding methods}\label{data-recoding-methods}}
-
-Feuda \emph{et al.}\textsuperscript{39} tackled a difficult phylogenetic
-problem -- placing the root of the animal phylogeny\textsuperscript{8}.
-In the past decade, some analyses have placed Ctenophora (comb jellies)
-and others Porifera (sponges) as the sister group to all other animals
-(Fig. Extended Data Fig. 4). Feuda \emph{et al.}\textsuperscript{39}
-present new phylogenetic analyses that they claim provide strong support
-for Porifera-sister. Their analyses consider datasets from studies by
-Chang \emph{et al.}\textsuperscript{55} and Whelan \emph{et
-al.}\textsuperscript{56}, both of which found support for
-Ctenophora-sister. Feuda \emph{et al.} were concerned that these
-Ctenophora-sister results were artefacts of lineage-specific differences
-in amino acid frequencies. In an attempt to reduce these differences,
-they recoded the full set of twenty amino acids into six groups of amino
-acids. These groups have more frequent evolutionary changes within them
-than between them, based on empirical observations in large protein
-datasets\textsuperscript{38}. The intent is to discard many
-lineage-specific changes, which are expected to fall within these
-groups. Rather than model compositional heterogeneity, as their title
-suggests, this approach discards heterogeneous information so that much
-simpler models with fewer states can be applied. They report that
-posterior predictive (PP) analyses\textsuperscript{30} indicate 6-state
-recoded analyses have better model adequacy than 20-state amino acid
-analyses, and ``Porifera-sister was favored under all recoding
-strategies''. Here we focus on two aspects of Feuda \emph{et al.} First,
-we point out that many of their recoded analyses are actually unresolved
-(\emph{i.e.}, without strong support for either Porifera-sister or
-Ctenophora-sister), and that the analyses with the best posterior
-predictive scores do not provide strong support for Porifera-sister.
-Second, we present new analyses that show the impact of recoding is
-largely due to discarding information, not accommodating variation in
-amino acid composition. These findings indicate that it is premature to
-accept Porifera-sister and reject Ctenophora-sister. They also show that
-recoding can be a problematic method for addressing compositional
-variation.
-
-Feuda \emph{et al.} examine support for Ctenophora-sister and
-Porifera-sister under all combinations of two models of molecular
-evolution, four datasets, and four coding schemes. This provides 32
-analyses that they report in their Table 3 and that we present
-graphically here as Extended Data Fig. 6. There is striking variation in
-support for Ctenophora-sister and Porifera-sister across these analyses
-(Extended Data Fig. 6). Feuda \emph{et al.} accept the results of some
-analyses and reject others based on posterior predictive (PP) analyses
-of model adequacy, which assess how well a model explains variation in
-the data\textsuperscript{30}. They considered five different posterior
-predictive statistics that capture different types of variation in the
-data. From this they conclude that their ``results strongly support
-sponges as the sister group of all other animals''.
-
-This conclusion does not follow from their own presented results. Only a
-single analysis with posterior predictive scores provides what could be
-considered strong support \textgreater{} 95 posterior probability) for
-Porifera-sister. Of the 32 analyses, posterior predictive scores were
-calculated for 16 (those for the full Whelan and Chang matrices). Based
-on posterior predictive scores, Feuda \emph{et al.} reject eight of
-these that were conducted under the GTR+G model (which all have strong
-support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses
-(Extended Data Fig. 5). Two of these eight are analyses of the original
-20-state amino acid data, both of which provide strong support for
-Ctenophora-sister. Of the six recoded analyses, five are unresolved.
-Only a single analysis for which posterior predictive scores are
-available provides strong support for Porifera-sister, the CAT-GTR+G
-analysis of the SR-6\textsuperscript{38} recoded
-Whelan\textsuperscript{56} matrix. Furthermore, this analysis does not
-have the best score according to any of the five posterior predictive
-statistics they considered (Extended Data Fig. 5). The only statistic
-that stands out for this one analysis is that it has the highest maxdiff
-(Extended Data Fig. 5), indicating that it did not converge as well as
-other analyses.
-
-Though their study does not provide strong support for Porifera-sister,
-the sensitivity of their results to recoding provides an opportunity to
-better understand and evaluate the impact of recoding more generally.
-This is important given the growing interest in
-recoding\textsuperscript{39}. Feuda \emph{et al.} hoped recoding would
-reduce potential artefacts due to differences across species in amino
-acid frequencies. They interpreted the fact that their analyses are
-sensitive to recoding as evidence that such an artefact exists and that
-they successfully addressed it by recoding. An alternative hypothesis is
-that recoding impacts phylogenetic analyses because it discards so much
-information. These two hypotheses can be tested by applying new recoding
-schemes that also reduce twenty states down to six, but are based on
-random grouping rather than empirical frequencies of amino acid
-exchange. Empirical and random recodings both discard the same amount of
-information, but only empirical recoding reduce the impact of amino-acid
-frequency as intended. Different results between empirical and random
-recoding would be consistent with the hypothesis that the empirical
-approach works as intended to accommodate compositional heterogeneity.
-Similar results would suggest that the impact of recoding is due to
-discarding information. Here we focus on their single analysis with a
-posterior predictive score that supports Porifera-sister, the CAT-GTR+G
-analysis of the SR-6 recoded Whelan data. we created four new random
-recoding schemes by shuffling the amino acids in the SR-6 scheme (see
-Supplemental Methods and analysis code at
-\url{https://github.com/caseywdunn/feuda_2017}). When we applied each of
-these randomized codes to the Whelan matrix and analyzed them under the
-CAT-GTR+G model with phylobayes-mpi\textsuperscript{37}, we observed
-similar results as for the empirical SR-6 recoding. Like SR-6 recoding,
-random recoding increases support for Porifera-sister and improves the
-apparent adequacy of models to explain heterogeneity of states across
-taxa (PP taxon hetero mean and max, Extended Data Fig. 4).
-
-These analyses suggest that the major impact of recoding on phylogenetic
-analyses is data reduction, not accommodation of compositional
-heterogeneity across species. This indicates that recoding may not be an
-effective tool for accommodating among-species differences in amino acid
-frequencies. Compositional heterogeneity would be better addressed with
-models of molecular evolution that explicitly describe such
-differences\textsuperscript{57}, if progress can be made on the
-considerable computational challenges of such complex models.
-
-\hypertarget{supplementary-methods}{%
-\subsection{Supplementary Methods}\label{supplementary-methods}}
-
-\hypertarget{matrix-mapping}{%
-\subsubsection{Matrix mapping}\label{matrix-mapping}}
-
-Taxa and partition correspondence across manuscripts was assessed by
-comparing all sequences for each taxon in each partition across all
-matrices with diamond blast. Based on inspection of sequence similarity,
-we excluded all comparisons with less than 99\% identity and greater
-than 10\textsuperscript{-25} e-value.
-
-\hypertarget{taxa-comparison-across-matrices}{%
-\subsubsection{Taxa comparison across
-matrices}\label{taxa-comparison-across-matrices}}
-
-The primary intent of comparing taxa across matrices was to validate our
-taxon name reconciliation across studies.
-
-We first considered pairwise similarity between the same species from
-different matrices in different studies.
-
-\hypertarget{partition-comparison-across-matrices}{%
-\subsubsection{Partition comparison across
-matrices}\label{partition-comparison-across-matrices}}
-
-The count for a partition pair can be much larger than the number of
-genes in the matrix, which suggests that the count is the number of hsps
-rather than the number of sequences with hits.
-
-There are 17 matrices. A gene that is perfectly sampled would form a
-cluster with this size. To our surprise, very few clusters, though, are
-this size. This suggests that intersection of genes between matrices is
-extreme low. This finding suggested genes used for exploring the
-position of the root of the animal phylogeny
-
-\hypertarget{supplementary-tables}{%
-\subsection{Supplementary Tables}\label{supplementary-tables}}
+\hypertarget{extended-tables}{%
+\subsection{Extended Tables}\label{extended-tables}}
+
+\begin{longtable}[]{@{}lllrrrllr@{}}
+\caption{Table 1. Overview of data matrices used in this
+study}\tabularnewline
+\toprule
+Study & Matrix & Clade & Number of gene partitions & Number of sites &
+Number of taxa & Infered sister lineage & Journal &
+Citations\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Study & Matrix & Clade & Number of gene partitions & Number of sites &
+Number of taxa & Infered sister lineage & Journal &
+Citations\tabularnewline
+\midrule
+\endhead
+Dunn2008 & Dunn2008 & Opisthokonta & 150 & 21152 & 64 &
+Ctenophora-sister & Nature & 1794\tabularnewline
+Philippe2009 & Philippe2009 & Opisthokonta & 129 & 30257 & 55 &
+Porifera-sister & Current Biology & 649\tabularnewline
+Hejnol2009 & Hejnol2009 & Opisthokonta & 1487 & 270580 & 94 &
+Ctenophora-sister & Proc. Biol. Sci. & 707\tabularnewline
+NA & Hejnol2009\_subsampled & Opisthokonta & 844 & 55594 & 94 &
+Ctenophora-sister & NA & NA\tabularnewline
+Nosenko2013 & Nosenko2013\_nonribo\_9187 & Opisthokonta & 35 & 9187 & 63
+& Ctenophora-sister & Molecular Phylogenetics and Evolution &
+221\tabularnewline
+NA & Nosenko2013\_ribo\_14615 & Opisthokonta & 87 & 14615 & 71 &
+Porifera-sister & NA & NA\tabularnewline
+Ryan2013 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 61 &
+Ctenophora-sister & Science & 534\tabularnewline
+Moroz2014 & Moroz2014\_3d & Opisthokonta & 114 & 22772 & 46 &
+Ctenophora-sister & Nature & 491\tabularnewline
+Whelan2015 & Whelan2015\_D1 & Opisthokonta & 251 & 59733 & 76 &
+Ctenophora-sister & PNAS & 220\tabularnewline
+NA & Whelan2015\_D10 & Opisthokonta & 210 & 59733 & 70 &
+Ctenophora-sister & NA & NA\tabularnewline
+NA & Whelan2015\_D20 & Opisthokonta & 178 & 81008 & 65 &
+Ctenophora-sister & NA & NA\tabularnewline
+Borowiec2015 & Borowiec2015\_Best108 & Choaimalia & 108 & 41808 & 36 &
+Ctenophora-sister & BMC Genomics & 104\tabularnewline
+NA & Borowiec2015\_Total1080 & Choaimalia & 1080 & 384981 & 36 &
+Ctenophora-sister & NA & NA\tabularnewline
+Pisani2015 & Ryan2013\_est & Opisthokonta & 406 & 88384 & 60 &
+Porifera-sister & PNAS & 206\tabularnewline
+NA & Whelan\_D6 & Opisthokonta & 115 & 33403 & 70 & Porifera-sister & NA
+& NA\tabularnewline
+Chang2015 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
+Ctenophora-sister & PNAS & 119\tabularnewline
+Simion2017 & Simion2017 & Opisthokonta & 1719 & 401632 & 97 &
+Porifera-sister & Current Biology & 220\tabularnewline
+NA & Simion2017\_subsampled & Opisthokonta & 1719 & 110602 & 97 &
+Porifera-sister & NA & NA\tabularnewline
+Feuda2017 & Chang2015 & Opisthokonta & 170 & 51940 & 77 &
+Ctenophora-sister & Current Biology & 119\tabularnewline
+NA & WhelanD20 & Opisthokonta & 178 & 81008 & 65 & Porifera-sister & NA
+& NA\tabularnewline
+Whelan2017 & Whelan2017b\_full & Holozoa & 212 & 68062 & 80 &
+Ctenophora-sister & Nature Ecology and Evolution & 79\tabularnewline
+NA & Whelan2017b\_strict & Choaimalia & 117 & 49388 & 76 &
+Ctenophora-sister & NA & NA\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 
+\textbf{Extended Data Table 1.} The models selected by ModelFinder for
+each matrix.
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+matrix & clade & result & model\_summary\tabularnewline
+\midrule
+\endhead
+Borowiec2015\_Best108 & Choanimalia & Ctenophora-sister &
+WAG+C60\tabularnewline
+Chang2015 & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
+Dunn2008 & Opisthokonta & Ctenophora-sister & WAG+C60\tabularnewline
+Moroz2014\_3d & Choanimalia & Unresolved & WAG+C60\tabularnewline
+Nosenko2013\_nonribo\_9187 & Opisthokonta & Unresolved &
+WAG+C60\tabularnewline
+Nosenko2013\_ribo\_11057 & Choanimalia & Unresolved &
+WAG+C60\tabularnewline
+Nosenko2013\_ribo\_14615 & Opisthokonta & Unresolved &
+WAG+C60\tabularnewline
+Philippe2009 & Opisthokonta & Ctenophora-sister & WAG+C60\tabularnewline
+Ryan2013\_est & Opisthokonta & Ctenophora-sister &
+WAG+C60\tabularnewline
+Whelan2015\_D1 & Opisthokonta & Ctenophora-sister &
+LG+C60\tabularnewline
+Whelan2015\_D10 & Opisthokonta & Ctenophora-sister &
+WAG+C60\tabularnewline
+Whelan2015\_D20 & Opisthokonta & Ctenophora-sister &
+WAG+C60\tabularnewline
+Whelan2017\_full & Holozoa & Ctenophora-sister & LG+C60\tabularnewline
+Whelan2017\_strict & Choanimalia & Ctenophora-sister &
+LG+C60\tabularnewline
+\bottomrule
+\end{longtable}
+
+\hypertarget{extended-figures}{%
+\subsection{Extended Figures}\label{extended-figures}}
+
+\begin{verbatim}
+## <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
+##     aspect: function
+##     backtransform_range: function
+##     clip: on
+##     default: FALSE
+##     distance: function
+##     expand: TRUE
+##     is_free: function
+##     is_linear: function
+##     labels: function
+##     limits: list
+##     modify_scales: function
+##     range: function
+##     ratio: 1
+##     render_axis_h: function
+##     render_axis_v: function
+##     render_bg: function
+##     render_fg: function
+##     setup_data: function
+##     setup_layout: function
+##     setup_panel_params: function
+##     setup_params: function
+##     transform: function
+##     super:  <ggproto object: Class CoordFixed, CoordCartesian, Coord, gg>
+\end{verbatim}
+
+\includegraphics{manuscript_files/figure-latex/figure_gene_composition,-1.pdf}
+
+\textbf{Extended Data Fig. XXfigure\_gene\_composition.} Annotation and
+representation of BUSCO and ribosomal genes in each data matrix. (A).
+The number of partitions with BUSCO annotations in each matrix, relative
+to the number of partitions. (B). The percentage BUSCO annotations in
+each matrix. (C). The number of partitions with annotations of ribosomal
+genes in each matrix, relative to the number of partitions. (D). The
+percentage ribosomal genes in each matrix.
+
+\includegraphics{manuscript_files/figure-latex/figure_matrix_overlap-1.pdf}
+
+\textbf{Extended Data Fig. XXfigure\_matrix\_overlap.} Pairwise overlap
+between each of the primary matrices considered here. Horizontal size is
+proportional to the number of genes sampled, vertical size to the number
+of taxa sampled. The horizontal intersection shows the proportions of
+shared genes, the vertical intersection shows the proportions of shared
+taxa.
+
+\includegraphics{figures/Figure_Pairwise_heatmap_animal_root.png}
+\textbf{Extended Data Fig. XXpairwise\_heatmap.} Pairwise intersections
+of the level of congruence of gene partitions that are supported in each
+hypothesis across different analyses in Whelan2017\_full matrices
+calculated by Euclidean distance. (A) Pairwise comparisons across genes
+that supported Ctenophora-sister in different analyses. (B) Pairwise
+comparisons across genes that supported Porifera-sister in different
+analyses. The color scale indicates the Pearson correlation, where 1 is
+positive linear correlation, 0 is no linear correlation, and −1 is
+negative linear correlation. The pairwise intersection heatmap is
+generated by Intervene\textsuperscript{36}.
+
+\includegraphics{manuscript_files/figure-latex/figure_phylogenetic_signal-1.pdf}
+\textbf{Extended Data Fig. XXphylogetic signal} The distribution of
+phylogenetic signal for two alternative topological hypotheses on the
+earliest-branching animal lineage with different models and outgroup
+choice. (A). Proportion of genes supporting each of two alternative
+hypotheses for each of two alternative hypotheses for Philippe2009
+matrix. (B). Proportion of genes supporting each of two alternative
+hypotheses for each of two alternative hypotheses for Ryan2013\_est
+matrix. (C) Proportion of genes supporting each of two alternative
+hypotheses for each of two alternative hypotheses for Whelan2017\_full
+matrix. The (\Delta\$GLS) values for the genes across each data matrix
+are provided in Supplementary Table 4.
+
+\includegraphics{figures/Figure_random_rocoding.png}
+
+\textbf{Extended Data Fig. XXrandom\_recoding.} (A) The two alternative
+hypotheses for deep animal relationships considered here. Relationships
+that are not part of these hypotheses are shown as unresolved
+polytomies. (B) Each of the six plots presents one statistic, which
+include Posterior probability of Porifera-sister and the five Posterior
+Predictive (PP) statistics considered by Feuda \emph{et al.} Within each
+plot, there are six lines for six different analyses. These six analyses
+are the published SR-6 analyses presented by Feuda \emph{et al.}
+(SR6-published), analyses obtained by applying the same methods to the
+same data to to confirm that I can reproduce their published results
+(SR6-reproduced), and four analyses based on randomized recoding
+matrices obtained by shuffling the SR-6 coding scheme (random-00 --
+random-03). Each analysis includes results for 20 states (the raw amino
+acid data, shown by the left point) and for 6 states (the 6-state
+recoded data, shown by the right point). For each statistic, the results
+obtained with the random recoding are similar to those of the SR6
+recoding. This indicates that the impact of recoding is dominated by
+discarding data when collapsing from 20 states to 6 states, not
+accommodating compositional heterogeneity across lineages.
+
+\includegraphics{figures/Figure_recoding_pp.png}
+
+\textbf{Extended Data Fig. XXrecoding\_pp.} The subset of eight
+CAT-GTR+G analyses with posterior predictive (PP) scores that is the
+focus of Feuda \emph{et al.}'s primary conclusions. These are a subset
+of the 32 analyses presented in their Table 3 and graphically here in
+the upper right pane. The eight analyses are for two datasets (Chang and
+Whelan) and four coding schemes. The coding schemes are the original 20
+state amino acid data, and three different six state recodings that
+group amino acids based on different criteria: KGB6, D6, and SR6. Cells
+are color coded as in Supplemental Fig XXRecoidng\_summary. Only one of
+these analyses, the SR6 coding of the Whelan matrix, has \textgreater{}
+95 support for Porifera-sister. The 20-state and 6-state points on the
+plots in Fig XXRecoidng\_summary of the main text correspond to the 20
+and SR6 Whelan cells shown here. The presented statistics for these
+cells are posterior probability of Porifera-sister, Maxdiff (with lower
+scores indicating better convergence of runs), and five posterior
+predictive statistics (where lower absolute value indicates better model
+adequacy). The only one of these eight analyses that provides strong
+support for Porifera-sister is not the most adequate analysis by any of
+the posterior predictive scores, and showed the poorest convergence
+according to Maxdiff.
+
+\includegraphics{figures/Figure_recoding_summary.png} \textbf{Extended
+Data Fig. XXrecoding\_summary.} A graphical representation of the
+posterior probabilities for the 32 analyses presented by Feuda \emph{et
+al.} in their Table 3. Cells are color coded by whether posterior
+probability is \textgreater{} 95 for Porifera-sister, \textgreater{} 95
+for Ctenophora-sister, or neither (unresolved). Posterior predictive
+(PP) statistics were estimated for the 16 analyses in the top two rows
+of this figure (the Chang and full Whelan matrices), but not the bottom
+two (the Whelan matrices with reduced outgroup sampling).
+
+\pagebreak
+
+\hypertarget{supplementary-information}{%
+\section{Supplementary Information}\label{supplementary-information}}
+
+\hypertarget{supplementary-results}{%
+\subsection{Supplementary Results}\label{supplementary-results}}
+
+\hypertarget{models-of-molecular-evolution-1}{%
+\subsubsection{Models of molecular
+evolution}\label{models-of-molecular-evolution-1}}
+
+Models of molecular evolution are assembled by selecting different
+options for all these different components. The models that are applied
+in practice are heavily influenced by computational costs, as well as
+convention. For example, on the questions considered here, CAT
+site-heterogeneous models of equilibrium frequency have only been used
+in combination with Poisson and GTR exchangeability matrices. LG and WAG
+exchangeability matrices have only been used with site homogeneous
+estimates of equilibrium frequency. This is further confused by the
+abbreviations that are used for models. Papers often discuss CAT and WAG
+models as if they are exclusive, but these particular terms apply to
+non-exclusive model components-- CAT refers to equilibrium frequency
+variation across sites and WAG a particular exchangeability matrix. In
+this literature, CAT is generally shorthand for Poisson+CAT and WAG is
+shorthand for WAG+homogeneous equilibrium frequency estimation. One
+could, though, run a WAG+CAT model, although the computation time is
+much longer than Poisson+CAT model. To avoid confusion on this point, we
+always specify the exchangeability matrix first, followed by modifiers
+that describe the accommodation of heterogeneity in equilibrium
+frequencies (\emph{e.g.}, CAT). Moreover, the Gamma-rate heterogeinity
+(G4) is used in almost every analysis conducted here. If there are no
+modifiers, then it is implied that site homogeneous models are used.
+
+The exchangeability matrix \(E\) describes the rate at which one amino
+acid changes to another. Exchangeability matrices have been used in the
+studies under consideration here include: Poisson (or F81), WAG, LG,
+GTR. While the exchangeability matrix describes the relative rate of
+different changes between amino acids, the actual rate can be further
+scaled. There are a couple approaches that have been used in the studies
+considered here:
+
+\begin{itemize}
+\item
+  F81\textsuperscript{37} corresponds to equal rates between all states.
+  The F81 matrix is also sometimes referred to as the Poisson matrix. It
+  has no free parameters to estimate since all off-diagonal elements are
+  set to 1.
+\item
+  WAG\textsuperscript{38} is an empirically derived exchangeability
+  matrix based on a dataset of 182 globular protein families. It has no
+  free parameters to estimate since all off-diagonal elements are set
+  according to values estimated from this particular sample dataset.
+\item
+  LG\textsuperscript{39}, like WAG, is an empirically derived
+  exchangeability matrix. It is based on a much larger set of genes, and
+  variation in rates across sites was taken into consideration when it
+  was calculated. It has no free parameters to estimate since all
+  off-diagonal elements are set according to values estimated from this
+  particular sample dataset.
+\item
+  GTR, the General Time Reversible exchangeability matrix, has free
+  parameters for all off-diagonal elements that describe the
+  exchangeability of different amino acids. It is constrained so that
+  changes are reversible, \emph{i.e.} the rates above the diagonal are
+  the same as those below the diagonal. This leaves 190 parameters that
+  must be estimated from the data long with the other model parameters
+  and the phylogenetic tree topology. This estimation requires a
+  considerable amount of data and computational power, but if successful
+  has the advantage of being based on the dataset at hand rather than a
+  different dataset (as for LG and WAG).
+\end{itemize}
+
+While the exchangeability matrix describes the relative rate of
+different changes between amino acids, the actual rate can be further
+scaled. There are a couple approaches that have been used in the studies
+considered here:
+
+\begin{itemize}
+\item
+  Site homogeneous rates. The rates of evolution are assumed to be the
+  same at all sites in the amino acid alignment.
+\item
+  Gamma rate heterogeneity. Each site is assigned to a different rate
+  class with its own rate value. This accommodates different rates of
+  evolution across different sites. Gamma is used so commonly that
+  sometimes it isn't even specified, making it difficult at times to
+  know if a study uses Gamma or not.
+\end{itemize}
+
+The vector of equilibrium frequencies \(\Pi\) describes the stationary
+frequency of amino acids. There are a few approaches that have been used
+across the studies considered here:
+
+\begin{itemize}
+\item
+  Empirical site-homogeneous. The frequency of each amino acid is
+  observed from the matrix under consideration and applied to
+  homogeneously to all sites in the matrix.
+\item
+  Estimated site-homogeneous. The frequency of each amino acid is
+  inferred along with other model parameters, under the assumption that
+  it is the same at all sites.
+\item
+  CAT site heterogeneous. Each site is assigned to a class with its own
+  equilibrium frequencies. The number of classes, assignment of sites to
+  classes, and equilibrium frequencies within the data are all estimated
+  in a Bayesian framework.
+\item
+  C10 to C60\textsuperscript{19}. 10 to 60-profile mixture models as
+  variants of the CAT model under the maximum-likelihood framework.
+\item
+  Recoding method: Amino acids are recoded into six groups based on
+  function to account for both compositional heterogeneity and
+  substitution saturation. Several recoding strategies have been
+  proposed, including Dayhoff 6-state recoding, S\&R 6-state recoding,
+  KGB 6-state recoding based on exchangeability matrix of models.
+\end{itemize}
+
+\hypertarget{details-of-published-analyses}{%
+\subsubsection{Details of published
+analyses}\label{details-of-published-analyses}}
+
+\hypertarget{dunn-et-al.-2008}{%
+\paragraph{\texorpdfstring{Dunn \emph{et al.}
+2008}{Dunn et al. 2008}}\label{dunn-et-al.-2008}}
+
+Dunn \emph{et al.}\textsuperscript{2} added Expressed Sequence Tag (EST)
+data for 29 animals. It was the first phylogenomic analysis that
+included ctenophores, and therefore that could test the relationships of
+both Ctenophora and Porifera to the rest of animals. It was also the
+first phylogenetic analysis to recover Ctenophora as the sister group to
+all other animals.
+
+The data matrix was constructed using a semi-automated approach. Genes
+were translated into proteins, promiscuous domains were masked, all gene
+sequences from all species were compared to each other with blastp,
+genes were clustered based on this similarity with
+TribeMCL\textsuperscript{40}, and these clusters were filtered to remove
+those with poor taxon sampling and high rates of lineage-specific
+duplications. Gene trees were then constructed, and in clades of
+sequences all from the same species all but one sequence were removed
+(these groups are often due to assembly errors). The remaining gene
+trees with more than one sequence for any taxon were then manually
+inspected. If strongly supported deep nodes indicative of paralogy were
+found, the entire gene was discarded. If the duplications for a a small
+number of taxa were unresolved, all genes from those taxa were excluded.
+Genes were then realigned and sites were filtered with
+Gblocks\textsuperscript{41}, resulting in a 77 taxon matrix. Some taxa
+in this matrix were quite unstable, which obscured other
+strongly-supported relationships. Unstable taxa were identified with
+leaf stability indices\textsuperscript{42}, as implemented in
+phyutility\textsuperscript{43}, and removed from the matrix. This
+resulted in the 64-taxon matrix that is the focus of most of their
+analyses. Phylogenetic analyses were conducted under the Poisson+CAT
+model in PhyloBayes, and under the WAG model in
+MrBayes\textsuperscript{44} and RAxML\textsuperscript{45}.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+The placement of ctenophores (comb jellies) as the sister group to all
+other sampled metazoans is strongly supported in all our analyses. This
+result, which has not been postulated before, should be viewed as
+provisional until more data are considered from placozoans and
+additional sponges.
+\end{quote}
+
+Note that there was, in fact, an exception to strong support. An
+analysis of the 40 ribosomal proteins in the matrix recovered
+Ctenophora-sister with only 69\% support. This study did not include
+\emph{Trichoplax}.
+
+\hypertarget{philippe-et-al.-2009}{%
+\paragraph{\texorpdfstring{Philippe \emph{et al.}
+2009}{Philippe et al. 2009}}\label{philippe-et-al.-2009}}
+
+Philippe \emph{et al.}\textsuperscript{22} assembled an EST dataset for
+55 species with 128 genes to explore phylogenetic relationship of early
+diverging animals by added 9 new species. Th data matrix were assembled
+based on The phylogenetic analysis using Poisson-CAT model strongly
+supported Porifera-sister, and ctenophores was sister to cnidarians to
+form the ``coelenterate'' clade. Gene trees were then constructed, and
+potentially paralogs were removed by a bootstrap threshold of 70.
+Ambiguously aligned regions were trimemed and only genes sampled for at
+least two-thirds of species were retained. The phylogenetic analyses
+were conducted under the Poisson+CAT model in PhyloBayes.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+The resulting phylogeny yields two significant conclusions reviving old
+views that have been challenged in the molecular era: (1) that the
+sponges (Porifera) are monophyletic and not paraphyletic as repeatedly
+proposed, thus undermining the idea that ancestral metazoans had a
+sponge-like body plan; (2) that the most likely position for the
+ctenophores is together with the cnidarians in a ``coelenterate'' clade.
+\end{quote}
+
+\hypertarget{hejnol-et-al.2009}{%
+\paragraph{Hejnol et al.~2009}\label{hejnol-et-al.2009}}
+
+Hejnol \emph{et al.}\textsuperscript{11} added EST sequences from seven
+taxa, and a total of 94 taxa were included in the final data matrix to
+explore animal phylogeny, especially the position of acoelomorph
+flatwroms. The orthology inference was largely similar to Dunn et
+al.~2008, with the exception of orthology genes were clustered by MCL.
+The final data matrix included 1497 genes, and then subsampled with 844,
+330 and 53 gens by different thresholds of gene occupancy. With the
+exception of 53 gene matrix, maximum likelihood analyses from all other
+datasets strongly supported Ctenophora-sister (models were selected by
+RaxML perl script).
+
+\hypertarget{pick-et-al.-2010}{%
+\paragraph{\texorpdfstring{Pick \emph{et al.}
+2010}{Pick et al. 2010}}\label{pick-et-al.-2010}}
+
+Pick \emph{et al.}\textsuperscript{46} sought to test whether
+Ctenophora-sister was an artefact of insufficient taxon sampling. They
+added new and additional published sequence data to the 64-taxon matrix
+of Dunn \emph{et al.}\textsuperscript{2}. The new taxa included 12
+sponges, 1 ctenophore, 5 cnidarians, and \emph{Trichoplax}. They further
+modified the matrix by removing 2,150 sites that were poorly sampled or
+aligned. They considered two different sets of outgroups:
+Choanoflagellata (resulting in Choanimalia) and the same sampling as
+Dunn \emph{et al.} (resulting in Opisthokonta).
+
+All their analyses were conducted under the F81+CAT+Gamma model in
+PhyloBayes, in both a Bayesian framework and with bootstrapping. All
+analyses have the same ingroup sampling and site removal so it isn't
+possible to independently assess the impact of these factors. Analyses
+with Choanimalia sampling recovered Porifera-sister with 72\% posterior
+probability (PP) and 91\% bootstrap support (BS). With broader
+Opisthokonta sampling, support for Porifera-sister is 84\% PP. This is
+an interesting case where increased outgroup sampling leads to increased
+support for Porifera-sister.
+
+The authors argue that previous results supporting Ctenophora-sister
+``are artifacts stemming from insufficient taxon sampling and
+long-branch attraction (LBA)'' and that ``this hypothesis should be
+rejected''. Although the posterior probabilities supporting
+Porifera-sister are not strong, they conclude:
+
+\begin{quote}
+Results of our analyses indicate that sponges are the sister group to
+the remaining Metazoa, and Placozoa are sister group to the Bilateria
+\end{quote}
+
+They also investigated saturation, and conclude that Dunn \emph{et
+al.}\textsuperscript{2} is more saturated than Philippe \emph{et al.}
+2009 {[}Philippe:2009hh{]}. Note that the Pick \emph{et
+al.}\textsuperscript{46} dataset is not reanalyzed here because
+partition data are not available, and due to site filtering the
+partition file from Dunn \emph{et al.}\textsuperscript{2} cannot be
+applied to this matrix.
+
+\hypertarget{nosenko-et-al.-2013}{%
+\paragraph{\texorpdfstring{Nosenko \emph{et al.}
+2013}{Nosenko et al. 2013}}\label{nosenko-et-al.-2013}}
+
+Nosenko \emph{et al.}\textsuperscript{47} added Expressed Sequence Tag
+(EST) data for 9 species of non-bilaterian metazoans (7 sponges). They
+constructed a novel matrix containing 122 genes and parsed them into two
+non-overlaping matrices (ribosomal and non-ribosomal genes) and found
+incongruent results of deep metazoan phylogeny. The other major finding
+was that ribosomal gene partitions showed significantly lower saturation
+than the non-ribosomal ones.
+
+Orthologs were constructed using the bioinfomatics pipeline
+OrthoSelect\textsuperscript{48}. They also evaluated level of
+saturation, leaf stability of sampled taxa, compositional heterogeinity
+and model comparison of each matrix. By modifying gene sampling, ingroup
+and outgroup sampling, three major topologies related to the position of
+animal-root were constructed (including Porifera+Placozoa sister,
+Ctenophora-sister and Porifera-sister). Phylogenetic analyses were
+conducted under the Poisson+CAT, GTR+CAT and GTR models in PhyloBayes.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+we were able to reconstruct a metazoan phylogeny that is consistent with
+traditional, morphology-based views on the phylogeny of non-bilaterian
+metazoans, including monophyletic Porifera and ctenophores as a
+sister-group of cnidarians.
+\end{quote}
+
+Note the main conclusion of this paper is baed on the selection of genes
+that evolved slowly ribosomal RNA genes. The Ctenophora-sister was
+rejected based on the most saturated dataset and not supported by the
+matrix with the most closely related outgroups (Choanomalia
+matrix)\textsuperscript{47}.
+
+\hypertarget{ryan-et-al.-2013}{%
+\paragraph{\texorpdfstring{Ryan \emph{et al.}
+2013}{Ryan et al. 2013}}\label{ryan-et-al.-2013}}
+
+Ryan \emph{et al.} sequenced the first ctenophore genome of
+\emph{Mnemiopsis leidyi}. With the genome resources of \emph{M. leidyi},
+the authors constructed two phylogenomic datasets: a ``Genome set''
+based on 13 animal genomes and a ``EST Set'' that also included 59
+animals. They analyzed both matrices by site-homogeneous GTR+Gamma and
+site-heterogeneous Poisson-CAT models with three sets of outgroup
+sampling to evaluate the effect of outgroup selection to the ingroup
+topology for the Ryan2013\_est matrix. The Orthologs were constructed
+based on the method of Hejnol \emph{et al.} 2009. For the
+Ryan2013\_genome matrix, they performed phylogenetic analyses with both
+gene content and sequence-baed analyses. Overall, their results strongly
+supported Ctenophora-sister in all datasets they analyzed using
+site-homogeneous model. The Poisson+CAT model of the genome dataset
+strongly supported of a clade of Ctenophora and Porifera as the sister
+group to all other Metazoa and Bayesian analysis on the EST dataset did
+not converge after 205 days (but strongly supported Porifera in
+Choaimalia matrix).
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+Our phylogenetic analyses suggest that ctenophores are the sister group
+to the rest of the extant animals.
+\end{quote}
+
+\hypertarget{moroz-et-al.-2014}{%
+\paragraph{\texorpdfstring{Moroz \emph{et al.}
+2014}{Moroz et al. 2014}}\label{moroz-et-al.-2014}}
+
+Moroz \emph{et al.}\textsuperscript{49} sequenced the second ctenophore
+genome \emph{Pleurobrachia bachei} to explore the phylogenetic
+relationship of Metazoa. All phylogenetic analyses strongly supported
+Ctenophora-sister with different taxon and gene sampling using WAG
+site-homogeneous model. Two phylogenomic matrices were generated, the
+first set was represented by two ctenophore species, whereas the other
+set contained improved ctenophore sampling (10 taxa, Moroz2013\_3d).
+Orthology determination employed in HaMStR\textsuperscript{50} using
+1,032 ``model organism'' single-copy orthologs. Sequence were then
+trimmed and aligned. This resulted in a final matrix of 170,871 amino
+acid positions across 586 genes with 44 taxa for the first matrix, and
+114 genes with 60 taxa for the second matrix. All the phylogenetic
+analyses were analyzed in RAxML under the WAG+CAT+F models (different
+than CAT models in PhyloBayes) to reduce the computational cost.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+Our integrative analyses place Ctenophora as the earliest lineage within
+Metazoa. This hypothesis is supported by comparative analysis of
+multiple gene families, including the apparent absence of HOX genes,
+canonical microRNA machinery, and reduced immune complement in
+ctenophores.
+\end{quote}
+
+It should be noted that only the Moroz\_3d matrix has been reanalyzed in
+other studies, although the support of Ctenophora-sister is quite low.
+
+\hypertarget{borowiec-et-al.-2015}{%
+\paragraph{\texorpdfstring{Borowiec \emph{et al.}
+2015}{Borowiec et al. 2015}}\label{borowiec-et-al.-2015}}
+
+Borowiec \emph{et al.}\textsuperscript{12} assembled a genome dataset
+comprised 1080 orthologs derived from 36 publicly available genomes
+representing major lineages of animals, although only one genome of
+sponge and ctenophore was included. The orthologs were constructed using
+OrthologID pipeline\textsuperscript{51}. After removal of spurious
+sequences and genes with more than 40\% of mission data, the final
+matrix included 1080 (Total 1080). The authors further filtered the full
+dataset to 9 sub-datasets by filtering genes with high long-branch
+scores; genes with high saturation; gene occupancy; fast evolving genes.
+The main conclusion of the paper was largely based on
+BorowiecTotal\_1080 and Borowiec\_Best108 matrices. Phylogenetic
+analyses were conducted under the GTR+CAT model in PhyloBayes in
+selected matrices, and under the data-partitioning methods in RAxML for
+all matrices.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+Our phylogeny supports the still-controversial position of ctenophores
+as sister group to all other metazoans. This study also provides a
+workflow and computational tools for minimizing systematic bias in
+genome-based phylogenetic analyses.
+\end{quote}
+
+It should be noted that the authors also employed recoding-method in the
+Borowiec\_Best108 matrix and found neither support of Porifera-sister or
+Ctenophora-sister\textsuperscript{12}.
+
+\hypertarget{whelan-et-al.2015}{%
+\paragraph{Whelan et al.~2015}\label{whelan-et-al.2015}}
+
+Whelan \emph{et al.} 2015\textsuperscript{52} constructed a new
+phylogenomic dataset by eight new transcripomic data and investigated a
+range of possible sources of systematic error under multiple analyses
+(e.g.~long-branch attraction, compositional bias, fast evolving genes,
+etc.). Putative orthologs were determined of each species using HaMStR
+using the model organism core ortholog set (same as Moroz \emph{et al.}
+2014) and subsequently removal of genes with too much missing data and
+potential paralogs. The authors further filtered the full dataset to 24
+sub-datasets by filtering genes with high long-branch scores; genes with
+high RSFV values; genes that are potential paralogs; fast evolving genes
+and progressively removal of outgroups. All the maximum likelihood
+analyses with site-homogeneous model and PartitionFinder strongly
+suggested Ctenophora-sister. CAT-GTR models only used in least saturated
+dataset 6 and 16 also strongly supported Ctenophora.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+Importantly, biases resulting from elevated compositional heterogeneity
+or elevated substitution rates are ruled out. Placement of ctenophores
+as sister to all other animals, and sponge monophyly, are strongly
+supported under multiple analyses, herein.
+\end{quote}
+
+Note that the authors also reanalyzed Philippe2009 matrix (with the
+removal of ribosomal genes) and recovered Porifera-sister with moderate
+support (pp=90).
+
+\hypertarget{chang-et-al.2015}{%
+\paragraph{Chang et al.~2015}\label{chang-et-al.2015}}
+
+Chang \emph{et al.}\textsuperscript{53} was originally used to explore
+phylogenetic position of Myxozoa in Cnidaria but also sampled broadly
+across the breadth of animal diversity. The authors constructed a
+dataset with 200 protein markers based on Philippe et
+al.~2011\textsuperscript{7} with 51,940 amino acids and 77 taxa. Both
+site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly
+supported Ctenophora-sister.
+
+Note that this data matrix has been extensively reanalyzed in many
+studies.
+
+\hypertarget{pisani-et-al.-2015}{%
+\paragraph{\texorpdfstring{Pisani \emph{et al.}
+2015}{Pisani et al. 2015}}\label{pisani-et-al.-2015}}
+
+Pisani \emph{et al.}\textsuperscript{6} reanalyzed representative
+datasets that supported Ctenophora-sister, including Ryan2013\_est,
+Moroz2014\_3d and Whelan2015 datasets. It was the first study showing
+that progressively removal of more distantly related outgroups could
+largely affect phylogenomic inference of the position of the root of
+animal phylogeny. The authors suggested that the inclusion of outgroups
+very distant from the ingroup can cause systematic errors due to
+long-branch attraction. Phylogenetic analyses were conducted under the
+Poisson+CAT and GTR models in PhyloBayes. They found GTR-CAT and
+Poisson-CAT models generally had better model-fit than site-homogeneous
+GTR models in these data matrice. Moreover, they found the support of
+Ctenophora-sister decreases when the exclusion of distantly related
+outgroups and the use of site-heterogeneous CAT models.
+
+Regarding the recovery of Porifera-sister, the authors concluded:
+
+\begin{quote}
+Our results reinforce a traditional scenario for the evolution of
+complexity in animals, and indicate that inferences about the evolution
+of Metazoa based on the Ctenophora-sister hypothesis are not supported
+by the currently available data.
+\end{quote}
+
+\hypertarget{feuda-et-al.-2017}{%
+\paragraph{\texorpdfstring{Feuda \emph{et al.}
+2017}{Feuda et al. 2017}}\label{feuda-et-al.-2017}}
+
+Feuda \emph{et al.}\textsuperscript{9} didn't generate any new data,
+instead they used the data-recoding methods to reanalyze two key
+datasets that support Ctenophora-sister (Whelan2015\_D20, Chang2015
+datasets). It was the first phylogenomic study that suggested recoding
+methods have better performance than non-recoding methods in the
+relationships at the root of the animal tree. The authors compared model
+adequacy using posterior predictive analyses from a set of
+site-homogeneous (WAG, LG, GTR, data-partitioning) and
+site-heterogeneous (CAT-GTR) models in non-recoding and recoding
+datasets. The results showed that data-recoding can significant reduce
+compositional heterogeneity in both datasets with CAT-GTR models and
+strongly supported Porifera-sister hypothesis (see more details in
+recoding section in supplementary text).
+
+Regarding the recovery of Porifera-sister, the authors concluded:
+
+\begin{quote}
+Because adequate modeling of the evolutionary process that generated the
+data is fundamental to recovering an accurate phylogeny, our results
+strongly support sponges as the sister group of all other animals and
+provide further evidence that Ctenophora-sister represents a tree
+reconstruction artifact.
+\end{quote}
+
+\hypertarget{whelan-and-halanych-2016}{%
+\paragraph{Whelan and Halanych 2016}\label{whelan-and-halanych-2016}}
+
+Whelan \emph{et al.}\textsuperscript{21} is the only study to evaluate
+performance of site-heterogeneous models and site-homogeneous model with
+data partitioning under the simulation framework. The simulation results
+suggested that Poisson+CAT model consistently performed worse than other
+models in simulation datasets. More importantly, the authors also showed
+that both Poisson+CAT and GTR+ CAT models could overestimated
+substitutional heterogeneity in many cases. They also reanalyzed
+datasets from Philippe 2009 and Nosenko 2013 using both CAT models and
+data partitioning with site-homogeneous model. The results indicated
+that Poisson + CAT model tends to recover less accurate trees and more
+importantly, both GTR + CAT and data partitioning strongly supported
+Ctenophora-sister in reanalyses.
+
+The authors concluded:
+
+\begin{quote}
+Practices such as removing constant sites and parsimony uninformative
+characters, or using CAT-F81 when CAT-GTR is deemed too computationally
+expensive, cannot be logically justified. Given clear problems with
+CAT-F81, phylogenies previously inferred with this model should be
+reassessed.
+\end{quote}
+
+\hypertarget{whelan-et-al.-2017}{%
+\paragraph{\texorpdfstring{Whelan \emph{et al.}
+2017}{Whelan et al. 2017}}\label{whelan-et-al.-2017}}
+
+Whelan \emph{et al.}\textsuperscript{13} added 27 new ctenophore
+transcriptomic data to explore animal-root position as well as
+relationships within Ctenophores. It significantly increased ctenophore
+taxon sampling than other studies. Putative orthologs were determined
+largely similar to Whelan2015, with the difference of a core Ctenophora
+core datasets were constructed here with more than 2000 genes. The
+subsequent filtering strategy was also similar to the previous study.
+All analyses using site-homogeneous and site-heterogeneous models
+strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model
+in choanoazoa dataset. The main conclusions of this paper were based on
+Whelan2017\_full and Whelan2017\_strict matrices.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+\textgreater{}Using datasets with reasonably high ctenophore and other
+non-bilaterian taxon sampling, our results strongly reject the
+hypothesis that sponges are the sister lineage to all other extant
+metazoans.
+
+\hypertarget{simion-et-al.-2017}{%
+\paragraph{\texorpdfstring{Simion \emph{et al.}
+2017}{Simion et al. 2017}}\label{simion-et-al.-2017}}
+
+Simion \emph{et al.}\textsuperscript{14} added transcriptomic data for
+21 new animals. The data matrix was constructed using a semi-automated
+approach to comprehensively detect and eliminate potential systematic
+errors. The resulting dataset comprises 1,719 genes and 97 species,
+including 61 non-bilaterian species. It was by far the largest
+phylogenomic dataset in terms of taxon and gene sampling related to the
+relationship at the root of animal phylogeny.
+
+The final matrix was first analyzed using the Poisson-CAT model.
+Different that other phylobayes analyses, Simion \emph{et al.} used a
+gene jackknife strategy based on 100 analyses to overcome the
+computational limitation because of the large data size. Each jackknife
+is baed on a random selection of \textasciitilde{} 25\% of the genes.
+The Phylobayes with site-heterogeneous model strongly supported the
+Porifera-sister, whereas site-homogeneous strongly supported
+Ctenophora-sister in all datasets. Importantly, the authors compared the
+behavior of long-branch effect between site-heterogeneous and
+site-homogeneous models by progressively removal taxa and concluded
+higher sensitivity of site-homogeneous models to LBA than CAT models.
+
+Regarding the recovery of Ctenophora-sister, the authors concluded:
+
+\begin{quote}
+Our dataset outperforms previous metazoan gene superalignments in terms
+of data quality and quantity. Analyses with a best-fitting
+site-heterogeneous evolutionary model provide strong statistical support
+for placing sponges as the sister-group to all other metazoans, with
+ctenophores emerging as the second-earliest branching animal lineage.
+\end{quote}
+
+It should be noted that all the PhyloBayes runs have not reached
+convergence due to the computational cost in these large matrices.
+
+\hypertarget{data-recoding-methods}{%
+\subsubsection{Data-recoding methods}\label{data-recoding-methods}}
+
+Feuda \emph{et al.}\textsuperscript{35} tackled a difficult phylogenetic
+problem -- placing the root of the animal phylogeny\textsuperscript{8}.
+In the past decade, some analyses have placed Ctenophora (comb jellies)
+and others Porifera (sponges) as the sister group to all other animals
+(Fig. Extended Data Fig. 3B). Feuda \emph{et al.}\textsuperscript{35}
+present new phylogenetic analyses that they claim provide strong support
+for Porifera-sister. Their analyses consider datasets from studies by
+Chang \emph{et al.}\textsuperscript{53} and Whelan \emph{et
+al.}\textsuperscript{54}, both of which found support for
+Ctenophora-sister. Feuda \emph{et al.} were concerned that these
+Ctenophora-sister results were artefacts of lineage-specific differences
+in amino acid frequencies. In an attempt to reduce these differences,
+they recoded the full set of twenty amino acids into six groups of amino
+acids. These groups have more frequent evolutionary changes within them
+than between them, based on empirical observations in large protein
+datasets\textsuperscript{34}. The intent is to discard many
+lineage-specific changes, which are expected to fall within these
+groups. Rather than model compositional heterogeneity, as their title
+suggests, this approach discards heterogeneous information so that much
+simpler models with fewer states can be applied. They report that
+posterior predictive (PP) analyses\textsuperscript{24} indicate 6-state
+recoded analyses have better model adequacy than 20-state amino acid
+analyses, and ``Porifera-sister was favored under all recoding
+strategies''. Here we focus on two aspects of Feuda \emph{et al.} First,
+we point out that many of their recoded analyses are actually unresolved
+(\emph{i.e.}, without strong support for either Porifera-sister or
+Ctenophora-sister), and that the analyses with the best posterior
+predictive scores do not provide strong support for Porifera-sister.
+Second, we present new analyses that show the impact of recoding is
+largely due to discarding information, not accommodating variation in
+amino acid composition. These findings indicate that it is premature to
+accept Porifera-sister and reject Ctenophora-sister. They also show that
+recoding can be a problematic method for addressing compositional
+variation.
+
+Feuda \emph{et al.} examine support for Ctenophora-sister and
+Porifera-sister under all combinations of two models of molecular
+evolution, four datasets, and four coding schemes. This provides 32
+analyses that they report in their Table 3 and that we present
+graphically here as Extended Data Fig. 6. There is striking variation in
+support for Ctenophora-sister and Porifera-sister across these analyses
+(Extended Data Fig. 6). Feuda \emph{et al.} accept the results of some
+analyses and reject others based on posterior predictive (PP) analyses
+of model adequacy, which assess how well a model explains variation in
+the data\textsuperscript{24}. They considered five different posterior
+predictive statistics that capture different types of variation in the
+data. From this they conclude that their ``results strongly support
+sponges as the sister group of all other animals''.
+
+This conclusion does not follow from their own presented results. Only a
+single analysis with posterior predictive scores provides what could be
+considered strong support \textgreater{} 95 posterior probability) for
+Porifera-sister. Of the 32 analyses, posterior predictive scores were
+calculated for 16 (those for the full Whelan and Chang matrices). Based
+on posterior predictive scores, Feuda \emph{et al.} reject eight of
+these that were conducted under the GTR+G model (which all have strong
+support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses
+(Extended Data Fig. 5). Two of these eight are analyses of the original
+20-state amino acid data, both of which provide strong support for
+Ctenophora-sister. Of the six recoded analyses, five are unresolved.
+Only a single analysis for which posterior predictive scores are
+available provides strong support for Porifera-sister, the CAT-GTR+G
+analysis of the SR-6\textsuperscript{34} recoded
+Whelan\textsuperscript{54} matrix. Furthermore, this analysis does not
+have the best score according to any of the five posterior predictive
+statistics they considered (Extended Data Fig. 5). The only statistic
+that stands out for this one analysis is that it has the highest maxdiff
+(Extended Data Fig. 5), indicating that it did not converge as well as
+other analyses.
+
+Though their study does not provide strong support for Porifera-sister,
+the sensitivity of their results to recoding provides an opportunity to
+better understand and evaluate the impact of recoding more generally.
+This is important given the growing interest in
+recoding\textsuperscript{35}. Feuda \emph{et al.} hoped recoding would
+reduce potential artefacts due to differences across species in amino
+acid frequencies. They interpreted the fact that their analyses are
+sensitive to recoding as evidence that such an artefact exists and that
+they successfully addressed it by recoding. An alternative hypothesis is
+that recoding impacts phylogenetic analyses because it discards so much
+information. These two hypotheses can be tested by applying new recoding
+schemes that also reduce twenty states down to six, but are based on
+random grouping rather than empirical frequencies of amino acid
+exchange. Empirical and random recodings both discard the same amount of
+information, but only empirical recoding reduce the impact of amino-acid
+frequency as intended. Different results between empirical and random
+recoding would be consistent with the hypothesis that the empirical
+approach works as intended to accommodate compositional heterogeneity.
+Similar results would suggest that the impact of recoding is due to
+discarding information. Here we focus on their single analysis with a
+posterior predictive score that supports Porifera-sister, the CAT-GTR+G
+analysis of the SR-6 recoded Whelan data. we created four new random
+recoding schemes by shuffling the amino acids in the SR-6 scheme (see
+Supplemental Methods and analysis code at
+\url{https://github.com/caseywdunn/feuda_2017}). When we applied each of
+these randomized codes to the Whelan matrix and analyzed them under the
+CAT-GTR+G model with PhyloBayes-MPI, we observed similar results as for
+the empirical SR-6 recoding. Like SR-6 recoding, random recoding
+increases support for Porifera-sister and improves the apparent adequacy
+of models to explain heterogeneity of states across taxa (PP taxon
+hetero mean and max, Extended Data Fig. 3B).
+
+These analyses suggest that the major impact of recoding on phylogenetic
+analyses is data reduction, not accommodation of compositional
+heterogeneity across species. This indicates that recoding may not be an
+effective tool for accommodating among-species differences in amino acid
+frequencies. Compositional heterogeneity would be better addressed with
+models of molecular evolution that explicitly describe such
+differences\textsuperscript{55}, if progress can be made on the
+considerable computational challenges of such complex models.
+
 \hypertarget{references}{%
@@
 
-\leavevmode\hypertarget{ref-chang2015genomic}{}%
-10.Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H.,
-Huchon, D. \& Cartwright, P. Genomic insights into the evolutionary
-origin of myxozoa within cnidaria. \emph{Proceedings of the National
-Academy of Sciences} \textbf{112}, 14912--14917 (2015).
-
-\leavevmode\hypertarget{ref-whelan2015error}{}%
-11.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
-signal, and the placement of ctenophora sister to all other animals.
-\emph{Proceedings of the National Academy of Sciences} \textbf{112},
-5773--5778 (2015).
-
 \leavevmode\hypertarget{ref-hernandez2019six}{}%
-12.Hernandez, A.M. \& Ryan, J.F. Six-state amino acid recoding is not an
+10.Hernandez, A.M. \& Ryan, J.F. Six-state amino acid recoding is not an
 effective strategy to offset the effects of compositional heterogeneity
@@
 \leavevmode\hypertarget{ref-hejnol2009assessing}{}%
-13.Hejnol, A., Obst, M., Stamatakis, A., Ott, M., Rouse, G.W.,
+11.Hejnol, A., Obst, M., Stamatakis, A., Ott, M., Rouse, G.W.,
 Edgecombe, G.D., Martinez, P., Baguñà, J., Bailly, X., Jondelius, U. \&
@@
 \leavevmode\hypertarget{ref-borowiec2015extracting}{}%
-14.Borowiec, M.L., Lee, E.K., Chiu, J.C. \& Plachetzki, D.C. Extracting
+12.Borowiec, M.L., Lee, E.K., Chiu, J.C. \& Plachetzki, D.C. Extracting
 phylogenetic signal and accounting for bias in whole-genome data sets
@@
 \leavevmode\hypertarget{ref-whelan2017ctenophore}{}%
-15.Whelan, N.V., Kocot, K.M., Moroz, T.P., Mukherjee, K., Williams, P.,
+13.Whelan, N.V., Kocot, K.M., Moroz, T.P., Mukherjee, K., Williams, P.,
 Paulay, G., Moroz, L.L. \& Halanych, K.M. Ctenophore relationships and
@@
 \leavevmode\hypertarget{ref-simion2017large}{}%
-16.Simion, P., Philippe, H., Baurain, D., Jager, M., Richter, D.J., Di
+14.Simion, P., Philippe, H., Baurain, D., Jager, M., Richter, D.J., Di
 Franco, A., Roure, B., Satoh, N., Queinnec, E., Ereskovsky, A. \& others
@@
 
-\leavevmode\hypertarget{ref-lanier2012recombination}{}%
-17.Lanier, H.C. \& Knowles, L.L. Is recombination a problem for
-species-tree analyses? \emph{Systematic Biology} \textbf{61}, 691--701
-(2012).
-
-\leavevmode\hypertarget{ref-salichos2013inferring}{}%
-18.Salichos, L. \& Rokas, A. Inferring ancient divergences requires
-genes with strong phylogenetic signals. \emph{Nature} \textbf{497}, 327
-(2013).
-
 \leavevmode\hypertarget{ref-shen2018tempo}{}%
-19.Shen, X.-X., Opulente, D.A., Kominek, J., Zhou, X., Steenwyk, J.L.,
+15.Shen, X.-X., Opulente, D.A., Kominek, J., Zhou, X., Steenwyk, J.L.,
 Buh, K.V., Haase, M.A., Wisecaver, J.H., Wang, M., Doering, D.T. \&
@@
 \leavevmode\hypertarget{ref-fernandez2018phylogenomics}{}%
-20.Fernández, R., Kallal, R.J., Dimitrov, D., Ballesteros, J.A., Arnedo,
+16.Fernández, R., Kallal, R.J., Dimitrov, D., Ballesteros, J.A., Arnedo,
 M.A., Giribet, G. \& Hormiga, G. Phylogenomics, diversification
@@
 
-\leavevmode\hypertarget{ref-boeckmann2003swiss}{}%
-21.Boeckmann, B., Bairoch, A., Apweiler, R., Blatter, M.-C., Estreicher,
-A., Gasteiger, E., Martin, M.J., Michoud, K., O'Donovan, C., Phan, I. \&
-others The swiss-prot protein knowledgebase and its supplement trembl in
-2003. \emph{Nucleic acids research} \textbf{31}, 365--370 (2003).
-
 \leavevmode\hypertarget{ref-nguyen2014iq}{}%
-22.Nguyen, L.-T., Schmidt, H.A., Haeseler, A. von \& Minh, B.Q. IQ-tree:
+17.Nguyen, L.-T., Schmidt, H.A., Haeseler, A. von \& Minh, B.Q. IQ-tree:
 A fast and effective stochastic algorithm for estimating
@@
 \leavevmode\hypertarget{ref-zhou2017evaluating}{}%
-23.Zhou, X., Shen, X.-X., Hittinger, C.T. \& Rokas, A. Evaluating fast
+18.Zhou, X., Shen, X.-X., Hittinger, C.T. \& Rokas, A. Evaluating fast
 maximum likelihood-based phylogenetic programs using empirical
@@
 \leavevmode\hypertarget{ref-si2008empirical}{}%
-24.Si Quang, L., Gascuel, O. \& Lartillot, N. Empirical profile mixture
+19.Si Quang, L., Gascuel, O. \& Lartillot, N. Empirical profile mixture
 models for phylogenetic reconstruction. \emph{Bioinformatics}
@@
 \leavevmode\hypertarget{ref-kalyaanamoorthy2017modelfinder}{}%
-25.Kalyaanamoorthy, S., Minh, B.Q., Wong, T.K., Haeseler, A. von \&
+20.Kalyaanamoorthy, S., Minh, B.Q., Wong, T.K., Haeseler, A. von \&
 Jermiin, L.S. ModelFinder: Fast model selection for accurate
@@
 \leavevmode\hypertarget{ref-whelan2016let}{}%
-26.Whelan, N.V. \& Halanych, K.M. Who let the cat out of the bag?
+21.Whelan, N.V. \& Halanych, K.M. Who let the cat out of the bag?
 Accurately dealing with substitutional heterogeneity in phylogenomic
@@
 \leavevmode\hypertarget{ref-Philippe:2009hh}{}%
-27.Philippe, H., Derelle, R., Lopez, P., Pick, K., Borchiellini, C.,
+22.Philippe, H., Derelle, R., Lopez, P., Pick, K., Borchiellini, C.,
 Boury-Esnault, N., Vacelet, J., Renard, E., Houliston, E., QuEinnec, E.,
@@
 \leavevmode\hypertarget{ref-shen2017contentious}{}%
-28.Shen, X.-X., Hittinger, C.T. \& Rokas, A. Contentious relationships
+23.Shen, X.-X., Hittinger, C.T. \& Rokas, A. Contentious relationships
 in phylogenomic studies can be driven by a handful of genes.
@@
 
-\leavevmode\hypertarget{ref-khan2017intervene}{}%
-29.Khan, A. \& Mathelier, A. Intervene: A tool for intersection and
-visualization of multiple gene or genomic region sets. \emph{BMC
-bioinformatics} \textbf{18}, 287 (2017).
-
 \leavevmode\hypertarget{ref-Bollback:2002to}{}%
-30.Bollback, J.P. Bayesian model adequacy and choice in phylogenetics.
+24.Bollback, J.P. Bayesian model adequacy and choice in phylogenetics.
 \emph{Molecular Biology and Evolution} \textbf{19}, 1171--1180 (2002).
 
+\leavevmode\hypertarget{ref-Lartillot:2013fg}{}%
+25.Lartillot, N., Rodrigue, N., Stubbs, D. \& Richer, J. PhyloBayes MPI:
+phylogenetic reconstruction with infinite mixtures of profiles in a
+parallel environment. \emph{Systematic biology} \textbf{62}, 611--615
+(2013).
+
 \leavevmode\hypertarget{ref-guang2016integrated}{}%
-31.Guang, A., Zapata, F., Howison, M., Lawrence, C.E. \& Dunn, C.W. An
+26.Guang, A., Zapata, F., Howison, M., Lawrence, C.E. \& Dunn, C.W. An
 integrated perspective on phylogenetic workflows. \emph{Trends in
@@
 \leavevmode\hypertarget{ref-mah2014choanoflagellate}{}%
-32.Mah, J.L., Christensen-Dalsgaard, K.K. \& Leys, S.P. Choanoflagellate
+27.Mah, J.L., Christensen-Dalsgaard, K.K. \& Leys, S.P. Choanoflagellate
 and choanocyte collar-flagellar systems and the assumption of homology.
@@
 \leavevmode\hypertarget{ref-nielsen2019early}{}%
-33.Nielsen, C. Early animal evolution: A morphologist's view.
+28.Nielsen, C. Early animal evolution: A morphologist's view.
 \emph{Royal Society open science} \textbf{6}, 190638 (2019).
@@
 \leavevmode\hypertarget{ref-zhao2019cambrian}{}%
-34.Zhao, Y., Vinther, J., Parry, L.A., Wei, F., Green, E., Pisani, D.,
+29.Zhao, Y., Vinther, J., Parry, L.A., Wei, F., Green, E., Pisani, D.,
 Hou, X., Edgecombe, G.D. \& Cong, P. Cambrian sessile, suspension
@@
 \leavevmode\hypertarget{ref-Buchfink:2014}{}%
-35.Buchfink, B., Xie, C. \& Huson, D.H. Fast and sensitive protein
+30.Buchfink, B., Xie, C. \& Huson, D.H. Fast and sensitive protein
 alignment using diamond. \emph{Nature Methods} \textbf{12}, 59 EP
@@
 
+\leavevmode\hypertarget{ref-Camacho:2009}{}%
+31.Camacho, C., Coulouris, G., Avagyan, V., Ma, N., Papadopoulos, J.,
+Bealer, K. \& Madden, T.L. BLAST+: Architecture and applications.
+\emph{BMC Bioinformatics} \textbf{10}, 421 (2009).
+
+\leavevmode\hypertarget{ref-boeckmann:2003swiss}{}%
+32.Boeckmann, B., Bairoch, A., Apweiler, R., Blatter, M.-C., Estreicher,
+A., Gasteiger, E., Martin, M.J., Michoud, K., O'Donovan, C., Phan, I. \&
+others The swiss-prot protein knowledgebase and its supplement trembl in
+2003. \emph{Nucleic acids research} \textbf{31}, 365--370 (2003).
+
 \leavevmode\hypertarget{ref-Hagberg:2008}{}%
-36.Hagberg, A.A., Schult, D.A. \& Swart, P.J. Exploring network
+33.Hagberg, A.A., Schult, D.A. \& Swart, P.J. Exploring network
 structure, dynamics, and function using networkx. \emph{Proceedings of
@@
 
-\leavevmode\hypertarget{ref-Lartillot:2013fg}{}%
-37.Lartillot, N., Rodrigue, N., Stubbs, D. \& Richer, J. PhyloBayes MPI:
-phylogenetic reconstruction with infinite mixtures of profiles in a
-parallel environment. \emph{Systematic biology} \textbf{62}, 611--615
-(2013).
-
 \leavevmode\hypertarget{ref-Susko:2007ds}{}%
-38.Susko, E. \& Roger, A.J. On reduced amino acid alphabets for
+34.Susko, E. \& Roger, A.J. On reduced amino acid alphabets for
 phylogenetic inference. \emph{Molecular Biology and Evolution}
@@
 \leavevmode\hypertarget{ref-Feuda:2017ew}{}%
-39.Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O.,
+35.Feuda, R., Dohrmann, M., Pett, W., Philippe, H., Rota-Stabelli, O.,
 Lartillot, N., Wörheide, G. \& Pisani, D. Improved Modeling of
@@
 
+\leavevmode\hypertarget{ref-khan2017intervene}{}%
+36.Khan, A. \& Mathelier, A. Intervene: A tool for intersection and
+visualization of multiple gene or genomic region sets. \emph{BMC
+bioinformatics} \textbf{18}, 287 (2017).
+
 \leavevmode\hypertarget{ref-Felsenstein:1981vk}{}%
-40.Felsenstein, J. Evolutionary trees from DNA sequences: a maximum
+37.Felsenstein, J. Evolutionary trees from DNA sequences: a maximum
 likelihood approach. \emph{Journal of Molecular Evolution} \textbf{17},
@@
 \leavevmode\hypertarget{ref-Whelan:2001ds}{}%
-41.Whelan, S. \& Goldman, N. A General Empirical Model of Protein
+38.Whelan, S. \& Goldman, N. A General Empirical Model of Protein
 Evolution Derived from Multiple Protein Families Using a
@@
 \leavevmode\hypertarget{ref-Le:2008fp}{}%
-42.Le, S.Q. \& Gascuel, O. An improved general amino acid replacement
+39.Le, S.Q. \& Gascuel, O. An improved general amino acid replacement
 matrix. \emph{Molecular Biology and Evolution} \textbf{25}, 1307--1320
@@
 \leavevmode\hypertarget{ref-Enright:2002uq}{}%
-43.Enright, A., Van Dongen, S. \& Ouzounis, C. An efficient algorithm
+40.Enright, A., Van Dongen, S. \& Ouzounis, C. An efficient algorithm
 for large-scale detection of protein families. \emph{Nucleic Acids
@@
 \leavevmode\hypertarget{ref-Castresana:2000vy}{}%
-44.Castresana, J. Selection of conserved blocks from multiple alignments
+41.Castresana, J. Selection of conserved blocks from multiple alignments
 for their use in phylogenetic analysis. \emph{Molecular Biology and
@@
 \leavevmode\hypertarget{ref-Thorley:1999kg}{}%
-45.Thorley, J. \& Wilkinson, M. Testing the phylogenetic stability of
+42.Thorley, J. \& Wilkinson, M. Testing the phylogenetic stability of
 early tetrapods. \emph{Journal of Theoretical Biology} \textbf{200},
@@
 \leavevmode\hypertarget{ref-Smith:2008gb}{}%
-46.Smith, S.A. \& Dunn, C.W. Phyutility: a phyloinformatics tool for
+43.Smith, S.A. \& Dunn, C.W. Phyutility: a phyloinformatics tool for
 trees, alignments and molecular data. \emph{Bioinformatics} \textbf{24},
@@
 \leavevmode\hypertarget{ref-Ronquist:2003hx}{}%
-47.Ronquist, F. \& Huelsenbeck, J.P. MrBayes 3: Bayesian phylogenetic
+44.Ronquist, F. \& Huelsenbeck, J.P. MrBayes 3: Bayesian phylogenetic
 inference under mixed models. \emph{Bioinformatics} \textbf{19},
@@
 \leavevmode\hypertarget{ref-Stamatakis:2006wc}{}%
-48.Stamatakis, A. RAxML-VI-HPC: maximum likelihood-based phylogenetic
+45.Stamatakis, A. RAxML-VI-HPC: maximum likelihood-based phylogenetic
 analyses with thousands of taxa and mixed models. \emph{Bioinformatics}
@@
 \leavevmode\hypertarget{ref-Pick:2010eb}{}%
-49.Pick, K.S., Philippe, H., Schreiber, F., Erpenbeck, D., Jackson,
+46.Pick, K.S., Philippe, H., Schreiber, F., Erpenbeck, D., Jackson,
 D.J., Wrede, P., Wiens, M., Alie, A., Morgenstern, B., Manuel, M. \&
@@
 \leavevmode\hypertarget{ref-nosenko2013deep}{}%
-50.Nosenko, T., Schreiber, F., Adamska, M., Adamski, M., Eitel, M.,
+47.Nosenko, T., Schreiber, F., Adamska, M., Adamski, M., Eitel, M.,
 Hammel, J., Maldonado, M., Müller, W.E., Nickel, M., Schierwater, B. \&
@@
 \leavevmode\hypertarget{ref-schreiber2009orthoselect}{}%
-51.Schreiber, F., Pick, K., Erpenbeck, D., Wörheide, G. \& Morgenstern,
+48.Schreiber, F., Pick, K., Erpenbeck, D., Wörheide, G. \& Morgenstern,
 B. OrthoSelect: A protocol for selecting orthologous groups in
@@
 \leavevmode\hypertarget{ref-moroz2014ctenophore}{}%
-52.Moroz, L.L., Kocot, K.M., Citarella, M.R., Dosung, S., Norekian,
+49.Moroz, L.L., Kocot, K.M., Citarella, M.R., Dosung, S., Norekian,
 T.P., Povolotskaya, I.S., Grigorenko, A.P., Dailey, C., Berezikov, E.,
@@
 \leavevmode\hypertarget{ref-ebersberger2009hamstr}{}%
-53.Ebersberger, I., Strauss, S. \& Haeseler, A. von HaMStR: Profile
+50.Ebersberger, I., Strauss, S. \& Haeseler, A. von HaMStR: Profile
 hidden markov model based search for orthologs in ests. \emph{BMC
@@
 \leavevmode\hypertarget{ref-chiu2006orthologid}{}%
-54.Chiu, J.C., Lee, E.K., Egan, M.G., Sarkar, I.N., Coruzzi, G.M. \&
+51.Chiu, J.C., Lee, E.K., Egan, M.G., Sarkar, I.N., Coruzzi, G.M. \&
 DeSalle, R. OrthologID: Automation of genome-scale ortholog
@@
 
+\leavevmode\hypertarget{ref-whelan2015error}{}%
+52.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
+signal, and the placement of ctenophora sister to all other animals.
+\emph{Proceedings of the National Academy of Sciences} \textbf{112},
+5773--5778 (2015).
+
 \leavevmode\hypertarget{ref-Chang:2015hl}{}%
-55.Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H.,
+53.Chang, E.S., Neuhof, M., Rubinstein, N.D., Diamant, A., Philippe, H.,
 Huchon, D. \& Cartwright, P. Genomic insights into the evolutionary
@@
 \leavevmode\hypertarget{ref-Whelan:2015jj}{}%
-56.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
+54.Whelan, N.V., Kocot, K.M., Moroz, L.L. \& Halanych, K.M. Error,
 signal, and the placement of Ctenophora sister to all other animals.
@@
 \leavevmode\hypertarget{ref-Blanquart:2008gl}{}%
-57.Blanquart, S. \& Lartillot, N. A Site- and Time-Heterogeneous Model
+55.Blanquart, S. \& Lartillot, N. A Site- and Time-Heterogeneous Model
 of Amino Acid Replacement. \emph{Molecular Biology and Evolution}
@@
 \leavevmode\hypertarget{ref-Foster:2004tw}{}%
-58.Foster, P.G. Modeling compositional heterogeneity. \emph{Systematic
+56.Foster, P.G. Modeling compositional heterogeneity. \emph{Systematic
 biology} \textbf{53}, 485--495 (2004).

### 2020-04-28 manuscript/manuscript.rmd
fixed a few formatting issues to reknit for pdf and html. Set html to default output and removed other output formats.

@@
 output:
+  html_document: default
   pdf_document: default
   word_document: default
-  html_document: default
   github_document: default
@@
 This is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components– CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 2D).
-We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
+We also annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
 Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig.
@@
 In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$lnL<2) (Fig.
 5B; Supplementary Table 4).
@@
 More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
-For example, the major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
+For example, the major transition occurred at 60 -- 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
 Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
@@
 Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
-A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
 For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
@@
 (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation.
+The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and -1 is negative linear correlation.
 The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
@@
 (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
-The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.
+The $\Delta$GLS values for the genes across each data matrix are provided in Supplementary Table 4.
 
@@
 (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03).
+These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
 Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
@@
 
-> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.
+> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a 'coelenterate' clade.
 

### 2020-04-28 manuscript/manuscript.rmd
new abstract

@@
 
-## Summary (ideally no more than 200 words, currently 251 words)
-XX This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards
-
-Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life.
-However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses.
-This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution.
-Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions.
-Our reanalyses are generally consistent with previous results.
-We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices.
-We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
-We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic.
-Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers.
-This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.
-
-## Main (The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text), currently is ~4500 words...)
+## Summary
+
+Considerable debate remains regarding the deepest split in the animal tree of life.
+This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
+Some previous analyses found support for the first split being between comb jellies and all other animals (the Ctenophora-sister hypothesis), while others found support the first split being between sponges and all other animals (Porifera-sister).
+These studies differed in a many respects, including the genes and species selected, molecular evolution models, and analysis software.
+Here we systematically explore this question under consistent conditions by synthesizing data and results from XX previous phylogenomic analyses and performing new standardized analyses.
+Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
+Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models.
+Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.
+
+## Main
+
 Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.

### 2020-04-29 manuscript/manuscript.rmd
updates to main

@@
 
-Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life.
-However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig.
-1) emerging as the two primary hypotheses.
-Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
+Historically, there was little debate about the root of the animal tree of life and Porifera-sister (Fig.
+1) was widely accepted though rarely tested.
 By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns.
-However, a wide consensus of animal origin has not been reached despite multiple efforts.
-
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
 Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
-It has become clear that this is a very difficult phylogenetic challenge.
 As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
 Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
-Here we synthesize data and results from phylogenomic analyses that tested Ctenophora-sister and Porifera-sister.
-We reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
-This provides an integrative overview of the challenges posed by this question and provides direction for future studies.
-We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
 
@@
 
-## Variation across studies
-
-### Models of molecular evolution
-
-Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$ , the rate of evolution, and the amino acid equilibrium frequencies $\Pi$.
+## Variation across published studies
+
+The published analyses considere here differ in a variety of ways, including the model of molecular evolution, gene selection, ingroup species selection, and outgroup species selection.
+
+Models of molecular evolution describe how probable different evolutionary changes are, and can have a major impact on the topology of inferred phylogenies.
+Models have several components that each consider different aspects of the evolutionary process.
+The models used in the studies considered here differ according to three primary components: the exchangeability matrix $E$ , rates of evolution, and amino acid equilibrium frequencies $\Pi$.
 The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR).
@@
 
-### Gene and taxa sampling
-
 High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
@@
 As a result, the genes selected for analysis can vary widely between studies.
+
+Ingroup selection. XXX
 
@@
 In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
-
-### Data-recoding
-
-Compositional heterogeneity (clade-specific and site-specific differences in amino acid frequencies) have been a concern in the study of phylogenetics, and has been of particular concern for the deep animal phylogeny [@philippe2011resolving].
-Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie].
-Feuda *et al*.
-2017 [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices.
-By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent.
-However, how the recoding methods impact phylogenetic inference has not been well explored.
-Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].
 

### 2020-04-30 manuscript/manuscript.rmd
format supplementary text

@@
 
+**Extended Data Table 1.** TTable 1. Overview of data matrices used in this study.
 ```{r table_study_summary}
 tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
-	kable(tableXXdistribution, caption= "Table 1. Overview of data matrices used in this study")
+	kable(tableXXdistribution)
 
@@
 
-```{r table_phylogenetic_signal}
-tableXXdistribution = read_tsv("../trees_new/AU_test/summary.tsv")
-	kable(tableXXdistribution, caption= "Supplementary Table 4. Distribution of phylogenetic signal for different models and outgroup choice")
-
-```
-
-**Extended Data Table 1.** The models selected by ModelFinder for each matrix.
+**Extended Data Table 2.** The models selected by ModelFinder for each matrix.
 ```{r table_modelfinder}
@@
 
-## Supplementary Results
-
-### Models of molecular evolution
+### S.1 Models of molecular evolution
 Models of molecular evolution are assembled by selecting different options for all these different components.
@@
 
-### Details of published analyses
+### S.2 Details of published analyses
 
@@
 
-### Data-recoding methods
+### S.3 Data-recoding methods
 

### 2020-05-19 manuscript/manuscript.rmd
reworked intro

@@
 
-Considerable debate remains regarding the deepest split in the animal tree of life.
+There has been considerable recent debate about the deepest split in the animal tree of life.
 This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
@@
 
-Historically, there was little debate about the root of the animal tree of life and Porifera-sister (Fig.
+Historically, there was little debate about the root of the animal tree of life.
+Porifera-sister (Fig.
 1) was widely accepted though rarely tested.
 By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
-Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1).
-As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
-Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
+The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant animal relative.
+Since then many more studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Table 1).
+As the problem has become better characterized, it has become an interesting test-case for addressing difficult phylogenetic questions, which has attracted methodological interest well beyond investigators concerned with early animal evolution.
+These studies have taken a wide variety of approaches and results have varied across these different analyses.
+The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result.
+But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses.
+This variance is often confounded, for example when a particular model is only available in one software package.
+To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework.
 

### 2020-05-19 manuscript/manuscript.rmd
worked on body

@@
 This variance is often confounded, for example when a particular model is only available in one software package.
+
 To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework.
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
+These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.
 
@@
 
-## Variation across published studies
+## Variation in models and sampling across published studies
 
@@
 
-High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
-Surprisingly, gene overlap between studies has rarely been assessed.
-Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
-These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
-As a result, the genes selected for analysis can vary widely between studies.
-
-Ingroup selection. XXX
-
-Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig.
-1C).
-Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
-It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
-In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.
-
-## Overview of published analyses
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
-These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partitions that described boundaries between genes.
-
-### Matrix taxon composition
-
-Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood.
-However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig.
+Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa.
+As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species.
+Ingroup taxon sampling across clades varies widely between studies (Extended Data Fig.
 1).
-These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]).
+The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed.
+More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here.
 Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
 1).
-The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies.
-In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage.
-Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.
+Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig.
+XXXX).
 
@@
 
-### Matrix gene sampling are largely different across studies
-
-We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig.
-2A).
-The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco].
-The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
-Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices.
-However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig.
-2B).
-Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
+Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
+There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede].
+To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
+Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+It has been suggested the inclusion of more distant related (e.g.
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
+
+High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
+Studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
+These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
+As a result, the genes selected for analysis varies widely between studies (Fig.
+XX).
+This reflects the different methods used to collect sequence data and different criteria for gene selection.
+There are several ways to summarize and assess this variation in gene sampling.
+We annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1).
+These annotations show XXX.
+We identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
 2D).
-We also annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
-Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig.
-2).
-Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.
-
-### Support for Porifera-sister and Ctenophora-sister
+We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig.
+2A).
+The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are broadly shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco].
+The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
+We found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20% (Extended Data Fig.
+2B).
+This is an indication that gene sampling criterial are not only based on breadth of sampling, but also on other criteria such as features of molecular evolution and sampling effects such as sifferences in expression level.
+
+## Variation in published results
 
@@
 
-## New analyses of published matrices are largely consistent with previous
+## New standardized analyses of published matrices
 
@@
 
-### Drastic changes of phylogenetic signal between site- homogeneous and heterogeneous models
+### Drastic differences in phylogenetic signal between site- homogeneous and heterogeneous models
 

### 2020-05-25 manuscript/manuscript.rmd
modify main figures and legends

@@
 To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework.
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1).
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Table 1).
 These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.
@@
 
-The published analyses considere here differ in a variety of ways, including the model of molecular evolution, gene selection, ingroup species selection, and outgroup species selection.
+The published analyses considered here differ in a variety of ways, including the model of molecular evolution, gene selection, ingroup species selection, and outgroup species selection.
 
@@
 A detailed description of each component is provided in Supplementary Results.
-
-Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa.
-As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species.
-Ingroup taxon sampling across clades varies widely between studies (Extended Data Fig.
-1).
-The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed.
-More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here.
-Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
-1).
-Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig.
-XXXX).
-
-```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
-
-no_subsets = unlist(lapply(sequence_matrices, function(x) !grepl("only|_D", x@matrix_name)))
-
-taxon_rectangles = lapply(
-	sequence_matrices[no_subsets],
-	function ( sequence_matrix ){
-		clade_rects(sequence_matrix)
-	}
-) %>%
-bind_rows()
-
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
-
-clade_colors = c("#e31a1c","#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#a6cee3","#fdbf6f","#cab2d6", "#ff7f00")
-names( clade_colors ) = clades
-
-ggplot(data=taxon_rectangles) +
-	scale_x_continuous(name="Gene sampling") +
-	scale_y_continuous(name="Taxa sampling") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
-	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
- scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
-	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
-	facet_wrap( ~ matrix )
-
-```
-
-**Fig.
-XXtaxon_rectangles.** Matrix taxa and gene sampling.
-Each of the primary matrices considered here, color coded by taxon sampling.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-
-Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
-There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede].
-To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
-It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
 
@@
 
+```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
+
+no_subsets = unlist(lapply(sequence_matrices, function(x) !grepl("only", x@matrix_name)))
+
+taxon_rectangles = lapply(
+	sequence_matrices[no_subsets],
+	function ( sequence_matrix ){
+		clade_rects(sequence_matrix)
+	}
+) %>%
+bind_rows()
+
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
+clade_colors = c("#e31a1c","#fb9a99","#fb9a99","#1e90ff","#E69F00","#009E73","#1f78b4","#cab2d6","#c71585")
+names( clade_colors ) = clades
+
+taxon_rectangles= taxon_rectangles %>% filter(matrix!="Nosenko2013_nonribo_9187") %>% filter(matrix!="Whelan2017_strict") %>% filter(matrix!="Whelan2015_D1") %>% filter(matrix!="Whelan2015_D10") %>% filter(matrix!="Nosenko2013_ribo_11057")%>% filter(matrix!="Borowiec2015_Best108")
+
+ggplot(data=taxon_rectangles) +
+	scale_x_continuous(name="Gene sampling") +
+	scale_y_continuous(name="Taxa sampling") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
+	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
+ scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
+	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
+	facet_wrap( ~ matrix )
+
+	#matrix = factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
+
+```
+
+**Fig.
+XXtaxon_rectangles.** An overview of previous studies.
+(A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node).
+(B) Matrix taxa and gene sampling for each study.
+Each of the primary matrices considered here, color coded by taxon sampling.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+
+Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa.
+As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species.
+Ingroup taxon sampling across clades varies widely between studies (figure_taxon_rectangles).
+The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed.
+More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here.
+Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
+1).
+Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig.
+XXXX).
+
+Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
+There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede].
+To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
+Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+It has been suggested the inclusion of more distant related (e.g.
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
+
 ## Variation in published results
@@
 ```{r figure_support_analyses_published}
-aanalyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
+analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
 analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
@@
 
-x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
+x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT", "GTR+CAT"))
 p2 = analyses_new %>%
@@
 **Fig.
-XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalyses conducted in this study.
-(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (Table XX).
-(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (Supplementary Table XX).
+XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
+Each shape represents an phylogenetic analysis that has been analyzed by ML (designated by the circle) or BI inference (designated by the triangle).
+Analyses with a support of bootstrap values less than 90% or a posterior probability less than 0.95 are considered as unresolved.
+(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (related to Supplementary Table XX).
+(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (related to Supplementary Table XX).
 
@@
 5).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) or site ($\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
@@
 XX).
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models.
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutionary models.
 Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.

### 2020-05-25 manuscript/manuscript.rmd
more changes to the main figures and legends

@@
 Thus, it easily causes confusions with the increase number of studies.
-Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models.
+Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] under a panel of evolutionary models.
 We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
@@
 We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices).
-We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1).
+We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
+We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1).
 We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
 Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
-Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
+Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
 3B; Supplementary Table 3).
@@
 This has been addressed with CAT models.
-IQtree provides a new family of C60 models that also address site heterogeneity.
+IQ-TREE provides a new family of C60 models that also address site heterogeneity.
 Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared C60 results to CAT results for a subset of matrices to see if they give consistent results.
@@
 The Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and  IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
+For the matrices have conflicted results between Poisson-CAT and  IQ-TREE, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
 We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices.
@@
 5).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
@@
 The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
-Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices.
+Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices.
 These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
@@
 XX).
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutionary models.
+Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQ-TREE and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutionary models.
 Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
@@
 ```
-**Fig. XXsensitive.**
+**Fig.
+XXsensitive.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes.
+Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
+Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
+Support values are from the sensitive analyses are shown in Supplementary table XX.
 
@@
 
-#### Phylogenetic analyses in IQtree
+#### Phylogenetic analyses in IQ-TREE
 To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
@@
 site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQtree v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
 We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
@@
 Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C60 models in IQtree and Poisson-CAT models in PhyloBayes were conducted.
-The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQtree.
+For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson-CAT models in PhyloBayes were conducted.
+The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
 Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8.
@@
 We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
-The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
+For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
+The two different phylogenetic hypotheses passed to IQ-TREE (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
 The constraint trees were conducted in a customized R script.
-The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al.
+The numbers of genes and sites supporting each hypothesis were calculated with IQ-TREE output and Perl scripts from Shen et al.
 2017 [@shen2017contentious].
@@
 
-![](figures/Figure_Pairwise_heatmap_animal_root.png)
-**Extended Data Fig.
-XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance.
-(A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses.
-(B) Pairwise comparisons across genes that supported Porifera-sister in different analyses.
-The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and -1 is negative linear correlation.
-The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].
-
 ```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
@@
 **Extended Data Fig.
-XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice.
-(A).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix.
-(B).
-Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix.
-(C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix.
-The $\Delta$GLS values for the genes across each data matrix are provided in Supplementary Table 4.
+XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
+The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with a log-likelihood difference of two as a gene with strong (lnl > 2) or weak (lnl < 2) phylogenetic signal for each hypothesis (Smith et al.
+2020).
 

### 2020-05-26 manuscript/manuscript.rmd
add Whelan2017_strict.phy_Poisson_CAT70.bpdiff

@@
 
-To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework.
+To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized and systematic framework.
 Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Table 1).
@@
 
-The published analyses considered here differ in a variety of ways, including the model of molecular evolution, gene selection, ingroup species selection, and outgroup species selection.
+The published analyses considered here differ in a variety of ways, including the model of molecular evolution, gene selection, and ingroup and outgroup species selection.
 

### 2020-05-26 manuscript/manuscript_kernel.R
added analyses_sensitive tibble w/ phylobayes Poisson+CATn parameter sweep data

@@ -207,13 +207,9 @@ file_names_iqtree = list.files( path = trees_path_iqtree, pattern = iqtree_ext, 
 trees_iq = foreach( tree_file = file_names_iqtree ) %dopar%
 	parse_tree_iqtree( tree_file, taxonomy_reference )
 
-# read phylobayes sensitivity analyses
+# read phylobayes 
 trees_path_sensitive = "../trees_new/sensitive"
 pb_tree_ext = "\\.con\\.tre$"
-
-file_names_sensitive = list.files( path = trees_path_sensitive, pattern = pb_tree_ext, full.names = TRUE )
-trees_sensitive = foreach( tree_file = file_names_sensitive ) %dopar%
-  parse_tree_pb( tree_file, taxonomy_reference )
 
 # read pb trees
 trees_path_pb = "../trees_new/phylobayes"
@@ -222,6 +218,38 @@ file_names_pb = list.files( path = trees_path_pb, pattern = pb_tree_ext, full.na
 trees_pb = foreach( tree_file = file_names_pb ) %dopar%
            parse_tree_pb( tree_file, taxonomy_reference )
 
+# sensitivity analyses
+file_names_sensitive = list.files( path = trees_path_sensitive, pattern = pb_tree_ext, full.names = TRUE )
+trees_sensitive = foreach( tree_file = file_names_sensitive ) %dopar%
+  parse_tree_pb( tree_file, taxonomy_reference )
+
+# sensitivity tibble
+analyses_sensitive = lapply(
+  trees_sensitive,
+  function( tree ){
+    data.frame(
+      matrix = tree$matrix,
+      model = tree$model,
+      clade = tree$sampling,
+      support_ctenophora_sister = tree$ctenophora_sister * 100,
+      support_porifera_sister = tree$porifera_sister * 100,
+      stringsAsFactors = FALSE
+    )
+  }
+) %>%
+  bind_rows()
+
+analyses_sensitive$inference = rep( "Bayesian", length( trees_sensitive ) )
+analyses_sensitive$result = "Unresolved"
+analyses_sensitive$result[ (analyses_sensitive$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
+analyses_sensitive$result[ (analyses_sensitive$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
[…]

### 2020-06-17 manuscript/manuscript.rmd
update figures page1; page2; page4' page6

@@
 taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
-clade_colors = c("#e31a1c","#fb9a99","#fb9a99","#1e90ff","#E69F00","#009E73","#1f78b4","#cab2d6","#c71585")
-names( clade_colors ) = clades
+#clade_colors = c("#e31a1c","#fb9a99","#fb9a99","#1e90ff","#E69F00","#009E73","#1f78b4","#cab2d6","#c71585")
+
+clade_colors = c('#e6194b', '#f032e6', '#ffe119', '#4363d8', "#E69F00","#009E73","#82d1f4", '#e6beff', '#911eb4')
 
 taxon_rectangles= taxon_rectangles %>% filter(matrix!="Nosenko2013_nonribo_9187") %>% filter(matrix!="Whelan2017_strict") %>% filter(matrix!="Whelan2015_D1") %>% filter(matrix!="Whelan2015_D10") %>% filter(matrix!="Nosenko2013_ribo_11057")%>% filter(matrix!="Borowiec2015_Best108")
+
+x1= factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
 
@@
 	facet_wrap( ~ matrix )
-
-	#matrix = factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
 
@@
 analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
+
 x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
@@
 ``` {r sensitive, fig.width=8, fig.height=8}
-sensitive = read_tsv("../data_processed/tables/sensitive.tsv")
-Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )
-Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT60" , "Poisson_CAT90","Poisson_CAT120","Poisson_CAT150","Poisson_CAT180"))
-p1<-ggplot(Philippe2009, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+
+Philippe2009=analyses_sensitive %>% filter(matrix == "Philippe2009_only_choanozoa" )
+Philippe2009$model = factor(Philippe2009$model, levels=c("Poisson+CAT60" , "Poisson+CAT90","Poisson+CAT120","Poisson+CAT150","Poisson+CAT180"))
+p1<-ggplot(Philippe2009, aes(model))+
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
   scale_fill_manual(values=c("#E69F00", "#ebd196")) +
@@
   theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanozoa")
-
-Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_Holozoa")
-Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT360", "Poisson_CAT400","Poisson_CAT440","Poisson_CAT480"))
-p2<-ggplot(Philippe2009, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanimalia")
+
+Philippe2009=analyses_sensitive %>% filter(matrix == "Philippe2009_only_Holozoa")
+Philippe2009$model = factor(Philippe2009$model, levels=c("Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400","Poisson+CAT440","Poisson+CAT480"))
+p2<-ggplot(Philippe2009, aes(model))+
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
   scale_fill_manual(values=c("#E69F00", "#ebd196")) +
@@
 
-Whelan2017_full_only_choanozoa = sensitive %>% filter(Matrix == "Whelan2017_full_only_choanozoa")
-Whelan2017_full_only_choanozoa$Model = factor(Whelan2017_full_only_choanozoa$Model, levels=c("Poisson_CAT60" , "Poisson_CAT340","Poisson_CAT380","Poisson_CAT420","Poisson_CAT460"))
-p3<-ggplot(Whelan2017_full_only_choanozoa, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+Whelan2017_full_only_choanozoa = analyses_sensitive %>% filter(matrix == "Whelan2017_full_only_choanozoa")
+Whelan2017_full_only_choanozoa$model = factor(Whelan2017_full_only_choanozoa$model, levels=c("Poisson+CAT60" ,"Poisson+CAT340","Poisson+CAT380","Poisson+CAT420","Poisson+CAT460"))
+p3<-ggplot(Whelan2017_full_only_choanozoa, aes(model))+
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
   scale_fill_manual(values=c("#E69F00", "#ebd196")) +
@@
 
-Whelan2017_strict = sensitive %>% filter(Matrix == "Whelan2017_strict" )
-Whelan2017_strict$Model = factor(Whelan2017_strict$Model, levels=c("Poisson_CAT60" , "Poisson_CAT70","Poisson_CAT80","Poisson_CAT90","Poisson_CAT100","Poisson_CAT110", "Poisson_CAT120","Poisson_CAT150","Poisson_CAT180", "Poisson_CAT360"))
-p4<-ggplot(Whelan2017_strict, aes(Model))+
-  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
+Whelan2017_strict = analyses_sensitive %>% filter(matrix == "Whelan2017_strict" )
+Whelan2017_strict$model = factor(Whelan2017_strict$model, levels=c("Poisson+CAT60" , "Poisson+CAT70","Poisson+CAT80","Poisson+CAT90","Poisson+CAT100","Poisson+CAT110", "Poisson+CAT120","Poisson+CAT150","Poisson+CAT180", "Poisson+CAT360"))
+p4<-ggplot(Whelan2017_strict, aes(model))+
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
   scale_fill_manual(values=c("#E69F00", "#ebd196")) +
@@
 ## Extended Figures
-
 ```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
-matrix_summary = merge(matrix_summary,n_ribo,by="matrix")
-
-p1 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) +  	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
-
-matrix_summary=matrix_summary %>%
-  mutate(matrix=recode(matrix,`NA`="Simion2017"))
-
-matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-
-p2 = ggplot( matrix_summary, aes(x=matrix, y=n_busco_partitions/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Busco genes") + ylim(0, 100)
-
-p3 = ggplot( matrix_summary ) +
-	geom_point( aes(x=n_partitions, y=n, col=manuscript) ) +
-	geom_abline(slope=1,intercept=0) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Number of gene partitions", y = "Number of Ribosomal genes")
-	coord_fixed()
-
-p4 = ggplot( matrix_summary, aes(x=matrix, y=n/n_partitions*100, fill=manuscript) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of Ribosomal genes")+ ylim(0, 100)
-
- ggarrange(p1, p2,p3, p4,
-                    labels = c("A", "B", "C", "D"),
-                    ncol = 2, nrow = 2,common.legend = TRUE, legend = "right", align = "hv")
+
+ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
+busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+
+ribo$matrix = factor( ribo$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+
+p1 = ggplot(ribo, aes(x=matrix, y=number_of_genes, fill=gene_category) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Number of genes")  + scale_fill_manual(values=c("grey", "black"), labels=c("Non-ribosomal genes", "Ribosomal genes"))
+
+busco$matrix = factor( busco $matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
+busco$gene_category = factor(busco$gene_category, levels=c("non_busco", "busco"))
+p2 = ggplot(busco, aes(x=matrix, y=number_of_genes, fill=gene_category) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Number of genes")  + scale_fill_manual(values=c("grey", "black"), labels=c("Non-BUSCO genes", "BUSCO genes"))
+
+ ggarrange(p1, p2, labels = c("A", "B"))
 

### 2020-06-18 manuscript/manuscript.rmd
update figure in page 3

@@
 
-x1= factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
+taxon_rectangles$matrix= factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
 
@@
 	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
-	facet_wrap( ~ matrix )
+	facet_wrap( ~ matrix ) + theme_classic()
 

### 2020-06-23 manuscript/manuscript.rmd
revisions focused up until line 130, numbered figures

@@
 
-There has been considerable recent debate about the deepest split in the animal tree of life.
+There has been considerable recent debate about the placement of the root in the animal tree of life.
 This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
-Some previous analyses found support for the first split being between comb jellies and all other animals (the Ctenophora-sister hypothesis), while others found support the first split being between sponges and all other animals (Porifera-sister).
-These studies differed in a many respects, including the genes and species selected, molecular evolution models, and analysis software.
-Here we systematically explore this question under consistent conditions by synthesizing data and results from XX previous phylogenomic analyses and performing new standardized analyses.
+Some phylogenomic analyses find support for a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others found support for a split between comb jellies and all other animals (Ctenophora-sister).
+These analyses differed in a many respects, including the genes and species considered, molecular evolution models, and software.
+Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses.
 Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
@@
 By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant animal relative.
-Since then many more studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Table 1).
-As the problem has become better characterized, it has become an interesting test-case for addressing difficult phylogenetic questions, which has attracted methodological interest well beyond investigators concerned with early animal evolution.
+The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative.
+Since then many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1).
 These studies have taken a wide variety of approaches and results have varied across these different analyses.
-The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result.
+The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result.
 But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses.
-This variance is often confounded, for example when a particular model is only available in one software package.
-
-To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized and systematic framework.
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Table 1).
+This mthodological variance is often confounded, for example when a particular model is only available in one software package.
+
+To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework.
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1).
 These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.
@@
 **Fig.
-XXFigure_overview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
+1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
 (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
@@
 
-Models of molecular evolution describe how probable different evolutionary changes are, and can have a major impact on the topology of inferred phylogenies.
-Models have several components that each consider different aspects of the evolutionary process.
-The models used in the studies considered here differ according to three primary components: the exchangeability matrix $E$ , rates of evolution, and amino acid equilibrium frequencies $\Pi$.
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR).
-The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites.
-Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
-Models of molecular evolution are assembled by selecting different options for all these different components.
-The models that are applied in practice are heavily influenced by computational costs, as well as convention.
-For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices.
-LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
-This is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
+The models used in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
+The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another.
+The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empically estimated rates (WAG or LG), or all independently estimated from the data (GTR).
+The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes.
+The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
+The analyses differ in whether they take a homongenous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites.
+Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
+The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention.
+For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices.
+Model effect is further confused by the abbreviations that are used for models.
+Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT).
-Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here.
-If there are no modifiers, then it is implied that site homogeneous models are used.
-A detailed description of each component is provided in Supplementary Results.
-
-High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1).
-Studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices.
-These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details).
+To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT).
+If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone.
+Gamma-rate heterogeneity, a scalar that acccomodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
+
+High-throughput sequencing allows investigators to readilly assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1).
+Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
+These take different approaches to sequence comparison, filtering, and orthology assessment.
 As a result, the genes selected for analysis varies widely between studies (Fig.
-XX).
-This reflects the different methods used to collect sequence data and different criteria for gene selection.
-There are several ways to summarize and assess this variation in gene sampling.
-We annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1).
+2, Extended Data Fig.
+1).
+To explore this variation in gene sampling, we annotated each gene with against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table XX).
 These annotations show XXX.
-We identified number of ribosomal RNA genes in each data matrix (Extended Data Fig.
-2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig.
-2D).
+We identified the number of ribosomal protein genes in each data matrix (Extended Data Fig.
+2A).
 We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig.
-2A).
-The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are broadly shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco].
-The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]).
-We found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20% (Extended Data Fig.
 2B).
-This is an indication that gene sampling criterial are not only based on breadth of sampling, but also on other criteria such as features of molecular evolution and sampling effects such as sifferences in expression level.
+The animal BUSCO dataset contains 978 genes that tend to occur as a single copy used to assess the completeness of genome assemblies [@waterhouse2017busco].
+All matrices have than 50% of the BUSCO genes, and several less than 20%.
+The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenetomic data are based on transcriptome data) and features of molecular evolution.
 
@@
 **Fig.
-XXtaxon_rectangles.** An overview of previous studies.
+2.** An overview of previous studies.
 (A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node).
@@
 
-Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa.
-As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species.
-Ingroup taxon sampling across clades varies widely between studies (figure_taxon_rectangles).
-The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed.
-More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here.
+Ingroup taxon sampling across clades varies widely between studies (Fig.
+2, Extended Data Fig.
+1).
+Sampling of ingroup taxa (animals) was initially biased toward Bilateria.
+Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups.
 Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
@@
 Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig.
-XXXX).
+1).
 
 Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
-There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede].
+There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation needes].
 To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
-It has been suggested the inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
 
@@
 **Fig.
-XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
+3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
 Each shape represents an phylogenetic analysis that has been analyzed by ML (designated by the circle) or BI inference (designated by the triangle).
@@
 3A and Supplementary Table 2).
-Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1).
+Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Extended Data Table 1).
 Moreover, three studies are secondary studies without any new data matrices.
@@
 
+Inclusion of more distant related (e.g.
+Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
+
 ## New standardized analyses of published matrices
@@
 
-Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1).
+Here we reanalyzed 17 matrices from 13 studies that are used as main conclusions from their original paper (Extended Data Table 1).
 We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
@@
 **Fig.
-XXsensitive.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes.
+4.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes.
 Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
@@
 
-We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
+We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository.
 We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
@@
 #### Phylogenetic analyses in IQ-TREE
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1).
 Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
@@
 
-**Extended Data Table 1.** TTable 1. Overview of data matrices used in this study.
+**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
 ```{r table_study_summary}
@@
 ## Extended Figures
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
+	unlist(recursive=FALSE) %>%
+	bind_rows()
+
+ggplot(data=matrix_overlap_rectangles) +
+	scale_x_continuous(name="Genes") +
+	scale_y_continuous(name="Species") +
+	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
+	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
+
+```
+
+**Extended Data Fig.
+1.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
+
 ```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
@@
 **Extended Data Fig.
-XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
 (A).
@@
 
-```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
-	unlist(recursive=FALSE) %>%
-	bind_rows()
-
-ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="Genes") +
-	scale_y_continuous(name="Species") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
-
-```
-
-**Extended Data Fig.
-XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
 ```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
@@
 **Extended Data Fig.
-XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
+3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
@@
 **Extended Data Fig.
-XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here.
+4.** (A) The two alternative hypotheses for deep animal relationships considered here.
 Relationships that are not part of these hypotheses are shown as unresolved polytomies.
@@
 **Extended Data Fig.
-XXrecoding_pp.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 **Extended Data Fig.
-XXrecoding_summary.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
 Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).

### 2020-06-24 manuscript/manuscript.rmd
rewrote code for figure 3, some text edits

@@
 
-## Variation in models and sampling across published studies
+## Variation in models and sampling across published analyses
 
@@
 Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
-These take different approaches to sequence comparison, filtering, and orthology assessment.
-As a result, the genes selected for analysis varies widely between studies (Fig.
+As a result, the genes selected for analysis vary widely (Fig.
 2, Extended Data Fig.
@@
 These annotations show XXX.
-We identified the number of ribosomal protein genes in each data matrix (Extended Data Fig.
-2A).
-We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig.
-2B).
+We identified the ribosomal protein and BUSCO genes [@waterhouse2017busco] in each data matrix (Extended Data Fig.
+2).
 The animal BUSCO dataset contains 978 genes that tend to occur as a single copy used to assess the completeness of genome assemblies [@waterhouse2017busco].
@@
 
-```{r figure_taxon_rectangles, fig.height = 18, fig.width = 20 }
+```{r figure_taxon_rectangles, fig.height = 7.2, fig.width = 7.2 }
 
@@
 
-## Variation in published results
-
-```{r figure_support_analyses_published}
-analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
-analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
-
-x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
-p1 = analyses_published_filter %>%
-	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
+## Variation in results across published analyses
+
+```{r figure_support_analyses, fig.width=7.2, fig.height=3.6}
+
+analyses_published_processed = 
+  analyses_published %>%
+  dplyr::filter( !grepl( 'Recoding', model_combined ) ) %>%
+  dplyr::rename( model = model_combined ) %>%
+  select( model, clade, final, inference ) %>%
+  dplyr::rename( result = final ) %>%
+  mutate( status = "published" )
+
+analyses_new_processed = 
+  analyses_new %>%
+  mutate(model = factor(model)) %>%
+  select( model, clade, result, inference ) %>%
+  mutate( status = "new" )
+
+analyses_processed = 
+  rbind( analyses_published_processed, analyses_new_processed ) %>%
+  # mutate(software=recode( software, `PhyloBayes MPI`="PhyloBayes") ) %>%
+  mutate(model=recode(model,
+                    `GTR20`="GTR",
+                    `CAT+F81`="Poisson+CAT",
+                    `Poisson + CAT`="Poisson+CAT",
+                    `GTR + CAT`="GTR+CAT")) %>%
+  mutate( 
+    model = factor( model, 
+      levels=c("WAG", "LG", "GTR", "data partitioning", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT", "GTR+CAT") )) %>%
+  mutate( status = factor(status, levels = c("published", "new")))
+
+p = analyses_processed %>%
+	ggplot( aes(x=model, y=clade, col=result, shape=inference) ) +
+    facet_wrap( ~ status ) +
 		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
 		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-   xlab("Model complexity") + ylab("Clade")
-
-analyses_new=analyses_new %>%
-mutate(model=recode(model,
-                    `GTR20`="GTR",
-                    `CAT+F81`="Poisson+CAT"))
-
-x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT", "GTR+CAT"))
-p2 = analyses_new %>%
-	ggplot( aes(x=x2, y=clade, col=result, shape=inference) ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
-		scale_colour_manual(values = result_colors) +
-		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
-       xlab("Model complexity") + ylab("Clade")
-
-ggarrange(p1, p2, ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")
+		theme_classic() + 
+    theme(axis.text.x = element_text(angle=45, hjust=1))+
+    xlab("Model") + 
+    ylab("Clade")
+
+p
 
 ```
+
 **Fig.
@@
 Analyses with a support of bootstrap values less than 90% or a posterior probability less than 0.95 are considered as unresolved.
-(A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (related to Supplementary Table XX).
-(B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (related to Supplementary Table XX).
-
-Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+(A) Analyses transcribed from the literature (related to Supplementary Table XX).
+(B) New phylogenomic analyses were conducted in this study (related to Supplementary Table XX).
+
+We parsed the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
 3A and Supplementary Table 2).
-Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Extended Data Table 1).
-Moreover, three studies are secondary studies without any new data matrices.
-Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used.
-The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015.
-It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes.
-The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices.
+Among these studies, the main conclusions of five studies strongly favor of Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
+Three studies are based entirely on previously published data, and the remainder add data for one or more species.
+
+All published analyses that use models with site-homogeneous equilibrium frequencies (*i.e.*, models that do not include CAT) strongly support Ctenophora-sister, regardless of other factors such as outgroup choice (Fig.
+3A).
+Some analyses that exclude more distant outgroups and use models with site-heterogeneious CAT equilibrium frequencies support Porifera-sister.
+Because CAT is so parameter rich, many analyses that use a CAT model do not converge, and some of the unconvered analyses have an unresolved animal root.
+The strong support of Porifera-sister can only be recovered from analyses analyses of protein sequences using Poisson+CAT.
 Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
@@
 
-``` {r sensitive, fig.width=8, fig.height=8}
+``` {r sensitive, fig.width=7.2, fig.height=7.2}
 

### 2020-06-25 manuscript/manuscript.rmd
small tweak

@@
 
-We parsed the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+We parsed `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
 3A and Supplementary Table 2).
-Among these studies, the main conclusions of five studies strongly favor of Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
+The conclusions of five studies strongly favor of Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
 Three studies are based entirely on previously published data, and the remainder add data for one or more species.
@@
 Because CAT is so parameter rich, many analyses that use a CAT model do not converge, and some of the unconvered analyses have an unresolved animal root.
-The strong support of Porifera-sister can only be recovered from analyses analyses of protein sequences using Poisson+CAT.
-Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
+Although model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
 3A).

### 2020-06-25 manuscript/manuscript.rmd
made pass through Variation in results across published analyses

@@
   dplyr::rename( model = model_combined ) %>%
-  select( model, clade, final, inference ) %>%
+  select( model, clade, final, inference, manuscript, matrix ) %>%
   dplyr::rename( result = final ) %>%
@@
   mutate(model = factor(model)) %>%
-  select( model, clade, result, inference ) %>%
+  select( model, clade, result, inference, matrix ) %>%
+  mutate( manuscript = NA ) %>%
   mutate( status = "new" )
@@
 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
-Each shape represents an phylogenetic analysis that has been analyzed by ML (designated by the circle) or BI inference (designated by the triangle).
-Analyses with a support of bootstrap values less than 90% or a posterior probability less than 0.95 are considered as unresolved.
+Each point represents a phylogenetic Analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
+The clades are orgainized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis.
 (A) Analyses transcribed from the literature (related to Supplementary Table XX).
-(B) New phylogenomic analyses were conducted in this study (related to Supplementary Table XX).
-
-We parsed `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig.
+(B) New phylogenomic analyses conducted in this study (related to Supplementary Table XX).
+
+We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig.
 3A and Supplementary Table 2).
@@
 
-All published analyses that use models with site-homogeneous equilibrium frequencies (*i.e.*, models that do not include CAT) strongly support Ctenophora-sister, regardless of other factors such as outgroup choice (Fig.
-3A).
-Some analyses that exclude more distant outgroups and use models with site-heterogeneious CAT equilibrium frequencies support Porifera-sister.
-Because CAT is so parameter rich, many analyses that use a CAT model do not converge, and some of the unconvered analyses have an unresolved animal root.
-Although model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig.
-3A).
-Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.
-
-Inclusion of more distant related (e.g.
-Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].
+As others have noted (XX), our summary of previous phylogenetic analyses (Fig.
+3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommondate site-heterogeneous equilibrium frequencies with CAT.
+The only analyses that support Porifera-sister ahveh reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
+Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because CAT is so parameter rich, many analyses that use a model with CAT do not converge.
+Some of the unconvered analyses have an unresolved animal root.
+The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.
 
@@
 Here we reanalyzed 17 matrices from 13 studies that are used as main conclusions from their original paper (Extended Data Table 1).
-We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C).
+We also progressively trimmed each data matrix with three level of outgroup choice if applicable, original outgroup sampling permitting (Fig 1C).
 We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices).

### 2020-06-26 manuscript/manuscript.rmd
some chunks that are not validated with code to verify assertions in text

@@
 
-As others have noted (XX), our summary of previous phylogenetic analyses (Fig.
+```{r eval=FALSE}
+
+# Code to double check and explore some of the assertions in the text
+
+# Simion2017 is the only Holozoa analysis to support Porifera-sister
+analyses_processed %>% filter(status=="published") %>% filter(clade=="Holozoa") %>% filter(result=="Porifera-sister")
+
+# Check out the CAT analyses that supported Ctenophora sister
+analyses_processed %>% filter(status=="published") %>% filter(result=="Ctenophora-sister") %>% filter( grepl("CAT", model))
+
+```
+
+Our summary of previous phylogenetic analyses (Fig.
 3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommondate site-heterogeneous equilibrium frequencies with CAT.
+This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview.
 The only analyses that support Porifera-sister ahveh reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
@@
 
-One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
-Thus, it easily causes confusions with the increase number of studies.
+One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across analyses.
+This extensive variation can make it difficult to identify the primary factors that lead to different results.
 Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig.
+We selected this tool because it has greater model flexibility than other maximum likelihood (ML) tools [@zhou2017evaluating] (Fig.
 3B; Supplementary Table 3).
-Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.
-
-Here we reanalyzed 17 matrices from 13 studies that are used as main conclusions from their original paper (Extended Data Table 1).
-We also progressively trimmed each data matrix with three level of outgroup choice if applicable, original outgroup sampling permitting (Fig 1C).
-We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
+Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.
+There are some implementation differences.
+For example, CAT estimates the number of distinct site categories, the assigment of sites to those categories, and the equilibrium frequencies of each of those categories, while the C models specify the number of categories *a priori* (the C10 model as 10 categories, etc...).
+
+```{r eval=FALSE}
+
+# Code to double check some of the assertions in the text
+
+# See results (and other fields) for new Cxx analyses, all are Ctenophora-sister with exception of Moroz2014_3d and Nosenko matrices
+analyses_processed %>% filter(status=="new") %>% filter( grepl("C\\d\\d",model) ) %>% print(n=40)
+
+```
+
+For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis.
+For each of these matrices, we progressively trimmed taxon sampling of to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
+This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.
+Each of these was then analyzed using several models.
+We used modelfinder to identify the
+
+We tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
 We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1).
 We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
-Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
+Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling.
 Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.

### 2020-06-26 manuscript/manuscript.rmd
linted code, did not rerun kernel

@@
   warning = FALSE,
-	echo=FALSE,
-	dpi=300,
-	cache=FALSE
+  echo = FALSE,
+  dpi = 300,
+  cache = FALSE
   )
@@
   library( gtable )
+
   theme_set(theme_pubr())
@@
   # A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
-	cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+  cbPalette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
@@
 
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
-taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"
-#clade_colors = c("#e31a1c","#fb9a99","#fb9a99","#1e90ff","#E69F00","#009E73","#1f78b4","#cab2d6","#c71585")
-
-clade_colors = c('#e6194b', '#f032e6', '#ffe119', '#4363d8', "#E69F00","#009E73","#82d1f4", '#e6beff', '#911eb4')
-
-taxon_rectangles= taxon_rectangles %>% filter(matrix!="Nosenko2013_nonribo_9187") %>% filter(matrix!="Whelan2017_strict") %>% filter(matrix!="Whelan2015_D1") %>% filter(matrix!="Whelan2015_D10") %>% filter(matrix!="Nosenko2013_ribo_11057")%>% filter(matrix!="Borowiec2015_Best108")
-
-taxon_rectangles$matrix= factor(taxon_rectangles$matrix, levels=c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080" , "Chang2015", "Whelan2017_full", "Simion2017"))
-
-ggplot(data=taxon_rectangles) +
-	scale_x_continuous(name="Gene sampling") +
-	scale_y_continuous(name="Taxa sampling") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
-	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
- scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
-	#scale_fill_discrete(guide = guide_legend(reverse=TRUE, legend = "bottom right")) +
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] = "Simion2017"
+taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] = "Hejnol2009"
+
+clade_colors = c("#e6194b", "#f032e6", "#ffe119", "#4363d8", "#E69F00", "#009E73", "#82d1f4", "#e6beff", "#911eb4")
+
+taxon_rectangles = 
+  taxon_rectangles %>% 
+  filter(matrix != "Nosenko2013_nonribo_9187") %>% 
+  filter(matrix != "Whelan2017_strict") %>% 
+  filter(matrix != "Whelan2015_D1") %>% 
+  filter(matrix != "Whelan2015_D10") %>% 
+  filter(matrix != "Nosenko2013_ribo_11057") %>% 
+  filter(matrix != "Borowiec2015_Best108")
+
+taxon_rectangles$matrix = 
+  factor( 
+    taxon_rectangles$matrix, 
+    levels = c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est", 
+             "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080", "Chang2015", 
+             "Whelan2017_full", "Simion2017"))
+
+taxon_rectangles %>%
+  ggplot() +
+    scale_x_continuous(name = "Gene sampling") +
+    scale_y_continuous(name = "Taxa sampling") +
+    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = clade )) +
+    scale_fill_manual(values = clade_colors, guide = guide_legend(reverse = TRUE)) +
     facet_wrap( ~ matrix ) + theme_classic()
@@
   analyses_published %>%
-  dplyr::filter( !grepl( 'Recoding', model_combined ) ) %>%
+  dplyr::filter( !grepl( "Recoding", model_combined ) ) %>%
   dplyr::rename( model = model_combined ) %>%
@@
   # mutate(software=recode( software, `PhyloBayes MPI`="PhyloBayes") ) %>%
-  mutate(model=recode(model,
-                    `GTR20`="GTR",
-                    `CAT+F81`="Poisson+CAT",
-                    `Poisson + CAT`="Poisson+CAT",
-                    `GTR + CAT`="GTR+CAT")) %>%
+  mutate(model = recode(model,
+                    `GTR20` = "GTR",
+                    `CAT+F81` = "Poisson+CAT",
+                    `Poisson + CAT` = "Poisson+CAT",
+                    `GTR + CAT` = "GTR+CAT")) %>%
   mutate( 
     model = factor( model, 
-      levels=c("WAG", "LG", "GTR", "data partitioning", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT", "GTR+CAT") )) %>%
+      levels = c("WAG", "LG", "GTR", "data partitioning", 
+                 "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT", "GTR+CAT") )) %>%
   mutate( status = factor(status, levels = c("published", "new")))
@@
 p = analyses_processed %>%
-	ggplot( aes(x=model, y=clade, col=result, shape=inference) ) +
+  ggplot( aes(x = model, y = clade, col = result, shape = inference) ) +
     facet_wrap( ~ status ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
+    geom_jitter( width = 0.20, height = 0.20, alpha = 0.7, size = 2 ) +
     scale_colour_manual(values = result_colors) +
     theme_classic() + 
-    theme(axis.text.x = element_text(angle=45, hjust=1))+
+    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     xlab("Model") + 
@@
 # Simion2017 is the only Holozoa analysis to support Porifera-sister
-analyses_processed %>% filter(status=="published") %>% filter(clade=="Holozoa") %>% filter(result=="Porifera-sister")
+analyses_processed %>% 
+  filter(status == "published") %>% 
+  filter(clade == "Holozoa") %>% 
+  filter(result == "Porifera-sister")
 
 # Check out the CAT analyses that supported Ctenophora sister
-analyses_processed %>% filter(status=="published") %>% filter(result=="Ctenophora-sister") %>% filter( grepl("CAT", model))
+analyses_processed %>% 
+  filter(status == "published") %>% 
+  filter(result == "Ctenophora-sister") %>% 
+  filter( grepl("CAT", model))
+
+analyses_processed %>% 
+  filter(status == "published") %>% 
+  filter(result == "Ctenophora-sister") %>% 
+  filter( grepl("CAT", model)) %>% 
+  filter( clade != "Opisthokonta")
 
@@
 # See results (and other fields) for new Cxx analyses, all are Ctenophora-sister with exception of Moroz2014_3d and Nosenko matrices
-analyses_processed %>% filter(status=="new") %>% filter( grepl("C\\d\\d",model) ) %>% print(n=40)
+analyses_processed %>% 
+  filter(status == "new") %>% 
+  filter( grepl("C\\d\\d", model) ) %>% 
+  print(n = 40)
 
@@
 
-Philippe2009=analyses_sensitive %>% filter(matrix == "Philippe2009_only_choanozoa" )
-Philippe2009$model = factor(Philippe2009$model, levels=c("Poisson+CAT60" , "Poisson+CAT90","Poisson+CAT120","Poisson+CAT150","Poisson+CAT180"))
-p1<-ggplot(Philippe2009, aes(model))+
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+Philippe2009 = 
+  analyses_sensitive %>% 
+  filter(matrix == "Philippe2009_only_choanozoa" )
+
+Philippe2009$model = 
+  factor(
+    Philippe2009$model, 
+    levels = c("Poisson+CAT60", "Poisson+CAT90", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180"))
+
+p1 = Philippe2009 %>% 
+  ggplot( aes(model) ) +
+  geom_line( aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1 ), color = ("#E69F00")) + 
+  geom_point( aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) + 
+  geom_line( aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) + 
+  geom_point( aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) + 
   theme_classic() + 
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanimalia")
-
-Philippe2009=analyses_sensitive %>% filter(matrix == "Philippe2009_only_Holozoa")
-Philippe2009$model = factor(Philippe2009$model, levels=c("Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400","Poisson+CAT440","Poisson+CAT480"))
-p2<-ggplot(Philippe2009, aes(model))+
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  labs(x = "Number of CAT substitutional categories", y = "Support") + 
+  ggtitle("Philippe2009_Choanimalia")
+
+Philippe2009 = 
+  analyses_sensitive %>% 
+  filter(matrix == "Philippe2009_only_Holozoa")
+
+Philippe2009$model = 
+  factor(
+    Philippe2009$model, 
+    levels = c("Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400", "Poisson+CAT440", "Poisson+CAT480"))
+
+p2 = Philippe2009 %>% 
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) + 
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) + 
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
   theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Holozoa")
-
-Whelan2017_full_only_choanozoa = analyses_sensitive %>% filter(matrix == "Whelan2017_full_only_choanozoa")
-Whelan2017_full_only_choanozoa$model = factor(Whelan2017_full_only_choanozoa$model, levels=c("Poisson+CAT60" ,"Poisson+CAT340","Poisson+CAT380","Poisson+CAT420","Poisson+CAT460"))
-p3<-ggplot(Whelan2017_full_only_choanozoa, aes(model))+
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") + 
+  ggtitle("Philippe2009_Holozoa")
+
+Whelan2017_full_only_choanozoa = 
+  analyses_sensitive %>% 
+  filter(matrix == "Whelan2017_full_only_choanozoa")
+
+Whelan2017_full_only_choanozoa$model = 
+  factor( 
+    Whelan2017_full_only_choanozoa$model, 
+    levels = c("Poisson+CAT60", "Poisson+CAT340", "Poisson+CAT380", "Poisson+CAT420", "Poisson+CAT460"))
+
+p3 = Whelan2017_full_only_choanozoa %>% 
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) + 
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) + 
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
   theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Whelan2017_full_Choanozoa")
-
-Whelan2017_strict = analyses_sensitive %>% filter(matrix == "Whelan2017_strict" )
-Whelan2017_strict$model = factor(Whelan2017_strict$model, levels=c("Poisson+CAT60" , "Poisson+CAT70","Poisson+CAT80","Poisson+CAT90","Poisson+CAT100","Poisson+CAT110", "Poisson+CAT120","Poisson+CAT150","Poisson+CAT180", "Poisson+CAT360"))
-p4<-ggplot(Whelan2017_strict, aes(model))+
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `support_ctenophora_sister` ), color=("#E69F00"))+
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `support_porifera_sister` ), color=("#009E73"))+
-  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") + 
+  ggtitle("Whelan2017_full_Choanozoa")
+
+Whelan2017_strict = 
+  analyses_sensitive %>% 
+  filter(matrix == "Whelan2017_strict" )
+
+Whelan2017_strict$model = 
+  factor( 
+    Whelan2017_strict$model, 
+    levels = c("Poisson+CAT60", "Poisson+CAT70", "Poisson+CAT80", "Poisson+CAT90", "Poisson+CAT100",
+             "Poisson+CAT110", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180", "Poisson+CAT360"))
+
+p4 = Whelan2017_strict %>% 
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) + 
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) + 
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
   theme_classic() +
-  theme(axis.text.x = element_text(angle=45, hjust=1)) +
-  labs(x ="Number of CAT substitutional categories", y = "Support")+ ggtitle("Whelan2017_strict")
-
-ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
-          p3,p4, labels = c("C", "D"))
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") + 
+  ggtitle("Whelan2017_strict")
+
+ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")), nrow = 3,
+          p3, p4, labels = c("C", "D"))
 ```
@@
 analyses_new %>%
-	filter(modelfinder==TRUE) %>%
+  filter(modelfinder == TRUE) %>%
   select( matrix, clade, result, model_summary ) %>%
@@
 matrix_overlap_rectangles =
-	lapply(sequence_matrices[no_subsets], function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x,y))) %>%
-	unlist(recursive=FALSE) %>%
+  lapply( 
+    sequence_matrices[no_subsets], 
+    function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x, y))) %>%
+  unlist(recursive = FALSE) %>%
   bind_rows()
 
-ggplot(data=matrix_overlap_rectangles) +
-	scale_x_continuous(name="Genes") +
-	scale_y_continuous(name="Species") +
-	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=MSA ), alpha=0.5) +
-	facet_grid( matrix_1 ~ matrix_2, scales="free" ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle = 0), strip.text.x = element_text(angle = 60))
+matrix_overlap_rectangles %>% 
+  ggplot( ) +
+    scale_x_continuous(name = "Genes") +
+    scale_y_continuous(name = "Species") +
+    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
+    facet_grid( matrix_1 ~ matrix_2, scales = "free" ) +
+    theme( axis.text.x = element_text(angle = 45, hjust = 1), 
+           strip.text.y = element_text(angle = 0), 
+           strip.text.x = element_text(angle = 60))
 
@@
 
-ribo$matrix = factor( ribo$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-
-p1 = ggplot(ribo, aes(x=matrix, y=number_of_genes, fill=gene_category) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Number of genes")  + scale_fill_manual(values=c("grey", "black"), labels=c("Non-ribosomal genes", "Ribosomal genes"))
-
-busco$matrix = factor( busco $matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", "Whelan2017_full", "Whelan2017_strict") )
-busco$gene_category = factor(busco$gene_category, levels=c("non_busco", "busco"))
-p2 = ggplot(busco, aes(x=matrix, y=number_of_genes, fill=gene_category) ) +geom_bar(stat = "identity")  + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Number of genes")  + scale_fill_manual(values=c("grey", "black"), labels=c("Non-BUSCO genes", "BUSCO genes"))
+ribo$matrix = 
+  factor( 
+    ribo$matrix, 
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057", 
+             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108", 
+             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", 
+             "Whelan2017_full", "Whelan2017_strict") )
+
+p1 = ribo %>% 
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) + 
+  geom_bar(stat = "identity") + 
+  theme_classic() + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  labs(x = "Matrix", y = "Number of genes") + 
+  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
+
+busco$matrix = 
+  factor( 
+    busco$matrix, 
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057", 
+               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108", 
+               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017", 
+               "Whelan2017_full", "Whelan2017_strict") )
+
+busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
+
+p2 = busco %>% 
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) + 
+  geom_bar(stat = "identity") + 
+  theme_classic() + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  labs(x = "Matrix", y = "Number of genes") + 
+  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
 
@@
 ```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-# fig5_bars <- au_tests %>%
+
 au_tests %>%
@@
   mutate(percent = prop.table(n)) %>%
-
-ggplot(aes(x = Matrix, y = percent, fill=result)) +
+  ggplot(aes(x = Matrix, y = percent, fill = result)) +
     geom_bar(stat = "identity") +
-  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73")) +
+    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) + 
     coord_flip() +
-  theme(axis.text.y=element_blank(), axis.title.y=element_blank())
-
-#fig5_tbl <- au_tests %>%
-#select(c(manuscript, matrix, model)) %>%
-#distinct()
-
-#fig_5_ylabs_grob <- tableGrob(fig5_tbl)
-#fig_5_ylabs$heights = unit(rep(1/nrow(fig_5_ylabs), nrow(fig_5_ylabs)), "npc")
-#fig_5_ylabs$widths = unit.pmax(fig_5_ylabs$widths, unit(2, "lines"))
-
-#fig5_bars_grob <- ggplotGrob(fig5_bars)
-#fig5 <- gtable_add_cols(fig5, sum(fig_5_ylabs$widths), pos=0)
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=ncol(fig5), b=6, r=ncol(fig5))
-#fig5 <- gtable_add_grob(fig5, grobs = fig_5_ylabs, t=6, l=1, b=6, r=ncol(fig5))
-
-#grob_grid <- matrix(list(fig_5_ylabs_grob, fig5_bars_grob), nrow = 1)
-# https://cran.r-project.org/web/packages/gridExtra/vignettes/gtable.html
-
-# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
-#                    widths = unit(c(4, 4), "cm"),
-#                    heights = unit(c(2, 5), c("in", "lines")))
-#
-# grid.newpage()
-# grid.draw(fig5)
+    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
+
 ```

### 2020-06-26 manuscript/functions.R

@@ -31,14 +31,24 @@ setClass(
 #' @param partition_map_global A dataframe with columns manuscript, matrix, partition, global_partition_id that maps the locally defined partition to a global partition
 #' @return A PartitionedMultipleAlignment object
 #' @export
-PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, taxon_map_global=NULL, partition_map_global=NULL, manuscript_name="", matrix_name="", alignment_format="phylip" ) {
-  object = readAAMultipleAlignment( filepath = alignment_file, format=alignment_format )
+PartitionedMultipleAlignment = function( 
+  alignment_file, 
+  partition_file=NULL, 
+  taxon_map_global=NULL, 
+  partition_map_global=NULL, 
+  manuscript_name="", 
+  matrix_name="", 
+  alignment_format="phylip" ) {
+  
+  object = readAAMultipleAlignment( 
+    filepath = alignment_file, 
+    format = alignment_format )
   class(object) = "PartitionedMultipleAlignment"
   
   object@manuscript_name = manuscript_name
   object@matrix_name = matrix_name
   
-  conn = file( partition_file, open="r")
+  conn = file( partition_file, open = "r")
   lines = readLines( conn )
   lines = str_replace( lines, "charset", "CHARSET" )
   lines = lines[ grepl("CHARSET", lines) ]
@@ -50,9 +60,9 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
       function( line ){
         fields = str_match(line, "CHARSET (.+?)\\s*=\\s*(\\d+)\\s*-\\s*(\\d+)")
         D = data.frame( 
-          partition=fields[2], 
-          start=as.integer(fields[3]), 
-          stop=as.integer(fields[4]),
+          partition = fields[2], 
+          start = as.integer(fields[3]), 
+          stop = as.integer(fields[4]),
           stringsAsFactors = FALSE
         )
         return( D )
@@ -60,24 +70,34 @@ PartitionedMultipleAlignment = function( alignment_file, partition_file=NULL, ta
     ) %>%
     bind_rows()
   
-  if( ! is.null(partition_map_global) ){
+  if ( ! is.null(partition_map_global) ){
     
     partitions %<>% 
[…]

### 2020-06-26 manuscript/manuscript_kernel.R

@@ -17,14 +17,12 @@ library( magrittr )
 library( igraph )
 library( doParallel )
 
-setwd("/animal_root/manuscript")
-
 source( "functions.R" )
 
 # Set system computational parameters
 cores = detectCores() - 1
 if ( cores < 1 ) {
-	cores = 1
+  cores = 1
 }
 
 # Register parallel workers for %dopar%
@@ -32,23 +30,22 @@ registerDoParallel( cores )
 
 # Set up constants
 focal_matrices =
-	read.table(text =
-						 	"manuscript matrix
-						 Borowiec2015	Total1080
-						 Chang2015 Chang2015
-						 Dunn2008  Dunn2008
-						 Hejnol2009 Hejnol2009
-						 Moroz2014 ED3a
-						 Nosenko2013 nonribosomal_9187_smatrix
-						 Nosenko2013 ribosomal_11057_smatrix
-						 Philippe2009	Philippe2009
-						 Ryan2013	est.opisthokonta
-						 Ryan2013  genome.opisthokonta
-						 Simion2017	supermatrix_97sp_401632pos_1719genes
-						 Whelan2015	Metazoa_Choano_RCFV_strict
-						 ",
-						 header = TRUE,
-						 stringsAsFactors = FALSE)
+  read.table(text =
+    "manuscript matrix
+    Borowiec2015 Total1080
+    Chang2015 Chang2015
+    Dunn2008  Dunn2008
+    Hejnol2009 Hejnol2009
+    Moroz2014 ED3a
+    Nosenko2013 nonribosomal_9187_smatrix
+    Nosenko2013 ribosomal_11057_smatrix
+    Philippe2009 Philippe2009
+    Ryan2013 est.opisthokonta
+    Ryan2013 genome.opisthokonta
+    Simion2017 supermatrix_97sp_401632pos_1719genes
+    Whelan2015 Metazoa_Choano_RCFV_strict",
+    header = TRUE,
+    stringsAsFactors = FALSE)
 
 bootstrap_threshold = 90
 posterior_prob_threshold = 95
@@ -58,8 +55,8 @@ posterior_prob_threshold = 95
 papers = read_tsv( "../data_processed/tables/previously_published_manuscripts.tsv" )
 datasets = read_tsv( "../data_processed/tables/previously_published_matrices.tsv" )
 analyses_published =
-	read_tsv( "../data_processed/tables/previously_published_analyses.tsv", col_types = 'icccccccccccclccccccc' ) %>%
[…]

### 2020-06-29 manuscript/manuscript.rmd
reknit and fix NA in Extended Data Table 1

@@
 But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses.
-This mthodological variance is often confounded, for example when a particular model is only available in one software package.
+This methodological variance is often confounded, for example when a particular model is only available in one software package.
 
@@
 ```{r table_study_summary}
-tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
-  kable(tableXXdistribution)
+table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
+opts <- options(knitr.kable.NA = "")
+knitr::kable(table_study_summary)
 

### 2020-06-29 manuscript/manuscript.rmd
fix table and figure orders related to issure #22; also add sources of how supplemental table excel files are generated next to the header near the end of the manuscript

@@
 The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another.
-The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empically estimated rates (WAG or LG), or all independently estimated from the data (GTR).
+The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR).
 The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes.
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
-The analyses differ in whether they take a homongenous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites.
+The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites.
 Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
@@
 If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone.
-Gamma-rate heterogeneity, a scalar that acccomodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
+Gamma-rate heterogeneity, a scalar that accommodate total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
 
@@
 1).
-To explore this variation in gene sampling, we annotated each gene with against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table XX).
+To explore this variation in gene sampling, we annotated each gene against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1).
 These annotations show XXX.
@@
 All matrices have than 50% of the BUSCO genes, and several less than 20%.
-The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenetomic data are based on transcriptome data) and features of molecular evolution.
+The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenomic data are based on transcriptome data) and features of molecular evolution.
 
@@
 The clades are orgainized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis.
-(A) Analyses transcribed from the literature (related to Supplementary Table XX).
-(B) New phylogenomic analyses conducted in this study (related to Supplementary Table XX).
+(A) Analyses transcribed from the literature (related to Supplementary Table 2).
+(B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).
 
@@
 We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
-Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling.
+Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2).
 Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
@@
 1A); T2: Porifera-sister (Fig.
-1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Fig.
-5).
+1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Extended Data Fig.
+3).
 By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
@@
 In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$lnL<2) (Fig.
-5B; Supplementary Table 4).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs.
-5A, C).
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$lnL<2) (Extended Data Fig.
+3; Supplementary Table 4).
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+3).
 We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
@@
 CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
-Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX).
-The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx).
-Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices.
+Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5).
+The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix.
+Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices (Supplementary Table 6).
 These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
@@
 We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories in the representative data matrices (Fig.
-XX).
+4; Supplementary Table 7).
 Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQ-TREE and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutionary models.
 Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
-XX).
+4).
 More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values are from the sensitive analyses are shown in Supplementary table XX.
+Support values are from the sensitive analyses are shown in Supplementary Table 7.
 
@@
 Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
-3B).
+5).
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
@@
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-5).
+6).
 Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
@@
 We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also used a table where manual entries were needed (Supplementary Table 5, manual_taxonomy_map.tsv), e.g.
+We also used a table where manual entries were needed (manual_taxonomy_map.tsv), e.g.
 authors of the original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see Supplementary Table 6 (taxon_table.tsv).
+For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
 
@@
 We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See Supplementary Table 7 [XX partition_network_summary table] for a summary tally of each part of the comparison.
+See partition_network_summary in Rdata for a summary tally of each part of the comparison.
 
@@
 ## References
+
+## Supplementary Tables
+
+Supplementary Table 1.
+Summary of annotations (Busco, Ensemble, go-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
+
+Supplementary_Table 2. Summary of a total of 164 phylogenomic analyses were transcribed from the literature (Table is converted from analyses_published in Rdata).
+
+Supplementary_Table 3. Summary of a total of 106 phylogenomic analyses conducted in this study (Table is converted from analyses_new in R data).
+
+Supplementary_Table 4.
+Distribution of phylogenetic signal of differnet models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
+
+Supplementary_Table 5. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
+
+Supplementary_Table 6. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
+
+Supplementary_Table 7. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).

### 2020-06-29 manuscript/manuscript.rmd
fix a couple typos and update the methods in phylogenetic signal

@@
 Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
-There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation needes].
+There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting].
 To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
 
@@
 Our summary of previous phylogenetic analyses (Fig.
-3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommondate site-heterogeneous equilibrium frequencies with CAT.
+3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
 This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview.
-The only analyses that support Porifera-sister ahveh reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
+The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
 Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because CAT is so parameter rich, many analyses that use a model with CAT do not converge.
-Some of the unconvered analyses have an unresolved animal root.
+Some of the uncovered analyses have an unresolved animal root.
 The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.
@@
 There are some implementation differences.
-For example, CAT estimates the number of distinct site categories, the assigment of sites to those categories, and the equilibrium frequencies of each of those categories, while the C models specify the number of categories *a priori* (the C10 model as 10 categories, etc...).
+For example, CAT estimates the number of distinct site categories, the assignment of sites to those categories, and the equilibrium frequencies of each of those categories, while the C models specify the number of categories *a priori* (the C10 model as 10 categories, etc...).
 
@@
 Each of these was then analyzed using several models.
-We used modelfinder to identify the
+We used ModelFinder to identify the
 
@@
 site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing.
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
 For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
@@
 
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG+G, GTR+G, poisson+C60+G and associated best-fit model identified above.
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, poisson+C60 and associated best-fit model identified above.
 Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
@@
 #### Phylogenetic distribution of support
-To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta).
-For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the same best-fit model identified above, and site-homogeneous WAG and LG models.
-The two different phylogenetic hypotheses passed to IQ-TREE (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans.
-The constraint trees were conducted in a customized R script.
-The numbers of genes and sites supporting each hypothesis were calculated with IQ-TREE output and Perl scripts from Shen et al.
-2017 [@shen2017contentious].
-
-For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
+To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
+We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanomalia and the full matrix).
+For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
+The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to rest of animals.
+The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].
+
+For Poisson-CAT models, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
 
@@
 (A).
-The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.
-(B).
-The percentage BUSCO annotations in each matrix.
+The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
 (C).
-The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions.
-(D).
-The percentage ribosomal genes in each matrix.
+The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
 
@@
 **Extended Data Fig.
-3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
+3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with a log-likelihood difference of two as a gene with strong (lnl > 2) or weak (lnl < 2) phylogenetic signal for each hypothesis (Smith et al.
-2020).
+Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models.
+The GLS values for each gene in each data matrix are provided in Table S5.
+We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal.
 

### 2020-06-29 manuscript/manuscript.rmd
minor changes to the text

@@
 In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$lnL<2) (Extended Data Fig.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig.
 3; Supplementary Table 4).
@@
 We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
-The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
+These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
 
@@
 
-```{r figure_gene_composition,, fig.height = 6, fig.width = 10}
+```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
 
@@
 
-```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-
-au_tests %>%
-  unite(Matrix, c(manuscript, matrix, model)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
-  ggplot(aes(x = Matrix, y = percent, fill = result)) +
-    geom_bar(stat = "identity") +
-    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
-    coord_flip() +
-    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
-
-```
+###```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+
+#au_tests %>%
+#  unite(Matrix, c(manuscript, matrix, model)) %>%
+#  group_by(Matrix, result) %>%
+#  count() %>%
+#  group_by(Matrix) %>%
+#  mutate(percent = prop.table(n)) %>%
+#  ggplot(aes(x = Matrix, y = percent, fill = result)) +
+#    geom_bar(stat = "identity") +
+#    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
+#    coord_flip() +
+#    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
+
+#```
+
+![](figures/Figure_phylogenetic_signal.png)
+
 **Extended Data Fig.

### 2020-06-30 manuscript/manuscript.rmd
show software in figure

@@
   dplyr::rename( model = model_combined ) %>%
-  select( model, clade, final, inference, manuscript, matrix ) %>%
+  select( model, clade, final, inference, manuscript, matrix, software ) %>%
   dplyr::rename( result = final ) %>%
@@
   mutate( manuscript = NA ) %>%
-  mutate( status = "new" )
+  mutate( status = "new" ) %>%
+  mutate( software = NA )
+
+analyses_new_processed$software[ analyses_new_processed$inference == "ML" ] = "IQTREE"
+
+analyses_new_processed$software[ analyses_new_processed$inference == "Bayesian" ] = "PhyloBayes"
 
@@
   rbind( analyses_published_processed, analyses_new_processed ) %>%
-  # mutate(software=recode( software, `PhyloBayes MPI`="PhyloBayes") ) %>%
+  mutate(software=recode( software, `PhyloBayes MPI`="PhyloBayes") ) %>%
   mutate(model = recode(model,
@@
 p = analyses_processed %>%
-  ggplot( aes(x = model, y = clade, col = result, shape = inference) ) +
+  ggplot( aes(x = model, y = clade, col = result, shape = software) ) +
     facet_wrap( ~ status ) +

### 2020-06-30 manuscript/manuscript.rmd
refined text on new analyses

@@
 This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.
-Each of these was then analyzed using several models.
-We used ModelFinder to identify the
-
-We tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
-We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1).
-We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
+
+For each matrix, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] with BIC in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
+In all cases, models with C60 fit these matrices better than the site-homogeneous models.
+We then inferred support under the best-fit model from IQ-TREE (Extended Data Table 2).
+We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
 Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2).
-Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig.
-3B; Supplementary Table 3).
-
-Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister.
-This has been addressed with CAT models.
-IQ-TREE provides a new family of C60 models that also address site heterogeneity.
-Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared C60 results to CAT results for a subset of matrices to see if they give consistent results.
-This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.
-
-Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices.
-Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices.
-Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015.
-Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large].
-The Ctenophora-sister were strongly supported in all other matrices.
-For the matrices have conflicted results between Poisson-CAT and  IQ-TREE, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model.
-We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices.
-These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT.
-Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.
-
-### Drastic differences in phylogenetic signal between site- homogeneous and heterogeneous models
-
-The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
-To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
-1A); T2: Porifera-sister (Fig.
-1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Extended Data Fig.
-3).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
-This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
-
-In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig.
-3; Supplementary Table 4).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-3).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
-These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
+
+All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the ctenophore-sister hypothesis (Fig.
+3B).
+No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
+3B lower right; Supplementary Table 3), the condiitions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig.
+3A).
+To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses.
+Consistent with published results, some of these supported Porifera-sister.
+
+Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results.
+This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct.
+Instead, there is something about PhyloBayes CAT analyses that lead to support for Porifera-sister.
 
 ### Increase CAT categories lead to different topologies
-The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
-CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix.
+
+There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here.
+In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
+Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.
+
+A major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
+CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable.
+Models with C60 are set to 60 categories.
+Differences in analysis results could therefore be due to differences in the number of categories.
+
 Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5).
@@
 
+### Differences in phylogenetic signal between site- homogeneous and heterogeneous models
+
+The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
+To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
+1A); T2: Porifera-sister (Fig.
+1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Extended Data Fig.
+3).
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
+This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
+
+In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
+Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig.
+3; Supplementary Table 4).
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+3).
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
+These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
+
 ### Bias of data-recoding
@@
 A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time.
 For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.

### 2020-06-30 manuscript/manuscript.rmd
done through Summary

@@
 
-All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the ctenophore-sister hypothesis (Fig.
+All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig.
 3B).
@@
 
-### Increase CAT categories lead to different topologies
-
+### Category number explains differences between heterogeneous analyses
+
+Our analyses above show that, while PhyloBayes CAT analyses of matrices with restricted outgroup sampling sometimes support Porifera-sister, IQ-TREE C60 analyses of the same matrices do not.
 There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here.
@@
 
-A major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices.
+```{r}
+
+cat_categories = read_tsv("../data_processed/tables/cat_categories.tsv")
+
+cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
+cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR-CAT`
+
+```
+
+A primary difference between site-heterogeneous models with C (like C60) and CAT are the number of substitutional categories that sites are assigned to.
 CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable.
@@
 Differences in analysis results could therefore be due to differences in the number of categories.
-
-Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5).
-The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix.
-Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices (Supplementary Table 6).
-These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
-
-We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories in the representative data matrices (Fig.
+We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5).
+(XX add mean number of ) The number of categories is considerable larger than 60 categories used in C60 models, with a mean of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
+This is a very large number of additional parameters that must be estimated from the data.
+When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria.
+The fact that CAT is selecting ten fold more than 60 is consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
+
+Once we established this large difference in category number, we sought to test its direct impact on animal root position.
+It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes.
+We therefore varied the number of categories in Phylobayes analyses (Fig.
 4; Supplementary Table 7).
-Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQ-TREE and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutionary models.
-Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig.
-4).
-More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices.
-For example, the major transition occurred at 60 -- 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix.
-Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].
-
-The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model.
-Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups.
-Interestingly, we found both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
+We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
+This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories.
+When we sweep the number of categories, we can observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
+4), all within PhyloBayes.
+For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.
+
+The GTR-CAT model often has a better model-fitting than Poisson-CAT model in published analyses, and it is also suggested a better performance than Poisson-CAT model (XX citations).
+Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling.
+We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
 Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
-Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.
+
+These results further clarify when analyses support Porifera-sister.
+In our analyses, outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used.
+Analyses under other conditions either support Ctenophora-sister or are unresolved.
 

### 2020-07-01 manuscript/manuscript.rmd
update supplementary orders in the manuscript and add codes to generate them in kernel

@@
 No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
-3B lower right; Supplementary Table 3), the condiitions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig.
+3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig.
 3A).
@@
 Differences in analysis results could therefore be due to differences in the number of categories.
-We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5).
+We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 4).
 (XX add mean number of ) The number of categories is considerable larger than 60 categories used in C60 models, with a mean of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
 This is a very large number of additional parameters that must be estimated from the data.
-When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria.
+When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5).
 The fact that CAT is selecting ten fold more than 60 is consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
@@
 We therefore varied the number of categories in Phylobayes analyses (Fig.
-4; Supplementary Table 7).
+4; Supplementary Table 6).
 We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values are from the sensitive analyses are shown in Supplementary Table 7.
+Support values are from the sensitive analyses are shown in Supplementary Table 6.
 
@@
 Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig.
-3; Supplementary Table 4).
+3; Supplementary Table 7).
 In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 3).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4).
+We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 7).
 These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
@@
 
-Supplementary_Table 4.
-Distribution of phylogenetic signal of differnet models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
-
-Supplementary_Table 5. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
-
-Supplementary_Table 6. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
-
-Supplementary_Table 7. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).
+Supplementary_Table 4. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
+
+Supplementary_Table 5. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
+
+Supplementary_Table 6. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).
+
+Supplementary_Table 7.
+Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).

### 2020-07-01 manuscript/manuscript_kernel.R

@@ -347,3 +347,12 @@ time_run = time_stop - time_start
 ## Write the results to prepare them for manuscript.rmd
 
 save.image("manuscript.RData")
+
+## Write the results from RData to prepare them for supplementary tables
+write_csv(partition_map_global, "./Supplementary_tables/Supplementary_Table_1.csv", na = "NA", quote_escape = "double")
+write_csv(analyses_published, "./Supplementary_tables/Supplementary_Table_2.csv", na = "NA", quote_escape = "double")
+write_csv(analyses_new, "./Supplementary_tables/Supplementary_Table_3.csv", na = "NA", quote_escape = "double")
+write_csv(analyses_sensitive, "./Supplementary_tables/Supplementary_Table_6.csv", na = "NA", quote_escape = "double")
+write_csv(au_tests, "./Supplementary_tables/Supplementary_Table_7.csv", na = "NA", quote_escape = "double")
+
+

### 2020-07-01 manuscript/manuscript.rmd
use ggsave to save raw figures for Figure2 and Extended fig3

@@
 
-taxon_rectangles %>%
+p =taxon_rectangles %>%
   ggplot() +
@@
 
-```
+ggsave("figures/Figure2_raw.pdf", width=7.2, height=7.2)
+
+```
+![](figures/Figure_2_manuscript.png)
 
@@
 
+ggsave("figures/Figure3_raw.pdf", width=7.2, height=3.6)
 ```
@@
           p3, p4, labels = c("C", "D"))
+
+ggsave("figures/Figure4_raw.pdf", width=7.2, height=7.2)
+
 ```
@@
 
+ggsave("figures/Extended_Data_Fig1_raw.pdf", width=20, height=12)
+
 ```
@@
 
+ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=4)
+
 ```
@@
 
-###```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-
-#au_tests %>%
-#  unite(Matrix, c(manuscript, matrix, model)) %>%
-#  group_by(Matrix, result) %>%
-#  count() %>%
-#  group_by(Matrix) %>%
-#  mutate(percent = prop.table(n)) %>%
-#  ggplot(aes(x = Matrix, y = percent, fill = result)) +
-#    geom_bar(stat = "identity") +
-#    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
-#    coord_flip() +
-#    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
-
-#```
-
-![](figures/Figure_phylogenetic_signal.png)
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+
+p=au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+  ggplot(aes(x = Matrix, y = percent, fill = result)) +
+    geom_bar(stat = "identity") +
+    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
+    coord_flip() +
+    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
+
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=6, height=12)
+
+```
+
+![](figures/Extended_Data_Fig3_manuscript.png)
 

### 2020-07-02 manuscript/manuscript.rmd
working on last section of main text

@@
 
-High-throughput sequencing allows investigators to readilly assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1).
+High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1).
 Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
@@
 To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
 
@@
 
-Our analyses above show that, while PhyloBayes CAT analyses of matrices with restricted outgroup sampling sometimes support Porifera-sister, IQ-TREE C60 analyses of the same matrices do not.
-There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here.
+There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models.
 In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
@@
 
-A primary difference between site-heterogeneous models with C (like C60) and CAT are the number of substitutional categories that sites are assigned to.
-CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable.
-Models with C60 are set to 60 categories.
+A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories.
+CAT models employ a Dirichlet process prior to infer number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
+IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60.
 Differences in analysis results could therefore be due to differences in the number of categories.
-We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 4).
-(XX add mean number of ) The number of categories is considerable larger than 60 categories used in C60 models, with a mean of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
-This is a very large number of additional parameters that must be estimated from the data.
-When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5).
-The fact that CAT is selecting ten fold more than 60 is consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].
+The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
+This requires is a very large number of additional parameters that must be estimated from the data.
+When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5).
+These findings are consistent with the concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
 
 Once we established this large difference in category number, we sought to test its direct impact on animal root position.
-It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes.
+It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT.
 We therefore varied the number of categories in Phylobayes analyses (Fig.
@@
 This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories.
-When we sweep the number of categories, we can observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
-4), all within PhyloBayes.
+When we sweep the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
+4).
 For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.
 
-The GTR-CAT model often has a better model-fitting than Poisson-CAT model in published analyses, and it is also suggested a better performance than Poisson-CAT model (XX citations).
+Published studies have found that the GTR-CAT model often has a better fit than Poisson-CAT, and investigators have suggested that it is more appropriate than the Poisson-CAT model for this question (XX citations).
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling.
@@
 **Fig.
-4.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes.
-Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
+4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
+Statistical support values (posterior probabilities) were obtained from three data matricies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
@@
 
-### Differences in phylogenetic signal between site- homogeneous and heterogeneous models
-
-The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models.
-To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig.
-1A); T2: Porifera-sister (Fig.
-1B)) to three representative data matrices with different outgroup choice and models in both ML and BI framework (Extended Data Fig.
+### Phylogenetic signal
+
+To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
 3).
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in sifte-homogeneous models.
-This finding is largely consistent with the results from past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].
-
-In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models.
-Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models.
+This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
+
+Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
+In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
 3; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 3).
-We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 7).
-These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.
-
-### Bias of data-recoding
-There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
-They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
-5).
-Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-6).
-Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
-These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).
+We found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from XXs in matrices with distant outgroups (Opisthokonta) to XXb when outgroups are restricted (Choanimalia, Supplementary Table 7).
+
+The fact that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
+
+### Other approaches to accommodating site heterogeneity
+
+There are other approaches to addressing heterogeneity in equilibrium frequencies.
+For example, Feuda *et al.* [@Feuda:2017ew] emplyed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids.
+These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
+This approach could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
+Our own analyses (Supplementary text, Supplementary Fig.
+XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation.
 
 ## The current state of understanding and future direction
-Resolving the root position of animal phylogeny that diverged ~ 650 million years ago has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
-By synthesizing all past phylogenomic studies and perform new analyses to characterize difference between studies, we found the support of Porifera-sister can only be recovered by site-heterogeneous CAT models with the most closely related outgroups, and Ctenophora-sister in all other cases.
+Resolving the placement of the root in the animal phylogeny has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
+By synthesizing past phylogenomic studies and performing new analyses to characterize difference between studies, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with sampling restricted to the most closely related outgroups.
 Four factors contribute significantly to the previous incongruent results based on our analyses.
@@
 The GLS values for each gene in each data matrix are provided in Table S5.
-We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal.
+We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
 
@@
 
-Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie].
-In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig.
-Extended Data Fig.
-3B).
+There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
+5).
+Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
+6).
+
 Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.

### 2020-07-03 manuscript/manuscript.rmd
finished first full pass at main text

@@
 
-There are other approaches to addressing heterogeneity in equilibrium frequencies.
-For example, Feuda *et al.* [@Feuda:2017ew] emplyed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids.
-These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
-This approach could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
+Other approaches have been taken to addressing heterogeneity in equilibrium frequencies.
+For example, Feuda *et al.* [@Feuda:2017ew] employed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids.
+These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds].
+Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
 Our own analyses (Supplementary text, Supplementary Fig.
 XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition.
-These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation.
+These findings indicate that recoding can be a problematic method for addressing heterogeneity.
 
 ## The current state of understanding and future direction
-Resolving the placement of the root in the animal phylogeny has proven particularly difficult, even with large genome-scale phylogenomic data matrices.
-By synthesizing past phylogenomic studies and performing new analyses to characterize difference between studies, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with sampling restricted to the most closely related outgroups.
-Four factors contribute significantly to the previous incongruent results based on our analyses.
-First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies.
-Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models.
-More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice.
-Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny.
-Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species.
-Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.
-
-Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised.
-Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution.
-First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available).
-Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated].
-Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem.
-More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work.
-Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian].
-We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.
+
+Resolving the placement of the root in the animal phylogeny has proven particularly difficult.
+By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some analyses.
+This demonstrates and important interaction between model selection and taxon sampling.
+Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
+Site homogenous models have a single category that is applied to all sites.
+C models, as implemented in IQ-TREE, have 10-16 categories.
+CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT.
+When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories.
+As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig.
+4).
+As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.
+
+This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence.
+Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
+XX
+
+Overcoming the challenges that face this difficult and important question goes beyond models.
+In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic perspective.
+In particular, there are very few chromosome-level genome assembles outside of Bilateria.
+There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
+This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and greatly reduce the possibility of contamination in any of the gene sequence datasets.
+We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
 

### 2020-07-03 manuscript/manuscript.rmd
second pass through text

@@
 This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
-Some phylogenomic analyses find support for a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others found support for a split between comb jellies and all other animals (Ctenophora-sister).
-These analyses differed in a many respects, including the genes and species considered, molecular evolution models, and software.
+Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
+These analyses differed in a many respects, including genes considered, species considered, molecular evolution models, and software.
 Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses.
-Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
-Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models.
+Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
+Our new analyses are consistent with this.
+Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories.
 Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.
@@
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative.
-Since then many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1).
-These studies have taken a wide variety of approaches and results have varied across these different analyses.
+Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1).
 The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result.
-But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses.
+But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity.
 This methodological variance is often confounded, for example when a particular model is only available in one software package.
@@
 Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1).
-These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.
+These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.
 
@@
 Testing these hypotheses requires evaluating the support for each of these alternative nodes.
-(C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node).
+(C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
 (D) A ctenophore.
@@
 
-The published analyses considered here differ in a variety of ways, including the model of molecular evolution, gene selection, and ingroup and outgroup species selection.
-
-The models used in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
+The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
 The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another.
@@
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
-The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites.
+The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites.
 Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
@@
 
-High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1).
+High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1).
 Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
-As a result, the genes selected for analysis vary widely (Fig.
+As a result, the paritcular genes selected for analysis vary widely (Fig.
 2, Extended Data Fig.
 1).
-To explore this variation in gene sampling, we annotated each gene against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1).
-These annotations show XXX.
-We identified the ribosomal protein and BUSCO genes [@waterhouse2017busco] in each data matrix (Extended Data Fig.
+The composition of of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig.
 2).
-The animal BUSCO dataset contains 978 genes that tend to occur as a single copy used to assess the completeness of genome assemblies [@waterhouse2017busco].
-All matrices have than 50% of the BUSCO genes, and several less than 20%.
-The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenomic data are based on transcriptome data) and features of molecular evolution.
 
@@
 2.** An overview of previous studies.
-(A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node).
-(B) Matrix taxa and gene sampling for each study.
-Each of the primary matrices considered here, color coded by taxon sampling.
+(A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister ( orange nodes) or Porifera-sister (green nodes).
+(B) Each of the primary matrices considered here, color coded by taxon sampling.
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
 
-Ingroup taxon sampling across clades varies widely between studies (Fig.
+Ingroup taxon sampling also varies widely between studies (Fig.
 2, Extended Data Fig.
@@
 Sampling of ingroup taxa (animals) was initially biased toward Bilateria.
-Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups.
-Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig.
+Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig.
 1).
-Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig.
+Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig.
 1).
-
-Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup.
+This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.
+
+Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup.
 There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting].
-To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
+Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
@@
 Our summary of previous phylogenetic analyses (Fig.
-3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
+3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
 This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview.
 The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
-Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because CAT is so parameter rich, many analyses that use a model with CAT do not converge.
-Some of the uncovered analyses have an unresolved animal root.
+Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
 The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.
@@
 
-One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across analyses.
+One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses.
 This extensive variation can make it difficult to identify the primary factors that lead to different results.
-Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] under a panel of evolutionary models.
-We selected this tool because it has greater model flexibility than other maximum likelihood (ML) tools [@zhou2017evaluating] (Fig.
+Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models.
+We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig.
 3B; Supplementary Table 3).
 Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.
-There are some implementation differences.
-For example, CAT estimates the number of distinct site categories, the assignment of sites to those categories, and the equilibrium frequencies of each of those categories, while the C models specify the number of categories *a priori* (the C10 model as 10 categories, etc...).
 
@@
 For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis.
-For each of these matrices, we progressively trimmed taxon sampling of to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
+For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
 This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.
 
-For each matrix, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] with BIC in IQ-TREE (except only LG+C60 models are used in 3 large matrices).
+For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE.
 In all cases, models with C60 fit these matrices better than the site-homogeneous models.
-We then inferred support under the best-fit model from IQ-TREE (Extended Data Table 2).
-We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60.
-Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2).
+We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60.
+We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Extended Data Table 2).
 
@@
 No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
-3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig.
+3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
 3A).
@@
 This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct.
-Instead, there is something about PhyloBayes CAT analyses that lead to support for Porifera-sister.
+Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
 
@@
 A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories.
-CAT models employ a Dirichlet process prior to infer number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
+CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
 IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60.
@@
 The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
-This requires is a very large number of additional parameters that must be estimated from the data.
-When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5).
-These findings are consistent with the concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
+This requires a very large number of additional parameters that must be estimated from the data.
+When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to ModelFinder (Supplementary Table 5).
+These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
 
@@
 We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
-This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories.
-When we sweep the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
+This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories.
+When we swept the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
 4).
@@
 We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
-Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.
+Moreover, we also found all results strongly supported Ctenophora-sister with GTR-nCAT60 models.
 
 These results further clarify when analyses support Porifera-sister.
-In our analyses, outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used.
+Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used.
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
@@
 3).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes).
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models.
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
 This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
@@
 Other approaches have been taken to addressing heterogeneity in equilibrium frequencies.
-For example, Feuda *et al.* [@Feuda:2017ew] employed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids.
+For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids.
 These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds].
@@
 
-Resolving the placement of the root in the animal phylogeny has proven particularly difficult.
-By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some analyses.
-This demonstrates and important interaction between model selection and taxon sampling.
+Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem.
+By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
+This demonstrates an important interaction between model selection and taxon sampling.
 Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
@@
 Overcoming the challenges that face this difficult and important question goes beyond models.
-In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic perspective.
+In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context.
 In particular, there are very few chromosome-level genome assembles outside of Bilateria.
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and greatly reduce the possibility of contamination in any of the gene sequence datasets.
+This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

### 2020-07-03 manuscript/manuscript.rmd
Thrid round or pass to main text and Extended Fig3

@@
 
-Yuanning Li^1,3^, Xingxing Shen^3^, Benjamin Evans^2^, Antonis Rokas^3^, and Casey W. Dunn^1^*
-
-^1^Department of Ecology and Evolutionary Biology, Yale University
-
-^2^Yale Center for Research Computing, Yale University
-
-^3^Department of Biological Science, Vanderbilt University
+Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Steven H.D. Haddock^5^, Antonis Rokas^3^ and Casey W. Dunn^1^*
+
+^1^Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06511, USA
+
+^2^Department of Agriculture, Zhejiang University, Hangzhou, Zhejiang, China
+
+^3^Department of Biological Science, Vanderbilt University, Nashville 37235, USA
+
+^4^Yale Center for Research Computing, Yale University, New Haven, CT 06511, USA
+
+^5^Monterey Bay Aquarium Research Institute, Moss Landing, CA 95039, USA
 
 \* Corresponding author, casey.dunn@yale.edu
+
+ORCIDs:
+
+Yuanning Li: 0000-0001-5765-1419
+Xing-Xing Shen: 0000-0002-8436-595X
+Benjamin Evans:
+Steven H.D. Haddock: 0000-0001-9420-4482
+Antonis Rokas: 0000-0002-7248-6551
+Casey W. Dunn: 0000-0003-0628-5150
 
@@
 Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
-These analyses differed in a many respects, including genes considered, species considered, molecular evolution models, and software.
-Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses.
+These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
+Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic analyses and performing new standardized analyses.
 Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
@@
 Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories.
-Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.
+Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
 
@@
 Historically, there was little debate about the root of the animal tree of life.
-Porifera-sister (Fig.
-1) was widely accepted though rarely tested.
+Porifera-sister (Figs.
+1B, 1D) was widely accepted though rarely tested.
 By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative.
+The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs.
+1A, 1C).
 Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1).
-The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result.
+The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig.
+1E).
 But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity.
@@
 To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework.
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1).
+Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 15 studies (Fig.
+2A, Extended Data Table 1).
 These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.
@@
 **Fig.
-1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
+1.** Two alternative phylogenetic hypotheses from previous phylogenomic studies.
+(A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
 (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
@@
 The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR).
-The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes.
+The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes.
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
@@
 If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone.
-Gamma-rate heterogeneity, a scalar that accommodate total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
+Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
 
@@
 Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
-As a result, the paritcular genes selected for analysis vary widely (Fig.
-2, Extended Data Fig.
+As a result, the particular genes selected for analysis vary widely (Fig.
+2B, Extended Data Fig.
 1).
-The composition of of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig.
+The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig.
 2).
@@
 **Fig.
-2.** An overview of previous studies.
-(A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister ( orange nodes) or Porifera-sister (green nodes).
+2.** An overview of previous phylogenomic studies.
+(A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study.
 (B) Each of the primary matrices considered here, color coded by taxon sampling.
@@
 Ingroup taxon sampling also varies widely between studies (Fig.
-2, Extended Data Fig.
+2B, Extended Data Fig.
 1).
@@
 There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting].
-Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic].
+Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
 Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
@@
 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
-Each point represents a phylogenetic Analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
-The clades are orgainized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis.
+Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
+The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis.
 (A) Analyses transcribed from the literature (related to Supplementary Table 2).
@@
 3A and Supplementary Table 2).
-The conclusions of five studies strongly favor of Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
+The conclusions of five studies strongly favor Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
 Three studies are based entirely on previously published data, and the remainder add data for one or more species.
@@
 3B; Supplementary Table 3).
-Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.
+Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.
 
@@
 
+Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
 For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis.
 For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
-This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.
-
-For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE.
+This produced `r analyses_new$matrix%>% unique()%>% length()`from 11 studies.
+
+For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE.
 In all cases, models with C60 fit these matrices better than the site-homogeneous models.
@@
 
-Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results.
+Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results.
 This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct.
@@
 
-cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
-cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR-CAT`
+cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson+CAT`
+cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR+CAT`
 
@@
 This requires a very large number of additional parameters that must be estimated from the data.
-When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to ModelFinder (Supplementary Table 5).
+Despite C60 models have a better fit than site-homogeneous mdoels in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probabaly already too high according to ModelFinder (Supplementary Table 5).
 These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
@@
 
-Published studies have found that the GTR-CAT model often has a better fit than Poisson-CAT, and investigators have suggested that it is more appropriate than the Poisson-CAT model for this question (XX citations).
-Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling.
-We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model.
-Moreover, we also found all results strongly supported Ctenophora-sister with GTR-nCAT60 models.
+Published studies have found that the GTR+CAT model often has a better fit than Poisson+CAT, and investigators have suggested that it is more appropriate than the Poisson+CAT model for this question [@simion2017large].
+Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
+We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
+Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
 
@@
 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
-Statistical support values (posterior probabilities) were obtained from three data matricies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
+Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories.
+(A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
@@
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
-This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
+This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
 
@@
 3; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 3).
-We found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from XXs in matrices with distant outgroups (Opisthokonta) to XXb when outgroups are restricted (Choanimalia, Supplementary Table 7).
+For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).
 
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-Our own analyses (Supplementary text, Supplementary Fig.
-XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition.
+Our own analyses (Supplementary text, Supplementary Figs.
+4-6) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition.
 These findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
-Site homogenous models have a single category that is applied to all sites.
-C models, as implemented in IQ-TREE, have 10-16 categories.
+Site-homogeneous models have a single category that is applied to all sites.
+C models, as implemented in IQ-TREE, have 10-60 categories.
 CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT.
@@
 In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context.
-In particular, there are very few chromosome-level genome assembles outside of Bilateria.
+In particular, there are very few chromosome-level genome assemblies outside of Bilateria.
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
+This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 
-All tree files, intermidiate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publicaton.
+All tree files, intermediate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publication.
 
@@
 Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson-CAT models in PhyloBayes were conducted.
+For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
-Moreover, bayesian inference with the site-heterogeneous poisson-CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, CAT-GTR models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
+To minimize computational burden, GTR+CAT models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
 For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
@@
 
-For Poisson-CAT models, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
+For Poisson+CAT models, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
 
@@
 To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson-CAT and GTR-CAT and minimize computational burden, GTR-CAT and GTR-CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers fo categories were used.
+To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different number of categories were used.
 
@@
 
-## Ackowledgements
+## Acknowledgements
 
@@
 Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species.
-Th data matrix were assembled based on The phylogenetic analysis using Poisson-CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
+Th data matrix were assembled based on The phylogenetic analysis using Poisson+CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
 Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70.
@@
 leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals.
-They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson-CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix.
+They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson+CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix.
 The Orthologs were constructed based on the method of Hejnol *et al.* 2009.
@@
 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
-Both site-heterogeneous Poisson-CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.
+Both site-heterogeneous Poisson+CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.
 
@@
 Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes.
-They found GTR-CAT and Poisson-CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
+They found GTR+CAT and Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
 Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
@@
 
-The final matrix was first analyzed using the Poisson-CAT model.
+The final matrix was first analyzed using the Poisson+CAT model.
 Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.

### 2020-07-09 manuscript/manuscript.rmd
updated revisions from Xingxing and Antonis

@@
 
-^2^Department of Agriculture, Zhejiang University, Hangzhou, Zhejiang, China
+^2^2State Key Laboratory of Rice Biology and Ministry of Agriculture Key Lab of Molecular Biology of Crop Pathogens and Insects, Institute of Insect Sciences, Zhejiang University, Hangzhou 310058, China
 
@@
 
-Yuanning Li: 0000-0001-5765-1419
+Yuanning Li: 0000-0002-2206-5804
 Xing-Xing Shen: 0000-0002-8436-595X
@@
 
-There has been considerable recent debate about the placement of the root in the animal tree of life.
+There has been considerable debate about the placement of the root in the animal tree of life.
 This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
 Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
-These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
-Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic analyses and performing new standardized analyses.
+[XXXIt would be helpful if you also stated why this is a problem: Antonis] These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
+Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses.
 Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
 Our new analyses are consistent with this.
-Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories.
-Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
+Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of substitutional categories.
+[XXXTo me this part is a diversion – I would state your key inference (“our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories”) upfront.
+You can then explain how your results fit / shed light on previous interpretations of certain analyses.]Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
 
@@
 Porifera-sister (Figs.
-1B, 1D) was widely accepted though rarely tested.
-By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
+1B, 1E) was widely accepted though rarely tested.
+By contrast, there has long been uncertainty about the placement of Ctenophora in the animal tree of life [@Wallberg:2004ws].
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs.
-1A, 1C).
-Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1).
+1A, 1D).
+Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1).
 The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig.
-1E).
-But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity.
+1).
+But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity.
 This methodological variance is often confounded, for example when a particular model is only available in one software package.
@@
 Model effect is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
+studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1).
-Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices.
+Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices.
 As a result, the particular genes selected for analysis vary widely (Fig.
@@
     taxon_rectangles$matrix,
-    levels = c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est",
-             "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080", "Chang2015",
-             "Whelan2017_full", "Simion2017"))
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_ribo_14615", "Ryan2013_est",
+             "Moroz2014_3d", "Borowiec2015_Total1080", "Whelan2015_D20", "Chang2015",
+             "Simion2017", "Whelan2017_full"))
 
@@
 1).
-Sampling of ingroup taxa (animals) was initially biased toward Bilateria.
+Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria.
 Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig.
@@
 1).
-This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.
+This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.
 
@@
 **Fig.
-3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study.
+3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study.
 Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
-The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis.
+The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parametersto the right on the horizontal axis.
 (A) Analyses transcribed from the literature (related to Supplementary Table 2).
@@
 3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
-This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview.
+This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination.
 The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
-Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
+Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge and thus renders the analysis untrustworthy / hard to interpret.
 The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.
@@
 One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses.
-This extensive variation can make it difficult to identify the primary factors that lead to different results.
+This extensive variation makes it difficult to identify the primary factors that lead to different results.
 Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models.
@@
 3B; Supplementary Table 3).
-Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.
+Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.
 
@@
 For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
-This produced `r analyses_new$matrix%>% unique()%>% length()`from 11 studies.
+This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.
 
@@
 
-Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results.
+Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results.
 This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct.
@@
 A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories.
-CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
-IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60.
+CAT models employ a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
+IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60).
 Differences in analysis results could therefore be due to differences in the number of categories.
@@
 This requires a very large number of additional parameters that must be estimated from the data.
-Despite C60 models have a better fit than site-homogeneous mdoels in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probabaly already too high according to ModelFinder (Supplementary Table 5).
+Despite C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5).
 These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
@@
 This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories.
-When we swept the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
+When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
 4).
@@
 
-To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
+To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
 3).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
@@
 3; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 3).
@@
 
-The fact that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
+The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
 
@@
 
-## The current state of understanding and future direction
+## The current state of understanding and future directions
 
@@
 
-Overcoming the challenges that face this difficult and important question goes beyond models.
-In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context.
+Overcoming the challenges associated with this difficult and important question goes beyond models.
+In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie].
 In particular, there are very few chromosome-level genome assemblies outside of Bilateria.
@@
 
-For Poisson+CAT models, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]
+For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency.
+Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
+For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
+Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15
 
@@
 This is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
+studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 
-Note the main conclusion of this paper is baed on the selection of genes that evolved slowly ribosomal RNA genes.
+Note the main conclusion of this study is baed on the selection of genes that evolved slowly ribosomal RNA genes.
 The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].
@@
 The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes.
-The main conclusion of the paper was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices.
+The main conclusion of the study was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices.
 Phylogenetic analyses were conducted under the GTR+CAT model in PhyloBayes in selected matrices, and under the data-partitioning methods in RAxML for all matrices.
@@
 All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset.
-The main conclusions of this paper were based on Whelan2017_full and Whelan2017_strict matrices.
+The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.
 

### 2020-07-16 manuscript/manuscript.rmd
add Dunn2008_choanozoa C60 results that previous not added
update manuscript with Nathan's comments

@@
 Yuanning Li: 0000-0002-2206-5804
+
 Xing-Xing Shen: 0000-0002-8436-595X
+
 Benjamin Evans:
 Steven H.D. Haddock: 0000-0001-9420-4482
+
 Antonis Rokas: 0000-0002-7248-6551
+
 Casey W. Dunn: 0000-0003-0628-5150
@@
 
-The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
+The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
 The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another.
-The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR).
-The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes.
+The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
+The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes.
+Or for data-partitioning analyses, some gene partitions use different site-homogeneous models.
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
-The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites.
+The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites.
 Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
 The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention.
-For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices.
-Model effect is further confused by the abbreviations that are used for models.
-studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
+While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 1).
-The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig.
+Gene sampling varies in several ways, including in the fractions of single copy orthologs (e.g., BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
 2).
@@
 **Fig.
-2.** An overview of previous phylogenomic studies.
+2.** An overview of previous phylogenomic studieson animals.
 (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study.
@@
 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study.
-Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
-The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parametersto the right on the horizontal axis.
+Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved.
+The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis.
 (A) Analyses transcribed from the literature (related to Supplementary Table 2).
@@
 Our summary of previous phylogenetic analyses (Fig.
-3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
+3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
 This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination.
@@
 This extensive variation makes it difficult to identify the primary factors that lead to different results.
-Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models.
+Here we first reanalyze the primary matrices from each study under the same conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models.
 We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig.
@@
 Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
-For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis.
+For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies.
 For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
@@
 4).
-For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.
+For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson-CAT model.
 
@@
 
-These results further clarify when analyses support Porifera-sister.
-Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used.
+These results further clarify when analyses support Porifera-sister (e.g., Philippe2009, Whelan2017).
+Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used.
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
@@
 
-ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")), nrow = 3,
-          p3, p4, labels = c("C", "D"))
+ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"))
 
@@
 Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
-In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
+In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
 3; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 3).
-For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).
+For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
 
@@
 
-Other approaches have been taken to addressing heterogeneity in equilibrium frequencies.
+Other approaches have been taken to addressing base compositional heterogeneity across taxa.
 For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids.
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-Our own analyses (Supplementary text, Supplementary Figs.
-4-6) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition.
-These findings indicate that recoding can be a problematic method for addressing heterogeneity.
+However, Our own analyses (Supplementary text, Supplementary Figs.
+4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
 
@@
 Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
-XX
+Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
 
@@
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
+Instead of continuing to use previous matrices that has limited taxa sampling (e.g., Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
+[XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling.
+Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded.
-If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded.
+If the duplications for a small number of taxa were unresolved, all genes from those taxa were excluded.
 Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix.
@@
 > we were able to reconstruct a metazoan phylogeny that is consistent with traditional, morphology-based views on the phylogeny of non-bilaterian metazoans, including monophyletic Porifera and ctenophores as a sister-group of cnidarians.
-
-Note the main conclusion of this study is baed on the selection of genes that evolved slowly ribosomal RNA genes.
-The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].
 
@@
 The Orthologs were constructed based on the method of Hejnol *et al.* 2009.
-For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-baed analyses.
+For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-based analyses.
 Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model.
@@
 #### Whelan et al. 2015
-Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
+Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
 long-branch attraction, compositional bias, fast evolving genes, etc.).
@@
 All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister.
-CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora.
+GTR-CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.
 
@@
 
-Note that this data matrix has been extensively reanalyzed in many studies.
+Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long branch in the tree.
 
@@
 Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes.
-They found GTR+CAT and Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
+They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
 Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
@@
 Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets).
-It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods in the relationships at the root of the animal tree.
+It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods based on recovering Porifera-sister hypothesis.
 The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets.
@@
 The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets.
-More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in many cases.
+More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in almost every cases.
 They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model.
-The results indicated that Poisson + CAT model tends to recover less accurate trees and more importantly,  both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.
+The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.
 
@@
 It significantly increased ctenophore taxon sampling than other studies.
-Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes.
+Putative orthologs were determined largely similar to Whelan2015.
 The subsequent filtering strategy was also similar to the previous study.
@@
 Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
-Each jackknife is baed on a random selection of ~ 25% of the genes.
+Each jackknife is based on a random selection of ~ 25% of the genes.
 The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets.
@@
 
-> Our dataset outperforms previous metazoan gene superalignments in terms of data quality and quantity.
+> Our dataset outperforms previous metazoan gene super alignments in terms of data quality and quantity.
 Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.

### 2020-07-16 manuscript/manuscript.rmd
make sure e.g. and et al. are used consistently throughout the manuscript

@@
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT).
+To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT).
 If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone.
@@
 1).
-Gene sampling varies in several ways, including in the fractions of single copy orthologs (e.g., BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
+Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
 2).
@@
 
-There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models.
+There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models.
 In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
@@
 4).
-For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson-CAT model.
+For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+CAT model.
 
@@
 
-These results further clarify when analyses support Porifera-sister (e.g., Philippe2009, Whelan2017).
+These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
 Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used.
@@
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-Instead of continuing to use previous matrices that has limited taxa sampling (e.g., Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
+Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
 [XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling.
@@
 We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also used a table where manual entries were needed (manual_taxonomy_map.tsv), e.g.
-authors of the original matrix indicate species name in original manuscript.
+We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript.
 For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
@@
 To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g.
-site-heterogeneous vs.
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
 site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
@@
 **Extended Data Fig.
-5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+5.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model.
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).
+To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT).
 Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here.
@@
 
-#### Hejnol et al. 2009
+#### Hejnol *et al.* 2009
 
 Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms.
-The orthology inference was largely similar to Dunn et al.
-2008, with the exception of orthology genes were clustered by MCL.
+The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL.
 The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy.
@@
 #### Whelan et al. 2015
-Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g.
-long-branch attraction, compositional bias, fast evolving genes, etc.).
+Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.).
 Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.
@@
 All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister.
-GTR-CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.
+GTR+CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.
 
@@
 Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
-The authors constructed a dataset with 200 protein markers based on Philippe et al.
-2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
+The authors constructed a dataset with 200 protein markers based on Philippe *et al.* 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
 Both site-heterogeneous Poisson+CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.
@@
 It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods based on recovering Porifera-sister hypothesis.
-The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets.
-The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).
+The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (GTR+CAT) models in non-recoding and recoding datasets.
+The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).
 
@@
 The subsequent filtering strategy was also similar to the previous study.
-All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset.
+All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in choanoazoa dataset.
 The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.
@@
 Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
@@
 Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
-This leaves eight CAT-GTR+G analyses (Extended Data Fig.
+This leaves eight GTR+CAT analyses (Extended Data Fig.
 5).
@@
 Of the six recoded analyses, five are unresolved.
-Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
+Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the GTR+CAT analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
 Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
@@
 Similar results would suggest that the impact of recoding is due to discarding information.
-Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data.
+Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the GTR+CAT analysis of the SR-6 recoded Whelan data.
 we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding.
+When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding.
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.

### 2020-07-16 manuscript/tests/testthat/test_pb.R
created testing framework for phylobayes parsing re #31

@@ -1,0 +1,15 @@
+library( ape )
+library( tidyverse )
+library( magrittr )
+library( stringr )
+
+
+source( "../../phylobayes.R" )
+
+context("testing framework")
+
+test_that("Testing works", {
+  expect_equal( 1, 1 )
+})
+
+context("phylobayes chain files")

### 2020-07-16 manuscript/phylobayes.R
created testing framework and started parsing re #31

@@ -1,0 +1,111 @@
+library(ape)
+
+
+setOldClass("phylo")
+
+#######################################
+## Classes
+
+
+#' An S4 class to store a single sample from a phylobayes chain file
+#' 
+#' @slot allocation  The allocation vector
+setClass(
+  "PhylobayesSample",
+  slots = c(
+    
+    path = "character",
+    generation = "numeric",
+    
+    # These are shown in the order of a phylobayes MPI record. One item per line
+    # unless otherwise noted
+    
+    tree = "phylo",
+    x2 = "numeric", # value 1 appears constant between iterations
+    x3 = "numeric", # could be NSPR kappa or kmax, not in any other file
+    x4 = "numeric", # appears to be alpha parameter, found in trace file
+    x5 = "numeric", # ncat?
+    x6 = "numeric", # 20 numbers, possibly some starting values for each amino acid (last tab is blank)
+    # Blank line
+    frequencies = "numeric", # number of columns is number of states (eg 20 for amino acids), number of rows is the number of categories (one line per category in the file)
+    allocation = "numeric" # integer that indicates the category of each site
+  )
+)
+
+#' Construct a PhylobayesSample from the text for a single generation from a phylobayes chain
+#'
+#' @param sample_text text for a single sample from a phylobayes chain file
+#' @param path path to the original file
+#' @param generation the generation of the sample in the original file
+#' @return A PhylobayesSample object
+#' @export
+PhylobayesSample = function( 
+  text, 
+  path = NA,
+  generation = NA
+  ) {
+  
+  object = new( 
+    "PhylobayesSample",
+    path = path, 
+    generation = generation,
+    tree = ape::read.tree( text=text[1] ),
+    x2 = as.numeric(text[2])
+  )
+  
+  object
+}
+
+#######################################
+## Functions
+
+#' Parse a phylobayes mpi chain file
+#'
+#' @param fine_name name of file to parse
+#' @return A PhylobayesSample object
+#' @export
+
+parse_phylobayes_chain = function( file_name ){
+  
+  generation = 0
+  samples = list()
+  
+  # This will be very inefficient for large files, but will work for a first pass
+  lines = read_lines( file_name )
+  n = 0
+  sample_text = character()
+  
+  for (line in lines) {
+    n = n+1
+    if( n == 1 ) {
+      if ( grepl( "\\(", line ) ) {
+        generation = generation + 1
+        sample_text = append( sample_text, line )
+      } else {
[…]

### 2020-07-16 manuscript/tests/testthat/test_pb.R

@@ -8,8 +8,32 @@ source( "../../phylobayes.R" )
 
 context("testing framework")
 
+# The example file was gcreated by manually editing 
+# Whelan2017_strict.phy_Poisson_CAT60_Chain_1_last500.chain
+# to obtain the last three generations of the chain
+
+example_name = "example.chain"
+
 test_that("Testing works", {
   expect_equal( 1, 1 )
 })
 
+
+context("single phylobayes chain record")
+
+lines = read_lines( "example_single.chain" )
+
+pb = PhylobayesSample( lines, "example_single.chain", 1 )
+
+test_that("Tree has the correct number of tips", {
+  expect_equal( pb@tree$Nnode, 74 )
+})
+
 context("phylobayes chain files")
+
+chain = parse_phylobayes_chain("example.chain")
+
+test_that("Tree has the correct number of tips", {
+  expect_equal( length(chain), 7 )
+})
+

### 2020-07-16 manuscript/phylobayes.R
parsing all fields

@@ -27,7 +27,7 @@ setClass(
     x5 = "numeric", # ncat?
     x6 = "numeric", # 20 numbers, possibly some starting values for each amino acid (last tab is blank)
     # Blank line
-    frequencies = "numeric", # number of columns is number of states (eg 20 for amino acids), number of rows is the number of categories (one line per category in the file)
+    frequencies = "matrix", # number of columns is number of states (eg 20 for amino acids), number of rows is the number of categories (one line per category in the file)
     allocation = "numeric" # integer that indicates the category of each site
   )
 )
@@ -50,8 +50,22 @@ PhylobayesSample = function(
     path = path, 
     generation = generation,
     tree = ape::read.tree( text=text[1] ),
-    x2 = as.numeric(text[2])
+    x2 = as.numeric(text[2]),
+    x3 = as.numeric(text[3]),
+    x4 = as.numeric(text[4]),
+    x5 = as.numeric(text[5]),
+    x6 = text[6] %>% str_split("\t") %>% unlist() %>% as.numeric(),
+    frequencies = text[8:(length(text)-1)] %>% str_split("\t") %>% lapply( as.numeric ) %>% do.call( rbind, . ), 
+    allocation = text[length(text)] %>% str_split("\t") %>% unlist() %>% as.numeric()
   )
+  
+  if( text[7] != "" ){
+    stop("File format invalid, expected blank line")
+  }
+  
+  if( x5 != nrow(frequencies) ){
+    stop("Unexpected number of category frequency profiles")
+  }
   
   object
 }

### 2020-07-16 manuscript/tests/testthat/test_pb.R
added some more tests

@@ -25,7 +25,12 @@ lines = read_lines( "example_single.chain" )
 
 pb = PhylobayesSample( lines, "example_single.chain", 1 )
 
+
 test_that("Tree has the correct number of tips", {
+  expect_equal( length(pb@tree$tip.label), 76 )
+})
+
+test_that("Tree has the correct number of internal nodes", {
   expect_equal( pb@tree$Nnode, 74 )
 })
 
@@ -33,7 +38,17 @@ context("phylobayes chain files")
 
 chain = parse_phylobayes_chain("example.chain")
 
-test_that("Tree has the correct number of tips", {
+test_that("Chain has correct number of samples", {
   expect_equal( length(chain), 7 )
+})
+
+test_that("Allocation vector lengths are correct", {
+  x = lapply(chain, function(x){length(x@allocation)}) %>% unlist()
+  expect_true( all(x == 49388) )
+})
+
+test_that("Number of tips in trees are correct", {
+  x = lapply(chain, function(x){length( x@tree$tip.label )}) %>% unlist()
+  expect_true( all(x == 76) )
 })
 

### 2020-07-16 manuscript/explorations/phylobayes_explorations.rmd
some explorations of equilibrium frequency heterogeneity

---
output:
  html_document: default
  pdf_document: default
  word_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  dpi = 300,
  cache = FALSE
  )
```

```{r preliminaries}
  library( tidyverse )
  library( magrittr )

  source( "../phylobayes.R" )

  # A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
  cbPalette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

  result_colors = c( cbPalette[2], cbPalette[4], cbPalette[1] )
  names( result_colors ) = c( "Ctenophora-sister", "Porifera-sister", "Unresolved" )

```

An exploration of phylobayes chain samples.

```{r}

chain = parse_phylobayes_chain("../tests/testthat/example.chain")

# Take just the last sample for now
pb = chain[[length(chain)]]

allocation_counts = table(pb@allocation)

# Building on https://www.statmethods.net/advstats/mds.html
d = dist( pb@frequencies )
fit = cmdscale(d,eig=TRUE, k=2)

categories = tibble(
  index = allocation_counts %>% names() %>% as.numeric(),
  count = c(allocation_counts),
  x = fit$points[,1],
  y = fit$points[,2]
)

categories %>% 
  ggplot() +
  geom_point(aes(x=x, y=y, size=count) ) +
  ggtitle("MDS plot of rate categories") +
  theme_classic()

```

### 2020-07-16 manuscript/phylobayes.R

@@ -63,7 +63,7 @@ PhylobayesSample = function(
     stop("File format invalid, expected blank line")
   }
   
-  if( x5 != nrow(frequencies) ){
+  if( object@x5 != nrow(object@frequencies) ){
     stop("Unexpected number of category frequency profiles")
   }
   

### 2020-07-16 manuscript/tests/testthat/test_pb.R

@@ -19,6 +19,7 @@ test_that("Testing works", {
 })
 
 
+
 context("single phylobayes chain record")
 
 lines = read_lines( "example_single.chain" )
@@ -33,6 +34,8 @@ test_that("Tree has the correct number of tips", {
 test_that("Tree has the correct number of internal nodes", {
   expect_equal( pb@tree$Nnode, 74 )
 })
+
+
 
 context("phylobayes chain files")
 

### 2020-07-17 manuscript/phylobayes.R
incomplete records return NULL rather than throw error

@@ -37,13 +37,25 @@ setClass(
 #' @param sample_text text for a single sample from a phylobayes chain file
 #' @param path path to the original file
 #' @param generation the generation of the sample in the original file
-#' @return A PhylobayesSample object
+#' @return A PhylobayesSample object, or NULL if the record is incomplete
 #' @export
 PhylobayesSample = function( 
   text, 
   path = NA,
   generation = NA
   ) {
+  
+  
+  if( text[7] != "" ){
+    # The seventh line should be blank
+    return( NULL )
+  }
+  
+  if ( ! grepl( "\\(", text[1] ) ) {
+    # Record does not start with a tree
+    return( NULL )
+  }
+  
   
   object = new( 
     "PhylobayesSample",
@@ -59,12 +71,9 @@ PhylobayesSample = function(
     allocation = text[length(text)] %>% str_split("\t") %>% unlist() %>% as.numeric()
   )
   
-  if( text[7] != "" ){
-    stop("File format invalid, expected blank line")
-  }
-  
   if( object@x5 != nrow(object@frequencies) ){
-    stop("Unexpected number of category frequency profiles")
+     # Unexpected number of category frequency profiles
+    return( NULL )
   }
   
   object
@@ -88,6 +97,12 @@ parse_phylobayes_chain = function( file_name ){
   lines = read_lines( file_name )
   n = 0
   sample_text = character()
+  
+  # Each record starts with a tree line, which starts with a (
+  # Since many files are quite large, it may be apprpriate to tail them,
+  # in which case the first record may be incomplete. Set a flag to note when the
+  # first complete record has started, and don't parse before then.
+  start = FALSE
   
   for (line in lines) {
     n = n+1

### 2020-07-17 manuscript/tests/testthat/test_pb.R

@@ -18,8 +18,6 @@ test_that("Testing works", {
   expect_equal( 1, 1 )
 })
 
-
-
 context("single phylobayes chain record")
 
 lines = read_lines( "example_single.chain" )
@@ -34,6 +32,20 @@ test_that("Tree has the correct number of tips", {
 test_that("Tree has the correct number of internal nodes", {
   expect_equal( pb@tree$Nnode, 74 )
 })
+
+
+test_that("Missing tree returns NULL", {
+  expect_null( PhylobayesSample( lines[-1], "example_single.chain", 1 )  )
+})
+
+test_that("Missing blank line returns NULL", {
+  expect_null( PhylobayesSample( lines[-7], "example_single.chain", 1 )  )
+})
+
+test_that("Missing last line returns NULL", {
+  expect_null( PhylobayesSample( lines[-length(lines)], "example_single.chain", 1 )  )
+})
+
 
 
 

### 2020-07-19 manuscript/explorations/phylobayes_explorations.rmd
incomplete records return NULL rather than throw error
parses unconstrained CAT chains

@@
   library( magrittr )
+  library( gridExtra )
 
@@
 
-chain = parse_phylobayes_chain("../tests/testthat/example.chain")
-
-# Take just the last sample for now
-pb = chain[[length(chain)]]
+plot_chain = function( pb ){
 
@@
 
+  gg_mds = 
   categories %>% 
@@
 
+  gg_rank = 
+    categories %>%
+    arrange(desc(count)) %>%
+    mutate(rank = 1:nrow(categories)) %>%
+    ggplot() +
+      geom_point(aes(x=rank, y=count) ) +
+      ggtitle("Ordered count") +
+      theme_classic()
+
+  p = grid.arrange(
+  	gg_rank,
+  	gg_mds,
+  	ncol = 2
+  )
+
+  return(p)
+}
+
 ```
+
+```{r}
+
+chain = parse_phylobayes_chain( "../tests/testthat/example.chain" )
+
+# Take just the last sample for now
+pb = chain[[length(chain)]]
+
+plot_chain( pb )
+
+```
+
+**Figure n60.** Plots for nCAT=60.
+
+```{r}
+
+chain = parse_phylobayes_chain( "Whelan2017_strict.phy_Poisson_CAT100_Chain_1_last500.chain" )
+
+# Take just the last sample for now
+pb = chain[[length(chain)]]
+
+plot_chain( pb )
+
+```
+
+**Figure n100.** Plots for nCAT=100.
+
+```{r}
+
+chain = parse_phylobayes_chain( "Whelan2017_strict.phy_Chain1_last50000.chain" )
+
+# Take just the last sample for now
+pb = chain[[length(chain)]]
+
+plot_chain( pb )
+
+```
+
+**Figure CAT.** Plots for CAT.

### 2020-07-19 manuscript/phylobayes.R

@@ -25,7 +25,8 @@ setClass(
     x3 = "numeric", # could be NSPR kappa or kmax, not in any other file
     x4 = "numeric", # appears to be alpha parameter, found in trace file
     x5 = "numeric", # ncat?
-    x6 = "numeric", # 20 numbers, possibly some starting values for each amino acid (last tab is blank)
+    x6 = "numeric", # present in CAT runs, but not NCAT
+    x7 = "numeric", # 20 numbers, possibly some starting values for each amino acid (last tab is blank)
     # Blank line
     frequencies = "matrix", # number of columns is number of states (eg 20 for amino acids), number of rows is the number of categories (one line per category in the file)
     allocation = "numeric" # integer that indicates the category of each site
@@ -45,9 +46,22 @@ PhylobayesSample = function(
   generation = NA
   ) {
   
+  # Chain files from nCAT and CAT runs have slightly different formats. Specifically, 
+  # nCAT files have a tree line, then 4 lines with one scalar each, then a line with a vector.
+  # CAT files are the same, but have 5 lines with one scalar each
+  text_6 = NA
+  text_7 = NA
+  freq_start = NA
   
-  if( text[7] != "" ){
-    # The seventh line should be blank
+  if( text[7] == "" ){
+    text_7 =text[6]
+    freq_start = 8
+  } else if ( text[8] == ""  ) {
+    text_6 = text[6]
+    text_7 = text[7]
+    freq_start = 9
+  } else {
+    # The seventh or eight line should be blank
     return( NULL )
   }
   
@@ -55,7 +69,6 @@ PhylobayesSample = function(
     # Record does not start with a tree
     return( NULL )
   }
-  
   
   object = new( 
     "PhylobayesSample",
@@ -66,8 +79,9 @@ PhylobayesSample = function(
     x3 = as.numeric(text[3]),
     x4 = as.numeric(text[4]),
     x5 = as.numeric(text[5]),
-    x6 = text[6] %>% str_split("\t") %>% unlist() %>% as.numeric(),
-    frequencies = text[8:(length(text)-1)] %>% str_split("\t") %>% lapply( as.numeric ) %>% do.call( rbind, . ), 
+    x6 = as.numeric(text_6),
+    x7 = text_7 %>% str_split("\t") %>% unlist() %>% as.numeric(),
[…]

### 2020-07-19 manuscript/tests/testthat/test_pb.R

@@ -8,17 +8,26 @@ source( "../../phylobayes.R" )
 
 context("testing framework")
 
-# The example file was gcreated by manually editing 
-# Whelan2017_strict.phy_Poisson_CAT60_Chain_1_last500.chain
-# to obtain the last three generations of the chain
-
-example_name = "example.chain"
-
 test_that("Testing works", {
   expect_equal( 1, 1 )
 })
 
-context("single phylobayes chain record")
+
+# example.chain was created by manually editing 
+# Whelan2017_strict.phy_Poisson_CAT60_Chain_1_last500.chain
+# to obtain the last seven generations of the chain. This is 
+# an nCAT=60 chain.
+#
+# example_single.chain was created by taking a single record
+# from example.chain
+#
+# example_cat_single.chain was created by taking a single 
+# record from Whelan2017_strict.phy_Chain1_last50000.chain.
+# It is an unconstrained CAT run
+
+
+
+context("single nCAT phylobayes chain record")
 
 lines = read_lines( "example_single.chain" )
 
@@ -49,7 +58,7 @@ test_that("Missing last line returns NULL", {
 
 
 
-context("phylobayes chain files")
+context("phylobayes nCAT chain files")
 
 chain = parse_phylobayes_chain("example.chain")
 
@@ -67,3 +76,13 @@ test_that("Number of tips in trees are correct", {
   expect_true( all(x == 76) )
 })
 
+
+context("single CAT phylobayes chain record")
+
+lines = read_lines( "example_cat_single.chain" )
+
+pb = PhylobayesSample( lines, "example_single.chain", 1 )
+
+test_that("Tree has the correct number of tips", {
+  expect_equal( length(pb@tree$tip.label), 76 )
+})

### 2020-07-19 manuscript/explorations/phylobayes_explorations.rmd
makes nCAT60, nCAT100, and cat plots

@@
   allocation_counts = table(pb@allocation)
+  index = allocation_counts %>% names() %>% as.numeric()
+
+  if( ! all( index %in% rownames(pb@frequencies) ) ){
+    stop("Allocations are outside available category range")
+  }
+
+  # Consider only the subset of frequencies that are allocated
+  frequencies = pb@frequencies[ match(index , rownames(pb@frequencies) ), ]
 
   # Building on https://www.statmethods.net/advstats/mds.html
-  d = dist( pb@frequencies )
+  d = dist( frequencies )
   fit = cmdscale(d,eig=TRUE, k=2)
@@
   categories = tibble(
-    index = allocation_counts %>% names() %>% as.numeric(),
+    index = index,
     count = c(allocation_counts),
@@
     ggplot() +
-    geom_point(aes(x=x, y=y, size=count) ) +
+    geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) +
     ggtitle("MDS plot of rate categories") +

### 2020-07-19 manuscript/phylobayes.R

@@ -8,6 +8,9 @@ setOldClass("phylo")
 
 
 #' An S4 class to store a single sample from a phylobayes chain file
+#' In an unconstrained CAT run, the number of distinct categories, and 
+#' therefore rows in the frequencies matrix, is 5000, but only a subset 
+#' of them are allocated in the allocation vector
 #' 
 #' @slot allocation  The allocation vector
 setClass(
@@ -84,6 +87,8 @@ PhylobayesSample = function(
     frequencies = text[freq_start:(length(text)-1)] %>% str_split("\t") %>% lapply( as.numeric ) %>% do.call( rbind, . ), 
     allocation = text[length(text)] %>% str_split("\t") %>% unlist() %>% as.numeric()
   )
+  
+  rownames( object@frequencies ) = 0:( nrow( object@frequencies ) - 1 )
   
   if( object@x5 != nrow(object@frequencies) ){
      # Unexpected number of category frequency profiles

### 2020-08-11 manuscript/explorations/phylobayes_explorations.rmd

@@
   source( "../phylobayes.R" )
+  source( "../functions.R" )
 
@@
 
+  aa = PartitionedMultipleAlignment ("../../data_processed/matrices/Whelan2017_strict.phy", "../../data_processed/matrices/Whelan2017_strict.nex")
+
+  m = as.matrix(aa)
+  site_occupancy = colSums(m != "-" ) / nrow(m)
+
 ```
@@
 
-plot_chain = function( pb ){
+plot_chain = function( pb, site_occupancy ){
 
@@
 
-  categories = tibble(
+  # Calculate the mean occupancy for each site in the original matrix
+  occupancy = sapply( index, function(i){
+    sites = which( pb@allocation == i  )
+    mean( site_occupancy[ sites ] )
+  })
+
+  categories = 
+    tibble(
       index = index,
       count = c(allocation_counts),
+      occupancy = occupancy,
       x = fit$points[,1],
       y = fit$points[,2]
-  )
+    ) %>%
+    arrange(desc(count))
+
+  categories %<>% mutate(rank = 1:nrow(categories))
 
@@
     geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) +
-    ggtitle("MDS plot of rate categories") +
+    ggtitle("MDS plot of frequency categories") +
     theme_classic()
@@
     categories %>%
-    arrange(desc(count)) %>%
-    mutate(rank = 1:nrow(categories)) %>%
     ggplot() +
-      geom_point(aes(x=rank, y=count) ) +
-      ggtitle("Ordered count") +
+      geom_point(aes(x=rank, y=cumsum(count)) ) +
+      ggtitle("Cumulative sum of sites") +
       theme_classic()
@@
 
-plot_chain( pb )
+plot_chain( pb, site_occupancy )
 

### 2020-08-11 manuscript/manuscript.rmd
add CAT60, and CAT270 results in Fig. 4

@@
     Philippe2009$model,
-    levels = c("Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400", "Poisson+CAT440", "Poisson+CAT480"))
+    levels = c("Poisson+CAT60", "Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400", "Poisson+CAT440", "Poisson+CAT480"))
 
@@
   labs(x = "Number of CAT substitutional categories", y = "Support") +
-  ggtitle("Whelan2017_full_Choanozoa")
+  ggtitle("Whelan2017_full_Choanimalia")
 
@@
 
-ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"))
+ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)
 

### 2020-08-12 manuscript/manuscript.rmd
minor rewording, re-executed manuscript code

--- 2020-07-16 manuscript/manuscript.rmd
@@
   source( "functions.R" )
+  source( "phylobayes.R" )
 
@@
 These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
-Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used.
+Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
+This is very consistent across published analyses and our new panels of analyses.
+The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these respects.
+If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
 
@@
 We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
+
+```{r n_cat_render, fig.width=6, fig.height=12}
+
+# gg_n_cat
+
+```
+
+**Extended Data Fig. 3.5** The impact of adding a large number of categories.
 

### 2020-08-12 manuscript/manuscript.rmd
added chain analyses

@@
 
+There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
+It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available.
+This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
+Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
+This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
+One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
+Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
+If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
+That is exactly what we see (Extended Data Fig.
+3.5 a-b).
+This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
+
+We can further assess how, when a add parameters for a very large number of categories, the categories that are allocated to a small number of sites differ from those that are allocated to a large number of sites.
+It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
+We find support for this second scenario (Extended Data Fig.
+3.5 d-c).
+
 ``` {r sensitive, fig.width=7.2, fig.height=7.2}
@@
 
-# gg_n_cat
+summarise_sample = function( pb ){
+
+  allocation_counts = table(pb@allocation)
+  index = allocation_counts %>% names() %>% as.numeric()
+
+  if( ! all( index %in% rownames(pb@frequencies) ) ){
+    stop("Allocations are outside available category range")
+  }
+
+  # Consider only the subset of frequencies that are allocated
+  frequencies = pb@frequencies[ match(index , rownames(pb@frequencies) ), ]
+
+  # Building on https://www.statmethods.net/advstats/mds.html
+  d = dist( frequencies )
+  fit = cmdscale(d,eig=TRUE, k=2)
+
+  categories = 
+    tibble(
+      index = index,
+      count = c(allocation_counts),
+      x = fit$points[,1],
+      y = fit$points[,2]
+    ) %>%
+    arrange(desc(count)) %>%
+    mutate( cumsum = cumsum(count))
+
+  categories %<>% mutate(rank = 1:nrow(categories))
+
+}
+
+phil_cat_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
+
+phil_cat60_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
+
+whel_cat_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
+
+whel_cat60_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
+
+pb_samples = list(
+  summarise_sample( phil_cat_c1 ) %>% 
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
+  summarise_sample( phil_cat60_c1 ) %>% 
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
+  summarise_sample( whel_cat_c1 ) %>% 
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
+  summarise_sample( whel_cat60_c1 ) %>% 
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
+) %>%
+  bind_rows() %>%
+  mutate( analysis = paste(matrix, model, sep=" ")) %>%
+  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
+
+pb_samples %>% 
+ggplot() + geom_line(aes(x = rank, y = cumsum, col=analysis))
+
+gg_rank_phil = 
+  pb_samples %>% 
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) + 
+  theme(legend.position="right")
+
+gg_rank_whel = 
+  pb_samples %>% 
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) + 
+  theme(legend.position="right")
+
+gg_mds_phil = 
+ pb_samples %>% 
+  filter( matrix=="Philippe2009" ) %>% 
+  filter( model=="Poisson+CAT" ) %>% 
+  ggplot() +
+  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
+  theme(legend.position="right")
+
+gg_mds_whel = 
+ pb_samples %>% 
+  filter( matrix=="Whelan2017_strict" ) %>% 
+  filter( model=="Poisson+CAT" ) %>% 
+  ggplot() +
+  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
+  theme(legend.position="right")
+
+gg_rank_phil + ggtitle("(A) Philippe2009")
+
+gg_rank_whel + ggtitle("(B) Whelan2017")
+
+gg_mds_phil + ggtitle("(C) Philippe2017 Poisson+CAT")
+
+gg_mds_phil + ggtitle("(D) Whelan2017 Poisson+CAT")
 

### 2020-08-12 manuscript/phylobayes.R

@@ -162,3 +162,9 @@ parse_phylobayes_chain = function( file_name ){
   samples
 }
 
+
+parse_phylobayes_last_sample = function( file_name ){
+  chain = parse_phylobayes_chain( file_name )
+  chain[[length(chain)]]
+}
+

### 2020-08-14 manuscript/manuscript.rmd
added global mds plots, did not knit

@@
 
-There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
-It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available.
-This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
-Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
-This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
-One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
-Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
-That is exactly what we see (Extended Data Fig.
-3.5 a-b).
-This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
-
-We can further assess how, when a add parameters for a very large number of categories, the categories that are allocated to a small number of sites differ from those that are allocated to a large number of sites.
-It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
-Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
-We find support for this second scenario (Extended Data Fig.
-3.5 d-c).
-
-``` {r sensitive, fig.width=7.2, fig.height=7.2}
-
-Philippe2009 =
-  analyses_sensitive %>%
-  filter(matrix == "Philippe2009_only_choanozoa" )
-
-Philippe2009$model =
-  factor(
-    Philippe2009$model,
-    levels = c("Poisson+CAT60", "Poisson+CAT90", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180"))
-
-p1 = Philippe2009 %>%
-  ggplot( aes(model) ) +
-  geom_line( aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1 ), color = ("#E69F00")) +
-  geom_point( aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
-  geom_line( aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) +
-  geom_point( aes(y = `support_porifera_sister` ), color = ("#009E73")) +
-  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Number of CAT substitutional categories", y = "Support") +
-  ggtitle("Philippe2009_Choanimalia")
-
-Philippe2009 =
-  analyses_sensitive %>%
-  filter(matrix == "Philippe2009_only_Holozoa")
-
-Philippe2009$model =
-  factor(
-    Philippe2009$model,
-    levels = c("Poisson+CAT60", "Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400", "Poisson+CAT440", "Poisson+CAT480"))
-
-p2 = Philippe2009 %>%
-  ggplot( aes(model)) +
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
-  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) +
-  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
-  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Number of CAT substitutional categories", y = "Support") +
-  ggtitle("Philippe2009_Holozoa")
-
-Whelan2017_full_only_choanozoa =
-  analyses_sensitive %>%
-  filter(matrix == "Whelan2017_full_only_choanozoa")
-
-Whelan2017_full_only_choanozoa$model =
-  factor(
-    Whelan2017_full_only_choanozoa$model,
-    levels = c("Poisson+CAT60", "Poisson+CAT340", "Poisson+CAT380", "Poisson+CAT420", "Poisson+CAT460"))
-
-p3 = Whelan2017_full_only_choanozoa %>%
-  ggplot( aes(model)) +
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
-  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) +
-  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
-  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Number of CAT substitutional categories", y = "Support") +
-  ggtitle("Whelan2017_full_Choanimalia")
-
-Whelan2017_strict =
-  analyses_sensitive %>%
-  filter(matrix == "Whelan2017_strict" )
-
-Whelan2017_strict$model =
-  factor(
-    Whelan2017_strict$model,
-    levels = c("Poisson+CAT60", "Poisson+CAT70", "Poisson+CAT80", "Poisson+CAT90", "Poisson+CAT100",
-             "Poisson+CAT110", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180", "Poisson+CAT360"))
-
-p4 = Whelan2017_strict %>%
-  ggplot( aes(model)) +
-  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
-  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
-  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) +
-  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
-  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Number of CAT substitutional categories", y = "Support") +
-  ggtitle("Whelan2017_strict")
-
-ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)
-
-ggsave("figures/Figure4_raw.pdf", width=7.2, height=7.2)
-
-```
-**Fig.
-4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
-Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories.
-(A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
-Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values are from the sensitive analyses are shown in Supplementary Table 6.
-
-### Phylogenetic signal
-
-To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
-3).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
-This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
-
-Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
-In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
-3; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-3).
-For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
-
-The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
-
-### Other approaches to accommodating site heterogeneity
-
-Other approaches have been taken to addressing base compositional heterogeneity across taxa.
-For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids.
-These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds].
-Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-However, Our own analyses (Supplementary text, Supplementary Figs.
-4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
-Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
-
-## The current state of understanding and future directions
-
-Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem.
-By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
-This demonstrates an important interaction between model selection and taxon sampling.
-Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
-Site-homogeneous models have a single category that is applied to all sites.
-C models, as implemented in IQ-TREE, have 10-60 categories.
-CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT.
-When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories.
-As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig.
-4).
-As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.
-
-This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence.
-Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
-Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
-
-Overcoming the challenges associated with this difficult and important question goes beyond models.
-In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie].
-In particular, there are very few chromosome-level genome assemblies outside of Bilateria.
-There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
-[XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling.
-Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
-We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
-
-## Methods
-
-All tree files, intermediate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publication.
-
-### Data selection and wrangling
-
-We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository.
-We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
-All changes made are tracked with git.
-
-### Matrix comparison and annotation
-
-#### Taxon name reconciliation
-
-We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
-
-#### Sequence comparisons
-
-Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
-Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
-
-We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
-We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
-We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
-We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
-
-#### Partition network
-
-We used the sequence similarity comparisons described above to compare partitions.
-
-We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
-We extracted each connected component from this network.
-We further split these components if the the most connected node (i.e.
-most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
-We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See partition_network_summary in Rdata for a summary tally of each part of the comparison.
-
-### Phylogenetic analyses
-
-#### Phylogenetic analyses in IQ-TREE
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
-For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
-For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
-
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, poisson+C60 and associated best-fit model identified above.
-Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
-Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
-To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
-For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
-2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
-It should be noted that the main conclusion of Simion *et al*.
-was also based on selection of 25% of genes for their jackknife approach.
-
-#### Outgroup taxa sampling with C60 and CAT models
-Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
-These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
-The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
-Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, GTR+CAT models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
-For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
-Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
-Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
-A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time.
-For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
-We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
-
-#### Phylogenetic distribution of support
-To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanomalia and the full matrix).
-For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
-The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to rest of animals.
-The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].
-
-For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency.
-Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
-For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
-Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15
-
-#### Sensitive analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different number of categories were used.
-
-#### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
-The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
-
-The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
-This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
-The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
-
-The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
-To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
-This creates new recodings that have the same sized groups as SR-6.
-The new recodings were, from `random-00` to `random-03`:
-
-    MPKE AY IDGQ TRC WNLF SVH
-    EIFT WL QVPG NKM SCHA RYD
-    LCQS GK WPTI VRD YEFN MAH
-    IWQV TY KDLM ESH APCF GRN
-
-To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
-
-The 20 state matrices are the same across all analyses since they are not recoded.
-Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-3B) give insight into the technical variance of the inference process.
-
-Each new matrix was analyzed with PhyloBayes-MPI.
-Analyses were run for 1000 generations, and a 200 generation burnin applied.
-The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
-
-The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
-
-## Acknowledgements
-
-We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
-
-## Author contributions
-
-## Extended Tables
-
-**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
-```{r table_study_summary}
-table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
-opts <- options(knitr.kable.NA = "")
-knitr::kable(table_study_summary)
-
-```
-
-**Extended Data Table 2.** The models selected by ModelFinder for each matrix.
-```{r table_modelfinder}
-analyses_new %>%
-  filter(modelfinder == TRUE) %>%
-  select( matrix, clade, result, model_summary ) %>%
-  kable()
-
-```
-
-## Extended Figures
-
-```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-  lapply(
-    sequence_matrices[no_subsets],
-    function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x, y))) %>%
-  unlist(recursive = FALSE) %>%
-  bind_rows()
-
-matrix_overlap_rectangles %>%
-  ggplot( ) +
-    scale_x_continuous(name = "Genes") +
-    scale_y_continuous(name = "Species") +
-    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
-    facet_grid( matrix_1 ~ matrix_2, scales = "free" ) +
-    theme( axis.text.x = element_text(angle = 45, hjust = 1),
-           strip.text.y = element_text(angle = 0),
-           strip.text.x = element_text(angle = 60))
-
-ggsave("figures/Extended_Data_Fig1_raw.pdf", width=20, height=12)
-
-```
-
-**Extended Data Fig.
-1.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
-```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
-
-ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
-busco = read_tsv("../data_processed/tables/busco_gene.tsv")
-
-ribo$matrix =
-  factor(
-    ribo$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-             "Whelan2017_full", "Whelan2017_strict") )
-
-p1 = ribo %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
-
-busco$matrix =
-  factor(
-    busco$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-               "Whelan2017_full", "Whelan2017_strict") )
-
-busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
-
-p2 = busco %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
-
- ggarrange(p1, p2, labels = c("A", "B"))
-
-ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=4)
-
-```
-
-**Extended Data Fig.
-2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
-(C).
-The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
-
-```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-
-p=au_tests %>%
-  unite(Matrix, c(manuscript, matrix, model)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
-  ggplot(aes(x = Matrix, y = percent, fill = result)) +
-    geom_bar(stat = "identity") +
-    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
-    coord_flip() +
-    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
-
-ggsave("figures/Extended_Data_Fig3_raw.pdf", width=6, height=12)
-
-```
-
-![](figures/Extended_Data_Fig3_manuscript.png)
-
-**Extended Data Fig.
-3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
-The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
-Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models.
-The GLS values for each gene in each data matrix are provided in Table S5.
-We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
-
-```{r n_cat_render, fig.width=6, fig.height=12}
+```{r n_cat, fig.width=6, fig.height=12}
 
@@
       y = fit$points[,2]
-    ) %>%
+    )
+
+  if( ! all( categories$index == rownames(frequencies) ) ){
+    stop("category indexes and frequency vector names do not match")
+  }
+
+  frequencies_tib = 
+    as_tibble( frequencies )
+
+  names( frequencies_tib ) = paste0( "aa.", names( frequencies_tib ) )
+
+  categories %<>% 
+    bind_cols( frequencies_tib ) %>%
     arrange(desc(count)) %>% 
-    mutate( cumsum = cumsum(count))
-
-  categories %<>% mutate(rank = 1:nrow(categories))
+    mutate( cumsum = cumsum(count)) %<>% 
+    mutate(rank = 1:nrow(categories))
 
@@
 
+pb_frequencies = pb_samples %>% select( starts_with("aa.") ) %>% data.matrix()
+
+fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
+
+pb_samples %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
+
 pb_samples %>% 
@@
   filter( matrix=="Philippe2009" ) %>% 
-  filter( model=="Poisson+CAT" ) %>% 
   ggplot() +
-  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) + 
   theme(legend.position="right")
@@
   filter( matrix=="Whelan2017_strict" ) %>% 
-  filter( model=="Poisson+CAT" ) %>% 
   ggplot() +
-  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
   theme(legend.position="right")
 
-gg_rank_phil + ggtitle("(A) Philippe2009")
-
-gg_rank_whel + ggtitle("(B) Whelan2017")
-
-gg_mds_phil + ggtitle("(C) Philippe2017 Poisson+CAT")
-
-gg_mds_phil + ggtitle("(D) Whelan2017 Poisson+CAT")
+gg_chain = 
+ggarrange(
+  gg_rank_phil, 
+  gg_rank_whel, 
+  gg_mds_phil, 
+  gg_mds_whel, 
+  ncol  = 2, 
+  nrow = 2, 
+  labels = c("A Philippe2009", "B Whelan2017", "C", "D"), 
+  common.legend = TRUE
+)
+
+```
+
+There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
+It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available.
+This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
+Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
+This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
+One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
+Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
+If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
+That is exactly what we see (Extended Data Fig.
+3.5 a-b).
+This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
+
+We can further assess how, when a add parameters for a very large number of categories, the categories that are allocated to a small number of sites differ from those that are allocated to a large number of sites.
+It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
+We find support for this second scenario (Extended Data Fig.
+3.5 d-c).
+
+``` {r sensitive, fig.width=7.2, fig.height=7.2}
+
+Philippe2009 =
+  analyses_sensitive %>%
+  filter(matrix == "Philippe2009_only_choanozoa" )
+
+Philippe2009$model =
+  factor(
+    Philippe2009$model,
+    levels = c("Poisson+CAT60", "Poisson+CAT90", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180"))
+
+p1 = Philippe2009 %>%
+  ggplot( aes(model) ) +
+  geom_line( aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1 ), color = ("#E69F00")) +
+  geom_point( aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line( aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) +
+  geom_point( aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") +
+  ggtitle("Philippe2009_Choanimalia")
+
+Philippe2009 =
+  analyses_sensitive %>%
+  filter(matrix == "Philippe2009_only_Holozoa")
+
+Philippe2009$model =
+  factor(
+    Philippe2009$model,
+    levels = c("Poisson+CAT60", "Poisson+CAT270", "Poisson+CAT360", "Poisson+CAT400", "Poisson+CAT440", "Poisson+CAT480"))
+
+p2 = Philippe2009 %>%
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 2), color = ("#009E73")) +
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") +
+  ggtitle("Philippe2009_Holozoa")
+
+Whelan2017_full_only_choanozoa =
+  analyses_sensitive %>%
+  filter(matrix == "Whelan2017_full_only_choanozoa")
+
+Whelan2017_full_only_choanozoa$model =
+  factor(
+    Whelan2017_full_only_choanozoa$model,
+    levels = c("Poisson+CAT60", "Poisson+CAT340", "Poisson+CAT380", "Poisson+CAT420", "Poisson+CAT460"))
+
+p3 = Whelan2017_full_only_choanozoa %>%
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) +
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") +
+  ggtitle("Whelan2017_full_Choanimalia")
+
+Whelan2017_strict =
+  analyses_sensitive %>%
+  filter(matrix == "Whelan2017_strict" )
+
+Whelan2017_strict$model =
+  factor(
+    Whelan2017_strict$model,
+    levels = c("Poisson+CAT60", "Poisson+CAT70", "Poisson+CAT80", "Poisson+CAT90", "Poisson+CAT100",
+             "Poisson+CAT110", "Poisson+CAT120", "Poisson+CAT150", "Poisson+CAT180", "Poisson+CAT360"))
+
+p4 = Whelan2017_strict %>%
+  ggplot( aes(model)) +
+  geom_line(aes(y = `support_ctenophora_sister`, colour = "Ctenophora-sister", group = 1), color = ("#E69F00")) +
+  geom_point(aes(y = `support_ctenophora_sister` ), color = ("#E69F00")) +
+  geom_line(aes(y = `support_porifera_sister`, colour = "Porifera-sister", group = 1), color = ("#009E73")) +
+  geom_point(aes(y = `support_porifera_sister` ), color = ("#009E73")) +
+  scale_fill_manual(values = c("#E69F00", "#ebd196")) +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Number of CAT substitutional categories", y = "Support") +
+  ggtitle("Whelan2017_strict")
+
+ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)
+
+ggsave("figures/Figure4_raw.pdf", width=7.2, height=7.2)
+
+```
+**Fig.
+4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
+Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories.
+(A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict.
+Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
+Support values are from the sensitive analyses are shown in Supplementary Table 6.
+
+### Phylogenetic signal
+
+To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
+3).
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
+Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
+This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
+
+Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
+In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
+3; Supplementary Table 7).
+In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+3).
+For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
+
+The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
+
+### Other approaches to accommodating site heterogeneity
+
+Other approaches have been taken to addressing base compositional heterogeneity across taxa.
+For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids.
+These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds].
+Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
+However, Our own analyses (Supplementary text, Supplementary Figs.
+4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
+
+## The current state of understanding and future directions
+
+Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem.
+By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
+This demonstrates an important interaction between model selection and taxon sampling.
+Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
+Site-homogeneous models have a single category that is applied to all sites.
+C models, as implemented in IQ-TREE, have 10-60 categories.
+CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT.
+When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories.
+As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig.
+4).
+As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.
+
+This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence.
+Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
+Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
+
+Overcoming the challenges associated with this difficult and important question goes beyond models.
+In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie].
+In particular, there are very few chromosome-level genome assemblies outside of Bilateria.
+There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
+Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
+[XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling.
+Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
+We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
+
+## Methods
+
+All tree files, intermediate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publication.
+
+### Data selection and wrangling
+
+We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository.
+We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
+All changes made are tracked with git.
+
+### Matrix comparison and annotation
+
+#### Taxon name reconciliation
+
+We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
+We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript.
+For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
+
+#### Sequence comparisons
+
+Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
+Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
+
+We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
+We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
+We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
+We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+
+#### Partition network
+
+We used the sequence similarity comparisons described above to compare partitions.
+
+We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
+We extracted each connected component from this network.
+We further split these components if the the most connected node (i.e.
+most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
+We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+See partition_network_summary in Rdata for a summary tally of each part of the comparison.
+
+### Phylogenetic analyses
+
+#### Phylogenetic analyses in IQ-TREE
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
+For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
+For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
+
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, poisson+C60 and associated best-fit model identified above.
+Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
+Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
+To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
+For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
+2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+It should be noted that the main conclusion of Simion *et al*.
+was also based on selection of 25% of genes for their jackknife approach.
+
+#### Outgroup taxa sampling with C60 and CAT models
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
+These datasets include the same set of genes but differ in the composition of outgroup species.
+Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
+For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
+The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
+Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
+To minimize computational burden, GTR+CAT models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
+Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
+Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
+A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time.
+For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
+We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
+
+#### Phylogenetic distribution of support
+To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
+We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanomalia and the full matrix).
+For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
+The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to rest of animals.
+The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].
+
+For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency.
+Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
+For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
+Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15
+
+#### Sensitive analyses with different number of substitutional categories
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different number of categories were used.
+
+#### Performance of data-recoding
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
+The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
+
+The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
+This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
+
+The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
+To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
+This creates new recodings that have the same sized groups as SR-6.
+The new recodings were, from `random-00` to `random-03`:
+
+    MPKE AY IDGQ TRC WNLF SVH
+    EIFT WL QVPG NKM SCHA RYD
+    LCQS GK WPTI VRD YEFN MAH
+    IWQV TY KDLM ESH APCF GRN
+
+To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
+
+The 20 state matrices are the same across all analyses since they are not recoded.
+Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
+3B) give insight into the technical variance of the inference process.
+
+Each new matrix was analyzed with PhyloBayes-MPI.
+Analyses were run for 1000 generations, and a 200 generation burnin applied.
+The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
+
+The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+
+## Acknowledgements
+
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
+
+## Author contributions
+
+## Extended Tables
+
+**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
+```{r table_study_summary}
+table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
+opts <- options(knitr.kable.NA = "")
+knitr::kable(table_study_summary)
+
+```
+
+**Extended Data Table 2.** The models selected by ModelFinder for each matrix.
+```{r table_modelfinder}
+analyses_new %>%
+  filter(modelfinder == TRUE) %>%
+  select( matrix, clade, result, model_summary ) %>%
+  kable()
+
+```
+
+## Extended Figures
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+  lapply(
+    sequence_matrices[no_subsets],
+    function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x, y))) %>%
+  unlist(recursive = FALSE) %>%
+  bind_rows()
+
+matrix_overlap_rectangles %>%
+  ggplot( ) +
+    scale_x_continuous(name = "Genes") +
+    scale_y_continuous(name = "Species") +
+    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
+    facet_grid( matrix_1 ~ matrix_2, scales = "free" ) +
+    theme( axis.text.x = element_text(angle = 45, hjust = 1),
+           strip.text.y = element_text(angle = 0),
+           strip.text.x = element_text(angle = 60))
+
+ggsave("figures/Extended_Data_Fig1_raw.pdf", width=20, height=12)
+
+```
+
+**Extended Data Fig.
+1.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
+
+```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
+
+ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
+busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+
+ribo$matrix =
+  factor(
+    ribo$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+             "Whelan2017_full", "Whelan2017_strict") )
+
+p1 = ribo %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
+
+busco$matrix =
+  factor(
+    busco$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+               "Whelan2017_full", "Whelan2017_strict") )
+
+busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
+
+p2 = busco %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
+
+ ggarrange(p1, p2, labels = c("A", "B"))
+
+ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=4)
+
+```
+
+**Extended Data Fig.
+2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+(A).
+The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
+(C).
+The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
+
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+
+p=au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+  ggplot(aes(x = Matrix, y = percent, fill = result)) +
+    geom_bar(stat = "identity") +
+    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
+    coord_flip() +
+    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
+
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=6, height=12)
+
+```
+
+![](figures/Extended_Data_Fig3_manuscript.png)
+
+**Extended Data Fig.
+3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
+Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models.
+The GLS values for each gene in each data matrix are provided in Table S5.
+We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
+
+```{r n_cat_render, fig.width=6, fig.height=12}
+
+gg_chain
 

### 2020-08-14 manuscript/phylobayes.R

@@ -91,11 +91,6 @@ PhylobayesSample = function(
   
   rownames( object@frequencies ) = 0:( nrow( object@frequencies ) - 1 )
   
-  if( object@ncat != nrow(object@frequencies) ){
-     # Unexpected number of category frequency profiles
-    return( NULL )
-  }
-  
   # The order of the amino acid vector in PhyloBayes is specified here:
   # https://github.com/bayesiancook/pbmpi/blob/ba03ab5146140c9beab673735e826a10ac0ea6ab/sources/BiologicalSequences.h#L41
   # The trailing gap character is then removed here:
@@ -103,7 +98,11 @@ PhylobayesSample = function(
   
   colnames( object@frequencies ) = 
     c( "A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","W","Y" )
-
+  
+  if( object@ncat != nrow(object@frequencies) ){
+     # Unexpected number of category frequency profiles
+    return( NULL )
+  }
   
   object
 }

### 2020-08-14 manuscript/manuscript.rmd
clean up code and refine text, did not rerun

--- 2020-08-12 manuscript/manuscript.rmd
@@
 
+```{r n_cat, fig.width=6, fig.height=12}
+
+# Return the allocation midpoint, ie the percent of sites that are allocated to the 
+# 50% of the least frequent categories
+allocation_midpoint = function ( pb_summary ) {
+  x = pb_summary %>% pull( count ) %>% sort(decreasing = FALSE)
+  x = x/ sum(x)
+  y = cumsum( x )
+  return ( round( median( y ) *100, 2 ) )
+}
+
+# Parse the last sample from chain 1 of each analysis
+phil_cat_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
+
+phil_cat60_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
+
+whel_cat_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
+
+whel_cat60_c1 = 
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
+
+# Create a single tibble with summaries of all analyses
+pb_summaries = list(
+  summarise_sample( phil_cat_c1 ) %>% 
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
+  summarise_sample( phil_cat60_c1 ) %>% 
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
+  summarise_sample( whel_cat_c1 ) %>% 
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
+  summarise_sample( whel_cat60_c1 ) %>% 
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
+) %>%
+  bind_rows() %>%
+  mutate( analysis = paste(matrix, model, sep=" ")) %>%
+  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
+
+pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
+
+# Identify the midpoint of each set of allocations
+mid_phil_cat = 
+  pb_summaries %>% 
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  allocation_midpoint()
+
+mid_phil_ncat60 = 
+  pb_summaries %>% 
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+nCAT60" ) %>%
+  allocation_midpoint()
+
+mid_whel_cat = 
+  pb_summaries %>% 
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  allocation_midpoint()
+
+mid_whel_ncat60 = 
+  pb_summaries %>% 
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+nCAT60" ) %>%
+  allocation_midpoint()
+
+n_categories_phil_cat =
+  pb_summaries %>% 
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  nrow()
+
+n_categories_whel_cat =
+  pb_summaries %>% 
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  nrow()
+
+# Perform global MDS analysys
+fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
+pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
+
+# Generate plots
+gg_rank_phil = 
+  pb_summaries %>% 
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) + 
+  theme(legend.position="right")
+
+gg_rank_whel = 
+  pb_summaries %>% 
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) + 
+  theme(legend.position="right")
+
+gg_mds_phil = 
+ pb_summaries %>% 
+  filter( matrix=="Philippe2009" ) %>% 
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) + 
+  theme(legend.position="right")
+
+gg_mds_whel = 
+ pb_summaries %>% 
+  filter( matrix=="Whelan2017_strict" ) %>% 
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
+  theme(legend.position="right")
+
+gg_chain = 
+ggarrange(
+  gg_rank_phil, 
+  gg_rank_whel, 
+  gg_mds_phil, 
+  gg_mds_whel, 
+  ncol  = 2, 
+  nrow = 2, 
+  labels = c("A Philippe2009", "B Whelan2017", "C", "D"), 
+  common.legend = TRUE
+)
+
+```
+
+There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
+It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available.
+This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
+Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
+This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
+One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
+Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
+If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
+That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig.
+3.5 a-b), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
+The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
+This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig.
+3.5 a-b).
+This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
+
+We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundred.
+It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
+We find support for this second scenario (Extended Data Fig.
+3.5 d-c).
+
+Taken together, these results indicate that when CAT models are run without constraint on the number of categories, they select a very large number of categories.
+A large number of these categories are allocated to a small number of sites that have small differences from other categories.
+Given the very large number of parameters that must be estimated for this large number of rare categories, this suggests that unconstrained CAT analyses are overparameterized.
+While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQtree that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
+In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of caetegories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.
+
 ``` {r sensitive, fig.width=7.2, fig.height=7.2}
@@
 
-# gg_n_cat
+gg_chain
 

### 2020-08-14 manuscript/phylobayes.R

@@ -167,3 +167,51 @@ parse_phylobayes_last_sample = function( file_name ){
   chain[[length(chain)]]
 }
 
+
+
+#' Summarise each category
+#'
+#' @param pb A PhylobayesSample object
+#' @return A tibble, with one row per equilibrium frequency category
+#' @export
+
+summarise_sample = function( pb ){
+  
+  allocation_counts = table(pb@allocation)
+  index = allocation_counts %>% names() %>% as.numeric()
+  
+  if( ! all( index %in% rownames(pb@frequencies) ) ){
+    stop("Allocations are outside available category range")
+  }
+  
+  # Consider only the subset of frequencies that are allocated
+  frequencies = pb@frequencies[ match(index , rownames(pb@frequencies) ), ]
+  
+  # Building on https://www.statmethods.net/advstats/mds.html
+  d = dist( frequencies )
+  fit = cmdscale(d,eig=TRUE, k=2)
+  
+  categories = 
+    tibble(
+      index = index,
+      count = c(allocation_counts),
+      x = fit$points[,1],
+      y = fit$points[,2]
+    ) 
+  
+  if( ! all( categories$index == rownames(frequencies) ) ){
+    stop("category indexes and frequency vector names do not match")
+  }
+  
+  frequencies_tib = 
+    as_tibble( frequencies )
+  
+  names( frequencies_tib ) = paste0( "aa.", names( frequencies_tib ) )
+  
+  categories %<>% 
+    bind_cols( frequencies_tib ) %>%
+    arrange(desc(count)) %>% 
+    mutate( cumsum = cumsum(count)) %<>% 
+    mutate(rank = 1:nrow(categories))
+  
+}

### 2020-08-14 manuscript/manuscript.rmd
added legend and refined figure

@@
 
-```{r n_cat, fig.width=6, fig.height=12}
+```{r n_cat }
 
@@
   geom_line(aes(x=rank, y=count, col=model)) + 
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
   theme(legend.position="right")
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) + 
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
   theme(legend.position="right")
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
   theme(legend.position="right")
@@
 
-```{r n_cat_render, fig.width=6, fig.height=12}
+```{r n_cat_render, fig.width=6, fig.height=7}
 
@@
 
-**Extended Data Fig. 3.5** The impact of adding a large number of categories.
+**Extended Data Fig.
+3.5** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+These plots are for singe sampled isolated from the end of PhyloBayes chain files.
+(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis.
+The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
+The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
+(C,D) MDS plots for the amino acid frequencies.
+Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies.
+A single MDS analysis was done for all analyses, so that the projections are the same in each plot.
+For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories.
+This adds a considerable number of parameters for describing a similar overall space.
 

### 2020-08-14 manuscript/manuscript.rmd
first pass to the manuscript, fix several NAs, update figure orders, fix a couple misspellings

@@
 **Fig.
-2.** An overview of previous phylogenomic studieson animals.
+2.** An overview of previous phylogenomic studies on animals.
 (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study.
@@
 
-cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson+CAT`
-cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR+CAT`
+cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
+cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR-CAT`
 
@@
 Differences in analysis results could therefore be due to differences in the number of categories.
-The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses.
+The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses.
 This requires a very large number of additional parameters that must be estimated from the data.
@@
 
-These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
-Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
-Analyses under other conditions either support Ctenophora-sister or are unresolved.
-This is very consistent across published analyses and our new panels of analyses.
-The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these respects.
-If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
-
-```{r n_cat }
-
+```{r n_cat, fig.width=6, fig.height=12}
 # Return the allocation midpoint, ie the percent of sites that are allocated to the
@@
 }
-
 # Parse the last sample from chain 1 of each analysis
@@
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
-
 phil_cat60_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
-
 whel_cat_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
-
 whel_cat60_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
-
 # Create a single tibble with summaries of all analyses
@@
   mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
-
 pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
-
 # Identify the midpoint of each set of allocations
@@
   allocation_midpoint()
-
 mid_phil_ncat60 =
@@
   allocation_midpoint()
-
 mid_whel_cat =
@@
   allocation_midpoint()
-
 mid_whel_ncat60 =
@@
   allocation_midpoint()
-
 n_categories_phil_cat =
@@
   nrow()
-
 n_categories_whel_cat =
@@
   nrow()
-
 # Perform global MDS analysys
@@
 pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
-
 # Generate plots
@@
   geom_line(aes(x=rank, y=count, col=model)) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right")
-
+  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
 gg_rank_whel =
@@
   geom_line(aes(x=rank, y=count, col=model)) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right")
-
+  theme(legend.position="right") + ggtitle("Whelan2017_strict")
 gg_mds_phil =
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
   theme(legend.position="right")
-
 gg_mds_whel =
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
   theme(legend.position="right")
-
 gg_chain =
@@
   nrow = 2,
-  labels = c("A Philippe2009", "B Whelan2017", "C", "D"), 
+  labels = c("A", "B", "C", "D"),
   common.legend = TRUE
 )
-
-```
+```
+These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
+Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
+Analyses under other conditions either support Ctenophora-sister or are unresolved.
+This is very consistent across published analyses and our new panels of analyses.
+The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these respects.
+If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
 
 There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
-It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available.
+It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available.
 This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
@@
 That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig.
-3.5 a-b), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
+3A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig.
-3.5 a-b).
-This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
+3A-B).
+This suggests that support for Porifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
 
 We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundred.
-It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
-Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
+It could be that the sites that are allocated to rate categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
 We find support for this second scenario (Extended Data Fig.
-3.5 d-c).
+3C-D).
 
@@
 Given the very large number of parameters that must be estimated for this large number of rare categories, this suggests that unconstrained CAT analyses are overparameterized.
-While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQtree that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
-In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of caetegories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.
+While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
+In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of  categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.
 
@@
 To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
-3).
+4).
 By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
@@
 In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
-3; Supplementary Table 7).
+4; Supplementary Table 7).
 In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-3).
+4).
 For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
@@
 For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
-Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15
+Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].
 
@@
 
+To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Posson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
+The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
+
 #### Performance of data-recoding
@@
 Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-3B) give insight into the technical variance of the inference process.
+5) give insight into the technical variance of the inference process.
 
@@
 
-The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+The statistics presented in  Extended Data Figs. 5 and 6 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
+
+```{r n_cat_render, fig.width=6, fig.height=7}
+gg_chain
+```
+
+**Extended Data Fig.
+3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+These plots are for singe sampled isolated from the end of PhyloBayes chain files.
+(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis.
+The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
+The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
+(C,D) MDS plots for the amino acid frequencies.
+Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies.
+A single MDS analysis was done for all analyses, so that the projections are the same in each plot.
+For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories.
+This adds a considerable number of parameters for describing a similar overall space.
 
@@
 **Extended Data Fig.
-3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
@@
 
-```{r n_cat_render, fig.width=6, fig.height=7}
-
-gg_chain
-
-```
+![](figures/Figure_random_rocoding.png)
 
 **Extended Data Fig.
-3.5** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
-These plots are for singe sampled isolated from the end of PhyloBayes chain files.
-(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis.
-The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
-The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
-(C,D) MDS plots for the amino acid frequencies.
-Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies.
-A single MDS analysis was done for all analyses, so that the projections are the same in each plot.
-For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories.
-This adds a considerable number of parameters for describing a similar overall space.
-
-![](figures/Figure_random_rocoding.png)
-
-**Extended Data Fig.
-4.** (A) The two alternative hypotheses for deep animal relationships considered here.
+5.** (A) The two alternative hypotheses for deep animal relationships considered here.
 Relationships that are not part of these hypotheses are shown as unresolved polytomies.
@@
 **Extended Data Fig.
-5.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+6.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 **Extended Data Fig.
-6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
 Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
@@
 They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig.
+Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
 5).
@@
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-6).
+7).
 
@@
 They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
-Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister.
+Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister).
 Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
@@
 This provides 32 analyses that they report in their Table 3 and that we present graphically here as Extended Data Fig.
-6.
+5.
 There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Extended Data Fig.
-6).
+5).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
-5).
+6).
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig.
-5), indicating that it did not converge as well as other analyses.
+6), indicating that it did not converge as well as other analyses.
 
@@
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-3B).
+7).
 

### 2020-08-14 manuscript/manuscript.rmd
updated extended fig.3 based on most recent commit

@@
 
-```{r n_cat, fig.width=6, fig.height=12}
+```{r n_cat}
 # Return the allocation midpoint, ie the percent of sites that are allocated to the
@@
 }
+
 # Parse the last sample from chain 1 of each analysis
@@
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
+
 phil_cat60_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
+
 whel_cat_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
+
 whel_cat60_c1 =
   parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
+
 # Create a single tibble with summaries of all analyses
@@
   mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
+
 pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
+
 # Identify the midpoint of each set of allocations
@@
   allocation_midpoint()
+
 mid_phil_ncat60 =
@@
   allocation_midpoint()
+
 mid_whel_cat =
@@
   allocation_midpoint()
+
 mid_whel_ncat60 =
@@
   allocation_midpoint()
+
 n_categories_phil_cat =
@@
   nrow()
+
 n_categories_whel_cat =
@@
   nrow()
+
 # Perform global MDS analysys
@@
 pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
+
 # Generate plots
@@
   geom_line(aes(x=rank, y=count, col=model)) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
   theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
+
 gg_rank_whel =
@@
   geom_line(aes(x=rank, y=count, col=model)) +
-  theme(legend.position="right") + ggtitle("Whelan2017_strict")
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
+
 gg_mds_phil =
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
   theme(legend.position="right")
+
 gg_mds_whel =
@@
   geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
   theme(legend.position="right")
+
 gg_chain =

### 2020-08-15 manuscript/manuscript.rmd
resolved some comments and other minor edits

@@
 
-There has been considerable debate about the placement of the root in the animal tree of life.
+There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree.
 This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
 Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
-[XXXIt would be helpful if you also stated why this is a problem: Antonis] These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
+These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
 Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses.
@@
 Our new analyses are consistent with this.
-Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of substitutional categories.
-[XXXTo me this part is a diversion – I would state your key inference (“our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories”) upfront.
-You can then explain how your results fit / shed light on previous interpretations of certain analyses.]Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
+Our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained models with a very high number of site categories.
+Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
 
@@
 1A, 1D).
-Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1).
+Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1).
 The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig.
@@
 The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
-The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes.
-Or for data-partitioning analyses, some gene partitions use different site-homogeneous models.
+The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites.
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
-The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites.
+The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites.
 Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny.
@@
 Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
+Some analyses partition the data by genes, and use different models for each gene.
 
@@
 
-The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.
-
 ### Other approaches to accommodating site heterogeneity
@@
 Overcoming the challenges associated with this difficult and important question goes beyond models.
-In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie].
-In particular, there are very few chromosome-level genome assemblies outside of Bilateria.
+In particular, there are many organisms that are highly relevant to this problem, in particular better sampling of outgroups, ctenophores, and sponges, for which no genome or transcriptome data are available [@King:2017ie].
+In particular, there are very few chromosome-level genome assemblies for animals outside of Bilateria.
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.
-[XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling.
-Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
+This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 **Extended Data Fig.
-5.** (A) The two alternative hypotheses for deep animal relationships considered here.
-Relationships that are not part of these hypotheses are shown as unresolved polytomies.
-(B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
+5.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses (XX there are only five, one is missing) are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that we can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
 Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
@@
 The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Cells are color coded as in Supplemental Fig XXRecoidng_summary.
 Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
@@
 
-# Supplementary Information
-
-### S.1 Models of molecular evolution
-Models of molecular evolution are assembled by selecting different options for all these different components.
-The models that are applied in practice are heavily influenced by computational costs, as well as convention.
-For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices.
-LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
-This is further confused by the abbreviations that are used for models.
-studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix.
-In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model.
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT).
-Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here.
-If there are no modifiers, then it is implied that site homogeneous models are used.
-
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
-Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR.
-While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
-There are a couple approaches that have been used in the studies considered here:
-
-- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
-The F81 matrix is also sometimes referred to as the Poisson matrix.
-It has no free parameters to estimate since all off-diagonal elements are set to 1.
-
-- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
-It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
-It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
-This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
-This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
-
-While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
-There are a couple approaches that have been used in the studies considered here:
-
-- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
-
-- Gamma rate heterogeneity.
-Each site is assigned to a different rate class with its own rate value.
-This accommodates different rates of evolution across different sites.
-Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
-
-The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
-
-- Empirical site-homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
-
-- Estimated site-homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
-
-- CAT site heterogeneous.
-Each site is assigned to a class with its own equilibrium frequencies.
-The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
-
-- C10 to C60 [@si2008empirical]. 10 to 60-profile mixture models as variants of the CAT model under the maximum-likelihood framework.
-
-- Recoding method:  Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation.
-Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.
-
-### S.2 Details of published analyses
+## Supplementary Information
+
+### S.1 Summaries of published analyses
+
+Narrative summaries of the studies considered here.
 
@@
 It should be noted that all the PhyloBayes runs have not reached convergence due to the computational cost in these large matrices.
+
+### S.2 Models of molecular evolution
+
+The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR.
+While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+There are a couple approaches that have been used in the studies considered here:
+
+- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
+The F81 matrix is also sometimes referred to as the Poisson matrix.
+It has no free parameters to estimate since all off-diagonal elements are set to 1.
+
+- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
+It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
+It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
+
+- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
+It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
+This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
+This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
+
+While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
+There are a couple approaches that have been used in the studies considered here:
+
+- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
+
+- Gamma rate heterogeneity.
+Each site is assigned to a different rate class with its own rate value.
+This accommodates different rates of evolution across different sites.
+Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
+
+The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
+
+- Empirical site-homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
+
+- Estimated site-homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
+
+- CAT site heterogeneous.
+Each site is assigned to a class with its own equilibrium frequencies.
+The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
+
+- C10 to C60 [@si2008empirical]. 10 to 60-profile mixture models as variants of the CAT model under the maximum-likelihood framework.
 
@@
 
-## References
-
 ## Supplementary Tables
@@
 Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
+
+## References

### 2020-08-16 manuscript/manuscript.rmd
another pass to the manuscript based on to-do list in google docs

@@
 Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
-Some analyses partition the data by genes, and use different models for each gene.
+Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).
 
@@
 This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
-Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
+Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters.
 This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
@@
 Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
-That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig.
+If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many sites.
+That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
 3A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
-The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
+The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values are from the sensitive analyses are shown in Supplementary Table 6.
+Support values from the sensitive analyses are shown in Supplementary Table 6.
 
@@
 4).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-However, Our own analyses (Supplementary text, Supplementary Figs.
-4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+However, Our own analyses (Supplementary Information section S3, Supplementary Figs.
+5-7) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
+This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcripomic data.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
-The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to rest of animals.
+The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
 The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
@@
 To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different number of categories were used.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers of categories were used.
 
@@
 #### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
-The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/animal_root/feuda_2017_response.
+The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.
 
@@
 Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-5) give insight into the technical variance of the inference process.
+7) give insight into the technical variance of the inference process.
 
@@
 
-ggsave("figures/Extended_Data_Fig3_raw.pdf", width=6, height=12)
-
-```
-
-![](figures/Extended_Data_Fig3_manuscript.png)
+ggsave("figures/Extended_Data_Fig4_raw.pdf", width=6, height=12)
+
+```
+
+![](figures/Extended_Data_Fig4_manuscript.png)
 
@@
 
-![](figures/Figure_random_rocoding.png)
+![](figures/Extended_Data_Fig5.png)
 
 **Extended Data Fig.
-5.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These six analyses (XX there are only five, one is missing) are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that we can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
-
-![](figures/Figure_recoding_pp.png)
+5.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+
+![](figures/Extended_Data_Fig6.png)
 
@@
 Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
+The 20-state and 6-state points on the plots in Extended Data Fig.
+5 correspond to the 20 and SR6 Whelan cells shown here.
 The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
@@
 
-![](figures/Figure_recoding_summary.png)
+![](figures/Extended_Data_Fig7.png)
 **Extended Data Fig.
-7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+7.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
 
@@
 
-Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms.
+Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatworms.
 The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL.
@@
 
-They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+They also investigated saturation, and concluded that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
 Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
@@
 Nosenko *et al.* [@nosenko2013deep] added Expressed Sequence Tag (EST) data for 9 species of non-bilaterian metazoans (7 sponges).
-They constructed a novel matrix containing 122 genes and parsed them into two non-overlaping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny.
+They constructed a novel matrix containing 122 genes and parsed them into two non-overlapping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny.
 The other major finding was that ribosomal gene partitions showed  significantly lower saturation than the non-ribosomal ones.
 
-Orthologs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect].
-They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix.
+Orthologs were constructed using the bioinformatics pipeline OrthoSelect [@schreiber2009orthoselect].
+They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeneity and model comparison of each matrix.
 By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister).
@@
 For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-based analyses.
-Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model.
+Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using a site-homogeneous model.
 The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).
@@
 This resulted in a final matrix of 170,871 amino acid positions across 586 genes with 44 taxa for the first matrix, and 114 genes with 60 taxa for the second matrix.
-All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different than CAT models in PhyloBayes) to reduce the computational cost.
+All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different from CAT models in PhyloBayes) to reduce the computational cost.
 
@@
 
-Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
+Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprising 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
 The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid].
@@
 #### Whelan et al. 2015
-Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.).
+Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcriptomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.).
 Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.
 The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups.
-All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister.
+All the maximum likelihood analyses with site-homogeneous models and PartitionFinder strongly suggested Ctenophora-sister.
 GTR+CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.
@@
 
-Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
+Chang *et al.* [@Chang:2015hl] was originally used to explore the phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
 The authors constructed a dataset with 200 protein markers based on Philippe *et al.* 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
@@
 
-Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long branch in the tree.
+Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long-branch in the tree.
 
@@
 Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes.
-They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
+They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrices.
 Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
@@
 The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (GTR+CAT) models in non-recoding and recoding datasets.
-The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).
+The results showed that data-recoding can significantly reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in Supplementary Information section S3).
 
@@
 
-Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework.
-The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets.
-More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in almost every cases.
+Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous models with data partitioning under the simulation framework.
+The simulation results suggested that the Poisson+CAT model consistently performed worse than other models in simulation datasets.
+More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimate substitutional heterogeneity in almost every case.
 They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model.
@@
 7).
-
-Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
-Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister.
-Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies.
-In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids.
-These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
-The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
-Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
-They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
-Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister).
-Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
-These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister.
-They also show that recoding can be a problematic method for addressing compositional variation.
 

### 2020-08-18 manuscript/manuscript.rmd
move some cat computation to kernel, re-run kernel

@@
 output:
+  word_document: default
+  pdf_document: 
+    latex_engine: xelatex
   html_document: default
-  pdf_document: default
-  word_document: default
   github_document: default
@@
 ```{r n_cat}
-# Return the allocation midpoint, ie the percent of sites that are allocated to the
-# 50% of the least frequent categories
-allocation_midpoint = function ( pb_summary ) {
-  x = pb_summary %>% pull( count ) %>% sort(decreasing = FALSE)
-  x = x/ sum(x)
-  y = cumsum( x )
-  return ( round( median( y ) *100, 2 ) )
-}
-
-# Parse the last sample from chain 1 of each analysis
-phil_cat_c1 =
-  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
-
-phil_cat60_c1 =
-  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
-
-whel_cat_c1 =
-  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
-
-whel_cat60_c1 =
-  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
-
-# Create a single tibble with summaries of all analyses
-pb_summaries = list(
-  summarise_sample( phil_cat_c1 ) %>%
-    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
-  summarise_sample( phil_cat60_c1 ) %>%
-    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
-  summarise_sample( whel_cat_c1 ) %>%
-    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
-  summarise_sample( whel_cat60_c1 ) %>%
-    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
-) %>%
-  bind_rows() %>%
-  mutate( analysis = paste(matrix, model, sep=" ")) %>%
-  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
-
-pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
-
-# Identify the midpoint of each set of allocations
-mid_phil_cat =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  filter( model=="Poisson+CAT" ) %>%
-  allocation_midpoint()
-
-mid_phil_ncat60 =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  filter( model=="Poisson+nCAT60" ) %>%
-  allocation_midpoint()
-
-mid_whel_cat =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  filter( model=="Poisson+CAT" ) %>%
-  allocation_midpoint()
-
-mid_whel_ncat60 =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  filter( model=="Poisson+nCAT60" ) %>%
-  allocation_midpoint()
-
-n_categories_phil_cat =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  filter( model=="Poisson+CAT" ) %>%
-  nrow()
-
-n_categories_whel_cat =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  filter( model=="Poisson+CAT" ) %>%
-  nrow()
-
-# Perform global MDS analysys
-fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
-pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
 

### 2020-08-18 manuscript/manuscript_kernel.R

@@ -18,6 +18,7 @@ library( igraph )
 library( doParallel )
 
 source( "functions.R" )
+source( "phylobayes.R" )
 
 # Set system computational parameters
 cores = detectCores() - 1
@@ -332,6 +333,80 @@ partition_network_summary =
 
 ### Summarize phylogenetic signal by genes
 au_tests = parse_au_gene_tests()
+
+### Summarize categories from pbmpi
+
+# Parse the last sample from chain 1 of each analysis
+phil_cat_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
+
+phil_cat60_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
+
+whel_cat_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
+
+whel_cat60_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
+
+# Create a single tibble with summaries of all analyses
+pb_summaries = list(
+  summarise_sample(phil_cat_c1) %>%
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"), 
+  summarise_sample(phil_cat60_c1) %>%
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"), 
+  summarise_sample(whel_cat_c1) %>%
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"), 
+  summarise_sample(whel_cat60_c1) %>%
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
+) %>%
+  bind_rows() %>%
+  mutate( analysis = paste(matrix, model, sep=" ")) %>%
+  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
+
+pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
+
+# Identify the midpoint of each set of allocations
[…]

### 2020-08-18 manuscript/manuscript.rmd
new colors for Extended Data Fig. 3

@@
 
-```{r n_cat}
-
-# Generate plots
-gg_rank_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
-
-gg_rank_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
-
-gg_mds_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-gg_mds_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-gg_chain =
-ggarrange(
-  gg_rank_phil,
-  gg_rank_whel,
-  gg_mds_phil,
-  gg_mds_whel,
-  ncol  = 2,
-  nrow = 2,
-  labels = c("A", "B", "C", "D"),
-  common.legend = TRUE
-)
-```
 These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
@@
 ```{r n_cat_render, fig.width=6, fig.height=7}
-gg_chain
+
+edf2_cols = c("#74c2e7", "#410058")
+names(edf2_cols) = c("Poisson+nCAT60", "Poisson+CAT")
+
+# Generate plots
+gg_rank_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  scale_colour_manual(values = edf2_cols) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
+
+gg_rank_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  scale_colour_manual(values = edf2_cols) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
+
+gg_mds_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  scale_colour_manual(values = edf2_cols) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+gg_mds_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  scale_colour_manual(values = edf2_cols) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+ggarrange(
+  gg_rank_phil,
+  gg_rank_whel,
+  gg_mds_phil,
+  gg_mds_whel,
+  ncol  = 2,
+  nrow = 2,
+  labels = c("A", "B", "C", "D"),
+  common.legend = TRUE
+)
+
 ```

### 2020-08-18 manuscript/manuscript.rmd
more uniform scales on Extended Data Fig. 1

@@
     geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
-    facet_grid( matrix_1 ~ matrix_2, scales = "free" ) +
+    facet_grid( matrix_1 ~ matrix_2) +
     theme( axis.text.x = element_text(angle = 45, hjust = 1),

### 2020-08-18 manuscript/manuscript.rmd
update manuscript based on Alex's comments from google doc

@@
 output:
-  word_document: default
   pdf_document:
     latex_engine: xelatex
+  word_document: default
   html_document: default
@@
 
-Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
+Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
 For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies.
@@
 By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
-Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
+Moreover, the outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models.
 This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-However, Our own analyses (Supplementary Information section S3, Supplementary Figs.
-5-7) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+However, our own analyses (Supplementary Information section S3, Supplementary Figs.
+5-7) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
-Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
+Moreover, taxa sampling also plays an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
 
@@
 
-#### Sensitive analyses with different number of substitutional categories
+#### Sensitivity analyses with different number of substitutional categories
 To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
 To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers of categories were used.
-
-To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Posson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
+All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.
+
+To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
 The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
@@
 
-We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.
+We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.
 
@@
 3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
-These plots are for singe sampled isolated from the end of PhyloBayes chain files.
-(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis.
-The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
+These plots are for singe sample isolated from the end of PhyloBayes chain files.
+(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis.
+The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
 The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
@@
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
-Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models.
+Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models.
 The GLS values for each gene in each data matrix are provided in Table S5.
@@
 An analysis of the 40 ribosomal proteins in the matrix recovered Ctenophora-sister with only 69% support.
-This study did not include *Trichoplax*.
+This study did not include the placozoan ingroup *Trichoplax*.
 
@@
 
-Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species.
-Th data matrix were assembled based on The phylogenetic analysis using Poisson+CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade.
+Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore the phylogenetic relationship among early diverging animals by adding 9 new species.
+The data matrix was assembled based on the phylogenetic analysis using Poisson+CAT model, which strongly supported Porifera-sister, and ctenophores were recovered as sister to cnidarians forming the "coelenterate" clade.
 Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70.
-Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained.
+Ambiguously aligned regions were trimmed and only genes sampled for at least two-thirds of species were retained.
 The phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes.
@@
 Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatworms.
-The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL.
+The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes which were clustered by MCL.
 The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy.
@@
 
-They also investigated saturation, and concluded that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+They also investigated saturation, and concluded that the Dunn *et al.* [@Dunn:2008ky] data matrix is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
 Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
@@
 The final matrix was first analyzed using the Poisson+CAT model.
-Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
+Different from other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
 Each jackknife is based on a random selection of ~ 25% of the genes.
 The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets.
-Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
+Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removing taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
 
@@
 Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the GTR+CAT analysis of the SR-6 recoded Whelan data.
-we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
+We created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017).
 When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding.

### 2020-08-21 manuscript/manuscript.rmd
minor edits

--- 2020-08-18 manuscript/manuscript.rmd
@@
 output:
+  pdf_document: default
+  html_document: default
   word_document: default
-  pdf_document: 
-    latex_engine: xelatex
-  html_document: default
   github_document: default
@@
 Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
-Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).
+Some analyses partition the data by genes, and use different models for each gene.
 
@@
 
+```{r n_cat}
+# Return the allocation midpoint, ie the percent of sites that are allocated to the
+# 50% of the least frequent categories
+allocation_midpoint = function ( pb_summary ) {
+  x = pb_summary %>% pull( count ) %>% sort(decreasing = FALSE)
+  x = x/ sum(x)
+  y = cumsum( x )
+  return ( round( median( y ) *100, 2 ) )
+}
+
+# Parse the last sample from chain 1 of each analysis
+phil_cat_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
+
+phil_cat60_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
+
+whel_cat_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
+
+whel_cat60_c1 =
+  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
+
+# Create a single tibble with summaries of all analyses
+pb_summaries = list(
+  summarise_sample( phil_cat_c1 ) %>%
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
+  summarise_sample( phil_cat60_c1 ) %>%
+    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
+  summarise_sample( whel_cat_c1 ) %>%
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
+  summarise_sample( whel_cat60_c1 ) %>%
+    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
+) %>%
+  bind_rows() %>%
+  mutate( analysis = paste(matrix, model, sep=" ")) %>%
+  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
+
+pb_frequencies = pb_summaries %>% select( starts_with("aa.") ) %>% data.matrix()
+
+# Identify the midpoint of each set of allocations
+mid_phil_cat =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  allocation_midpoint()
+
+mid_phil_ncat60 =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+nCAT60" ) %>%
+  allocation_midpoint()
+
+mid_whel_cat =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  allocation_midpoint()
+
+mid_whel_ncat60 =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+nCAT60" ) %>%
+  allocation_midpoint()
+
+n_categories_phil_cat =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  nrow()
+
+n_categories_whel_cat =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  filter( model=="Poisson+CAT" ) %>%
+  nrow()
+
+# Perform global MDS analysys
+fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
+pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
+
+# Generate plots
+gg_rank_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
+
+gg_rank_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
+
+gg_mds_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+gg_mds_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+gg_chain =
+ggarrange(
+  gg_rank_phil,
+  gg_rank_whel,
+  gg_mds_phil,
+  gg_mds_whel,
+  ncol  = 2,
+  nrow = 2,
+  labels = c("A", "B", "C", "D"),
+  common.legend = TRUE
+)
+```
 These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
@@
 This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
-Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters.
+Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter.
 This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
@@
 Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many sites.
-That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
+If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites.
+That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig.
 3A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
-The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
+The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values from the sensitive analyses are shown in Supplementary Table 6.
+Support values are from the sensitive analyses are shown in Supplementary Table 6.
 
@@
 4).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
 Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models.
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-However, Our own analyses (Supplementary Information section S3, Supplementary Figs.
-5-7) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+However, Our own analyses (Supplementary text, Supplementary Figs.
+4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcripomic data.
+This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
-The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
+The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to rest of animals.
 The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
@@
 To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers of categories were used.
+All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different number of categories were used.
 
@@
 #### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/animal_root/feuda_2017_response.
-The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response.
+The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.
 
@@
 Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-7) give insight into the technical variance of the inference process.
+5) give insight into the technical variance of the inference process.
 
@@
 ```{r n_cat_render, fig.width=6, fig.height=7}
-
-edf2_cols = c("#74c2e7", "#410058")
-names(edf2_cols) = c("Poisson+nCAT60", "Poisson+CAT")
-
-# Generate plots
-gg_rank_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  scale_colour_manual(values = edf2_cols) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
-
-gg_rank_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  scale_colour_manual(values = edf2_cols) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
-
-gg_mds_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  scale_colour_manual(values = edf2_cols) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-gg_mds_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  scale_colour_manual(values = edf2_cols) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-ggarrange(
-  gg_rank_phil,
-  gg_rank_whel,
-  gg_mds_phil,
-  gg_mds_whel,
-  ncol  = 2,
-  nrow = 2,
-  labels = c("A", "B", "C", "D"),
-  common.legend = TRUE
-)
-
+gg_chain
 ```
@@
 
-ggsave("figures/Extended_Data_Fig4_raw.pdf", width=6, height=12)
-
-```
-
-![](figures/Extended_Data_Fig4_manuscript.png)
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=6, height=12)
+
+```
+
+![](figures/Extended_Data_Fig3_manuscript.png)
 
@@
 
-![](figures/Extended_Data_Fig5.png)
+![](figures/Figure_random_rocoding.png)
 
 **Extended Data Fig.
-5.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-
-![](figures/Extended_Data_Fig6.png)
+5.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These six analyses (XX there are only five, one is missing) are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that we can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+![](figures/Figure_recoding_pp.png)
 
@@
 Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Extended Data Fig.
-5 correspond to the 20 and SR6 Whelan cells shown here.
+The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here.
 The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
@@
 
-![](figures/Extended_Data_Fig7.png)
+![](figures/Figure_recoding_summary.png)
 **Extended Data Fig.
-7.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
 
@@
 
-Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatworms.
+Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms.
 The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL.
@@
 
-They also investigated saturation, and concluded that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
+They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
 Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
@@
 Nosenko *et al.* [@nosenko2013deep] added Expressed Sequence Tag (EST) data for 9 species of non-bilaterian metazoans (7 sponges).
-They constructed a novel matrix containing 122 genes and parsed them into two non-overlapping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny.
+They constructed a novel matrix containing 122 genes and parsed them into two non-overlaping matrices (ribosomal and non-ribosomal genes) and found incongruent results of deep metazoan phylogeny.
 The other major finding was that ribosomal gene partitions showed  significantly lower saturation than the non-ribosomal ones.
 
-Orthologs were constructed using the bioinformatics pipeline OrthoSelect [@schreiber2009orthoselect].
-They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeneity and model comparison of each matrix.
+Orthologs were constructed using the bioinfomatics pipeline OrthoSelect [@schreiber2009orthoselect].
+They also evaluated level of saturation, leaf stability of sampled taxa, compositional heterogeinity and model comparison of each matrix.
 By modifying gene sampling, ingroup and outgroup sampling, three major topologies related to the position of animal-root were constructed (including Porifera+Placozoa sister, Ctenophora-sister and Porifera-sister).
@@
 For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-based analyses.
-Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using a site-homogeneous model.
+Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model.
 The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).
@@
 This resulted in a final matrix of 170,871 amino acid positions across 586 genes with 44 taxa for the first matrix, and 114 genes with 60 taxa for the second matrix.
-All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different from CAT models in PhyloBayes) to reduce the computational cost.
+All the phylogenetic analyses were analyzed in RAxML under the WAG+CAT+F models (different than CAT models in PhyloBayes) to reduce the computational cost.
 
@@
 
-Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprising 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
+Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included.
 The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid].
@@
 #### Whelan et al. 2015
-Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcriptomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.).
+Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.).
 Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.
 The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups.
-All the maximum likelihood analyses with site-homogeneous models and PartitionFinder strongly suggested Ctenophora-sister.
+All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister.
 GTR+CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.
@@
 
-Chang *et al.* [@Chang:2015hl] was originally used to explore the phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
+Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity.
 The authors constructed a dataset with 200 protein markers based on Philippe *et al.* 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa.
@@
 
-Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long-branch in the tree.
+Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long branch in the tree.
 
@@
 Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes.
-They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrices.
+They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice.
 Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
@@
 The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (GTR+CAT) models in non-recoding and recoding datasets.
-The results showed that data-recoding can significantly reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in Supplementary Information section S3).
+The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).
 
@@
 
-Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous models with data partitioning under the simulation framework.
-The simulation results suggested that the Poisson+CAT model consistently performed worse than other models in simulation datasets.
-More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimate substitutional heterogeneity in almost every case.
+Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework.
+The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets.
+More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in almost every cases.
 They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model.
@@
 7).
+
+Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister.
+Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister.
+Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies.
+In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids.
+These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
+The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
+Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
+They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies".
+Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister).
+Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister.
+They also show that recoding can be a problematic method for addressing compositional variation.
 

### 2020-08-21 manuscript/manuscript.rmd

--- 2020-08-18 manuscript/manuscript.rmd
@@
 output:
+  html_document: default
   pdf_document:
@@
   word_document: default
-  html_document: default
   github_document: default

### 2020-08-24 manuscript/manuscript.rmd
update manuscript based on Jas's comments, all are typos or grammers. Did not re-knit the manuscript

@@
 
-We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundred.
-It could be that the sites that are allocated to rate categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundreds.
+It could be that the sites that are allocated to rare categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
 Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
@@
 site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
+We included models that are commonly used in previous analyses, including site-homogeneous Poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
 For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
@@
 
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, poisson+C60 and associated best-fit model identified above.
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, Poisson+C60 and associated best-fit model identified above.
 Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
@@
 For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
-2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included the most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
 It should be noted that the main conclusion of Simion *et al*.
@@
 #### Outgroup taxa sampling with C60 and CAT models
-Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia , Holozoa and Opisthokonta.
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta.
 These datasets include the same set of genes but differ in the composition of outgroup species.
@@
 Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, GTR+CAT models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 
@@
 To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanomalia and the full matrix).
+We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanimalia and the full matrix).
 For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
@@
 #### Sensitivity analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_Choanimalia (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.
@@
 The 20 state matrices are the same across all analyses since they are not recoded.
-Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-7) give insight into the technical variance of the inference process.
+Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
+7) gives insight into the technical variance of the inference process.
 
@@
 4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
-The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green).
+The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
 Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models.
@@
 
->Our phylogenetic analyses suggest that ctenophores are the sister group to the rest of the extant animals.
+> Our phylogenetic analyses suggest that ctenophores are the sister group to the rest of the extant animals.
 
@@
 The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid].
-After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080  (Total 1080).
+After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080  (Total 1080) genes.
 The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes.
@@
 Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.
-The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups.
+The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressive removal of outgroups.
 All the maximum likelihood analyses with site-homogeneous models and PartitionFinder strongly suggested Ctenophora-sister.
@@
 Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets.
-It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny.
+It was the first study showing that progressive removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny.
 The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction.
@@
 They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrices.
-Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.
+Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups are excluded and the use of site-heterogeneous CAT models are used.
 
@@
 The subsequent filtering strategy was also similar to the previous study.
-All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in choanoazoa dataset.
+All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in Choanimalia dataset.
 The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.
@@
 The final matrix was first analyzed using the Poisson+CAT model.
-Different from other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
+Different from other PhyloBayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size.
 Each jackknife is based on a random selection of ~ 25% of the genes.
-The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets.
+The PhyloBayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets.
 Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removing taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.
@@
 It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
-This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
+This leaves 190 parameters that must be estimated from the data along with the other model parameters and the phylogenetic tree topology.
 This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
@@
 
-There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
+There is growing interest in data-recoding and a previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
 They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
@@
 Supplementary Table 1.
-Summary of annotations (Busco, Ensemble, go-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
+Summary of annotations (BUSCO, Ensemble, GO-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
 
@@
 
-Supplementary_Table 6. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).
+Supplementary_Table 6. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices (Table is converted from analyses_sensitive in Rdata).
 

### 2020-09-08 manuscript/manuscript.rmd
add two sentences of importance of heterogeneous

@@
 Moreover, taxa sampling also plays an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
+Importantly, we want to emphasize that our concern here is not with the heterogeneous analyses but with unrealistic large number of unconstrained substitutional categories.
+The heterogeneity is important to consider in future phylogenomic studies (especially for exploring the deep relationships), but the results should be carefully examined.
 

### 2020-09-11 manuscript/phylobayes.R
older phylobayes changes

@@ -86,7 +86,7 @@ PhylobayesSample = function(
     rr = as.numeric(text_6),
     dirweight = text_7 %>% str_split("\t") %>% unlist() %>% as.numeric(),
     frequencies = text[freq_start:(length(text)-1)] %>% str_split("\t", 20) %>% lapply( as.numeric ) %>% do.call( rbind, . ), 
-    allocation = text[length(text)] %>% str_split("\t") %>% unlist() %>% as.numeric()
+    allocation = text[length(text)] %>% str_trim() %>% str_split("\t") %>% unlist() %>% as.numeric()
   )
   
   rownames( object@frequencies ) = 0:( nrow( object@frequencies ) - 1 )

### 2020-09-11 manuscript/manuscript.rmd
another pass at text

--- 2020-08-24 manuscript/manuscript.rmd
@@
 Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses.
-Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions.
-Our new analyses are consistent with this.
-Our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained models with a very high number of site categories.
-Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
+It had previously been suggested that site heterogeneous models favor Profera-sister, but we find that this is not the case.
+Profera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data.
+Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
 
@@
 Porifera-sister (Figs.
-1B, 1E) was widely accepted though rarely tested.
-By contrast, there has long been uncertainty about the placement of Ctenophora in the animal tree of life [@Wallberg:2004ws].
+1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested.
+By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws].
 The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs.
 1A, 1D).
-Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1).
-The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig.
-1).
+Since then, many more relevant phylogenomic studies have been published (Fig.
+2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1).
+The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig.
+3A).
 But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity.
-This methodological variance is often confounded, for example when a particular model is only available in one software package.
+This is because methodological variance is often confounded, for example when a particular model is only available in one software package.
 
@@
 2A, Extended Data Table 1).
-These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.
+These include all such phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations.
+Among these 15 studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
 
@@
 (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
-(D) A ctenophore.
-(E) A sponge.
+(D) A ctenophore, commonly known as comb jellies.
+(E) A poriferan, commonly known as sponges.
 
@@
 
-The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$.
-The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another.
+The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $R$ and amino acid equilibrium frequencies $\Pi$.
+The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another.
 The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
@@
 (B) Each of the primary matrices considered here, color coded by taxon sampling.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of species sampled.
 
@@
 The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
-Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge and thus renders the analysis untrustworthy / hard to interpret.
+Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
 The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.
@@
 
-Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
-For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies.
+For each of the published studies, we selected the matrix that was the primary focus of the manuscript, or has been reanalyzed extensively in other studies, for further analysis.
 For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
-This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.
+This produced `r analyses_new$matrix%>% unique()%>% length()` data matrices from 11 studies that presented new sequence data and for which partition data were available.
 
@@
 In all cases, models with C60 fit these matrices better than the site-homogeneous models.
+This is consistent wwith the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
 We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60.
@@
 3B).
-No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
+No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
 3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
@@
 There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models.
-In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
+In published analyses of the animal root, these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
 Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.
@@
 A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories.
-CAT models employ a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
+The standard CAT model employs a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq].
 IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60).
@@
 This requires a very large number of additional parameters that must be estimated from the data.
-Despite C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5).
+Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5).
 These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
 
-Once we established this large difference in category number, we sought to test its direct impact on animal root position.
+We examined the specific impact of these large difference in category number on animal root position.
 It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT.
@@
 We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
-This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories.
-When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
+This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories.
+When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig.
 4).
-For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+CAT model.
-
-Published studies have found that the GTR+CAT model often has a better fit than Poisson+CAT, and investigators have suggested that it is more appropriate than the Poisson+CAT model for this question [@simion2017large].
+For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model.
+
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
-We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
+We found that for both Whelan2017 and Philippe2009 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
 Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
@@
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
-This is very consistent across published analyses and our new panels of analyses.
-The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these respects.
+This is consistent across published analyses and our new panels of analyses.
+The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects.
 If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
@@
 Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many sites.
+If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories.
 That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
@@
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
-This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig.
+This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig.
 3A-B).
-This suggests that support for Porifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.
-
-We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundreds.
+This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.
+
 It could be that the sites that are allocated to rare categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
@@
 
-Taken together, these results indicate that when CAT models are run without constraint on the number of categories, they select a very large number of categories.
-A large number of these categories are allocated to a small number of sites that have small differences from other categories.
-Given the very large number of parameters that must be estimated for this large number of rare categories, this suggests that unconstrained CAT analyses are overparameterized.
-While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
+Taken together, this suggests that unconstrained CAT analyses are overparameterized.
+While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
 In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.
@@
 
-## The current state of understanding and future directions
-
-Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem.
+## Conclusions
+
+Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem [@laumer2019revisiting].
 By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
@@
 As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.
-
-This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence.
-Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many?
-Moreover, taxa sampling also plays an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.
 
@@
 
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another.
+The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to others.
 Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR.
@@
 
-There is growing interest in data-recoding and a previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies.
-They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
+Feuda *et al.* [@feuda2017improved] were concerned that the Ctenophora-sister results Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj] were artefacts of lineage-specific differences in amino acid frequencies.
+In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids.
+These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds].
+The intent is to discard many lineage-specific changes, which are expected to fall within these groups.
+Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied.
+
+Feuda *et al.* [@feuda2017improved] report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
+Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
 5).
-Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods).
-When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-7).
+Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+These findings indicate that recoding can be a problematic method for addressing compositional variation.
 

### 2020-09-11 manuscript/manuscript.rmd
minor tweeks

@@
 
-Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Steven H.D. Haddock^5^, Antonis Rokas^3^ and Casey W. Dunn^1^*
+Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Antonis Rokas^3^ * and Casey W. Dunn^1^*
 
@@
 
-^5^Monterey Bay Aquarium Research Institute, Moss Landing, CA 95039, USA
-
-\* Corresponding author, casey.dunn@yale.edu
+\* Corresponding authors, antonis.rokas@vanderbilt.edu, casey.dunn@yale.edu
 
@@
 
-Benjamin Evans:
-Steven H.D. Haddock: 0000-0001-9420-4482
+Benjamin Evans: 0000-0003-2069-4938
 

### 2020-09-22 manuscript/manuscript.rmd
change Philippe2009 tree names

@@
 ## References
+
+branch_lengths = read.csv("../data_processed/tables/branch_lengths.tsv", header=TRUE, sep="\t")
+
+branch_lengths$Clade =
+    factor(
+        branch_lengths$Clade,
+        levels = c("Bilateria", "Cnidaria", "Placozoa", "Porifera", "Ctenophora", "Choanoflagellata"))
+
+p=ggplot (branch_lengths, aes(x=Clade, y=branch_length, fill=Clade)) + geom_violin(trim=FALSE) + geom_boxplot(width=0.1) + theme_classic()+  labs(x="Clade", y = "Branch Lengths")
+
+p+scale_fill_manual(values=c("#E6BEFF", "#911EB4", "#82D1F4", "#009E73", "#E69F00", "#4363D8"))
+
+Bilateria_branch_mean = branch_lengths %>% filter(Clade=="Bilateria") %>% summarise(mean = mean(branch_length), n = n())
+Porifera_branch_mean = branch_lengths %>% filter(Clade=="Porifera") %>% summarise(mean = mean(branch_length), n = n())
+Ctenophora_branch_mean = branch_lengths %>% filter(Clade=="Ctenophora") %>% summarise(mean = mean(branch_length), n = n())
+
+taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
+outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
+Bilateria= taxa_map %>% filter (clade_assignment=="Bilateria")
+Cnidaria= taxa_map %>% filter (clade_assignment=="Cnidaria")
+Placozoa= taxa_map %>% filter (clade_assignment=="Placozoa")
+Porifera= taxa_map %>% filter (clade_assignment=="Porifera")
+Ctenophora= taxa_map %>% filter (clade_assignment=="Ctenophora")
+
+tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+tree1_rooted =  root(tree1, outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree1_rooted) +
+  geom_cladelabel(node=90, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=114, label="Porifera", color="#009E73") +
+  geom_hilight(node=90, fill="#E69F00") +
+  geom_hilight(node=114, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 100)) + ggtitle("    WAG+C60")
+
+tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
+tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree2_rooted) +
+  geom_cladelabel(node=119, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=100, label="Porifera", color="#009E73") +
+  geom_hilight(node=119, fill="#E69F00") +
+  geom_hilight(node=100, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 100)) + ggtitle("    Poisson+nCAT60")
+
+tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
+tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree3_rooted) +
+  geom_cladelabel(node=100, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=130, label="Porifera", color="#009E73") +
+  geom_hilight(node=100, fill="#E69F00") +
+  geom_hilight(node=130, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 100)) + ggtitle("    Poisson+nCAT90")
+p3 = flip(p3, 100, 98)
+
+tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
+tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree4_rooted) +
+  geom_cladelabel(node=100, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=130, label="Porifera", color="#009E73") +
+  geom_hilight(node=100, fill="#E69F00") +
+  geom_hilight(node=130, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 100)) + ggtitle("    Poisson+CAT")
+p4 = flip(p4, 100, 98)
+
+tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
+tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
+p5=ggtree(tree5_rooted) +
+  geom_cladelabel(node=119, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=100, label="Porifera", color="#009E73") +
+  geom_hilight(node=119, fill="#E69F00") +
+  geom_hilight(node=100, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 100)) + ggtitle("    GTR+CAT")
+
+ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

### 2020-09-23 manuscript/manuscript.rmd
add GTR_CAT60 results

--- 2020-09-11 manuscript/manuscript.rmd
@@
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
-We found that for both Whelan2017 and Philippe2009 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
+We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
 Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
@@
 ## References
+
+branch_lengths = read.csv("../data_processed/tables/branch_lengths.tsv", header=TRUE, sep="\t")
+
+branch_lengths$Clade =
+    factor(
+        branch_lengths$Clade,
+        levels = c("Bilateria", "Cnidaria", "Placozoa", "Porifera", "Ctenophora", "Choanoflagellata"))
+
+p=ggplot (branch_lengths, aes(x=Clade, y=branch_length, fill=Clade)) + geom_violin(trim=FALSE) + geom_boxplot(width=0.1) + theme_classic()+  labs(x="Clade", y = "Branch Lengths")
+
+p+scale_fill_manual(values=c("#E6BEFF", "#911EB4", "#82D1F4", "#009E73", "#E69F00", "#4363D8"))
+
+Bilateria_branch_mean = branch_lengths %>% filter(Clade=="Bilateria") %>% summarise(mean = mean(branch_length), n = n())
+Porifera_branch_mean = branch_lengths %>% filter(Clade=="Porifera") %>% summarise(mean = mean(branch_length), n = n())
+Ctenophora_branch_mean = branch_lengths %>% filter(Clade=="Ctenophora") %>% summarise(mean = mean(branch_length), n = n())
+
+taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
+outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
+Bilateria= taxa_map %>% filter (clade_assignment=="Bilateria")
+Cnidaria= taxa_map %>% filter (clade_assignment=="Cnidaria")
+Placozoa= taxa_map %>% filter (clade_assignment=="Placozoa")
+Porifera= taxa_map %>% filter (clade_assignment=="Porifera")
+Ctenophora= taxa_map %>% filter (clade_assignment=="Ctenophora")
+
+tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+tree1_rooted =  root(tree1, outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree1_rooted) +
+  geom_cladelabel(node=90, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=114, label="Porifera", color="#009E73") +
+  geom_hilight(node=90, fill="#E69F00") +
+  geom_hilight(node=114, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95)) + ggtitle("    WAG+C60")
+
+tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
+tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree2_rooted) +
+  geom_cladelabel(node=119, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=100, label="Porifera", color="#009E73") +
+  geom_hilight(node=119, fill="#E69F00") +
+  geom_hilight(node=100, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1)) + ggtitle("    Poisson+nCAT60")
+
+tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
+tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree3_rooted) +
+  geom_cladelabel(node=100, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=130, label="Porifera", color="#009E73") +
+  geom_hilight(node=100, fill="#E69F00") +
+  geom_hilight(node=130, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1)) + ggtitle("    Poisson+nCAT90")
+p3 = flip(p3, 100, 98)
+
+tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
+tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree4_rooted) +
+  geom_cladelabel(node=100, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=130, label="Porifera", color="#009E73") +
+  geom_hilight(node=100, fill="#E69F00") +
+  geom_hilight(node=130, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) <1)) + ggtitle("    Poisson+CAT")
+p4 = flip(p4, 100, 98)
+
+tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
+tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
+p5=ggtree(tree5_rooted) +
+  geom_cladelabel(node=119, label="Ctenophora", color="#E69F00") +
+  geom_cladelabel(node=100, label="Porifera", color="#009E73") +
+  geom_hilight(node=119, fill="#E69F00") +
+  geom_hilight(node=100, fill="#009E73") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1)) + ggtitle("    GTR+CAT")
+
+ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
+
+taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
+outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
+
+tree1= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+tree1_rooted =  root(tree1, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree1_rooted) +
+  geom_cladelabel(node=85, label="Ctenophora") +
+  geom_cladelabel(node=76, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=33, label="Placozoa") +
+  geom_cladelabel(node=53, label="Bilateria") +
+  geom_cladelabel(node=51, label="Cnidaria") +
+  geom_hilight(node=85, fill="#E69F00") +
+  geom_hilight(node=76, fill="#009E73") +
+  geom_hilight(node=33, fill="#82D1F4") +
+  geom_hilight(node=53, fill="#E6BEFF") +
+  geom_hilight(node=51, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95), hjust=-.3, size=2) + ggtitle("    WAG+C60")
+
+tree2=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
+tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree2_rooted) +
+  geom_cladelabel(node=83, label="Ctenophora") +
+  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=32, label="Placozoa") +
+  geom_cladelabel(node=58, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=83, fill="#E69F00") +
+  geom_hilight(node=85, fill="#009E73") +
+  geom_hilight(node=32, fill="#82D1F4") +
+  geom_hilight(node=58, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT60")
+
+tree3=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
+tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree3_rooted) +
+  geom_cladelabel(node=83, label="Ctenophora") +
+  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=32, label="Placozoa") +
+  geom_cladelabel(node=58, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=83, fill="#E69F00") +
+  geom_hilight(node=85, fill="#009E73") +
+  geom_hilight(node=32, fill="#82D1F4") +
+  geom_hilight(node=58, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT90")
+
+tree4=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
+tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree4_rooted) +
+  geom_cladelabel(node=58, label="Ctenophora") +
+  geom_cladelabel(node=84, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=35, label="Placozoa") +
+  geom_cladelabel(node=60, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=58, fill="#E69F00") +
+  geom_hilight(node=84, fill="#009E73") +
+  geom_hilight(node=35, fill="#82D1F4") +
+  geom_hilight(node=60, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+CAT")
+
+tree5=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
+tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
+p5=ggtree(tree5_rooted) +
+  geom_cladelabel(node=58, label="Ctenophora") +
+  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=38, label="Placozoa") +
+  geom_cladelabel(node=61, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=58, fill="#E69F00") +
+  geom_hilight(node=85, fill="#009E73") +
+  geom_hilight(node=38, fill="#82D1F4") +
+  geom_hilight(node=61, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    GTR+CAT")
+p5 = flip(p5, 85, 38)
+
+ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

### 2020-09-24 manuscript/manuscript.rmd
add two tree files and update orders of extended figures

--- 2020-09-11 manuscript/manuscript.rmd
@@
   library( gtable )
+  library( ggtree )
+  library( ape )
 
@@
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
-We found that for both Whelan2017 and Philippe2009 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
+We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model (Extended Data Fig.
+3E).
 Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
 
-These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017).
+These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig.
+3; Philippe2009 Extended Data Fig.
+4).
 Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
@@
 That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
-3A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
+5A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig.
-3A-B).
+5A-B).
 This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.
@@
 Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
-We find support for this second scenario (Extended Data Fig.
+We find support for this second scenario (Extended Data Fig5.
 3C-D).
@@
 To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
-4).
+6).
 By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
@@
 In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
-4; Supplementary Table 7).
+6; Supplementary Table 7).
 In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-4).
+6).
 For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
@@
 Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-7) gives insight into the technical variance of the inference process.
+9) gives insight into the technical variance of the inference process.
 
@@
 
-The statistics presented in  Extended Data Figs. 5 and 6 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+The statistics presented in  Extended Data Figs. 7 and 8 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
+
+```{r figure_Whelan2017_trees,, fig.height = 8, fig.width = 10}
+
+taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
+outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
+
+tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+tree1_rooted =  root(tree1, outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree1_rooted) +
+  geom_cladelabel(node=90, label="Ctenophora", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=114, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=28, label="Placozoa") +
+  geom_cladelabel(node=109, label="Bilateria") +
+  geom_cladelabel(node=95, label="Cnidaria") +
+  geom_hilight(node=90, fill="#E69F00") +
+  geom_hilight(node=114, fill="#009E73") +
+  geom_hilight(node=28, fill="#82D1F4") +
+  geom_hilight(node=109, fill="#E6BEFF") +
+  geom_hilight(node=95, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95), hjust=-.3, size=2) + ggtitle("    WAG+C60")
+
+tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
+tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree2_rooted) +
+  geom_cladelabel(node=119, label="Ctenophora", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=100, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=22, label="Placozoa") +
+  geom_cladelabel(node=93, label="Bilateria") +
+  geom_cladelabel(node=86, label="Cnidaria") +
+  geom_hilight(node=119, fill="#E69F00") +
+  geom_hilight(node=100, fill="#009E73") +
+  geom_hilight(node=22, fill="#82D1F4") +
+  geom_hilight(node=93, fill="#E6BEFF") +
+  geom_hilight(node=86, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT60")
+
+tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
+tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree3_rooted) +
+  geom_cladelabel(node=100, label="Ctenophora", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=130, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=22, label="Placozoa") +
+  geom_cladelabel(node=93, label="Bilateria") +
+  geom_cladelabel(node=86, label="Cnidaria") +
+  geom_hilight(node=100, fill="#E69F00") +
+  geom_hilight(node=130, fill="#009E73") +
+  geom_hilight(node=22, fill="#82D1F4") +
+  geom_hilight(node=93, fill="#E6BEFF") +
+  geom_hilight(node=86, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT90")
+p3 = flip(p3, 100, 98)
+
+tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
+tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree4_rooted) +
+geom_cladelabel(node=100, label="Ctenophora", angle=90, hjust='center', offset.text=.01) +
+geom_cladelabel(node=130, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+geom_cladelabel(node=22, label="Placozoa") +
+geom_cladelabel(node=93, label="Bilateria") +
+geom_cladelabel(node=86, label="Cnidaria") +
+geom_hilight(node=100, fill="#E69F00") +
+geom_hilight(node=130, fill="#009E73") +
+geom_hilight(node=22, fill="#82D1F4") +
+geom_hilight(node=93, fill="#E6BEFF") +
+geom_hilight(node=86, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) <1), hjust=-.3, size=2) + ggtitle("    Poisson+CAT")
+p4 = flip(p4, 100, 98)
+
+tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
+tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
+p5=ggtree(tree5_rooted) +
+geom_cladelabel(node=119, label="Ctenophora", angle=90, hjust='center', offset.text=.01) +
+geom_cladelabel(node=100, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+geom_cladelabel(node=22, label="Placozoa") +
+geom_cladelabel(node=93, label="Bilateria") +
+geom_cladelabel(node=86, label="Cnidaria") +
+geom_hilight(node=119, fill="#E69F00") +
+geom_hilight(node=100, fill="#009E73") +
+geom_hilight(node=22, fill="#82D1F4") +
+geom_hilight(node=93, fill="#E6BEFF") +
+geom_hilight(node=86, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 1), hjust=-.3, size=2) + ggtitle("    GTR+CAT")
+
+ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
+
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=8)
+```
+
+**Extended Data Fig.
+3.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix.
+(A).
+WAG+C60.
+(B).
+Poisson+nCAT60.
+(C).
+Poisson+nCAT90.
+(D).
+Poisson+CAT.
+(E).
+GTR+CAT.
+
+```{r figure_Philippe2009_trees,, fig.height = 8, fig.width = 10}
+
+taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
+outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
+
+tree1= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+tree1_rooted =  root(tree1, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree1_rooted) +
+  geom_cladelabel(node=85, label="Ctenophora") +
+  geom_cladelabel(node=76, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=33, label="Placozoa") +
+  geom_cladelabel(node=53, label="Bilateria") +
+  geom_cladelabel(node=51, label="Cnidaria") +
+  geom_hilight(node=85, fill="#E69F00") +
+  geom_hilight(node=76, fill="#009E73") +
+  geom_hilight(node=33, fill="#82D1F4") +
+  geom_hilight(node=53, fill="#E6BEFF") +
+  geom_hilight(node=51, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95), hjust=-.3, size=2) + ggtitle("    WAG+C60")
+
+tree2=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
+tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree2_rooted) +
+  geom_cladelabel(node=83, label="Ctenophora") +
+  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=32, label="Placozoa") +
+  geom_cladelabel(node=58, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=83, fill="#E69F00") +
+  geom_hilight(node=85, fill="#009E73") +
+  geom_hilight(node=32, fill="#82D1F4") +
+  geom_hilight(node=58, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT60")
+
+tree3=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
+tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree3_rooted) +
+  geom_cladelabel(node=83, label="Ctenophora") +
+  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=32, label="Placozoa") +
+  geom_cladelabel(node=58, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=83, fill="#E69F00") +
+  geom_hilight(node=85, fill="#009E73") +
+  geom_hilight(node=32, fill="#82D1F4") +
+  geom_hilight(node=58, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT150")
+
+tree4=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
+tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree4_rooted) +
+  geom_cladelabel(node=58, label="Ctenophora") +
+  geom_cladelabel(node=84, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+  geom_cladelabel(node=35, label="Placozoa") +
+  geom_cladelabel(node=60, label="Bilateria") +
+  geom_cladelabel(node=52, label="Cnidaria") +
+  geom_hilight(node=58, fill="#E69F00") +
+  geom_hilight(node=84, fill="#009E73") +
+  geom_hilight(node=35, fill="#82D1F4") +
+  geom_hilight(node=60, fill="#E6BEFF") +
+  geom_hilight(node=52, fill="#911EB4") +
+  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+CAT")
+
+  tree5=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
+  tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
+  p5=ggtree(tree5_rooted) +
+    geom_cladelabel(node=58, label="Ctenophora") +
+    geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
+    geom_cladelabel(node=38, label="Placozoa") +
+    geom_cladelabel(node=61, label="Bilateria") +
+    geom_cladelabel(node=52, label="Cnidaria") +
+    geom_hilight(node=58, fill="#E69F00") +
+    geom_hilight(node=85, fill="#009E73") +
+    geom_hilight(node=38, fill="#82D1F4") +
+    geom_hilight(node=61, fill="#E6BEFF") +
+    geom_hilight(node=52, fill="#911EB4") +
+    geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    GTR+CAT")
+  p5 = flip(p5, 85, 38)
+
+ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
+
+ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=8)
+```
+**Extended Data Fig.
+4.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix.
+(A).
+WAG+C60.
+(B).
+Poisson+nCAT60.
+(C).
+Poisson+nCAT150.
+(D).
+Poisson+CAT.
+(E).
+GTR+CAT.
 
@@
 **Extended Data Fig.
-3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+5** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
 These plots are for singe sample isolated from the end of PhyloBayes chain files.
@@
 
-ggsave("figures/Extended_Data_Fig4_raw.pdf", width=6, height=12)
+ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=12)
 
@@
 **Extended Data Fig.
-4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+6** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
@@
 **Extended Data Fig.
-5.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
 Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
@@
 **Extended Data Fig.
-6.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+8.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 **Extended Data Fig.
-7.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+9.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
 These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
@@
 Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
-5).
+7).
 Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
@@
 This provides 32 analyses that they report in their Table 3 and that we present graphically here as Extended Data Fig.
-5.
+7.
 There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Extended Data Fig.
-5).
+7).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 This leaves eight GTR+CAT analyses (Extended Data Fig.
-5).
+7).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
@@
 Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
-6).
+8).
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig.
-6), indicating that it did not converge as well as other analyses.
+8), indicating that it did not converge as well as other analyses.
 
@@
 Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-7).
+9).
 

### 2020-10-05 manuscript/manuscript.rmd
small changes to the manuscript based on Nicolas's comments

@@
 
-p
-
 ggsave("figures/Figure3_raw.pdf", width=7.2, height=3.6)
@@
 
+![](figures/Figure3_published.png)
 **Fig.
@@
 Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
+Moreover, we found that the branch lengths of Ctenophora taxa are generally longer than other animals with both site-homogeneous and site-heterogeneous models in many data matrices (*e.g.* Extended Data Figs.
+3, 4).
 
@@
 
-Taken together, this suggests that unconstrained CAT analyses are overparameterized.
+Taken together, although further analysis are warranted, this suggests that unconstrained CAT analyses might be overparameterized.
 While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
@@
   analyses_sensitive %>%
+  filter(model != "GTR+CAT60") %>%
   filter(matrix == "Philippe2009_only_choanozoa" )
@@
   analyses_sensitive %>%
+  filter(model != "GTR+CAT60") %>%
   filter(matrix == "Whelan2017_full_only_choanozoa")
@@
 
-ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)
+p=ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)
 
@@
 ```
+![](figures/Figure4_manuscript.png)
 **Fig.
@@
 There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcripomic data.
+This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcriptomic data.
 We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
@@
 
-ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
+p=ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
 
@@
 ```
-
+![](figures/Extended_Data_Fig3_manuscript.png)
 **Extended Data Fig.
@@
 
-ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
+p= ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)
 
@@
 ```
+![](figures/Extended_Data_Fig4_manuscript.png)
 **Extended Data Fig.

### 2020-10-12 manuscript/manuscript.rmd
addressed feedback from Nicolas

@@
 
-Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Antonis Rokas^3^ * and Casey W. Dunn^1^*
+Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*
 
@@
 This is because methodological variance is often confounded, for example when a particular model is only available in one software package.
+There are a variety of things that make this a particularly challenging problem.
+For one, the relevant nodes are the deepest in the animal tree of life.
+Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long branch attraction to outgroups [@pisani2015genomic].
+While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs.
+3, 4), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].
 
@@
 Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
-Moreover, we found that the branch lengths of Ctenophora taxa are generally longer than other animals with both site-homogeneous and site-heterogeneous models in many data matrices (*e.g.* Extended Data Figs.
-3, 4).
 
@@
 One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
-Each new category adds the same number of parameters, but that site can be allocated to any number of sites.
+Each new category adds the same number of parameters, but that category can be allocated to any number of sites.
 If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories.
@@
 3C-D).
-
-Taken together, although further analysis are warranted, this suggests that unconstrained CAT analyses might be overparameterized.
-While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites.
-In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.
 
@@
 
-Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long-branch in the tree.
-
 #### Pisani *et al.* 2015

### 2020-10-13 manuscript/manuscript.rmd
update minor changes of text and figure orders

@@
 csl: nature.csl
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
 output:
-  html_document: default
   pdf_document:
     latex_engine: xelatex
+  html_document: default
   word_document: default
@@
 Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses.
-It had previously been suggested that site heterogeneous models favor Profera-sister, but we find that this is not the case.
-Profera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data.
+It had previously been suggested that site heterogeneous models favor Porifera-sister, but we find that this is not the case.
+Porifera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data.
 Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
@@
 For one, the relevant nodes are the deepest in the animal tree of life.
-Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long branch attraction to outgroups [@pisani2015genomic].
+Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
 While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs.
-3, 4), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].
+1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].
 
@@
 Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
-2).
+3).
 
@@
 2B, Extended Data Fig.
-1).
+4).
 Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria.
 Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig.
-1).
+4).
 Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig.
-1).
+4).
 This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.
@@
 We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model (Extended Data Fig.
-3E).
+5E).
 Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
@@
 These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig.
-3; Philippe2009 Extended Data Fig.
-4).
+1; Philippe2009 Extended Data Fig.
+2).
 Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
@@
 That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
-5A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories.
+5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
@@
 These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi.
+Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
 For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
@@
 
-For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency.
+For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe2009 and Whelan2017 matrices due to computational efficiency.
 Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
@@
 #### Sensitivity analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_Choanimalia (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60).
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanimalia (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60).
 To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
@@
 The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
+
+#### Cross-validation analyses
+Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score.
+For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.
 
@@
 
+We thank members of the Dunn and Rokas laboratories for discussions and comments.
+We are grateful to Nicolas Lartillot, Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript.
 We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.
+Yuanning Li was previously supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad.
 
@@
 opts <- options(knitr.kable.NA = "")
-knitr::kable(table_study_summary)
+knitr::kable(table_study_summary) %>%
+  kable_styling(full_width = F)
 
@@
 ## Extended Figures
-
-```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
-
-matrix_overlap_rectangles =
-  lapply(
-    sequence_matrices[no_subsets],
-    function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x, y))) %>%
-  unlist(recursive = FALSE) %>%
-  bind_rows()
-
-matrix_overlap_rectangles %>%
-  ggplot( ) +
-    scale_x_continuous(name = "Genes") +
-    scale_y_continuous(name = "Species") +
-    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
-    facet_grid( matrix_1 ~ matrix_2) +
-    theme( axis.text.x = element_text(angle = 45, hjust = 1),
-           strip.text.y = element_text(angle = 0),
-           strip.text.x = element_text(angle = 60))
-
-ggsave("figures/Extended_Data_Fig1_raw.pdf", width=20, height=12)
-
-```
-
-**Extended Data Fig.
-1.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
-```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
-
-ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
-busco = read_tsv("../data_processed/tables/busco_gene.tsv")
-
-ribo$matrix =
-  factor(
-    ribo$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-             "Whelan2017_full", "Whelan2017_strict") )
-
-p1 = ribo %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
-
-busco$matrix =
-  factor(
-    busco$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-               "Whelan2017_full", "Whelan2017_strict") )
-
-busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
-
-p2 = busco %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
-
- ggarrange(p1, p2, labels = c("A", "B"))
-
-ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=4)
-
-```
-
-**Extended Data Fig.
-2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
-(C).
-The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
 
@@
 
-ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=8)
-```
-![](figures/Extended_Data_Fig3_manuscript.png)
+ggsave("figures/Extended_Data_Fig1_raw.pdf", width=10, height=8)
+```
+![](figures/Extended_Data_Fig1_manuscript.png)
 **Extended Data Fig.
-3.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix.
+1.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix.
 (A).
@@
 
-tree1= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
-tree1_rooted =  root(tree1, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
-p1 = ggtree(tree1_rooted) +
+tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+tree6_rooted =  root(tree6, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
+p1 = ggtree(tree6_rooted) +
   geom_cladelabel(node=85, label="Ctenophora") +
@@
 
-tree2=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
-tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
-p2= ggtree(tree2_rooted) +
+tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
+tree7_rooted =  root(tree7, outgroup$relabelled_name, resolved.root=TRUE)
+p2= ggtree(tree7_rooted) +
   geom_cladelabel(node=83, label="Ctenophora") +
@@
 
-tree3=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
-tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
-p3=ggtree(tree3_rooted) +
+tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
+tree8_rooted =  root(tree8, outgroup$relabelled_name, resolved.root=TRUE)
+p3=ggtree(tree8_rooted) +
   geom_cladelabel(node=83, label="Ctenophora") +
@@
 
-tree4=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
-tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
-p4=ggtree(tree4_rooted) +
+tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
+tree9_rooted =  root(tree9, outgroup$relabelled_name, resolved.root=TRUE)
+p4=ggtree(tree9_rooted) +
   geom_cladelabel(node=58, label="Ctenophora") +
@@
 
-  tree5=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
-  tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
-  p5=ggtree(tree5_rooted) +
+  tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
+  tree10_rooted =  root(tree10, outgroup$relabelled_name, resolved.root=TRUE)
+  p5=ggtree(tree10_rooted) +
     geom_cladelabel(node=58, label="Ctenophora") +
@@
 
-ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=8)
-```
-![](figures/Extended_Data_Fig4_manuscript.png)
+ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=8)
+```
+![](figures/Extended_Data_Fig2_manuscript.png)
 **Extended Data Fig.
-4.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix.
+2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix.
 (A).
@@
 GTR+CAT.
+
+```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
+
+ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
+busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+
+ribo$matrix =
+  factor(
+    ribo$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+             "Whelan2017_full", "Whelan2017_strict") )
+
+p1 = ribo %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
+
+busco$matrix =
+  factor(
+    busco$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+               "Whelan2017_full", "Whelan2017_strict") )
+
+busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
+
+p2 = busco %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
+
+ ggarrange(p1, p2, labels = c("A", "B"))
+
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=4)
+
+```
+
+**Extended Data Fig.
+3.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
+(A).
+The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
+(C).
+The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
+
+```{r figure_matrix_overlap, fig.height = 12, fig.width = 20}
+
+matrix_overlap_rectangles =
+  lapply(
+    sequence_matrices[no_subsets],
+    function(x) lapply(sequence_matrices[no_subsets], function(y) overlap_rects(x, y))) %>%
+  unlist(recursive = FALSE) %>%
+  bind_rows()
+
+matrix_overlap_rectangles %>%
+  ggplot( ) +
+    scale_x_continuous(name = "Genes") +
+    scale_y_continuous(name = "Species") +
+    geom_rect( mapping = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = MSA ), alpha = 0.5) +
+    facet_grid( matrix_1 ~ matrix_2) +
+    theme( axis.text.x = element_text(angle = 45, hjust = 1),
+           strip.text.y = element_text(angle = 0),
+           strip.text.x = element_text(angle = 60))
+
+ggsave("figures/Extended_Data_Fig4_raw.pdf", width=20, height=12)
+
+```
+
+**Extended Data Fig.
+4.** Pairwise overlap between each of the primary matrices considered here.
+Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
+The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
 
@@
 
+ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=7)
 ```
@@
 
-![](figures/Extended_Data_Fig4_manuscript.png)
+![](figures/Extended_Data_Fig6_manuscript.png)
 
@@
 
-![](figures/Extended_Data_Fig5.png)
+![](figures/Extended_Data_Fig7.png)
 
@@
 
-![](figures/Extended_Data_Fig6.png)
+![](figures/Extended_Data_Fig8.png)
 
@@
 
-![](figures/Extended_Data_Fig7.png)
+![](figures/Extended_Data_Fig9.png)
 **Extended Data Fig.
@@
 More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimate substitutional heterogeneity in almost every case.
-They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model.
+They also reanalyzed datasets from Philippe2009 and Nosenko2013 using both CAT models and data partitioning with site-homogeneous model.
 The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.

### 2020-10-13 manuscript/manuscript.rmd
move all read.tree and read_tsv to kernel

@@
 
-cat_categories = read_tsv("../data_processed/tables/cat_categories.tsv")
-
 cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
@@
 ```{r table_study_summary}
-table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
+
 opts <- options(knitr.kable.NA = "")
@@
 
-taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
-outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
-
-tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+outgroup= taxa_map_whelan %>% filter (clade_assignment=="Choanoflagellida")
+
 tree1_rooted =  root(tree1, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
 tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
 tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
 tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
 tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-taxa_map=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
-outgroup= taxa_map %>% filter (clade_assignment=="Choanoflagellida")
-
-tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+outgroup= taxa_map_philippe %>% filter (clade_assignment=="Choanoflagellida")
+
 tree6_rooted =  root(tree6, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
 tree7_rooted =  root(tree7, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
 tree8_rooted =  root(tree8, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
 tree9_rooted =  root(tree9, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-  tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
   tree10_rooted =  root(tree10, outgroup$relabelled_name, resolved.root=TRUE)
@@
 
-ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
-busco = read_tsv("../data_processed/tables/busco_gene.tsv")
-
 ribo$matrix =

### 2020-10-13 manuscript/manuscript_kernel.R

@@ -430,4 +430,24 @@ write_csv(analyses_new, "./Supplementary_tables/Supplementary_Table_3.csv", na =
 write_csv(analyses_sensitive, "./Supplementary_tables/Supplementary_Table_6.csv", na = "NA", quote_escape = "double")
 write_csv(au_tests, "./Supplementary_tables/Supplementary_Table_7.csv", na = "NA", quote_escape = "double")
 
+## Move all read functions to kernel so all vairables are stored in Rdata
+
+cat_categories = read_tsv("../data_processed/tables/cat_categories.tsv")
+table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
+taxa_map_whelan=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
+taxa_map_philippe=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
+tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
+tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
+tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
+tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
+tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
+tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
+tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
+tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
+ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
+busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+
+
 

### 2020-10-13 manuscript/manuscript.rmd
add cross_validation codes to the manuscript

@@
 ```{r preliminaries}
+  library( rstatix )
   library( tidyverse )
@@
 
+![](figures/Extended_Data_Fig9.png)
+**Extended Data Fig. 10.**
+cross_validation =
+  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
+  gather("model", "score", `nCAT60`, `Poisson-CAT`)
+
+```{r cross_validation, fig.width=6, fig.height=7}
+p_w = cross_validation %>%
+  filter(matrix=="Whelan2017_strict") %>%
+  ggplot(aes(x=model, y=score)) +
+  geom_boxplot() +
+  ggtitle("Whelan2017") +
+  theme_classic()
+p_p = cross_validation %>%
+  filter(matrix=="Philippe2009_Chanimalia") %>%
+  ggplot(aes(x=model, y=score)) +
+  geom_boxplot() +
+  ggtitle("Philippe2009") +
+  theme_classic()
+ggarrange(
+  p_w,
+  p_p,
+  ncol=2, labels = c("A", "B"))
+
+```
+```{r}
+# https://www.datanovia.com/en/lessons/t-test-in-r/
+cross_validation %>%
+  group_by(matrix,model) %>%
+  get_summary_stats(score, type = "mean_sd")
+# Whelan
+Dw = cross_validation %>%
+  filter(matrix=="Whelan2017_strict")
+
+stat_test_w =
+  t.test(score ~ model, var.equal = TRUE, data=Dw)
+stat.test_w$p.value
+
+# Philippe
+Dp = cross_validation %>%
+  filter(matrix=="Philippe2009_Chanimalia")
+
+stat.test_p =
+    t.test(score ~ model, var.equal = TRUE, data=Dp)
+stat.test_p
+stat.test_p$p.value
+```
+
 \pagebreak

### 2020-10-13 manuscript/manuscript_kernel.R

@@ -448,6 +448,9 @@ tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.
 tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
 ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
 busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+cross_validation =
+  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
+  gather("model", "score", `nCAT60`, `Poisson-CAT`)
 
 
 

### 2020-10-13 manuscript/manuscript.rmd
add cross_validation table
minor changes to the code

@@
 
+# https://www.datanovia.com/en/lessons/t-test-in-r/
+cross_validation %>%
+  group_by(matrix,model) %>%
+  get_summary_stats(score, type = "mean_sd")
+# Whelan
+Dw = cross_validation %>%
+  filter(matrix=="Whelan2017_strict")
+
+stat_test_w =
+  t.test(score ~ model, var.equal = TRUE, data=Dw)
+P_value_Whelan2017_strict=stat.test_w$p.value
+
+# Philippe
+Dp = cross_validation %>%
+  filter(matrix=="Philippe2009_Chanimalia")
+
+stat.test_p =
+    t.test(score ~ model, var.equal = TRUE, data=Dp)
+
+P_value_Philippe2009=stat.test_p$p.value
 ```
@@
 
-![](figures/Extended_Data_Fig7.png)
+![](figures/Extended_Data_Fig10.png)
 
 **Extended Data Fig.
-7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3.
+7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3 and the subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+(A).
 Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
 Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-
-![](figures/Extended_Data_Fig8.png)
-
-**Extended Data Fig.
-8.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+(B).
 These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
@@
 
-![](figures/Extended_Data_Fig9.png)
-**Extended Data Fig. 10.**
 cross_validation =
@@
 ```
-```{r}
-# https://www.datanovia.com/en/lessons/t-test-in-r/
-cross_validation %>%
-  group_by(matrix,model) %>%
-  get_summary_stats(score, type = "mean_sd")
-# Whelan
-Dw = cross_validation %>%
-  filter(matrix=="Whelan2017_strict")
-
-stat_test_w =
-  t.test(score ~ model, var.equal = TRUE, data=Dw)
-stat.test_w$p.value
-
-# Philippe
-Dp = cross_validation %>%
-  filter(matrix=="Philippe2009_Chanimalia")
-
-stat.test_p =
-    t.test(score ~ model, var.equal = TRUE, data=Dp)
-stat.test_p
-stat.test_p$p.value
-```
+
+**Extended Data Fig. 10.** Cross validation..
 

### 2020-10-13 manuscript/manuscript.rmd
successfully reknit when move the read functions back to the manuscript

--- 2020-10-13 manuscript/manuscript.rmd
@@
 output:
+  html_document: default
   pdf_document:
     latex_engine: xelatex
-  html_document: default
   word_document: default
@@
   names( result_colors ) = c( "Ctenophora-sister", "Porifera-sister", "Unresolved" )
+  cat_categories = read_tsv("../data_processed/tables/cat_categories.tsv")
+table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
+taxa_map_whelan=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
+taxa_map_philippe=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
+tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
+tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
+tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
+tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
+tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
+tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
+tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
+tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
+tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
+tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
+ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
+busco = read_tsv("../data_processed/tables/busco_gene.tsv")
+cross_validation =
+  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
+  gather("model", "score", `nCAT60`, `Poisson-CAT`)
 
@@
 
+# https://www.datanovia.com/en/lessons/t-test-in-r/
+cross_validation %>%
+  group_by(matrix,model)
+# Whelan
+Dw = cross_validation %>%
+  filter(matrix=="Whelan2017_strict")
+
+stat_test_w =
+  t.test(score ~ model, var.equal = TRUE, data=Dw)
+P_value_Whelan2017_strict=stat_test_w$p.value
+
+# Philippe
+Dp = cross_validation %>%
+  filter(matrix=="Philippe2009_Chanimalia")
+
+stat_test_p =
+    t.test(score ~ model, var.equal = TRUE, data=Dp)
+
+P_value_Philippe2009=stat_test_p$p.value
 ```
@@
 ```{r table_study_summary}
-
+table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
 opts <- options(knitr.kable.NA = "")
-knitr::kable(table_study_summary) %>%
-  kable_styling(full_width = F)
+knitr::kable(table_study_summary)
 
@@
 This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
+
+cross_validation =
+  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
+  gather("model", "score", `nCAT60`, `Poisson-CAT`)
+
+```{r cross_validation, fig.width=6, fig.height=3.5}
+p_w = cross_validation %>%
+  filter(matrix=="Whelan2017_strict") %>%
+  ggplot(aes(x=model, y=score)) +
+  geom_boxplot() +
+  ggtitle("Whelan2017") +
+  theme_classic()
+p_p = cross_validation %>%
+  filter(matrix=="Philippe2009_Chanimalia") %>%
+  ggplot(aes(x=model, y=score)) +
+  geom_boxplot() +
+  ggtitle("Philippe2009") +
+  theme_classic()
+ggarrange(
+  p_w,
+  p_p,
+  ncol=2, labels = c("A", "B"))
+
+```
+
+**Extended Data Fig. 10.** Cross validation..
 

### 2020-10-21 manuscript/explorations/phylobayes_explorations.rmd

@@
 
-chain = parse_phylobayes_chain( "Whelan2017_strict.phy_Chain1_last50000.chain" )
+chain_whel_cat = parse_phylobayes_chain( "../../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain" )
 
 # Take just the last sample for now
-pb = chain[[length(chain)]]
+pb_whel_cat = chain_whel_cat[[length(chain_whel_cat)]]
 
-plot_chain( pb, site_occupancy )
+plot_chain( pb_whel_cat, site_occupancy )
 
@@
 
-**Figure CAT.** Plots for CAT.
+**Figure CAT.** Plots for Whelan CAT.
+
+```{r}
+
+chain_phil_cat = parse_phylobayes_chain( "../../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain" )
+
+# Take just the last sample for now
+pb_phil_cat = chain_phil_cat[[length(chain_phil_cat)]]
+
+plot_chain( pb_phil_cat, site_occupancy )
+
+```
+
+**Figure Philippe CAT.**
+
+## Assessing consistency of allocations
+
+The goal of these analyses is to investigate the consistency of site allocation to rare or frequent categories across generations and chains.
+The metric we looked at was, for each site, the mean and distribution of the mean number of sites allocated to the category the site belongs to across generations.
+
+First, we extracted these statistics from each chain ising `allocation_job.sh`, which runs `pb_allocation.py`
+
+```{r}
+
+allocations_phil_cat_chain1 = read_delim( "Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.category.summary", delim=" ", col_names=c("mean_chain1", "std_chain1") )
+allocations_phil_cat_chain2 = read_delim( "Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_2.category.summary", delim=" ", col_names=c("mean_chain2", "std_chain2") )
+
+allocations_phil_cat = 
+  bind_cols( allocations_phil_cat_chain1, allocations_phil_cat_chain2 ) %>% 
+  mutate( site=1:nrow(allocations_phil_cat_chain1)) %>%
+  mutate( sample=pb_phil_cat@allocation) %>%
+  mutate( sample_1=chain_phil_cat[[length(chain_phil_cat)-1]]@allocation)
+
+allocations_phil_cat %>% ggplot() + geom_point(aes(x=mean_chain1, y=mean_chain2), alpha=0.1) + ggtitle("Philippe2009")
+
+allocations_phil_cat %>% ggplot() + geom_point(aes(x=mean_chain1, y=sample_1), alpha=0.1) + ggtitle("Philippe2009")
+
+allocations_phil_cat %>% ggplot() + geom_point(aes(x=sample_1, y=sample), alpha=0.1) + ggtitle("Philippe2009")
+
+allocations_whel_cat_chain1 = read_delim( "Whelan2017_strict.phy_Poisson_CAT_Chain1.category.summary", delim=" ", col_names=c("mean_chain1", "std_chain1") )
+allocations_whel_cat_chain2 = read_delim( "Whelan2017_strict.phy_Poisson_CAT_Chain2.category.summary", delim=" ", col_names=c("mean_chain2", "std_chain2") )
+
+allocations_whel_cat = bind_cols( allocations_whel_cat_chain1, allocations_whel_cat_chain2 ) %>% mutate( site=1:nrow(allocations_whel_cat_chain1))
+
+allocations_whel_cat %>% ggplot() + geom_point(aes(x=mean_chain1, y=mean_chain2), alpha=0.1) + ggtitle("Whelan2017")
+
+```

### 2020-10-21 manuscript/manuscript.rmd

@@
   library( knitr )
+  options(digits=2)
   library( gridExtra )
@@
   cat_categories = read_tsv("../data_processed/tables/cat_categories.tsv")
-table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
-taxa_map_whelan=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Whelan2017/strict.phy")
-taxa_map_philippe=read_tsv("../reconciliation/taxonomy_info/taxon_table.tsv") %>% filter(original_matrix=="../considered_data/Philippe2009/Philippe2009.nex")
-tree1= read.tree("../trees_new/iqtree/Whelan2017_strict.model_test.treefile")
-tree2=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT60.con.tre")
-tree3=read.tree("../trees_new/sensitive/Whelan2017_strict.phy_Poisson_CAT90.con.tre")
-tree4=read.tree("../trees_new/phylobayes/Whelan2017_strict.bpcomp.con.tre")
-tree5=read.tree("../trees_new/phylobayes/Whelan2017_strict.phy_GTR_CAT.con.tre")
-tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
-tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
-tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
-tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
-tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
-ribo = read_tsv("../data_processed/tables/ribosomal_gene.tsv")
-busco = read_tsv("../data_processed/tables/busco_gene.tsv")
-cross_validation =
-  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
-  gather("model", "score", `nCAT60`, `Poisson-CAT`)
 
@@
 
-^2^2State Key Laboratory of Rice Biology and Ministry of Agriculture Key Lab of Molecular Biology of Crop Pathogens and Insects, Institute of Insect Sciences, Zhejiang University, Hangzhou 310058, China
-
-^3^Department of Biological Science, Vanderbilt University, Nashville 37235, USA
+^2^State Key Laboratory of Rice Biology and Ministry of Agriculture Key Lab of Molecular Biology of Crop Pathogens and Insects, Institute of Insect Sciences, Zhejiang University, Hangzhou 310058, China
+
+^3^Department of Biological Science, Vanderbilt University, Nashville, TN 37235, USA
 
@@
 
-Xing-Xing Shen: 0000-0002-8436-595X
+Xing-Xing Shen: 0000-0001-5765-1419
 
@@
 2B, Extended Data Fig.
-1).
+4).
 Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
@@
 This is consistent wwith the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
-We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60.
-We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Extended Data Table 2).
+We then inferred support under the best-fit model (Supplementary Table 1), except for the three largest matrices where we used LG+C60.
+We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 1).
 
@@
 # https://www.datanovia.com/en/lessons/t-test-in-r/
-cross_validation %>%
-  group_by(matrix,model)
+
 # Whelan
@@
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
-We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model (Extended Data Fig.
-5E).
+We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
 Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.
@@
 This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.
+Moreover, we also conducted Bayesian cross-validation experiments employed in PhyloBayes to evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices.
+Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig.
+6).
+These results are further corroborated with our *t*-test analysis that there is no significant differences between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).
 
@@
 To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
-6).
+7).
 By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
@@
 In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
-6; Supplementary Table 7).
+7; Supplementary Table 7).
 In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-6).
+7).
 For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
@@
 However, our own analyses (Supplementary Information section S3, Supplementary Figs.
-5-7) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+8-9) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 
-All tree files, intermediate results and scripts/commands associated with this study are available at https://github.com/caseywdunn/animal_root, doi will be provided upon publication.
+### Data and code availability
+
+The main data and results associated with the main text and supplementary materials are available in the GitHub data repository at https://github.com/dunnlab/animal_tree_root.
+All tree files, intermediate results and scripts/commands associated with this study are available in the Figshare data repository at https://doi.org/10.6084/m9.figshare.13085081.v1.
 
@@
 
-The statistics presented in  Extended Data Figs. 7 and 8 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+The statistics presented in  Extended Data Fig. 8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 We thank members of the Dunn and Rokas laboratories for discussions and comments.
-We are grateful to Nicolas Lartillot, Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript.
+We are grateful to Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript.
 We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.
-Yuanning Li was previously supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad.
+Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad.
 
 ## Author contributions
+C.W.D., Y.L.
+conceived this study.
+C.W.D.
+and A.R.
+supervised the project.
+Y.L., C.W.D., X-X.S.
+and B.E.
+conducted analyses.
+Y.L., C.W.D.
+B.E.
+wrote the paper.
+All authors discussed the results and implications and commented on the manuscript at all stages.
 
@@
 knitr::kable(table_study_summary)
-
-```
-
-**Extended Data Table 2.** The models selected by ModelFinder for each matrix.
-```{r table_modelfinder}
-analyses_new %>%
-  filter(modelfinder == TRUE) %>%
-  select( matrix, clade, result, model_summary ) %>%
-  kable()
 
@@
 This adds a considerable number of parameters for describing a similar overall space.
-
-```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
-
-p=au_tests %>%
-  unite(Matrix, c(manuscript, matrix, model)) %>%
-  group_by(Matrix, result) %>%
-  count() %>%
-  group_by(Matrix) %>%
-  mutate(percent = prop.table(n)) %>%
-  ggplot(aes(x = Matrix, y = percent, fill = result)) +
-    geom_bar(stat = "identity") +
-    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
-    coord_flip() +
-    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
-
-ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=12)
-
-```
-
-![](figures/Extended_Data_Fig6_manuscript.png)
-
-**Extended Data Fig.
-6** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
-The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
-Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models.
-The GLS values for each gene in each data matrix are provided in Table S5.
-We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
-
-![](figures/Extended_Data_Fig10.png)
-
-**Extended Data Fig.
-7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3 and the subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
-(A).
-Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
-Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
-(B).
-These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
-The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
-The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
-Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Extended Data Fig.
-5 correspond to the 20 and SR6 Whelan cells shown here.
-The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
-The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
-
-![](figures/Extended_Data_Fig9.png)
-**Extended Data Fig.
-9.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
-These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
-Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
-For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
-This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
-
-cross_validation =
-  read_tsv("../data_processed/tables/cross_validation.tsv") %>%
-  gather("model", "score", `nCAT60`, `Poisson-CAT`)
 
@@
 
-```
-
-**Extended Data Fig. 10.** Cross validation..
+ggsave("figures/Extended_Data_Fig6_raw.pdf", width=6, height=7)
+
+```
+
+**Extended Data Fig. 6.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+
+```{r figure_phylogenetic_signal, fig.width=6, fig.height=12}
+
+p=au_tests %>%
+  unite(Matrix, c(manuscript, matrix, model)) %>%
+  group_by(Matrix, result) %>%
+  count() %>%
+  group_by(Matrix) %>%
+  mutate(percent = prop.table(n)) %>%
+  ggplot(aes(x = Matrix, y = percent, fill = result)) +
+    geom_bar(stat = "identity") +
+    scale_fill_manual(values = c("#E69F00", "#ebd196", "#a2e8d5", "#009E73")) +
+    coord_flip() +
+    theme(axis.text.y = element_blank(), axis.title.y = element_blank())
+
+ggsave("figures/Extended_Data_Fig7_raw.pdf", width=6, height=8)
+
+```
+
+![](figures/Extended_Data_Fig7_manuscript.png)
+
+**Extended Data Fig.
+7** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
+Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models.
+The GLS values for each gene in each data matrix are provided in Table S5.
+We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.
+
+![](figures/Extended_Data_Fig8.png)
+
+**Extended Data Fig.
+8.** Graphical representations of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3 and the subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+(A).
+Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved).
+Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
+(B).
+These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane.
+The eight analyses are for two datasets (Chang and Whelan) and four coding schemes.
+The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6.
+Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
+The 20-state and 6-state points on the plots in Extended Data Fig.
+8A correspond to the 20 and SR6 Whelan cells shown here.
+The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
+The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
+
+![](figures/Extended_Data_Fig9.png)
+**Extended Data Fig.
+9.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
+Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point).
+For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding.
+This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional hetezrogeneity across lineages.
 
@@
 ## Supplementary Information
+
+## Supplementary tables
+
+**Supplementary Table 1.** The models selected by ModelFinder for each matrix.
+```{r table_modelfinder}
+analyses_new %>%
+  filter(modelfinder == TRUE) %>%
+  select( matrix, clade, result, model_summary ) %>%
+  kable()
+
+```
 
@@
 Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
-7).
+8A).
 Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
@@
 This provides 32 analyses that they report in their Table 3 and that we present graphically here as Extended Data Fig.
-7.
+8A.
 There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Extended Data Fig.
-7).
+8A).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 This leaves eight GTR+CAT analyses (Extended Data Fig.
-7).
+8A).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
@@
 Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
-8).
+8B).
 The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig.
-8), indicating that it did not converge as well as other analyses.
+8B), indicating that it did not converge as well as other analyses.
 

### 2020-10-21 manuscript/scrap.rmd

--- 2018-10-19 manuscript.rmd
@@
----
-csl: plos.csl
-output:
-  github_document: default
-  html_document: default
-  pdf_document: default
-bibliography: manuscript.bib
----
+# Scrap
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-	message = FALSE,
-	warning = FALSE,
-	echo=FALSE,
-	dpi=300,
-	cache=TRUE
-	)
-```
-
-```{r preliminaries}
-library( tidyverse )
-library( magrittr )
-library(viridis)
-
-source( "functions.R" )
-```
-
-```{r}
-
-# A colorblind friendly palette from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette:
-cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-
-result_colors = c( cbPalette[2], cbPalette[4], cbPalette[1] )
-names( result_colors ) = c( "Ctenophora-sister", "Porifera-sister", "Unresolved" )
-
-papers = read_tsv( "tables/papers.tsv" )
-datasets = read_tsv( "tables/datasets.tsv" )
-analyses = 
-	read_tsv( "tables/analyses.tsv", col_types = 'iccccccccccclnnnncc' ) %>%
-	mutate( taxon_sampling = factor( taxon_sampling, levels= c( 'Choanimalia', 'Holozoa', 'Opisthokonta' ) ) )
-
-bootstrap_threshold = 90
-posterior_prob_threshold = 95
-
-analyses$result = "Unresolved"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_porifera_sister >= posterior_prob_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "Bayesian" ) & (analyses$support_ctenophora_sister >= posterior_prob_threshold) ] = "Ctenophora-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_porifera_sister >= bootstrap_threshold) ] = "Porifera-sister"
-analyses$result[ ( analyses$inference == "ML" ) &       (analyses$support_ctenophora_sister >= bootstrap_threshold) ] = "Ctenophora-sister"
-analyses$result = factor( analyses$result )
-
-analyses$model_summary = analyses$model_rate_matrix
-long_summary = paste(analyses$model_rate_matrix, analyses$model_equilibrium, sep="+")
-analyses$model_summary[ ! is.na(analyses$model_equilibrium) ] = long_summary[ ! is.na(analyses$model_equilibrium) ]
-analyses$model_summary[ grepl("WAG", analyses$model_summary) ] = "WAG" # Simplify WAG
-analyses$model_summary = factor( analyses$model_summary, levels=c("GTR+CAT", "F81+CAT", "LG", "WAG") )
-
-```
-
-# Rooting the animal tree of life
-
-Casey W. Dunn^1^*, Benjamin Evans^2^, and Yuanning Li^1^
-
-^1^Department of Ecology and Evolutionary Biology, Yale University
-
-^2^Yale Center for Research Computing, Yale University
-
-\* Corresponding author, casey.dunn@yale.edu
-
-## Abstract
-
-## Introduction
-
-In the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses.
-Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested.
-In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].
-The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative.
-Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither.
-As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem.
-Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support.
-Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies.
-We hope that this provides an integrative overview of the challenge and provides direction for future studies.
-We also hope that the work we have done here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.
-
-![](figures/Figure_overview.png)
-**Fig XXOverview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
-(B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade.
-Testing these hypotheses requires evaluating the support for each of these alternative nodes.
-(C) A ctenophore.
-(D) A sponge.
-
-## Variation across studies
-
-### Models of molecular evolution
-
-Models of molecular evolution have several components that each consider different aspects of the evolutionary process.
-The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.
-
-The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include:
-
-- F81 [@Felsenstein:1981vk] corresponds to equal rates between all states.
-The F81 matrix is also sometimes referred to as the Poisson matrix.
-It has no free parameters to estimate since all off-diagonal elements are set to 1.
-
-- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix.
-It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated.
-It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.
-
-- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids.
-It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal.
-This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology.
-This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).
-
-While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled.
-There are couple approaches that have been used in the studies considered here:
-
-- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.
-
-- Gamma rate heterogeneity.
-Each site is assigned to a different rate class with its own rate value.
-This accommodates different rates of evolution across different sites.
-Gamma is used so commonly that sometimes it isn't even specified, making it difficult at times to know if a study uses Gamma or not.
-
-The vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids. There are a few approaches that have been used across the studies considered here:
-
-- Empirical site homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.
-
-- Estimated site homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.
-
-- CAT site heterogeneous [@Lartillot:2004dq].
-Each site is assigned to a class with its own equilibrium frequencies.
-The number of classes, assignment of sites to classes, and equilibrium frequencies within the data are all estimated in a Bayesian framework.
-
-Models can be assembled by selecting different options for all these different components.
-The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention.
-For example, on the questions considered here F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency.
-LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency.
-This is further confused by the abbreviations that are used for models.
-Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix.
-CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
-One could, though, run a WAG+CAT model.
-
-To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma).
-If there are no modifiers, then it is implied that site homogeneous models are used.
-
-### Gene sampling
-
-### Outgroup taxon sampling
-
-XXX
-
-**Fig XXOutgroup.** The animals and their outgroups, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
-
-Choanimalia, Holozoa, Opisthokonta
-
-### Ingroup taxon sampling
-
-Sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.
-
-## Overview of published analyses
-
-### Alignment composition
-
-```{r taxonomy}
-
-clades = c( "Fungi", "Ichthyosporea", "Filasterea", "Choanoflagellida", "Ctenophora", "Porifera", "Placozoa", "Bilateria", "Cnidaria" )
-
-clade_colors = viridis( length(clades) )
-names( clade_colors ) = clades
-
-taxa = 
-	read_tsv("taxonomy_info/taxon_table.tsv") %>% 
-	distinct( relabelled_name, clade_assignment, ncbi_tax_id ) %>%
-	dplyr::rename( taxon=relabelled_name, clade=clade_assignment ) %>%
-	mutate( clade = factor( clade, levels=clades )  )
-
-```
-
-```{r}
-
-a_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.phy"
-p_file = "~/Desktop/animal_root/blast/phylips/Chang2015_Chang2015.nex"
-msa = PartitionedMultipleAlignment( a_file, p_file, taxa )
-
-D=clade_rects(msa)
-
-ggplot() + 
-	scale_x_continuous(name="x") + 
-	scale_y_continuous(name="y") +
-	geom_rect(data=D, mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) + 
-	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
-	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
-	scale_fill_discrete(guide = guide_legend(reverse=TRUE))
-```
-
-XXX
-
-**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by taxon sampling.
-Horizontal size is proportional to the number of genes (XXOr should it be sites?) sampled, vertical size to the number of taxa sampled.
-
-XXX
-
-**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc).
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-
-### Alignment overlap
-
-XXX
-
-**Fig XXAlignment overlap.** Pairwise overlap between each of the primary matrices considered here.
-Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
-The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
-
-### Support for Porifera-sister and Ctenophora-sister
-
-```{r support}
-
-analyses %>% 
-	ggplot( aes(x=model_summary, y=taxon_sampling, col=result) ) +
-		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
-		scale_colour_manual(values = result_colors) +
-		theme_classic()
-```
-
-A total of `r nrow(analyses)` analyses were transcribed from the literature.
-
-## New analyses of published matrices
-
-One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways.
-Here we reanalyze the primary matrices from each study under consistent conditions with iqtree.
-We selected this tool because it has greater model flexibility than other tools and is very fast.
-
-We first tested a variety of models for each matrix, and inferred support under the selected model. We then analyzed every matrix under a panel of standard models, including XXX.
-
-### Comparison of iqtree and phylobayes results
-
-Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister.
-This has been addressed with CAT models.
-iqtree provides a new family of C models that also address site heterogeneity.
-Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results.
-This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.
-
-## New analyses of new matrices
-
-Based on the variation across analyses, we constructed new matrices with altered taxon and gene sampling to test specific hypotheses about differences in support.
-
-## The current state of understanding
-
-### Interpretting variation support
-
-External criteria, eg posterior predictive scores, model fit etc
-
-## Next steps
-
-## Conclusion
-
-## Methods
-
-All files associated with this analysis are available at https://github.com/caseywdunn/animal_root
-
-### Data wrangling
-
-### Matrix comparison and annotation
-
-### Phylogenetic analyses
-
-## Ackowledgements
-
-## Author contributions
-
-\pagebreak
-
-# Supplemental Information
-
-## Details of published analyses
-
-### Dunn *et al.* 2008
-
-Dunn *et al.* [@Dunn:2008ky] added Expressed Sequence Tag (EST) data for 29 animals.
-It was the first phylogenomic analysis that included ctenophores, and therefore that could test the relationships of both Ctenophora and Porifera to the rest of animals.
-It was the first phylogenetic analysis to recover Ctenophora as the sister group to all other animals.
-
-The data matrix was constructed using a semi-automated approach.
-Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications.
-Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors).
-The remaining gene trees with more than one sequence for any taxon were then manually inspected.
-If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded.
-If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded.
-Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix.
-Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships.
-Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix.
-This resulted in the 64-taxon matrix that is the focus of most of their analyses.
-Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].
-
-Regarding the recovery of Ctenophora-sister, the authors concluded:
-
-> The placement of ctenophores (comb jellies) as the sister group to all other sampled metazoans is strongly supported in all our analyses.
-This result, which has not been postulated before, should be viewed as provisional until more data are considered from placozoans and additional sponges.
-
-Note that there was, in fact, an exception to strong support.
-An analysis of the 40 ribosomal proteins in the matrix recovered Ctenophora-sister with only 69% support.
-This study did not include *Trichoplax*.
-
-### Philippe *et al.* 2009
-
-Philippe *et al.* 2009 [Philippe:2009hh]...
-
-### Hejnol et al. 2009
-
-### Pick *et al.* 2010
-
-Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling.
-They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky].
-The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*.
-They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
-They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
-
-All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping.
-All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
-Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
-With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
-This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.
-
-The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected".
-Although the posterior probabilities supporting Porifera-sister are not strong, they conclude:
-
-> Results of our analyses indicate that sponges are the sister group to the remaining Metazoa, and Placozoa are sister group to the Bilateria
-
-They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh].
-Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
-
-## Matrix mapping
-
-Taxa and partition correspondence across manuscripts was assessed by comparing all sequences for each taxon in each partition across all matrices with diamond blast.
-Based on inspection of sequence similarity, we excluded all comparisons with less than 99% identity and greater than 10^-25^ e-value.
+Misc code removed fromt he main manuscript that may still be of use later.
 
@@
 
-```{r taxon_mapping_across_studies}
+```{r taxon_mapping_across_studies, eval=FALSE}
 
 Dtaxaxman = 
-	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/cross_mscript_taxa.tsv") %>%
+	read_tsv("../reconciliation/blast/graphs/cross_mscript_taxa.tsv") %>%
 	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
@@
 
-```{r taxon_mapping}
+```{r taxon_mapping, eval=FALSE}
 
 Dtaxa = 
-	read_tsv("~/Dropbox/repo/animal_root/blast/graphs/taxon_graph.tsv") %>%
+	read_tsv("../blast/graphs/taxon_graph.tsv") %>%
 	arrange( mscript1, dataset1, taxon1, mscript2, dataset2, taxon2 )
@@
 
-taxonomy_reference = read_tsv("taxonomy_info/taxon_table.tsv")
-
 ```
-
-### Partitions
-
-## References

### 2020-10-24 manuscript/manuscript.rmd
updaed comments from Antonis

@@
 output:
-  html_document: default
+  word_document: default
   pdf_document:
     latex_engine: xelatex
-  word_document: default
+  html_document: default
   github_document: default
@@
 
-Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*
+Yuanning Li^1,3^, Xing-Xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*
 
@@
 
-^3^Department of Biological Science, Vanderbilt University, Nashville, TN 37235, USA
+^3^Department of Biological Sciences, Vanderbilt University, Nashville, TN 37235, USA
 
@@
 
-There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree.
-This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
+There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics.
+Resolving the debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
 Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
-These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software.
-Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses.
-It had previously been suggested that site heterogeneous models favor Porifera-sister, but we find that this is not the case.
-Porifera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data.
-Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
+These analyses differ in many respects, including in the genes considered, species considered, molecular evolution models, and software.
+Here we systematically explore the rooting of the animal tree of life under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing a comprehensive set of new standardized analyses.
+It has previously been suggested that site-heterogeneous models favor Porifera-sister, but we find that this is not the case.
+Rather, Porifera-sister is only obtained under a narrow set of conditions when the categories of sites are unconstrained in number and include a large number of categories that apply to a small fraction of data.
+Our analyses shed light on the sources of variation between studies aimed at resolving a highly contentious and important branch of the tree of life, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
 
@@
 Historically, there was little debate about the root of the animal tree of life.
-Porifera-sister (Figs.
-1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested.
-By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws].
-The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs.
-1A, 1D).
-Since then, many more relevant phylogenomic studies have been published (Fig.
+Porifera-sister (Fig.
+1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig.
+1B), was widely accepted though rarely tested.
+By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig.
+1D) in the animal tree of life [@Wallberg:2004ws].
+The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig.
+1A).
+Since then, many more phylogenomic studies have been published (Fig.
 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1).
-The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig.
+The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig.
 3A).
 But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity.
-This is because methodological variance is often confounded, for example when a particular model is only available in one software package.
-There are a variety of things that make this a particularly challenging problem.
-For one, the relevant nodes are the deepest in the animal tree of life.
-Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
-While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs.
+Several factors make resolution of the root of the animal tree a particularly challenging problem.
+For one, the branch in question is the deepest in the animal tree of life.
+Another factor that has been invoked is that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
+While some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs.
 1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].
 
-To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework.
-Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 15 studies (Fig.
+To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework.
+Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig.
 2A, Extended Data Table 1).
-These include all such phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations.
+This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations.
 Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
@@
 **Fig.
-1.** Two alternative phylogenetic hypotheses from previous phylogenomic studies.
+1.** Two alternative phylogenetic hypotheses for the root of the animal tree.
 (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade.
@@
 (D) A ctenophore, commonly known as comb jellies.
-(E) A poriferan, commonly known as sponges.
+(E) A poriferan, commonly known as a sponge.
 
@@
 The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
-The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites.
+The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), which means that the same matrix is used for all sites.
 The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others.
@@
 The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention.
-While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
+While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of species sampled.
+Only 11 matrices were shown here since three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
 
@@
 Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria.
-Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig.
+Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig.
 4).
@@
 4).
-This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.
+This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.
 
@@
 This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination.
-The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT.
+The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT.
 Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
@@
 Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results.
-This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct.
+This indicates that the traditional framing of the problem, that accommodating site heterogeneity leads to support for Porifera-sister, is not correct.
 Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
 
-### Category number explains differences between heterogeneous analyses
-
-There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models.
+### Category number explains differences between site-heterogeneous analyses
+
+There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site-heterogeneous models.
 In published analyses of the animal root, these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes.
@@
 Differences in analysis results could therefore be due to differences in the number of categories.
-The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses.
+The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices.
 This requires a very large number of additional parameters that must be estimated from the data.
-Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5).
+Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model is overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 categories are probably already too high according (Supplementary Table 5).
 These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
 
-We examined the specific impact of these large difference in category number on animal root position.
-It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT.
+We examined the specific impact of this large difference in category number on animal root position.
+It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT.
 We therefore varied the number of categories in Phylobayes analyses (Fig.
@@
 
-These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig.
-1; Philippe2009 Extended Data Fig.
-2).
-Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
+These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig.
+1; Philippe2009: Extended Data Fig.
+2): when outgroup sampling is restricted (Choanimalia), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
 This is consistent across published analyses and our new panels of analyses.
-The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects.
-If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
-
-There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
-It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available.
+The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects.
+If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
+
+There are several effects of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
+It could be that there is a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available.
 This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
@@
 Each new category adds the same number of parameters, but that category can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories.
+If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain.
 That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
-5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories.
+5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, repectively.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
@@
 This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.
-Moreover, we also conducted Bayesian cross-validation experiments employed in PhyloBayes to evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices.
+To formally evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments employed in PhyloBayes.
 Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig.
-6).
-These results are further corroborated with our *t*-test analysis that there is no significant differences between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).
+6); a t-test analysis shows that there is no significant difference between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).
 
@@
 **Fig.
-4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
+4.**  Sensitivity analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
 Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories.
@@
 
-To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
+To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
 7).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE.
-Moreover, the outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrixin each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis.
+Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models.
 This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].
@@
 Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
-In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
+In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
 7; Supplementary Table 7).
-In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
+In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
 7).
-For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
+For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
 
@@
 
-Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem [@laumer2019revisiting].
-By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
+Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting].
+By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
 This demonstrates an important interaction between model selection and taxon sampling.
@@
 
-Overcoming the challenges associated with this difficult and important question goes beyond models.
-In particular, there are many organisms that are highly relevant to this problem, in particular better sampling of outgroups, ctenophores, and sponges, for which no genome or transcriptome data are available [@King:2017ie].
-In particular, there are very few chromosome-level genome assemblies for animals outside of Bilateria.
-There will be important advances in genome sequencing in coming years that will have a major impact on this problem.
-This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcriptomic data.
-We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
+Overcoming the challenges associated with resolving this difficult and important branch of the tree of life goes beyond models.
+For example, there are many organisms that are highly relevant to this problem, in particular outgroup, ctenophore, and sponge taxa, for which no genome or transcriptome data are available [@King:2017ie].
+More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria.
+But resolution of the debate will likely require more than just more data – for example, high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction, reduce the possibility of contamination, and address occupancy issues in transcriptomic data.
+We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
 
@@
 
-- CAT site heterogeneous.
+- CAT site-heterogeneous.
 Each site is assigned to a class with its own equilibrium frequencies.

### 2020-10-26 manuscript/manuscript.rmd
reknit

@@
 output:
-  word_document: default
+  html_document: default
   pdf_document:
     latex_engine: xelatex
-  html_document: default
+  word_document: default
   github_document: default

### 2020-10-26 manuscript/manuscript.rmd
rewording overparameterization parts

@@
 There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics.
-Resolving the debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
+This debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
 Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister).
@@
 It has previously been suggested that site-heterogeneous models favor Porifera-sister, but we find that this is not the case.
-Rather, Porifera-sister is only obtained under a narrow set of conditions when the categories of sites are unconstrained in number and include a large number of categories that apply to a small fraction of data.
-Our analyses shed light on the sources of variation between studies aimed at resolving a highly contentious and important branch of the tree of life, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.
+Rather, Porifera-sister is only obtained under a narrow set of conditions when the number of site-heterogeneous categories is unconstrained and range into the hundreds.
+Site-heterogenous models with a fixed number of dozens of categories support Ctenophora-sister, and cross-validation indicates that such models fit the data just as well as the unconstrained models.
+Our analyses shed light on an important source of variation between phylogenomic analyses of the animal root.
+The datasets and analyses consolidated here will also be a useful test-platform for development of phylogenomic methods for this and other difficult problems.
 
@@
 Several factors make resolution of the root of the animal tree a particularly challenging problem.
-For one, the branch in question is the deepest in the animal tree of life.
-Another factor that has been invoked is that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
-While some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs.
-1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].
+For one, the nodes in question are the deepest in the animal tree of life.
+Another factor that has been invoked is branch lengths (*e.g.* Extended Data Figs.
+1, 2), which are impacted by both divergence times and shifts in rate of evolution.
+Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages.
+The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans.
+The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
 
@@
 The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another.
-The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
+The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different.
+If different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
 The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), which means that the same matrix is used for all sites.
@@
 The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention.
-While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
+While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components -- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix.
 In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation.
@@
 In all cases, models with C60 fit these matrices better than the site-homogeneous models.
-This is consistent wwith the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
+This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
 We then inferred support under the best-fit model (Supplementary Table 1), except for the three largest matrices where we used LG+C60.
@@
 
-All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig.
+All IQ-TREE analyses, apart from unresolved analyses (for Moroz2014_3d and all Nosenko2013 matrices), supported Ctenophora-sister (Fig.
 3B).
@@
 This indicates that the traditional framing of the problem, that accommodating site heterogeneity leads to support for Porifera-sister, is not correct.
-Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
+Instead, there is something about the PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.
 
@@
 The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices.
-This requires a very large number of additional parameters that must be estimated from the data.
-Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model is overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 categories are probably already too high according (Supplementary Table 5).
-These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].
+This requires a very large number of additional estiamted parameters.
+ModelFinder selects site heterogeneous C models over site homogenous models, but often selects site heterogeneous modelds with fewer than 60 categories (Supplementary Table 5).
+This indicates that BIC, the criterion used to select models in ModelFinder, finds even 60 categories to be too many.
 
@@
 Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig.
-6); a t-test analysis shows that there is no significant difference between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).
+6); a t-test analysis shows that there is no significant difference between the model-fit of CAT and nCAT60 models (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$).
 

### 2020-10-27 manuscript/manuscript.rmd
made pass through whole paper considering feedback from Nicolas

@@
 
+Casey W. Dunn: 0000-0003-0628-5150
+
 Antonis Rokas: 0000-0002-7248-6551
-
-Casey W. Dunn: 0000-0003-0628-5150
 
@@
 This requires a very large number of additional estiamted parameters.
-ModelFinder selects site heterogeneous C models over site homogenous models, but often selects site heterogeneous modelds with fewer than 60 categories (Supplementary Table 5).
-This indicates that BIC, the criterion used to select models in ModelFinder, finds even 60 categories to be too many.
 
@@
 For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model.
-
 Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
@@
 
-There are several effects of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister).
-It could be that there is a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available.
-This would suggest that we should favor the result under a very large number of categories, Porifera-sister.
-Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters.
-This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister.
-One way to assess this is to look at the allocation of sites to equilibrium frequency categories.
-Each new category adds the same number of parameters, but that category can be allocated to any number of sites.
-If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain.
-That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
+The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories.
+ModelFinder selects site heterogeneous C60 models (Supplementary Table 5), but IQtree often gives a warning under C60 that the model may be overfit, with too many categories, because some mixture weights are close to 0.
+While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data.
+Cross Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes].
+It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister reseults obtained in unconstrained CAT analyses.
+To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes.
+We found nearly identical distributions of cross validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
+6).
+A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$).
+This indicates that 60 categories in a Poisson+nCAT60 model are as as adequate at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model.
+This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.
+
+In a site-heterogeneous model, each site is allocated to one of the categories in each sample.
+Different categories can have very different numbers of sites.
+We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
 5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, repectively.
@@
 5A-B).
-This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.
-To formally evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments employed in PhyloBayes.
-Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig.
-6); a t-test analysis shows that there is no significant difference between the model-fit of CAT and nCAT60 models (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$).
-
-It could be that the sites that are allocated to rare categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
+The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on predictive power of these far more complex models.
+
+It could be that the sites that are allocated to rare categories are very different from the sites in the common categories, in which case the added categories are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
 Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
-We find support for this second scenario (Extended Data Fig5.
-3C-D).
+Plots of amino acid space are consistent wit this second scenario (Extended Data Fig5.
+C-D).
 
@@
 Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting].
-By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses.
-This demonstrates an important interaction between model selection and taxon sampling.
-Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency.
-Site-homogeneous models have a single category that is applied to all sites.
-C models, as implemented in IQ-TREE, have 10-60 categories.
-CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT.
-When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories.
-As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig.
+By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some such analyses.
+Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency (Fig.
 4).
-As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.
-
-Overcoming the challenges associated with resolving this difficult and important branch of the tree of life goes beyond models.
+The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though.
+This suggests that we shouldn't priviledge these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.
+
+Pin-pointing category number as an issue with large effect on analyses of the animal root will help guide future analyses that address this question.
+We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and that this problem can be a test-bed to develop methods and tools that will help with other difficult phylogenetic problems as well.
+Advances on the question of the animal root will come from progress on other fronts as well.
 For example, there are many organisms that are highly relevant to this problem, in particular outgroup, ctenophore, and sponge taxa, for which no genome or transcriptome data are available [@King:2017ie].
 More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria.
-But resolution of the debate will likely require more than just more data – for example, high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction, reduce the possibility of contamination, and address occupancy issues in transcriptomic data.
-We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
+It will be critical for this to change in coming years.
 

### 2020-10-27 manuscript/manuscript.rmd
removed extended fig 5c-d

@@
 
-It could be that the sites that are allocated to rare categories are very different from the sites in the common categories, in which case the added categories are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain.
-Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value.
-Plots of amino acid space are consistent wit this second scenario (Extended Data Fig5.
-C-D).
-
 ``` {r sensitive, fig.width=7.2, fig.height=7.2}
@@
 
-```{r n_cat_render, fig.width=6, fig.height=7}
+```{r n_cat_render, fig.width=6, fig.height=3.5}
 
@@
   gg_rank_whel,
-  gg_mds_phil,
-  gg_mds_whel,
   ncol  = 2,
-  nrow = 2,
-  labels = c("A", "B", "C", "D"),
+  labels = c("A", "B"),
   common.legend = TRUE
@@
 
-ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=7)
+ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=3.5)
 ```
@@
 The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
-(C,D) MDS plots for the amino acid frequencies.
-Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies.
-A single MDS analysis was done for all analyses, so that the projections are the same in each plot.
-For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories.
-This adds a considerable number of parameters for describing a similar overall space.
 
@@
 **Extended Data Fig.
-7** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
@@
 
-Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore the phylogenetic relationship among early diverging animals by adding 9 new species.
+Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore the deepest animal phylogenetic relationships by adding 9 new species.
 The data matrix was assembled based on the phylogenetic analysis using Poisson+CAT model, which strongly supported Porifera-sister, and ctenophores were recovered as sister to cnidarians forming the "coelenterate" clade.

### 2020-10-27 manuscript/manuscript.rmd
cleaned up figure 2

@@
 2.** An overview of previous phylogenomic studies on animals.
-(A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study.
-(B) Each of the primary matrices considered here, color coded by taxon sampling.
+The horizontal axis is a time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study.
+Each of the primary matrices considered here is shown above the axis, color coded by taxon sampling.
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of species sampled.

### 2020-10-27 manuscript/manuscript.rmd
small cleanups, animal_root -> animal_tree_root

@@
 #### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/animal_root/feuda_2017_response.
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017.
 The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.

### 2020-10-27 manuscript/manuscript_kernel.R

@@ -1,5 +1,5 @@
 # The computationally intensive analyses for the manuscript
-# presented at https://github.com/caseywdunn/animal_root/tree/master/manuscript .
+# presented at https://github.com/dunnlab/animal_tree_root/tree/master/manuscript .
 #
 # Executing this code generates manuscript.RData, which contains analysis
 # results. That file is then read by manuscript.rmd for rendering and

### 2020-10-27 manuscript/manuscript.rmd
final pass to the manuscript and update figure and table orders

@@
 
-There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics.
+There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most challenging problems in animal phylogenetics.
 This debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system.
@@
 Site-heterogenous models with a fixed number of dozens of categories support Ctenophora-sister, and cross-validation indicates that such models fit the data just as well as the unconstrained models.
-Our analyses shed light on an important source of variation between phylogenomic analyses of the animal root.
-The datasets and analyses consolidated here will also be a useful test-platform for development of phylogenomic methods for this and other difficult problems.
+Our analyses shed light on an important source of variation between phylogenomic studies of the animal root.
+The datasets and analyses consolidated here will also be a useful test-platform for the development of phylogenomic methods for this and other difficult problems.
 
@@
 The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans.
-The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].
+The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artifact of long-branch attraction to outgroups [@pisani2015genomic].
 
@@
 Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig.
-2A, Extended Data Table 1).
+2, Extended Data Table 1).
 This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations.
-Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].
+Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data, and gene-partition data is not available from one study [@Pick:2010eb].
 
@@
 The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another.
-The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different.
+The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81), or different.
 If different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR).
@@
 If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone.
-Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
+Gamma-rate heterogeneity, a scalar that accommodates the total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.
 Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).
@@
 As a result, the particular genes selected for analysis vary widely (Fig.
-2B, Extended Data Fig.
+2, Extended Data Fig.
+3).
+Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
 4).
-Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
-3).
 
@@
 Ingroup taxon sampling also varies widely between studies (Fig.
-2B, Extended Data Fig.
-4).
+2, Extended Data Fig.
+3).
 Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria.
 Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig.
-4).
+3).
 Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig.
-4).
+3).
 This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.
@@
 Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).
 
@@
 The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis.
-(A) Analyses transcribed from the literature (related to Supplementary Table 2).
-(B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).
+(A) Analyses transcribed from the literature (related to Supplementary Table 1).
+(B) New phylogenomic analyses conducted in this study (related to Supplementary Table 2).
+Over 1,011,765 CPU hours (~115 CPU years) were used for this study.
 
 We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig.
-3A and Supplementary Table 2).
-The conclusions of five studies strongly favor Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1).
-Three studies are based entirely on previously published data, and the remainder add data for one or more species.
+3A and Supplementary Table 1).
+The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Extended Data Table 1).
+Three studies are based entirely on previously published data, and the remainder add data for one or more species.s
 
@@
 
-Our summary of previous phylogenetic analyses (Fig.
+OOur summary of previous phylogenetic analyses (Fig.
 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
 This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination.
-The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT.
+The only analyses that support Porifera-sister have reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT.
 Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
@@
 This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
-We then inferred support under the best-fit model (Supplementary Table 1), except for the three largest matrices where we used LG+C60.
-We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 1).
+We then inferred support under the best-fit model (Supplementary Table 3), except for the three largest matrices where we used LG+C60.
+We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 3).
 
@@
 No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
-3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
+3B lower right; Supplementary Table 2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
 3A).
@@
 We therefore varied the number of categories in Phylobayes analyses (Fig.
-4; Supplementary Table 6).
+4; Supplementary Table 5).
 We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
@@
 The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories.
-ModelFinder selects site heterogeneous C60 models (Supplementary Table 5), but IQtree often gives a warning under C60 that the model may be overfit, with too many categories, because some mixture weights are close to 0.
+ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6).
 While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data.
-Cross Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes].
-It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister reseults obtained in unconstrained CAT analyses.
+Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes].
+It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses.
 To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes.
-We found nearly identical distributions of cross validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
-6).
+We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
+5).
 A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$).
-This indicates that 60 categories in a Poisson+nCAT60 model are as as adequate at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model.
+This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model.
 This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.
@@
 We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
-5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, repectively.
+6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig.
-5A-B).
-The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on predictive power of these far more complex models.
+6A-B).
+The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.
 
@@
 **Fig.
-4.**  Sensitivity analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
+4.**  Sensitivity analyses with representative data matrices were analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes.
 Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values from the sensitive analyses are shown in Supplementary Table 6.
+Support values from the sensitivity analyses are shown in Supplementary Table 5.
 
@@
 7).
-By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrixin each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis.
+By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis.
 Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models.
@@
 We extracted each connected component from this network.
-We further split these components if the the most connected node (i.e.
+We further split these components if the most connected node (i.e.
 most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
@@
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
-Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
+Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
 To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, Bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
 For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
@@
 Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad.
+This work was conducted in part using the resources of the Advanced Computing Center for Research and Education (ACCRE) at Vanderbilt University.
+Research in A.R.'s lab is supported by the National Science Foundation (DEB-1442113), the National Institutes of Health / National Institute of Allergy and Infectious Diseases (1R56AI146096-01A1), the Guggenheim Foundation, and the Burroughs Wellcome Fund.
 
@@
 supervised the project.
-Y.L., C.W.D., X-X.S.
-and B.E.
+Y.L., B.E., X-X.S., and C.W.D.
 conducted analyses.
-Y.L., C.W.D.
-B.E.
+Y.L., B.E.
+and C.W.D.
 wrote the paper.
@@
 GTR+CAT.
+All trees were visualized and annotated in R package ggtree [@yu2018two].
 
@@
 GTR+CAT.
-
-```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
-
-ribo$matrix =
-  factor(
-    ribo$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-             "Whelan2017_full", "Whelan2017_strict") )
-
-p1 = ribo %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
-
-busco$matrix =
-  factor(
-    busco$matrix,
-    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
-               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
-               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
-               "Whelan2017_full", "Whelan2017_strict") )
-
-busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
-
-p2 = busco %>%
-  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
-  geom_bar(stat = "identity") +
-  theme_classic() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = "Matrix", y = "Number of genes") +
-  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
-
- ggarrange(p1, p2, labels = c("A", "B"))
-
-ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=4)
-
-```
-
-**Extended Data Fig.
-3.** Annotation and representation of BUSCO and ribosomal genes in each data matrix.
-(A).
-The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
-(C).
-The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
+All trees were visualized and annotated in R package ggtree [@yu2018two].
 
@@
 
-ggsave("figures/Extended_Data_Fig4_raw.pdf", width=20, height=12)
+ggsave("figures/Extended_Data_Fig3_raw.pdf", width=20, height=12)
 
@@
 **Extended Data Fig.
-4.** Pairwise overlap between each of the primary matrices considered here.
+3.** Pairwise overlap between each of the primary matrices considered here (related to Supplementary Table 8).
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
@@
 
-```{r n_cat_render, fig.width=6, fig.height=3.5}
-
-edf2_cols = c("#74c2e7", "#410058")
-names(edf2_cols) = c("Poisson+nCAT60", "Poisson+CAT")
-
-# Generate plots
-gg_rank_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  scale_colour_manual(values = edf2_cols) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
-
-gg_rank_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_line(aes(x=rank, y=count, col=model)) +
-  scale_colour_manual(values = edf2_cols) +
-  xlab( "category rank" ) +
-  ylab( "site count" ) +
-  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
-
-gg_mds_phil =
-  pb_summaries %>%
-  filter( matrix=="Philippe2009" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  scale_colour_manual(values = edf2_cols) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-gg_mds_whel =
-  pb_summaries %>%
-  filter( matrix=="Whelan2017_strict" ) %>%
-  ggplot() +
-  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
-  scale_colour_manual(values = edf2_cols) +
-  xlim( -0.5, 0.5) +
-  ylim( -0.6, 0.6) +
-  xlab( "MDS1" ) +
-  ylab( "MDS2" ) +
-  theme(legend.position="right")
-
-ggarrange(
-  gg_rank_phil,
-  gg_rank_whel,
-  ncol  = 2,
-  labels = c("A", "B"),
-  common.legend = TRUE
-)
-
-ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=3.5)
+```{r figure_gene_composition,, fig.height = 4, fig.width = 10}
+
+ribo$matrix =
+  factor(
+    ribo$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+             "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+             "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+             "Whelan2017_full", "Whelan2017_strict") )
+
+p1 = ribo %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual(values = c("grey", "black"), labels = c("Non-ribosomal genes", "Ribosomal genes"))
+
+busco$matrix =
+  factor(
+    busco$matrix,
+    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187", "Nosenko2013_ribo_11057",
+               "Nosenko2013_ribo_14615", "Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108",
+               "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "Simion2017",
+               "Whelan2017_full", "Whelan2017_strict") )
+
+busco$gene_category = factor(busco$gene_category, levels = c("non_busco", "busco"))
+
+p2 = busco %>%
+  ggplot( aes(x = matrix, y = number_of_genes, fill = gene_category) ) +
+  geom_bar(stat = "identity") +
+  theme_classic() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Matrix", y = "Number of genes") +
+  scale_fill_manual( values = c("grey", "black"), labels = c("Non-BUSCO genes", "BUSCO genes") )
+
+ ggarrange(p1, p2, labels = c("A", "B"))
+
+ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=4)
+
 ```
@@
 **Extended Data Fig.
-5** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
-These plots are for singe sample isolated from the end of PhyloBayes chain files.
-(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis.
-The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
-The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
+4.** Annotation and representation of BUSCO and ribosomal genes in each data matrix (related to Supplementary Table 8).
+(A).
+The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions.
+(C).
+The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.
 
@@
 
-ggsave("figures/Extended_Data_Fig6_raw.pdf", width=6, height=7)
-
-```
-
-**Extended Data Fig. 6.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=7)
+
+```
+
+**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+
+```{r n_cat_render, fig.width=6, fig.height=3.5}
+
+edf2_cols = c("#74c2e7", "#410058")
+names(edf2_cols) = c("Poisson+nCAT60", "Poisson+CAT")
+
+# Generate plots
+gg_rank_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  scale_colour_manual(values = edf2_cols) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
+
+gg_rank_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_line(aes(x=rank, y=count, col=model)) +
+  scale_colour_manual(values = edf2_cols) +
+  xlab( "category rank" ) +
+  ylab( "site count" ) +
+  theme(legend.position="right")+ ggtitle("Whelan2017_strict")
+
+gg_mds_phil =
+  pb_summaries %>%
+  filter( matrix=="Philippe2009" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  scale_colour_manual(values = edf2_cols) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+gg_mds_whel =
+  pb_summaries %>%
+  filter( matrix=="Whelan2017_strict" ) %>%
+  ggplot() +
+  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
+  scale_colour_manual(values = edf2_cols) +
+  xlim( -0.5, 0.5) +
+  ylim( -0.6, 0.6) +
+  xlab( "MDS1" ) +
+  ylab( "MDS2" ) +
+  theme(legend.position="right")
+
+ggarrange(
+  gg_rank_phil,
+  gg_rank_whel,
+  ncol  = 2,
+  labels = c("A", "B"),
+  common.legend = TRUE
+)
+
+ggsave("figures/Extended_Data_Fig6_raw.pdf", width=6, height=3.5)
+```
+
+**Extended Data Fig.
+6** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+These plots are for singe sample isolated from the end of PhyloBayes chain files.
+(A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis.
+The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data.
+The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.
 
@@
 
-## Supplementary tables
-
-**Supplementary Table 1.** The models selected by ModelFinder for each matrix.
-```{r table_modelfinder}
-analyses_new %>%
-  filter(modelfinder == TRUE) %>%
-  select( matrix, clade, result, model_summary ) %>%
-  kable()
-
-```
-
 ### S.1 Summaries of published analyses
@@
 
-Supplementary Table 1.
-Summary of annotations (BUSCO, Ensemble, GO-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
-
-Supplementary_Table 2. Summary of a total of 164 phylogenomic analyses were transcribed from the literature (Table is converted from analyses_published in Rdata).
-
-Supplementary_Table 3. Summary of a total of 106 phylogenomic analyses conducted in this study (Table is converted from analyses_new in R data).
-
-Supplementary_Table 4. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
-
-Supplementary_Table 5. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
-
-Supplementary_Table 6. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices (Table is converted from analyses_sensitive in Rdata).
-
-Supplementary_Table 7.
-Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
+**Supplementary Table 1.** Summary of a total of 164 phylogenomic analyses were transcribed from the literature (Table is converted from analyses_published in Rdata).
+
+**Supplementary Table 2.** Summary of a total of 106 phylogenomic analyses conducted in this study (Table is converted from analyses_new in R data).
+
+**Supplementary Table 3.** The models selected by ModelFinder for each matrix in IQ-TREE.(Table is converted from analyses_new in Rdata).
+
+**Supplementary Table 4.** Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
+
+**Supplementary Table 5.** Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices (Table is converted from analyses_sensitive in Rdata).
+
+**Supplementary Table 6.** Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
+
+**Supplementary Table 7.** Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
+
+**Supplementary Table 8.** Summary of annotations (BUSCO, Ensemble, GO-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
 

### 2020-10-27 manuscript/manuscript.rmd
fix table1 wideth when converted to PDF, also fix several typos

@@
   library( ape )
+  library(kableExtra)
 
@@
 The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though.
-This suggests that we shouldn't priviledge these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.
+This suggests that we shouldn't privilege these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.
 
@@
 For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
-The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
+The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
 The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
@@
 
-**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
+**Extended Data Table 1.** Overview of data matrices used in this study.
 ```{r table_study_summary}
@@
 opts <- options(knitr.kable.NA = "")
-knitr::kable(table_study_summary)
+knitr::kable(table_study_summary, digits = 3, format.args = list(big.mark = ",",
+  scientific = FALSE)) %>%
+  kable_styling(latex_options="scale_down")
 

### 2020-10-27 manuscript/manuscript_kernel.R

@@ -409,11 +409,16 @@ fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
 pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
 
 ## Write the results from RData to prepare them for supplementary tables
-write_csv(partition_map_global, "./Supplementary_tables/Supplementary_Table_1.csv", na = "NA", quote_escape = "double")
-write_csv(analyses_published, "./Supplementary_tables/Supplementary_Table_2.csv", na = "NA", quote_escape = "double")
-write_csv(analyses_new, "./Supplementary_tables/Supplementary_Table_3.csv", na = "NA", quote_escape = "double")
-write_csv(analyses_sensitive, "./Supplementary_tables/Supplementary_Table_6.csv", na = "NA", quote_escape = "double")
+write_csv(partition_map_global, "./Supplementary_tables/Supplementary_Table_8.csv", na = "NA", quote_escape = "double")
+write_csv(analyses_published, "./Supplementary_tables/Supplementary_Table_1.csv", na = "NA", quote_escape = "double") 
+write_csv(analyses_new, "./Supplementary_tables/Supplementary_Table_2.csv", na = "NA", quote_escape = "double")
+write_csv(analyses_sensitive, "./Supplementary_tables/Supplementary_Table_5.csv", na = "NA", quote_escape = "double")
 write_csv(au_tests, "./Supplementary_tables/Supplementary_Table_7.csv", na = "NA", quote_escape = "double")
+modelfinder = analyses_new %>%
+  filter(modelfinder == TRUE) %>%
+  select( matrix, clade, result, model_summary)
+write_csv(modelfinder, "./Supplementary_tables/Supplementary_Table_3.csv", na = "NA", quote_escape = "double")
+
 
 ## Move all read functions to kernel so all vairables are stored in Rdata
 

### 2020-11-03 manuscript/manuscript.rmd
fix Fig.2 and update comments from Nicole

@@
 Testing these hypotheses requires evaluating the support for each of these alternative nodes.
-(C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.
-(D) A ctenophore, commonly known as comb jellies.
+(C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanozoa, Holozoa, and Opisthokonta.
+(D) A ctenophore, commonly known as a comb jelliy.
 (E) A poriferan, commonly known as a sponge.
@@
 Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanozoa (animals plus most closely related Choanoflagellata), Holozoa (Choanozoa plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).
 
@@
 
-OOur summary of previous phylogenetic analyses (Fig.
+Our summary of previous phylogenetic analyses (Fig.
 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT.
 This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination.
-The only analyses that support Porifera-sister have reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT.
+The only analyses that support Porifera-sister have reduced outgroup sampling (Choanozoa, Holozoa) and site-heterogeneous models with CAT.
 Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge.
@@
 For each of the published studies, we selected the matrix that was the primary focus of the manuscript, or has been reanalyzed extensively in other studies, for further analysis.
-For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling.
+For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanozoa versions, where permitted by original outgroup sampling.
 This produced `r analyses_new$matrix%>% unique()%>% length()` data matrices from 11 studies that presented new sequence data and for which partition data were available.
@@
 For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model.
-Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling.
+Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanozoa sampling.
 We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model.
@@
 1; Philippe2009: Extended Data Fig.
-2): when outgroup sampling is restricted (Choanimalia), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
+2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
@@
 The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects.
-If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
+If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanozoa taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.
 
@@
 It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses.
-To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes.
+To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa matrices, we conducted Bayesian cross-validation experiments with PhyloBayes.
 We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
 5).
-A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$).
+A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$).
 This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model.
-This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.
+This suggests there is no reason to favor the narrow analysis conditions under which Porifera-sister are obtained.
 
@@
 Different categories can have very different numbers of sites.
-We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig.
+We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig.
 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively.
-The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
+The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
 This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig.
@@
 7).
-For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).
+For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Supplementary Table 7).
 
@@
 #### Outgroup taxa sampling with C60 and CAT models
-Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta.
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanozoa, Holozoa and Opisthokonta.
 These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
-For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
+Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
+For each Choanozoa data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
 The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
 Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+To minimize computational burden, GTR+CAT models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 
@@
 To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanimalia and the full matrix).
+We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanozoa and the full matrix).
 For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
@@
 #### Sensitivity analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanimalia (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanozoa (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
 All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.
 
-To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
+To compare the allocation of frequency categories across sites in the Philippe2009_Choanozoa and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
 The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
@@
 #### Cross-validation analyses
-Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
 Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score.
@@
 **Extended Data Fig.
-2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix.
+2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanozoa data matrix.
 (A).
@@
 
-**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia data matrices.
+**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
 
@@
   ylab( "site count" ) +
-  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
+  theme(legend.position="right") + ggtitle("Philippe2009_Choanozoa")
 
@@
 **Extended Data Fig.
-6** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+6** The allocation of frequency categories across sites in the Philippe2009_Choanozoa matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
 These plots are for singe sample isolated from the end of PhyloBayes chain files.
@@
 **Extended Data Fig.
-7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices).
+7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
@@
 They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
-They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
+They considered two different sets of outgroups: Choanoflagellata (resulting in Choanozoa) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
 
@@
 All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors.
-Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
+Analyses with Choanozoa sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS).
 With broader Opisthokonta sampling, support for Porifera-sister is 84% PP.
@@
 The subsequent filtering strategy was also similar to the previous study.
-All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in Choanimalia dataset.
+All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in Choanozoa dataset.
 The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

### 2020-11-10 manuscript/manuscript.rmd
change Choanoflagellata to Choanoflagellatea

@@
 Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic].
-Three progressively more inclusive clades have often been investigated: Choanozoa (animals plus most closely related Choanoflagellata), Holozoa (Choanozoa plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).
+Three progressively more inclusive clades have often been investigated: Choanozoa (animals plus most closely related Choanoflagellatea), Holozoa (Choanozoa plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).
 
@@
 3B).
-No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig.
+No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellatea and use models with site-heterogeneous equilibrium frequencies (Fig.
 3B lower right; Supplementary Table 2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
@@
 These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
+Choanozoa only includes Choanoflagellatea outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
 For each Choanozoa data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
@@
 They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned.
-They considered two different sets of outgroups: Choanoflagellata (resulting in Choanozoa) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
+They considered two different sets of outgroups: Choanoflagellatea (resulting in Choanozoa) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).
 

### 2020-11-12 manuscript/manuscript.rmd
update acknowledgement
Updated model description

@@
 
-The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories.
-ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6).
-While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data.
-Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes].
-It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses.
-To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa matrices, we conducted Bayesian cross-validation experiments with PhyloBayes.
-We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
+The first questions regard model fit.
+ModelFinder selects site heterogeneous C60 models according to BIC (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6).
+In PhylobBayes, Cross-validation is a reliable and suggested approach to model evaluation [@lartillot2009phylobayes].
+We elvaluated  Poisson+CAT and Poisson+nCAT60 models with cross-validation in PhyloBayes for the Whelan2017_strict and Philippe2009_Choanozoa matrices.
+For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
 5).
 A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$).
-This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model.
-This suggests there is no reason to favor the narrow analysis conditions under which Porifera-sister are obtained.
+Cross-validation therefore does not indicate that the unconstrained CAT models are a better fit than the models with 60 categories.
+This suggests there is not a reason to favor the narrow analysis conditions (Choanozoa taxon sampling, Poisson exchange matrix, and unconstrained CAT models that use hundreds of categories) under which Porifera-sister are obtained.
 
@@
 6A-B).
-The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.
+The fact that so many categories apply to such a small fraction of data may help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.
 
@@
 4).
-The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though.
+The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model fit according to cross-validation, though.
 This suggests that we shouldn't privilege these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.
@@
 More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria.
-It will be critical for this to change in coming years.
+Future analyses focused on complete genomes rather than transcriptomes and partial genomes will have multiple advantages.
+Data matrices derived from these more complete sources will have a lower fraction of missing sequences.
+Complete gene sampling within each species will also greatly improve analyses of gene duplciation and loss, a crititical step in building phylogenomic matrices such as those presented here.
+Analyses of this new generation of matrices derived from complete genomes will be well served my understandind the sources of analysis variation in the generation of matrices that came before them.
 
@@
 We thank members of the Dunn and Rokas laboratories for discussions and comments.
-We are grateful to Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript.
+We are grateful to Steve Haddock, Kyle David, and Nathan Whelan for feedback on an earlier version of this manuscript.
+We are grateful to Nicole King, Daniel Richter, Paul Lewis, Warren Francis and Joseph Ryan for feedback on the preprint version of this manuscript.
 We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.

### 2020-11-22 manuscript/manuscript.rmd
suceefully rerun the manuscript and update Fig2 taxon name
updated manuscript format

@@
 output:
-  html_document: default
   pdf_document:
     latex_engine: xelatex
+  html_document: default
   word_document: default
@@
 
-## Summary
+## Abstract
 
@@
 Since then, many more phylogenomic studies have been published (Fig.
-2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1).
+2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Table 1, Supplementary Information section S1).
 The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig.
@@
 For one, the nodes in question are the deepest in the animal tree of life.
-Another factor that has been invoked is branch lengths (*e.g.* Extended Data Figs.
-1, 2), which are impacted by both divergence times and shifts in rate of evolution.
+Another factor that has been invoked is branch lengths (*e.g.* Figs.
+S1, S2), which are impacted by both divergence times and shifts in rate of evolution.
 Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages.
@@
 Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig.
-2, Extended Data Table 1).
+2, Table 1).
 This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations.
@@
 
-High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1).
+High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Table 1).
 Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices.
 As a result, the particular genes selected for analysis vary widely (Fig.
-2, Extended Data Fig.
-3).
-Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig.
-4).
+2, Fig.
+S3).
+Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Fig.
+S4).
 
@@
 Ingroup taxon sampling also varies widely between studies (Fig.
-2, Extended Data Fig.
-3).
+2, Fig.
+S3).
 Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria.
-Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig.
-3).
-Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig.
-3).
+Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Fig.
+S3).
+Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Fig.
+S3).
 This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.
@@
 The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis.
-(A) Analyses transcribed from the literature (related to Supplementary Table 1).
-(B) New phylogenomic analyses conducted in this study (related to Supplementary Table 2).
+(A) Analyses transcribed from the literature (related to Table S1).
+(B) New phylogenomic analyses conducted in this study (related to Table S2).
 Over 1,011,765 CPU hours (~115 CPU years) were used for this study.
@@
 We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig.
-3A and Supplementary Table 1).
-The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Extended Data Table 1).
+3A and Table S1).
+The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Table 1).
 Three studies are based entirely on previously published data, and the remainder add data for one or more species.s
@@
 We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig.
-3B; Supplementary Table 3).
+3B; Table S3).
 Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.
@@
 This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved].
-We then inferred support under the best-fit model (Supplementary Table 3), except for the three largest matrices where we used LG+C60.
-We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 3).
+We then inferred support under the best-fit model (Table S3), except for the three largest matrices where we used LG+C60.
+We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Table S3).
 
@@
 No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellatea and use models with site-heterogeneous equilibrium frequencies (Fig.
-3B lower right; Supplementary Table 2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
+3B lower right; Table S2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig.
 3A).
@@
 Differences in analysis results could therefore be due to differences in the number of categories.
-The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices.
+The number of categories inferred by CAT in PhyloBayes can be very high (Table S4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices.
 This requires a very large number of additional estiamted parameters.
@@
 We therefore varied the number of categories in Phylobayes analyses (Fig.
-4; Supplementary Table 5).
+4; Table S5).
 We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister.
@@
 
-These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig.
-1; Philippe2009: Extended Data Fig.
-2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
+These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Fig.
+S1; Philippe2009: Fig.
+S2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories).
 Analyses under other conditions either support Ctenophora-sister or are unresolved.
@@
 The first questions regard model fit.
-ModelFinder selects site heterogeneous C60 models according to BIC (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6).
+ModelFinder selects site heterogeneous C60 models according to BIC (Table S3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Table S6).
 In PhylobBayes, Cross-validation is a reliable and suggested approach to model evaluation [@lartillot2009phylobayes].
 We elvaluated  Poisson+CAT and Poisson+nCAT60 models with cross-validation in PhyloBayes for the Whelan2017_strict and Philippe2009_Choanozoa matrices.
-For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig.
-5).
+For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Fig.
+S5).
 A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$).
@@
 Different categories can have very different numbers of sites.
-We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig.
-6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively.
+We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Fig.
+S 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively.
 The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix.
-This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig.
-6A-B).
+This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Fig.
+S6A-B).
 The fact that so many categories apply to such a small fraction of data may help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.
@@
 Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively.
-Support values from the sensitivity analyses are shown in Supplementary Table 5.
+Support values from the sensitivity analyses are shown in Table S5.
 
@@
 
-To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig.
-7).
+To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Fig.
+S7).
 By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis.
@@
 Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models.
-In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig.
-7; Supplementary Table 7).
-In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig.
-7).
-For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Supplementary Table 7).
+In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Fig.
+S7; Table S7).
+In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Fig.
+S7).
+For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Table S7).
 
@@
 Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew].
-However, our own analyses (Supplementary Information section S3, Supplementary Figs.
-8-9) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
+However, our own analyses (Supplementary Information section S3, Figs.
+S8-S9) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition.
 Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.
@@
 
-## Methods
+## Supplementary materials
+
+## Materials and Methods
 
@@
 
-We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository.
+We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
 We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
@@
 #### Phylogenetic analyses in IQ-TREE
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1).
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
 Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
@@
 The 20 state matrices are the same across all analyses since they are not recoded.
-Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig.
-9) gives insight into the technical variance of the inference process.
+Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Fig.
+S9) gives insight into the technical variance of the inference process.
 
@@
 
-The statistics presented in  Extended Data Fig. 8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
+The statistics presented in  Fig. S8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 We thank members of the Dunn and Rokas laboratories for discussions and comments.
-We are grateful to Steve Haddock, Kyle David, and Nathan Whelan for feedback on an earlier version of this manuscript.
+We are grateful to Steve Haddock, Kyle David, Stephen Haddock and Nathan Whelan for feedback on an earlier version of this manuscript.
 We are grateful to Nicole King, Daniel Richter, Paul Lewis, Warren Francis and Joseph Ryan for feedback on the preprint version of this manuscript.
@@
 
-## Extended Tables
-
-**Extended Data Table 1.** Overview of data matrices used in this study.
+## Tables
+
+**Table 1.** Overview of data matrices used in this study.
 ```{r table_study_summary}
@@
 
-## Extended Figures
+## Supplementary Figures
 
@@
 ![](figures/Extended_Data_Fig1_manuscript.png)
-**Extended Data Fig.
-1.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix.
+**Fig.
+S1.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix.
 (A).
@@
 ![](figures/Extended_Data_Fig2_manuscript.png)
-**Extended Data Fig.
-2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanozoa data matrix.
+**Fig.
+S2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanozoa data matrix.
 (A).
@@
 
-**Extended Data Fig.
-3.** Pairwise overlap between each of the primary matrices considered here (related to Supplementary Table 8).
+**Fig.
+S3.** Pairwise overlap between each of the primary matrices considered here (related to Table S8).
 Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.
@@
 
-**Extended Data Fig.
-4.** Annotation and representation of BUSCO and ribosomal genes in each data matrix (related to Supplementary Table 8).
+**Fig.
+S4.** Annotation and representation of BUSCO and ribosomal genes in each data matrix (related to Table S8).
 (A).
@@
 
-**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
+**Fig. S5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
 
@@
 
-**Extended Data Fig.
-6** The allocation of frequency categories across sites in the Philippe2009_Choanozoa matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
+**Fig.
+S6** The allocation of frequency categories across sites in the Philippe2009_Choanozoa matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color).
 These plots are for singe sample isolated from the end of PhyloBayes chain files.
@@
 
-**Extended Data Fig.
-7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
+**Fig.
+S7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices).
 The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green).
@@
 
-**Extended Data Fig.
-8.** Graphical representations of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3 and the subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
+**Fig.
+S8.** Graphical representations of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3 and the subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions.
 (A).
@@
 Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister.
-The 20-state and 6-state points on the plots in Extended Data Fig.
-8A correspond to the 20 and SR6 Whelan cells shown here.
+The 20-state and 6-state points on the plots in Fig.
+S8A correspond to the 20 and SR6 Whelan cells shown here.
 The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy).
@@
 ![](figures/Extended_Data_Fig9.png)
-**Extended Data Fig.
-9.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
+**Fig.
+S9.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses.
 These five analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03).
@@
 Feuda *et al.* [@feuda2017improved] report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices.
-Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Extended Data Fig.
-8A).
+Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister) (Fig.
+S8A).
 Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition.
@@
 Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes.
-This provides 32 analyses that they report in their Table 3 and that we present graphically here as Extended Data Fig.
-8A.
-There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Extended Data Fig.
-8A).
+This provides 32 analyses that they report in their Table 3 and that we present graphically here as Fig.
+S8A.
+There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Fig.
+S8A).
 Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to].
@@
 Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister).
-This leaves eight GTR+CAT analyses (Extended Data Fig.
-8A).
+This leaves eight GTR+CAT analyses (Fig.
+S8A).
 Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister.
@@
 Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the GTR+CAT analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix.
-Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig.
-8B).
-The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig.
-8B), indicating that it did not converge as well as other analyses.
+Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Fig.
+S8B).
+The only statistic that stands out for this one analysis is that it has the highest maxdiff (Fig.
+S8B), indicating that it did not converge as well as other analyses.
 
@@
 When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding.
-Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig.
-9).
+Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig.
+S9).
 
@@
 
-**Supplementary Table 1.** Summary of a total of 164 phylogenomic analyses were transcribed from the literature (Table is converted from analyses_published in Rdata).
-
-**Supplementary Table 2.** Summary of a total of 106 phylogenomic analyses conducted in this study (Table is converted from analyses_new in R data).
-
-**Supplementary Table 3.** The models selected by ModelFinder for each matrix in IQ-TREE.(Table is converted from analyses_new in Rdata).
-
-**Supplementary Table 4.** Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
-
-**Supplementary Table 5.** Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices (Table is converted from analyses_sensitive in Rdata).
-
-**Supplementary Table 6.** Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
-
-**Supplementary Table 7.** Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
-
-**Supplementary Table 8.** Summary of annotations (BUSCO, Ensemble, GO-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
+**Table S1.** Summary of a total of 164 phylogenomic analyses were transcribed from the literature (Table is converted from analyses_published in Rdata).
+
+**Table S2.** Summary of a total of 106 phylogenomic analyses conducted in this study (Table is converted from analyses_new in R data).
+
+**Table S3.** The models selected by ModelFinder for each matrix in IQ-TREE.(Table is converted from analyses_new in Rdata).
+
+**Table S4.** Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).
+
+**Table S5.** Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices (Table is converted from analyses_sensitive in Rdata).
+
+**Table S6.** Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).
+
+**Table S7.** Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
+
+**Table S8.** Summary of annotations (BUSCO, Ensemble, GO-terms, ribosomal protein genes) and network analyses for each partition from all phylogenomic all matrices used in this study (Table is converted from partition_map_globle in Rdata).
 

### 2020-11-22 manuscript/manuscript.rmd
more on formats

@@
 bibliography: manuscript.bib
-csl: nature.csl
+csl: science.csl
 header-includes:
@@
 
+**Short title:** Rooting the animal tree of life
+
+**One sentence Summary**: We explore the root of animal tree by synthesizing past work, perform new phylogenomic and sensitivity analyses, and found that the primary source of variation between results is the number of categories used in site-heterogeneous models.
+
 ## Abstract
@@
 
-## Main
+## Main Text
 
@@
 
-## Supplementary materials
-
-## Materials and Methods
-
-### Data and code availability
-
-The main data and results associated with the main text and supplementary materials are available in the GitHub data repository at https://github.com/dunnlab/animal_tree_root.
-All tree files, intermediate results and scripts/commands associated with this study are available in the Figshare data repository at https://doi.org/10.6084/m9.figshare.13085081.v1.
-
-### Data selection and wrangling
-
-We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
-We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
-All changes made are tracked with git.
-
-### Matrix comparison and annotation
-
-#### Taxon name reconciliation
-
-We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
-We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript.
-For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
-
-#### Sequence comparisons
-
-Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
-Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
-
-We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
-We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
-We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
-We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
-
-#### Partition network
-
-We used the sequence similarity comparisons described above to compare partitions.
-
-We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
-We extracted each connected component from this network.
-We further split these components if the most connected node (i.e.
-most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
-We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
-See partition_network_summary in Rdata for a summary tally of each part of the comparison.
-
-### Phylogenetic analyses
-
-#### Phylogenetic analyses in IQ-TREE
-To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
-Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
-site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
-We included models that are commonly used in previous analyses, including site-homogeneous Poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
-For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
-For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
-
-We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, Poisson+C60 and associated best-fit model identified above.
-Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
-Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
-To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
-For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
-2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included the most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
-It should be noted that the main conclusion of Simion *et al*.
-was also based on selection of 25% of genes for their jackknife approach.
-
-#### Outgroup taxa sampling with C60 and CAT models
-Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanozoa, Holozoa and Opisthokonta.
-These datasets include the same set of genes but differ in the composition of outgroup species.
-Choanozoa only includes Choanoflagellatea outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
-For each Choanozoa data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
-The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
-Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
-To minimize computational burden, GTR+CAT models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-
-For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, Bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
-For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
-Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
-Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
-A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
-Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time.
-For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
-We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
-
-#### Phylogenetic distribution of support
-To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
-We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanozoa and the full matrix).
-For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
-The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
-The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
-By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].
-
-For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe2009 and Whelan2017 matrices due to computational efficiency.
-Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
-For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
-Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].
-
-#### Sensitivity analyses with different number of substitutional categories
-To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanozoa (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60).
-To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
-All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.
-
-To compare the allocation of frequency categories across sites in the Philippe2009_Choanozoa and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
-The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
-
-#### Cross-validation analyses
-Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
-Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score.
-For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.
-
-#### Performance of data-recoding
-All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017.
-The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.
-
-The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
-This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
-The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
-
-The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
-To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
-This creates new recodings that have the same sized groups as SR-6.
-The new recodings were, from `random-00` to `random-03`:
-
-    MPKE AY IDGQ TRC WNLF SVH
-    EIFT WL QVPG NKM SCHA RYD
-    LCQS GK WPTI VRD YEFN MAH
-    IWQV TY KDLM ESH APCF GRN
-
-To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
-
-The 20 state matrices are the same across all analyses since they are not recoded.
-Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Fig.
-S9) gives insight into the technical variance of the inference process.
-
-Each new matrix was analyzed with PhyloBayes-MPI.
-Analyses were run for 1000 generations, and a 200 generation burnin applied.
-The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
-
-The statistics presented in  Fig. S8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
-
 ## Acknowledgements
@@
 All authors discussed the results and implications and commented on the manuscript at all stages.
+
+## List of Supplementary materials
+Materials and methods
+Table S1 - S8
+Fig. S1 - S9
+Supplementary text S1 - S3
+
+## Supplementary Materials
+
+## Materials and Methods
+
+### Data and code availability
+
+The main data and results associated with the main text and supplementary materials are available in the GitHub data repository at https://github.com/dunnlab/animal_tree_root.
+All tree files, intermediate results and scripts/commands associated with this study are available in the Figshare data repository at https://doi.org/10.6084/m9.figshare.13085081.v1.
+
+### Data selection and wrangling
+
+We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository.
+We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks.
+All changes made are tracked with git.
+
+### Matrix comparison and annotation
+
+#### Taxon name reconciliation
+
+We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix.
+We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript.
+For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.
+
+#### Sequence comparisons
+
+Using the original partition files for each matrix, we separated each sequence for each taxon from each partition.
+Because many of the matrices had been processed by the original authors to remove columns that are poorly sampled or highly variable, these matrix-derived sequences can have deletions relative to the actual gene sequences.
+
+We used DIAMOND v0.9.26 [@Buchfink:2014] to compare each sequence to all others using default diamond Blastp parameters.
+We further filtered DIAMOND results such that we retained hits for 90% of partitions (pident > 50.0, eValue < 1e-5, no self vs self).
+We ran BUSCO with default parameters for all sequences against the provided Metazoa gene set.
+We also ran a BLAST+ v2.8.1 [@Camacho:2009] blastp search against the SwissProt [@boeckmann:2003swiss] database, filtering results such that we retain at least one hit for ~97% of partitions (pident > 50.0, eValue < 1e-15).
+
+#### Partition network
+
+We used the sequence similarity comparisons described above to compare partitions.
+
+We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions.
+We extracted each connected component from this network.
+We further split these components if the most connected node (i.e.
+most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components.
+We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes.
+See partition_network_summary in Rdata for a summary tally of each part of the comparison.
+
+### Phylogenetic analyses
+
+#### Phylogenetic analyses in IQ-TREE
+To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1).
+Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs.
+site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option).
+We included models that are commonly used in previous analyses, including site-homogeneous Poisson, WAG, LG, GTR models plus C10-C60 models in the model testing.
+For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters.
+For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.
+
+We then reanalyzed each matrix under a panel of evolutionary models, including WAG, GTR, Poisson+C60 and associated best-fit model identified above.
+Nodal support was assessed with 1000 ultrafast bootstrap replicates for each analysis.
+Because of the large size of Hejnol2009 and Simion2017, it was not computationally feasible to analyze the whole matrix using the C60 model or CAT site-heterogeneous models.
+To circumvent his limitation, we reduced the data size from their full matrices to facilitate computational efficiency for site-heterogeneous models.
+For Hejnol2009 matrix, we instead used the 330-gene matrix constructed by Hejnol *et al*.
+2009, since the main conclusion for their study is based on this subsampled matrix; For Simion2017 matrix, we only included the most complete 25% of genes (genes that were present in less than 79 taxa were removed; 428 genes were kept).
+It should be noted that the main conclusion of Simion *et al*.
+was also based on selection of 25% of genes for their jackknife approach.
+
+#### Outgroup taxa sampling with C60 and CAT models
+Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanozoa, Holozoa and Opisthokonta.
+These datasets include the same set of genes but differ in the composition of outgroup species.
+Choanozoa only includes Choanoflagellatea outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi.
+For each Choanozoa data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted.
+The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE.
+Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8.
+To minimize computational burden, GTR+CAT models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+
+For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, Bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above.
+For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation.
+Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in.
+Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp).
+A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability.
+Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time.
+For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks.
+We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
+
+#### Phylogenetic distribution of support
+To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices.
+We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanozoa and the full matrix).
+For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model.
+The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals.
+The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious].
+By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].
+
+For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe2009 and Whelan2017 matrices due to computational efficiency.
+Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl).
+For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300).
+Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].
+
+#### Sensitivity analyses with different number of substitutional categories
+To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanozoa (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60).
+To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.
+All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.
+
+To compare the allocation of frequency categories across sites in the Philippe2009_Choanozoa and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above.
+The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.
+
+#### Cross-validation analyses
+Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanozoa data matrices.
+Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score.
+For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.
+
+#### Performance of data-recoding
+All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017.
+The randomized recoding analyses are in the `recoding/alternative` subdirectory of the git repository.
+
+The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group.
+This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix.
+The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].
+
+The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data.
+To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled.
+This creates new recodings that have the same sized groups as SR-6.
+The new recodings were, from `random-00` to `random-03`:
+
+    MPKE AY IDGQ TRC WNLF SVH
+    EIFT WL QVPG NKM SCHA RYD
+    LCQS GK WPTI VRD YEFN MAH
+    IWQV TY KDLM ESH APCF GRN
+
+To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.
+
+The 20 state matrices are the same across all analyses since they are not recoded.
+Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Fig.
+S9) gives insight into the technical variance of the inference process.
+
+Each new matrix was analyzed with PhyloBayes-MPI.
+Analyses were run for 1000 generations, and a 200 generation burnin applied.
+The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.
+
+The statistics presented in  Fig. S8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.
 
@@
 
-## Supplementary Information
-
-### S.1 Summaries of published analyses
+## Supplementary text
+
+### S1 Summaries of published analyses
 
@@
 
-### S.2 Models of molecular evolution
+### S2 Models of molecular evolution
 
@@
 
-### S.3 Data-recoding methods
+### S3 Data-recoding methods
 

### 2022-08-10 scripts_mine/add_comp_nrs.py
scrips,etc

@@ -11,7 +11,19 @@ from Bio import SeqIO
 from compare_comp_nrs import add_co_nr,process_df
 from glob import glob
 import numpy as np
+# read in blast results, function copied from blast_partitions_graph.py
+def read_blast_results(dirpattern, colnames, pident_cutoff, evalue_cutoff):
+    df = None
+    for i, infile in enumerate(sorted(glob(dirpattern))):
+        tmp_df = pd.read_csv(infile, sep='\t', names=colnames)
 
+        tmp_df = tmp_df[((tmp_df['pident']>pident_cutoff) & 
+                         (tmp_df['evalue']<evalue_cutoff))]
+        if i ==0:
+            df = tmp_df
+        else:
+            df = pd.concat((df, tmp_df),ignore_index=True)
+    return df
 #add component number
 def edit_df(df,partition_df):
     for i,name in zip(df.index.values.astype(int), df.name):
@@ -35,6 +47,28 @@ def read_blast_query(dirpattern,colnames):
 #            df = pd.concat((l, tmp_l),ignore_index=True)
     df = pd.DataFrame({colnames[0]:l1,colnames[1]:l2})
     return df
+
+#Read blast results, add component nrs only for result not for query
+ava_pident_cutoff = 50
+ava_evalue_cutoff = 1e-5
+ava_colnames = ['qseqid', 'sseqid', 'pident', 'length', 
+                'mismatch', 'gapopen', 'qstart', 'qend', 
+                'sstart', 'send', 'evalue', 'bitscore']
+ava_df = read_blast_results('../add_new_data_test/diamond_results/*_permissive.txt',   #read only Dunn and Philippe) 
+                            ava_colnames, 
+                            ava_pident_cutoff, 
+                            ava_evalue_cutoff)
+# filter out self-matches
+ava_df_sm = ava_df[ava_df['qseqid'] != ava_df['sseqid']]
+
+#unify names with partition map
+ava_df_sm = ava_df["qseqid"].str.split(":", n = 3, expand = True)
+ava_df["qseqid"]= ava_df_sm[0]+ava_df_sm[3]
+ava_df_sm = ava_df["sseqid"].str.split(":", n = 3, expand = True)
+ava_df["sseqid"]= ava_df_sm[0]+ava_df_sm[3]
+
+#keep only columns with names
+ava_df = ava_df[['sseqid','qseqid']]
 
 #import fasta headers from blast queries to use as name
 infile = r'..\reconciliation\blast\queries\dunn_philippe\*.fa'

### 2022-08-10 scripts_mine/compare_comp_nrs.py

@@ -48,7 +48,7 @@ def set_co_nr(name,partition_df):
 def add_co_nr(i,df,col_name,col_name2,name,partition_df):
     com_nr = set_co_nr(name,partition_df)
     df.at[i,col_name]=com_nr
-    df.at[i,col_name2]=name
+    #df.at[i,col_name2]=name
     return df
 
 #Add component numbers for both search and query in all rows of dataframe, input df must have columns with names qreqid and sseqid
@@ -95,7 +95,7 @@ ava_evalue_cutoff = 1e-5
 ava_colnames = ['qseqid', 'sseqid', 'pident', 'length', 
                 'mismatch', 'gapopen', 'qstart', 'qend', 
                 'sstart', 'send', 'evalue', 'bitscore']
-ava_df = read_blast_results('../reconciliation/blast/diamond_results/dunn_philippe/*_permissive.txt',   #read only Dunn and Philippe) 
+ava_df = read_blast_results('../add_new_data_test/diamond_results/*_permissive.txt',   #read only Dunn and Philippe) 
                             ava_colnames, 
                             ava_pident_cutoff, 
                             ava_evalue_cutoff)
@@ -115,7 +115,7 @@ ava_df = ava_df[['sseqid','qseqid']]
 
 #load partition maps
 supp_mat_df = pd.read_excel('../Li_etal_Supplementary_Tables.xlsx', sheet_name="Table S9",)
-partitions_df_expanded_r = pd.read_csv("../manuscript/part_map_glob.csv", sep="\t")
+partitions_df_expanded_r = pd.read_csv("../files_mine/part_map_glob.csv", sep="\t")
 
 #filter out hits with only one match
 #supp_mat_df = supp_mat_df[supp_mat_df["nodes_in_component"] !=1] 

### 2022-08-10 shape_partition.py

@@ -1,21 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Tue Aug  9 14:16:08 2022
-
-@author: matil
-"""
-import pandas as pd
-
-
-partitions = pd.read_csv("dosReisDonoghueYang_2015/dosReis_etal_superalignment.nex", sep="\t")
-partitions['new'] = "CHARSET"
-partitions['new1'] = partitions.protein
-partitions['new2'] = "="
-partitions['new3'] = partitions.start.astype(str)+'-'+partitions.end.astype(str)+';'
-
-partitions_new = partitions[['new','new1','new2','new3']]
-
-partitions_new.loc[0] = ['BEGIN SETS;',None, None, None]
-partitions_new.loc[partitions_new.shape[0]] = ['END SETS;',None, None, None]
-                      
-partitions_new.to_csv("dosReisDonoghueYang_2015/dosReis_etal_superalignment_new.nex",sep="\t",index=False,header=False,)
